### 2013-09-30 JournalToSubmit2013/index.tex
inicial --martin

%% This is file `elsarticle-template-1a-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
%%
\documentclass[preprint,12pt]{elsarticle}
\usepackage{geometry}
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

\journal{Journal of Systems and Software}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
the Presence of Non-Functional Properties}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[inst1]{Pl\'acido A. Souza Neto}
\author[inst2]{Genoveva Vargas-Solar}
\author[inst3]{Martin A. Musicante}
\author[inst4]{Valeria~de~Castro}
\author[inst3]{Umberto Costa}

\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
Brazil}

\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}

\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
Brazil}

\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}

\begin{abstract}
This paper presents\ldots
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%%
%% Start line numbering here if you want
%%
% \linenumbers

%% main text

Service oriented computing is at the origin of an evolution in the field of software development. 
An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}. The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.  In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.

Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.  Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.

The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase. Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS). 
%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.

Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.  
%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it. They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method). 
Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}. Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation. As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).

The remainder of the paper is organized as follows. Section \ref{sec:motivation} gives an overview of our approach. It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify. Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
%
Section \ref{sec:implementation} describes implementation and validation issues.
Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms. Section \ref{sec:conclusions} concludes the paper and discusses future work.

%*********************************************************************************************************
\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
%*********************************************************************************************************
Consider for instance the following scenario. An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts. Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
\begin{itemize}
\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item Facebook and Twitter services export a method for  updating the status of a given user:
\begin{itemize} 
%\item{\sf\small get-Status ( usedid ): String ; 
\item {\sf\small update-Status ( userid, new-status ): String}; 
\end{itemize}
\end{itemize}
%
%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
%the scenario. The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed. 
%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users. The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription). Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).

Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
The "To Publish Music" scenario 
%(showed by doted line) 
starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\begin{figure}
\centering
\includegraphics[width=0.55\textwidth]{figs/SC}
\caption{BPMN model of the "To Publish Music" scenario}
\label{fig:E3valuemodel}
\end{figure}

Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition. The challenge being  ensuring the compliance between the specification and the resulting application. Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Modeling a services' based application}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.  Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach. Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system. 
\begin{figure} [htpb]
\centering
\includegraphics[width=0.65\textwidth]{figs/SODM}
\caption{SOD-M development process}
\label{fig:sodm}
\end{figure} 
As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}. Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) 
and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
%provide
%is an evolving approach to software development that deals with the provision of 
%models, transformations between them and code generators to address software development. 
%One of the main advantages of model-driven approaches is the provision of 
%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.

Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Modeling non-functional constraints of services' based applications}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.

Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse. Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods. Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols. Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented. In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
%\subsection{$\pi$-SOD-M}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

%
%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints. 
%
Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
%This method is described in the following sections.
As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level. Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).  $\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}). Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).

%*********************************************************************************************************
\section{$\pi$ services' composition meta-model}\label{sec:piscm}
%*********************************************************************************************************

The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform. This model is represented by means of a UML activity diagram. Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).

\begin{itemize}
\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions. They are graphically presented as a partition in the activity diagram. A collaborator can be either internal or external to the system being modelled. When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.

\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity. Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled. There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).

\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
\end{itemize}

 \begin{figure}[htpb]
\centering
\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
\label{fig:e-scomposition-metamodel}
\end{figure}

To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
Figure \ref{fig:servicecompositionmodel}). There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}. Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.

Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process. We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types. In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).

The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part. 
%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
An {\em A-policy} groups a set of rules. It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts. An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .

\begin{figure}[htpb]
\centering
\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
\caption{Services' composition model for the business service "To publish music"}
\label{fig:servicecompositionmodel}
\end{figure}

Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space. A call to such services must be part of the authentication protocol required by these services.
In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}). In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
 {\sf\small OAuth}  implements the open authentication protocol.
As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service. This variable type is imported through the library {\sf\small OAuth.Token}. The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}). Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.

{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.  As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. On the notification of an event of that type, a credential is obtained using the username and password values. The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.

Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code. For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.

Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation. The following Section describes the $\pi$-PEWS  meta-model that supports this representation.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
consisting of  classes representing:
\begin{itemize}
\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
 choice ($\ + \ $) among services,
the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).

\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
\end{itemize}
%
\begin{figure}
\centering
\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Metamodel}
\label{fig:metamodel}
\end{figure}

As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).  It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.  Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
\begin{itemize}
\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}. A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules. The value of a variable is represented using the type {\sc Variable}. The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
\end{itemize}

%An authentication {\em A-policy} represents the situation where an invocation in
%an activity occurs until its sender and/or its recipient have been
%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
%

Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules. The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\section{Transformation rules}\label{sec:mmrules}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model. There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
%\noindent

%{\bf\em 
\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}. A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}. When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
\begin{itemize}
\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
 \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
\end{itemize}

In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}. Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
\begin{figure}
\centering{
%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
\label{fig:transformations}
\end{figure}

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
%\noindent

%{\bf\em 
\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model. The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.   The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.

As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}. 
The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.

%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.

%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}. The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.

%
%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario. Thus in both cases the events are of type {\sf ActivityPrepared}. These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}. In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
%\begin{figure}[htpb]
%\centering{
%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
%\label{fig:p-scim}
%\end{figure}

%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
%\begin{figure}
%\centering{
%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
%\caption{Pews program implementing the "To Publish Music" application}
%\label{fig:pewsexpression}
%\end{figure}

%*********************************************************************************************************
\section{Implementation issues}\label{sec:implementation}
%*********************************************************************************************************
This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions. For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application. Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.

%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}. From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{$\pi$-SOD-M Development Environment}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it. The environment implements the abstract architecture shown in Figure \ref{fig:sodm}. Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
\begin{figure}[htpb]
	\begin{center}
		\includegraphics[width=0.60\textwidth]{figs/architecture}
	\end{center}
		\caption{$\pi$-SOD-M Development Environment}
   \label{fig:policymanager}
\end{figure}
\begin{itemize}
\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}. Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).

\item	 We used  ATL \footnote{http://eclipse.org/atl/. An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).

\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in. We coded the pews.mt program  that implements the model to text transformation for generating executable code. It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the 
{\em A-policy} based services' composition execution environment.

%\item Finally, we created a chain execution  to execute the model to text transformation.
\end{itemize}
%
%
%

%
As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.  The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions. The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).

The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
Each activity is seen as the process where the service method call is executed.
The execution of an activity has four states: prepared, started, terminated, and failure.
The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.

At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).  Policies associated to a scope are activated when the execution of its scope starts. A {\em A-policy} will have to be executed only if one or several of its rules is triggered. If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section. 
%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
If rules belonging to several policies are triggered then policies are also ordered according to an execution plan. The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
%The order of policies has implications on the global order of the rules to be executed.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Validation}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

We validated our approach  by implementing the "To Publish Music" application. It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.  Once the services' composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose. Our implementation includes a  code generator that generates executable code for a reliable services' composition.
The model represents an services' composition expression and its associated policies.

We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized. Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available. We also implemented a reference coordination without policies. We dealt with the availability exceptions by hand and the authentication protocols were embedded within the activities code. During our validation Twitter changed the authentication protocol and the maintenance of the application this implied re-programming the reference application. Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.

%*********************************************************************************************************
\section{Related works}\label{sec:related}
%*********************************************************************************************************
Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions. They also include methods and approaches for modeling non-functional constraints.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
%\noindent{\bf\em 
\subsection{Programming non-functional properties for services}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols. WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services. For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction. The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer. As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process. This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. Other works, like \cite{helga2}
introduce a model for transactional services composition based on
an advanced transactional model. \cite{samy} proposes an approach
that consists of a set of algorithms and rules to assist designers
to compose transactional services. In \cite{Vid04} the model
introduced in \cite{schuldt-etal-TODS} is extended to web services
for addressing atomicity.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
%\noindent{\bf\em 
\subsection{Modeling non-functional properties}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
There are few methodologies and approaches that address the explicit modeling  of non functional
properties for service based applications.   Software process methodologies for
building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.

 \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed. A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called. Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
 The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
description language, for defining contracts for services. There are an associated architecture framework, design standards and a methodology, for developing applications using services.  A services' based application  specification is generated after several
 \cite{AbrialLNSS91} B-machines refinements that describe the services and their
compositions.
\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
oriented business process development methodology with phases for business process development. The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}. SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
%specification, realization, implementation and deployment monitoring and management.

 %\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes. Activities in business process can be performed by services so that the model of business process
%represents a composition of services. It classifies services   as public utilities, business services or
%composition services. Software development that uses services involves creating programs for composing and configuring services to create new composite services. The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
%\noindent{\bf\em 
\subsection{Discussion}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence). In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way. Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse. In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.

Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.

%*********************************************************************************************************
\section{Conclusions and future work}\label{sec:conclusions}
%*********************************************************************************************************
This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications. We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.  We extended the SOD-M method, particularly the $\pi$-SCM (services' composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code. We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.

Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services. We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications. Having such business rules expressed and then translated and associated to the services' composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.

Programming non-functional properties is not an easy task, so we are defining a set of predefined {\em A-policy} types with the associated use rules for guiding the programmer when she associates them to a concrete application. {\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \cite{key}          ==>>  [#]
%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
%%   \citet{key}         ==>>  Author [#]

%% References with bibTeX database:

\bibliographystyle{plain}
\bibliography{biblio}

%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use model1a-num-names.bst.

%% References without bibTeX database:

% \begin{thebibliography}{00}

%% \bibitem must have the following form:
%%   \bibitem{key}...
%%

% \bibitem{}

% \end{thebibliography}

\end{document}

%%
%% End of file `elsarticle-template-1a-num.tex'.

### 2013-09-30 RE_Track2013/src/paperRE_tosubmit.tex
inicial --martin

\documentclass{sig-alternate}
  \pdfpagewidth=8.5truein
  \pdfpageheight=11truein

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{SAC'13}{March 18-22, 2013, Coimbra, Portugal.}
\CopyrightYear{2013} % Allows default copyright year (2002) to be over-ridden - IF NEED BE.
\crdata{978-1-4503-1656-9/13/03}  % Allows default copyright data (X-XXXXX-XX-X/XX/XX) to be over-ridden.
% --- End of Author Metadata ---

\title{Non-Functional Requirements for Service-Based Applications : A Systematic Review}

% \subtitle{[Extended Abstract]
% \titlenote{A full version of this paper is available as
% \textit{Author's Guide to Preparing ACM SIG Proceedings Using
% \LaTeX$2_\epsilon$\ and BibTeX} at
% \texttt{www.acm.org/eaddress.htm}}}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in  
% the body of your article BEFORE References or any Appendices.

\numberofauthors{5} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor Valeria de Castro\\
       \affaddr{Universidad Rey Juan Carlos}\\
       \affaddr{M\'{o}stoles, Spain}\\
       \email{Valeria.deCastro@urjc.es}
\alignauthor
       Martin A. Musicante\\
       \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
       \affaddr{Natal-RN, Brazil}\\
       \email{mam@dimap.ufrn.br}
\alignauthor
       Umberto Souza da Costa\\
       \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
       \affaddr{Natal-RN, Brazil}\\
       \email{umberto@dimap.ufrn.br}        
\and
\alignauthor
       Pl\'acido A. Souza Neto\\
	   \affaddr{Federal Institute of Rio Grande do Norte (IFRN)},\\
       \affaddr{Natal-RN, Brazil}\\
       \email{placido.neto@ifrn.edu.br}
\alignauthor Genoveva Vargar-Solar\\
       \affaddr{French Council of Scientific Research}\\
       \affaddr{LIG-LAFMIA }\\
       \affaddr{Saint Martin d'H\`{e}res, France}\\
       \email{Genoveva.Vargas-Solar@imag.fr}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\date{21 September 2013}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
This paper presents a systematic literature review of non-functional
requirements (NFRs) for service-based applications. 
We propose the classification of non-functional requirements into 3 levels:
\textit{business}, \textit{service} and \textit{system}. 
During the development process, the NFRs can be refined from 
one level to a more concrete level (\textit{i.e.}, from business to service
level or from service to system level). Along with the classification, we also
propose a specific nomenclature for this type of modelling. 
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]

%\terms{Delphi theory}

\keywords{Non-functional requirements, Classification, Sevice-based
development, Contraints, Contract, Policy}

\section{Introduction}

Non-functional aspects concerning services and applications semantics are often
expressed as requirements and constraints.
It is not uncommon to these aspects not not to be fully considered during the development of applications.
In many cases, they are considered once the
application has been implemented, in order to ensure some
level of reliability (e.g., data privacy, exception handling,
atomicity, data persistence). 
This situation leads to service-based
applications that are partially specified and that are thereby
partially compliant with their requirements.

In systems and requirements engineering, a non-functional
requirement (NFR) specifies criteria about the behaviour of a
system. 
These criteria may be not related to the results produced by the system,
but to other conditions of its performance. 
Non-functional requirements are often called qualities of a system. 
They are also referred as ``constraints'', ``quality attributes'', ``quality goals'', ``quality of
service requirements'' and ``non-behavioural requirements'' \cite{Stellman2005}. 
In the case of service-based applications, non-functional requirements concern
the application itself as well as its component services.

As services are independent components, ensuring non-functional properties is a
challenge.
Different
studies~\cite{Babamir2010,AgarwalLS09,CholletL09,GutierrezRF10,XiaoCZBOLH08,JeongCL09,TsadimasNA12}
try to associate non-functional requirements to services using different approaches. 
Associating non-functional requirements to services
compositions can help to ensure that the resulting application is compliant to
the user requirements and also with the characteristics of the services it uses.

%  in developing this type of application. To adapt quality properties in
% the web service development context can provide a greater software reuse.

 Attempts to solve these problems have led to the development of a great deal of
 work in non-functional attributes and non-functional requirements. Although
 they are used as synonyms in most NFR approaches \cite{MairizaZN10}, we will
 distinguish the concepts of non-functional requirements and non-functional attributes.
 NFRs will be defined as a group of semantically correlated
 non-functional attributes (NFA). For example, \textit{security} is an
 non-functional requirement that comprises attributes such as
 \textit{confidentiality} and \textit{integrity}. A non-functional
 attribute describes the characteristics of a functional requirement.

In this context, this paper presents \textit{(i)} a systematic review
 of the treatment given to non-functional requirements for web service based applications, and
 \textit{(ii)} proposes a classification of the non-functional requirements for
these kind of applications.

% The goal of this chapter is presents, based the concepts obtained from
% non-functional requirements and software engineering methodologies' works,
% a systematic analysis concerning \textit{(i) a towards NFR
% classification for reliable web service development, (ii) its application on the modeling web services composition and
% (iii) how to apply these set of concepts through a proposal for a reliable web
% service based methodology development}.

The analysis will be conducted in the form of a systematic review
%\footnote{A systematic review provides means of identifying,
% evaluating, and interpreting the literature relevant to a particular research
% question or topic area} 
\cite{Kitchenham08}.
 The result of this review can be useful to summarize the existing treatment of NFRs and to provide a background to new research activities.

%  We present an analysis into 2 parts to facilitate the understanding of the
%  areas and related works and what their importance in the context of
%  service-based development.  Our goal is to identify a common nomenclature for existing NFR
%  and web service based methodologies and propose a specification of these
%  concepts in the context of reliable web services development. 
%  
%  Since the focus of this work is the development of reliable service-based
%    applications, we analyzed what are, where and how non-functional requirements
%    are modeled in the development of this type of application. Allied to this,
%    we analyzed which methodologies aim the development of service, and if
%    consider modeling of non-functional aspects.

The paper is organized as follow: Section \ref{sec:nfrs} presents a systematic
literature review of non-functional requirements for service-based application.
In Section \ref{sec:classification} we propose a classification of the NFRs
considering the analysis of works made in the systematic review section. Section
\ref{sec:conclusion} concludes the paper.

\section{Non-Functional Requirements\\ for Service-Based Applications}   
\label{sec:nfrs}

This section presents a systematic review of the non-func\-tion\-al
requirements and properties used in the context of service-oriented
development. The purpose of our analysis is:
\begin{itemize}
  \item to identify the concepts, properties and notations used in the
  service-based systems development;
  \item to find if there is any pattern or relationship between non-functional
requirement concepts in different modelling levels.
\end{itemize}

For each phase of the development, the requirements and non-functional
requirements are refined. As a result of this process, the granularity of the
concepts described by the requirement becomes thinner and more precise.

% The focus of a modeling-based non-functional requirements becomes important
% because it reveals a continuing need on the development of distributed
% service-oriented. Thus, it is necessary identify how non-functional requirements
% are classified in different development process phases and iterations.

% Thus, this section also presents the collected data and results
% from the research work analysis according to some research questions. Once the
% sources' extraction execution had been completed, it was important to be sure
% that multiple publications of the same approach were not included in the data analysis. Tables \ref{tab:result02}
% and \ref{tab:result03} show the results of each work under each question
% described in section \ref{subsec:question}.

% % Among the properties presented we highlight the security
% % and performance properties. All users want to access data
% % securely and quickly. Most studies analyzed have both properties. Reliability is
% % also an important non-functional requirement presented in some papers and needed
% % to end users.

% 
% The analisys was carried out by effectuating the following activities:
% \textit{(i) question formulation; (ii) source selection; (iii) selection
% process; (iv) information extraction;} and \textit{(v) results}.

We propose 7 research questions ($RQ_1$ to $RQ_7$) to guide our analysis about non-functional
requirements. 
For each question we define a set of possible answers in order to guide
the analysis.
These possible answers are defined from our knowledge on each topic. 
The questions are closely related to service-oriented development with a NFR focus.
They are:
\begin{itemize} 
  \item \textbf{\texttt{$RQ_1$:}} How are NFR modelled by existing
  methodologies for developing reliable Web services?
  \begin{itemize}
	  \item \textit{Answer is specific to each proposal}
	\end{itemize}  
 \item \textbf{\texttt{$RQ_2$:}} Which are the NFR that are more frequently
 by methodologies developing web services?
 	\begin{itemize}
	  \item \textit{Possible answers: security / availability / portability / \ldots / reliability /
	  performance}
	\end{itemize}
  \item \textbf{\texttt{$RQ_3$:}} What is the software
  development approach used in the paper?
	\begin{itemize}
	  \item Possible answers: \textit{Model driven approach (*MDD) / Ontology (*Ont) / Formal method
	  (*FM) / Artificial intelligence (*AI) / Business Process Modeling (*BP)
	  Traditional (*TDT)}
	\end{itemize} 
  \item \textbf{\texttt{$RQ_4$:}} What is the discipline (application domain)
  of the ``\textit{non-functional requirements}'' / ``\textit{non-functional
  properties}'' used in the work?
  \begin{itemize}
	  \item Possible answers: \textit{Software architecture / QoS model / Language definition /
	  Methodology / etc}
	\end{itemize}	
  \item \textbf{\texttt{$RQ_5$:}} Does the paper proposes a (meta)model
  describing and analyzing NFR? Is there any relationship between
  the non-functional requirements (meta)model proposed and business services? 
\begin{itemize}
	  \item Possible answers: \textit{yes / no} -- \textit{yes / no}
	\end{itemize}  
  \item \textbf{\texttt{$RQ_6$:}} Do the non-functional aspects are treated in
  an independent way?
\begin{itemize}
	  \item Possible answers: \textit{single / composition}
	\end{itemize}
\item \textbf{\texttt{$RQ_7$:}} Which is the publication year of the paper?
	\begin{itemize}
	  \item Possible answers: \textit{Year of publication}
	\end{itemize}	   
\end{itemize}

% \subsection{Source Analysis}
% \label{subsec:souce_analysis}

% In the development of systems, for each phase of development,
% the requirements are being refined and the granularity becomes tiner after each
% iteration. With the analysis, we can identify how these NFRs have been
% classified.

The research questions identify characteristics
 of the ser\-vice-oriented application development, especially the
modelling of non-functional requirements/properties, how they are addressed and
if there is any classification.

\bigskip

We used the approach described in \cite{Kitchenham08} for searching,
collecting and selecting works related with non-functional requirements for
developing service-based applications. Based on these data we analyzed concepts used
in each work for the description of NFRs and the difference among them.

We selected the sources proposed by \cite{Kitchenham08} for searching primary
studies. These sources contain the works published in journals, conferences and
workshops which are of recognized quality within the research community. The
search engines are: \textit{(i) IEEE Computer; (ii) ACM Digital Library;} and
\textit{(iii) Science Direct}. In the selected sources, we used the following search
query criteria:

\begin{center}
\texttt{(((non functional properties) OR (non functional requirements))
AND web service AND composition))}
\end{center}

After define the sources' selection, we apply the
process used to identify those works that provided direct
evidence with regard to the research questions. Deciding
for the inclusion and exclusion criteria for filtering the corpus works
selection, we selected those related to non-functional requirements/properties,
and quality for web service based applications. Initially, the selection
criteria were interpreted liberally and clear exclusions were only made with
regard to title, abstract and introduction.

Based on the guidelines mentioned in \cite{Kitchenham08}, we established a
multi-step process made up of three steps with different selection
criteria:

\begin{itemize}
\item Step 1 - the search string must be run on the selected
search engine. An initial set of studies was obtained by filtering
of title, abstract, and if necessary, introduction. All the studies were
selected according to the inclusion and exclusion criteria. Studies which
were not clearly related to any aspect of the research questions
were excluded.
\item Step 2, the exclusion criteria were based on the following
practical issues: non-English papers, non-
International Conference papers and non-International Workshop
papers. Specifically in the case of ACM library, we considered only the
transaction journal works.
\item Step 3, the papers selection process was based on detailed
research questions.
\end{itemize}

The information for each step was collected considering the 3 searchers
and the the query presented previously. the results of each step were:
\textit{(i)} for each source a list of all the studies that
fulfilled the query; \textit{(ii)} a list
of studies for each source which contained all the works that did not fulfill the
second stage inclusion criteria; and
\textit{(iii)} the last step produced a list of works for each source which
contained all the studies that fulfilled the second step.

The notation used to refer non-functional requirements will be highlighted in
\textit{italic}, as well as the set of values that can be associated with each
concept. For example, a notation used by a work can be \textit{quality property}
or \textit{non-functional concern}, and the values associated with
its notation are \textit{security}, \textit{reliability}, \textit{transaction},
etc. Thus, considering the research question, we analyze a set of works in order
to identify any pattern or divergence between the analyzed literature.

\bigskip

Babamir et al.\cite{Babamir2010} ranks services \textit{quality properties}
in three categories (business level, service level, system level).
\textit{Quality properties} are associated with \textit{quality constraints},
that are defined as assertions or propositional logic formulas.
\textit{Non-functional attributes, composition mo\-del entity} and \textit{mo\-del entity} are the notations used by
Xiao et al. \cite{XiaoCZBOLH08} for classifying the different concepts for
non-functional requirements modeling. Non-functional attributes (NFAs) describe
the non-function aspects in the abstract process models. The framework to
model NFRs proposes to annotate composition models with NFAs.

D'Ambrogio \cite{DAmbrogio06} uses the term \textit{quality characteristics}
to group similar characteristics into \textit{quality categories}. Each
\textit{quality characteristic} is quantified by \textit{quality dimensions}.
\textit{Quality characteristic} is a quantified QoS aspect, for example
\textit{latency, throughput, reliability, availability}, etc. \textit{Quality
characteristics} of a common subject are grouped into abstract \textit{quality
categories}, for example \textit{performance} (for latency and throughput characteristics) and
\textit{dependability} (for reliability and availability characteristics).

The terms \textit{category, sub-category}
and \textit{property}  are used by Yeom
et al.\cite{Yeom2006} to classify non-functional requirements.
\textit{Business}, \textit{service} and \textit{system} are the possible values to be associated
with the proposed terms, and a \textit{sub-category} can be \textit{security,
value, interoperability}, etc. The work in \cite{Yeom2006} defines a \textit{web
services quality model}, which considers non-functional properties in several
aspects. In this model, web services qualities are classified in categories and
sub-categories. Chollet et al.\cite{CholletL09} uses only two terms to
classify and relate quality properties with services, they are:
\textit{activity} and \textit{quality property}. Each activity represents a
functional property that can be divided in sub-activities, depending on its
granularity. For non-functional requirements, the work in \cite{CholletL09}
describes the possibility of creating different meta-models for each
\textit{quality property}, and then relate them with activities.

Schmeling et al.\cite{SchmelingCM11} uses the 
\textit{non-functional concerns (NFC)} notation to describe NFRs. This term
encompasses two aspects: the specification of NFCs and their realization.
\cite{SchmelingCM11} defines that a \textit{functional concern (FC)} is a
consistent piece of functionality that is part of a software system.
\textit{Non-functional concern (NFC)} is a general term describing a matter of interest or importance that does correspond
to a non-functional requirement of a system, for example, \textit{security,
reliability, transactional behavior}, etc. A \textit{non-functional action} represents some
behavior that results in a \textit{non-functional attribute}. An example of
\textit{non-functional action} is \textit{encryption}, which realizes the \textit{non-functional attribute},
\textit{confidentiality}. A \textit{non-functional activity} is also used as a
term, which means to encapsulate the control flow of \textit{non-functional
action} that apply to the same subject. The term is used in analogy to activity
and action in UML2.

Ceri et al.\cite{CeriDMF07} uses the terms \textit{policy, rule,
condition} and \textit{action model} to specify NFRs, and in a similar way,
Agarwal et al.\cite{AgarwalLS09} also uses the concept of \textit{service
policy} associated with the concept of service. Each service is also associated
with a \textit{service property}, which may have a specific value
(\textit{security, reliability}, and so on). Each service is also associated
with a \textit{unit function}, that represents one or more
requirements.

Ovaska et al.\cite{OvaskaEHPA10} uses
\textit{quality attribute, category, conceptual layer} and \textit{importance}
to organize and classify the NFRs, Pastrana et al.\cite{PastranaPK11} uses
the term \textit{contract} to describe non-functional requirements. In a
\textit{contract} it is possible to define pre-conditions, post-conditions and
invariants. A \textit{web service} can have many \textit{contracts}, that
defines many \textit{assertions} and are associated with \textit{quality
properties}.

Some related papers do not have any
nomenclature to classify non-functional requirements. However, some of them uses
the \textit{attribute} notation \cite{ZhangPSP05,BasinDL06,JeongCL09}, other
uses \textit{properties} \cite{Fabra2011}, other \textit{factors}
\cite{MohantyRP10,GutierrezRF10}, \textit{characteristics}
\cite{DiamadopoulouMPS08}, \textit{quality level} \cite{ModicaTV09} or
\textit{values} \cite{ThissenW06,BasinDL06}.

\subsection{Analysis}
\label{sec:proposal}

% The purpose of this analysis is to identify the properties and non-functional
% requirements used in the development of systems, looking for patterns
% or relationships between non-functional concepts in different
% modeling levels.

Although there exist many different types of notation used for classify
non-functional requirements, in general, the values associated to them are the
same, \textit{i.e.} \textit{security, performance, reliability, usability, availability}, etc. What
distinguishes the different approaches is the adoption of different NFR
hierarchies in order to prioritize or classify the quality requirements. 
Another interesting factor
is that the each work uses different approaches to model these requirements.
Thus, there are different kinds of notations for non-functional
requirements.

\begin{table*}[ht!]
\centering
%\scriptsize
\tiny
\begin{tabular}{l|l|l|l|l}
  \hline 
  \hline
   \textbf{Reference} & \textbf{\textit{NFR concepts}} &
   \textbf{\textit{NFR values}} & \textbf{\textit{Approach}} &
   \textbf{\textit{Domain / Scope}} 
   \\
  \hline
  \hline  
  Babamir et al. \cite{Babamir2010} &  property / & responsiveness /
  availability   & TDT & Software architecture
 \\  
  & category /  & performance / sla properties   &  & 
 \\
 &  constraint &    &  & 
 \\ 
  \hline

  Yeom et al. \cite{Yeom2006} & category / & business value / performance /
   & TDT & QoS model\\ 
   & sub-category / & stability /manageability /   & & \\
   & property & security / business processing  & & \\
    &  & interoperability & & \\ 
  \hline

  Xiao et al. \cite{XiaoCZBOLH08} &  NF attribute  & time /
cost / resource  & BP  & Business processes
   \\
 	&   &   &  & modeling
   \\   
  \hline 
  D'Ambrogio \cite{DAmbrogio06} &  characteristics /  & availability /
  reliability / & MDD & WSDL \\ &  category /  &  access control  &  &
  extension
  \\
  &  dimension &    &  & 
 \\
   \hline
  Chollet et al. \cite{CholletL09} & activity / & security  & MDD 
  & Orchestration Framework
   \\
  &  NF attribute &    &  & 
  \\

  \hline
  Schmeling et al. \cite{SchmelingCM11} & NF concern / & security &
  MDD &  Web service 
  \\
  & NF attribute / &  &
   & composition process
    \\
  &  NF action / &    &  &
   \\
  &  NF activity	 &    &  &
  \\ 
  \hline
    Thi{\ss}en et al. \cite{ThissenW06} & NF value  &  performance /
  reliability &  FM & Software architecture \\  
   & & cost / availability &  & \\  
  \hline
  Zhang et al. \cite{ZhangPSP05} & attribute /& security & FM &  Access control 
   \\
  &  predicate &    &  &
  \\

  \hline
  Basin et al. \cite{BasinDL06} &  attribute & security  & MDD & System
  architecture
  \\
  \hline
  Ceri et al. \cite{CeriDMF07} & police / rule  & n.a. & TDT & Context-aware
  applications
   \\
  &  condition / action &    &  &

  \\
  \hline
  Fabra et al. \cite{Fabra2011} & property & storage /
  processing & MDD & Web service methodology 
  \\
   &   & (*case study) &  & 
  \\
  \hline
  Modica et al. \cite{ModicaTV09} & quality level & sla
  properties & TDT &  Service oriented 
  \\
   &  &  &  & architecture \\
  \hline
  Ovaska et al. \cite{OvaskaEHPA10} & attribute & security / reliability
  & MDD &  Model development
  \\

  &  category &    &  &
  \\
  \hline
  Agarwa et al. \cite{AgarwalLS09} & property / & \textit{not explicitly
  defined} & Ont & Policy language
  \\

  &  policy / &    &  &
  \\

  &  function &    &  &
  \\
  \hline
   &  & operation cost /&  &  \\
  &  & performance / availability /  &  &  \\
  Jeong et al. \cite{JeongCL09} & NF attribute & accessibility / security /
  & AI  & Service oriented   \\ 
   &  &
  interoperability / usability / & & \\ &  & user satisfaction &  & architecture
  \\
  \hline
   &  & performance / reliability /  &  & \\
   & NF property / & scalability / capacity /&  & \\
  Pastrana et al. \cite{PastranaPK11} & contract  / & robustness /
  precision / & Ont & Web service methodology\\
  & assertion / & accessibility
  / availability / &  & \\ 
   & NF behaviour & interoperability / security  &  & \\    
  \hline
  Diamadopoulou et al. \cite{DiamadopoulouMPS08} & NF characteristic &
  user' subjective & TDT & Web service selection  \\ 
   &  & perception  & &  \\
  \hline
  Gutierrez et al. \cite{GutierrezRF10} & NF factor & Security  & BP &  Web
  service
  \\
   &  NF sub-factor&   &  &  development process
  \\  
  \hline
  Mohanty et al. \cite{MohantyRP10} & NF attribute or & reliability /
  performance / & & Artificial intelligent /\\ 
  & NF factor & integrity / usability  & AI  & Web services classification	 \\
  &  & response time / documentation &   &  \\

  \hline  
  \hline  
\end{tabular} 
\caption{Research question results - $RQ_1$, $RQ_2$, $RQ_3$, $RQ_4$.}
\label{tab:result02}
\end{table*}

A number of approaches use
\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} MDD
(Model Driven Development) for designing and developing systems. Fabra
\textit{et al.} \cite{Fabra2011} presents a complete methodology that does not
mainly focus on non-functional requirements. Fabra \textit{et al.} \cite{Fabra2011} also describes the importance of the use of MDD in the development of
service-oriented applications. \cite{ThissenW06,ZhangPSP05} use formal methods
to define a development process based on NFR for web services,
\cite{AgarwalLS09,PastranaPK11} use ontologies for the definitions and
modeling of non-functional requirements and \cite{XiaoCZBOLH08,GutierrezRF10} use Business Process Modeling (BPM) for
system specification, including NFR. Most works focus on composition
service modeling while others define requirements models to
represent the properties used.

In \cite{Babamir2010,Yeom2006} non-functional properties for web services
are classified into three categories such as
\textit{service level view, system level view} and \textit{business level view}.
In the \textit{\textit{business}} level the quality properties are:
\textit{service charge, compensation rate, penalty rate} and
\textit{reputation}; at the \textit{\textit{service}} level the quality
properties are: \textit{performance} and \textit{stability}; and at the
\textit{system} level the quality properties are classified into:
\textit{manageability, interoperability, business processing} and
\textit{security}.

In the method defined in \cite{XiaoCZBOLH08}, each task in the process model is
annotated with the NFAs (\textit{non-functional attributes}). During the design phase, the
service composition and the definition of NFAs are separated. Then,
each task in the process model is annotated with the corresponding NFA. The
attributes are related with tasks or data item. For data, the NFAs are:
\textit{value} and \textit{range}; and for tasks the NFAs are: \textit{cost,
time, resources} and \textit{expression}.

D'Ambrogio \cite{DAmbrogio06} presents a WSDL extension for describing the QoS
of web services. The process is based on MDA. The work presents a catalog
of \textit{QoS characteristics} for the web service domain and the Q-WSDL
(Quality WSDL) meta-model for modeling QoS properties in web services. The properties
presented are: \textit{availability, reliability} and \textit{access
control}. \cite{CholletL09} presents a security meta-model for web service
composition. The non-functional requirements considered concern
\textit{authentication, integrity} and \textit{confidentiality}. Each property is related with a service activity. In \cite{SchmelingCM11}, authors
present an approach and a toolset for specifying and implementing the
composition of several non-functional properties of Web services. The
non-functional attributes described in \cite{SchmelingCM11} are: \textit{confidentiality, integrity
(security concern)} and \textit{response time (performance concern)}.

The work presented in \cite{ThissenW06} describes
steps to design a selection mechanism of services identified as candidates for
participating in a composition, considering their quality properties. The steps
are: \textit{(i)} identification of relevant QoS information; \textit{(ii)} identification of basic composition patterns and
QoS aggregation rules for these patterns; and \textit{(iii)} definition of a
selection mechanism of service candidates. As QoS properties considered in
\cite{ThissenW06} are: \textit{performance, cost, reliability} and
\textit{availability}.

% We highlight the following works: \cite{CeriDMF07, 
% Yeom2006,JeongCL09,PastranaPK11}. These research works present a relevant
% detail of some NFR and a good detail of them in each stage of development.

Among the properties presented we highlight \textit{security}
and \textit{performance}. Users usually need to access data
securely and quickly. Most studies have both properties as the most
important non-functional requirements. \textit{Reliability} is also an important
non-functional requirement presented in some works and required by end users.

We can now relate the research questions presented in section \ref{sec:nfrs},
with the works described above. Tables \ref{tab:result02} and \ref{tab:result03}
show some results.

The vocabulary used for naming and characterizing NFR is not stable, 5 works out
360 (1.6\%). 19 of the total of works chosen (26.31\%) propose a
classification of non-functional requirements.

% The many different types of NFR terms used
% in these works, there is still no clear classification of NFRs
% and a hierarchy or dependency between them. Only 5 of the works present a hierarchy
% among the different non-functional requirement notations, representing only
% 1.6\% of the total (over 306 works - appendix \ref{append:analysis}) and 26.31\% of total of
% works chosen (over 19 works - appendix \ref{append:analysis}). This hierarchy
% is used according to the modeling level and process phase.

% Among these works, we highlight
% \cite{Yeom2006,DAmbrogio06,CeriDMF07}, because it presents the types of NFR and
% their importance at each modeling level.

\begin{table}[ht!]
\centering
\tiny
\begin{tabular}{l|l|l|l}
  \hline 
  \hline
   \textbf{Reference} & \textbf{\textit{Service model}} --
   & \textbf{\textit{Service type}}  &
   \textbf{\textit{Year of}} 
   \\
    &   \textbf{\textit{Business services}} &   &
   \textbf{\textit{publication}} 
   \\
  \hline
  \hline  
  Babamir et al. \cite{Babamir2010} & no -- yes  & composition & 2010   
 \\  
  \hline   
  Yeom et al. \cite{Yeom2006} & yes -- yes & single   & 2006  \\  \hline
  Xiao et al. \cite{XiaoCZBOLH08} & no -- no & composition    & 2008
   \\ 
  \hline 
  D'Ambrogio \cite{DAmbrogio06} & yes  -- no & composition  & 2006 \\
   \hline
  Chollet et al. \cite{CholletL09} & yes -- yes & composition  &  2009 \\
  \hline 
  Schmeling et al. \cite{SchmelingCM11} & no -- no & composition &  2011 \\ 
  \hline
   Thi{\ss}en et al. \cite{ThissenW06} & no -- yes & composition & 2006
   \\
  \hline
  Zhang et al. \cite{ZhangPSP05} & no -- no  & single  & 2005 \\ 
  \hline
  Basin et al. \cite{BasinDL06} & yes -- no & single / composition  & 
  2006\\
  \hline 
  Ceri et al. \cite{CeriDMF07} & no -- no  & single  &  2007\\ 
  \hline 
  Fabra et al. \cite{Fabra2011} & yes -- yes & composition  &  2011\\
  \hline
  Modica et al. \cite{ModicaTV09} & no -- no & composition & 2009\\ 
  \hline
  Ovaska et al. \cite{OvaskaEHPA10} & yes -- no & single & 2010\\
  \hline
  Agarwa et al. \cite{AgarwalLS09} & yes -- no & single / composition  &  
  2009\\
  \hline
  Jeong et al. \cite{JeongCL09} & no -- no  & composition &  2009\\
  \hline
  Pastrana et al. \cite{PastranaPK11} & yes -- no & composition &  2011 \\
  \hline
  Diamadopoulou et al. \cite{DiamadopoulouMPS08} & no -- no & composition 
  & 2008\\
  \hline
  Gutierrez et al. \cite{GutierrezRF10} & no -- no &  single / composition 
  & 2010\\
  \hline
  Mohanty et al. \cite{MohantyRP10} & no -- no  & single  & 2010\\
  \hline
  \hline  
\end{tabular}
\caption{Research question results - $RQ_5$, $RQ_6$, $RQ_7$.}
\label{tab:result03}
\end{table}

There are other works
\cite{ieee_1998,CysneirosLN01,Cleland-HuangSZS06,Glinz05rethinkingthe,sommerville08}
that consider and propose non-functional requirements classifications. A number
of software attributes are defined as requirements in
\cite{sommerville08,ieee_1998}. According to \cite{ieee_1998}, the most common
non-functional properties are: \textit{performance}; interface; operational;
resource; verification; acceptance; \textit{maintainability}; documentation;
\textit{security}; \textit{portability}; quality; \textit{reliability};
\textit{usability}; and \textit{safety}. We highlight those that are used in
most analyzed classifications. \cite{ieee_1998} proposes an NFR specification
and a template for specifying requirements, considering both, functional and
non-functional properties.

The NFRs are classified in 4 main concepts by \cite{CysneirosLN01}. The
classification uses \textit{quality properties}. These properties are:
\textit{performance; security; cost;} and \textit{usability}. Pastrana
\cite{PastranaPK11} describes an ontology based methodology and uses a NFR
classification based in some properties, most already described in other works.
However that work is the only one to use \textit{scalability, capacity}
and \textit{precision} properties. Considering web services development,
these properties are not very frequently used, however, considering data
processing, these properties can be important in cases of large data processing
requests through services.

Sommerville \cite{sommerville08} classifies requirements,
either functional or non-functional, in three main blocks:
\textit{process, product} and \textit{external}. \cite{sommerville08} 
also defines a sub-classification considering the system domain.
Software requirements are classified as: \textit{usability; reliability; safety;
efficiency; performance;} and \textit{capacity}.

% Based on the works presented in \cite{Cleland-HuangSZS06,Glinz05rethinkingthe},
% and considering the previous analysis of NFR, the properties highlighted are
% security, performance and reliability. Although the works have
% different focuses and the NFR classification and definition are made in
% accordance with different arguments, the more present in NFR ratings
% requirements are that ones presented in Figures \ref{fig:NRFmodel} and
% \ref{fig:proposalNRF}.

Some requirements may determine the system design, for example, \textit{(i)}
Keep certain functions in separate modules; \textit{(ii)} check data integrity
for critical variables; and \textit{(iii)} permit only limited communication
(requester / provider), are examples of restrictive requirements
\cite{ieee_1998}. These examples are related with some NFR concepts as: \textit{security, reliability} and
\textit{availability}.

% Thus, in the context of service-based
% development, it is important that non-functional requirements are described
% through abstract modeling levels.

%    Thus, based on theses works and the
% results obteined in the systematic review, the results presented in tables
% \ref{tab:result02} and \ref{tab:result03}, we will propose a NFR model and
% classification for service based application development.

% \subsection{NFR Metamodel:} 
% \label{sec:servicedefinitionNFR}

In the service-based development there is a clear difference between the
business, service and system levels. The quality requirements are treated
differently in each of these levels. Despite the different nomenclatures they
can be described in general as \textit{non-functional concern/requirement} and
\textit{non-functional attribute}.

Considering \cite{ieee_1998} and \cite{sommerville08}, we notice that
restrictions are usually related to use cases and functional requirements, and
in most cases, a service activity can be represented as an use case. It is in
this way that non-functional requirements with web services are related. They
don't propose a specific way to design constraints to services, but assuming
that a service is modeled as a use case, and it has quality requirements.

At the service level, each service activity is related in some way with the
concept of contracts. Contracts can be grouped into policies and their
rules. Policies are directly related to concepts like quality,
security, performance and availability; contracts are associated with non-functional attributes.

\bigskip

% In spite of the variety of tools and new approaches for web service development,
% there is not (yet) a consensus on a software process methodology for web
% services
% \cite{PapazoglouH06,Papazoglou03,cdl2006,MilanovicM06,FeuerlichtM05,Ramollari_asurvey,somet2005}.
% %  In order to develop and use such a methodology, it is of paramount importance to
% % provide adequate descriptions of its expected capabilities and competences.  
%  
% Works proposing methodologies can be
% classified into two types: (\textit{i}) those that propose new approaches for non-functional
% properties guarantees; and (\textit{ii}) those proposing service-oriented
% development methodologies for the whole development process.
%     
% \cite{HL05TACoS} proposes \textit{Design by Contract} for web services. It is
% possible to describe three levels for specifying contracts: \textit{implementation level,
% XML level} and \textit{model level}. Design by Contract applied to web services
% allow the verification services web through the use runtime checkers, before the
% deployment, such as \textit{jmlrac} \cite{LeavensCCRC02}, by adding behavioral
% information to the specification of services, using JML \cite{LeavensCCRC02}.
% 
% CDL (\textit{Contract Definition Language}) \cite{cdl2006} is a XML-based
% description language, for describing contracts for services. The
% development with CDL offers an architecture framework, design standards and a
% methodology \cite{MilanovicM05,Milanovic05,Milanovic06,MilanovicM06}, that can
% be easy understood and applied to the development of real applications. The
% greatest difficulty is that the language only
% represents contracts for services. Its specification is generated by
% refining B \cite{AbrialLNSS91} machines that describe
% the services and their compositions.

% \begin{table*}
% \centering
% \caption{Some Typical Commands}
% \begin{tabular}{|c|c|l|} \hline
% Command&A Number&Comments\\ \hline
% \texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
% \texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
% \texttt{{\char'134}table}& 300 & For tables\\ \hline
% \texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
% \end{table*}
% end the environment with {table*}, NOTE not {table}!

\begin{table}[ht!]
\tiny
\begin{tabular}{l|c|c|c|c}
  \hline
  \hline
   & IEEE & ACM  & Science Direct & Total \\
  \hline
  \hline
  Total results & 65 & 271 (75 \footnote{We considered only 75 transactions
  journal works}) & 166 & 502 (306 \footnote{considering the 75 ACM
  transactions journal works}) \\
  \hline
  Step one  & 19 & 10 & 20  & 49 \\
  Step one (\%) & 29.23\% & 13.33\% & 12\% &
  16\% \\
  \hline
  Step two & 7 & 3 & 9 & 19\\
  Step two (\%)  & 10.76\% & 4\% & 5.42\% & 6.20\%\\
  \hline
  \hline
\end{tabular}
\caption{Summary of the studies selected at each step.}
\label{tab:result01}
\end{table}

The extraction of information was based on the research questions, and each work extraction question included the following
items: \textit{(i)} where the paper was found; \textit{(ii)} identification of the title and main
subjects; \textit{(iii)} summary of the research; \textit{(iv)} inclusion and
exclusion criteria; \textit{(v)} objective and result; and \textit{(vi)}
subjective results.

% The systematic literature review took place between December 2011 and January
% 2012 and we did not filter works within a specific year interval. Although the
% studies which were concretely analyzed were published between 2005 and 2011.
Table \ref{tab:result01} shows a summary of the studies selected in each stage
of the selection procedure for each source. The ``Total results'' were obtained
by running the search string on the selected sources. The next four rows show
the results obtained after applying stages one (2 first rows) and two (2 last
rows) of the studies selection procedure.

\begin{figure}
\centering
\includegraphics[width=.49\textwidth]{figs/data.png}
\caption{Publications per year.}
\label{fig:statistics}
\end{figure}

In the first step, respecting the filters described, the 65 articles collected
from IEEE, only 29.23\% of them were in accordance with the criteria
described early, representing 19 articles. In ACM Library, from 75 works
collected, only 13.33\% passed in the first stage filter, representing 10 articles. In Science Direct
had the lowest percentage, totaling 20 of the 166 articles collected by the
query, thus representing 12\% of the total. Despite being the lowest relative
value, the Science Direct had the largest absolute result, with 20 works
in the first step. In the second stage, the percentage dropped further, and the
relevant works and with accordance to the criteria have been collected as the
final result. The results were respectively, 10.76\%, 4\% and 5.42\% of total
from the IEEE, ACM and Science Direct. The highest percentage was among the
 works from  IEEE, while the largest number od results, in absolute terms,
 was collected from Science Direct. The approaches resulting from this last
 stage were studied in depth and information concerning the detailed research
 questions and other fields of the extraction forms was extracted from each
 eork we selected. 49 works were selected in the first stage, and, only 19
 works in the second stage. It represents  6.20\% of the total amount of works.

Figure \ref{fig:statistics} shows the publications per year, from
2005 to 2011. 12 of 19 articles were selected in the systematic review published
in 2006, 2009 and 2010, being four in each year and 6 in Science Direct source,
5 in IEEE, and only 1 at the ACM. All nine of Science Direct were
published in the last 4 years. Figure also
shows that the number of publications that consider classification of NFR once
again increased from 2008.

\section{Classification of Non-Functional Requirements}
\label{sec:classification}

According to the previous analysis, we defined the main non-functional
requirement concepts that are associated with service-based development. We also
present a synthesis of the concepts common to the analyzed approaches used for
modelling non-functional requirements of service-based applications. The NFRs
have been classified into three levels, \textit{business, services} and
\textit{systems}. According to each modeling level, non-functional requirements
are being refined from the business level to system level.

One classification of NFRs can be seen in \cite{Yeom2006}.
In that classification, NFRs are not applied on data but just to functions and
 service performance. 
We consider that it is important to classify the
 requirements of business and data (value) restrictions, since data accessible to web
 services depend on the application contexts.

 In sections \ref{sec:nfr-metamodel} and \ref{sec:nfr-classification} we present
 which concepts (NFR meta-model) and values (NFR classification) we consider
 important to a reasonable modeling of non-functional requirements for service
 applications.

\begin{figure*}[ht!]  
\centering  
\includegraphics[width=0.70\textwidth]{figs/contractModel.pdf}
\caption{Service-Based Non-Functional Requirement Model.}
\label{fig:NRFmodel} 
\end{figure*}

 \subsection{NFR Meta-Model} 
\label{sec:nfr-metamodel}

The model present in figure \ref{fig:NRFmodel} shows the relationship between
those concepts
%\footnote{The appendix
%\ref{app:nfr-concepts} provides a description of each concept presented in
%figure \ref{fig:NRFmodel}.} 
we consider important for quality requirements used
in service-based development.

% One requirement, whether functional or non-functional, can be represented by one
% or more use cases. For example, a paying process can be modeled through the use
% cases that represent withdrawal and deposit of money in transaction of customer
% and service provider accounts. A payment process also requires the guarantee of
% a complete transaction and data security. Thus several use cases are modeled to
% a single requirement.

%Each concept represents:

One \texttt{Requirement}, whether functional or non-functional, can be
represented by one or more use cases. 
Each use case represents a \texttt{Service
Activity}. 
For example, a paying process can be modeled through the use cases
that represent withdrawal and deposit of money in transaction of customer and
service provider accounts. A payment process also requires the guarantee of a
complete transaction and data security. Thus several use cases are modeled to a
single requirement. All these use case can be implemented as services.
% 
% A composite use case is a type of use case. This type of
% use case abstracts a more macro function, and may contain extra functions to be
% performed so that the process is complete. For example, to publish a song, you
% must choose which song you want to publish, , authenticate to twitter or facebook,
% so that the music you're listening is published. We consider this type of use
% case, a use case compound.

Each use case has \textit{business} or
\textit{value} constraints. \textit{Business constraints} are restriction on
functions and how they may be implemented. The \textit{value constraints} are restrictions on
the service interface, which the desired values for input and output data. Each
constraint is associated with NFAs.

A \texttt{Contract} is a set of constraints for the same function. For example,
a contract for the payment operation. The constraints for payment are: (i) the
value amount should not be less than 10 euros and (ii) the user should always
receive a purchase confirmation by phone message. This restrictions are
grouped into a single contract for payment verification.

% These constraints
% are also associated with non-functional attributes, e.g., reliability,
% transactions and data privacy card.

An \texttt{Exceptional Behavior} happens when a contract is not respected. When
this happens a new function is called or the process is stopped. For example, if
the bank does not authorize the payment, the system offers alternative forms
of payment such as PayPal.

Finally a \texttt{Policy} groups similar contracts. For example, security
contracts are grouped into a security policy and performance contracts are
grouped into a performance policy.

\subsection{NFR Classification}
\label{sec:nfr-classification}

\begin{figure*}[ht!]  
\centering  
\includegraphics[width=0.70\textwidth]{figs/nfrRelationship.pdf}
\caption{Relationship of the NFR Concepts.}
\label{fig:nfr-relationship}   
\end{figure*}

Table \ref{tab:result04} shows the NFR classification we propose, organized
 as follows (by column): (i) the level of abstraction, (ii) the proper term for
 this kind of abstraction and (iii) possible values to be used. The rows
 represent the abstraction level for modeling the NFR. The highest level is the
 \textit{Business level}, the intermediary one is the \textit{Service level};
 and finally the lowe level, called \textit{System level}.

\begin{table}[ht!]
\centering
%\scriptsize
\tiny
\begin{tabular}{l|l|l}
  \hline 
  \hline
   \textbf{Modeling Level} & \textbf{\textit{Concept / Notation}} & \textbf{\textit{NFR / NFA}}  
   \\
  \hline
  \hline   
  Business Level & Constraint  & Business Constraint,    
 \\  
  &   & Value Constraint\\
   \hline    
   &  & Integrity, Transaction,  \\
   &  & Accessibility, Encryption, \\
   &  & Cost, Time Constraint, \\
  Service Level & Contract & Encryption, Platform, \\
   &  & Privacy, Authentication, \\
   &  & Resource, Capacity, \\
   &  & Privacy, Confidentiability 
   \\   \hline
   &  & Security, Performance,\\ 
   & & Interoperability, Scalability,\\
  System Level & Policy & Reliability, Usability,\\
   & & Transactional Behaviour,\\
   & & Availability \\ 
   \hline
  \hline  
\end{tabular}
\caption{Non-Functional Requirements Classification.}
\label{tab:result04}
\end{table}

At the business modeling level, non-functional requirements are classified
into business restrictions. We have adopted, \textit{business} and \textit{value}
constraints to address the most abstract levels of restrictions in business
modeling.

The relationship between non-functional requirements and attributes help
identify groups of restrictions and contracts to develop specific policies. The
level of service shows non-functional requirements, and system level presents a
finer granularity, describing the non-functional attributes. Thus, a set of
contracts that are related to the same attribute form specific policies for the
system. The relationship between the concepts in business, service and system
levels are presented in figure \ref{fig:nfr-relationship}.

% \begin{itemize}
%   \item Security:
% 	\begin{itemize}
%   		\item Access control; Transaction; Privacy; Cost; Authentication;
%   		Integrity; and Confidentiability.
% 	\end{itemize}
%   \item Performance:
% 	\begin{itemize}
%   		\item Time constraint; Capacity; and Transaction.
% 	\end{itemize}
% 	\item Availability:
% 	\begin{itemize}
%   		\item Capacity; and Time constraint.
% 	\end{itemize}
% 	\item Interoperability:
% 	\begin{itemize}
%   		\item Conformability; and Platform.
% 	\end{itemize}
% 	\item Usability: 
% 	\begin{itemize}
%   		\item Accessibility; and Platform.
% 	\end{itemize}
% 	\item Reliability:
% 	\begin{itemize}
%   		\item Access control; Confidentiability; Accessibility; Integrity; and
%   		Resource.
% 	\end{itemize}
% \end{itemize}

In our work we adopt \textit{security, performance, availability,
interoperability, usability} and \textit{reliability} as NFRs (these are
system level properties), and we consider as NFAs \textit{access control, time
constraint, privacy, accessibility} and so on (which are service level
properties).

\section{Conclusions}
\label{sec:conclusion}

This paper presented a systematic review of NFRs associated with web services.
We grouped the NFRs according to their characteristics and analyzed them in each
context of application. We also reviewed some existing NFR classification for service-based
development.
% In general, after reviewing the current research works related
% with the problem of service-oriented development, we could obtain
% some conclusion.

To the best of our knowledge, there are no proposals that define a
service-oriented approach for the \textit{complete} development of systems,
considering non-functional requirements. Although many efforts have been made to support the new
technological proposals for the Web such as web services, in general, existing
approaches focus their processes on traditional software engineering
methodologies, with emphasis to the functional aspects of the application.

Thus, our analysis and classification can help to improve the development of
service-based applications which have as their main effort the
guarantee of quality requirements and the generation of reliable systems,
considering NFRs. We believe that the early modelling of NFRs during the
software process helps users and developers to improve the quality of the
solution.

\bibliographystyle{abbrv}
\bibliography{biblio} 
\end{document}

### 2013-09-30 SBCARS2013/src/aux_files/methodology.tex
inicial --martin

\chapter{$\pi$SOD-M: A Methodology for Building Reliable Service Based
Applications}
\label{chapter:methodology}

\epigraph{``\textit{There are two ways of constructing a software design: One way is to make it so
simple that there are obviously no deficiencies, and the other way is to make it
so complicated that there are no obvious deficiencies. The first method is far
more difficult.}''}{C.A.R. Hoare}

% Software construction is moving towards the use of methodologies for controlling
% the software construction process and facilitating the specification of software
% functional and non-functional requirements (NFRs), and its maintenance and
% evolution for integrating new functionalities. 
% There exist several methodologies, methods and new
% approaches\cite{CastroMV11,PapazoglouH06,sommerville08,Gervais02,somf,MilanovicM06}
% for software engineering (such as presented and analysed in the previous
% chapter), improving software development and modeling non-functional
% requirements. These methods can help to ensure the coherence between
% functional and non-functional properties necessary to the system, particularly
% when information systems include complex business processes calling web services
% or legacy applications exported as services. The challenge is to ensure
% reliability properties for the complete application. There are still a number of
% problems to be solved in the development of system, specially for service-based
% development, \textit{e.g.}, automatic composition; NFRs specification and
% modelling; and specific development methodologies for service-based applications.

% Software is moving towards models, methodologies, software process and
% architectures that should be easily modeled and developed, accommodate changes
% and integrate new functionality. Functional and non-functional
% requirements (NFRs) may demand such a robust method to provide the necessary
% support for the modeling and development software process.

This chapter presents $\pi$SOD-M (\textit{Policy-based
Service Oriented Development Methodology}), a methodology for modeling and
developing reliable web service based applications using a MDA approach. We
propose guidelines for building service compositions that include non-functional
requirements (NFRs). The methodology focuses on the development of
service-based system, considering non-functional aspects of the software being
developed. These aspects are implemented as policies for the application
services. The design of this aspects integrates both functional and
non-functional requirements.

This chapter is organized as follows. Section
\ref{sec:pisodm} introduces the concepts, general structure and motivations
to define a specific methodology for policy-based web service development.
We use a motivational example for
illustrating the concepts and principle of $\pi$SOD-M. Sections
\ref{sec:pim-pisodm}, \ref{sec:psm-pisodm} and \ref{sec:models-tranformation} describe the development
models and transformations of our methodology. They also how to apply the
methodology for building concrete applications. Section \ref{sec:pisodm_conclusion} concludes the chapter.

\section{$\pi$SOD-M}  
\label{sec:pisodm}

$\pi$SOD-M is a MDA (Model Driven Architecture) based methodology. It provides
a \correctingText{environment} for building service compositions considering
their non-functional requirements. $\pi$SOD-M extends the SOD-M \cite{valeriaThesis} method by adding
the concept of \textit{Policy} \cite{Espinosa-OviedoVZC09,Espinosa-Oviedo2011a}
for representing NFR associated to service-based applications. $\pi$SOD-M also
proposes the generation of a set of models at different abstraction levels, as
well as transformations between these models.

% Our approach extends SOD-M with the notion
% of \textit{Policy} \cite{Espinosa-OviedoVZC09,Espinosa-Oviedo2011a} for
% representing non-functional aspects associated to service-based applications.

% A model driven development process needs both models at different
% abstraction levels and transformations between them. The
% transformations allow us to relate elements from an input source model to an
% output target model (different from each other). At each level, a transformation
% between models is performed to refine the specification of the system into a
% more concrete version of it.

$\pi$SOD-M's models represent both the cross-cutting aspects of the application
being modelled, as well as the constraints associated to services. The systems
cross-cutting concerns affect functional concerns, such as availability;
recovery; and persistence aspects. Constraints are restrictions that must be
respected during the execution of the application, for example the fact that a
service requires an authentication for executing system functions.

\newText{
$\pi$-SOD-M supports the construction of service-oriented applications that implement business processes.
Therefore, it proposes a development process based on the definition of models
(instances of the meta-modes) and transformations for semi-automatically
generating different levels of abstraction that describe a service-oriented
application from abstract platform independent views (CIM and PIM level) to
platform dependent views and the PSM and implementation levels. We extended the
Business and Service views of the SOD-M method \cite{CastroMV11}. The Business
view defines  the concepts for modeling a business process, while the Service
view defines the concepts  for designing services compositions. Our methodology
introduces concepts (e.g. NF-requirement, constraint, assertion, contract,
policy) in the Policy view for describing constraints associated to services and
non-functional properties associated to service processes. }

$\pi$SOD-M defines a service oriented approach providing a set of guidelines to
build service based information systems (SIS) and proposes to use services as
first-class objects for the whole system development process. $\pi$SOD-M extends
the SOD-M models to include capabilities to model NFRs. The
SOD-M models that are being extended are: \textit{use case model, extended use case model, service
process model} and \textit{service composition model}. $\pi$SOD-M
provides a conceptual structure to: (i) capture the system requirements and specification in
high-level abstraction models (computation independent models, CIMs); (ii)
obtain the PIMs from such models (specification documents). The
platform independent models (PIMs) are designed to specify the system
details; (iii) transform such models into platform specific models
(PSMs)\footnote{\correctingText{$\pi$-PEWS language.}} that bundles the
specification of the system with the details of the targeted platform; and (iv)
serialize such model into the working-code that implements the system.

% As one of the main aims of MDA is to separate design from architecture and
% technologies, the $\pi$SOD-M's models describe the system behaviour and its
% restrictions, do not considering definition of standard architecture, because
% the devising of the system's architecture is outside the scope of our work. We
% consider that all applications to be modeled are services.

\subsection{General Overview}

% The SOD-M \textit{use case} and \textit{extended use case} models were extended
% to a single model named \textit{$\pi$-UseCase} (figure
% \ref{fig:sodmExtensions}).

% The original models (\textit{use case} and  \textit{extended use case} models) do not define
% concepts to design non-functional restrictions, which are considered in the
% \textit{$\pi$-UseCase} model.

% \textit{service process} model of SOD-M was
% extended to the

 Figure \ref{fig:sodmExtensions} presents the SOD-M models in the context of MDA
 architecture, and defines which models are extended by our approach. The
 \textit{$\pi$-UseCase} model describes services requirements, constraints and quality requirements. The \textit{$\pi$-ServiceProcess} model defines the
concept of service contract for representing the service input and output data restrictions, as well as
function restrictions. In this model we propose the definition of service
contracts. The \textit{$\pi$-ServiceProcess} model groups the constraints
described in the \textit{$\pi$-UseCase} model into contracts that are associated
with services. The  \textit{$\pi$-ServiceComposition} model provides the concept of \textit{Policy}
 that groups contracts with similar non-functional requirements. For
example, security restrictions, such as contracts for authentication, privacy
data access or transactions are grouped into a security policy.

% \textit{service composition} model of SOD-M was extended to the

%  . The contracts specified in the \textit{$\pi$-ServiceProcess} model are
%  refined into policies in the \textit{$\pi$-ServiceComposition} model. \textit{Policy}
% 
%

\begin{figure}[ht]
\centering
\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/ArquiteturaPiSOD-M}
\caption{SOD-M and $\pi$SOD-M Extension Models.}
\label{fig:sodmExtensions}
\end{figure}

% The services restrictions are being refined into each level of modeling.

In the \textit{$\pi$-UseCase} model, use cases restrictions are designed as
constraints. In the \textit{$\pi$-ServiceProcess}, the constraints are grouped
into service contracts, and use cases are refined to services or functions.
Then, contracts are grouped into policies.

 The $\pi$SOD-M methodology proposes the \textit{$\pi$-PEWS} model as platform
 specific model. An instance of the \textit{$\pi$-PEWS} model is generated from
 the \textit{$\pi$-ServiceComposition} model, and it is a representation of a
 \textit{$\pi$-PEWS} specification \cite{BaCAM05,Placido2010LTPD}. From
 a \textit{$\pi$-PEWS} model it is possible to generate a service composition
 code.

  $\pi$SOD-M also defines model-to-model transformation rules (starting from
  the \textit{$\pi$-UseCase} model) to \textit{$\pi$-ServiceComposition} model;
  and uses model-to-text transformations to generate the corresponding
  implementation code in the \textit{$\pi$-PEWS} language.

%   Non-functional
%   requirements are modelled by specifying high-level platform
%   independent models, and then the specific platform code is generated.

\subsection{Development Process}

$\pi$SOD-M uses the concept of MDA viewpoint, used as a technique of
abstraction to focus on a particular aspect of the issue or proposed system.
MDA defines, specifically, three viewpoints (figure
\ref{fig:developmentProcess}):

\begin{itemize}
  \item \textbf{Computation Independent Viewpoint:} This level focusses the
  environment of the system, as well as on its business and requirements
  specifications. At this moment of the development, the structure and system
  processing details are still unknown or undetermined. In $\pi$SOD-M, this
  level is represented as a list of business services from a requirements and
  business specification document. 
  \item \textbf{Platform Independent Viewpoint:} This level focusses the system
  functionality, hiding the details of any particular platform. This
  specification defines those parts of the system that do not
  change from one platform to another. In $\pi$SOD-M, this level is modelled by the system use case, service
  process and service composition models.
  \item \textbf{Platform Specific Viewpoint:} This level focusses the
  functionality, without hiding the details of a particular platform,
  combining the platform independent view with the specific aspects of the platform to
  implement the system.  In $\pi$SOD-M, the result of this level is the
  $\pi$PEWS specification \cite{Placido2010LTPD} which represents as a platform
  specific model.
\end{itemize}

\begin{figure} [ht!]
\centering
\includegraphics[width=0.8\textwidth]{fig/PiSOD-MProcess}
\caption{$\pi$SOD-M Development Process and Models.}
\label{fig:developmentProcess}
\end{figure}

 Computation Independent Models (CIM) aim to represent the business view, while
 Platform Independent Models (PIM) and Platform Specific Models (PSM) aim
 to represent the information system view and detail the information system to
 be implemented to fulfill the requirements of a business environment.

 Figure \ref{fig:developmentProcess} presents the $\pi$SOD-M development
 process, which defines a service oriented approach providing the guidelines for
 building service-based information systems (SIS), that was the result of the
 SOD-M extension described in figure \ref{fig:sodmExtensions}.

%  Notice that it is
%  possible to have different platform models at the PSM level (figure
%  \ref{fig:sodmExtensions}), such as BPEL or WSDL. However,
%  $\pi$SOD-M defines the \textit{$\pi$-PEWS} model as platform specific model. We
%  choose \textit{$\pi$-PEWS} as platform  specific model because this language
%  can express in their specification, service compositions and restrictions on
%  services through contracts. Besides being a language developed by our research
%  group.

Next section presents the concepts that lead to modeling applications in
$\pi$SOD-M. These concepts form the basis of our methodology for modeling the
reliable service-based applications.

\subsection{Methodology Concepts} 
\label{sec:concepts}

% $\pi$SOD-M proposes a new approach for defining non-functional
% aspects in service-oriented development.

The concepts presented in figure \ref{fig:pisodm-concepts} represent the main
elements of the methodology used for system application modeling. The $\pi$SOD-M's
meta-model\footnote{\textit{Meta-model} is the construction of a collection of
``concepts'' (things, terms, etc.) within a certain domain. A model is an
abstraction of phenomena in the real world; a meta-model is yet another abstraction, highlighting properties of the model
itself. A model conforms to its meta-model in the way that a computer program
conforms to the grammar of the programming language in which it is written.}
concepts are used in all methodology meta-models. They
describe the key concepts that must be modelled in a service-oriented
application. The $\pi$SOD-M meta-model consists of the set of concepts for
modelling and development applications that use MDA. These concepts represent the reliable
service-based system development, and are present in the modelling any
application.

Notice that the three $\pi$SOD-M methodology views (\textit{Business, System}
and \textit{Policy}) are different to those levels proposed by the traditional
MDA literature: \textit{CIM, PIM} and \textit{PSM}. In the $\pi$SOD-M
meta-model, at different MDA levels, there are elements from the $\pi$SOD-M
concepts.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/PiSOD-M.pdf}
\caption{$\pi$SOD-M Concepts.}
\label{fig:pisodm-concepts}
\end{figure}

 All the proposed concepts and models for business modeling (CIM
 level), and for information system and policy modeling at PIM and PSM
 levels comprise the backbone for the $\pi$SOD-M based applications design.
 These concepts are important to identify the system requirements and properties throughout the
 development. Thus, the purpose of this section is to present a description of
 each concept, so that when applied and used, it can be seen in their
 proper modeling context.

The methodology relies on a
set of concepts (represented as a meta-model) for modeling
reliable applications. These concepts are structured in three views:
\textit{Business, System} and \textit{Policy views} (figure
\ref{fig:pisodm-concepts}):

\begin{itemize}
  \item \textbf{Business view:} focuses on the business features
  that are the basis of an information system.  The concepts that
correspond to the \textit{Business View} describe business elements. There are
some concepts which correspond to both views (\textit{Business and System
Views}) and will be analyzed from both perspectives.

  \item \textbf{System View:} focuses on the main features and
  processes for the information system development. The \textit{System View}
  concepts are used to describe the functionality and processing system.

	\item \textbf{Policy View:} focuses on non-functional requirements and
	business constraints of the information system. The concepts that correspond to
	the \textit{Policy View} describe the NFRs and constraints related to the
	system functionalities. $\pi$SOD-M's main goal is to model non-functional
	requirements considering the models proposed by the SOD-M methodology.
\end{itemize}

% Figure \ref{fig:pisodm-concepts} shows a diagram that includes the
% set of concepts proposed by $\pi$SOD-M (which are organized in defined views),
% as well as the relationship between these concepts.

%  The SOD-M concepts do not provide this
% level of modeling. Thus, our proposal add non-functional
% concepts for modeling reliable software from early stages
% of service development (specification).

\newText{
The organization into different levels of abstraction of non-functional properties proposed by $\pi$-SOD-M relies on its concept of non-functional Requirement (NFR) defined  as a group of semantically correlated non-functional
attributes (NFA). For example, \textit{security} is an NFR that
comprises attributes such as \textit{confidentiality} and \textit{integrity}.
NFR are used for modeling global constraints, NFAs model non-fuctional characteristics that can be precisely measured, such as
 \textit{performance} \cite{RosaC04}. This kind of characteristics are closer to
 implementation. Then, the concept non-functional property proposed by
 \cite{RosaC04,Chung2009,MylopoulosBook99} is used for modeling  different
 types of software constraints. Thus, it generalizes the concepts NFR and NFA of
 our methodology. In $\pi$SOD-M, NFR and NFA are  modelled along the different
 modeling levels  as \textsc{constraint types} that can be grouped and 
 transformed into \textsc{contract types}, that can be transformed into
 \textsc{policy types}. A constraint type is used for representing  a
 non-functional property. }

 The $\pi$SOD-M concepts\footnote{The concepts marked with $^\ast$ are also part of
 the  \textbf{System view}} are:

\begin{itemize}
   \item \textbf{Business view:}
		\begin{itemize}
		  \item \textsc{Business process -} Represents the aggregation of
		  logically related tasks that are carried out to achieve a given business result.
		  \item \textsc{End consumer$^\ast$ -} Represents an entity that needs and
		  consumes business services. End consumers are those who pay (either with money or using
		  any other kind of value) to obtain a service that they use
		  themselves.
		  \item \textsc{Business service$^\ast$ -} Represents a result of a
		  business process (or part of it) providing value for an end consumer.
		  \item \texttt{Business task$^\ast$ -} Represents a business function
		  performed by a business collaborator as a part of a business process.
		  \item \textsc{Business collaborator$^\ast$ -} Represents an entity that
		  collaborates in the business processes of an organization, performing certain tasks needed
		  to provide a business service.

		\end{itemize}

\item \textbf{System view:}	
	\begin{itemize}
	  \item \textsc{Requirement -} Represents a super type for functional and
	  non-functional requirements. Thus, the use cases can be related to both types
	  of requirements.	
	  \item \textsc{Use case -} Represents a set of actions performed by the system
	  to carry out part of a business service.
	  \item \textsc{Composite use case -} Represents a set of actions performed by
	  the system to carry out part of a business service, which can be broken down
	  into different use cases, which may in turn be basic or composite.
	  \item \textsc{Service composite process -} Represents a set of logically
	  related activities necessary for carrying out a business service.
	  \item \textsc{Service activity -} Represents a behaviour (set of individual
	  actions) forming part of the execution flow of a business service.
	  \item \textsc{Action -} Represents a fundamental behaviour unit that is part
	  of a service activity and describes some transformation or processing in the
	  system being modelled.
	\end{itemize}

\item \textbf{Policy view:}		
	\begin{itemize}
		  \item \textsc{Non-Functional (NF) Attribute -} An attribute that describes
		the quality or characteristics of a functional requirement. For example
		\textit{confidentiality} and \textit{privacy} is an example for a non-functional
		attribute for the functional requirement \textit{user registration}.
		  \item \textsc{Non-Functional (NF) Requirement -} A group of semantically
		correlated non-functional attributes (NFA). For example, \textit{security} is
		an NF Requirement that comprises attributes such as \textit{confidentiality}
		and \textit{integrity}.
		  \item \textsc{Constraint -} A constraint prevents the system
		  from achieving more than its goal. With the definition of constraints, the
		  system  can be more robust, and unexpected problems can be solved before they
		  happen. For example, in a banking system, the customer can only withdraw
		  money if they have a positive balance in the account.
		    \item \textsc{Constraint Type -} Represents the types of constraints, that
		    may be on a system function or values. This entity can have their
		    property setted as a business and data (value) constraint (expressed as an
		    attribute). 
		  \item \textsc{Contract -} Is the formalization of obligations (requires) and
		  benefits (ensures) of a function, service activity or component.
		  The following questions can be used to define contracts: \textit{What does it
		  expect? What does it guarantee?} Contracts are crucial to software correctness
		  and therefore they should be part of the design process. An interface is a
		  kind of contract.
		  \item \textsc{Assertion -} Represents a predicate or a state of the
		  application before it runs (its preconditions), or the state when it is
		  finished running (post-conditions);
		  \item \textsc{Exceptional Behaviour -} Are alternative execution paths if
		  any condition or restriction is not respected. For example, if a user's
		  password is not correct after three attempts, the user account is locked for
		  security reasons.
		  \item \textsc{Policy -} A policy is a set of rules applied to a particular
		  scope. This scope can be defined as an action, an activity, a function or a
		  workflow. A policy is a composition of contracts applied to a non-functional
		  application requirement. For example, a security policy of a system constraint
		  may include authentication, access and data privacy properties.

		  \item \textsc{Rule -} Represents information about an event, condition, and
		  action where the event part represents the moment in which a constraint can be
		  evaluated according to a condition represented by the condition part and the
		  action to be executed for reinforcing it represented by the action part.
		  \item \textsc{Variable -} Represents a symbolic name given to some known
		  information. A variable are related with a Policy. A Policy can have one or
		  many variables. Each Variable has a name and a type.
	\end{itemize}
\end{itemize}

% 
% For example, in the \textit{$\pi$-UseCase} model (\textit{PIM} level) uses the
% {\sc Use Case\footnote{We use {\sc capitals} for referring to meta-models' classes.}},
% {\sc Constraint, Business Service, End Costumer} and {\sc NF Requirements}
% concepts to represent what is being modeled. {\sc Business Service} and
% {\sc End Costumer} are elements that model \textit{Business} concepts, while
% {\sc Use Case} models \textit{System} concepts and, {\sc Constraint} and {\sc
% NF Requirement} model \textit{Policy} concepts. The same is true for all other
% $\pi$SOD-M models, \textit{$\pi$- ServiceProcess,
% $\pi$-ServiceComposition} and \textit{$\pi$-PEWS}. The relationships
% among these $\pi$SOD-M concepts are also presented in figure
% \ref{fig:pisodm-concepts}, and especially the {\sc Policy} level for quality
% requirements modeling.
% 
% \bigskip
% \bigskip
% 
% Considering the set of concepts presented, each of them aims model
% important parts of the application. The services-based systems represent
% business relationships through {\sc Business Services},  in a more abstract
% level. {\sc Business Services} make up different application services, in order
% to perform a business relationship. Services are system features being
% modeled by {\sc Use Cases}. {\sc Use Cases} are refined at each
% development iteration, where s set of restrictions are
% specified for each function or service. System restrictions are represented
% through {\sc Constraints}, {\sc Contracts} or {\sc Policies}. {\sc Constraints}
% are restrictions on {\sc  Use Cases}, while {\sc Contracts} are restrictions on
% {\sc Actions}, or their specific functions, and {\sc Policy} is a restriction
% on a group of services that can be represented as {\sc Service Activity} or a
% {\sc Service Composite Process} . Thus, these terms specify particulars of the
% application at different modeling levels.
% 
% 
% Based on this, so that all the system features needed are specified, these
% concepts relate to each other in order to describe each use case, function,
% service or  application process. One {\sc Requirement}, whether functional or
% non-functional, represents by one or more {\sc Use Cases}. For example, a
% payment process can be modelled through the use cases that represent withdrawal
% and deposit of money in transaction of customer accounts. A payment process also
% requires the guarantee of a complete transaction and data security. Thus, some
% use cases can compose a single requirement.
% 
% A {\sc Composite Use Case} represents a type of {\sc Use Case}. This type of
% use case abstracts a more macro function, and may contain extra functions to be
% performed so that the process is complete. For example, to publish a song in a
% social network, you must choose which song you want to publish, authenticate, so
% that the music you're listening is published. We consider this type of use case,
% a use case compound.
% 
% Each {\sc Use Case} has many {\sc Constraints} (business or
% value constraints). Business constraints represents restrictions expressed on
% functions and how they may be implemented. Value constraints are restrictions on the
% service interface, which detail the desired values  for input and output
% data. Each constraint is associated with {\sc Non-Functional Attributes} (NFA).
% 
% A {\sc Contract} consists of a set of {\sc Constraints}. As a {\sc Contract}
% represents a composition of assertions, each {\sc Constraint} are
% refined into {\sc Assertion} that will compose one {\sc Contract}. For
% example, a contract for the payment operation by credit card, the value should
% not be less than 10 euros and the user should always receive a purchase
% confirmation by email or by phone message. This set of restrictions will be
% composed into a single contract for payment transactions. These constrains are
% also associated with non-functional attributes, e.g., \textit{reliability,
% transactions and privacy data}.
% 
% An {\sc Exceptional Behavior} represents a non excepted system behavior or
% a contract violation. When this happens other function is called
% or the application process is stopped. In $\pi$SOD-M, this entity represents the
% contract violation. For example, if the bank does not authorize the payment, the
% system or service offers alternative forms of payment such as a bank invoice or
% purchase via PayPal.
% 
% 
% Finally a {\sc Policy} groups similar {\sc Contracts}, \textit{e.g.}
% contracts for the application safety, or contracts on the system performance are
% grouped as policies. For example, payment and authentication contracts are
% grouped into \textit{security} policy, as well as contracts that require a
% time-out processing, are grouped into \textit{performance} policy. Each policy
% is associated with a non-functional requirement, a set of {\sc Rules} and
% {\sc Variables}. Each {\sc Non-Functional Requirement} is associated with
% one or more {\sc Non-Functional Attributes}.

 The following next sections will describe the example scenario to better
 explain the $\pi$SOD-M meta-model and the methodology concepts.

\subsection{Case Study}
\label{sec:example}

We use a case study in order to better describe specific details of the
methodology proposed in this work.

% We will specify the
% application of concepts and models in this case study so that each detail of the
% methodology is covered. We cover all $\pi$SOD-M features through examples at
% the same time in which the application is being developed, the concepts,
% activities and models of our methodology are explained.

Consider the following scenario: An organization
wants to provide the service-based application ``\textit{To
Publish Music}'' that monitors the music a person is listening
during some periods of time and sends the song title to this
person's Twitter or Facebook accounts. More specifically, this social
network user will have their status synchronized in both Twitter\footnote{Twitter is an online social networking service and micro-blogging service that enables its
users to send and read text-based posts of up to 140 characters, known as
``\textit{tweets}''.} and Facebook\footnote{Facebook is a social networking
service and website. Users may join common-interest user groups, organized by
workplace, school or college, or other characteristics, and categorize their
friends into lists}, with the title of the music the user is listening
in Spotify\footnote{Spotify is a music streaming service offering digitally
restricted streaming of selected music from a range of major and independent
record labels. Spotify allows registered users to integrate their account with
existing Facebook and Twitter accounts.}. The user may
also want to download music. For this, the user needs to process the purchase of music
they want to download. Payments will be processed via PayPal or credit card. All services
are available via Spotify and our application needs to interact with Spotify
users, so they can listen music, publish on Facebook and Twitter, and
purchase music online.  The corresponding use case model is illustrated in in
figure \ref{fig:example_usecase}.

% Lets consider the use case model illustrated in figure
% \ref{fig:example_usecase}.
%  Figure \ref{fig:example_usecase} shows the use case model for the ``\textit{To
% Publish Music}'' application.

\begin{figure}[ht!]
\centering
\includegraphics[width=.95\textwidth]{chapters/methodology/figs/runningExampleUCSingle.pdf}
\caption{Use Case Model ``\textit{To Publish Music}'' - Spotify Music Service.}
\label{fig:example_usecase}
\end{figure}

% As long as the user listens the music, they can
% publish the music information on social networks. For downloading music, it is
% necessary to purchase the song the user wants. To publish songs, the user must 
% login and authenticate on the social networks services, and for that there are
% constraints to be modelled and implemented. The same is true for purchasing
% music, where some integrity checks and confidentiality of the credit card
% data are required.

Figure \ref{fig:example_serviceprocess} shows the activity model for our
scenario. It starts by contacting the music service Spotify for retrieving the
user's musical status.  If the user wishes to download a particular song, they
can download the song to purchase (activities \textit{Buy Music} and
\textit{Download Music}). Finally the user can publish the music. The activity
\textit{publish music} can also be modelled as a parallel workflow for updating
different social networks, as follows: Twitter and Facebook services can be
contacted in parallel for updating the user's status with the corresponding song title.

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{chapters/methodology/figs/runningExampleSPSingle.pdf}
\caption{Service Process Model ``\textit{To Publish Music}''.}
\label{fig:example_serviceprocess}
\end{figure}

The execution flow is described as follows. Each action from the activity
diagram represents one or more functions in the application. Each function or
service has an interface, consisting of its input and output, and the
specification of how to call this function/service. An activity must
be executed according to the rules identified in the use case model. For example, the
``\textit{publish music}'' action (figure \ref{fig:example_serviceprocess}),
can be refined into two flows: (i) a flow to publish on Twitter and (ii) another to
update the Facebook, since each service has different rules for update and
login.

Given a set of services with their exported methods (known
in advance or provided by a service provider), building service-based
applications is a task that implies expressing an application logic as a service
composition together with some restrictions.

\begin{figure}[ht!] 
\centering
\includegraphics[width=.95\textwidth]{chapters/methodology/figs/runningExampleUCSingle_detail.pdf}
\caption{Traditional UML Concepts.}
\label{fig:umlConcepts} 
\end{figure}

Both models presented in figures \ref{fig:example_usecase} and \ref{fig:example_serviceprocess} use the
 traditional UML to design requirements in use case diagram and workflow actions
in the activity diagram. Figure \ref{fig:umlConcepts} presents
some of the main UML concepts that are used to model system diagrams. The
structure of each $\pi$SOD-M model will be detailed in accordance with their
respective meta-model and transformations necessary for achieving the aim of modeling
reliable applications.

% This example scenario will be used to exemplify the
% concepts of the $\pi$SOD-M methodology as well as to model the system properties from the
% use case to service composition model.

\section{Platform Independent Models}
\label{sec:pim-pisodm}

The \textit{Platform Independent Models (PIM)} are used for modeling the
system functionality, abstracting technological details.

% These models can be used on different deployment platforms. Since $\pi$SOD-M
% defines a method for developing service-oriented applications, such models focus on identifying the services to be
% offered by the system, as well as on identifying general functionalities,
% processes and constraints.

The models proposed at $\pi$SOD-M's PIM level are (figure
\ref{fig:developmentProcess}): \textit{ $\pi$-UseCase, $\pi$-ServiceProcess} and \textit{
$\pi$-ServiceComposition}. They provide concepts for expressing the design of:
(\textit{i}) system requirements and constraints; (\textit{ii}) external
services and their compositions; (\textit{iii}) execution flow; (\textit{iv})
input and output service restrictions; (\textit{v}) service contracts and
actions; and (\textit{vi}) policies for the implementation of business services.
We use these models to provide a representation of the system and to
specify its restrictions.

%Each of the three Platform Independent Models are presented next.

\subsection{\textit{$\pi$-UseCase} Model}

The \textit{$\pi$-UseCase} (figure \ref{fig:usecasemodel}) is used
to describe the functionalities (services and other functionalities) of the system. The
services are represented in this model as use cases. We design this model by
using the UML use cases, with a small extension for considering constraints.
Application constraints are represented by stereotyped use cases and comments
related to them. Each constraint describes what will be checked for a
particular application function be executed. Figure \ref{fig:piUseCaseExample}
shows\footnote{The annotation in this figure are not part of
the model. They are included for didactic explanation only. The same applies for
the other figures in this section which present the same type of annotation.} an
example of how to describe constraints on use cases in $\pi$SOD-M. We use part
of our case study to present them. Each use case may have multiple constraints.
In this example the \textit{publish music} use case has an authentication
constraint (\textit{user must authenticate}). For a song to be published on any
social network it is necessary to provide data privacy. The constraint type
concerns the authentication values and the constraint has an non-functional
attributes. The use case is also related with requirement, a business service
and a non-functional requirement properties.

% The main difference between the \textit{extended use case} diagram proposed by
% SOD-M\cite{valeriaThesis} and our \textit{$\pi$-UseCase} diagram is the
% inclusion of concepts that represent the modeling of non-functional requirements (\textit{policy view}).

\begin{figure}[ht!] 
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/piUseCase-Example_detail.pdf}
\caption{$\pi$-UseCase model Example.}
\label{fig:piUseCaseExample} 
\end{figure}

% This representation will be described in detail throughout this section, however
% the general \textit{$\pi$-UseCase} concepts are presented in figure
% \ref{fig:piUseCaseExample}. This representation shows only an overview of how
% to model system features and restrictions, and describe the concepts in this
% level of abstraction.
%

 \subsubsection{\textit{$\pi$-UseCase} Diagram, Terms and Concepts}

The \textit{$\pi$-UseCase} model describes the
functional aspects of the application being developed and the restrictions
that applies to each of them. These restrictions are modeled using the concepts:
(i) constraint type and (ii) non-functional requirement with its
specific attribute.

 A \textit{$\pi$-UseCase} model is a way to represent features of an
 application, services, constraints, and consumers of application functionality.

%   What differentiates
%  this model from others is its particular content and the form of their
%  presentation.
%   
%  This model is particularly essential for the organization and modeling of
%  application behavior. The behavior is represented by a normal and 
%  exceptional flows. In both cases there are restrictions that must be verified.
%  The restrictions may be on variables, values or function calls.

 \begin{figure}[ht!] 
\centering
\includegraphics[width=.4\textwidth]{chapters/methodology/figs/constraints_detail.pdf}
\caption{Constraint Representation.}
\label{fig:constraint_detail} 
\end{figure}

 All constraints are described by stereotyped use cases. Every relationship
 between an use case and a constraint is done through a dependency relationship
 (from UML). All use cases that are related with constraint must be
 represented such as described in figure \ref{fig:constraint_detail}. A
 constraint is defined by a stereotype, containing its \textit{type} and
 \textit{name}. The {\sc Constraints} are represented by a stereotyped use case
(\textit{<<constraint>>}).

 \begin{figure}[ht!] 
\centering
\includegraphics[width=.75\textwidth]{chapters/methodology/figs/nfr_detail.pdf}
\caption{Non-functional Requirement Representation.}
\label{fig:nfr_detail} 
\end{figure}

Each use case is related with one or more system requirements. Moreover, these
use cases are related to non-functional requirements necessary for the proper
execution of the application function. Similarly, the function described by an
case of use can be part of a business service (in the case of service remote
calls). These information to a specific use case may be described through
comments as described in figure \ref{fig:nfr_detail}. All these information
must be associated with the use case and its constraints.

 \begin{figure}[ht!] 
\centering
\includegraphics[width=.85\textwidth]{chapters/methodology/figs/nfa_detail.pdf}
\caption{Non-functional Attribute Representation.}
\label{fig:nfa_detail} 
\end{figure}

The constraints are always related to non-functional requirements. More
specifically, each constraint describes a non-functional attribute,
\textit{e.g.} \textit{authentication} (presented in figure
\ref{fig:nfa_detail}). A constraint consists of the constraint description and
the identification of variables be verified (by an \textit{@}, for example
\textit{@song}), the identification of the non-functional attribute and its
description.

% Considering the $\pi$SOD-M concepts, the \textit{$\pi$-UseCase} model
% describes the representation of system functionalities to be modelled in this
% early stage of the design. The system features and its interactions are
% presented using meta-model entities.

These concepts that are described in the $\pi$SOD-M  meta-model and used in
\textit{$\pi$-UseCase} represent the concepts we identified necessary
for reliable service-based modeling.

%  are presented in each abstraction level
% in order to better describe and refine properties of the applications that are
% being developed.
%  
%  \bigskip 
%  \textbf{Properties} 
%  \bigskip 
%

 \subsubsection{Meta-model}

The concepts modelled in the \textit{$\pi$-UseCase} meta-model are: {\sc
Business Service, End Consumer, Requirement, Use Case, Composite Use Case, Non-Functional Requirement,
Non-Functional Attribute} and {\sc Constraint}.  Figure
\ref{fig:usecasemodel} presents the \textit{$\pi$-UseCase} meta-model and the
relationship between each concept. We highlight the elements that represent the
\textit{policy view}.

% The \textit{$\pi$-UseCase} elements aim to describe system properties and system
% functions in a abstract way, with a coarser granularity. Without considering any
% application details, they specify the general features of the system and its
% restrictions, as services or other functions. For example, i

In the \textit{$\pi$-UseCase} an {\sc End Consumer} is represented by an {\sc
Actor} (figure \ref{fig:usecasemodel}). An {\sc Actor} is related to {\sc Use Cases}
(from the original UML definition), while a {\sc Composite Use Case} is a set of
actions performed by the system which can be broken into different {\sc Use Cases}.

{\sc Business Service} aggregates several {\sc Use Cases} (figure
\ref{fig:usecasemodel}), simply put, a service can be expressed by one or more
use cases. Notice that not always an use case will be part of a business
service. However, using our approach, we can identify in the early stages of
modeling, which use cases are related to business services.

\begin{figure}[ht!]
\centering
\includegraphics[width=1.0\textwidth]{chapters/methodology/figs/PiUseCaseMetamodel.pdf}
\caption{$\pi$-UseCase Concepts (Meta-model).}
\label{fig:usecasemodel}
\end{figure}

The {\sc Business Collaborators} concept are represented in the
\textit{$\pi$-UseCase} model through {\sc Packages}. A {\sc Business
Collaborator} represents an external service or a system that interact
with the application that are being modelled. Each {\sc Business Collaborators}
combines the features described in each {\sc Package}. Thus, the
services functions can be specified being grouped into packages (figure
\ref{fig:businessCollaborator}).

\begin{figure}[ht!]
\centering
\includegraphics[width=.75\textwidth]{chapters/methodology/figs/businessCollaborator.pdf}
\caption{Business Collaborator / Package Representation.}
\label{fig:businessCollaborator}
\end{figure}

The {\sc Non-Functional Requirement} and {\sc Non-Functional Attribute} concepts
are represented as part of description of {\sc Use cases} and {\sc Constraints}.
For model this concepts the designer will use special tags to represent them  (figures
\ref{fig:nfr_detail} and \ref{fig:nfa_detail}).

% In summary, the main concepts to
% be modelled at this level are: {\sc Constraints}, and its respective {\sc
% Constraint Type}, and all {\sc Non-Functional Attribute} and {\sc Non-Functional
% Requirement} associated to each {\sc Constraint}. These concepts are not
% modelled in the original SOD-M extended use case model. Thus, we add these
% constraint concepts in the \textit{$\pi$-UseCase} model.

An {\sc Use Case} may have several {\sc Constraints}. Each
{\sc Constraint} has its name, description, and which ones should be
checked during the execution of the application. Each {\sc Constraint} is
represented as a stereotyped (\texttt{<<constraint>>}) use case. A
restriction may be a restriction on data ({\sc Value Constraint}),
represented as the stereotype \texttt{<<value>>}  (figure
\ref{fig:constraint_detail}); or on business rules (\textit{Business Constraint}), represented as the stereotype
\texttt{<<business>>}; and {\sc Exceptional Behaviour} constraints, which
represents constraint violation for the use case execution, represented as the stereotype
\texttt{<<exceptional\_behaviour>>}.

The \textit{$\pi$-UseCase} meta-model
also represents the use of include and extend relations among the different use
cases identified. The semantics are the same as in the traditional UML use case
model. An include relation specifies the existence of a flow of events in which
the base use case includes the behaviour of the other use case and an extend
relation specifies that the behaviour of a base use case can be optionally
extended by the behaviour of another use case.

 \subsubsection{UML Concepts Representation}

\begin{figure}[ht!]
\centering
\includegraphics[width=0.65\textwidth]{chapters/methodology/figs/umlRepresentation.pdf}
\caption{\textit{$\pi$-UseCase} Concepts Represented as UML Elements.}
\label{fig:umlrepresentation}
\end{figure}

Once described the meaning of each \textit{$\pi$-UseCase} concept in the
meta-model, it is important to know how specific models can be
created from this general representation. Every \textit{$\pi$-UseCase} model
created to describe application features must follow the
meta-model concepts. The \textit{$\pi$-UseCase} meta-model describes what can be
specified in each model.

\begin{figure}[ht!]
  \centering
  \subfloat[\textit{Use Case} and \textit{Actor} Model]
  {\label{fig:piusecase1}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piusecase/pi_uc_usecase-actor_final}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  ~
  \\
   \subfloat[\textit{Extended Use Case} Model]
  {\label{fig:piusecase2}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piusecase/pi_uc_extend_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
 \subfloat[\textit{Include Use Case} Model]
  {\label{fig:piusecase3}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piusecase/pi_uc_include_final}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  ~
  \caption{\textit{$\pi$-UseCase} Model Representation.}
  \label{fig:piuse_representation1}
\end{figure}

\begin{figure}[ht!]
  \centering
  \subfloat[\textit{Package} Model]
  {\label{fig:piusecase4}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piusecase/pi_uc_package_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[\textit{Use Case}, \textit{Requirement} and
  \textit{Non-Functional Requirement} Model]
  {\label{fig:piusecase5}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piusecase/pi_uc_usecase-comments_final}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  ~
  \\
  \subfloat[\textit{Constraint} Model]
  {\label{fig:piusecase6}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piusecase/pi_uc_constraint_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)

  \caption{\textit{$\pi$-UseCase} Model Representation (2).}
  \label{fig:piuse_representation2}
\end{figure}

The \textit{$\pi$-UseCase} concepts are modelled as described in
figure \ref{fig:umlrepresentation} and the relationship between these concepts
is also applied to the UML elements. Figures \ref{fig:piuse_representation1} and
\ref{fig:piuse_representation2} detail how to represent the relationship
between the meta-model concepts in a \textit{$\pi$-UseCase} model.

The \textit{loop arrow} symbol ($\looparrowright$) is used in figures
\ref{fig:piuse_representation1} and \ref{fig:piuse_representation2} to
relate the concepts of the meta-model and the UML elements used in the
representation of the concepts.

\correctingText{This means how the \textit{$\pi$-UseCase} meta-model concepts (left
side) can be modelled as an UML model (right side).}

An {\sc Actor} is an abstraction of the {\sc End
Costumer} concept, presented in $\pi$SOD-M meta-model. Thus, it is possible to
associate several {\sc Actors} with an {\sc Use Case} and many {\sc Use Cases} with an {\sc
Actor} ({\sc End Costumer}). The original use case meta-model defines
that {\sc Classifier} is a superclass of {\sc Actor}. A {\sc Classifier} is a
category of UML elements that have some common features, such as
\textit{attributes} or \textit{methods}. The relationship between these entities
is many to many (figure \ref{fig:piusecase1}). An  {\sc Use Case} can also be
related with an {\sc Extends} or an {\sc Include} entity in the \textit{$\pi$-UseCase} model, and
this model preserve the original UML standard semantic. {\sc
Extends} and {\sc Include} are types of stereotyped dependence relationships
(\texttt{<<extend>>} and \texttt{<<include>>} in figures \ref{fig:piusecase2}
and \ref{fig:piusecase3}, respectively). The {\sc Extends} relationship can modify the behaviour of the base use case. Suppose someone
wants to close a bank account, however still have some money there. Before
closing the account, there should be a withdrawal or transfer of money in this
account, so that the balance is zero. This process is modelled with a {\sc
Extends} dependence between the \textit{close account} use case and
the \textit{withdrawal money} and \textit{transfer money} use cases. The {\sc Include} relationship includes
the steps from one use case into another.

A {\sc Package} clause and its relationship with {\sc Use Case} and {\sc Actor}
 can be modelled in \textit{$\pi$-UseCase} model. The original semantics was
 preserved and it is possible have different levels of packages, due to  {\sc
 Package} entity auto reference, one to many. Each package can have many use
 cases and actors (\ref{fig:piusecase4}).

The {\sc Requirements}, {\sc Non-Functional Requirements} and {\sc Business
Services} must be associated with a specific {\sc Use Case}. At this level of modelling, these information are represented as
comment \textit{tags}, they are: \textit{\$ \ldots \$} for {\sc Requirements},
\textit{\# \ldots \#} for {\sc Non-Functional Requirements}, and \textit{\%
\ldots \%} for {\sc Business Service}. A {\sc
Business Service} is a aggregation of {\sc Use Case}, and each {\sc Use Case}
must be related with a system requirement {\sc Requirements}. Figure
\ref{fig:piusecase5} presents the relationship of this concepts, and how
represent this information in UML.

All {\sc Use Cases} may also be associated with a set of {\sc Constraints}. A
{\sc Constraint} is a stereotyped use case that describes the {\sc Constraint Type} and its
restriction. Additional information is given by comment tags, \textit{!
\ldots !} for {\sc Non-Functional Attribute}, and \textit{@
\ldots @} for candidate variables or values that must be checked during the
system execution, as presented in figure \ref{fig:piusecase6}. The relationship
between {\sc Use Cases} and  {\sc Constraint} is made by a non-stereotyped
dependence arrow.

\subsubsection{\textit{To Publish Music} Use Case}

Considering the example scenario, the ``\textit{To Publish Music}'' use case can
update the Facebook or Twitter user status. Therefore, it is necessary to
perform a social network authentication with the user's data. Each social
network uses different services and different forms of authentication. The
\textit{authentication} constraint is required to update a music status. The
restriction is stereotyped as a \texttt{<<value>>} constraint, because the
\textit{user's id} and \textit{password} are verified. The {\sc Non-Functional
Requirement} that is being modelled for this use case is \textit{Security},
because, considering  table \ref{tab:result04}, ``\textit{authentication}'' is
a {\sc Non-Functional Attribute} of \textit{Security}. These information will be
refined and detailed in the following phases.

% 
% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.9\textwidth]{chapters/methodology/figs/piusecase/application/packages}
% \caption{$\pi$-UseCase Scenario Example.}
% \label{fig:usecasemodelpackages}
% \end{figure}

Figure \ref{fig:usecaseContraint} shows
the model represented in figure \ref{fig:example_usecase} for the \textit{buy
music, download music, listen music} and \textit{pay} use cases. The model was
designed considering the meta-model representation detailed in figures
\ref{fig:piuse_representation1} and \ref{fig:piuse_representation2}, and also
the \textit{$\pi$-UseCase} meta-model. Notice that from  \textit{$\pi$-UseCase}
concepts is possible to describe the actors, use cases, constraints,
non-functional requirements and their attributes. For example, the \textit{pay}
use case (figure \ref{fig:usecaseContraint}), there are extended use cases, constraints, packages, non-functional requirements and non-functional
attributes related to this function. The modeling was based on the described
representation of the \textit{$\pi$-UseCase} meta-model. The extended use case
representation presented in figure \ref{fig:piusecase2} is used in the model to
expressed the \textit{pay by PayPal} and \textit{pay by card} extended use
cases. The include use case representation presented in figure \ref{fig:piusecase3} is
used in the model to express the included relation of \textit{buy music} and
\textit{pay} use cases. The package representation presented in figure
\ref{fig:piusecase4} is expressed in the example by the \textit{app.bank}
package and the \textit{pay} use case relation. The representation described in figures
\ref{fig:piusecase5} and \ref{fig:piusecase6} is expressed by the \textit{pay}
use case properties' description. The combination of a \textit{$\pi$-UseCase}
meta-model's entities means that all entities presented in figures
\ref{fig:piUseCaseExample} and \ref{fig:usecaseContraint} can be modelled to
express the quality requirements.

The use cases that have restrictions are \textit{buy music} and
\textit{pay}. The process of buying a song requires the user's private data
for a Spotify account, and also a secure connection, represented as
\texttt{<<value>>} and \texttt{<<business>>} constraint, respectively. For
payment, the user must provide the data of payment card or PayPal account login
and password, represented as \texttt{<<value>>} stereotype . Other restriction
is that the minimum payment value is 2 euros. The {\sc Non-Functional
Requirement} that are being modelled for this use cases are \textit{Security}
and \textit{Reliability}. The specific {\sc Non-Functional Attributes} for this
restriction is \textit{Transaction} and \textit{Confidentiality} (see figure
\ref{fig:usecaseContraint}).

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/piusecase/application/packages2.pdf}
\caption{Use Case Model With Constraint Definition - Scenario Example Refined.}
\label{fig:usecaseContraint}
\end{figure}

% Each {\sc Constraint} may be related to {\sc Non-Functional
% Attributes}. These types are the attributes related to {\sc Non-Functional
% Requirement}. For example, \textit{security} is an {\sc Non-Functional Requirement} that
% comprises attributes such as \textit{Confidentiality} and \textit{Privacy}
% ({\sc Non-Functional Attribute}). Thus the use case constraints are always
% related to a {\sc Non-Functional Requirement} (figure
% \ref{fig:usecasemodel}).

\bigskip
\bigskip

After modeling the system services and features, and its
restrictions, it is necessary to model the services process and its
execution restrictions. For this, the methodology uses a service process diagram
to describe the services execution and contracts. Next we present the
\textit{$\pi$-ServiceProcess} model. This model refines the application
constraints through the workflow model \correctingText{(as presented in Figure
\ref{fig:developmentProcess})}.

\subsection{\textit{$\pi$-ServiceProcess} Model}

A \textit{Service Process} is a collection of related, structured activities that produce a specific service or product for
a particular customer. It can be visualized with a flowchart
as a sequence of activities with interleaving decision points.  A set of
linked activities that take an input and transform it to create an output.
Ideally, the transformation that occurs in the process should add value to the
input and create an output that is more useful and effective. To represent these
properties our methodology proposes the \textit{$\pi$-ServiceProcess} model.

% The main difference between the service process diagram proposed by SOD-M
% \cite{valeriaThesis} and our \textit{$\pi$-ServiceProcess}  diagram is the
% inclusion of concepts that represent the modeling of non-functional
% requirements, as well as in \textit{$\pi$-UseCase} model. 
The non-functional requirements are specified in this model using the concepts
of contracts and assertions over functions. These contracts may represent the
control of privacy information, performance, reliability, and so on.

 \subsubsection{\textit{$\pi$-ServiceProcess} Diagram, Terms and Concepts}

% In the context of reliable service composition development, the
% \textit{$\pi$-ServiceProcess} perform different third party functions services,
% so that something new can be offered.

Systems that use third party services add value to the application because the
services are already available and validated. If the services do not offer any
quality warranty, these quality requirements are the responsibility of the
application designer. Thus, the contracts of services will be modeled as a
wrapper which describes restrictions on input and output information on
particular functions or a set of them.

$\pi$SOD-M proposes modeling \textit{Service Process} diagram
(\textit{$\pi$-ServiceProcess} model), that is defined considering the
functional and non-functional requirements designed in a
\textit{$\pi$-UseCase} model.

The \textit{$\pi$-ServiceProcess} model represents the service processes and its
contracts. It models the set of system activities and describe contracts for
each activity or the process as a whole. Each service identified in the previous
model (\textit{$\pi$-UseCase}) is mapped to an action.

% Figure \ref{fig:serviceprocess} shows
% the \textit{$\pi$ServiceProcess} meta-model proposed by $\pi$SOD-M method for
% model \textit{Service Process}.

%  \bigskip 
%  \textbf{Properties} 
%  \bigskip

%  What differentiates this model
%  from others is its particular content and how to represent restrictions to be
%  verified during the workflow execution.
%  
  \begin{figure}[ht!]
\centering
\includegraphics[width=.8\textwidth]{chapters/methodology/figs/assertion_detail.pdf}
\caption{Assertion Representation.}
\label{fig:assertion_representation}
\end{figure}

 A \textit{$\pi$-ServiceProcess} model represents the system process,
 actions, assertions and contracts. This model describes the workflow for the modelled application
 behaviour. The behaviour is represented by each action and the restrictions by
 the assertion description. All assertions are described by stereotyped actions.

   \begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/contract_detail.pdf}
\caption{Contract Representation.}
\label{fig:contract_representation}
\end{figure}

Figures \ref{fig:assertion_representation} and \ref{fig:contract_representation}
present the assertion and contract representation in this
model. An Assertion is represented in UML, and a set of assertions forms a
contract. Assertions describe restrictions on a single action, but actions may
have different restrictions. Figure \ref{fig:assertion_representation} shows how an assertion is modelled, with their property type, name and acceptable values (expressed through
expressions), while figure \ref{fig:contract_representation} shows how this set
of assertions form a single action contract.

   \begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/otherObjects_detail.pdf}
\caption{Node, Activity Edge and Contract Representation.}
\label{fig:other_representation}
\end{figure}

Figure \ref{fig:other_representation} shows the representation of the general
concepts of this model, nodes, activity edges, actions and contracts. This figure also refines the information
for publish a song in a social network.

 \subsubsection{Meta-model}

Considering the $\pi$SOD-M concepts (figure \ref{fig:pisodm-concepts}), the
\textit{$\pi$-ServiceProcess} meta-model (figure \ref{fig:serviceprocess})
describes the system functionalities to be modelled in a contract based service
process. The $\pi$SOD-M concepts modelled in the \textit{$\pi$-ServiceProcess}
meta-model are: {\sc Contract, Assertion, Exceptional Behaviour, Activity,
Service Activity, Action} and {\sc Constraint}.

Restrictions on activities and services are not described in the
original SOD-M proposal. The concepts of {\sc Contract, Assertion} and
{\sc Exceptional Behaviour} make it possible to model restrictions on services
execution, ensuring greater reliability application. Entities highlighted in
\textit{$\pi$-ServiceProcess} meta-model (figure \ref{fig:serviceprocess})  are
the difference from the original SOD-M service process model.

 \begin{figure}[ht!]
\centering
\includegraphics[width=1.0\textwidth]{chapters/methodology/figs/PiServiceProcess.pdf}
\caption{$\pi$-ServiceProcess Concepts (Meta-model).}
\label{fig:serviceprocess}
\end{figure}

A specific \textit{$\pi$-ServiceProcess} model shows the set of logically
related activities that need to be performed in a service-based system. So, the
activities of this model represent a behaviour that is part of the
system workflow. $\pi$SOD-M proposes representing this model using
the UML activity diagram, with an extension for {\sc Contract} design,
which encapsulates several {\sc Constraints} modelled in \textit{$\pi$-UseCase}
model. In \textit{$\pi$-ServiceProcess}, three main features are
represented: (i) \textit{service process}, (ii) \textit{service activity} and
(iii) \textit{activity contract}. The service process is represented by the
model itself. A service activity represents a behaviour that is part of the
execution flow, and is identified in this model as an {\sc Action}. A activity
contract represents the {\sc Non-Functional Requirement} behaviour that is also
part of the execution flow of a service, and is identified in this model as an
stereotyped activity (\texttt{<<assertion>>}). All {\sc Assertion} related with
an {\sc Action} compose a {\sc Contract}. From the {\sc Contract} and {\sc
Assertion} concepts is possible to define the interface specifications
for each activity service, such as pre-conditions and post-conditions.

\subsubsection{UML Concepts Representation}

To formalize the concept of {\sc Contract} by rules, we use the definition of
{\sc Assertion} (figure \ref{fig:pisp1}). An {\sc Assertion} is represented by a
stereotyped action (\texttt{<<assertion>>}), and {\sc Assertion Properties}
through comments. An {\sc Assertion} may be a \textit{pre-condition},
\textit{post-condition} or a \textit{time restriction} over the service
execution ({\sc Assertion Property}). These information are represented as
the comment \textit{tags} \textit{\$ \ldots \$}. The assertion predicate is
modelled as \textit{\# \ldots \#} tag. Figure \ref{fig:pisp1} shows how these
concepts are represented in UML.

\begin{figure}[ht!]
  \centering
  \subfloat[\textit{Assertion} Model]
  {\label{fig:pisp1}\includegraphics[width=0.5\textwidth]{chapters/methodology/figs/piserviceprocess/assertion_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \subfloat[\textit{Contract} Model]
  {\label{fig:pisp2}\includegraphics[width=0.5\textwidth]{chapters/methodology/figs/piserviceprocess/contract_final}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[\textit{Exceptional Behaviour} Model]
  {\label{fig:pisp3}\includegraphics[width=0.5\textwidth]{chapters/methodology/figs/piserviceprocess/exceptionalBeh_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \subfloat[\textit{Action} Model]
  {\label{fig:pisp4}\includegraphics[width=0.5\textwidth]{chapters/methodology/figs/piserviceprocess/action_final}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[\textit{Service Activity} Model]
  {\label{fig:pisp5}\includegraphics[width=0.5\textwidth]{chapters/methodology/figs/piserviceprocess/serviceActivity_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)

  \caption{\textit{$\pi$-ServiceProcess} Model Representation.}
  \label{fig:pisp_representation}
\end{figure}

\begin{figure}[ht!]
  \centering
  \subfloat[\textit{Control Nodes} Model]
  {\label{fig:pisp6}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piserviceprocess/controlNode_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[\textit{Activity Edge - Control Flow} Model]
  {\label{fig:pisp7}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piserviceprocess/activityEdge_final}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  ~
  \\
  \subfloat[\textit{Activity Edge - Control Object} Model]
  {\label{fig:pisp8}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piserviceprocess/activityEdge2_final}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)

  \caption{\textit{$\pi$-ServiceProcess} Model Representation (2).}
  \label{fig:pisp_representation2}
\end{figure}

A set of {\sc Assertion} composes a service {\sc Contract} related with an {\sc
Action} (figure \ref{fig:pisp2}).  An Assertion describes a value (\textit{value
constraint} from $\pi$-UseCase model) restriction, setting maximum and minimum
limits, or business restrictions, setting conditions for function calls
((\textit{business constraint} from $\pi$-UseCase model)). The
business restrictions are defined over service functions. Every service contract
may have an exception behaviour, if a constraint is not obeyed. For example, if the
user card information are incorrect, the service is called again until the limit
of 3 calls, otherwise, execution continues normally. In this case, the exception
is the new call to the same service. For this, we also use the concept of {\sc
Exceptional Behaviour} (figure \ref{fig:pisp3}). A {\sc Exceptional Behaviour}
is also a stereotyped action (\texttt{<<exceptional behaviour>>}) and its
properties are modelled through comment tags. The condition to be verified is
modelled as \textit{\# \ldots \#} tag. Figure \ref{fig:pisp1} shows how these
concepts are represented in UML. If the condition is \textit{false}, the
predicate becomes true, because of the expression negation
(\textit{not}). Thus the action is performed. The action to be performed is
modelled as \textit{! \ldots !} tag. The {\sc Action} concept is modelled in UML
as the action component (figure \ref{fig:pisp4}), and a {\sc Service Activity}
represents a set of {\sc Action} (figure \ref{fig:pisp5}). 
We highlight the models in figures \ref{fig:pisp1}, \ref{fig:pisp2} and
\ref{fig:pisp3}, because they present how an assertion action, a contract, and a
exceptional behaviour can be designed.

Figure\ref{fig:pisp_representation2}
shows the UML representation of the {\sc Control Nodes} and {\sc Activity Edges}
concepts in $\pi$-ServiceProcess meta-model. This representation follows the
original UML specification. Figure \ref{fig:pisp6} presents the
{\sc Control Nodes} and how its can be applied in a real modelling, and finally
figures \ref{fig:pisp7} and \ref{fig:pisp8} show the {\sc Activity Edges} and
its application.

% These model concepts is very important, because it is from
% the restrictions described in the previous model ($\pi$-UseCase) that each
% constraint begins to be refined and applied to real application services.

\subsubsection{\textit{To Publish Music} Process}

Considering the example scenario, the contract based process of activities
execution is shown in figure \ref{fig:serviceprocessContract}.
The activities flow in figure \ref{fig:example_serviceprocess} presents any restriction. Thus, it is necessary
to represent each {\sc Constraint} modelled in a more representative {\sc
Contract} based service activity diagram.

Based on the concepts described and its representation in UML, we describe how
to model a service process using the assertions as restrictions for each action.
The difference of the model shown in figure \ref{fig:example_serviceprocess}
and the figure \ref{fig:serviceprocessContract} is that restrictions are added
on the actions in order to describe quality requirements.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/piserviceprocess/runningExampleSP_final.pdf}
\caption{Service Process Model with Contract Definition - Complete Example.}
\label{fig:serviceprocessContract}
\end{figure}

% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.89\textwidth]{chapters/methodology/figs/piserviceprocess/runningExampleSP_part.pdf}
% \caption{Service Process Model with Contract Definition}
% \label{fig:serviceprocessContractPart}
% \end{figure}

Each {\sc Use Case} modelled in the use case diagram are refined into {\sc
Actions} in the service process diagram. The set of {\sc Constraints} are
transformed into {\sc Assertions} and {\sc Contracts}. The \textit{buy music}
and \textit{publish music} services (update Twitter and Facebook) has pre- and
post-conditions assertions that are composed into a contract for each service. 
 The  \textit{buy music} pre-conditions are: (i) verify if the User
data are correct; (ii) if the User is already logged in Spotify; (iii) if bank
account information are correct and; (iv) if there is enough money to make the
payment. As post-condition, it ensures the complete transaction and verify if a
notification were sent to the user and Spotify, about the payment authorization.
There are four assertions for the \textit{buy music} action, and each assertion
has been detailed with the assertion property and predicate that must be verified. 
To \textit{update} services, depending of each service, there may be different
restrictions. As an example, a new verification of user data and message format
is appropriate (maximum 140 characters), in case of Twitter. In the case of Facebook, it is required that
the user is already logged on Spotify and these data are the same as
Facebook. As post-condition, it ensures that the Facebook service send a
notification of success. To update Twitter a
pre-condition is required, while to update Facebook is necessary
a pre-condition and a confirmation notice is modelled as post-condition. As a
pre-condition for ``\textit{twitter update}'' it is necessary that \textit{(i)}
the music format be correct and (\textit{ii})  the twitter login and password be
correct for the update.

\bigskip 
\bigskip

After modeling of service process, its features, and the
service contracts, it is necessary to model the service composition process and
its policies. $\pi$SOD-M uses a service composition diagram
to describe the third party services and each policy allied to
each composition \correctingText{(as presented in Figure
\ref{fig:developmentProcess})}. In the next section we present this model and their
respective policy concept .

\subsection{$\pi$-ServiceComposition Model}

The service composition model represents the workflow needed to execute a
service or a set of them, but in a more detailed way, \textit{i.e.}, identifying
those entities that collaborate with the service process workflow, performing
services that are necessary to execute the actions (in the business
collaborators). Considering the workflow described in the
\textit{$\pi$-ServiceProcess} diagram, this model identifies the services and
functions that correspond to each action in the business process.

Each action describes a fundamental behaviour unit that is part of a service
activity and represents some behaviour in the system being
modelled. The \textit{$\pi$-ServiceComposition} model refines the concepts
described in the \textit{$\pi$-ServiceProcess} model, grouping contracts
into policy (figure \ref{fig:serviceprocess}).

% The main difference between the \textit{service composition} diagram proposed by
% SOD-M \cite{valeriaThesis} and our \textit{$\pi$-ServiceComposition}  model is
% the inclusion of concepts that represent policy and rules related with service
% compositions.

\subsubsection{\textit{$\pi$-ServiceComposition} Diagram, Terms and Concepts}

% \bigskip
%  \textbf{Properties} 
%  \bigskip 
%  
 A \textit{$\pi$-ServiceComposition} model represents service
 compositions and their related business collaborators (external services),
 policies, rules associated with a policy and the whole application
 process and functionalities.

%  What differentiates this model from the
%  \textit{$\pi$-ServiceProcess} model is its particular content and the form of
%  their presentation.
%  
 This model is essential for the organization and modeling of
 services and their location. We refine the representation of packages
 described in the  \textit{$\pi$-UseCase} model into business
 collaborator. The set of use cases described in a package
 (\textit{$\pi$-UseCase} model) will be represented by an external service
 functions.

 Figure \ref{fig:businesscollaborator_detail} shows how to relate the actions
 described in the \textit{$\pi$-ServiceProcess} model with services of a specific business
 collaborator. A service is represented by an stereotyped action 
 (\textit{<<service>>}), and their connection is made by an edge (\textit{Object
 Flow}). Thus each action described in the application workflow may be
 replaced by a service or a composition of them. If there is no relationship
 between the actions described in the application workflow (\textit{External
 false} business collaborator) with external services, means that function
 is native to the application being developed.

\begin{figure}[ht!] 
\centering
\includegraphics[width=0.99\textwidth]{chapters/methodology/figs/serviceComposition_detail}
\caption{Business Collaborator Representation.}
\label{fig:businesscollaborator_detail}
\end{figure}

\begin{figure}[ht!] 
\centering
\includegraphics[width=0.99\textwidth]{chapters/methodology/figs/policy_detail}
\caption{Policy Representation.}
\label{fig:policy_detail}
\end{figure}

The main properties described in the  \textit{$\pi$-ServiceComposition} model
for the ``\textit{to publish music}'' example are: (i) description of workflow
application (\textit{External false}), where the business collaborator \textit{Application} points out that it is not an
external workflow, (ii) the external services (\textit{business collaborators})
and (iii) the application policies. Figures
\ref{fig:businesscollaborator_detail} and \ref{fig:policy_detail} presents how
to model the \textit{$\pi$-ServiceComposition} concepts (figure
\ref{fig:servicecomposition}).

Figure \ref{fig:businesscollaborator_detail} shows how each action is connected to an external service
(through \textit{Activity Edge - Object Flow}), and how to define the
description of each business process, while Figure  \ref{fig:policy_detail} shows each
policy property and how it should be presented in the model.

Each business collaborator encompasses a set of services that can
be performed by the application. These external services are used by the 
application must be described in this model, and not in the
\textit{$\pi$-ServiceProcess} model. The contracts described for each action are
refined into policies, considering the non-functional requirement of each
contract.

 \subsubsection{Meta-model}

Considering the $\pi$SOD-M concepts (figure \ref{fig:pisodm-concepts}), the
\textit{$\pi$-ServiceComposition} meta-model (figure
\ref{fig:servicecomposition}) describes the service compositions
to be modelled and the policies related with each service composition.
The $\pi$SOD-M concepts modelled in the \textit{$\pi$-ServiceComposition}
meta-model are: {\sc Policy, Rule, Variable, Exceptional Behaviour, Service
Activity, Action, Business Collaborator} and {\sc Non-Functional
Requirement}.

 $\pi$SOD-M proposes representing this model using the a extension of UML
 activity diagram technique. This extension describes the relationship of each {\sc Action} for the
respective {\sc Service Activities}. Thus, as shown in figure
\ref{fig:servicecomposition}, the meta-model includes typical modeling elements
of the activity diagram such as {\sc Activity Nodes}, {\sc Initial
Nodes} and {\sc Final Nodes}, {\sc Decision Nodes}, etc., along with
new elements defined by $\pi$SOD-M such as {\sc Business Collaborator},
{\sc Service Activity} and {\sc Action} (see
figures \ref{fig:servicecomposition} and \ref{fig:servicecompositionPolicy}).

All \textit{$\pi$-ServiceComposition} models
describe the set of services and its relation with the service process and its
activities that need to be performed. So, the external services
are represented by the {\sc Business Collaborator} concept. All service
composition may specify a {\sc Policy} as restriction. A {\sc Policy} is the
set of {\sc Contracts} related with the service process expressed in the
previous model (\textit{$\pi$-ServiceProcess}).

Policy over service activities and business collaborators are not described in
the original SOD-M method. The rule concept grouped into policy and
exceptional behaviour make possible to model restrictions on service composition. Entities
highlighted in the \textit{$\pi$-ServiceComposition} (figure
\ref{fig:servicecomposition}) meta-model are the difference from the original
SOD-M service composition model.

\begin{figure}[ht!]
\centering
\includegraphics[width=1.0\textwidth]{chapters/methodology/figs/PiServiceComposition}
\caption{\textit{$\pi$-ServiceComposition} Concepts (Meta-model).}
\label{fig:servicecomposition}
\end{figure}

A {\sc Business Collaborator} element represents those
entities that collaborate in the business processes by
performing some of the required actions. They are graphically
presented as a partition in the activity diagram.
A collaborator can be either internal or external to the
system being modelled. When the collaborator of the
business is external to the system, the attribute \textit{IsExternal} of the
collaborator is set to \textit{true}.

{\sc Action}, a kind of {\sc Executable Node}, are represented
in the model as an activity. Each action identified in
the model describes a fundamental behaviour unit which
represents some type of transformation or processing
in the system being modelled. There are two types of
actions: i) a Web Service (attribute Type is \textit{WS}); and ii) a
simple operation that is not supported by a Web Service,
called an {\sc Activity Operation} (attribute Type is \textit{AOP}).

The {\sc Service Activity} element is a composed activity
that must be carried out as part of a business service and
is composed of one or more executable nodes.

The policy based service composition model refines the concept contract of
service process model. A policy assemble a set of contracts and can be
applied to more than one activity. The restriction on service composition model
is marked with the stereotype \texttt{<<policy>>}.

An {\em Policy} groups a set of rules. It describes global variables and
operations that can be shared by the rules and that can be used for expressing
their Event  and Condition parts. An {\em Policy} is associated to the
elements {\sc Business Collaborator}, {\sc Service Activity} and, {\sc Action}  of
the $\pi$-ServiceComposition meta-model (see figure
\ref{fig:servicecomposition}) .

% Instead of programming different protocols within the application logic, we
% propose to include the modeling of non-functional requirements like
% transactional behaviour, security and adaptability at the early stages of the
% services' composition engineering process.

 \subsubsection{UML Concepts Representation}

The representation described in figure \ref{fig:picomposition_representation}
presents how the user of the methodology's user apply the concepts described in
the \textit{$\pi$-ServiceComposition} meta-model. In the specific case
of \textit{$\pi$-ServiceComposition} meta-model, those concepts will be used to
describe how {\sc Business Collaborator, Action, Service Activity, Policy and
Rules} can be specified. Figure \ref{fig:picomposition_representation}
also describes how the proposed concepts can be modelled in real system
development cases. All \textit{$\pi$-ServiceComposition} based models
(presented in figure \ref{fig:picomposition_representation}) must follow the
meta-model description and its constraints.

\begin{figure}[ht!]
  \centering
  \subfloat[\textit{Service Activity} and \textit{Business Collaborator}  Model]
  {\label{fig:pisc1}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piservicecomposition/serviceComposition1}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[\textit{Policy} and \textit{Rule}  Model]
  {\label{fig:pisc2}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piservicecomposition/serviceComposition3}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  ~
  \\
  \subfloat[\textit{Business Collaborator - External false} Model]
  {\label{fig:pisc3}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piservicecomposition/serviceComposition2}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[\textit{Business Collaborator - External true} Model]
  {\label{fig:pisc4}\includegraphics[width=0.8\textwidth]{chapters/methodology/figs/piservicecomposition/serviceComposition4}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \caption{\textit{$\pi$-ServiceComposition} Representation Models.}
  \label{fig:picomposition_representation}
\end{figure}

Figure \ref{fig:pisc1} presents the representation for the relation between
{\sc Action} and {\sc Business Collaborator}. As a {\sc Service Activity} are
modelled as a set of {\sc Actions} (service functions), they are represented as
\textit{<<External true>>} {\sc Business Collaborator}. It means that an
external service is invoked by the application's function. Each {\sc Business
Collaborator} can represents an external service. Figure \ref{fig:pisc2}
presents the representation for the relation between {\sc Action}, {\sc Policy} and {\sc
Rule}. Figure \ref{fig:pisc3} presents the representation of the original
\textit{Service Process} associated to the \textit{<<External false>>} {\sc
Business Collaborator} and figure \ref{fig:pisc4} presents how to associate the
\textit{Service Process}' {\sc Action} with real services.

Notice that all \textit{$\pi$ServiceComposition} meta-model concepts can be
represented in the real application model, as we illustrate next using the
scenario example.

\subsubsection{\textit{Publish Music} Service Composition}

To illustrate the use of the \textit{$\pi$-ServiceComposition} model we used it for
defining the policy based composition model of the example scenario (see figure
\ref{fig:servicecompositionPolicy}). There are four external {\sc Business
Collaborators}, they are: \textit{Bank, Spotify, Twitter} and \textit{Facebook}.
The model also shows the business process of the application that consists of
six {\sc Service Activities} (see figure \ref{fig:example_serviceprocess}):
\textit{search music, select music, buy music, download music, listen music} and
\textit{publish music}. Note that the action \textit{publish music} of the
application calls the actions of two service collaborators namely Facebook and
Twitter, and the action \textit{buy music} of the application calls two actions
of the service collaborator Bank.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/runningExampleServiceComposition.pdf}
\caption{Service Composition Model with Policy Definition - Spotify Music
Service.}
\label{fig:servicecompositionPolicy}
\end{figure}

The {\em Facebook} and {\em Twitter} services require authentication
protocols in order to execute methods that will read and update the users'
space. A call to such services must be part of the authentication protocol
required by these services. In this example we  associate two authentication
policies, one for the open authentication protocol, represented by the class
{\sf\small Twitter AuthPolicy} that will be associated to the activity
{\sf\small UpdateTwitter} (see figure \ref{fig:servicecompositionPolicy}). In
the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http
authentication protocol will be associated to the activity {\sf\small
UpdateFacebook}. {\sf\small OAuth}  implements the open authentication protocol.
As shown in figure \ref{fig:servicecompositionPolicy}, the {\em Policy} has a
variable {\sf\small Token} that will be used to store the authentication token
provided by the service. This variable is imported through the library
{\sf\small Auth.Token}. The {\em Policy}  defines two rules, both can be
triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has
been associated to the variable {\sf\small token}, stated in by the condition of
rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small
R$_1$}); (ii) if the token has expired, stated in the condition of rule
{\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}). Note
that the code in the actions profits from the imported  {\sf\small Auth.Token}
for transparently obtaining or renewing a token from a third party.

HTTP-Auth implements the HTTP-Auth protocol.  As shown in figure
\ref{fig:servicecompositionPolicy}, the {\em Policy} imports an http protocol
library and it has two variables {\sf\small username} and {\sf\small password}.
The event of type {\sf\small ActivityPrepared} is the triggering event of the
rule {\sf\small R$_1$}. On the notification of an event of that type, a
credential is obtained using the username and password values. The object
storing the credential is associated to the scope, i.e., the activity that will
then use it for executing the method call.

\bigskip
\bigskip

Thanks to rules and policies it is possible to model and associate
non-functional properties to services' compositions  and then generate the code.
For example, the atomic integration of information retrieved from different
social network services, automatic generation of an integrated view of the
operations executed in different social networks or for providing security in
the communication channel when the payment service is called.

Back to the  definition process of a SIS, once the {\em Policy} based services'
composition model has been defined, then it can be transformed into a model
(i.e., $\pi$-PEWS model, Section \ref{sec:psm-pisodm}) that can support then
executable code generation.

% The following Section describes the $\pi$-PEWS
% meta-model that supports this representation.

\section{$\pi$-PEWS Platform Specific Models}
\label{sec:psm-pisodm}

Platform specific models (PSMs) are used to combine the specifications in
the PIM with the details of the chosen implementation platform. PSM
models are used to implement the system (generating code automatically), and
they must provide all information needed to build the system and its
components. A PSM model can also run as a model to be used for further
refinements by others PSM models.

$\pi$SOD-M platform model proposed at PSM level combines
specific details of the service-based platforms. We use a $\pi$-PEWS language
meta-model for representing the specification of a service composition on the
language. A model can be then used to generate the corresponding code.
$\pi$-PEWS~\cite{Placido2010LTPD} is a extension of the PEWS language that
provides constructs for specifying policies for services through contracts clauses (see Appendix \ref{append:pews_language}).

% From models described in PEWS, the code can be automatically generated. We extend the PEWS language, and named the its extension as
% $\pi$-PEWS (see Appendix \ref{append:pews_language}), and aggregate to it
% constructs for specifying policies for services through contracts clauses.

\subsection{$\pi$-PEWS Specification, Terms and Concepts}

 A \textit{$\pi$-PEWS} model represents the system
 specification. This model is essential for the system specification,
 detailing the services, compositions and system's constraints. A
 \textit{$\pi$-PEWS} specification describes the behavioural aspects of the
 system.

 We extended the PEWS language~\cite{Placido2010LTPD} to support the notion
\textit{policy} through contracts definitions (see figure
\ref{fig:contractRepresentation} and appendix \ref{append:pews_language}). The
extension does not modify the syntax of existing PEWS programs, but complements
it by specifying (non-functional) restrictions in a separate block. This feature
is intended to enhance reusability and allows the programmer to adequate a
program to specific requirements of the application being developed. This means
that the program developer can reuse the control part of the program and add
application-specific requirements as contract or time constraints.

 \begin{figure}[ht!]
\centering
\includegraphics[width=.9\textwidth]{chapters/methodology/figs/pewsEspecification.pdf}
\caption{$\pi$-PEWS Specification Representation.}
\label{fig:pewsRepresentation}
\end{figure}

 Behavioural web service interface languages describe not only the input/output
interface of a web service but also the expected behaviour of its components.
This behaviour can be specified as a workflow, defining the \textit{order} in
which the components of a service will be executed, so that the workflow
specifies the functional behaviour of the compound web service (figures \ref{fig:pewsRepresentation} and
\ref{fig:contractRepresentation}).

% This is what
% $\pi$-PEWS programs do (figures \ref{fig:pewsRepresentation} and
% \ref{fig:contractRepresentation}).

\begin{figure}[ht!] 
\centering
\includegraphics[width=.9\textwidth]{chapters/methodology/figs/pewsEspecificationContract.pdf}
\caption{$\pi$-PEWS Contract Representation.}
\label{fig:contractRepresentation}
\end{figure}

Additionally, we can specify non-functional behaviour for a service;
\textit{i.e}, to impose some additional restrictions which are separated from 
the application's workflow. This can be done by
using the notion of \textit{contract} to be added to each particular instantiation of
the PEWS program. The result of this is to allow a given service (workflow) to
have different restrictions in different contexts.

The policy model proposed here follows the main ideas presented
in~\cite{Espinosa-OviedoVZC09,PortillaHE08}, where contracts are added to a
given composition, as a separate concern. The properties defined by a contract
should be verified at runtime. Recovery actions, defined by the contract, are to
be performed in case of failure of the contract's conditions. Recovery actions
are defined by the contract itself.

\subsection{Meta-model}

% , a programming language that lets the service designer  combine the methods or
% subprograms that implement each operation of a service, in order to achieve the
% desired application logic.

The idea of the $\pi$-{\sc Pews} meta-model is based on the services'
composition approach provided by the language PEWS
\cite{BHM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}).
Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model consisting of  classes representing:

\begin{itemize}
\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc Composite Operation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
A {\sc Path} can be an {\sc Operation} or a {\sc Composite Operation}
denoted by an identifier. A {\sc Composite Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
choice ($\ + \ $) among services,
the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).

\item {\em Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
\end{itemize}

\begin{figure}[ht!]
\centering
\includegraphics[width=1.0\textwidth]{chapters/methodology/figs/PEWSMetamodel}
\caption{$\pi$-PEWS Meta-model.}
\label{fig:metamodel}
\end{figure}

As shown in the diagram an {\sc Policy} is applied to a {\sc Scope} that can be
either an {\sc Operation} (e.g., an authentication protocol associated to a
method exported by a service),  an {\sc Operator} (e.g., a temporal constraint
associated to a sequence of operators, the authorized delay between reading a
song title in Spotify and updating the walls must be less then 30 seconds), and
a {\sc Path} (e.g., executing the walls' update under a strict atomicity
protocol -- all or noting).  It groups a set of ECA rules, each rule having a
classic semantics, i.e, {\em when an event of type E occurs if  condition C is
verified then execute the action A}.  Thus, an {\em Policy} represents a set of
reactions to be possibly executed if one or several triggering events of its
rules are notified.

\begin{itemize}
\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
\item The class {\sc Policy} represents a recovery strategy implemented by ECA
rules of the form {\sc Event} - {\sc Condition} - {\sc Action}. A {\em Policy}
has variables that represent the view of the execution state of its associated
scope, that is required for executing the rules. The value of a variable is
represented using the type {\sc Variable}. The class {\sc Policy} is specialized
for defining specific constraints, for instance authentication policies.
\end{itemize}

Given a $\pi$-ServiceComposition model of a specific service-based application
(expressed according to the $\pi$-ServiceComposition meta-model), it is possible
to generate its corresponding $\pi$-PEWS model by using transformation
rules. The following Section describes the transformation rules between the
$\pi$-ServiceComposition and $\pi$-PEWS meta-models of our method.

The $\pi$-PEWS language (extension of PEWS) is described in the Appendix
\ref{append:pews_language}. The generated code by $\pi$SOD-M, as end product, is
based on the meta-model shown in figure \ref{fig:servicecomposition} and the
language syntax.

% \subsection{\textit{Publish Music} $\pi$-PEWS Specification}

\section{Model Transformations}
\label{sec:models-tranformation}

A model transformation usually specifies which models are acceptable as input,
and if appropriate what models it may produce as output, by specifying the
meta-model to which model must conform. Model transformations can be seen
as processes that take models as input and produces models as output. There
is a wide variety of kinds of model transformation and uses of them, which differ in their inputs and outputs
and also in the way they are expressed \cite{miller}. A model transformation is
also a way of ensuring that a family of models is consistent, in a precise sense
which the software engineer can define. The aim of using a model transformation
is to save effort and reduce errors by automating the building and modification
of models where possible \cite{atl_manual}.

The $\pi$SOD-M model transformations occur in the two levels, PIM and PSM,
we defined:  PIM-to-PIM and PIM-to-PSM transformations. Transformations in the
same level are considered ``horizontal transformations''. Transformations
between different levels are called ``vertical transformations''. There are 3
model transformations defined in $\pi$SOD-M: $\pi$-UseCase to
$\pi$-ServiceProcess; $\pi$-ServiceProcess to $\pi$-ServiceComposition; and
$\pi$-ServiceComposition to $\pi$-PEWS.

\begin{figure}[ht!]
  \centering
  \subfloat[From one A to one B]
  {\label{fig:rule01}\includegraphics[width=0.33\textwidth]{chapters/implementation/figs/rule01}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  ~
   \subfloat[From many A to many B]
  {\label{fig:rule02}\includegraphics[width=0.33\textwidth]{chapters/implementation/figs/rule02}}
  ~
   \subfloat[From many A to one B]
  {\label{fig:rule03}\includegraphics[width=0.33\textwidth]{chapters/implementation/figs/rule06}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
 \subfloat[From one A to one B, and one C]
  {\label{fig:rule04}\includegraphics[width=0.33\textwidth]{chapters/implementation/figs/rule03}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  ~
   \subfloat[From many A to many B, and one C]
  {\label{fig:rule05}\includegraphics[width=0.33\textwidth]{chapters/implementation/figs/rule04}}
  ~
   \subfloat[From many A to many B, and one C]
  {\label{fig:rule06}\includegraphics[width=0.33\textwidth]{chapters/implementation/figs/rule05}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \caption{Entities' Model Transformation Rules.}
  \label{fig:rules}
\end{figure}

 All $\pi$SOD-M transformation rules are described in natural language.
 These transformations ensure consistency between the concepts being refined and
 processed at different levels. Figure \ref{fig:rules} presents the set of rules
we defined to apply for each type of transformation. The entities of the
left-hand side represent the source model, and those on the right-hand side
represent the target model. Figure \ref{fig:rule01} shows the transformation
rule in which a single source entity is transformed in a target model entity.
Figure \ref{fig:rule02} describes a many to many entity transformation: a set of
objects from the source model is transformed into a set of objects in the target
model. Figure \ref{fig:rule03} presents a many to one transformation rule: many
objects in the source model it will be transformed into a single object element
in the target model. The next 3 rules are advanced rules, which transform single element in the source model into a set of different elements
in the target model. In figure \ref{fig:rule04}, a source element is
transformed in two different ones, \textit{e.g.} from a ``\textit{A}'' element,
it generates a ``\textit{B}'' and a ``\textit{C}'' element in the target model.
In figure \ref{fig:rule05}, a set of source elements is
transformed into a set of elements ``\textit{B}'' and one ``\textit{C}'' element
in the target model. Finally, figure \ref{fig:rule06} presents the rule which
transforms source elements into a set of elements
``\textit{B}'' and one element ``\textit{C}'' in the target model.

\subsection{From $\pi$-UseCase to
$\pi$-ServiceProcess}

The PIM to PIM transformation process between $\pi$-UseCase and
$\pi$-ServiceProcess (Table \ref{tab:transformationUseCaseToServiceProcess})
models defines how the application requirements are represented as a service
process. At this level, services and general functions are represented as simple
or composite use cases.

Every {\sc Use Case} is transformed into
an {\sc Action} in the \textit{$\pi$-ServiceProcess} model, and every {\sc
Extend} association identified in the \textit{$\pi$-UseCase} model is
transformed into a {\sc Fork Node}, \newText{because several extends relations
implies a conditional verification in the execution workflow}. If the extend
association has only one {\sc Use Case}, the fork will present the {\sc Action} as an alternative flow, and later, both flows will meet. If the extend association has several source
{\sc Use Case}, the fork will present different {\sc Actions} as mutual
alternatives flows, and later, all these flows will meet. A {\sc Activity
Services} (from $\pi$-ServiceProcess meta-model) consists of a composition of
one or more {\sc Actions}. {\sc Action} can be viewed as a method of
the system application. Thus, services are represented as a set of functions in
the \textit{$\pi$-ServiceProcess} meta-model, because {\sc Contracts} modelled
are associated to functions ({\sc Actions}).

All {\sc Constraints} in the source model (\textit{$\pi$-UseCase model}) are
transformed into a {\sc Contract}. A {\sc Contract} is represented as a set of {\sc Constraints}. Each
constrain description is directly transformed into {\sc Assertion}. A set of
{\sc Constraint} of a {\sc Use Case} or a {\sc Composite Use Case} is
transformed into a set of {\sc Assertion} used to define a
{\sc Contract}. All {\sc Use Cases} are associated to a {\sc Non-Functional
Requirement} and all {\sc Constraints} are associated with a {\sc Non-Functional Attribute}. Thus,
{\sc Constraint} representing the same {\sc Use Case} is transformed into a
{\sc Contract}. A {\sc Contract} can have exceptional cases. The negation of the
{\sc Constraint} predicate are treated as {\sc Exceptional Behaviour}. In this
particular case, a {\sc Exceptional Behaviour} will be generated in the target
model.

The {\sc Constraint Type} can have different transformations, depending of the
following values:

\begin{itemize}
\item If the value of {\sc Constraint Type} is \textit{BUSINESS}, it will be
transformed into none {\sc Assertion};
\item If the value of {\sc Constraint Type} is \textit{VALUE} with the
{\sc Assertion} \textit{isExceptionalBehaviour} attribute setted to
\textit{false}, it will be transformed into one {\sc Assertion};
\item If the value of {\sc Constraint Type} is \textit{VALUE} with the
{\sc Assertion} \textit{isExceptionalBehaviour} attribute setted to
\textit{true}, it will be transformed into a {\sc Exceptional Behaviour};
\end{itemize}

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Transformation Rules: From $\pi$-UseCase to $\pi$-ServiceProcess.}
\label{tab:transformationUseCaseToServiceProcess}
\centering
\begin{tabular}{l|l|l}
    \hline
    Source  &  Mapping Rules & Target\\
    $\pi$-UseCase meta-model  &   & $\pi$-ServiceProcess meta-model\\
    \hline
    \hline

	Constraint & - All \textbf{Constraints} in the source model will be transformed
	into a \textbf{Assertion} in the target model.  & Assertion, Contract, \\
	 & All \textbf{Use Cases} are associated to a \textbf{Non-Functional
	 Requirement} and all \textbf{Constraints} & Exceptional Behaviour   \\
	 &are associated with a \textbf{Non-Functional Attribute}. Thus,
	 \textbf{Constraint} representing the same Use Case,  &  \\
	& is transformed in a \textbf{Contract}. & \\
	& - A \textbf{Contract} can also have a exceptional cases. The negation of the
	\textbf{Constraint} predicate are treated &\\

	& as the type \textit{EXCEPTIONALBEHAVIOUR}. Thus, all the \textbf{Assertions}
	related to this \textbf{Constraint},  &\\

	&in this particular case, will be transformed in a \textbf{Exceptional
	Behaviour}.&\\
	& - The set of {\sc Assertions} of a {\sc Use Case} are associated with a
	{\sc Contract} & \\
	& - The {\sc Contract} name is formed by ``<useCaseName>'' + ``Contract''
	token. & \\
\hline

	 Constraint Type & - The \textbf{Constraint Type} in source model can have the
	 following transformation: & Assertion\\

	   &  $\bullet$ \textbf{BUSINESS} type is transformed into none
     \textbf{Assertion}; &  Exceptional Behaviour\\

      &  $\bullet$ \textbf{VALUE} with the
     \textbf{Assertion} \textit{isExceptionalBehaviour} attribute setted to false
      is transformed into& \\
     & a \textbf{Assertion} in the target model;&\\

      &  $\bullet$ \textbf{EXCEPTIONALBERAVIOR} with the
      \textbf{Assertion} \textit{isExceptionalBehaviour} attribute setted  & \\
      &to true is transformed into a \textbf{Exceptional Behaviour} in the target
      model.&\\

    & - Both, the  value and exceptional restriction are based on in accordance
    with the values&\\
    & of \textbf{Assertions}&\\

    & - \textbf{Constraint Type} defines the type of restriction on a
    \textbf{Use Case}. &  \\

% \hline
%
% 	Assertion, & Every \textbf{Assertion} associated to an element (\textbf{Assertion
% 	Property} and \textbf{Constraint}) in the source  & Assertion, \\
% 	Assertion Property & model becomes an \textbf{Assertion} associated to the
% 	corresponding element in the target model. & Assertion Property \\
%
\hline

	Non-Functional & Every \textbf{Non-Functional Attribute} associated to an
	element (\textbf{Constraint} and \textbf{Non-Functional} &
	Non-Functional \\ Attribute &  \textbf{Requirements}) in the source model
	becomes a \textbf{Non-Functional Attribute} associated to the  &
	Attribute \\
	&corresponding element (\textbf{Contract}) in the target model. &\\
\hline
    Use Case   &  For every \textbf{Use Case} there will be a
    \textbf{Action} in the service delivery process model & Action  \\
    & describing the Business Service &   \\

    \hline

    Extend (Use Case)    & - Every \textbf{Extend} association identified in the
     \textit{$\pi$-UseCase model} will be represented in the target model   & Control Node, \\
     &by a \textbf{Fork Node}.&  Fork Node,\\
      & - The \textbf{Service Activity} corresponding to the source
      \textbf{ Use Case} of the extend association must be  & Service
      Activity
    \\

      & previous to the \textbf{Service Activity} corresponding to the
      target \textbf{ Use Case} of the extend association. &\\

        &  - If the extend association has only one source \textbf{
        Use Case}, the fork will present the \textbf{Service} & \\
        &\textbf{ Activity} as an alternative to another flow with no \textbf{Service Activity}. Later, both flows will meet;&\\

       &  - If the extend association has several source \textbf{
        Use Case}, the fork will present the different \textbf{Service} & \\
        &\textbf{Activities} as mutual alternatives to another flow with no
        \textbf{Service Activity}. Later, all these flows will&\\
        &meet;&\\
    \hline

      Include (Use Case)    & - An \textbf{Include} association is found in the
      \textit{$\pi$-UseCase model},  the \textbf{Service Activity} corresponding to the    &
      Service Activity \\
       &source must be subsequent to the \textbf{Service Activity} corresponding
       to the target \textbf{ Use Case} of the & \\
      &include association;&\\

        &  - If the include association has several targets, the designer must
        decide the appropriate sequence & \\
        & for the different \textbf{Service Activities} corresponding to the
        target  \textbf{ Use Case} (which will obviously&\\

        &  be previous to the \textbf{Service Activity} corresponding to the
        source \textbf{ Use Case}).&\\

\hline

	Requirement & - A \textbf{Requirement} in the source model is transformed into
	\textbf{Service Activity} of the target model. & Service Activity
	\\ & - The rules for a \textbf{Requirement} to be transformed into a  &\\
	& service activity is analyzed after the transformations of included and
	extended use cases.&\\
	 &- A \textbf{Requirement} associated with only one \textbf{Use Case} in the
	 source model is transformed in a.&\\
	 &\textbf{Service Activity} and an \textbf{Action}, respectively. &\\
	&- All \textbf{Use Cases} must be associated with a \textbf{Requirement}.&\\

\hline

\end{tabular}
\end{table}

\begin{figure}[ht!]
  \centering
  \subfloat[From \textit{Use Case} and \textit{Constraint} to \textit{Action}
  and \textit{Contract}]
  {\label{fig:trans01}\includegraphics[width=0.5\textwidth]{chapters/implementation/figs/trans01}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
   \subfloat[From \textit{Constraint} to \textit{Contract} (2)]
  {\label{fig:trans02}\includegraphics[width=0.5\textwidth]{chapters/implementation/figs/trans06}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[From \textit{Composite Use Case} to \textit{Service Activity}]
  {\label{fig:trans03}\includegraphics[width=0.5\textwidth]{chapters/implementation/figs/trans03}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank
   % line to force the subfig onto a new line)
  \subfloat[From \textit{Requirements} to \textit{Service Activity}]
  {\label{fig:trans04}\includegraphics[width=0.5\textwidth]{chapters/implementation/figs/trans02}}
  \caption{$\pi$-UseCase2$\pi$-ServiceProcess Model Transformation Rules.}
  \label{fig:modelRulesUS2SP}
\end{figure}

Given the \textit{$\pi$-UseCase} model, to all {\sc Constraint} entity related
with a {\sc Use Case}, there is a {\sc Contract} that compounds a set of {\sc
Assertions} entity (figure \ref{fig:trans01}) and the {\sc Use Case} is refined
in a service {\sc Action}.

\begin{exampl}
Considering the scenario example presented in Section
\ref{sec:example}, the transformation of the ``\textit{listen music}'' {\sc Use
Case} is transformed into a service {\sc Action}. This action is a Spotify
service function that can be invoked to play the music. For ``\textit{publish music}''
{\sc Use Case}, all {\sc Constrains} are transformed in a set of
{\sc Assertions} that are grouped in a {\sc Contract}
(``\textit{publishMusicContract}''). The
``\textit{publishMusicContract}'' is related with the ``\textit{publishMusic}''
{\sc Action}.
\end{exampl}

\begin{figure}[ht!]
\centering
\subfloat[From \textit{Extend} to \textit{Fork/Join Flow}]
{\label{fig:trans05}\includegraphics[width=0.9\textwidth]{chapters/implementation/figs/trans04}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\\
\subfloat[From \textit{Include} to \textit{Activity Node Flow}]
{\label{fig:trans06}\includegraphics[width=0.9\textwidth]{chapters/implementation/figs/trans05}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\caption{$\pi$-UseCase2$\pi$-ServiceProcess Model Transformation Rules (2).}
\label{fig:modelRulesUS2SP2}
\end{figure}

A {\sc Constrain} transformation means that from {\sc Constraint} and {\sc
Constraint Type} entities are generated detailed {\sc Contract} information,
that are refined into {\sc Exceptional Behaviours} and {\sc Assertions} entities
(figure \ref{fig:trans02}). {\sc Constraint} is related with a {\sc Constraint Type}, and there are
different cases for the transformation of this concept. A \textit{Business}
type are transformed into an {\sc Assertion} that does not consider value
attributes information like \textit{maxValue} and \textit{minValue} (see figure \ref{fig:usecasemodel}). These {\sc Assertion} information
attributes are considered if the {\sc Constraint Type} is a \textit{Value}
type. This is semi-automatic, because there is not enough
information in a $\pi$-UseCase model to run a complete automatic transformation.
If the {\sc Constraint Type} is a \textit{Value} type, the designer must specify
the variable information, and its boundary values after the transformation. For this, the designer must consider the {\sc Constraint} and {\sc Use
Case} descriptions. By default, \textit{value constraints} are transformed into
\textit{pre-conditions} and  \textit{business constraints} are transformed into
\textit{post-conditions}. This rule may be adjusted by the designer at the time
of transformation.

A {\sc Service Activity} and the {\sc Actions} are generated from two different
elements in the source model ($\pi$-UseCase), from a {\sc Requirement} or a {\sc
Composite Use Case}. Figures \ref{fig:trans03} and \ref{fig:trans04} present
the transformation schema for {\sc Requirement} and {\sc
Composite Use Case} concepts. Both elements are related with many {\sc Use
Cases} that are transformed in many {\sc Actions}. A {\sc Use Case} are related
with, either, a {\sc Requirement} and a {\sc Composite Use Case}. {\sc Actions}
are related with a {\sc Service Activity}.

\begin{exampl}
For the execution of the ``\textit{download music}'' use case is necessary
process with the payment process. Thus, the set of {\sc Use Cases} that include the ``\textit{download
music}'' process are transformed in {\sc Actions}, and a {\sc Service Activity}
that aggregates all these {\sc Actions} is also generated. 
\end{exampl}

\begin{figure} [ht!]
\centering

\subfloat[Extended Use Case Example]
{\label{fig:extend01}\includegraphics[width=0.3\textwidth]{chapters/implementation/figs/ucExtended}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\subfloat[Extended Use Case Example (2)]
{\label{fig:extend02}\includegraphics[width=0.3\textwidth]{chapters/implementation/figs/ucExtended2}}
\\
\subfloat[Equivalent Service Process Workflow]
{\label{fig:servProc01}\includegraphics[width=0.3\textwidth]{chapters/implementation/figs/servProc}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\subfloat[Equivalent Service Process Workflow (2)]
{\label{fig:servProc02}\includegraphics[width=0.3\textwidth]{chapters/implementation/figs/servProc2}}

\caption{Extended Transformation Examples.}
\label{fig:modelRulesUS2SP}
\end{figure}

The transformations for {\sc Extend} and {\sc Include} dependence elements are
not as simple as the previous transformations (figures \ref{fig:trans05}.
The workflow generated for each \textit{<<extends>>} relationship with just two
use cases (figure \ref{fig:extend01}) is described by figure
\ref{fig:servProc01}. The generated workflow contains: one {\sc Fork Node},
one {\sc Join Node}, and four {\sc Control Flow} elements, and also two {\sc
Action} elements, one for each {\sc Use Case}, if there are less than 2 extended
{\sc Use Case}. When there is more than one  \textit{<<extends>>} relationship
among different use cases (figures \ref{fig:extend02} and
\ref{fig:servProc02}), the transformation is proceed as: adding two more {\sc
Control Flow} for each new {\sc Use Case}, and one {\sc Action} for each new
extended {\sc Use Case}.

\begin{exampl}
Considering the scenario example, it is possible to apply
the same rule for the ``\textit{publish music}'' use case, which has two
extended use cases, ``\textit{public twitter}'' and ``\textit{public
Facebook}'', applying the illustration expressed in figures \ref{fig:extend02}
and \ref{fig:servProc02}.  
\end{exampl}

For the {\sc Include} use case elements, the transformation is represented as a
{\sc Action} sequence, as shown in figure \ref{fig:trans06}. For each {\sc Use
Case} element an {\sc Action} element  is generated. For a set of \textit{n}
{\sc Use Cases} we generate \textit{n-1} {\sc Object Flow} elements. Each
{\sc Control Flow} links two {\sc Actions}.

\begin{exampl}
Given the ``\textit{download music}'' use case from the scenario example, it
includes payment process to buy the music. It is represented as a include in the $\pi$-UseCase model and is
transformed in a sequence flow in the $\pi$-ServiceProcess model (as presented
in figure \ref{fig:include}).
\end{exampl}

All this transformation rules from \textit{$\pi$-UseCase} model to
\textit{$\pi$-ServiceProcess} model were validated with the scenario example.
All transformations described in this section are completely automatic, however
information about values and its boundaries are inserted by the designer after
the transformation process.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.5\textwidth]{chapters/methodology/figs/include.pdf}
\caption{Include Transformation Example.}
\label{fig:include}
\end{figure}

\subsection{From $\pi$-ServiceProcess to
$\pi$-ServiceComposition}

The PIM to PIM transformation process from \textit{$\pi$-ServiceProcess} to
\textit{$\pi$-ServiceComposition} model will refine further the application
design. The main goal of this transformation is to group all {\sc Contracts} and
{\sc Actions} in {\sc Policies} and {\sc Service Activities}, respectively.

All {\sc Actions} entities in the source model will be transformed into an {\sc
Action} in the target model, and every {\sc Service Activity} in the source
model will be transformed into a {\sc Service Activity} in the target model.
This happens because \textit{$\pi$-ServiceComposition} is an extension of
\textit{$\pi$-ServiceProcess}.

For the non-functional specifications, the {\sc Contract} entity with the same
{\sc Non-Functional Requirement} in the source model
(\textit{$\pi$-ServiceProcess} model) will be transformed into a {\sc Policy} in
the target model ($\pi$-ServiceComposition model).  Each {\sc Assertion} will be
transformed into a {\sc Rule:condition} attribute in the target model, but if
the {\sc Assertion} has a value type, the \textit{name}
and the attributes in the source model will be transformed into a
\textsc{Variable} in the target model. The other {\sc Assertion
Property} values remains unchanged.

The {\sc  Assertion:aProperty} attribute can have different transformations,
depending of the following values:

\begin{itemize}
\item \textit{POST-CONDITIONS} are transformed into \textit{POST} in
the $\pi$-ServiceComposition model;
\item \textit{PRECONDITIONS} are transformed into \textit{PRE}
in the $\pi$-ServiceComposition model;
\item \textit{TIMERESTRICTIONS} are transformed into \textit{TIME}
in the $\pi$-ServiceComposition model.
\end{itemize}

The {\sc Exceptional Behaviour} entities will be
transformed into an {\sc Action} in the \textit{$\pi$-ServiceComposition} model,
and every {\sc Non-Functional Attribute} associated to an element
({\sc Contract} and {\sc Non-Functional Requirement}) in the
$\pi$-ServiceProcess model becomes a \textsc{Non-Functional Attribute}
associated to the corresponding element ({\sc Policy}) in the \textit{$\pi$-ServiceComposition}
model.

Table \ref{tab:transformationServiceProcessToComposition} describes the
transformations between \textit{$\pi$-ServiceProcess} and
\textit{$\pi$-ServiceComposition} meta-models.

\begin{table} [ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Transformation Rules: From $\pi$-ServiceProcess to
$\pi$-ServiceComposition.}
\label{tab:transformationServiceProcessToComposition}
\centering
\begin{tabular}{l|l|l}
    \hline
    Source  &  Mapping Rules & Target\\
    $\pi$-ServiceProcess meta-model &   & $\pi$-ServiceComposition\\
      &&  meta-model\\
    \hline
    \hline

   Action & All \textbf{Actions} in the source model will be
   transformed in a \textbf{Action} in the target model. &Action\\
   \hline
   Service Activity&  - All \textbf{Service Activity} in the source model will
   be transformed in a \textbf{Service Activity} in the target model. &Service Activity,\\
   \hline
   Contract, & All \textbf{Contract} with the same \textbf{Non-Functional
   Requirement}, in the source model, will be transformed in& Policy\\
   Non-Functional Attribute&  a \textbf{Policy} in the target
   model.&\\
   \hline
   Assertion & Each \textbf{Assertion} will be transformed in a
   \textbf{Rule:condition} attribute in the target model. & Rule:condition\\
   \hline

   Assertion:name, & If the \textbf{Assertion} has the type equal to
   \textit{VALUE}, the \textit{name} and the \textit{type} attribute in the source
   model will & Variable\\
   Assertion:type & be transformed in a \textbf{Variable} in the target model.&
   \\
   \hline
   Exceptional Behaviour & An \textbf{Exceptional Behaviour} and its attributes in
   the source model will be transformed & Rule:action\\
   &in a \textbf{Rule:action} in the target model. &\\
   \hline
   Assertion:aProperty &- Depending on the \textbf{assertion Property} type in
   the source model, it will be transformed in a \textbf{Rule:event}&
   Rule:Event,\\ & in the target model. The transformation rules
    are:& Event Type\\
    &$\bullet$ \textit{PRECONDITION} type is transformed into \textit{PRE}
    in the target model;&\\
    &$\bullet$ \textit{POST-CONDITION} type is transformed into \textit{POST} in
    the target model;&\\
    &$\bullet$ \textit{TIMERESTRICTION} type is transformed into \textit{TIME}
    in the target model.&\\
   \hline

	Non-Functional & Every \textbf{Non-Functional Attribute} associated to an
	element (\textbf{Contract} and \textbf{Non-Functional} &
	Non-Functional \\ Attribute &  \textbf{Requirements}) in the source model
	becomes a \textbf{Non-Functional Attribute} associated to the  &
	Attribute \\
	&corresponding element (\textbf{Policy}) in the target model. &\\
	\hline
\end{tabular}
\end{table}

\begin{figure} [ht!]
  \centering
  \subfloat[From \textit{Contract} and \textit{Assertion} to \textit{Policy}
  and \textit{Rule}]
  {\label{fig:trans07}\includegraphics[width=0.5\textwidth]{chapters/implementation/figs/trans07}}
   %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
   \subfloat[From \textit{Action} to \textit{Service Activity} (2)]
  {\label{fig:trans08}\includegraphics[width=0.5\textwidth]{chapters/implementation/figs/trans08}}
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
  \\
  \subfloat[From \textit{Exceptional Behaviour} to \textit{Action}]
  {\label{fig:trans09}\includegraphics[width=0.6\textwidth]{chapters/implementation/figs/trans09}}
  \caption{$\pi$-ServiceProcess2$\pi$-ServiceComposition Model Transformation
  Rules.}
  \label{fig:modelRulesSp2SC}
\end{figure}

Figure \ref{fig:modelRulesSp2SC} presents the main transformation rules for
this level. The main feature of this
transformation is the generation of a \textsc{Policy} from of a set of {\sc
Contracts}. But, \textit{what is the adopted criterion for this transformation?} The main criteria is that all
{\sc Contracts} from the same {\sc Non-Functional Requirement} will be
transformed in one {\sc Policy}. Thus, we can have {\sc Policies} on:
\textit{Safety, Performance, Reliability}, and so on.

% A set of {\sc Assertions} are composed in one {\sc
% Policy}, transformed directly into {\sc Rules}, and many {\sc Contracts} turn
% into one {\sc Policy}. This happens because all {\sc Assertions} that represent
% a contract, become directly in {\sc Rules}.

Each \textsc{Assertion} of a\textsc{Contract} ( \textit{$\pi$-UseCase}
model) is transformed into a \textsc{Rule} in the \textit{$\pi$-ServiceProcess}
model. The set of {\sc Constracts} generates a {\sc Policy} composed by those
{\sc Rules}. Thus, the set of {\sc Rules} make up
a single {\sc Policy}. It is important to highlight that only the {\sc
Contracts} of the same {\sc Non-Functional Requirement} is transformed into the
same {\sc Policy}. Different NFRs' {\sc Contracts}, are transformed into
different {\sc Policies}. Figure \ref{fig:trans07} show how this transformation
is made.

Similar to the transformation between {\sc Contract} and
{\sc Policy}, a set of {\sc Actions} is transformed into a single {\sc Service
Activity} entity. {\sc Actions} are related with {\sc Business Collaborator},
and all information related to a {\sc Business Collaborator} are transformed
from {\sc Packages} information ($\pi$-UseCase model). All {\sc Packages}
are transformed into {\sc Business Collaborator}. Figure \ref{fig:trans08}
presents the transformation between {\sc Action} and {\sc Service Activity}
elements.

An exceptional behaviour is triggered by a failure in the verification of the
condition of a rule. An {\sc Exceptional Behaviour}
is performed to preserve the state of the application. Figure
\ref{fig:trans09} presents how an {\sc Exceptional Behaviour} that can be
transformed into an {\sc Action} call.

As the \textit{$\pi$-ServiceComposition} model refines the
\textit{$\pi$-ServiceProcess} concepts at PIM level, a service
previously defined as actions (source model) is refined as composition of
those actions (target model) that are necessary to represent a business
service, identifying who are the partners involved in the realization ({\sc
Business Collaborators}). In addition, $\pi$SOD-M defines a platform specific
model based on web services composition. This model is explicitly
indicates those actions which are (or will be, if not yet implemented) supported
by web services.

\begin{exampl}
Considering the scenario example, the {\sc
Action}'s \textit{update music} {\sc Contract} is transformed is a {\sc Policy}
with its {\sc Rules}. All contract {\sc Assertions} are transformed in {\sc
Rule} and its attributes, \textit{e.g.} the login and password verification.  The
``\textit{securityLoginPolicy}'' is all set of {\sc Rules} that were transformed
from the {\sc Assertions} in $\pi$-ServiceProcess model. The {\sc Non-Functional
Requirement} information will be used to the {\sc Policy} generation comes from
the initial use case model. Also the {\sc Business Collaborator}
\textit{Facebook} and \textit{Spotify} information came from {\sc Package}
$\pi$-UseCase entity element. All {\sc Contracts} of the same {\sc
Non-Functional Requirement} are composed in a {\sc Policy}.
\end{exampl}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%% FROM --> SERVICE COMPOSITION; TO PEWS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{From $\pi$-ServiceComposition to
$\pi$-PEWS}

Table \ref{tab:transformationServiceCompositionToPEWS} describes the PIM to PSM
transformations between $\pi$-ServiceComposition and $\pi$-PEWS meta-models. We
propose two groups of rules: those that transform services composition
elements of the $\pi$-ServiceComposition into $\pi$-PEWS meta-model elements;
and those that transform rules grouped by policies into A-policy types.

A named action of the $\pi$-ServiceComposition represented by  {\sc\em Action}
and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a
corresponding attribute name {\sc Operation:name}. A  named service activity
represented by the elements {\sc\em Service Activity}  and  {\sc\em
Service Activity:name} of the $\pi$-ServiceComposition, are  transformed into a
named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc
Composite Operation} and {\sc Composite Operation:name}. When more than one action
is called, according to the following  composition patterns expressed using the
operators {\sc\em merge, decision, fork and join} in the $\pi$-ServiceComposition
the corresponding transformations, according to the PEWS operators presented
above, are:
\begin{itemize}
\item   $op_1 . op_2$ if no {\sc\em Control Node} is specified
\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
\item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
\end{itemize}

The {\em A-policies} defined for the elements of the $\pi$-ServiceComposition are
transformed into {\sc A-Policy} classes, named according to the names expressed
in the source model. The transformation of the rules expressed in the
$\pi$-ServiceComposition is guided by the event types associated to these rules.
The variables associated to an {\em A-policy} expressed in the
$\pi$-ServiceComposition as {\sc\em $<$Variable:name, Variable:type$>$} are
transformed into elements of type {\sc Variable} with attributes {\sc name} and
{\sc type} directly specified from the elements {\sc\em  Variable:name} and
{\sc\em Variable:type} of the $\pi$-ServiceComposition model.

\begin{table}[ht!]
\tiny
\renewcommand{\arraystretch}{1.3}
\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS.}
\label{tab:transformationServiceCompositionToPEWS}
\centering
\begin{tabular}{l|l|l}
    \hline
    Source  &  Mapping Rules & Target\\
    $\pi$-ServiceComposition meta-model  &   & $\pi$-PEWS meta-model\\
    \hline
    \hline

    Action   &  - An \textit{Action} in the source model corresponding to an
    external \textbf{Business Collaborator} is mapped  & Operation:alias\\
       &   to an Operation in target model.   & \\
       &   - The \textbf{Action:name} in the source model is transformed into
       \textbf{Operation:name} in the target model.  & \\
    \hline

    Service Activity    & - The \textbf{Service Activity} in the source model is
    mapped to a \textbf{Composite Operation} in target  & Type Operation,\\
        &  model when more than one \textbf{Actions} are called.   &
        Composition Operation\\

        & - If \textbf{Composite Operation} is generated for a given
        \textit{Service Activity} then the \textbf{Service Activity:name} & \\
        & in the source model is mapped to
        \textbf{Composition Operation:name} in the target model. &\\
    \hline

    Control Nodes   &  - The \textbf{Control Node} in the source model is mapped
    to a \textbf{Operator} in target model. According to the & Operator\\

    & type of \textbf{Control Node} (merge, decision, join, fork)
    the expression of the \textbf{Composite Operation} is: & \\

    &  $\bullet$ Sequence if no \textbf{Control Node} is specified;  & \\
    &  $\bullet$ Parallel - Sequence for  a \textbf{Control Nodes} pattern fork -
    join; & \\ &
    $\bullet$ Choice - Sequence for a \textbf{Control Node} pattern decision -
    merge &
    \\

    \hline

    Business Collaborator   & A \textbf{Business Collaborator:isExternal} in the
    source model generates a \textbf{Namespaces} in the target model  &
    Namespace\\
    \hline

    Rule:event   &  The \textbf{Rule}\'s attribute event in the source model is
    transformed into an \textbf{Event:type} of the target  & Event Type,\\
    & model. In this case attribute is mapped to an
entity with an attribute. The \textbf{Event Type} of a & Event\\
  & \textbf{Rule} in the target model is
determined by the Rule type: & \\
     &  $\bullet$ \textbf{Event Type} of a \textit{Precondition} \textbf{Rule} is
     \textit{ActivityPrepared}; & \\
     &  $\bullet$ \textbf{Event Type} of a \textit{Postcondition} \textbf{Rule}
     is \textit{TermActivity}; & \\
    &  $\bullet$ \textbf{Event Type} of a \textit{TimeRestriction} \textbf{Rule}
     is \textit{Activity}; & \\
    \hline

    Rule:condition   &  The \textbf{Rule}\'s attribute condition in the source
    model is transformed into a \textbf{Condition:expression}  & Condition\\
    &  in the target model.  In this case, an attribute is
mapped into an entity with an attribute  & \\
    \hline

    Rule:action   & The \textbf{Rule:action} in the source model is
    transformed in an \textbf{Action:act} in the target model.  & Action\\
    & The attribute action is mapped to an entity with an attribute. In the
    target model an action & \\
    & is executed according to the rule condition value (true/false). & \\

    \hline

    Policy   & - Every \textbf{Policy} associated to an element
    (\textbf{Business Collaborator, Service, Activity, Action}) in the   &
    APolicy\\
    & source model becomes an \textbf{APolicy} associated to the corresponding
element in the target model. & \\
   & - The name attribute of a \textbf{Policy} in the source model becomes an
\textbf{Apolicy:name} of the target model. & \\
    \hline
    Variable   &  Every \textbf{Variable}, and its attributes, associated to a
    \textbf{Policy} in the source model
    becomes a \textbf{Variable}   & Variable\\
    &associated to an \textbf{APolicy} in the target
    model. The variables can be used in an \textbf{APolicy}\'s Condition &\\
    & of the target model.&\\
    \hline
    Rule:event &  For a \textbf{Rule} in the source
    model, depending on the \textbf{Event Type}, the corresponding transformation
    & Precondition \\ &   in the target model is: \textbf{Precondition,
    Postcondition} or \textbf{TimeRestriction Rule}  & Postcondition\\
    &&TimeRestriction\\
    &&Rule\\
    \hline
\end{tabular}
\end{table}

As shown in Table \ref{tab:transformationServiceCompositionToPEWS}, for an event
of type {\sc\em Pre} the corresponding transformed rule is of type {\sc
Precondition}; for an event of type {\sc\em Post} the corresponding transformed
rule is of type {\sc Post-condition}; finally, for an event of type {\sc\em
TimeRestriction} the corresponding transformed rule is of type {\sc Time}. The
condition expression of a rule in the $\pi$-ServiceComposition ({\sc\em
Rule:condition}) is transformed into a class {\sc\em Condition:expression} where
the attributes of the expression are transformed into elements of type {\sc
Attribute}.

Figures \ref{fig:pewsRepresentation} and \ref{fig:contractRepresentation}
present the \textit{$\pi$-PEWS} specification resulting from the
\textit{$\pi$-ServiceComposition} model transformation of our scenario example.

\section{Conclusions}
\label{sec:pisodm_conclusion}

This chapter presented $\pi$-SOD-M for specifying and designing reliable service
based applications, using a MDA approach. As one of the main aims of MDA is to
separate design from architecture and technologies, the $\pi$SOD-M's models
describe the system behaviour and its restrictions, without considering
definition of a standard architecture.

Non-functional constraints are related to business rules associated to the
general semantics of the application. In the case of service-based
applications, this type of applications also concern the use constraints imposed
by the services. We worked on the definition of a method for explicitly expressing
such properties in the early stages of the specification of services based
applications. Having such business rules expressed and then translated and
associated to the services' composition can help to ensure that the resulting
application is compliant to the user requirements and also to the
characteristics of the services it uses.

\newText{
Our approach considers that it is possible to model both aspects, functional and
non-functional, in an service-oriented application development throughout the
whole development process.

A business process specification can be modelled with the concepts {\sc Use
Case} and {\sc Requirement}. In the same level, the NFRs can be modelled using
the concepts of {\sc Constraint}, {\sc Constraint Type} and {\sc NF-Requirment}.
The concepts used to represent the functional aspects and the business process were modeled by SOD-M method.

In order to fill the gap between the abstract specification and the concrete
implementation, the Business process activities that can be done  by external
entities can be modelled  with the concepts {\sc Action}, {\sc Business
Collaborator}. In order to express NFP from NFRs we proposed the concepts of:
{\sc Contract}, {\sc Assertion (property)} and  {\sc Exceptional behaviour}.

Since our objective is also to provide a platform dependent specification, We
also propose a platform specific meta-model, named $\pi$-PEWS, wich represents
the $\pi$-PEWS specification language for define web services path expressions.
 }

Designing and programming non-functional properties is not an easy task, so we
are defining a set of predefined {\em Policy} types with the associated use rules for guiding
the programmer when she associates them to a concrete application. {\em Policy}
types  that can also serve as patterns for programming or specializing the way
non-functional constraints are programmed.  We model and associate policies to 
service-based applications that represent both systems' cross-cutting aspects and use
constraints stemming from the services used for implementing them.

An advantage of our methodology is the use of high-level models,
which by means of model transformations, helps software developers to make the
most of the knowledge for specifying and developing business services.

% Before entering in detail in the models and the
% proposed process by $\pi$SOD-M, in the next section presents the methodology in
% the context of MDA architecture, which supports the proposed method.

### 2013-09-30 SOAP_Track2013/src/paperSOAP_tosubmit.tex
inicial --martin

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
+\documentclass{sig-alternate}
+  \pdfpagewidth=8.5truein
+  \pdfpageheight=11truein
 
 \begin{document}
+%
+% --- Author Metadata here ---
+\conferenceinfo{SAC'13}{March 18-22, 2013, Coimbra, Portugal.}
+\CopyrightYear{2013} % Allows default copyright year (2002) to be over-ridden - IF NEED BE.
+\crdata{978-1-4503-1656-9/13/03}  % Allows default copyright data (X-XXXXX-XX-X/XX/XX) to be over-ridden.
+% --- End of Author Metadata ---
 
-\begin{frontmatter}
+\title{Reliable Web Services Composition: A MDD Approach}
 
-%% Title, authors and addresses
+% \subtitle{[Extended Abstract]
+% \titlenote{A full version of this paper is available as
+% \textit{Author's Guide to Preparing ACM SIG Proceedings Using
+% \LaTeX$2_\epsilon$\ and BibTeX} at
+% \texttt{www.acm.org/eaddress.htm}}}
+%
+% You need the command \numberofauthors to handle the 'placement
+% and alignment' of the authors beneath the title.
+%
+% For aesthetic reasons, we recommend 'three authors at a time'
+% i.e. three 'name/affiliation blocks' be placed beneath the title.
+%
+% NOTE: You are NOT restricted in how many 'rows' of
+% "name/affiliations" may appear. We just ask that you restrict
+% the number of 'columns' to three.
+%
+% Because of the available 'opening page real-estate'
+% we ask you to refrain from putting more than six authors
+% (two rows with three columns) beneath the article title.
+% More than six makes the first-page appear very cluttered indeed.
+%
+% Use the \alignauthor commands to handle the names
+% and affiliations for an 'aesthetic maximum' of six authors.
+% Add names, affiliations, addresses for
+% the seventh etc. author(s) as the argument for the
+% \additionalauthors command.
+% These 'additional authors' will be output/set for you
+% without further effort on your part as the last section in
+% the body of your article BEFORE References or any Appendices.
 
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
+\numberofauthors{1} %  in this sample file, there are a *total*
+% of EIGHT authors. SIX appear on the 'first-page' (for formatting
+% reasons) and the remaining two appear in the \additionalauthors section.
+%
+\author{
+% You can go ahead and credit any number of authors here,
+% e.g. one 'row of three' or two rows (consisting of one row of three
+% and a second row of one, two or three).
+%
+% The command \alignauthor (no curly braces needed) should
+% precede each author name, affiliation/snail-mail address and
+% e-mail address. Additionally, tag each line of
+% affiliation/address with \affaddr, and tag the
+% e-mail address with \email.
+% 
+% 1st. author
+% \alignauthor
+% Pl\'acido A. Souza Neto\\
+% 	   \affaddr{Federal Institute of Rio Grande do Norte (IFRN)},\\
+%        \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
+%        \affaddr{Natal-RN, Brazil}\\
+%        \email{placido.neto@ifrn.edu.br}
+% \alignauthor Genoveva Vargar-Solar\\
+%        \affaddr{French Council of Scientific Research}\\
+%        \affaddr{LIG-LAFMIA }\\
+%        \affaddr{Saint Martin d'H\`{e}res, France}\\
+%        \email{Genoveva.Vargas-Solar@imag.fr}
+% \alignauthor Valeria de Castro\\
+%        \affaddr{Universidad Rey Juan Carlos}\\
+%        \affaddr{M\'{o}stoles, Spain}\\ 
+%        \email{Valeria.deCastro@urjc.es}
+% \and
+% \alignauthor Martin A. Musicante\\
+%        \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
+%        \affaddr{Natal-RN, Brazil}\\
+%        \email{mam@dimap.ufrn.br}
+% \alignauthor Jos\'e-Luis Zechinelli-Martini\\
+%        \affaddr{Fundaci\'on Universidad de las Am\'ericas Puebla (UDLAP)}\\
+%        \affaddr{San Andr\'es Cholula, M\'exico}\\
+%        \email{joseluis.zechinelli@udlap.mx}
+% \alignauthor Christine Collet\\
+%        \affaddr{Grenoble Institute of Technology}\\
+%        \affaddr{Saint Martin d'H\`{e}res, France}\\
+%        \email{Christine.Collet@grenoble-inp.fr}
+}
+% There's nothing stopping you putting the seventh, eighth, etc.
+% author on the opening page (as the 'third row') but we ask,
+% for aesthetic reasons that you place these 'additional authors'
+% in the \additional authors block, viz.
 
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
+% \additionalauthors{Additional authors: Javier A. Espinosa-Oviedo (Grenoble
+% Institute of Technology, email: {\texttt{Javier.Espinosa@imag.fr}}) and
+% Esperanza Marcos (Universidad Rey Juan Carlos, email: {\texttt{esperanza.marcos@urjc.es}}).}
+ \date{21 September 2012}
 
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
+% Just remember to make sure that the TOTAL number of authors
+% is the number that will appear on the first page PLUS the
+% number that will appear in the \additionalauthors section.
 
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
-
+\maketitle
 \begin{abstract}
-This paper presents\ldots
+This paper presents an approach for modeling and associating {\em Policies} to 
+services' based applications. It proposes to extend the SOD-M model driven
+method with (i) the $\pi$-{\sc SCM}, a {\em Policy} services' composition
+meta-model for representing non-functional constraints associated to services'
+based applications; (ii) the $\pi$-{\sc Pews} meta-model providing guidelines
+for expressing the composition and the policies; and, (iii)  model to model and
+model to text transformation rules for semi-automatizing the implementation of
+reliable services' compositions. As will be shown within our environment
+implementing these meta models and rules, one may represent both systems'
+cross-cutting aspects (e.g., exception handling for describing what to do when a
+service is not available, recovery, persistence aspects) and constraints
+associated to services, that must be respected for using them (e.g., the fact
+that a service requires an authentication protocol for executing a method).
 \end{abstract}
 
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
+% A category with the (minimum) three required fields
+%\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
+%performance measures]
 
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
+%\terms{Delphi theory}
 
-\end{keyword}
+\keywords{Methodology, $\pi$SOD-M, Sevice Composition, Policy}
 
-\end{frontmatter}
+%*********************************************************************************************************
+\section{Introduction}
+%*********************************************************************************************************
 
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
+Service oriented computing is at the origin of an evolution in the field of
+software development. An important challenge of service oriented development is 
+to ensure the alignment between IT systems and the business logic.
+%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
+Thus, organizations are seeking for mechanisms to deal with the gap between the
+systems developed and business needs \cite{bell}. The literature stresses the
+need for methodologies and techniques for service oriented analysis and design,
+claiming that they are the cornerstone  in the development of meaningful
+services' based applications \cite{5}.  In this context, some authors argue that
+the convergence of model-driven software development, service orientation and
+better techniques for documenting and improving business processes are the key
+to make real the idea of rapid, accurate development of software that serves,
+rather than dictates, software users' goals \cite{watson}.
 
-%% main text
+Service oriented development methodologies providing mod\-els, best practices,
+and reference architectures to build services' based applications mainly address 
+functional aspects \cite{1,2,decastro1,PapazoglouH06}.  Non-functional aspects
+concerning services' and application's "semantics", often expressed as
+requirements and constraints in general purpose methodologies, are not fully
+considered or they are added once the application has been implemented in order 
+to ensure some level of reliability (e.g., data privacy, exception handling,
+atomicity, data persistence). This leads to services' based applications that
+are partially specified and that are thereby partially compliant with
+application requirements.
 
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
+The objective of this work is to model non-functional constraints and associate
+them to  services' based applications  early during the services' composition
+modeling phase. Therefore this paper presents $\pi$SOD-M, a model-driven
+meth\-od that extends the SOD-M  \cite{decastro1} for building reliable  services' based
+information systems (SIS).
 %SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
 
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
+This work (i) proposes to extend the SOD-M \cite{decastro1} method with 
+the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing
+non-func\-tion\-al constraints associated to services' based applications; (ii)
+defines the $\pi$-{\sc Pews}  meta-model providing
+guidelines for expressing the composition and the {\em A-policies}; and finally,
+ (iii) defines model to model transformation rules for generating the $\pi$-{\sc
+ Pews} model of a reliable services' composition starting from the extended
+ services' composition model; and, model to text transformations for generating
+ the corresponding implementation. As will be shown within our environment
+ implementing these meta models and rules, one may represent both systems'
+ cross-cutting aspects (e.g., exception handling for describing what to do when
+ a service is not available, recovery, persistence aspects) and constraints
+ associated to services, that must be respected for using them (e.g., the fact
+ that a service requires an authentication protocol for executing a method).
 
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
+The remainder of the paper is organized as follows. Section \ref{sec:motivation}
+gives an overview of our approach. It describes a motivation example that
+integrates and synchronizes well-known social networks services namely Facebook,
+Twitter and, Spotify. Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and
+\ref{sec:mmrules} describe respectively the three key elements of our proposal,
+namely the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation
+rules that support the semi-automatic generation of reliable services'
+compositions.
 %describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
@@
 %*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
+\section{Modeling reliable services' \\ com\-po\-si\-tions with
+$\pi$SOD-M}\label{sec:motivation}
 %*********************************************************************************************************
-Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
+Consider for instance the following scenario. An organization wants to provide
+the services' based application ``To Publish Music'' that monitors the music a
+person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
 Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
@@
 Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
+The ``To Publish Music'' scenario 
 starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
@@
 \centering  
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
+\includegraphics[width=0.32\textwidth]{figs/SC} 
+\caption{BPMN model of the ``To Publish Music'' scenario}
 \label{fig:E3valuemodel}   
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+ Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach  
+ providing  a set of guidelines to build services' based information systems
+ (SIS) \cite{decastro1,decastro2}.  Therefore, SOD-M proposes to use services as
+ first-class objects for the whole process of the SIS  development and it 
+ follows a Model Driven Architecture (MDA) \cite{miller}  approach. Extending
+ from the highest level of abstraction of the MDA, SOD-M provides  a conceptual
+ structure to: first, capture the system requirements and specification in
+ high-level abstraction models (computation independent models, CIM's); next, 
+ starting from such models build platform independent models (PIM's) specifying
+ the system details; next transform such models into platform specific models
+ (PSM's) that bundles the specification of the system with the details of the
+ targeted platform; and finally, serialize such model into the working-code that
+ implements the system.
+
 \begin{figure} [htpb]
 \centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
+\includegraphics[width=0.5\textwidth]{figs/SODM}
 \caption{SOD-M development process}
@@
 \end{figure}
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
+
+As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by
+building the high-level computational independent models and enables specific
+models for a service platform to be obtained as a result  \cite{decastro1}.
+Referring to the ``To Publish Music'' application, using SOD-M the designer
+starts defining an E3value model \footnote{The E3 value model is a business model that
+represents a business case and allows  to understand the environment in which
+the services' composition will be placed \cite{e3value}.}  at the CIM level and
+then the corresponding models of the PIM are generated leading to a services'
+composition model (SCM).
+
 %SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
@@
 
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
+Now, consider that besides the services' composition that represents the order
+in which the services are called for implementing the application ``To Publish
+Music'' it is necessary to model  other requirements that represent the (i)
+conditions imposed by services for being contacted, for example the fact the
+Facebook and Twitter require authentication protocol in order to call their
+methods for updating the wall; (ii) the conditions stemming from the business
+rules of the application logic, for example the fact that the walls in Facebook
+and Twitter must show the same song title and if this is not possible then none
+of them is updated.
 
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
+Adding non-functional requirements and services con\-straints in the services'
+composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
 
@@
 %
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
+This work extends SOD-M for building applications by modeling the application
+logic and its associated non-func\-tion\-al constraints and thereby ensuring the
+generation of reliable services' composition.
 %This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
+As a first step in our approach, and for the sake of simplicity we started
+modeling non-functional constraints at the PSM level. Thus, in this paper we
+propose the $\pi$-SCM, the services' composition meta-model extended with {\em
+A-policies} for modeling non-functional constraints (highlighted in Figure 
+\ref{fig:sodm} and described in Section \ref{sec:piscm}).  $\pi$SOD-M defines
+the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the
+services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
 Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
@@
 %*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
+\section{$\pi$-services' composition \\ meta-\-model}\label{sec:piscm}
 %*********************************************************************************************************
@@
 The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
+represents  a workflow needed to implement a services' composition, identifying
+those entities that collaborate in the business processes (called {\sc Business
+Collaborators} \footnote{We use {\sc capitals} for referring to meta-models'
+classes.}) and the {\sc Actions} that  they perform. This model is represented
+by means of a UML activity diagram. Thus, as shown in Figure
+\ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling
+elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes}
+and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined
+by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc
+Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
 
@@
 
- \begin{figure}[htpb]
+ \begin{figure*}[htpb] 
 \centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
+\includegraphics[width=0.80	\textwidth]{figs/E-service-composition-metamodel}
 \caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
 \label{fig:e-scomposition-metamodel}
-\end{figure}
+\end{figure*}
 
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
+To illustrate the use of the $\pi$-SCM meta-model we used it for defining the
+{\em A-policy} based composition model of the ``To Publish Music'' scenario (see
+Figure \ref{fig:servicecompositionmodel}). There are three external business
+collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em
+italics} to refer to concrete values of the classes of a model that are derived
+from the classes of a meta-model.}). It also shows the business process  of the 
+``To Publish Music'' application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
 Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
@@
 An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
 
@@
 
+\begin{figure}[ht!]
+\centering
+\includegraphics[width=0.50\textwidth]{figs/e-composition-model}
+\caption{Services' composition model for the business service ``To publish
+music''}
+\label{fig:servicecompositionmodel}
+\end{figure}
 Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
+\section{$\pi$-PEWS  meta-model}\label{sec:pewsmetamodel}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
+The idea of the $\pi$-{\sc Pews} meta-model is based on the services'
+composition approach provided by the language PEWS~\cite{BaAM06,Placido2010LTPD}
+(\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
+Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model consisting of  classes representing:
 \begin{itemize}
@@
 %     
-\begin{figure}
+\begin{figure}[ht!] 
 \centering 
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
+\includegraphics[width=0.50\textwidth]{figs/PiPEWSMetamodel}
 \caption{$\pi$-{\sc Pews} Metamodel}
@@
 
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
+Table \ref{tab:transformation} shows the transformation principle between the
+elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
 There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
@@
 %{\bf\em  
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
+\subsection{Transformation of the services' composition elements of the
+$\pi$-SCM to the $\pi$-PEWS elements}
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
+A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em
+Action:name} is transformed to a  named class {\sc Operation} with a
+corresponding attribute name {\sc Operation:name}. A  named service activity
+represented by the elements {\sc\em ServiceActivity}  and  {\sc\em
+ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
+When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Table \ref{tab:transformation}):
 \begin{itemize}
@@
 
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
+In the scenario ``To Publish Music'' the service activity {\sf PublishMusic} of
+the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em
+UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook}
+and {\em Twitter}. Given that no {\sf ConstrolNode} is specified by the $\pi$-SC
+model, the corresponding transformation is the expression that defines a {\sf
+Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of
+the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
+
+\begin{table}
+\tiny
+\renewcommand{\arraystretch}{1.3}
+\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
+\label{tab:transformation}
+\centering
+\begin{tabular}{l|l|l} 
+    \hline
+    Source:  &  Mapping Rules & Target:\\
+    $\pi$-SCM  &   & $\pi$-PEWS\\
+    \hline 
+    \hline
+
+    Action   &  - An \textit{Action} in the source model    & Operation:\\
+       &  corresponding to an external \textbf{Business}  &alias\\
+       & \textbf{Collaborator} is mapped to an Operation&\\
+       & in target model.&\\
+       &   - The \textbf{Action:name} in the source model 
+       & \\ 
+       &   is transformed into \textbf{Operation:name}  & \\ 
+       &  in the target model. & \\
+    \hline
+
+        Service     & - The \textbf{ServiceActivity} in the source model & Type\\
+        Activity &  is mapped to a \textbf{Composite Operation} & Operation,\\ 
+        & in target model when more than one \textbf{Actions}& Composite\\
+        & are called.&Operation\\
+
+        & - If \textbf{Composite Operation} is generated for a  &  \\
+        &  given \textbf{Service Activity} then the &\\
+        & \textbf{ServiceActivity:name} in the source model&\\
+        &  is  mapped to \textbf{CompositeOperation:name}&\\
+        & in the target model.& \\
+    \hline
+
+    Control &  - The \textbf{Control Node} in the source model & Operator\\
+    Nodes & is mapped to a \textbf{Operator} in target model. &\\ 
+    & According to the type of \textbf{Control Node}  &\\ 
+    &(merge, decision, join, fork) the expression &\\
+    &of the  \textbf{Composite Operation} is:&\\
+    &  $\bullet$ Sequence if no \textbf{ControlNode} is specified;  & \\
+    &  $\bullet$ Parallel - Sequence for  a \textbf{ControlNodes} & \\ 
+    &pattern fork -
+    join; &\\
+    & $\bullet$ Choice - Sequence for a \textbf{ControlNodes} &
+    \\
+    &pattern decision - merge &\\
+    \hline
+
+    Business    & A \textbf{BusinessCollabortor:isExternal} in the &Namespace\\
+    Collaborator  &source model generates a \textbf{Namespaces} &\\
+     & in the target model&\\
+    \hline
+
+    Rule:event   &  The \textbf{Rule}'s attribute event in the source & Event
+    Type,\\ & model is transformed into an \textbf{Event:type} &Event\\
+    & of the target model. In this case attribute is & \\
+    &mapped to an entity with an attribute. &\\
+    &The \textbf{Event Type} of a \textbf{Rule} in the target& \\
+    & model is determined by the Rule type: & \\
+     &  $\bullet$ \textbf{Event Type} of a \textit{Precondition} \textbf{Rule} 
+     & \\ &is \textit{ActivityPrepared}; &\\
+     &  $\bullet$ \textbf{Event Type} of a \textit{Postcondition} \textbf{Rule} 
+     & \\ & is \textit{TermActivity};&\\
+    &  $\bullet$ \textbf{Event Type} of a \textit{TimeRestriction} \textbf{Rule}
+    & \\ & is \textit{Activity}; &\\
+    \hline
+
+      Rule:  &  The \textbf{Rule}’s attribute condition in the
+      & Condition\\
+      condition &  source model is transformed into a&\\
+      & \textbf{Condition:expression}in the target model.   &\\ 
+      &In this case, an attribute is mapped into an  &
+      \\
+      &entity with an attribute &\\
+    \hline
+
+    Rule:action   & The \textbf{Rule:action} in the source model is & Action\\
+    &transformed in an \textbf{Action:act} in the target &\\
+    &  model.  The attribute action is mapped to & \\
+    & an entity with an  attribute. In the target
+    & \\ 
+    &  model an action is executed according to& \\
+
+    & the rule condition value (true/false).&\\
+    \hline
+
+    Policy   & - Every \textbf{Policy} associated to an element &
+    APolicy\\
+    & (\textbf{Business Collaborator,} \textbf{Service, Activity, }&\\
+    &
+     \textbf{Action}) in the source model becomes an 
+    &\\ 
+    & \textbf{APolicy}  associated to the corresponding & \\ 
+    & element in the target model. &\\
+     & - The name attribute of a \textbf{Policy} in the & \\
+     & source model becomes an \textbf{Apolicy:name} & \\ 
+     & of the  target model.&\\
+    \hline
+
+    Variable   &  Every \textbf{Variable}, and its attributes, &
+    Variable\\
+    & associated to a  \textbf{Policy} in the source &\\
+    &model becomes a \textbf{Variable} associated to an &\\
+     &\textbf{APolicy} in the target model.  &\\
+     &The variables can be used in an \textbf{APolicy}'s &\\
+     &Condition of the target model.&\\
+    \hline
+
+    Rule:event &  For a \textbf{Rule} in the source
+    model, depending & Precondition, \\
+    &on the \textbf{Event Type}, the corresponding& Postcondition, \\
+    &  transformation in the target model is:
+     &Time\\
+    & \textbf{Precondition,} \textbf{Postcondition} or  &Restriction, \\
+    &\textbf{Time Restriction Rule}&Rule\\
+    \hline
+
+ \end{tabular} 
+\end{table}
+
+%  \begin{table}
+% \tiny
+% \renewcommand{\arraystretch}{1.3}
+% \caption{Transformation Rules: From $\pi$ServiceComposition to $\pi$PEWS}
+% \label{tab:transformation}
+% \centering
+% \begin{tabular}{l|l|l}
+% 
+%     
+%     Policy   & - Every \textbf{Policy} associated to an element
+%     (\textbf{BusinessCollaborator, Service, Activity, Action}) in the   &
+%     APolicy\\
+%     & source model becomes an \textbf{APolicy} associated to the corresponding
+% element in the target model. & \\
+%    & - The name attribute of a \textbf{Policy} in the source model becomes an
+% \textbf{Apolicy:name} of the target model. & \\ 
+%     \hline
+
+%     Variable   &  Every \textbf{Variable}, and its attributes, associated to a
+%     \textbf{Policy} in the source model
+%     becomes a \textbf{Variable}   & Variable\\
+%     &associated to an \textbf{APolicy} in the target 
+%     model. The variables can be used in an \textbf{APolicy}’s Condition &\\
+%     & of the target model.&\\
+%     \hline
+
+%     Rule:event &  For a \textbf{Rule} in the source
+%     model, depending on the \textbf{EventType}, the corresponding transformation
+%     & Precondition \\ &   in the target model is: \textbf{Precondition,
+%     Postcondition} or \textbf{TimeRestriction Rule}  & Postcondition\\
+%     &&TimeRestriction\\
+%     &&Rule\\
+%     \hline
+% \end{tabular}
+% \end{table}
+
+% \begin{figure*}[ht!]
+% \centering{
+% %\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
+% \includegraphics[width=0.70\textwidth]{figs/Mapping-1}}
+% \caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
+% \label{fig:transformations}
+% \end{figure*}
 
@@
 %{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
+\subsection{Transformation of rules grouped by A-pol\-i\-cies   in the $\pi$-SCM
+to A-Policies of  $\pi$-PEWS}
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
@@
 
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
+As shown in Table \ref{tab:transformation}, for an event of type {\sc\em Pre}
+the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
 The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
@@
 %Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
+In the scenario ``To Publish Music'' the {\sf Policies} {\em OAuthPolicy} and
+{\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
 Thus in both cases the events are of type {\sf ActivityPrepared}.
@@
 %*********************************************************************************************************
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
+This section  describes the $\pi$SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
 For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
+Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
 
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-SOD-M Development Environment}
+\subsection{$\pi$SOD-M Development Environment}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 
-Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
+Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
 The environment implements the abstract architecture shown in Figure \ref{fig:sodm}.
@@
 	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
+		\includegraphics[width=0.43\textwidth]{figs/architecture}
 	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
+		\caption{$\pi$SOD-M Development Environment}
    \label{fig:policymanager}
@@
 \begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
+\item We  used the Eclipse Modeling Framework (EMF)\footnote{The EMF project is
+a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
 Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
@@
 An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
+for  developing the mapping plug-in implementing the  mappings between models
+($\pi$-Ser\-viceCom\-po\-si\-tion\-2\-$\pi$-PEWS Plug-in).
 
@@
 
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Validation}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-We validated our approach  by implementing the "To Publish Music" application.
+% %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+% \subsection{Validation}
+% %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+% 
+% We validated our approach  by implementing the "To Publish Music" application.
 It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.
@@
 Our implementation includes a  code generator that generates executable code for a reliable services' composition.
-The model represents an services' composition expression and its associated policies.
-
-We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
+% The model represents an services' composition expression and its associated policies. 
+% 
+% 
+% We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
 Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available.
@@
 %\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
+%\subsection{Programming non-functional properties for services}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
+Current standards in services' composition implement functional, non-functional
+constraints and communication aspects by combining different languages and
+protocols. WSDL and SOAP among others are languages used respectively for
+describing services' interfaces and message exchange protocols for calling
+methods exported by such services.
+For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTrans\-ac\-tion.
 The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
@@
 This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
+Other works, like \cite{helga2} introduce a model for transactional services composition based on an advanced transactional model.
+\cite{samy} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services.
+In \cite{Vid04} the model
 introduced in \cite{schuldt-etal-TODS} is extended to web services
@@
 %\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
+%\subsection{Modeling non-functional properties}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
@@
 %\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
+% \subsection{Discussion}
+% %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+% As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
 In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
@@
 In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
+% 
+% Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
 
@@
 %*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
+This paper presented $\pi$SOD-M for specifying and designing reliable service based applications.
 We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
@@
 
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
+Non-functional constraints are related to business rules associated to the
+general ``semantics'' of the application and in the case of services' based
+applications, they also concern the use constraints imposed by the services.
 We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications.
@@
 
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
+\bibliographystyle{abbrv}
 \bibliography{biblio} 
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
 \end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-09-30 SistReview/paperToSubmit.tex
inicial --martin

\documentclass{sig-alternate}
  \pdfpagewidth=8.5truein
  \pdfpageheight=11truein

\begin{document} 
%
% --- Author Metadata here ---
\conferenceinfo{SAC'14}{March 24-28, 2014, Gyeongju, Korea.}
\CopyrightYear{2014} % Allows default copyright year (2002) to be over-ridden - IF NEED BE.
\crdata{978-1-4503-2469-4/14/03}  % Allows default copyright data (X-XXXXX-XX-X/XX/XX) to be over-ridden.
% --- End of Author Metadata ---

\title{Non-Functional Requirements for Service-Based Applications: A Systematic Review}

% \subtitle{[Extended Abstract]
% \titlenote{A full version of this paper is available as
% \textit{Author's Guide to Preparing ACM SIG Proceedings Using
% \LaTeX$2_\epsilon$\ and BibTeX} at
% \texttt{www.acm.org/eaddress.htm}}}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in  
% the body of your article BEFORE References or any Appendices.

\numberofauthors{1} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
%\alignauthor
%       Pl\'acido A. Souza Neto\\
%	   \affaddr{Federal Institute of Rio Grande do Norte (IFRN)},\\
%       \affaddr{Natal-RN, Brazil}\\
%       \email{placido.neto@ifrn.edu.br}
%\alignauthor
%       Martin A. Musicante\\
%       \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
%       \affaddr{Natal-RN, Brazil}\\
%       \email{mam@dimap.ufrn.br}
%\alignauthor Genoveva Vargar-Solar\\
%       \affaddr{French Council of Scientific Research}\\
%       \affaddr{LIG-LAFMIA }\\
%       \affaddr{Saint Martin d'H\`{e}res, France}\\
%       \email{Genoveva.Vargas-Solar@imag.fr}
%\alignauthor Valeria de Castro\\
%       \affaddr{Universidad Rey Juan Carlos}\\
%       \affaddr{M\'{o}stoles, Spain}\\
%       \email{Valeria.deCastro@urjc.es}
%\and
%\alignauthor
%       Umberto Souza da Costa\\
%       \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
%       \affaddr{Natal-RN, Brazil}\\
%       \email{umberto@dimap.ufrn.br}        
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\date{21 September 2013}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract} 
This paper presents a systematic literature review of non-functional
requirements (NFRs) for service-based applications. 
The main goal of the review is to identify the most common terms used to refer to NFRs.
We also propose a classification of the most common terms, as well as, a model to integrate those terms.
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]

%\terms{Delphi theory}

\keywords{Non-functional requirements, Service-based development, Service compositions, Web services, Systematic review.}

\section{Introduction}

Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.
Functional properties are not the only crucial aspect in the software development process. 
Other properties need to be addressed to fit in the application with its context.
These other aspects are called Non-Functional Properties.

Non-Functional Requirements (NFRs) specify those properties that are not addressed by the functional  specification.
They are often called \textit{qualities} of the software system.
%They are also referred as ``constraints'', ``quality attributes'', ``quality goals'', ``quality of
%service requirements'' and ``non-behavioural requirements'' \cite{Stellman2005}.
Non-Functional Requirements may specify response time, security constraints or quality of the solution, among others.

Service-Oriented Computing~\cite{Papazoglou2007}, is a software development paradigm where pre-existing services are combined to produce more complex applications. 
%The selection of services is usually guided by the functional requirements of the application. 
%In this scenario, there exists the need to provide support for the specification of non-functional requirements, such as
%security, reliability, and efficiency. 
The development of service-based applications can benefit from the inclusion of NFRs to the software process from its early stages.
Failure to comply with this inclusion means that the final application is obtained from a partial specification, making the deployment a difficult task.
%Ideally, non-functional requirements
%would be considered along with all the stages of the software development. 
The adoption of non-functional specifications from the early states of development
can help the developer to produce applications that are capable of dealing with
their context.
Non-functional properties of service oriented applications have been
addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
Different proposals~\cite{Babamir2010,AgarwalLS09,CholletL09,GutierrezRF10,XiaoCZBOLH08,JeongCL09,TsadimasNA12}
support non-functional requirements in the context of web service development.

Most software development methods define software processes that use the notion of refinement.
Software process begins with the formulation of an abstract specification, which is successively refined to yield the implementation of the system.
Methods for the development of web service applications are no exception to this rule.
At least two levels of abstraction can be distinguished: a \textit{Business Level}, including the abstract specification, and a \textit{System Level}, including actual computer programs that implements the system.

In the case of web service applications we will distinguish two separate layers of the implementation.
The \textit{Composition Layer} is the upper layer of the implementation. 
It defines the workflow of the system, in terms of individual service calls.
The \textit{Service Layer} defines those services that are called by the composition.

\bigskip

In this work we investigate the extent in which NFRs are considered by the development methods proposed in the literature.
We have conducted a systematic review~\cite{Kitchenham08} to summarize the approaches that support NFRs.
As a result, we propose a classification of NFRs for web services.

\bigskip

This paper is organized as follows: 
Section~\ref{sec:sr} presents a systematic literature review of non-functional requirements for service-based applications.
The findings of the systematic review are presented in Section~\ref{sec:nfindings} and used in Section~\ref{sec:classification} to propose a classification and a model for NFRs in the context of web service development. 
Section~\ref{sec:conclusion} concludes the paper.

\section{A Systematic Review}  
\label{sec:sr}

In this work we develop a systematic literature review of NFRs for web services. 
Our review considers the abstraction levels and implementation layers defined above. 
%
%We present a systematic review of the non-func\-tion\-al
%requirements and properties used in the context of service-oriented
%development. 
The purpose of our analysis is to: 
\textit{(i)} Identify concepts, properties and notations related to NFRs and used in service-based system development; and 
\textit{(ii)} Define a classification for these concepts, properties and notations, at different  levels of abstraction.

%For each phase of the software process, the requirements and non-functional
%requirements are refined. 
%As a result of this process, the granularity of the
%concepts described by the requirement becomes thinner and more precise.

% The focus of a modeling-based non-functional requirements becomes important
% because it reveals a continuing need on the development of distributed
% service-oriented. Thus, it is necessary identify how non-functional requirements
% are classified in different development process phases and iterations.

% Thus, this section also presents the collected data and results
% from the research work analysis according to some research questions. Once the
% sources' extraction execution had been completed, it was important to be sure
% that multiple publications of the same approach were not included in the data analysis. Tables \ref{tab:result02}
% and \ref{tab:result03} show the results of each work under each question
% described in section \ref{subsec:question}.

% % Among the properties presented we highlight the security
% % and performance properties. All users want to access data
% % securely and quickly. Most studies analyzed have both properties. Reliability is
% % also an important non-functional requirement presented in some papers and needed
% % to end users.

% 
% The analisys was carried out by effectuating the following activities:
% \textit{(i) question formulation; (ii) source selection; (iii) selection
% process; (iv) information extraction;} and \textit{(v) results}.

We propose seven research questions ($RQ_1$ to $RQ_7$) to guide our analysis about non-functional
requirements for web services. 
For each question we define a set of possible answers in order to guide
the analysis.
%These possible answers are defined from our knowledge on each topic. 
The questions are:
\begin{itemize} 
  \item \textbf{\texttt{$RQ_1$:}} How NFRs are modelled by existing
  methodologies for Web services?
  %\begin{trivlist}
%	  \item 
Possible answers: \textit{Answer is specific to each proposal.}
	%\end{trivlist}  
 \item \textbf{\texttt{$RQ_2$:}} Which kinds of NFRs are considered more frequently?
 	%\begin{trivlist}
	  %\item 
	  Possible answers: \textit{Security / availability / portability / \ldots / reliability /
	  performance.}
	%\end{trivlist}
  \item \textbf{\texttt{$RQ_3$:}} What is the main underlying approach used by the proposal?
	%\begin{trivlist}
%	  \item 
Possible answers: \textit{Model driven approach (MDD) / Ontologies (Ont) / Formal methods (FM) / Artificial intelligence (AI) / Business Process Modeling (BP) / Traditional (TDT).}
	%\end{trivlist} 
  \item \textbf{\texttt{$RQ_4$:}} What is the scope of the proposal?
  %\begin{trivlist}
%	  \item 
Possible answers: \textit{Software architecture / QoS model / Language definition / Methodology / etc.}
	%\end{trivlist}	
  \item \textbf{\texttt{$RQ_5$:}} Does the paper propose a (meta)model
  for NFRs? Does the Business Level specification include NFRs?  
%\begin{trivlist}
%	  \item 
Possible answers: \textit{yes / no} -- \textit{yes / no.}
%	\end{trivlist}  
  \item \textbf{\texttt{$RQ_6$:}} Are non-functional aspects considered at the composition or single service level?
%\begin{trivlist}
%	  \item 
Possible answers: \textit{single / composition.}
%	\end{trivlist}
\item \textbf{\texttt{$RQ_7$:}} What is the publication year of the paper?
%	\begin{trivlist}
%	  \item 
Possible answers: \textit{Year of publication.}
%	\end{trivlist}	   
\end{itemize}

We use the approach described in~\cite{Kitchenham08} for searching,
collecting and selecting works related with NFRs for service-based applications. 
The results are used to identify and compare key concepts related to NFRs.
The search engines used in our review include journals, conferences and
workshops of recognized quality. 
We used the following search engines: \textit{(i) IEEE Computer; (ii) ACM Digital Library;} and
\textit{(iii) Science Direct}.

The query used to perform the search is defined as: 
 \begin{center}
\texttt{(((non functional properties) OR (non functional requirements))
AND web service AND composition))}
\end{center}

The results of our query for each search engine were filtered in accordance to the following criteria:
\begin{enumerate}
\item[1.] Papers written in other languages than English were excluded.
\item[2.] Papers published in non-international conferences and workshops were also excluded.
\item[3.] Papers whose title and abstract do not refer to NFRs were excluded. 
\end{enumerate}

After this filtering process, the remaining papers were analyzed in full.
A brief account of our findings is given in the next section.

\section{Our Findings}   
\label{sec:nfindings}

In this section we analyze the results of the search query for our systematic review.
The search was performed on three different scientific sources, namely IEEE Computer, ACM Digital Library (only journal papers) and Science Direct.
We first show the contribution of each search engine, in terms of the number of relevant papers and year of publication.
Next, we summarize the answers we have obtained to the questions of our systematic review.

Table \ref{tab:result01} shows the number of papers obtained at each stage
of the selection procedure, for each source. 
The first row shows the total number of papers returned by each engine, for the search query (Section~\ref{sec:sr}).
The next row of the table shows the results obtained after filtering the papers by considering their titles and abstracts.
The percentage figures indicate the proportion of remaining papers, for each source.
The last row of the table shows the number of relevant papers obtained, after considering the complete text of each paper.
The percentages are relative to the total, for each source.
The distribution of publication year for the 26 selected papers is shown in Figure~\ref{fig:statistics}.

\begin{table}[ht!]
\tiny
\centering
\begin{tabular}{l|c|c|c|c}
  \hline
  \hline
   & IEEE & ACM  & Science Direct & Total \\
  \hline
  \hline
  Total  & 76 &  88  & 248 & 412 \\
  \hline
  Title/Abstract & 25 (32.9\%) & 12 (13.6\%) & 30 (12.1\%)  & 67 (16.3\%) \\
  \hline
  Full text & 11 (14.5\%) & 3 (3.4\%) & 12 (4.8\%) & 26 (6.3\%)\\
  \hline
  \hline
\end{tabular}
\caption{Number of papers per source.}
\label{tab:result01}
\end{table}

\begin{figure} 
\centering
\includegraphics[width=.49\textwidth]{figs/NumberPapersYear.png}
\caption{Publications per year.}
\label{fig:statistics}
\end{figure}

For each of the 26 papers, we obtained answers for each research question.
Our findings are presented in Tables~\ref{tab:result02} and~\ref{tab:result03}.

\begin{table*}[ht!]
\centering
%\scriptsize
\tiny
\begin{tabular}{l|l|l|c|l}
  \hline 
  \hline
   \textbf{Reference} & \textbf{$RQ_1$:\textit{NFR concepts}} &
   \textbf{$RQ_2$:\textit{Properties}} & \textbf{$RQ_3$:\textit{Approach}} &
   \textbf{$RQ_4$:\textit{Domain / Scope}} 
   \\
  \hline
  \hline  
  Babamir et al. \cite{Babamir2010} &  property / category / constraint & responsiveness /
  availability   & TDT & Software architecture
 \\  
  &  & performance / sla properties   &  & 
 \\ 
  \hline

  Yeom et al. \cite{Yeom2006} & category / sub-category / & business value / performance / stability /
   & TDT & QoS model\\ 
   & property & manageability /security / business processing  & & \\
    &  & interoperability & & \\ 
  \hline

  Xiao et al. \cite{XiaoCZBOLH08} &  NF attribute  & time /
cost / resource  & BP  & Business processes modeling
   \\   
  \hline 
  D'Ambrogio \cite{DAmbrogio06} &  characteristics / category / dimension & availability /
  reliability / access control & MDD & WSDL extension
 \\
   \hline
  Chollet et al. \cite{CholletL09} & activity / NF attribute & security  & MDD 
  & Orchestration Framework
   \\

  \hline
  Schmeling et al. \cite{SchmelingCM11} & NF concern / NF attribute /& security &
  MDD &  Web service 
  \\
  & NF action / NF activity &  &
   & composition process
  \\ 
  \hline
    Thi{\ss}en et al. \cite{ThissenW06} & NF value  &  performance /
  reliability / &  FM & Software architecture \\  
   & & cost / availability &  & \\  
  \hline
  Zhang et al. \cite{ZhangPSP05} & attribute / predicate & security & FM &  Access control 
   \\
  \hline
  Basin et al. \cite{BasinDL06} &  attribute & security  & MDD & System
  architecture
  \\
  \hline
  Ceri et al. \cite{CeriDMF07} & police / rule / condition / action & n.a. & TDT & Context-aware
  applications
  \\
  \hline
  Fabra et al. \cite{Fabra2011} & property & storage /
  processing & MDD & Web service methodology 
  \\
  \hline
  Modica et al. \cite{ModicaTV09} & quality level & sla
  properties & TDT &  Service oriented architecture
  \\
  \hline
  Ovaska et al. \cite{OvaskaEHPA10} & attribute / category & security / reliability
  & MDD &  Model development
  \\
  \hline
  Agarwa et al. \cite{AgarwalLS09} & property / policy / function & \textit{not explicitly
  defined} & Ont & Policy language
  \\
  \hline
  Jeong et al. \cite{JeongCL09} & NF attribute & operation cost / performance / availability /& AI & Service oriented architecture\\
  &  &  accessibility / security / interoperability / &  &  \\
   &  & usability / user satisfaction & &
\\
  \hline
 Pastrana et al. \cite{PastranaPK11}  & NF property / contract  / & performance / reliability / scalability / & Ont & Web service methodology\\
   &assertion / NF behaviour &  capacity / robustness / precision / security /&  & \\
  &  & accessibility / availability / interoperability    &  & \\    
  \hline
  Diamadopoulou et al. \cite{DiamadopoulouMPS08} & NF characteristic &
  user' subjective perception & TDT & Web service selection  \\ 
  \hline
  Gutierrez et al. \cite{GutierrezRF10} & NF factor / NF sub-factor & Security  & BP &  WS development process
  \\  
  \hline
  Mohanty et al. \cite{MohantyRP10} & NF attribute / NF factor & reliability /
  performance / integrity / & AI & Artificial intelligence /\\ 
  &  &  usability / response time / documentation &   & Web services classification	 \\
  \hline  
   Karunamurthy et al. \cite{Karunamurthy2012787} & Non-functional parameters &
   cost / response time / availability /  security /   & BP &  DSL (NFSL)
\\
   &  & availability / reliability / reputation   &  & 
  \\  
  \hline  
   Liu et al. \cite{Liu20121080} & QoS parameter &
   cost / execution duration / accuracy /  & FM  &  QoS model
\\
   &  &  security / integrity / availability / reliability   &   & 
  \\
  \hline

   Tran et al. \cite{Tran2012531} & QoS policies &
   performance / availability / security /  & MDD  &  Language definition /
\\
   &  &  response time /  SLA aspects  &   &  QoS model
  \\  
  \hline

   Wang et al. \cite{Wang2012} & Non-functional properties &
   response time / price / reliability /   & N/A  &  Genetic Algorithm /
\\
   &  & availability / platform / location / provider   &   &  QoS model
  \\  
  \hline

     Li et al. \cite{Li2013} & Dimensions / &
  execution time / storage / reliability /  & TDT  &   QoS model
\\
   & QoS parameters & service cost / communication time /    &   &  
  \\

   &  & message length / availability    &  & 
  \\  
  \hline

      Kamalabad et al. \cite{Kamalabad2012} & QoS Attributes &
   reliability / availability / response time /   & Ont  &  Business

\\
   &  & performance / stability / accuracy /    &   &  Specification
  \\

   &  & capacity / robustness / cost /   &  & 
  \\
  &  & scalability / throughput /   &  &
  \\
  &  & efficiency / accessibility / successability /   &  & 
  \\
  &  & reputation / consistency / delivery time    &  & 
  \\  
  \hline

   Rumpel et al. \cite{Rumpel2012} & Quality Requirement /  &
  runtime / cost management / security   & N/A  &  Requirement definition 
\\
   & Quality Property  &  &   &  
  \\  
  \hline  
  \hline

\end{tabular} 
\caption{Research question results - $RQ_1$, $RQ_2$, $RQ_3$, $RQ_4$.}
\label{tab:result02}
\end{table*}

Let us now present a brief description of the terminology used by each of the analyzed papers to refer to the different aspects of NFRs.
In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view, namely,  
\textit{service level}, \textit{system level} and \textit{business level}.
In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}. The work also defines a \textit{web service quality model}, which considers non-functional properties.

%At the business level, the quality properties considered are 
%\textit{service charge}, \textit{compensation rate}, \textit{penalty rate} and    \textit{reputation}.
%At the service level, the quality properties are 
%\textit{performance} and \textit{stability}.
%At the system level, the quality properties are 
%\textit{manageability}, \textit{interoperability}, \textit{business processing} and
%\textit{security}.

In~\cite{XiaoCZBOLH08} the authors use the terms  
\textit{non-functional attributes}, \textit{composition mo\-del entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
The notion of non-functional attribute is used to describe NFRs of the abstract process model. In the lower level, the composition is annotated with non-functional attributes.

D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}. 
\textit{Quality dimensions} are used to quantify an individual characteristic.
For instance, the quality category \textit{performance} groups characteristics such as
\textit{latency} and \textit{throughput}. 
The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services. A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties as \textit{availability}, \textit{reliability} and \textit{access control}.

Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolset for specifying and implementing web service compositions with support to several NFRs. The term \textit{non-functional concern} (NFC) is used to denote  NFRs. 
%Two aspects are considered: specification and realization.
\textit{Non-functional concern} is a general term used to describe non-functional requirements. 
For instance, \textit{security}, \textit{reliability}, \textit{transactional behavior} are non-functional requirements. 
A \textit{non-functional action} represents some behavior that implements \textit{non-functional attributes}. 
An example of \textit{non-functional action} is \textit{encryption}, which provides the implementation of the \textit{non-functional attribute} \textit{confidentiality}. 
Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.

Pastrana et al.~\cite{PastranaPK11} use the term \textit{contract} to describe non-functional requirements. 
\textit{Contracts} may have pre-conditions, post-conditions and invariants. 
Each contract defines \textit{assertions} associated with \textit{quality properties}. 
Each service may have as many associated \textit{contracts} as needed.

\begin{table}[ht!]
\centering
\tiny
\begin{tabular}{l|l|l|l}
  \hline 
  \hline
   \textbf{Reference} & $RQ_5$:\textbf{\textit{WS model}} --
   & $RQ_6$:\textbf{\textit{Service type}}  &
   $RQ_7$: 
   \\
    &   \textbf{\textit{Business services}} &   & \textbf{\textit{Year}}
%   \textbf{\textit{publication}} 
   \\
  \hline
  \hline  
  Babamir et al. \cite{Babamir2010} & no -- yes  & composition & 2010   
 \\  
  \hline   
  Yeom et al. \cite{Yeom2006} & yes -- yes & single   & 2006  \\  \hline
  Xiao et al. \cite{XiaoCZBOLH08} & no -- no & composition    & 2008
   \\ 
  \hline 
  D'Ambrogio \cite{DAmbrogio06} & yes  -- no & composition  & 2006 \\
   \hline
  Chollet et al. \cite{CholletL09} & yes -- yes & composition  &  2009 \\
  \hline 
  Schmeling et al. \cite{SchmelingCM11} & no -- no & composition &  2011 \\ 
  \hline
   Thi{\ss}en et al. \cite{ThissenW06} & no -- yes & composition & 2006
   \\
  \hline
  Zhang et al. \cite{ZhangPSP05} & no -- no  & single  & 2005 \\ 
  \hline
  Basin et al. \cite{BasinDL06} & yes -- no & single / composition  & 
  2006\\
  \hline 
  Ceri et al. \cite{CeriDMF07} & no -- no  & single  &  2007\\ 
  \hline 
  Fabra et al. \cite{Fabra2011} & yes -- yes & composition  &  2011\\
  \hline
  Modica et al. \cite{ModicaTV09} & no -- no & composition & 2009\\ 
  \hline
  Ovaska et al. \cite{OvaskaEHPA10} & yes -- no & single & 2010\\
  \hline
  Agarwa et al. \cite{AgarwalLS09} & yes -- no & single / composition  &  
  2009\\
  \hline
  Jeong et al. \cite{JeongCL09} & no -- no  & composition &  2009\\
  \hline
  Pastrana et al. \cite{PastranaPK11} & yes -- no & composition &  2011 \\
  \hline
  Diamadopoulou et al. \cite{DiamadopoulouMPS08} & no -- no & composition 
  & 2008\\
  \hline
  Gutierrez et al. \cite{GutierrezRF10} & no -- no &  single / composition 
  & 2010\\
  \hline
  Mohanty et al. \cite{MohantyRP10} & no -- no  & single  & 2010\\
  \hline
   Karunamurthy et al. \cite{Karunamurthy2012787}& no -- no  & composition  &
   2012\\

  \hline
   Liu et al. \cite{Liu20121080}  & no -- no & single / composition  &  
  2012\\
  \hline
  Tran et al. \cite{Tran2012531} & yes -- yes  & composition &  2012\\
  \hline
    Wang et al. \cite{Wang2012} & no -- no & composition &  2012 \\
  \hline
   Li et al. \cite{Li2013} & no -- no & composition 
  & 2013\\
  \hline
    Kamalabad et al. \cite{Kamalabad2012} & yes -- no &  composition 
  & 2013\\
  \hline
   Rumpel et al. \cite{Rumpel2012}  & yes -- no  & composition  & 2012\\
  \hline
  \hline  
\end{tabular}
\caption{Research question results - $RQ_5$, $RQ_6$, $RQ_7$.}
\label{tab:result03}
\end{table}

\begin{figure*}[ht!]  
\centering  
\includegraphics[width=0.70\textwidth]{figs/nfrRelationship.png}
\caption{Relationship between  NFR Concepts.}
\label{fig:nfr-relationship}   
\end{figure*}

Chollet et al.~\cite{CholletL09} associate (non-functional) \textit{quality properties} to 
(functional) activities. The authors present a security meta-model for web service
composition. The NFRs considered are \textit{authentication}, \textit{integrity} and \textit{confidentiality}. Each NFR is associated with a service activity.

Ceri et al.\cite{CeriDMF07} uses the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services. 
Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}. 
Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
Other authors do not define specific terms to refer to NFRs. 
They use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09}, 
\textit{property}~\cite{Fabra2011}, 
\textit{factor}~\cite{MohantyRP10,GutierrezRF10}, 
\textit{characteristic}~\cite{DiamadopoulouMPS08}, 
\textit{quality level}~\cite{ModicaTV09}, and
\textit{value}~\cite{ThissenW06,BasinDL06}.

% The purpose of this analysis is to identify the properties and non-functional
% requirements used in the development of systems, looking for patterns
% or relationships between non-functional concepts in different
% modeling levels.

Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
However, distinct hierarchies and models are proposed for NFRs,  according to different points of view.
We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.

%Although there exist many different types of notation used for classify
%non-functional requirements, in general, the values associated to them are the
%same, \textit{i.e.} \textit{security, performance, reliability, usability, availability}, etc. What
%distinguishes the different approaches is the adoption of different NFR
%hierarchies in order to prioritize or classify the quality requirements. 
%Another interesting factor
%is that the each work uses different approaches to model these requirements.
%Thus, there are different kinds of notations for non-functional
%requirements.

Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications. This work  presents a complete development methodology, although this methodology is not centered on NFRs.
The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account. 
In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs, 
whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling (BPM) is used for
system specification, including NFRs. 
The majority of the authors concentrate on the  modeling of service compositions, although a significant number of approaches is focused on the definition of NFR models.

In the method defined in~\cite{XiaoCZBOLH08}, tasks in the process model can be 
annotated with \textit{non-functional attributes} (NFAs). 
NFAs are defined apart  and are concerned with data items or tasks. 
NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.

The proposal in~\cite{ThissenW06} presents steps for  selecting services 
by taking QoS information into account. The proposed steps are: 
\textit{(i)} identification of relevant QoS information; 
\textit{(ii)} identification of basic composition patterns and 
QoS aggregation rules for these patterns; and 
\textit{(iii)} definition of a selection mechanism of services. 
The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
\textit{availability}.

Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.  
The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.

Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.  
In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.

Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.  
For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.

\section{Classification of NFR}
\label{sec:classification}

Our systematic review is useful to identify those aspects of the NFR treatment in web service development that are consensual among authors.
These aspects include terminology to denote concepts as well as the relationships among them.

Most works agree on distinguishing three points of view, namely the point of view of the organization (or Business view), of the individual service providers (or Service view) and of the composition designer (or System view).
The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
Service and System views are concerned with the implementation of the software solution: The Service level is concerned with the building blocks of the application.
It may use web services provided by third party sources.
The System level is concerned with the coordination of services, to implement the business logic.

Table \ref{tab:result04} shows the NFR classification we propose, in accordance to these points of view.
The second column of the table presents the term we suggest for NFRs at each level.
The last column contains the most common terms found in the literature for specific classes of NFRs at each level.

\begin{table}[ht!]
\centering
%\scriptsize
\small
\begin{tabular}{l|l|l}
  \hline 
  \hline
   \textbf{\textit{Viewpoint}} & \textbf{\textit{Term}} & \textbf{\textit{Concepts}}  
   \\
  \hline
  \hline   
  Business View & Constraint  & Business Constraint,    
 \\  
  &   & Value Constraint\\
   \hline

   &  & Security, Performance,\\ 
   & & Interoperability, Scalability,\\
  System View & Policy & Reliability, Usability,\\
   & & Transactional Behaviour,\\
   & & Availability \\ 
   \hline
     &  & Integrity, Transaction,  \\
   &  & Accessibility, Encryption, \\
   &  & Cost, Time Constraint, \\
  Service View & Contract & Encryption, Platform, \\
   &  & Privacy, Authentication, \\
   &  & Resource, Capacity, \\
   &  & Privacy, Confidentiability 
   \\   \hline
  \hline  
\end{tabular}
\caption{NFR Classification.}
\label{tab:result04}
\end{table}

The relationship between concepts of the three levels is depicted in Figure~\ref{fig:nfr-relationship}.
The \textit{Contracts} bound to services should be devised to implement \textit{Policies} (at the System level).
These policies are used to guarantee business \textit{Constraints}.

Another classification of NFRs can be found in~\cite{Yeom2006}, which does not considers NFRs over data (but just over functions and service performance). 
We consider that it is important to classify the requirements of business and data (value) restrictions, since data processing is an essential part of the web service execution.

In the next subsection we present a NFR-centred meta-model for web service applications.

\begin{figure}[ht!]  
\centering  
\includegraphics[width=0.45\textwidth]{figs/metamodelo.png}
\caption{NFR Model.}
\label{fig:NRFmodel} 
\end{figure}

\subsection{NFR Meta-Model} 
\label{sec:nfr-metamodel}

The model presented in Figure~\ref{fig:NRFmodel} shows the relationship between
those concepts
we consider important for quality requirements used
in service-based development.

In this model, one \texttt{Requirement} (functional or non-func\-tion\-al) can be represented by one or more use cases. 
Each use case represents a \texttt{Service Activity}. 
Each use case has \textit{business} or
\textit{value} constraints. 
\textit{Business constraints} are restriction on functions and how they may be implemented. The \textit{value constraints} are restrictions on
the service interface, which the desired values for input and output data. Each
constraint is associated with NFAs.

A \texttt{Contract} is a set of constraints for the same function. For example,
a contract for the payment operation. The constraints for payment are: (i) the
value amount should not be less than 10 euros and (ii) the user should always
receive a purchase confirmation by phone message. This restrictions are
grouped into a single contract for payment verification.

% These constraints
% are also associated with non-functional attributes, e.g., reliability,
% transactions and data privacy card.

An \texttt{Exceptional Behavior} happens when a contract is not respected. When
this happens a new function is called or the process is stopped. For example, if
the bank does not authorize the payment, the system offers alternative forms
of payment such as PayPal.

Finally a \texttt{Policy} groups similar contracts. For example, security
contracts are grouped into a security policy and performance contracts are
grouped into a performance policy.

\section{Conclusions}
\label{sec:conclusion}

Notice that \textit{security} and \textit{performance} are the most frequently considered NFRs.
\textit{Reliability} is also present in most proposals.

This paper presented a systematic review of NFRs associated with web services.
We grouped the NFRs according to their characteristics and analyzed them in each
context of application. 
We identified the main NFR concepts associated with service-based development. 
We also presented a synthesis of the most common concepts found in the literature. 
We suggest a characterization of NFR-related concepts and a classification by considering three points of view: \textit{business}, \textit{services} and \textit{system}.

Our analysis and classification can help to improve the development of
service-based applications, by focusing on the guarantee of quality requirements.

We believe that the specification of NFRs from the early stages of the software process
can improve the quality of the solution.

\bibliographystyle{abbrv}
\bibliography{biblio}

\end{document}

### 2013-09-30 WEH2012/src/weh'12.tex
inicial --martin

\documentclass[10pt, conference, compsocconf]{IEEEtran}

% Add the compsocconf option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

\usepackage{balance}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Handling Contract Violations in Java Card Using Explict Exception Channels}

% author names and affiliations
% use a multiple column layout for up to two different
% affiliations

\author{\IEEEauthorblockN{Juliana Ara\'ujo, Rafael Souza, N\'elio Cacho,
Anamaria Moreira} 
\IEEEauthorblockA{Computing and Applied Mathematics Department (DIMAp)\\ 
Federal University of Rio Grande do Norte (UFRN)\\
 Natal-RN, Brazil\\
 \{juliana,rafael,neliocacho,anamaria\}@dimap.ufrn.br}
\and

\IEEEauthorblockN{Pl\'acido A. Souza Neto}
\IEEEauthorblockA{Information Technology Department (DIATINF)\\
Federal Institute of Rio Grande do Norte (IFRN)\\
Natal-RN, Brazil\\
placido.neto@ifrn.edu.br}
}

% make the title area
\maketitle

\begin{abstract}
Java Card is a version of Java developed to run on devices with severe storage and processing restrictions. The applets that run on these devices are frequently intended for use in critical, highly distributed, mobile conditions. This requires runtime verification approach based on Design by Contract to improve the safety of Java Card applications. However handling contract violation in Java Card applications is challenging due to their communication structure and platform restrictions. Additionally the Java Card exception handling mechanism requires that developers understand the source of an exception, the place where it is handled, and everything in between. As system development evolves, exceptional control flows become less well-understood, with negative consequences for the program maintainability and robustness. In this paper, we claim that such problem can be addressed by implementing an innovative exception handling model which provides abstractions to explicitly describe global views of exceptional control flows.
\end{abstract}

% \begin{IEEEkeywords}
% component; formatting; style; styling
% 
% \end{IEEEkeywords}

\IEEEpeerreviewmaketitle

\section{Introduction}

The recent advances on microelectronics have enabled the development of a wide
variety of smart card applications, which are able to easily and securely
perform complicated computations without being influenced by external factors
\cite{Rankl}. It is possible to develop totally new approaches to support
electronic purse, health care control, mobile communications, id verification,
access control, and so forth. As a result, a number of approaches have started
to emerge to support the implementation of smart card applications
\cite{Krakatoa} \cite{CostaMMN09} \cite{CostaMMN12}.

Java Card applets \cite{Chen:2000} is a kind of smart card application that are usually
used and deployed in highly distributed and mobile situations and tend to be
used in critical applications. To reduce financial and/or human risks, rigorous
verification of such applets is often required to guarantee the intended
behavior of the system to which these applets belong. Considering the need for
correctness in critical systems and bundled applications on cards with memory
constraint, Java Card Modeling Language (JCML) \cite{CostaMMN09}
\cite{CostaMMN12} has been proposed to support the specification of contracts between modules in accordance with the
Design by Contract \cite{Meyer92} principles.

Contracts in JCML are defined by means of annotations which express assertions,
such as method preconditions and postconditions and class invariants. JCML
annotations are automatically translated into runtime assertion checking code
that can run on smart cards. Whenever an assertion (contract) is violated, an
exception is signaled. Unfortunately, the smart card environment provided by the
Java Card platform has some limitations to deal with exceptions.

First, there is no guarantee that the exception signaled by one part of the
application is going to reach the other part of the same application. Second,
it is not possible to know which contract was violated once exceptions do not
provide stack trace information. Finally, the Java Card Exception handling is
not able to identify all exceptions signaled by the card application.

In this context, the contributions of this paper are twofold. First, we discuss
   the liabilities of conventional exception handling mechanisms to deal with
   exceptions in smart cards (Section \ref{sec:limitation}). Second, we present the EJCFlow
   (Section \ref{sec:ChannelInJC}), an implementation of the EFlow \cite{Cacho:2008}
   \cite{Cacho:2008b} model that supports explicit representation of exception
   flows in Java Card applications. The proposed implementation exploits existing JCML annotations to make the
   association of handlers with normal code more flexible.

\section{Background}
\label{sec:background}

\subsection{Java Card Description}

Java Card provides a secure environment for applications that run on smart cards
and other devices with very limited memory and processing. Due to the nature of
these devices, the platform is quite limited. Java Card applications are called
applets. In order to run one of these applets in a Java Card device, one must:
(i) write the Java Card applet; (ii) compile the applet; (iii) convert the
binary classes into a converted applet CAP file; (iv) install the CAP file on
the card; (v) run the applet. The main differences between Java Virtual Machine
(JVM) and Java Card Virtual Machine (JCVM) standards are the exclusion of some
important JVM features such as many primitive types, dynamic classes and
threads.

Because of these restrictions, a typical Java Card application will be very
limited, and only some basic functionalities will be provided on-card. The Java
Card API specification \cite{apiJavaCard} \cite{ortiz} \cite{sun2008} defines a
small subset of the traditional Java programming language API which is
supported by JCVM. On top of this, the Java Card
framework defines its own set of core classes that
specifically support Java Card applications. The main
packages \cite{Chen:2000} are: \textbf{\textit{javacard.framework}};
\textbf{\textit{javacard.security}}; \textbf{\textit{java.oi}};
\textbf{\textit{java.lang}}; and \textbf{\textit{java.rmi}}.

Most of the heavier processing is executed on the socalled
host side, the program that runs on the terminal to
which the card is temporarily connected. However, for
safety reasons, some card data may not be seen by the host
application. Such sensible data must be manipulated on the
card, safely and correctly. That is one of the advantages of
smart cards with respect to magnetic strip cards: their oncard
code may be used to ensure the safety and correctness
of data apart from the host application. Thus, it
is important to send clear messages to card users and host
applications, and handle exceptions more effectively.

\subsection{Exceptions in Java Card}

The Java Card platform does not support all of the
exception types found in the Java technology core
packages \cite{apiJavaCard}. For instance, exceptions related to
unsupported features (such as thread) are naturally not
supported.

Likewise Java, the exception types on Java Card
platform can be checked or unchecked. All checked
exceptions extend \textit{CardException} and all unchecked
exceptions are a specialization of \textit{CardRuntimeException}.

Unchecked exceptions may represent execution errors,
programming errors or errors in the communication
process between the application and the smart card. All
checked exceptions must be caught by the applet \cite{Chen:2000}. As a
consequence, no applet method can specify a checked
exception in its method signature. Whenever this happen,
the Java compiler issues a compilation error.

As Java Card platform does not support the String
class type, thus it is not possible to use string messages as
an exception reason. As a quick alternative, Java Card
supports the definition of a reason code to provide a
meaning for the signaled exceptions.

Due to memory restrictions of Java Card devices, the
JCVM strongly recommends to reuse exception objects.
Figure \ref{1} shows how the Java Card Runtime Environment
(JCRE) implements exceptions in a manner that
maximizes object recycling by using a singleton each time
the \textit{ISOException throwIt()} method is used.

The \textit{ISOException} is thrown by calling the \textit{throwIt()}
method with a reason (\textit{sw}) for the exception. The
\textit{throwIt()} method is declared static so that there is no need
to instantiate an ISOException object in order to throw the
exception. Instead, simply call \textit{throwIt()} and customize
each call with a reason code. The \textit{throwIt()} method in turn
invokes throw on the exception object.

\begin{figure}

\end{figure}

\subsection{JCML - Java Card Modeling Language Description}

JCML \cite{CostaMMN09} \cite{CostaMMN12} annotates Java Card programs to produce
runtime verification code which can be performed on
devices with severe memory and processing restrictions.
JCML includes all JML \cite{jml} \cite{Leavens_2007} \cite{Bhorkar} constructs
which can be translated into Java Card compliant code.

JCML specifications are Java Card programs annotated
with specification constructs. The specification part of a
JCML program is defined using a special kind of
comments. Java Card constructs defined within the
annotations are treated by JCML compiler and can be
used in the pre- and post-conditions and invariants.

The code generated by JCML compiler is smaller and
faster than equivalent code generated by the original JML
compiler. This is due the Java Card limitations, and
because of that, the JCML have some limitations.

For each condition in the JCML annotations, a
checking method is generated. The compiler uses the
wrapper approach proposed in \cite{Cacho:2008}. In this approach, the
code of each (annotated) method of the program is
embedded in a new method, whose task is to verify the
assertions and call the original method. The embedded
methods are renamed and made private. The wrapper
method has the same name and signature as the original
method it wraps. The wrapper method checks the
preconditions and invariants and then executes the
original method. After that, the wrapper checks the
invariant and the existing post-conditions. The auxiliary
methods are generated for the wrapper and will signal
exceptions whenever the assertions are violated.

JCML defines three reason codes to be passed as
parameter to the \textit{throwIt()} method of the \textit{ISOException}
class. The \textit{throwIt()} is invoked whenever any violation
of pre/pos condition and invariant occurs. This can be
clarified by the example of a Java class \textit{UserAccess} to
manage user access to credit, which can be either
professor or student. One of its methods is void
\textit{addCredits(short)} For which were defined in JCML the
following notations:

\begin{figure}[ht!]
\centering  
\scriptsize 
\begin{verbatim}
/*@
1 requires value >= 0 && 
2          (value + getCredits()) <= MAX_CREDITS && 
3          (userType == STUDENT ==> (value + getCredits()) 
4          <= STUDENT_MAX_CREDITS);
5 ensures printerCredits >= value;
@*/
\end{verbatim}
\end{figure}

This specification defines as pre-conditions that the
value to be added cannot be negative or exceed the
maximum number of credits, \textit{e.g.} 200 Euros, and if user is
a student, the value should not exceed the maximum
credit defined for students, \textit{e.g.} 100 Euros. As postcondition,
the specification defines the current credit
value is greater than or equal to the value recently added.

The addCredits wrapper method (Figure \ref{2}), generated
by JCML compiler, wraps the original method call in a
try-catch block that (i) checks the invariant and
precondition; (ii) calls the original method and (iii) checks
the invariant and postcondition. Figure \ref{2} also shows the
generated code for \textit{checkPre\$addCredits\$} and
\textit{checkPost\$addCredits\$}.

\begin{figure}

\end{figure}

As the initial goal of JCML language and
compiler were the generation of runtime verification
code for Java Card programs, the treatment of error
messages and condition violation are limitations of it.

\section{Limitations of Handing Contract Violations in Java Card}
\label{sec:limitation}

As described in the previous section, every time a
contract is violated, an \textit{ISOException} is signaled by the applet with a specific reason code. In order to show the
flow of control created by the propagation of an
\textit{ISOException}, Figure \ref{3} depicts part of the architecture of
a Java Card application. A black arrow from square box \textbf{a}
to square box \textbf{b} indicates that method a invoked method
\textbf{b}. As would be expected, each arrow also indicates that
control flow is passed from one method to the other.

According to Figure \ref{3}, the host application
(\textit{App.addCredits} method) calls the
\textit{CardChannel.transmit()} method via a \textit{Proxy} class with a
\textit{CommandAPDU} as parameter. Notice that smart cards
communicate using a packet mechanism called
Application Protocol Data Units (APDUs). Hence, the
host application sends a \textit{CommandAPDU} to the card
application which performs the processing requested and
returns a response APDU. If an exception is thrown on
card side, the status word field of the \textit{ResponseAPDU}
contains the error code which can be verified by the Host
Application.

At this point, we identify some limitations to handle
contract violation in Java Card applications.

First, there is no guarantee that the exception signaled
by the card application is going to be handled by the host
application. This may occur because immediately upon
return from invoking \textit{transmit} method (inside
\textit{Proxy.setField} at line 3), the developer must check the
status word of the \textit{ResponseAPDU} to identify the
occurrence of an exception. If this check is not performed,
the exception is ignored by the \textit{setField} method.
Therefore, the Java Card specification demands that the
error handling code must be extensively intermingled with
the code for normal processing, making the logic for both
cases harder to maintain.

Second, it is not possible to know which contract was
violated once the \textit{ResponseAPDU} does not provide the
stack trace of the exception signaled by the card
application. Stack traces are a useful programming
construct to support developers in debugging tasks.
Consequently, the lack of stack traces makes the task of
identifying the cause of a defect/violation even more
difficult in Java Card applications.

A more serious difficulty occurs if any other
exception (apart from \textit{ISOException}) flows out of
\textit{Applet.process()} method. In this case, the JCRE catches
this outgoing exception and returns a ResponseAPDU
with \textit{ISO7816.SW\_UNKNOWN} as the status word field. In
order words, the developer knows that something goes
wrong but not specifically what.

In our view, all the aforementioned problems stem
from the fact that the Java Card exception handling
mechanism does not appropriately take global \cite{RobillardM00}
\cite{Robillard:2003} exceptions into account. In other words, the Java Card
exception handling mechanism provides some constructs
for raising, propagating and handling exceptions.
However, not much explicit support is provided to the
tasks of defining and understanding the (un)desirable paths of global exceptions
from the raising site to the final handler.

\begin{figure}

\end{figure}

The main consequence of this limitation of Java Card
exception handling is that global exceptions introduce
implicit control flows. An implicit control flow occurs
whenever a developer can not directly observe what code
will be executed during run-time. As we described above,
implicit control flow created by the Java Card exception
handling mechanism makes it difficult to explicitly
analyze: if an exception is going to be handled and where,
and which components are affected by an exception
control flow.

\section{Explicit Exception Channels in Java Card}
\label{sec:ChannelInJC}

In order to address the problems described above, we
propose a new implementation for the EFlow model.
EFlow \textit{Cacho:2008,Cacho:2008b} is a platform-independent model for
exception handling whose major goal is to make
exception flow explicit, safe, and understandable by
means of explicit exception channels and pluggable
handlers. Section \ref{subsec:eflow} describes the concepts that
underpin EFlow model. In the following, Section \ref{subsec:limitEflow}
illustrates how the abstractions defined by EFlow model are concretized in terms
of JCML annotations. Finally, Section \ref{subsec:detailEflow} describes some
implementation details of the proposed solution.

\subsection{EFlow Model}
\label{subsec:eflow}

EFlow model is grounded on mechanisms to explicitly
represent global exceptional behavior. This specification
enables software developers to establish constraints
governing the implementation of non-local exceptions
flows. An exception handling specification is composed
of two abstractions: \textit{explicit exception channel} and
\textit{pluggable handlers}.

An \textit{explicit exception channel} (\textit{channel}, for short) is
an abstract duct through which exceptions flow from a
raising site to a handling site. More precisely, an explicit
exception channel \textit{EEC} is a 5-tuple consisting of: (1) a set
of exception types \textit{E}, (2) a set of raising sites \textit{RS}; (3) a set
of handling sites \textit{HS}; (4) a set of intermediate sites IS; and
(5) a function \textit{EI} that specifies the channel's exception
interface.

\textit{Exception types}, as the name indicates, are types that,
at runtime, are instantiated to exceptions that flow through
the channel. The \textit{raising sites} are loci of computation
where exceptions from E can be raised. The actual
erroneous condition that must be detected to raise an
exception depends on the semantics of the application and
on the assumed failure model. For reasoning about
exception flow, the fault that caused an exception to be
raised is not important, just the fact that the exception was
raised. The \textit{handling sites} of an explicit exception channel
are loci of computation where exceptions from E are
handled, potentially being re-raised or resulting in the
raising of new exceptions. In languages such as Java, both
raising and handling sites are methods, the program
elements that throw and handle exceptions.

If an explicit exception channel has no associated
handlers for one or more of the exceptions that flow
through it, it is necessary to define its exception interface.
The latter is a statically verifiable list of exceptions that a
channel signals to its enclosing context, similarly to
Java's \textit{throws} clause. In our model, the exception
interface is defined as a function ($Ex_1 \mapsto Ex_2$) that
translates exceptions flowing ($Ex_1$) through the channel to
exceptions signaled ($Ex_2$) to the enclosing EHC.

Raising and handling sites are the two ends of an
explicit exception channel. Handling sites can be
potentially any node in the method call graph that results
from concatenating all maximal chains of method
invocations starting in elements from HS and ending in
elements from RS. All the nodes in such graph that are
neither handling nor raising sites are considered
\textit{intermediate sites}. Intermediate sites comprise the loci of
computation through which an exception passes from the
raising site on its way to the handling site. Intermediate
sites in Java are methods that indicate in their interfaces the exceptions that
they throw, \textit{i.e.} exceptions are just propagated through them, without
side effects to program behavior. Note that the notions of handling, raising, and
intermediate site are purely conceptual and depend on the specification of the
explicit exception channel. They are also inherently recursive. For example, an
intermediate site of an explicit exception channel can be considered the raising
site of another channel.

A \textit{pluggable handler} is an exception handler that can be associated to
arbitrary EHCs, thus separating error handling code from normal code. A single pluggable
handler can be associated, for example, to a method call in a class \textit{C1},
two different method declarations in another class, \textit{C2}, and all methods in
a third class \textit{C3}. In this sense, they are an improvement over
traditional notions of exception handler. Another difference is that a pluggable handler exists
independently of the EHCs to which it is associated. Therefore, these handlers
can be reused both within an application and across different applications.

We can define several different explicit exception channels in terms of the
  elements of Figure \ref{2}. For example, let \textit{EEC1} be the explicit
  exception channel defined by the tuple (\textit{\{ISOException\},\{UserAccessJCML.
  checkPre\$addCredits\$\},\{App.addCredits\},\{Proxy.addCredits\},\{\}}). Only
  exception \textit{ISOException} flows through this channel. Explicit exception
  channel \textit{EEC1} has method \textit{checkPre\$addCredits\$} as its sole
  raising site and Proxy.addCredits as its only intermediate site. Method App.addCredits is handling site, since one
  pluggable handler can be bound to it. \textit{EEC1} does not define an
  exception interface, as it includes handlers for all of its exceptions.

\subsection{Proposed Implementation of EFlow Model}
\label{subsec:limitEflow}

This section presents EJCFlow, a JCML extension that implements the EFlow model
to support Java Card applications. EJCFlow provides means for developers to
define explicit exception channels and pluggable handlers in terms of the
abstractions supported by JCML.

In order to support the definition of Explicit Exception Channels and Pluggable
Handlers, EJCFlow provides a notation to define \textit{contexts}, which in
practice are methods that can represent raising, intermediate or handling sites. Contexts are
defined as the following:

\begin{figure}[ht!]
\centering  
\scriptsize 
\begin{verbatim}
/*@ econtext(RS, withincode(
                 UserAccessJCML.checkPre$addCredits()) 
@*/

/*@econtext(HS, withincode(App.addCredits) @*/
\end{verbatim}
\end{figure}

Here we defined a context called RS bounded to the method
\textit{checkPre\$addCredits\$()} and a context called \textit{HS} bounded to
the method \textit{addCredits()}.

After defining context, EJCFlow provides a way to define Explicit Exception
Handling by a notation that takes 3 parameters: (1) the channels's name; (2) the
exception type; and (3) the raising context. Explicit channels are defined as
the following:

\begin{figure}[ht!]
\centering  
\scriptsize

\begin{verbatim}
/*@ echannel(ECC, ISOException, RS) @*/
\end{verbatim}
\end{figure}

The above example defines one explicit exception channel called \textit{ECC},
which represents the exception \textit{ISOException} flowing out of the context
\textit{RS}, previously defined.

In order to specify the handling site of an explicit exception channel, EJCFlow
provides a notation that consists of: (1) the associated channel's name; (2) the
handling exception type (3) the handling context; and (4) the body of the actual
handler implementation. The following code snippet presents a simple example:

\begin{figure}[ht!]
\centering  
\scriptsize

\begin{verbatim}
/*@ ehandler(ECC, ISOException, HS) @*/ 
public static void aHandler(ISOException e) {
   //do something 
}
\end{verbatim}

\end{figure}

The above code snippet defines a method \textit{aHandler()} which handles
exceptions of type \textit{ISOException} flowing through explicit exception
channel \textit{EEC} specifically when the exception reaches the context HS.

When a program cannot handle all the exceptions that flow through an explicit
exception channel, it is necessary to declare these exceptions in the channel's
exception interface. The einterface annotation serves this purpose. The
following code snippet illustrates the definition of exception interfaces:

\begin{figure}[ht!]
\centering  
\scriptsize

\begin{verbatim}
/*@ econtext (rS1, withincode 
              (UserAccessJCML.addCredits( short ))
              ) @*/
/*@ econtext (rI1, withincode( Applet.addCredits)) @*/

/*@ echannel (ECC1, ISOException , rS1) @*/
/*@ einterface (ECC1 , ISOException, rI1) @*/
\end{verbatim}
\end{figure}

This above explicitly indicates the exception to be declared in the exception
interface of the channel.

\subsection{Implementation Details}
\label{subsec:detailEflow}

EJCFlow consists of two parts: (1) a parser to support EFlow model on Java Card
environment; and (2) a bytecode analyzer to make the needed transformations on
the applications bytecode. The first is built utilizing the tool JavaCC (Java
Compiler Compiler) version 5.0 and the latter is built on top of Soot \cite{Vallee-RaiGHLPS00}.

The entire process (Figure \ref{fig:sootEchannel}), from a pure JavaCard program to an
transformed program by JCML-Compiler and EJCFlow can be divided in three stages: (1)
annotate the Java Card application with JCML and run JCMLc, producing new .class
files with runtime condition verifications; (2) specify the contexts, channels
and handlers, which will be parsed and verified by our JavaCC-based tool, in
order to ensure the validity and correlation of application code with EFlow
definitions, generating an representation of EFlow model data (channels,
contexts and handlers); and (3) with the EFlow data, host side application
binaries and card side JCMLc generated binaries, our soot-based tool will make
the bytecode needed transformations, generating new card side and host side
application binaries, which are able to communicate and handle properly JCML
specified behavior by means of EFlow definitions.

\begin{figure}[ht!] 
\centering  
\includegraphics[width=0.45 \textwidth]{figs/SootEchannelProcess} 
\caption{High-level overview of the components of the implementation}
\label{fig:sootEchannel}   
\end{figure}

\section{Related Works}
\label{sec:relatedWorks}

There are a few approaches that try to handle contract violation by means of
improving the exception handling behavior. The Krakatoa tool \cite{Krakatoa} for
certification of java and Java Card programs translate the java program,
annotated with JML, into the WHY \cite{FilliatreM07} input language. The WHY input language
is an ML-like minimal language, with very limited imperative features, such as
references and exceptions. On the other hand, \cite{Rebelo} defines a set of types of
faults that can happen on the interplay between java exception handling
mechanism and runtime assertions checkers (RACs) for JML. It presents an error
recovery technique for RACs that tackles such limitations. The proposed approach
uses of aspect-orientation to monitor the exceptions that represent contract
violations and force such exceptions to be signaled, preventing exceptional
handling contexts from negatively affecting RAC generated exceptions.

The EJCFlow implementation has three advantages in terms of the aforementioned
related works. First, EJCFlow abstractions allow the explicit representations of
properties governing global exception behaviors. As a consequence, the
separation between normal and exceptional behavior is preserved, resulting in
untangled software artefacts, reusable handlers, and improved maintainability.
Second, EJCFlow abstractions make it possible to understand the potential paths
of global exceptions from an end-to-end perspective by looking at a single part
of the program. Third, EJCFlow improves reliability in evolving software by
enforcing that harmful exceptional behaviors are not added between the two ends
of a given explicit exception channel. Also, the EFlow model allows to verify
whether or not all exceptions flowing through an explicit exception channels are
correctly handled at pluggable handlers.

\section{Conclusion} 
\label{sec:conclusion}

This paper has presented an implementation of the EFlow model whose main purpose
is to improve the exception handling mechanisms of Java Card Applications. We
leverage the JCML specification-level annotations to promote improved separation
between normal and error handling code, while keeping track of exception control
flows. Our ongoing work encompasses the empirical evaluation of the error
proneness on the use of EJCFlow when compared to conventional proposals
discussed in this paper.

% conference papers do not normally have an appendix

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% Better way for balancing the last page:

\balance

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{plain}
\bibliography{weh'12}

% that's all folks
\end{document}

### 2013-09-30 WESOA2013/source/auxiliar/enviroment.tex
inicial --martin

\chapter{\textit{$\pi$SOD-M} Environment} 
\label{chap:environment}

\epigraph{``\textit{The computer was born to solve problems that did not exist
before.}''}{Bill Gates}

This chapter describes the $\pi$SOD-M environment for developing reliable
service based systems using the methodology we propose. The $\pi$SOD-M
environment was developed to support the $\pi$SOD-M methodology. The $\pi$SOD-M
based process consists in generating an application starting from a
\textit{$\pi$-UseCase} model and then transforming it into series of models at
the PIM and PSM levels before finally generating the code that implements the
application.

%  that implements 
% the generation of {\em Policy} based services' compositions. For a given
% services' based application, the process consists in generating an application
% starting from a $\pi$-UseCase modeling. From a $\pi$-UseCase model,
% transformations between models are made until the platform specific model level
% to generate service composition specification code. Four model to model
% transformations are performed, three transformations between models, and one
% model to text transformation generates the service composition specification in
% $\pi$-PEWS language. All transformations are done through an Eclipse plugin. All
% models definition are the result of a general process defined by the $\pi$SOD-M
% method in which a set of models are built following a service oriented approach.

$\pi$SOD-M environment is built on the top of Eclipse framework
(http://www.eclipse.org), which is a framework to build Integrated Development
Environments (IDEs). We also use the Eclipse
Modelling Framework (EMF), a meta-modelling framework that was devised to be
extended and provides with the utilities needed to define, edit and handle
(meta)-models. To automate the transformation models we use the ATL Language\footnote{ATL uses EMF to handle models (that is to serialize, to navigate and to modify
them). Once the ATL transformation is executed, the ATL engine matches the
source pattern and the source model and instantiates each pattern into the
target model.} \cite{atl_manual}, a model transformation language framed in
Eclipse. Another language used for model transformation is Acceleo \cite{acceleo}, a
model to text engine. $\pi$SOD-M uses this environment~\footnote{\newText{The environmental requirements are the Eclipse Modelling (Galilleo or higher)
and all proposed plugins, \textit{$\pi$-UseCase, $\pi$-ServiceProcess,
$\pi$-SerciveComposition} and \textit{$\pi$-PEWS} editors; and also the
transformation plugins. Because the environment is proposed in the
context of Eclipse, the environmental requirements are also the same needed to
utilize the Eclipse tool.}} to generate service
composition specification code in \textit{$\pi$-PEWS}.

% then for each matching, the target pattern is instantiated in the target model, replacing the match found in the
% source model.

% Thus, ATL works perfectly with the models defined over EMF, as well as
% $\pi$SOD-M plugin. So one, ATL compiler provides features like multiples source
% patterns, rule inheritance, and endpoint rules.

%  After the user model the application and do the
% transformations, as final result, he can have the correspondent code for
% the service application. Thus, the methodology development environment
% is composed of these two platforms (ATL and Acceleo) that assist in the development
% of $\pi$SOD-M based applications.

% It is not necessary the methodology's final user knows Acceleo or ATL
% for modeling an application. These languages have been used only for the
% tools development. The tool's user also does not need have extensive knowledge
% in MDA to use the tool. Knowledge of basic MDA's concepts and what means
% transformation between models are sufficient to adapt and use the tool in a real
% development environment.

The remainder of the chapter is organized as follows. Section
\ref{sec:architecture} describes the environment 
architecture, including the definition of meta-models; and, the implementation
of the model transformation process and the code generation engine. In section
\ref{sec:install} describes how to install and use the environment. Section
\ref{sec:extendingEnvironment} describes how to extend the
environment, for adding new components for generation of system specification
in different languages. Section \ref{sec:env_conclusion} concludes the chapter.

\section{General Architecture}
\label{sec:architecture}

Figure \ref{fig:policymanager} presents the general architecture for the
$\pi$SOD-M environment and details how components interact with each other. 
Each model has an editor, and transformation tools that support the application
development.

% This figure
% presents, in general, the same components described in figure
% \ref{fig:environmentComponents}, although emphasizing the context of Eclipse.
% 
% The meta-models and their plugins for editing are at the highest
% level. In an intermediate level we have the ATL based transformation
% components for the models mapping, and finally, the code generation is
% accomplished through the mapping described in Acceleo, generating the
% specification to be executed.

\begin{figure} [ht!]   
\centering
\includegraphics[width=1.0\textwidth]{chapters/methodology/figs/FrameworkOverviewFigure}
\caption{$\pi$SOD-M Development Environment.}
\label{fig:policymanager} 
\end{figure}

The $\pi$SOD-M environment architecture is organized in
three layers: \textit{(i) Meta-model level, (ii) PIM-to-PIM Transformations
level level, (iii) PIM-to-PSM Transformations level
level} and \textit{(iv)  Code generation level}.
 Each layer comprises a set of components that
together support the $\pi$SOD-M environment. 
This figure presents how those components are implemented by our tool: (i) the
\textit{Meta-model component} (figure \ref{fig:environmentComponents})
represents the \textit{Model plugins module} of figure \ref{fig:policymanager} (together with all methodology
meta-models); (ii) the \textit{PIM-to-PIM Transformations and PIM-to-PSM
Transformations components} (figure \ref{fig:environmentComponents}) represent
the \textit{Mapping plugins module} of figure \ref{fig:policymanager}; 
(iii) the \textit{Code Transformation} component of the figure
\ref{fig:environmentComponents} is implemented by the \textit{Code generation
module} of figure \ref{fig:policymanager}; and (iv) the \textit{Execution
engine} is represented by the lower tier of figure \ref{fig:policymanager}.

% The main components
% are: \textit{(I)} - \textit{$\pi$-UseCase} ecore meta-model,
% \textit{$\pi$-ServiceProcess} ecore meta-model, \textit{$\pi$-ServiceComposition} ecore meta-model and
% \textit{$\pi$-PEWS} ecore meta-model, that are
% sub-components of the \textit{Meta-model component} (\textit{EMF Ecore
% meta-model level}); \textit{(II)} - \textit{$\pi$-UseCase2$\pi$-ServiceProcess} component and
% \textit{$\pi$-ServiceProcess2$\pi$-ServiceComposition} component, that are
% sub-components of the \textit{PIM-to-PIM Transformation component} (\textit{Model transformation
% level}); \textit{(III)} - \textit{$\pi$-ServiceComposition2$\pi$-PEWS}
% component, that are sub-components of the \textit{PIM-to-PSM Transformation
% component}; \textit{(IV)} - the \textit{Code Transformation} component
% (\textit{Code generation level}); and \textit{(V)} - the \textit{Execution Engine} component, comprising both,
% \textit{APolicy Engine} and \textit{PEWS Engine} components.

\begin{figure} [ht!]   
\centering
\includegraphics[width=1.0\textwidth]{chapters/implementation/figs/componentsWOE}
\caption{Environment Components.} 
\label{fig:environmentComponents} 
\end{figure}

The \textit{Models plugin module} comprises the components that describe
the $\pi$SOD-M meta-models, and how their models must be created. There are
four meta-model components. All components of the \textit{Mapping plugin module}
depend of the definitions made in \textit{Models plugin}.

When a model transformation is made, the models must comply with their
respective meta-model. The process is executed as follows: every time a
transformation is performed, a consistency check of both, source and target
models is performed. After all transformation are made, the PSM model is
translated into code of a particular platform, in the case of $\pi$SOD-M,
\textit{$\pi$-PEWS} is the chosen platform.  The transformation of PSM model in
code is the last stage of transformation. The component of the \textit{Code
generation module}  depend of the PSM generated by the last model
transformation. Finally, the \textit{Execution Engine} component performs the
execution of the service-based specification code.

% The following three sections will detail each particular feature of the
% $\pi$SOD-M development environment.

\subsection{Ecore Meta-models (\textit{Models Plugin Module})}

The implementation of each meta-model is defined in \textit{Ecore}\footnote{The 
\textit{Models Plugin Module} is a set of Ecore files that represents all $\pi$SOD-M meta-models. These meta-models are the sources for the development of
each of the methodology's model. All models designed for an application must
obey their meta-model specification. This module is composed by all the proposed
methodology's models in EMF (\textit{.ecore extension}).} files, they are:
\textit{$\pi$-UseCase, $\pi$-ServiceProcess, $\pi$-ServiceComposition} and
\textit{$\pi$-PEWS}.  All meta-models have a related
\textit{genmodel}\footnote{A \textit{.genmodel} is a intermediary file format
used to produce the syntax editor for each meta-model.} definition. The plugin
editor can be created from the \textit{genmodel} definition, so that models can
be specified.

% The Ecore Tools component provides a complete environment to create, edit and
% maintain Ecore models. This component eases handling of Ecore models with a
% Graphical Ecore Editor and bridges to other existing Ecore tools.  
% From a model specification, EMF provides tools and runtime support to produce a
% set of classes for the model, a set of adapter classes that enable viewing and
% command-based editing of the model, and a basic editor. Models can be specified
% using annotated UML, XML documents, or modeling tools, then imported into
% EMF. Most important of all, EMF provides the foundation for interoperability
% with other EMF-based tools and applications.

% Although there are editors for each model of the methodology, it is still
% necessary components that can make a transformations between them. Thus the
% editing process need not be complete at all levels, but only at the highest
% level, \textit{i.e.} $\pi$-UseCase model, and then make automatic generation for
% the other models.

%After the creation of each meta-model were created their respective plugins.

Using these set of tools (model editors) it is possible to create models
at different $\pi$SOD-M levels. There are editors for all $\pi$SOD-M models:
\textit{$\pi$-UseCase editor, $\pi$-ServiceProcess editor,
$\pi$-ServiceComposition} editor and \textit{$\pi$-PEWS} editor.

Although there are editors for each methodology model, it is still
necessary components to perform transformations among them. Thus, the
model specification process can be made in all methodology levels. However it
can be made only at the highest level, \textit{i.e.} \textit{$\pi$-UseCase}
model, and then execute automatic transformation to generate lowest level
models.

 \subsection{Model Transformation (\textit{Mapping Plugin Module})}

 The \textit{Model transformation level} has a set of components for processing
 and transforming models. The model transformation components are based on
 the source models for generating the equivalent target model. For example, from
 a \textit{$\pi$-UseCase} model is generated the
\textit{$\pi$-ServiceProcess} model, from a \textit{$\pi$-ServiceProcess}
model is generated the \textit{$\pi$-ServiceComposition} model, and finally,
from a \textit{$\pi$-ServiceComposition} model is generated a 
\textit{$\pi$-PEWS} model. After the model generation the
designer can perform refinements. Refinement can improve and adjust elements
that require a more detailed description at this modeling level.
% Figure \ref{fig:policymanager} shows the environment components for the
% processing models transformation.

$\pi$SOD-M model transformation process is based on the rules described in
chapter \ref{chapter:methodology}. The implementation process requires
additional, more specific information to be taken into account. We had to
consider the representation of our concepts in the Eclipse and ATL environments
for MDA-based application development, \textit{e.g.}, aspects of plugin
generation; design model properties; compliance of the designed
meta-model with the generated model; the specific model implementation; and etc.

% \begin{itemize}
%   \item Each meta-model is modeled using the EMF environment. During the
%   meta-model definition, a beta plugin version was being generated to verify the
%   model development consistence and coherence. For example the $\pi$-UseCase
%   meta-model, during its definition, short use cases models examples were
%   developed to validate our definition;
%   \item A stable plugin version was generated and then a complete example was
%   modeled to validate the usability of each $\pi$SOD-M model plugin. The
%   sequence of the model plugin development were:
%   \textit{$\pi$-ServiceComposition, $\pi$-PEWS, $\pi$-UseCase} and
%   \textit{$\pi$-ServiceProcess}. We first generated the low level plugins to
%   validate the code generation from a service composition specification.
%   Immediately after we developed the high level model to the complete $\pi$SOD-M
%   support.  
%   \item  With all models well developed and assembled in the tool for each
%   aspect of service model development, it was necessary proceed with the
%   automatic transformation. Each transformation represents, from a specific
%   $\pi$SOD-M model, that it is possible to generate a equivalent lower level
%   $\pi$SOD-M model, \textit{e.g.}, (i) from the $\pi$-UseCase model, the
%   $\pi$-ServiceProcess can be generated; (ii) from the $\pi$-ServiceProcess
%   model a $\pi$-ServiceComposition can be generated; (iii) from the
%   $\pi$-ServiceComposition model, the $\pi$-PEWS model can be generated;  and
%   (iv) from the $\pi$-PEWS model, the system service composition code can be
%   generated.
%   \item By the end, we configure a $\pi$SOD-M Eclipse version for research use.
% \end{itemize}

We used the Eclipse Modeling Framework (EMF) for implementing the
meta-models  \textit{$\pi$-UseCase, $\pi$-ServiceProcess,
$\pi$-ServiceComposition} and \textit{$\pi$-PEWS}. The ATL language was used
for developing the mapping rules for transformation of models
(\textit{$\pi$-UseCase2$\pi$-ServiceProcess},
\textit{$\pi$-ServiceProcess2$\pi$-ServiceComposition} and
\textit{$\pi$-ServiceComposition2$\pi$-PEWS} plug-ins). 
This plugin takes as input a \textit{$\pi$-PEWS} model implementing a specific
service composition and it generates the code to be executed by the {\em Policy} based
\textit{Execution Engine} (figure \ref{fig:environmentComponents}).

 \begin{figure}  [ht!]
\centering
\includegraphics[width=1.0\textwidth]{chapters/methodology/figs/modelTrasformation}
\caption{ATL Model to Model Transformation in $\pi$SOD-M.}
\label{fig:modelTomodelTransfomation}
\end{figure}

 Figures \ref{fig:modelTomodelTransfomation} and 
 \ref{fig:modelToTextTransformation} present a general view of the $\pi$SOD-M
 models transformation, showing the set of plug-ins developed to
 implement it. Figures \ref{fig:modelTomodelTransfomation} presents the
 model-to-model transformations, while figure
 \ref{fig:modelToTextTransformation} shows the model-to-text transformation
 schema. The environment implements the abstract architecture shown in figure
 \ref{fig:policymanager}. This environment consists of plug-ins implementing the
 \textit{$\pi$-UseCase, $\pi$-ServiceProcess, $\pi$-ServiceComposition} and
 \textit{$\pi$-PEWS} meta-models used for defining models; and ATL rules for
 transforming PIM and PSM models (model to model transformation) and finally
 generating code (model to text transformation) with Acceleo.

\begin{figure} [ht!]  
\centering
\includegraphics[width=0.7\textwidth]{chapters/methodology/figs/AcceleoTransformationFigure}
\caption{Acceleo Model to Text Transformation in $\pi$SOD-M.}
\label{fig:modelToTextTransformation} 
\end{figure}

In order to proceed with a model transformation it is necessary to configure the
transformation environment. Figure \ref{fig:configurationATL} presents a
standard screen for configuring each transformation (in this case for the ``to
publish music'' scenario). From any source model, for example,
\textit{$\pi$-UseCase, $\pi$-ServiceProcess} or
\textit{$\pi$-ServiceComposition}, the system can perform the automatic
transformation. Using the transformation tool of figure
\ref{fig:configurationATL} requires from the user: (1) to indicate the
ATL transformation rules file; (2) to choose the source meta-model reference
(.ecore file); (3) to choose the target meta-model reference (.ecore file); (4) 
to choose the source model that will be transformed (\textit{e.g.
music.piusecase} file); (5) to choose the target model that will be generated
(\textit{e.g. music.piserviceprocess} file); and finally, (6) to run the tool.
The same must be performed for all model-to-model transformations. For model-to-text
transformation, the rule file to be chosen should be the
Acceleo file.

\begin{figure}[ht!]
\centering
\includegraphics[width=.90\textwidth]{chapters/implementation/figs/ATL-trasnformationPiUseCase2PiServiceProcess.pdf}
\caption{ATL Configuration for $\pi$SOD-M Transformation.}
\label{fig:configurationATL}
\end{figure}

As an example, figure \ref{fig:configurationATL} shows the configuration for the
transformation from \textit{$\pi$-UseCase} (source) to
\textit{$\pi$-ServiceProcess} (target). Notice that, in this case, the reference
meta-models must be the same. For further transformations, this process must
follow the same sequence, changing only the models and reference meta-models.

\subsubsection{\textit{$\pi$-UseCase2$\pi$ServiceProcess} Transformation Rules}

The transformation rules describe how models are transformed. As a general rule, 
the automatic transformation of models favors a faster development of
applications. In most cased, the environment or the designer should verify if
the transformations are valid or not. In the $\pi$SOD-M environment, the
consistency of the transformations must be performed by the designer. If a
problem on the transformation is identified, the target model can be modified manually.

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:exampleRules,caption=ATL Example Rule.]

1  rule ruleName{	 
2	  from		
3		  var_sourceName: SourceModel!Entity
4	  to
5		  var_targetName: TargetModel!Entity(
6			  atribute_1 <- var_sourceName.atribute_a,
7			  atribute_2 <- var_sourceName.atribute_b	
8		  )
9  }
\end{lstlisting}
\label{fig:exampleRule} 
\end{figure}

In ATL, there exist two different kinds of rules that correspond to the two
different programming modes provided by ATL (e.g. declarative and imperative
programming). The \textit{matched rules}\footnote{The matched rules constitute
the core of an ATL declarative transformation since they make it possible to
specify 1) for which kinds of source elements target elements must be generated,
and 2) the way the generated target elements have to be initialized. A matched
rule is identified by its name. It matches a given type of source model element,
and generates one or more kinds of target model elements. The rule specifies the
way generated target model elements must be initialized from each matched source
model element\cite{atl_manual}.}  (declarative programming) and the
\textit{called rules}\footnote{The called rules provide ATL developers with convenient
imperative programming facilities. Called rules can be seen as a particular type
of helpers: they have to be explicitly called to be executed and they can accept
parameters. However, as opposed to helpers, called rules can generate target
model elements as matched rules do. A called rule has to be called from an
imperative code section, either from a match rule or another called
rule\cite{atl_manual}.} (imperative programming) \cite{atl_manual}. For the $\pi$SOD-M environment, we
use both types of rules. Listing \ref{list:exampleRules} shows a general example of an
ATL \textit{matched rule}. According to this, we will present the main ATL rules used to
implement the model transformation. As described in listing
\ref{list:exampleRules}, a rule consists of a name (line 1), a source entity
(\textit{from} clause in lines 2-3), and one or more target entities (\textit{to} clause in line 4-5). Each entity has a name, such as \textit{var\_sourceName} and
\textit{var\_targetName}. The transformation is performed by each entity
attribute and general rules established in each case.

% The rules for automatic \textit{$\pi$-UseCase2$\pi$ServiceProcess}
% transformation comply with the description made in section
% \ref{sec:models-tranformation}, and we present some rules in ATL for model
% transformation in $\pi$SOD-M.

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:useCase2Action,caption=ATL -
piUseCase2piServiceProcess : useCase2action Rule. ]

1  rule useCase2action {
2  	 from usecase : PiUseCase!FunctionalBasicUseCase(		
3   	 usecase.extend->size() >= 2	
4  	 )
5  	 to sp_action :  PiServiceProcess!Action (
6		  name <- usecase.name
7	 ), sp_serviceActivity : PiServiceProcess!ServiceActivity(
8		 action <- sp_action,
9		 name <- usecase.name + 'SA',
10		 serviceProcess <- thisModule.raiz_piUseCase
11	 ), sp_contract: PiServiceProcess!Contract(					
12		 name <- usecase.name + 'Contract',
13		 assertionInContract <- usecase.UCContraint,
14		 action <- sp_action,
15		 serviceProcess <- thisModule.raiz_piUseCase
16	 ) 	
17 }
\end{lstlisting}
%\label{fig:pewscontract} 
\end{figure}

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:constraint2contract,caption=ATL -
piUseCase2piServiceProcess : constraint2contract Rule. ]

1  rule constraint2contract{	 
2	  from		
3		  constraint: PiUseCase!UseCaseConstraint
4	  to
5		  assertion: PiServiceProcess!Assertion(
6			  name <- constraint.name,
7			  description <- constraint.description	
8		  )
9  }
\end{lstlisting}
%\label{fig:pewscontract} 
\end{figure}

Listing \ref{list:useCase2Action} shows the ATL transformation rule between a
{\sc Use Case} and an {\sc Action}, while listing \ref{list:constraint2contract}
shows the transformation rule between {\sc Constraint} and {\sc Assertion}.
There is a restriction in the rule presented in listing
\ref{list:useCase2Action} (line 3), which defines that there must be more than
two {\sc Extend} relation between use cases. In this case, all use cases are
transformed into actions, and are grouped into a {\sc Service Activity} (lines
7-10). And all the constraints are transformed in a set of assertions that are
automatically associated with a {\sc Contract} (lines 11-15). This
transformation (listing \ref{list:useCase2Action}) describes that from a {\sc
Use Case} (lines 2), {\sc Actions}, {\sc Service Activity} and {\sc Contract}
can be generated (lines 5, 7 and 11). It depends on the existence of {\sc
Constraint} or dependence elements ({\sc Incllude} or {\sc Extend}) related to
this use case (line 3).

The listing \ref{list:constraint2contract} describes that all {\sc Constraint}
(\textit{from} clause in lines 2-3) are directly transformed into an
{\sc Assertion} (\textit{to} clause in lines 4-5) (there is no restriction in
the this clause). The restrictions are treated in separate rules such as in
listing \ref{list:useCase2Action}.

\subsubsection{\textit{$\pi$-ServiceProcess2$\pi$-ServiceComposition}
Transformation Rules}

As the \textit{$\pi$-ServiceComposition} model is a
refinement of some concepts defined in \textit{$\pi$-ServiceProcess}, such as 
{\sc Contract} and {\sc Activity Services}, most of the transformation
between these models are through direct transformation, without restrictions.
We present the ATL transformation (listing
\ref{list:rooServiceProcess2rootServiceComposition}) of the root element that
comprises the main elements of both models.

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:rooServiceProcess2rootServiceComposition,caption=ATL -
piServiceProcess2piServiceComposition : root Rule. ]

1  rule root {
2 	from		
3 		root: piServiceProcess!ServiceProcess
4	to
5		root_service: piServiceComposition!CompositionServiceModel(
6			activities <- root.activity,
7			edges <- root.edge,
8			compositionPolices <- root.contract
9		)
10 }

\end{lstlisting} 
\label{fig:pewscontract} 
\end{figure}

Listing \ref{list:rooServiceProcess2rootServiceComposition} shows the
transformation rule for the main elements from \textit{$\pi$-ServiceProcess} 
to \textit{$\pi$-ServiceComposition} model. The {\sc Activities} and its
{\sc Edges}, and {\sc Contracts}, which are transformed into {\sc Policy} (lines
6-8). Other rules that describe the transformation between {\sc Assertions} into
policy {\sc Rule} are other type of transformation. However, other rules
describe that all {\sc Contracts} must belong to a specific {\sc Non-functional
Requirement}, for example, all contracts for the performance restrictions, will
be grouped into a single  performance {\sc Policy}. This listing is an ATL
example for \textit{$\pi$-ServiceProcess} model transformation into 
to \textit{$\pi$-ServiceComposition} model.

\subsubsection{\textit{$\pi$-ServiceComposition2$\pi$-PEWS} Transformation
Rules}

The transformation \textit{$\pi$-ServiceComposition2$\pi$-PEWS} is
unique among PIM and PSM levels, however there is no difference in the
rules description in ATL, since all rules are defined in terms of
meta-models rather than specific models. Thus, PIM-to-PSM transformation rules
in ATL follow the same syntax and semantics of the PIM-to-PIM transformation
rules.

The model to be generated in this transformation, namely \textit{$\pi$-PEWS}
model, will be used as input for the code generation component, so that the
application code specification is generated. The listings \ref{list:rootSC2PEWS} and \ref{list:rulePre} present two
transformation rules in ATL. The first describes the transformations of the main
elements for the description of a service composition, the main path, the name
of the specification, the services and policies (lines 6 - 9), while
listing \ref{list:rulePre} describes a policy {\sc Rule} and its related concepts, such as
{\sc Action}, {\sc Event} and {\sc Condition}. The ATL rule has a constraint
to be checked (line 4), what kind of {\sc Rule} is being translated for the
specific language, because depending on the type, the transformation will be
change. The {\sc Rule} element (line 15) consists of all the properties necessary to
create a {\sc Pre-condition}, such as {\sc Action}, {\sc Event} and {\sc Condition} (lines 16-18),
and to which {\sc Policy} the {\sc Pre-condition} is related (line 19).

\begin{figure} 
\tiny
\centering

\begin{lstlisting}[label=list:rootSC2PEWS,caption=ATL -
piServiceComposition2piPEWS : root Rule. ]

1  rule root{ 
2	 from sCM : pisc!CompositionServiceModel
3	 to 
4	 	path: pipews!Path (),
5	 	pews : pipews!PEWSCTSpec (
6			name <- 'newModelPEWSpecName',
7			has <- path,
8			contains <- sCM.partition,
9			defines <- 	thisModule.policies
10		)
11  }
\end{lstlisting}
\label{fig:pewscontract} 
\end{figure}   
\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:rulePre,caption=ATL -
piServiceComposition2piPEWS : Pre-condition Rule. ]

1  rule rulePre{
2 	 from
3		r: pisc!Rule (
4			r.event = #PRE)
5	 to
6		rAct: pipews!Action(
7			act <- r.action
8		),
9		rEvt: pipews!Event( 
10			type <- #ActivityPrepered
11		),
12		rCond: pipews!Condition(
13			expression <- r.condition
14		),
15		rRule: pipews!Precondition(			
16			calls <- rAct,
17			defines <- rCond,
18			hasSome <- rEvt,
19			policy <- r.policy
20		)
21  }

\end{lstlisting}
\label{fig:pewscontract} 
\end{figure}

% The next section will show how the code is generated from a \textit{$\pi$-PEWS}
% model.

\subsection{Code Generation  (\textit{Code Generation Module})}
 }

\newText{ 
 The $\pi$SOD-M methodology and its environment use $\pi$-PEWS\footnote{Other meta-models and
 transformations can be defined for supporting  code generation into WS-BPEL,
 XAML and other languages adapted for describing business/service process.} as
 intermediate language for supporting the generation of executable code. 
 $\pi$-PEWS is a behavioral language for defining web services interfaces by
 specifying the input/output interface of a web service exported method but also
 the expected behaviour of its methods (e.g., dependencies, exchanged messages).
 The extension  $\pi$-PEWS extension was inspired by languages like JCML
 \cite{CostaMMN12}, JML \cite{LeavensKP07} and Eiffel \cite{Meyer92b}, which
 introduce the notion of  contract for specifying the behavior of a function. In
 $\pi$-PEWS this behaviour can be specified by defining the order in which the
 methods of  services will be executed within a business process and methods
 input/output restrictions. 
 }

The $\pi$SOD-M architecture's code generation component (\textit{\textit{Code
generation level}}) is a \textit{$\pi$-PEWS} specification generator. The
code is produced from a \textit{$\pi$-PEWS} model, after it be generated by a
model transformation from \textit{$\pi$-ServiceComposition} model. This
component was implemented using Acceleo \cite{acceleo}. Figure
\ref{fig:sequenceDiagram} presents a sequence diagram describing the model
transformation process and how the designer interacts with the environment to
specify each $\pi$SOD-M model until the specification code is generated.

\begin{figure}[ht!]
\centering
\includegraphics[width=1.0\textwidth]{chapters/implementation/figs/codeGeneration.pdf}
\caption{Acceleo Specification for \textit{$\pi$-PEWS} Code Generation.}
\label{fig:acceleoCode}
\end{figure}

Figure \ref{fig:acceleoCode} presents the \textit{$\pi$-PEWS} meta-model
developed using EMF and some pieces of Acceleo specification. This figure shows
the specification for the \textit{Namescape, Operation, Service} and
\textit{Contract} code generation for \textit{$\pi$-PEWS}. After the code
transformation process, a \textit{.pews} file is created. The listing
\ref{list:namespace}, \ref{list:service} and \ref{list:contract} present parts of Acceleo code for the \textit{$\pi$-PEWS} code generation. The code are the same presented in figure
\ref{fig:acceleoCode}, which presents the relation between the meta-model
concepts and the Acceleo code. The code generation follow the language syntax
described in appendix \ref{append:pews_language}.

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:namespace,caption=Acceleo - Namespace and
Operation Code Specification.] 
1  <%script type="PiPEWSMetamodel.PEWSSpec" name="default"
2  file="<%name%>.pews"%>
3  //-----------------------------------------------------------------
4  //------------ <%name%>.pews Service Specification ----------------
5  //-----------------------------------------------------------------

7  //Namespaces
8  <%for (contains){%>
9    namespace <%name%> = <%WDSLAdress%>
10  <%}%>

12  //Operations
13  <%for (has.eAllContents("Operation")){%>
14    alias <%alias%> = portType/<%name%> in <%isDefinedIn.name%> 
15  <%}%>
\end{lstlisting}
\label{fig:pewscontract} 
\end{figure}

In listing \ref{list:namespace} (lines 1-2) references the 
meta-model, the root element (\textit{PEWSSpec}), and the
name of the generated file (\textit{<\%name\%>.pews}). Lines 3-5 presents the
name of the service specification. Lines 7-10 describes a \textit{for}
integration in the \textit{contains} relationship between \textit{PEWSSpec} and
\textit{Namespace}, and lines 12-15 all operations defined in each
\textit{Namespace} is generated. This operations came from the
\textit{isDefinedIn} relationship between \textit{Namespace} and
\textit{Operation} entities.

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:service,caption=Acceleo - Service Code
Specification.] 
1   //Services
2   <%for (has.eAllContents("CompositeOperation")) {%>	
3     service <%name%> = <%contains.left.filter("Operation").alias%> 
4     <%if ((contains.nameOperator.toString().equalsIgnoreCase("sequence"))) {%> 
5         . 
6     <%}else if((contains.nameOperator.toString().equalsIgnoreCase("parallel"))){%> 
7         || 
8     <%}%> 
9    <%contains.right.filter("Operation").alias%>
10  <%}%>
\end{lstlisting}
\label{fig:pewscontract} 
\end{figure}

Listing \ref{list:service} presents the Acceleo specification for the
\textit{$\pi$-PEWS} model transformation into code. A service is an alias for
one or more operations. Listing \ref{list:contract} specifies the contract generation, using the
\textit{defines} relationship between \textit{APolicy} and \textit{Rule}. Each
contract have a set of rules for the specification.

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:contract,caption=Acceleo - PEWS Contract
Code Specification.] <%script type="PEWSMetamodel.PEWSCTSpec" 
1  //Contract
2  <%for (defines) {%>	
3  defContract <%name%>Contract{
4	 isAppliedTo: <%isRelatedWith.filter("Operation").alias%>;
5	 <%for (defines.filter("Precondition")) {%>
6	 requires: <%defines.expression%>
7  		(OnFailureDo: <%calls.act.toString()%>);
8	 <%}%>
9	 <%for (defines.filter("Postcondition")) {%>
10	 ensures: <%defines.expression%>
11 		(OnFailureDo: <%calls.act.toString()%>); 
12	 <%}%>
13	 <%for (defines.filter("TimeRestriction")) {%>
14	 timeConstraints: <%defines.expression%>
15	 <%}%>
16 }

 <%}%>
\end{lstlisting}
\label{fig:pewscontract} 
\end{figure}

The generated code can be executed in both, \textit{$\pi$-PEWS}
and A-Policy engines\footnote{ Both engines are currently under development. The
\textit{$\pi$-PEWS} engine is being developed at UFRN anf the A-Policy engine in
being developed at Grenoble/France.}  These 2 engines are not native components
in the $\pi$SOD-M plugin. The environment supports the process design to
generate code in any language. New language editor components, like BPEL or XAML, can be easily
coupled to the environment. Therewith, it is necessary to add the language
meta-model and make the transformation process. Thus, from a
\textit{$\pi$-ServiceComposition} model, different models and codes can be
generated (not only \textit{$\pi$-PEWS}). This requires only the definition of
equivalent meta-models, and the corresponding code transformation rules.

The composition engine manages the life cycle of the composition. Once a
composition instance is activated, the engine schedules the composition
activities according to the composition control flow. Each activity is seen as
a process where the service method call is executed. The execution of an
activity has four states: prepared, started, terminated, and failure. The
execution of the control flow (sequence, and/or split and join) can also be
prepared, started, terminated and raise a failure.

At execution time, the evaluation of policies done by the {\sc Policy} manager
must be synchronized with the execution of the services composition (i.e., the
execution of an activity or a control flow).  Policies associated to a scope are
activated when the execution of its scope starts. A {\sc Policy} will have to
be executed only if one or several of its rules is triggered. If several rules
are triggered, the {\em Policy} manager first builds an execution plan that
specifies the order in which such rules will be executed according to the
strategies defined in the following section.        
%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
If rules belonging to several policies are triggered then policies are also
ordered according to an execution plan. The execution of policies is out of the
scope of this thesis, the interested reader can refer to
\cite{Espinosa-Oviedo2011a} for further details.

\begin{landscape}
\begin{figure}
\includegraphics[width=25cm,height=15cm]{chapters/implementation/figs/sequenceDiagram.pdf}
\caption{Model Transformation Process.}
\label{fig:sequenceDiagram}
\end{figure}
\end{landscape}

%  The next section will describe how the creation and generation of each model in
%  SOD-M  is done, for better understanding the transformations process
%  represented in figure \ref{fig:sequenceDiagram} .

\section{Defining Reliable Service Based Applications}
\label{sec:install}

% \footnote{The environment download
% can be performed at http://www3.ifrn.edu.br/\~placido/piSOD-M}

The $\pi$SOD-M environment development
starts with the creation of a project and then the definition of a
\textit{$\pi$-UseCase} model\footnote{To create a model, the user must execute the sequence: \textit{File > New
> Other > EMF Model Wizard}, and choose one of the methodology's model.}, supposing that the business and requirement
specification document have been previously completed.  Figure \ref{fig:screanPiSODM} presents the views provided by the environment:
 \textit{Project view, Menu view, Editor view} and \textit{Properties view}.

% As $\pi$SOD-M environment is based on Eclipse, the developer and analyst
% need not struggle to adapt to the modeling environment. Figure
% \ref{fig:screanPiSODM} presents a tool overview, which has four views:

\begin{figure}[ht!]

\includegraphics[width=.99\textwidth]{chapters/implementation/figs/telaPiSOD-M.pdf}
\caption{$\pi$SOD-M Eclipse Plugin Environment.}
\label{fig:screanPiSODM}
\end{figure}

% For editing is important to follow the concepts presented in $\pi$SOD-M
% meta-models, because it must follow rigorously its structure.

\subsection{\textit{$\pi$-UseCase} Model}

The goal of creating a \textit{$\pi$-UseCase} is to represent the functions and
system services described in the requirements specification and
business specification documents, which are the requirements input for this
phase. In accordance to the process described in figure \ref{fig:sequenceDiagram}, the designer receives the
documents as input and creates the \textit{$\pi$-UseCase} model. With this
model created, the transformation component generates the
\textit{$\pi$-ServiceProcess} model as output.

 To create the \textit{$\pi$-UseCase} model, it is necessary
choose the root element (\textit{Model}) as the starting point of modeling,
during the process of create a ``\textit{new .piusecase file}'' (
using the sequence, \textit{File > New > Other > EMF Model Wizard}). From
this point the model is created as a tree of elements with its specific
references, as shown in Figure \ref{fig:pisodmToolModel}, which shows the model elements and its equivalence in the graphical $\pi$-UseCase
model, each element is built in an iterative way and must obey the hierarchy
and its relationships.

Each model element has a number of childs or siblings to which it is related to,
\textit{e.g.}, an {\sc Use Case} relates {\sc Constraint}, {\sc Extend} and
{\sc Include} elements, as well as {\sc NFRs} relates {\sc NFAs}.
Figure \ref{fig:pisodmToolModel} shows how to create the model elements for the
``\textit{to publish music}'' scenario.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create_model-UseCase.pdf}
\caption{$\pi$-UseCase Model Definition in $\pi$SOD-M Eclipse Plugin.}
\label{fig:pisodmToolModel}
\end{figure}

\begin{exampl}
Items 1 and 2 in Figure \ref{fig:pisodmToolModel} show how to create an
{\sc Actor} and a {\sc Requirement}, item 3 presents the \textit{update music}
{\sc Use Case} and how to create a {\sc Constraint} related to this element.
Item 4 makes reference to a {\sc Constraint} and finally, the items 5 and 6 are
equivalent to a {\sc NFR} and a {\sc Package} elements, respectively. Each item
in the model refers to a equivalent element in the graphical model.
\end{exampl}

Figure \ref{fig:pisodmToolModelProperties} presents some configuration
properties. After creating an element, it is necessary to set its properties.
All elements have properties that describes their relationship and specification
values. These properties are used to give specification details, as
well as future transformations.

% All properties must be set, except of the
% properties are not used in the specification of one specific application, for
% example, there is no exceptional behaviour for an action. In this case it is not
% necessary to set this property. Another example that is no necessary set a
% element property is an actor or use case are not related with a package.

\begin{exampl}
Item 1 in Figure \ref{fig:pisodmToolModelProperties} shows the {\sc Use Case}
element properties for the ``\textit{to publish music}'' application. After the
creation of a use case it is necessary to create the actor, its \textit{name}, which \textit{requirement} the \textit{use case}
belongs, and the respective \textit{package}. Item 2 shows the properties of a {\sc Constraint} element. In a
constraint, its type must be explicit (\textit{VALUE, BUSINESS} or
\textit{EXCEPTIONALBEHAVIOUR}), its \textit{description}, with all candidate
\textit{variables} described with an `\textit{@}', its \textit{name} and which
{\sc Non-Functional Attributes} element this constraint is related to,
\textit{e.g.}, \textit{Authentication}, is an attribute of the
{\sc Non-Functional Requirements}, \textit{Security}. Items 3 and 4 show the
 ``\textit{To Publish Music}'' requirement properties and package
``\textit{app}'', respectively. The properties of each element obey the elements
described in the \textit{$\pi$-UseCase} meta-model.

\end{exampl}

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-UseCase.pdf}
\caption{\textit{$\pi$-UseCase} Properties in $\pi$SOD-M Eclipse Plugin.}
\label{fig:pisodmToolModelProperties}
\end{figure}

\subsection{\textit{$\pi$-ServiceProcess} Models}

The goal of creating a \textit{$\pi$-ServiceProcess} model is to represent the
system execution flow. The designer receives the
\textit{$\pi$-UseCase} model as input to generate the
\textit{$\pi$-ServiceProcess} model. After the \textit{$\pi$-ServiceProcess}
model be generated, the designer calls again the model transformation component
to generate the \textit{$\pi$-ServiceComposition} model as output (figure
\ref{fig:sequenceDiagram}).

 To create the \textit{$\pi$-ServiceProcess} model, it is necessary
choose the root object (\textit{Service Process}) as the starting point of
modeling, during the process of create a ``\textit{new .piserviceprocess file}''
option. From this point on, the model is created as a tree with its specific
references.

%  Figure \ref{fig:piserviceProcessToolModel} 
% presents the service process model, a graphical model reference and detail of
% how create each model element. It is possible to create the $\pi$-ServiceProcess
% model in isolation, however this model is generated from the previous model
% descriptions.

As this model is a refinement of the concepts described in the previous model,
its information is part of the information and properties of the
\textit{$\pi$-UseCase} model, but they focus on the process workflow
and application functions.

\begin{exampl}
Figure \ref{fig:piserviceProcessToolModel} shows the complete model of the example scenario and its main components. From
the root element the user can create the following elements: {\sc Service
Activity, Object Flow, Control Flow, Contract} and {\sc Non-Functional
Attribute}. In Figure \ref{fig:piserviceProcessToolModel}, item marked with 1
highlights the creation of the {\sc Control Flow }. Items 2 and 3 show the elements {\sc Fork
Node} and {\sc Join Node}, respectively, essential for the execution flow
description. Items 4 show the {\sc Action} element, which describes the
\textit{download music, listen music} and \textit{publish twitter} actions and
finally, items 5 highlight the {\sc Assertion} element, which is used to
describe the restrictions over each {\sc Action}.   
\end{exampl}

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create_model-ServiceProcess.pdf}
\caption{\textit{$\pi$-ServiceProcess} Model Definition in $\pi$SOD-M Eclipse
Plugin.}
\label{fig:piserviceProcessToolModel}
\end{figure}

The designer must also configure the properties of each
\textit{$\pi$-ServiceProcess} model element to complete it by specifying all
constraints that compose the contracts and the application process. Figure
\ref{fig:piserviceProcessToolModelProperties} presents the properties of the
main model elements for our example scenario. These properties complement the
modeling of the execution flow.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-ServiceProcess.pdf}
\caption{\textit{$\pi$-ServiceProcess} Properties in $\pi$SOD-M Eclipse Plugin.}
\label{fig:piserviceProcessToolModelProperties}
\end{figure}

\begin{exampl}
Items 1 in Figure \ref{fig:piserviceProcessToolModelProperties}
describe the properties of a {\sc Fork Node}, which is associated with {\sc Edge Activities}
({\sc Control Flow} and {\sc Object Flow}) that connect the {\sc Action}
elements. For example, the {\sc Fork Node} \textit{fn\_toPublishMusic} has an
input edge (\textit{cf1}) and three output streams edges (\textit{cf2, cf3} and
\textit{cf5}). All these elements are {\sc Control Flow} entities (items 3).
Items 2 show the {\sc Action} properties, and items 4 and 5 describe a {\sc
Contract} and its {\sc Assertions} properties, respectively. A {\sc Contract}
has a \textit{name} and the information of an {\sc Action}.
{\sc Assertions} are {\sc Contract}'s child element. Each  {\sc Contract} can
have many  {\sc Assertions}. Each  {\sc Assertion} has six properties, they are:
(i) \textit{AProperty}, which describes the runtime verification execution time,
(ii) \textit{description}, (iii) \textit{variable name}, (iv and v) a maximum
and minimum allowed to be checked (\textit{MaxValue} and \textit{MinValue}) and
the variable (vi) \textit{type}. In cases of non numeric types, \textit{e.g.}
\textit{Short Int, Flot} and \textit{Double}, only \textit{MinValue} property
value is considered property  and\textit{MaxValue} is described as
\textit{null}.
\end{exampl}

\subsection{\textit{$\pi$-ServiceComposition} Models}

The goal of creating a \textit{$\pi$-ServiceComposition} model is to represent
the system service compositions and to reference the external services
that are called by the application. The designer receives the
\textit{$\pi$-ServiceProcess} model as input to generate the
\textit{$\pi$-ServiceComposition} model. After the
\textit{$\pi$-ServiceComposition} model is generated, the designer calls again
the model transformation component to generate the \textit{$\pi$-PEWS} model as
output (figure \ref{fig:sequenceDiagram}). The output model describes the system
specification in a specific platform.

To create the \textit{$\pi$-ServiceComposition} model, it is necessary to choose
the root object (\textit{Composition Service Model}) as the starting point of modeling,
during the process of creating a ``new .piservicecomposition file'' option. From
this point on, the model is created as a tree with its specific references.

Recall that the \textit{$\pi$-ServiceComposition} model is a refinement of
\textit{$\pi$-ServiceProcess} model, most elements are the same, except for {\sc
Business Collaborator, Policy, Rules} and {\sc Variable}. Thus, we will detail these elements in the
model editor description.

\begin{exampl}
Figure \ref{fig:piserviceCompositionToolModel} shows
the relationship of these elements for our example scenario. Each {\sc Action}
has an specific {\sc Business Collaborator} (item 3), this elements express the external service
provider, such as Facebook, Twitter or Bank. Another example of a {\sc Business
Collaborator} definition is a WSDL specification. The WSDL namespaces represents
a kind of {\sc Business Collaborator}. Besides {\sc Business Collaborator}
element, the main children root element (\textit{Composition Service Model})
are: {\sc Policy} and {\sc Service Activity} (items 1), and each {\sc Policy}
(items 2) is directly related with {\sc Service Activities}. From a {\sc Policy}
it is possible to create a set of {\sc Rules} (item 4). Figure
\ref{fig:piserviceCompositionToolModel} also presents an equivalent graphic
model of the application. In this figure, the \textit{Application}'s {\sc
Business Collaborator} presents only the main {\sc Actions} (\textit{buy music, listen music, publish
Facebook} and \textit{publish twitter}) that are related with a {\sc Service
Activities} that have one {\sc Policy}. The \textit{Application}'s {\sc Business
Collaborator} model is the same presented in the previous section
(\textit{$\pi$-ServiceProcess} model).

\end{exampl}

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create-model-Serviceconposition.pdf}
\caption{\textit{$\pi$-ServiceComposition} Model Definition in $\pi$SOD-M
Eclipse Plugin.}
\label{fig:piserviceCompositionToolModel}
\end{figure}

The properties of \textit{$\pi$-ServiceComposition} elements are
configured as the same way the other two previous editors, and the properties
described in the \textit{$\pi$-ServiceComposition} meta-model
(Figure \ref{fig:servicecomposition}). Thus, to configure the properties of this model,
it is necessary simply choose the desired element and to modify its values.

\subsection{\textit{$\pi$-PEWS} Models}

The goal of creating a \textit{$\pi$-PEWS} model is to represent
the application in a specific platform. The designer receives the
\textit{$\pi$-ServiceComposition} model as input to generate the
\textit{$\pi$-PEWS} model. After the
\textit{$\pi$-PEWS} model be generated, the designer calls again
the model transformation component to generate the \textit{$\pi$-PEWS}
specification code as output (figure \ref{fig:sequenceDiagram}). This code will
be executed.

To create the \textit{$\pi$-PEWS} model, it is necessary to choose the root
object (\textit{PEWS Spec}) as the starting point of modeling. From
this point on, the model is created as a tree with its specific references.

The \textit{$\pi$-PEWS} meta-model is a representation of the
\textit{$\pi$-PEWS} language. Each \textit{$\pi$-PEWS} model is a
program/specification written in that language. The elements described in this
model represent parts of language's grammar constructs. Each entity in the model
represents a piece of \textit{$\pi$-PEWS} code.

\begin{figure}[ht!]
\centering
\includegraphics[width=.90\textwidth]{chapters/implementation/figs/create_model-PEWS.pdf}
\caption{\textit{$\pi$-PEWS} Model Definition in $\pi$SOD-M Eclipse Plugin.} 
\label{fig:piPEWSToolModel}
\end{figure}

Figures \ref{fig:piPEWSToolModel} and \ref{fig:piPEWSToolModelProperties}
show how the elements that compose a \textit{$\pi$-PEWS} model can be specified
in our tool. This model is generated automatically from the
\textit{$\pi$-ServiceComposition} model.

\begin{figure}[ht!]
\centering
\includegraphics[width=.90\textwidth]{chapters/implementation/figs/create_model-PEWSProperties.pdf}
\caption{\textit{$\pi$-PEWS} Model Properties in $\pi$SOD-M Eclipse Plugin.} 
\label{fig:piPEWSToolModelProperties}
\end{figure}

\textit{$\pi$-PEWS} is the last model generated before the code generation. At
this stage, it is spected that the designer proceeds with a general (manual)
check of consistency of the model. It is
important to remark that the environment alone does not replace the
modeling work required to design and develop services based applications.

\section{Extending the Environment} 
\label{sec:extendingEnvironment}

Both, the $\pi$SOD-M environment and methodology can be extended, in order
to improve the components that describe and implement the methodology.
Extensions can be done in two different levels: (i) adding new models to the
existing infra-structure, and (ii) considering more abstract levels.

The extension may be in terms of language: new meta-models for other
languages can be described and coupled to the environment. The extension process
should take place as follows: new languages meta-models may be designed and
coupled in the environment architecture (such as BPEL, XAML, WSDL and XLANG).
After creating the desired meta-model, a mapping must be done. The mapping must
respect the \textit{$\pi$-ServiceComposition} meta-model. It is also necessary
to describe the rules for code generation, using a code generator engine such
Acceleo.

% Thus, new
% languages may be added for the environment improvements and for improve the
% methodology flexibility.

Considering the more abstract level of the methodology, new
meta-models can be described. For instance, the computing independent level
(CIM level) of SOD-M may be added to our methodology to describe the
requirements and business restrictions in terms of models.

% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.95\textwidth]{chapters/implementation/figs/componentes.pdf}
% \caption{$\pi$SOD-M Environment Extension Components.}
% \label{fig:extension}
% \end{figure}
% 
% 
%  
% 
% Figure \ref{fig:extension} presents how the environment can be extended.
% The components that can be inserted components of CIM-to-PIM transformation
% models (item 1), new languages meta-models description (item 2) and
% components which represent the transformations between the
% \textit{$\pi$-ServiceComposition} model and each language (item 3). Thus, it is
% possible to extend the development environment of SOD-M. The extension must comply with
% the methodology concepts and environment architecture.

\section{Conclusion} 
\label{sec:env_conclusion}

This chapter introduced the implementation
 of the $\pi$SOD-M methodology environment. We also presented a representation
 for our model description. The implementation includes all (i) meta-models of
 $\pi$SOD-M, (ii) editors for each model for the applications being developed,
 and (iii) the plugins for the transformations defined by the methodology. An
 example was also presented to describe the environment features.

%  This graphical representation can be better refined
%  in future work, focussing on processing models or on the development of a
%  visual tool for processing models and code generation.

% It was
% presented and described the general environment architecture, its properties, language used, its execution environment and how to extends the environment. The
% development environment of $\pi$SOD-M supports the methodology for a better use of their concept.

% Using the proposed graphical nomenclature, we described all the transformation
% rules that were used for the development of the environment's plugin responsible
% for the transformation of all $\pi$SOD-M models. This chapter also presented, in
% detail, all $\pi$SOD-M meta-models and editors, such as how to extend the
% environment and use it. Finally, after the design description, were
% introduced the execution environment example for the development each models in
% $\pi$SOD-M context.

### 2013-09-30 WESOA2013/source/auxiliar/validation.tex
inicial --martin

\chapter{Evaluation}
\label{chapter:validation}

\epigraph{``\textit{Let us change our traditional attitude to the construction
of programs: Instead of imagining that our main task is to instruct a computer
what to do, let us concentrate rather on explaining to human beings what we want
a computer to do.}''}{Donald E. Knuth}

This chapter describes the experimentation we conducted for
\correctingText{evaluating} the $\pi$SOD-M methodology. The chapter presents
three \correctingText{examples}. The first two \correctingText{examples} concern simplified
service-based applications that focus on the application's logic and explicit business rules. They show
how $\pi$SOD-M can be used for developing applications and exemplify the
consideration of non-functional aspects along the whole
development process. The last \correctingText{example}
(\textit{GesIMED}~\cite{valeriaThesis} application) \correctingText{aims to perform an
evaluation of our approach, analyzing the main differences and particularities
between SOD-M and $\pi$SOD-M.}

\newText{The objective of the experimentation was to evaluate the features and
concepts proposed by $\pi$SOD-M, considering the implementation of some
examples, and perform a comparison with SOD-M. This comparison focuses on
verifying which of those proposed aspects favored the development of
service-oriented applications in the presence of non-functional requirements. 
In this direction, our intention is: (i) analyze if it is possible to model
non-functional requirements from initial levels of abstraction in the
service-oriented development, (ii) and verify what are the advantages of using
the proposed approach, compared with the original proposal of SOD-M.}

% O objetivo principal deste capítulo é avaliar as características e conceitos
% adicionados em piSOD-M e fazer uma comparação com SOD-M, com a intenção de
% destacar que é possível modelar requisitos não funcionais desde níveis iniciais
% de abstração no desenvolvimento orientado a serviços.

%  used for
% validating the specification and generation of the models proposed in
% $\pi$SOD-M. The validation split the thesis into parts, each one can be
% validated individually. This parts are: (i) the modelling of each $\pi$SOD-M
% model, (ii) the model transformations and (iii) the results.
% 
%  
% The validation is performed through three case studies. The first case study
% aims to perform a qualitative evaluation of our approach, analyzing the main
% differences and particularities between SOD and $\pi$SOD-M. The others case
% studies aim to validate and verify the use of $\pi$SOD-M and its
% environment, analyzing the positive results and the benefits to 
% model non-functional requirements from early stage of development.

%The thesis' validation of the  follows from the sub-validations.

%\bigskip

% The \textit{IEEE Standard Computer Dictionary} \cite{ieeeDict} defines
% validation as \textit{the process of evaluating a system or component during or
% at the end of the development process to determine whether it satisfies specified
% requirements}. If we apply this definition to a methodology: \textit{Validation
% is the process of evaluating a method during or at the end of the development process to determine whether it
% satisfies specified requirements and eases the development of a system}. This
% chapter validates the methodology proposed in this thesis in order to verify
% whether it promotes a better development of reliable
% services composition applications.

% The steps during the thesis were analyzed by application of laboratory case
% studies with an iterative and incremental approach. Thus, while the $\pi$SOD-M
% concepts have been defined, the concepts application have been made in each case
% studies to verify its validity.

% The validation process has been performed in a continuous and
% progressive approach along the definition of $\pi$SOD-M.

This chapter is organized as follows: section
\ref{sec:toPublishMusic_inpractice} shows the \textit{To Listen Music}
\correctingText{example}, section \ref{sec:crimes_inpractice} introduces the
\correctingText{example} \textit{Crime Map} and section \ref{sec:siw_gesimed}
introduces the \textit{GesIMED} \correctingText{example} that we used for
validating the methodology. Section \ref{sec:leassonsLearned} presents the lessons learned
with the validation process. Finally, Section \ref{sec:conclusion_validation} concludes
the chapter and discusses on validation results.

% From strategy for
% validation, the process was carried out with the development resolution step as
% part of an iterative process and feedback cyclic, allowing solve problems of
% concepts and its application while the methodology was being defined. Thus, in
% each step of methodology the validation was applied iteratively
% on two different case studies.
% 
% The process defined in the research method validation was therefore not
% a linear process, but iterative and incremental, thus providing the opportunity
% to go strengthen the results and learning from each research step.  
% 
% Following, each developed case study is described, indicating both the needs
% identified, and new contributions emerged from them, which are also shown
% graphically, using the methodology models. The case studies resolution were
% performed using the proposed approach.

% \section{$\pi$SOD-M in Practice}
% \label{sec:execution}

% This section presents, in detail, the design and implementation of two case
% studies conducted to validate the method proposed in this
% thesis. Both case studies were developed using the models proposed by
% $\pi$SOD-M.

% Following all models will be developed for this case and then
% describe the main aspects of the implementation of each of them.

\section{Example 1: To Publish Music}
\label{sec:toPublishMusic_inpractice}

An organization wants to provide the service-based
 application that monitors the music a person is listening during some periods
 of time and sends the song title to this person's Twitter or Facebook accounts.
 Thus, this social network user will have her status synchronized in Twitter or
 Facebook. The scenario begins by get through the music service Spotify for
 fetching the user's musical status. The social network services are then
 contacted to modify the user's status with the music information. The user can
 also download a specific music, he can proceed with the download process, which
 includes the payment process.

The user can download a music, after having proceed with the
 payment via PayPal or credit card. All services are available via Spotify and
 this application needs to interact with Spotify users, so they can listen
 music, publish them on Facebook and Twitter, or buy music online. The scenario
 to monitor and publish songs has the User and Spotify actors.

The functionality of the application are: \textit{Search music;
Choose music; Download music; Listen Music; Buy music} and \textit{Publish music}.

 \subsection{$\pi$-UseCase Model}
\label{sec:usecase_topublishMusic}

Figure \ref{fig:publishMusic_usecase} shows the main part of the use case model
for the \textit{To Publish Music} application\footnote{The complete
\textit{$\pi$-UseCase} model can be found in appendix \ref{append:diagrams}.}.
The $\pi$-UseCase model describes the system functions and constraints. Figures
\ref{fig:toolpublishMusic_usecase} and \ref{fig:toolpublishMusic_usecase2}
detail the model and configuration properties for the described example
(actual view taken from the $\pi$SOD-M environment).
Three restrictions are specified on the use cases described by the
model, two on the \textit{buy music} use case and a restriction on the
\textit{pay} use case. These restrictions are value and business constraints.
\textit{Buy music} restrictions are related to secure connection and to
ensure that the user needs to have an account in the music network. They are
marked using \textit{\#\ldots\#}, figures \ref{fig:publishMusic_usecase},
\ref{fig:toolpublishMusic_usecase} and \ref{fig:toolpublishMusic_usecase2}.
Notice that those NFRs and constraints will be the base to define assertions and
contracts by the transformations of models.

\begin{figure}[ht!]    
\centering
\includegraphics[width=.99\textwidth]{chapters/methodology/figs/piusecase/application/packages2.pdf}
\caption{\textit{To Publish Music} $\pi$-UseCase.}
\label{fig:publishMusic_usecase}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create_model-UseCase.pdf}
\caption{\textit{To Publish Music} $\pi$-UseCase Environment Detail.}
\label{fig:toolpublishMusic_usecase}
\end{figure}

Figures \ref{fig:toolpublishMusic_usecase} and
\ref{fig:toolpublishMusic_usecase2} present the update music model and show how
the properties are defined in the $\pi$SOD-M environment. This example describes
the properties over the \textit{update music} (item 4, figure
\ref{fig:toolpublishMusic_usecase}) use case and the restriction over the
authentication process (item 3, figure \ref{fig:toolpublishMusic_usecase}). To
update a music in a social network, the user must be authenticated.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-UseCase.pdf}
\caption{\textit{To Publish Music}$\pi$-UseCase Environment Detail.}
\label{fig:toolpublishMusic_usecase2}
\end{figure}

The modeling of the application properties using the environment $\pi$SOD-M
are the same presented in the $\pi$-UseCase model, as described in Figure
\ref{fig:toolpublishMusic_usecase2}. For example, the authentication restriction
 details the type of constraint (\textit{VALUE}), a \textit{description},
 \textit{name} and a \textit{non-functional attribute}
 (\textit{authentication}). These properties are described in all constraints
 (item 2, figure \ref{fig:toolpublishMusic_usecase2}).

Figure \ref{fig:servicesToPublishMusic} shows the service components that must
interact with the \textit{application} and the interaction among them. There are
four services that are invoked to produce results for the \textit{Publish Music} application:
\textit{Spotify, Facebook, Twitter} and \textit{Bank}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.8\textwidth]{chapters/validation/figs/toPublishMusic}
\caption{To Publish Music Services.}
\label{fig:servicesToPublishMusic}
\end{figure}

 \subsection{$\pi$-ServiceProcess Model}
\label{sec:serviceProcess_topublishMusic}

Figure \ref{fig:publishmusic_piserviceProcessToolModelProperties} shows the
\textit{$\pi$-ServiceProcess} diagram for the execution of activities based on
contracts. We present here the part of the diagram produced from the concepts of the figure
\ref{fig:toolpublishMusic_usecase2}. The complete \textit{$\pi$-ServiceProcess}
diagram of the application is shown in figure \ref{fig:serviceprocessContract}.
Notice that it is necessary to represent each {\sc Constraint} modelled in a {\sc Contract} based service activity diagram. Each constraint
described in the previous model is transformed into a contract in the
$\pi$-ServiceProcess model.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-ServiceProcess.pdf}
\caption{\textit{To Publish Music} $\pi$-ServiceProcess.}
\label{fig:publishmusic_piserviceProcessToolModelProperties}
\end{figure}

\textit{Publish Twitter} and \textit{publish Facebook} (figure
\ref{fig:publishmusic_piserviceProcessToolModelProperties}) have pre- and
post-conditions assertions that are grouped into a contract for each service:
(i) verify if the user data are correct; (ii) if the user is already logged in
Spotify; As post-condition, it ensures the complete transaction and
verifies whether the payment authorization notification was sent to the user and to Spotify.
There may be different restrictions depending on the services participating in
the publish process. The user must be logged in Spotify in order to access
Facebook, and Facebook must send a ``successful login'' acknowledge which is
verified by the post-condition. For publishing a message, Twitter imposes
pre-conditions on the length of the message and user data login. In contrast,
Facebook only sends a confirmation acknowledge. Items 4 and 5 in figure
\ref{fig:publishmusic_piserviceProcessToolModelProperties} show the
pre-conditions for ``\textit{publish Twitter}'' concerning \textit{(i)} the format
of the music file and (\textit{ii}) access authorization credentials, namely
login and password.

\correctingText{The development of the $\pi$-ServiceProcess refine
the constraints into contracts for the services as shown
in Figure \ref{fig:publishmusic_piserviceProcessToolModelProperties}.}

% , about the payment authorization. To \textit{update} services, depending of each
% service, there may be different restrictions.

%  The Facebook service requires that the user is already logged on
%  Spotify and these data are the same as Facebook. As post-condition, it ensures
%  that the Facebook service send a notification of success. The Twitter service
%   requires pre-conditions over the message length and user privacy data, while
%   to update the Facebook is a necessary precondition and a confirmation notice
%   as post-condition. As a precondition for ``\textit{publish twitter}'' (itens
%   4 and 5, figure \ref{fig:publishmusic_piserviceProcessToolModelProperties}) it is necessary that
%   \textit{(i)} the music format is correct and (\textit{ii})  the twitter login and password is correct for the update.

 \subsection{$\pi$-ServiceComposition Model}
\label{sec:serviceComposition_topublishMusic}

 There are four external {\sc Business
Collaborators} in this application: \textit{Bank, Spotify, Twitter} and
\textit{Facebook} (items 3, figure
\ref{fig:topublishMusic_piserviceCompositionToolModel}). Figure \ref{fig:topublishMusic_piserviceCompositionToolModel} shows the application
services that consist of a set of functions: \textit{search music, select
music, buy music, download music, listen music} and \textit{publish music}. The
\textit{publish music} action calls two service collaborators (Facebook and
Twitter), and the \textit{buy music} action calls two functions of the Bank
service collaborator.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create-model-Serviceconposition.pdf}
\caption{\textit{To Publish Music} $\pi$-ServiceComposition Model.}
\label{fig:topublishMusic_piserviceCompositionToolModel}
\end{figure}

The \textit{To Publish Music} $\pi$-ServiceComposition model defines three
policies: \textit{transactionPolicy} (item 4, figure
\ref{fig:topublishMusic_piserviceCompositionToolModel}), \textit{httpAuthPolicy}
(item 2) and \textit{authPolicy} (item 6). The \textit{trasactionPolicy}
verifies the Bank data restrictions over the payment and confirmation process.
The \textit{httpAuthPolicy} policy specifies rules over the
Facebook function for the authentication and notification processes. These rules
are executed during the publish massage function request. The
\textit{authPolicy} policy implements the open authentication protocol
required for accessing the Twitter service.

From the \textit{To Publish Music} rules and policies it is possible to model
and associate non-functional properties to services' compositions,
\textit{e.g.} the integration of information retrieved from different
social network services, automatic generation of an integrated view of the
operations executed in different social networks or for providing security in
the communication channel when the payment service is called.

From this stage, the \textit{$\pi$-PEWS} model (PSM) depicted in figures
\ref{fig:pewsRepresentation} and \ref{fig:contractRepresentation} can be
generated.

%  \subsubsection{$\pi$-PEWS Model}
% \label{sec:pews_topublishMusic}

% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.90\textwidth]{chapters/implementation/figs/create_model-PEWSProperties.pdf}
% \caption{$\pi$-PEWS Model Properties in $\pi$SOD-M Eclipse Plugin.} 
% \label{fig:piPEWSToolModelProperties}
% \end{figure}

% \section{Results}
% \label{sec:results}

\section{Example 2: Crime Map}
\label{sec:crimes_inpractice}

The \textit{Crime Map} \correctingText{example} addresses the design and
implementation of a system that deals with statistical data on criminal activities. The system works
in conjunction with third-party services, in order to process a search supplied
by a user. The system must provide a public utility service to inform
citizens about the real crime situation in a particular location or region.
Making use of third party services, the application must provide a statistics crime portal. The result depends on user queries and upon a user
demand results can be presented on a map. The system also interacts with
micro-blogging, for looking up messages containing specific crime
information. The presentation of the data concerning criminal activities is
provided by the \textit{Police service}. The presentation of a locality map
crime is made by the \textit{Google Maps service} and messages presentation on
the micro-blogging is done by the \textit{Twitter service}. The system also has
a \textit{Post service}, for searching an address location that can be then
presented in a map by Google maps. The scenarios to query data and crime
statistics have the following actors: \textit{(i) User, (ii) Map service, (iii) Post service, (iv) Micro-blogging
service} and \textit{(v) Police service}, to query the criminal activities
(figure \ref{fig:usecaseCrimeMap_total}).

Considering the application functions, the crimes may be searched by types,
\textit{e.g.} \textit{theft, murder, kidnapping}, etc. The search can also be
made by city' district or region, for example, \textit{center, west, south,
north} or \textit{east}, and it can process search temporal criteria, for
example, crimes of the previous day of the last weeks or months. Finally, the
service can also process conjunctive queries, \textit{e.g.}, \textit{the number of kidnapping of a specific district in the last month}.

Whenever a new crime is published by the police, the system should post a
 message on Twitter stating the new crime, with a specific hashtag,
 \textit{e.g.} \textit{\#crime043\_kidnapping\_center\_05-08-2012}. The hashtag
 structure contains the crime number, type, local and the day it happens. From
 this, users can comment about the crime and these comment will appear in the
 application interface. All users must have an account to use the application.

 The application functionalities are: \textit{Choose search type of crime; see
 crime information; Receive information on the chosen query; See map; List crimes;}
 and \textit{Comment about a crime.}

 Non-functional requirements for this application include: \textit{queries
 response performance, crime information validation}, \textit{crime
 information presented in a map} and \textit{ensuring crimes data security.}

\subsection{$\pi$-UseCase Model}
\label{sec:usecase_crimesMap}

% The modeling of the system starts with the $\pi$-UseCase model. This involves
% determining the services identified in the list of requirements, which final
% consumers identified in the list of business services, and which of the will be
% end consumers or service users. Thus, it is possible to identify the modeling
% elements that are part of the use case model.

 Figure \ref{fig:usecaseCrimeMap_total} presents the Crime Map $\pi$-UseCase
model. This model includes examples of \textit{business, value}
and \textit{exceptional behaviours} constraints. The application accesses
data from third party services to provide the user an easy way to visualize new
information, these services are: \textit{Map, Police, Postal} and
\textit{Microb-logging Services}.

The user interacts with the application that will perform the services
for obtain and publish details about the crimes. The services are modeled as
actors (they are external applications). Each service has a set of functions
that can be performed. The $\pi$-UseCase model describes both the external services
and the functionality of the application (see figure
\ref{fig:usecaseCrimeMap_total}).

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase}
\caption{Crime Map $\pi$-UseCase.}
\label{fig:usecaseCrimeMap_total}
\end{figure}

The user can search for crimes and view its occurrences in a map. The search may be performed by \textit{type of
crime, region} or \textit{period}. The user can also perform a search that
combines these criteria. If the application's user does not choose any
option, the search will retrieve up to 30 recent occurrences. If the location
maps service is unavailable or there is no response from it after 5 seconds,
only the crimes information will be presented, without a map view.

Figure \ref{fig:usecaseCrimeMap_total} also
presents the $\pi$-UseCase model tree representation generated by the $\pi$SOD-M
environment model .Figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2} detail the
use cases and constraints presented in Figure
\ref{fig:usecaseCrimeMap_total}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase1}
\caption{\textit{Search crime} $\pi$-UseCase Detail.}
\label{fig:usecaseCrimeMap1}
\end{figure}

The \textit{search crimes} use case (Figure \ref{fig:usecaseCrimeMap1})
requests the \textit{Police} service to verify the crimes according to the
parameters supplied by the user. The restriction of this use case depends on
the parameters selected by the user. If the user chooses one or several types of
specific crime's type, the system will present those crimes that have more than
five occurrences. For \textit{search by address}, the address must comply with
the restrictions of the police service, the research can be done by street,
neighborhood, city or district. \textit{Search by period} will provide only
occurrences within the last six months.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase2}
\caption{\textit{See crime information} $\pi$-UseCase Detail.}
\label{fig:usecaseCrimeMap2}
\end{figure}

Figure \ref{fig:usecaseCrimeMap2} describes the design of the use cases:
\textit{see crime information, see statistics, show map} and \textit{share
information}. To share information of a crime, it is necessary to access the
Twitter micro-blogging service to proceed with the authentication and posting.
The service requires the user id and password, however the
application must ensure the User privacy information while performing
authentication. To see statistics of crimes, the application should process a
larger volume of data to generate statistical results. Thus, it is required to
provide resources to ensure the data processing performance. The restriction on
map presentation is over response time and accessibility of the \textit{Postal}
and \textit{Google maps} services. The response time limit is 5 seconds. If the
map is not processed, only the information of the crime are presented. Figures
\ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2} also show the
properties used in the $\pi$SOD-M environment and the equivalent concepts in the
model.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.8\textwidth]{chapters/validation/figs/crimesMap}
\caption{Crime Map Services.}
\label{fig:servicesCrimeMap}
\end{figure}

After the identification the various functions that are required by the system
to perform the business services, the $\pi$-ServiceProcess model is used to
represent the workflow necessary to perform a service within the system.

\subsection{$\pi$-ServiceProcess Model}
\label{sec:serviceprocess_crimesMap}

The $\pi$-ServiceProcess model is used to represent the workflow necessary to
perform the system service. The service processes execution is described in two
steps. Figures \ref{fig:serviceProcessCrimeMap} and \ref{fig:serviceProcessCrimeMap2} show the
flow of functions execution described in the $\pi$-UseCase models shown
in Figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}. Each
diagram can represent different business services (see figures
\ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}).

The diagrams of figures \ref{fig:serviceProcessCrimeMap} and
\ref{fig:serviceProcessCrimeMap2} were obtained by applying the
\textit{$\pi$-UseCase2$\pi$-ServiceProcess} transformation to the diagrams of
figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}, respectively.

%  Thus, it makes easier to represent the performance of services. It is also possible to describe the
% process in a large workflow, but for applications running multiple
% services, it is better to have smaller processes to better represent the
% service execution. Thus we present the services business through the different
% processes as presented in Figures \ref{fig:usecaseCrimeMap1} and
% \ref{fig:usecaseCrimeMap2}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceProcess}
\caption{\textit{Crime information} $\pi$-ServiceProcess.}
\label{fig:serviceProcessCrimeMap}
\end{figure}

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceProcess_crime1}
\caption{\textit{Crime information}
$\pi$-ServiceProcess Environment.}
\label{fig:toolserviceProcessCrimeMap}
\end{figure}

Each constraint of figures \ref{fig:serviceProcessCrimeMap} and
\ref{fig:serviceProcessCrimeMap2} are transformed into assertions. For example,
the assertions of address format,  crime types and time are restrictions over 
the action \textit{search crime}, being one pre-condition and two
post-conditions. These restrictions form the contract on the \textit{search
crime} function. Similarly, for viewing maps, if the answer to the request
takes longer than 5 seconds (5000 milliseconds), the map display is suspended and only 
the information is presented. Each assertion in this $\pi$-ServiceProcess model
stems from the $\pi$-UseCase model constraints. These restrictions are shown
in Figure \ref{fig:serviceProcessCrimeMap}. Figure
\ref{fig:toolserviceProcessCrimeMap} presents the equivalent model
generated by the methodology transformation in the $\pi$SOD-M environment for
the process detailed in Figure \ref{fig:serviceProcessCrimeMap}. Each node
defines the process element and its properties.

In Figure \ref{fig:toolserviceProcessCrimeMap}, \textit{i\_node} corresponds
to the initial process node. {\sc Actions} are grouped into  {\sc Service
Activities}, such as \textit{choose crimes property} action that is part of
the \textit{search crime} service activity. A {\sc Control Flow} is an edge that
links two nodes, for example \textit{cf1} connect the initial node
\textit{i\_node} (source) with the \textit{choose crimes property} action
(target). Figure \ref{fig:toolserviceProcessCrimeMap} also presents the
assertions described in Figure \ref{fig:serviceProcessCrimeMap} grouped into 
{\sc Contracts}. The contracts are: \textit{searchCrimeContract} and
\textit{seeCrimeInformationContract}. These contracts are related with the
actions \textit{search crime} and \textit{see crimes information}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceProcess2}
\caption{\textit{See crime statistic and share information} $\pi$-ServiceProcess
Detail.}
\label{fig:serviceProcessCrimeMap2}
\end{figure}

Figure \ref{fig:serviceProcessCrimeMap2} presents the assertions over the
 \textit{see statistics} and \textit{share information} actions. Both have 2 
pre-conditions. To \textit{share information} on Twitter it is necessary
to verify the format of crime information (140 characters) and the Twitter id
and password, for authentication. Regarding the \textit{see statistics} action, there is a
business restriction over data volume, and a value restriction over the crime
information format. Thus, the presentation of the crimes statistics must be done
after the contract verification. Figure
\ref{fig:toolserviceProcessCrimeMap2} presents the equivalent model
generated by the methodology transformation in the $\pi$SOD-M environment for
the process detailed in Figure \ref{fig:serviceProcessCrimeMap2}. Each node
defines the process element and its properties. Figure
\ref{fig:toolserviceProcessCrimeMap2} also presents the assertions grouped
into three contracts that are related with its specific actions. The contracts
are: \textit{seeStatisticsContract}, \textit{seeCrimeInformationContract} and
\textit{shareInformationContract}. The contracts are related with the action
\textit{see statistics}, \textit{see crimes information} and \textit{share
information}, respectively.

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceProcess_crime2}
\caption{\textit{Crime statistic and share information}
$\pi$-ServiceComposition Environment.}
\label{fig:toolserviceProcessCrimeMap2}
\end{figure}

These models provide an overview of the business processes from the requirements
described in the $\pi$-UseCase model. These models also provide a more detailed
view of the execution of business processes and the possible
interaction with external services. The description of this interaction is the
result of the refinement of these models ($\pi$-ServiceProcess models) through
the $\pi$-ServiceComposition model.

\subsection{$\pi$-ServiceComposition Model}
\label{sec:servicecomposition_crimesMap}

Figures \ref{fig:serviceCompositionCrimeMap} and
\ref{fig:serviceCompositionCrimeMap2} show the $\pi$-ServiceComposition models,
which represent the service composition processes of each
business service and additional indication of which are members of the business
that execute the action. Actions are derived from service activities,
identifying the set of actions that are necessary for the completion of each service.

The main partition expresses the \textit{Application} execution
(\textit{External false}) that represents the general process. Actions have the equivalent
service relation, which realize the external service, such as \textit{Police}, \textit{Google Maps} or
\textit{Twitter} services. They are expressed by a {\sc Business Collaborator}.
Both, Figure \ref{fig:serviceCompositionCrimeMap} and
\ref{fig:serviceCompositionCrimeMap2} present two external {\sc Business
Collaborators} describing the application services that are invoked. These
$\pi$-ServiceComposition models refine the $\pi$-ServiceProcess models, matching
action with real service functions that may be executed by the system
application.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceComposition}
\caption{\textit{Crime information}
$\pi$-ServiceComposition.}
\label{fig:serviceCompositionCrimeMap} 
\end{figure}

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceCOmposition_crime1}
\caption{\textit{Crime information}
$\pi$-ServiceComposition Environment.}
\label{fig:toolserviceCompositionCrimeMap}
\end{figure}

For each $\pi$-ServiceComposition model, the contracts described in the
$\pi$-ServiceProcess model are grouped in policies. Figure
\ref{fig:serviceCompositionCrimeMap} describes the \textit{performancePolicy}
and \textit{reliabilityPolicy} policies, both associated with a service activity
and its actions. It is important to highlight that all policies are applied to the
entire actions associated with a service activity, for example, the
\textit{performancePolicy} verifies the request time for presenting the Google
maps. If the time exceeds 5000 milliseconds, the maps call is ignored and only
the crime information is presented. The \textit{reliabilityPolicy} is
associated with crime data verification. There are three rules over crime
address format, types of crime and the period of time in which it happens. The
crime presentation may obey the pre- and post-condition described in this policy. Figure
\ref{fig:toolserviceCompositionCrimeMap} presents the $\pi$SOD-M tool
description for the model described in Figure
\ref{fig:serviceCompositionCrimeMap}.

\begin{figure}[ht!]   
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceComposition2}
\caption{\textit{Crime statistic and share information}
$\pi$-ServiceComposition.}
\label{fig:serviceCompositionCrimeMap2}
\end{figure}

\begin{figure}[ht!]    
\centering
\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceCOmposition_crime2}
\caption{\textit{See crime statistic and share information}
$\pi$-ServiceComposition Environment Detail.}
\label{fig:toolserviceCompositionCrimeMap2}
\end{figure}

Figure \ref{fig:serviceCompositionCrimeMap2} refines the $\pi$-ServiceProcess
model described in Figure \ref{fig:serviceProcessCrimeMap2} and Figure
\ref{fig:toolserviceCompositionCrimeMap2} presents the $\pi$SOD-M tool
description for the model described in Figure
\ref{fig:serviceCompositionCrimeMap2}. This model details the policies for
\textit{crime statistic} and \textit{share information} services. The
policies are \textit{statisticPolicy} and \textit{authenticationPolicy},
respectively.

These models provide an overview of the external services and its composition
from the actions described in the $\pi$-ServiceProcess model. These models also
provide a more detailed view of the system restrictions, applying policies
over external services. This model refines the $\pi$-ServiceProcess models
detailing the Business Collaborators that are expressed as external services,
and grouping contract into policies.

% From these composition models, next
% step is generate the specification system code in $\pi$-PEWS.

% \subsubsection{$\pi$-PEWS Model}
% \label{sec:pews_crimesMap}

\begin{figure}[ht!]
\tiny
\centering

\begin{lstlisting}[label=list:crimeInformationPEWS,caption=pi-PEWS
Specification: Crime information. ]

1  ns police =  "http:\\www.pm.rn.gov.br/service"
2  ns googleMaps =  "http://maps.googleapis.com/maps/api/"

4  alias chooseCrimesProperties = portType/chooseCrimesProperties in police
5  alias searchCrimes = portType/searchCrimes in police
6  alias seeMap = portType/showMap in googleMaps
7  alias seeCrimesInformation = portType/getCrimeDetail in police

9 (chooseCrimesProperties . searchCrimes)* . seeMap . seeCrimesInformation

11 def contract searchCrimesContract{
12  isAppliedTo: searchCrimes;    
13  requires: addressFormat == ?? 
14     (onFailureDo: call(chooseCrimesProperties) );  
15    ensures : !(getCrimes(types).length > 5) ==> false &&
16              result[i].period < 1800
17       (onFailureDo: show(getCrimesInformation())); 
18 }

20 def contract showMapContract{
21  isAppliedTo: seeMap;    
22    ensures : serviceResponse.getTime() < 5000
23       (onFailureDo: skip); 
24 }
\end{lstlisting}
\label{fig:crimeInformationPEWS}  
\end{figure}

The result of the transformation of this model to the \textit{$\pi$-PEWS} model
is gives in the listing \ref{list:crimeInformationPEWS} and
\ref{list:crimeStatisticsPEWS}.

\begin{figure}[ht!]
\tiny
\centering
\begin{lstlisting}[label=list:crimeStatisticsPEWS,caption=pi-PEWS Specification: Crime
statistic and share information. ]

1  ns police =  "http:\\www.pm.rn.gov.br/service"
2  ns twitter =  "https://dev.twitter.com/docs/api/"

5  alias searchCrimes = portType/searchCrimes in police
7  alias seeCrimesInformation = portType/getCrimeDetail in police
4  alias seeStatistics = portType/getCrimeStatistics in police
6  alias shareInformation = portType/publishMessage in twitter

9  searchCrimes . ((seeCrimesInformation . shareInformation) |
10					seeStatistics)

12 def contract searchCrimesContract{
13  isAppliedTo: searchCrimes;    
14  requires: addressFormat == ?? 
15     (onFailureDo: call(chooseCrimesProperties) );  
16    ensures : !(getCrimes(types).length > 5) ==> false &&
17              result[i].period < 1800
18       (onFailureDo: show(getCrimesInformation())); 
19 }

21 def contract shareInformationContract{
22  isAppliedTo: shareInformation;    
23  requires: name == ?? AND password == ?? &&
24              crimes[i].Format == ?? 
25     (onFailureDo: call(shareInformation) );  
26 }

28 def contract seeStatisticsContract{
29  isAppliedTo: seeStatistics;    
30  requires: crimes[i].Format == ?? 
31     (onFailureDo: call(seeStatistics) );  
32 }
\end{lstlisting}
\label{fig:crimeStatisticsPEWS}  
\end{figure}

\section{Example 3: GesIMED Application}
\label{sec:siw_gesimed}

Aiming to perform a qualitative analysis and the validation of $\pi$SOD-M
methodology, we present the \textit{GesIMED Application}\footnote{If necessary, a detailed description of this case
 study can be found in \cite{valeriaThesis}.}
 case study.

% This case
% study was described in~\cite{valeriaThesis} with the goal of validating the
% SOD-M methodology. In
% this section we present the different produced by using the SOD-M and
% $\pi$SOD-M, as well as the contributions generated by of our proposal. We will
% model the same business case for both methods and see the difference between them.

This \correctingText{example} was originally developed in \cite{valeriaThesis} for the SOD-M
methodology. Here , we adapt the requirements of the system to include NFR. This
system is a management Web system that processes medical images through the Web \cite{valeriaThesis}. The objective is to
manage information about the creation and maintenance of scientific studies
for neuroscience research. The application is designed to be used primarily by
researchers in neuroscience, such as neurologists, neuropsychologists and
neuroradiologists who conducting research in this area.

The application requirements are the following: (i) a
database of medical images that can be accessed by the user; (ii) it has an
interface for querying the database; (iii)  implement standard  procedures for analyzing
and processing stored images; and (iv) the images analysis and processing
results must be also stored, so they can be used in future studies.

The  Medical Image Analysis Laboratory from the Universidad Rey Juan Carlos
(LAIM) offers three specific services to researchers in neuroscience. The access
to services has a cost and financial revenues are assigned to LAIM. The services
are:

\begin{itemize}
  \item Storage and retrieval medical imaging service (SACim);
  \item Image processing service (SPim);
  \item Image visualization service (SVim);
\end{itemize}

From the functions offered by these services we will detail the application
characteristics. The application uses these services offered by LAIM for
modeling the application.

% From this case study description we present the modeling performed in SOD-M, and
% soon after perform the modeling in $\pi$SOD-M. From the business case
% \textit{perform image processing}, we will differentiate the characteristics of
% each methodology.

% \subsection{$\pi$SOD-M Modeling}
% \label{sec:GesIMED-pisodm}
\bigskip
\bigskip

Given the description of the business requirements and the expected
functionality of the system, $\pi$SOD-M is used for developing this application.

% The modeling of \textit{GesIMED} application in $\pi$SOD-M has
% the same input considered in SOD-M method, the description of business
% requirements and system functionality. Thereafter, we follow the development of
% each model according to specification of the $\pi$SOD-M  methodology. Our goal
% in this section is to highlight the particularities of $\pi$SOD-M regarding
% SOD-M.

We will detail the \textit{perform image processing} business service. In $\pi$SOD-M, the \textit{$\pi$-UseCase} model
describes the application functions, restrictions and its non-functional
attributes. This business service (\textit{perform image processing}) is 
sufficiently rich to describe the peculiarities of $\pi$SOD-M
compared to the original model of SOD-M.

\subsection{\textit{$\pi$-UseCase} Model}

Figure \ref{fig:piUseCase-imageProcessing} shows the \textit{$\pi$-UseCase}
model. This model defines four constraints: \textit{authenticate, response time,
data format} and \textit{validate payment card}. These constraints are restrictions over the functions of the
\textit{GesIMED} application. In the description of each constraint it is
necessary to describe the constraint type (\textit{business} or \textit{value}).
The model has one business constraint and three value constraints. It is
possible to describe constraints for each use case application.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-piuc-01.pdf}
\caption{$\pi$-UseCase: Perform Image Processing.}
\label{fig:piUseCase-imageProcessing}
\end{figure}

% Notice that the other use cases are the same of those that were
% modeled in SOD-M. The difference is that they are now systematized in packages.
% Our approach guides the systematic functions through packages, because these
% packages represent business collaborators.

Besides the constraints, the model details the non-functional attributes
related to each constraint, for example, the \textit{response time} constraint.
For value constraints, it is necessary identify candidates variables to be
verified, through the use of \textit{@}. For example, the
authentication constraint, has the \textit{@id} and \textit{@password}
identifiers that must be verified.

% It is possible to identify a real difference between the use case models in
% SOD-M and $pi$SOD-M, especially with regard of constraints, in a way to describe
% details of quality requirements.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-pisp-01.pdf}
\caption{$\pi$-ServiceProcess: Perform Image Processing.}
\label{fig:piServiceProcess-imageProcessing}
\end{figure}

The main feature of our approach is the association of non-functional
requirements with use cases. For example, the user can identify that the
quality requirement related to a payment process is
\textit{transactional} or \textit{data privacy}. So, through the process, it is
possible to refine this information, considering other details provided.

The result of the \textit{$\pi$-UseCase} model is a detailed list of
constraints, its respective use case, system requirement information and
non-functional attributes required for each service.

\subsection{\textit{$\pi$-ServiceProcess} Model}

Figure \ref{fig:piServiceProcess-imageProcessing} shows the
\textit{$\pi$-ServiceProcess} model. This model transforms the use cases into actions,
and constraints into assertions. The resulting workflow describes pre or
post-conditions on workflow execution. The assertions are wrappers around the
action. All the assertions related to a specific action become a
contract. For example, the \textit{proceed payments service} action has a
pre- and a post-condition. The pre-condition is a restriction over the payment
information (\textit{card number, name} and \textit{value}), while the
post-condition defines the final state after the payment. The other assertions
of this model are the result of the transformation of each constraint described
in the \textit{$\pi$-UseCase} model.

%  Notice that the workflow (figure \ref{fig:piServiceProcess-imageProcessing})
%  is the same as described in SOD-M, with the difference of having assertions
%  during the process, so that the constraints are verified.

 \begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-pisc-01.pdf}
\caption{$\pi$-ServiceComposition: Perform Image Processing.}
\label{fig:piServiceComposition-imageProcessing}
\end{figure}

\subsection{\textit{$\pi$-ServiceComposition} Model}

The \textit{$\pi$-ServiceComposition} model presented in figure
\ref{fig:piServiceComposition-imageProcessing} refines the 
\textit{$\pi$-ServiceProcess} model. This model
describes the real services and their associated collaborators. The information
of the business collaborator comes from the package description. Contracts with
the same non-functional requirements become service policies. The policies
define rules, which are a direct transformation of the assertions in the
previous model.

% Another difference
% of our approach to the original approach of SOD-M, is that we do not treat each
% collaborator as a business process.

\begin{figure}
\tiny
\centering

\begin{lstlisting}[label=list:pewsNeuro,caption=pi-PEWS Specification: Perform
Image Processing. ]

1  ns app =  "http:\\www.neuro.laim.org"
2  ns LAIM =  "http:\\www.neuro.laim.org/laim-service.wsdl"
3  ns SACim = "http:\\www.neuro.laim.org/sacim-service.wsdl"
4  ns SPim =  "http:\\www.neuro.laim.org/spim-service.wsdl"

6  alias authenticate = portType/verifyUserandPassword in LAIM
7  alias proceedServicePayment = portType/payment in LAIM
8 alias provideImage = portType/provideImage in app
9  alias receiveImage = portType/receiveImage in SACim
10 alias processData = portType/processData in SPim
11 alias verifyImageFormatAndResult = portType/verifyImageFormatAndResult in
12 SACim alias getResults = portType/verifyExecutionTipe in LAIM
13 alias downloadResults = portType/downloadFiles in SACim

15 service performProcessing = 
16            receiveImage . processData . verifyImageFormatAndResult

18 authenticate . proceedServicePayment . provideImage . 
19   performProcessing . [act(getResults).time > 5]getResults .
20   {downloadResults}

21 def contract authenticateContract{
22  isAppliedTo: authenticate;    
23  requires: id == ?? && password = ??;
24     (onFailureDo: call(authenticate) )  
25 }

27 def contract paymentContract{
28  isAppliedTo: proceedServicePayment;    
29  requires: card_name =?? && 
30  			user_name == ?? && 
31  			value == ??;  
32  ensures : isPaid == true ;
33 }

35 def contract performProcessingContract{
36  isAppliedTo: performProcessing;    
37  requires: imageExtension == '.png' || imageExtension == '.pdf'
38    (onFailureDo: skip );    
39  ensures : resultValue <= 2 ;
40 }
\end{lstlisting}
\label{fig:pewsNeuro}  
\end{figure}

Listing \ref{list:pewsNeuro} shows the \textit{$\pi$-PEWS} specification that
was generated from the \textit{$\pi$-ServiceComposition} model. This
specification considers the operations (lines 6-16), its business collaborators
(namespaces in the lines 1-4), the main workflow (lines 18-20) and services
restrictions (described as contracts in the lines 21-40). The final result of
our approach is the specification that describes the structure and sequence of
execution to be performed by an orchestrator. \textit{$\pi$-PEWS}
specification defines a set of contracts to represent the application restrictions. Thus,
using $\pi$SOD-M it is possible to refine the models until the code generation.

\section{Lessons Learned}
\label{sec:leassonsLearned}

This section summarizes the experience we obtained after the use of $\pi$SOD-M
for the three \correctingText{examples}. \newText{In this section we also present some
results collected from the evaluation made ​​between the two approaches.}

\subsection{Example 1: To Publish Music}

In this case study we can see that the restrictions on web services can be
modeled from the early stages of development, not leaving it to the programming
language, or to the programmer to resolve this type of problem. More
specifically, this case study had restrictions \textit{performance} and \textit{security}. Since
the \textit{$\pi$-UseCase} model to the \textit{$\pi$-ServiceComposition} model,
non-functional requirements and their attributes might be described without
serious impact on the modeling of other application requirements.

The restrictions on the functions provided by
the \textit{Spotify} service was basically \textit{performance}. This type of
restriction is not easy to represent in the implementation, because it relies on
the network available to the user. We described the
constraints and refined them into a \textit{ performance} policy for updating
the music. The security restrictions have been described for the service that provides banking information. The security policy generated was described
for two functions of the Bank service, \textit{pay} and \textit{send
confirmation} functions. This type of specification can be refined incrementally
by applying the modelling steps defined by the $\pi$SOD-M methodology.

\correctingText{It} does not cover all the possible restrictions in
an application but it shows that it is possible to guarantee
quality requirements refining information at each iteration of the modeling,
such as constraints of \textit{performance} and \textit{security}.

The difficulty in modeling this \correctingText{example} was the definition of policies for
each type of non-functional requirement. For example, the \textit{httpAuthPolicy} policy for
the authentication on Facebook service has restrictions from
both types, \textit{performance} and \textit{security}.

\subsection{Example 2: Crime Map}

The most important process when designing a system is the requirement analysis
process, especially identifying the non-functional requirements and the system
restrictions. In order to ease the coding process and let the developer focus on
technical issues, the system architecture must be automatically generated, from
the system design.

Using $\pi$SOD-M has enabled the explicit specification of functional and
non-functional requirements and their refinement along the different phases of
the development process.

% its structure
% may be automatically generated from the system design. If we carry out with some technical problem while coding, you can refer the document for the
% programming language you are using and solve it quickly, however problems with
% architecture or requirements analysis does not take longer to resolve. So, a
% good methodology is more important than is good language, and we could see this
% on the validation of $\pi$SOD-M.

In this \correctingText{example} we can see the restrictions on four web services:
\textit{Police, Google Maps, Twitter} and \textit{Post} service. As these
services are independent, the constraints of the application being developed
must comply with the service APIs. If a restriction is inconsistent with a
function, there is no guarantee that it will be respected. For example, if the police offers crimes information on a
particular region by type or day, it is not possible to present the restrictions
of a particular time of day, for example, crimes that happen in the afternoon.

This \correctingText{example} has two \textit{reliability} policies and one \textit{security}
policy. Policies on data are common, because they are restrictions on
the service interface. We notice in the development of this \correctingText{example} that
function restrictions, especially the execution time and time
response are more difficult to ensure.

\subsection{Example 3: GesIMED}

The $\pi$SOD-M methodology helps to define a detailed description of the
application, its restrictions and non-functional requirements associated with each service.
The refinement of the quality requirements of each level helps to improve the
development detail in order to produce a more reliable application. The NFRs are
not only described in a general way, but the restrictions are designed to each
service individually.

Considering the modeling of this \correctingText{example} in SOD-M, the first step of the is
to list the global requirements (business case). Figure \ref{fig:sodm_usecase}
shows the use case model for the application. This model is very simple and describes the main application requirements. The actor is represented as the ``neuroscience
researcher'', which is the final consumer of the services to be implemented.
Furthermore, the model is represented as use cases, which describes the main
system business services: ``perform image processing\footnote{During our
analysis, we will detail the \textit{perform image processing} requirement, as a
way to better understand the particularities of both methods.}'', ``perform
image view'' and ``perform query''.

\begin{figure}[ht!]
\centering
\includegraphics[width=.6\textwidth]{chapters/validation/figs/neuro-useCase.pdf}
\caption{Use Case Model~\cite{valeriaThesis}.}
\label{fig:sodm_usecase}
\end{figure}

The next model is the the extended use cases model. This model is used to model
the functionality required by the system to perform each business services.
Business services described in the previous model, become more detailed in the
extended use case model. In this model are described both the application's
features such as the possible services to be performed. Figure
\ref{fig:sodm_extendedusecase} presents the extended use case model for
\textit{perform image processing} business service. This model is detailed in 6
use cases that are necessary for the image processing. The use cases are: (i)
perform image processing; (ii) proceed with the service payment; (iii) provide
images; (iv) authenticate; (v) get results; and (vi) download results.

\begin{figure}[ht!] 
\centering
\includegraphics[width=.89\textwidth]{chapters/validation/figs/neuro-extendedUseCase.pdf}
\caption{Extended Use Case Model~\cite{valeriaThesis}.}
\label{fig:sodm_extendedusecase}
\end{figure}

Once identified the different functionalities that are required by the
system to perform business services, the service process model is used to
represent the workflow necessary to perform a business service. SOD-M represents
each business service in different diagrams, which are shown in the following
figures. Figure \ref{fig:sp-imageProcessing}, for the business service ``
\textit{perform image processing}''. The activities represented in these figures
are derived from the basic use cases represented in the model extended use case.

\begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sp01.pdf}
\caption{Image Processing - Service Process Diagram~\cite{valeriaThesis}.}
\label{fig:sp-imageProcessing}
\end{figure}

% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-03.pdf}
% \caption{Image Visualization - Service Process Diagram~\cite{valeriaThesis}.}
% \label{fig:sp-imageVisualization}
% \end{figure}
% 
% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-02.pdf}
% \caption{Perform Query - Service Process Diagram~\cite{valeriaThesis}.}
% \label{fig:sp-query}
% \end{figure}

The service composition models represent the processes of composition of the
various actions that are necessary to perform each business service, detailing
which business collaborator performing each action.

 \begin{figure}[ht!]
\centering
\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sc-01.pdf}
\caption{Image Processing - Service Composition Diagram~\cite{valeriaThesis}.}
\label{fig:sc-imageProcessing}
\end{figure}

% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sc-02.pdf}
% \caption{Image Visualization - Service Composition
% Diagram~\cite{valeriaThesis}.}
% \label{fig:sc_example02}
% \end{figure}
% 
% \begin{figure}[ht!]
% \centering
% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sc-03.pdf}
% \caption{Perform Query - Service Composition Diagra~\cite{valeriaThesis}.}
% \label{fig:sc_example03}  
% \end{figure}

Figure \ref{fig:sc-imageProcessing} present the details of the services composition
for the business service we are modelling. It is important to realize that the
service processes described previously, are refined by the execution of these service.
Notice that SOD-M creates a workflow for each business service, independently.
This workflow is a sequence of services that can be run
considering the system features. Figure
\ref{fig:sc-imageProcessing} presents that the workflow
uses the services offered by LAIM to perform each function, according the SACim, SPim and SVim
services. It is also important to note that the external
application services are marked with the stereotype \textit{<<WS>>}.

The result of applying the SOD-M method is a detailed service composition model
from the business use case, in our specific case the ``\textit{perform image
processing}'' business service.

 %\bigskip

% Notice that the non-functional properties is not modelled by SOD-M, in contract
% to $\pi$SOD-M models.

\section{Measurement and evaluation of the $\pi$SOD-M methodology}

Based on \cite{basili:1985}, the measurement and evaluation process requires a
mechanism for determining what data is to be collected; why it is to be
collected; and how the collected data is to be interpreted. The purpose of the measurement and
evaluation flows from the needs of the organization; the need to better
understand resource utilization in order to improve cost estimation; or the need
to evaluate the benefits, advantages and drawbacks of a research project.

The data collection process itself requires a basic paradigm that traces the
goals of the collected process, \textit{i.e.} the reasons the data are being
collected. \cite{basili:1985,Kontio:1996} describe that the need for information
must be quantified whenever possible and the quantification analyzed as to whether or not it
satisfies the needs. This quantification of the goals should then be mapped into
a set of data that can be collected.

Each evaluation \textit{goal} generates a set of questions that attempt to
define and quantify the specific goal which is at the root of its goal tree. The
\textit{goal} is only as well defined as the \textit{question} that it
generates. Each \textit{question} generates a set of \textit{metrics} or
\textit{distributions of data}.

The paradigm proposed by \cite{basili:1985} defines a set of six steps to the
data collection:

\begin{enumerate}
  \item Generate a set of goals;
  \item Derive a set of questions of interest or hypotheses which quantify those
  goals;
  \item Develop a set of data metrics and distributions which provide the
  information needed to answer the question of interest;
  \item Define a mechanism for collecting the data;
  \item Perform a evaluation of the data; 
  \item Analyze the data collected to answer the question posed.
\end{enumerate}

\bigskip

Aiming to collect data that help in an initial analysis of our methodology
through three examples described in this chapter, we will use the approach
proposed by Basili \cite{basili:1985} to extract these data and analyze if the goals of
our proposal have been achieved and how it is possible improve the $\pi$SOD-M
methodology. 
% 
% Com o objetivo de coletar dados que auxiliem em uma análise inicial da
% metodologia proposta através dos 3 exemplos descritos neste capítulo, iremos
% utilizar a abordagem proposta por Basili para extrair esses dados e analisar se
% os objetivos da nossa proposta foram alcançados e de que forma é possível
% melhorar a metodologia $\pi$SOD-M.

\paragraph{Generate a set of goals:} First we present the goals we have, when we
intend to measurement and evaluate the $\pi$SOD-M methodology. The goals are:

% Primeiramente apresentamos os objetivos que almejamos ao realizar essa medição e
% avaliação da metodologia. Os objetivos são.
% 
% Objetivos:

\begin{enumerate}
  \item  Verify if the identification of non-functional requirements at initial
  levels of development improves the modeling of reliable web services.

%   Verificar se a identificação de requisitos não-funcionais nos níveis iniciais de desenvolvimento melhora a modelagem de serviços web confiáveis.

  \item  Verify how to refine application constraints into specification code,
  and if it improves the software quality.

%   Verificar se é possível refinar restrições de qualidade em código de especificação.

  \item  Verify if it is possible to reuse code generated for requirements
non-functional.

% Verificar se é possível a reutilização de código gerado para requisitos
% não-funcionais.

  \item  Verify how the non-functional requirements are specified in the
  methodology.

\end{enumerate}

\paragraph{Set of questions of interest which quantify the goals:} Following the
approach proposed in Basili \cite{basili:1985}, the questions are related to each goal. Table \ref{tab7:questions} present the questions related to the goals of the $\pi$SOD-M evaluation analysis.

\begin{table}
\centering
\caption{$\pi$SOD-M's Goals and Questions}
\label{tab7:questions}

\begin{tabular}{|l|l|}

\hline 
\hline \textbf{Goals} & \textbf{Questions} \\

\hline 1. Verify if the identification of non-functional & 1. What are the
levels of modeling  \\

requirements  at initial levels of development   & presented in the
methodology?  \\

 improves the modeling of reliable web services. &    And how the NFRs are
 modelled?  \\

 &  \\

 &  2. What is the goal of each modeling level?\\

&  \\

 &  3. Which kind of artifact is generated at\\
 & each level of development?\\
\hline
2. Verify how to refine application constraints & 1. What kind of modelling
approach\\

 into specification code, and if it improves &  is used in the methodology? \\

the software quality. & \\

& 2. How are modeled the non-functional\\
& requirements?\\

\hline

3. Verify if it is possible to reuse code  &1. Which level of development the code \\
generated for requirements non-functional. &  is generated?  and what kind of\\ 
 &code is generated?\\
&\\
& 3. The code is generated automatically \\
& or manual?\\

\hline

4. Verify how the non-functional requirements & 1. How are classified the
non-functional \\ 
are specified in the  methodology. & requirements?  \\

\hline
\hline

\hline

\end{tabular}

\end{table}
% 
% Seguindo a abordagem proposta em Basile, as perguntas referentes a cada objetivo
% são.

% OBJ1 - QUESTIONS
% 1 - Quais os níveis de modelagem que a metodologia piSOD-M apresenta? E de que
% forma os NFR são representados?
% 2 - Qual o objetivo de cada nível de modelagem
% da metodologia piSOD-M? 
% 3 - Que tipo de artefato é gerado em cada nível de desenvolvimento?
% 
% OBJ2 - QUESTIONS
% 1 - Qual abordagem de desenvolvimento é utilizada na metodologia?
% 2 - Como são modelados os requisitos não-funcionais?
% 
% OBJ3 - QUESTIONS
% 1 - Em qual nível de desenvolvimento o código é gerado? E que tipo de código é
% gerado? 
% 2 - O código é gerado de maneira automática ou manual?
% 
% OBJ4 - QUESTIONS
% 1 - Como são classificados os requisitos não-funcionais?
% 2 - Existe algum tipo de classificação dos RNF?

% Ainda seguindo a abordagem proposta por Basili, identificamos métricas com o
% objetivo de quantificar as perguntas, de forma a gerar dados que possam ser
% analisado e avaliados com o objetivo de verificar se os objetivos da avaliação
% foi satisfeito. As métricas são por listadas por objetivos/questões (por
% exemplo, #1.2. Esta métrica está relacionada ao objetivo 1, e a questão 2
% identificados anteriormente), dessa forma as medidas a serem quantificadas são
% apresentadas na tabela X.

\paragraph{Set of data metrics and distributions:} Still following the approach
proposed by Basili \cite{basili:1985}, we identify metrics with the objective of quantify the questions in order to generate data that can be
analyzed and evaluated in order to verify that the goals of the evaluation
were satisfied, or not. The metrics are listed by goal/question (for
example \#1.2. This metric is related to \textit{goal} 1, and \textit{question}
2 previously identified). The statements for measure the data are
shown in table \ref{tab8:metrics}.

\begin{table}
\centering
\caption{$\pi$SOD-M's Metrics/Distributions}
\label{tab8:metrics}

\begin{tabular}{|c|l|}

\hline 
\hline \textbf{Goal/Question} & \textbf{Metrics/Distributions} \\

\hline \#1.1 & 1. How many levels the methodology defines?  \\
& 2. How many models are generated at each level?\\
\hline \#1.2 & 1. How many restrictions were modeled (by level)?  \\
\hline \#1.3 & 1. How many artifacts were generated (by level)?  \\
\hline \#2.1 & 1. How long (in days) was spent for the modeling   \\
&  of each case study? \\
\hline \#2.2 & 1. How many RNFs were modeled using the approach?  \\
\hline \#3.1 & 1. How many lines of code were generated for \\
& specify the non-functional requirement identified? \\

\hline \#3.2 & 1. What is the total of line of code generated? \\

\hline \#4.1 &  1. How many types of non-functional requirements \\
& were identified?\\

 &  2. Which non-functional requirements \\
& were identified?\\

\hline
\hline

\hline

\end{tabular}

\end{table}

% Metricas
% 
% 1.1.1 Quantos níveis a metodologia define?
% 1.1.2 Quantos modelos são gerados em cada nível?
% 
% 1.2.1 Quantas restrições são modelada?
% 
% 1.3.1 Quantos artefatos podem ser gerados?
% 
% 2.1.1 Qual o tempo médio para o desenvolvimento de cada estudo de caso?
% 2.2.1 Quantos RNF foram modelados?
% 
% 3.1.1 Quantas línhas de código foram geradas para cada requisito não funcional
% identificado?
% 3.2.1 Qual a soma total de linha de código geradas?
% 
% 4.1.1 e 4.1.2 Quantos tipos de requisitos não-funcionais foram identificados?
%

\paragraph{Collecting the data:} After the development of metrics and the types
of data to be collected, the extraction was performed considering the modeling
steps of $\pi$SOD-M for each case study presented in this chapter. Table
\ref{tab9:data} presents the results of the data collection.

\paragraph{Evaluation of the data:} Tables \ref{tab8:metrics} and
\ref{tab9:data} present the questions and data generated from the evaluation of
each example. All examples were modeled in accordance with the proposed
methodology and yielded the following results.

% As tabelas 9 e 10 apresentam as perguntas e os dados gerados a partir da
% avaliação de cada exemplo. Todos os exemplos foram modelados de acordo com a
% metodologia e gerou os seguintes resultados.

\begin{table} 
\centering
\caption{Results: $\pi$SOD-M's Metrics/Distributions}
\label{tab9:data}

\begin{tabular}{|c|c|c|c|}
\hline 
\cline{2-4}
& \multicolumn{3}{ |c| }{$\pi$SOD-M} \\
\hline  
Metrics/Distributions & Example 1 & Example 2 & Example 3 
\\
\hline 
\hline \#1.1.1 & 4 & 4 & 4 \\
\hline \#1.1.2 & 1-1-1-1 & 2-2-2-2 & 3-1-1-1 \\
\hline \#1.2.1 & 5c-8a-6r-2co & 7c-9a-7r-5co & 4c-6a-6r-3co\\
\hline \#1.3.1 & 1 & 1 & 1 \\
\hline \#2.1.1 & 8 & 10 & 5\\
\hline \#2.2.1 & 2& 3& 3\\
\hline \#3.1.1 &9 & 37 & 20 \\
\hline \#3.2.1 & 29 & 56& 40\\
\hline \#4.1.1 & 2& 3& 3\\
%\hline \#4.1.2 & & & \\

\hline

\end{tabular}
\end{table}

The methodology defines 4 levels of modeling (\textit{\#1.1.1}), however the
case studies generated a different number of models ($\pi$-UseCase,
$\pi$-ServiceProcess, $\pi$-ServiceComposition e $\pi$-PEWS). In \textit{Example
1} was generated 1 model of each type, in \textit{Example 2} were generated two
models of each type, and in \textit{Example 3} were generated 3 $\pi$-UseCase
models and 1 referring to the others (\textit{\#1.1.2}). We found that this
variation has no involvement with the identification of the number of non-functional
requirements.

% A metodologia define 4 níveis de modelagem, no entanto os estudos de caso
% geraram um número diferente na quantidade de modelos ($\pi$-UseCase,
% $\pi$-ServiceProcess, $\pi$-ServiceComposition e $\pi$-PEWS). No exemplo 1 foram
% gerados 1 modelo de cada, No exemplo 2 foram gerados 2 modelos de cada tipo, e
% por fim, o exemplo 3 gerou 3 modelo de caso de uso e 1 referente aos demais.
% verificamos que essa variação não tem nenhuma implicação com a identificação dos
% requisitos.

Regarding the types of restrictions modeled at each level, we find that
the number constraints identified implies an increase in the number of
lines of code. The  \textit{Example 1} set five \textit{constraints} in the
$\pi$-UseCase model, 8 \textit{assertions} in the $\pi$-ServiceProcess model, 6
\textit{rules} in the $\pi$-ServiceComposition model and 2 \textit{contracts} in
the $\pi$-PEWS model (\textit{5c-8a-6r-2Co} \footnote{5c =
5 constraints, assertions 8th = 8; 6r = 6 rules; 2Co contracts = 2}). This
example generated 29 lines of code specification. However, the \textit{Example
2} presents the following data \textit{7c-9a-7r-5co}, generating 56 lines of
specification code. It was observed that the number of restriction increases the
number of lines specification. The \textit{Example 3} presents the following
data, \textit{4c-6a-6r-3co}, generating 40 code lines specification
(\textit{\#1.2.1}, \textit{\#3.1.1} and \textit{\#3.2.1}).

% Em relação aos tipos de resquições modeladas em cada nível, verificamos que
% quanto mais restrições são identificadas implica em um aumento no número de
% lineas de código. O exemplo defini 5 constraints no modelo de $\pi$-UseCase, 8
% assertions no modelo de $\pi$-ServiceProcess, 6 regras no modelo de
% $\pi$-ServiceComposition e 2 contratos no modelo de $\pi$-PEWS
% (\textit{5c-8a-6r-2co}\footnote{5c = 5 constraints; 8a = 8 assertions; 6r = 6
% regras; 2co = 2 contratos}). Esse exemplo tem 29 linhas de código de
% especificação. Enquanto o exemplo 2 apresenta o seguinte dado, 7c-9a-7r-5co,
% gerando 56 linhas de código de especificação. Percebe-se que o número de
% restrições aumenta o número de linhas de especificação. O exemplo 3 tem as
% seguintes resquições, 4c-6a-6r-3co, gerando 40 linhas de código de
% especificação.

According to the data analysis, we also found that the number of lines of code
for the definition of contracts increase according to the numbers of rules
identified (\textit{\#1.2.1}).

The number of days spent for modeling also varied
according to the number of constraints. Thus, the time spent with
identification and modeling of the constraints implies a greater number of days
of development. \textit{Example 2} was modeled in 10 days (\textit{\#2.1.1}) and
this example showed a greater number of restrictions and non-functional
requirements. We assume that if these restrictions were not modeled since the initial stages, the time spent on
maintenance could be higher. \textit{Example 3} had the fewest
number of restrictions, and therefore took less time to be modeled. Thus the
number of non-functional requirements of an application implies the time spent
for the project development using $\pi$SOD-M (\textit{\#2.2.1} and
\textit{\#4.1.1}).

% Segundo a análise, verificamos também que o número de linha de especificação em
% apenas para a definiçào dos contratos, cresce de acordo com o números de regras
% identificadas. A quantidade de dias gastos para a modelagem também variou de
% acordo com o número de restrições da aplicação. Dessa forma, o tempo gasto com a
% identificação e modelagem das restrições do exemplo teve uma implicação direta
% com o número de dias gastos. O exemplo 2 foi modelago em 10 dias e este
% apresentou um número maior de restrições, e requisitos não funcionais. Podemos
% supor que se essas restrições não fossem modeladas desde as etapas iniciais, o
% tempo gasto com manutenção poderia ser maior. O exemplo 3 teve o menor número de
% restrições, e consequentemente levou menos tempo para ser modelado. Dessa forma
% o número de requisitos não-funcionais de uma aplicação implica no tempo gasto
% no desenvolvimento do projeto.

\paragraph{Conclusions:}

We conclude that the identification of NFRs in initial levels of modelling helps
the development of services, mainly because the restrictions can be
refined and transformed into code specification. This favors the
prior identification of errors, and help in the classification of the
non-functional requirements.

Classify restrictions as constraints, then refine this concept into assertions,
and then group it into policy improves identifying types of non-functional
requirements, providing reusability of services and better use of the code
specification.

We also found that the use of the method provides identification of different
types of NFRs. These requirements are translated into contracts and can be
checked at runtime. We can see that the number of lines of code generated
for each type of non-functional requirement does not have a regular variation,
and it does not generates apparent side effects.

\subsection{Comparison of the SOD-M and $\pi$SOD-M approaches}

% A tabela X apresenta dados em relacao aos 3 exemplos utilizados na avaliacao. O
% resultado reflete as diferencaos entre as duas abordagem com a intencao de
% verificar como a extensao de SOD-M pode melhorar o processo de modelagem de
% aplicacoes orientadas a servicos na presenta de requisitos nao funcionais. 
\newText{
We also perform a comparative analysis between $\pi$SOD-M and SOD-M,
considering other aspects. Table \ref{tab6:comparison} presents the data related with the
three examples used for the $\pi$SOD-M evaluation: (i) To Publish Music example (\textit{Ex. 1}),  (ii) Crime Map example (\textit{Ex. 2}), and  (iii) GesIMED
example (\textit{Ex. 3}). The result reflects the difference between the two
approaches (SOD-M and $\pi$SOD-M) with the intention of verifying how $\pi$SOD-M
can improve the modeling process of service-oriented applications in the presence of
non-functional requirements. }

% Os aspectos analisados foram: (i) a quantidade de use cases necessarios para a
% modelagem de mais alto nivel tanto dos requisitos como das restricoes da
% aplicacao; (ii) quantas actions sao necessarias para a representacao do processo
% de negocio; (iii) quando business collaborators sao necessarios para a
% execucao completa do workflow; (iv) quantos requisitos nao funcionais foram
% identificados durante a modelagem da aplicacao; (v) quantas linhas de codigo
% podem ser geradas a partir do modelo de composicao de servicos; e (vi) quantas
% transformacoes de modelos sao realizadas durante todo o processo de
% desenvolvimento.
\newText{
The aspects analyzed and presented in table \ref{tab6:comparison} are: (i) the
number of use cases required for modeling both, the
restrictions and requirements of the application, (ii) how many actions are required for the
representation of the business process, (iii) how many business collaborators
are needed to the complete execution of the workflow, (iv) the number of
non-functional requirements that were identified during the modeling of the
application, (v) how many lines of code were generated from the service
composition model of each example, and (vi) how many model transformations held
throughout the development process.
}

% Considerando o numero de casos de uso necessarios para a modelagem dos exemplos
% apresentados neste capitulo, os exemplo modelados com piSOD-M apresentam mais
% casos de uso pois e possivel modelar as restricoes da aplicacao atraves de casos
% de uso esteriotipados com o esteriotipo costraint. Devido a este motivo o
% exemplo 1 apresenta 9 casos de uso usando SOD-M, enquanto que utilizando a
% abordagem piSOD-M se tem 13 casos de uso. A diferenca (4 casos de uso)
% representa a quantidade de restricoes que foi identificada neste nivel atraves
% do modelo de pi-UseCase. No exemplo 2 temos 8 casos de uso utilizando SOD-M
% enquanto temos 13 utilizando piSOD-M. Uma diferenca de 5 casos de uso. E por fim
% o exemplo 3 temos 5 casos de uso utilizando SOD-M, e 9 casos de uso atraves da
% modelagem com piSOD-M.  piSOD-M apresenta uma quantidade maior de casos de uso
% pelo fato deste modelo tambem representar as restricoes da aplicacao, mesmo no
% nivel mais abstrado. Verificamos que apesar, de um esforco relativamente maior
% de modelagem, a representatividade do modelo no que diz respeito aos requisitos
% funcionais e nao-funcionais e maior em piSOD-M que em SOD-M.

\newText{
Considering the number of use cases for modeling the examples presented in this
chapter, the models from the $\pi$SOD-M approach present more use cases then
the models in SOD-M, because using $\pi$SOD-M it is possible to model the
restrictions of the application through stereotyped use cases
(\textit{constraint}). Due to this reason, example 1 has 9
use cases using SOD-M, while using $\pi$SOD-M, the model has 13 use cases. The difference (4 use cases) between
both approaches represents the amount of restrictions that have been identified
at this level through the $\pi$-UseCase model. In example 2, it has 8 use cases
using SOD-M while using $\pi$SOD-M it has 13 use cases. A difference of 5 use
cases. In the third example, it presents 5 use cases using SOD-M, and 9 cases
through $\pi$SOD-M. $\pi$SOD-M presents a greater number of use cases because
this model also represent the constraints of the application, even in the most
abstract level. We found that in spite of a relatively greater effort modeling,
$\pi$SOD-M describes functional and non-functional aspects, while SOD-M models
only functional aspects.
}

\begin{table}
\centering
\caption{Comparison: SOD-M and $\pi$SOD-M}
\label{tab6:comparison}

\begin{tabular}{r|c|c|c||c|c|c|}

\cline{2-7}
& \multicolumn{3}{ |c|| }{SOD-M} & \multicolumn{3}{ |c| }{$\pi$SOD-M} \\

\cline{2-7} & Ex. 1 & Ex. 2 & Ex. 3 & Ex.
1 & Ex. 2 & Ex. 3
\\
\hline 
\hline Use Case &9 &8 & 5& 13 & 14 &9\\
\hline Action &7 & 7 &5 &15 &12 &11\\
\hline Business  & 4&3 &3 &4 &3 &3\\
       Collaborator & & & & & &\\
\hline NFRs & - & - & - &2 &3 &3\\
	   modelled &  &  &  & & &\\
\hline Lines & 20 & 19 &20 & 29 & 56&40\\
	   of Code & & & & & &\\
\hline Transformations  & 2 & 2 & 2 & 3 & 3 & 3\\

\hline

\end{tabular}

\end{table}

% Em relacao ao numero de actions, a nossa abordagem apresenta um numero maior
% pois e necessário descrever pre- e pos-condições para a execuçao do workflow. O
% numero maior de actions em piSOD-M favorece que a execução do workflow obedeça a
% restrições do sistema. Estas actions são resultado das trasnformações dos casos
% de uso. Os casos de uso marcados com esteriotipos constraint, sao transformados
% em actions com esteriotipos assertion. É importante perceber na tabela, que no
% primeiro exemplo, o numero de action varia de 7 para 15, no segundo de 7 para 12
% e no terceiro exemplo varia de 5 para 11. A verificação de restrições atravéd de
% assertions favorece a representacao de requisitos nao-funcionais necessários
% para a execução do workflow.

\newText{
In relation to the number of actions, the $\pi$SOD-M approach provides a larger
number, because it needs to describe pre- and post-conditions for the
services. The larger number of actions in $\pi$SOD-M favors the verification of
restrictions during the execution of the workflow. These
actions are the result of a model transformation. The use cases marked
with \textit{constraint} stereotype are transformed into actions with
the \textit{assertion} stereotype. It is important to realize in the table that
in the first example, the number of action ranges from 7 to 15,  the second 7
to 12 and in the third example ranges from 5 to 11. The constraint checking
through assertions favors the representation of non-functional requirements
necessary for the workflow execution.
}

% Considerando o número de Business Collaborator, esse valor nao sofre nenhuma
% variação entre as abordagens, pelo fado dos serviços a serem executados nos 3
% exemplos e nas 2 abordagens serem os mesmos. Apesar do número de casos de uso e
% actions variarem isso não implica na variação da quantidade de serviços externos
% (business collaborator) necessários para a execução da aplicação.

\newText{Considering the number of \textit{Business Collaborator}, this value does not
suffer any variation between both approaches, because of the services to be
performed in the 3 examples in both approaches are the same. Although the number
of use cases and actions vary, it does not imply the variation of the amount of external
services (\textit{Business Collaborator}) required for execute the application.  
}

% Como o objetivo de ambas as abordagens são destintos, verificar a quantidade de
% requisitos não-funcionais não é comparável, no entanto aponta para uma melhoria
% e evolução da abordagem SOD-M. Usando SOD-M não é possível identificar nenhum
% requisito não funcional durante a modelagem. No entanto em piSOD-M é possível
% modelar distintos requisitos não-funcionais e refina-los durante a modelagem,
% fazendo com que em um nível mais baixo sejam representados como propriedades não
% funcionais \cite{MylopoulosBook99}. Apesar de não ser comparável pois ambas as
% abordagens tem focos de desenvolvimento diferentes, a proposta apresentada em
% piSOD-M abraga um valor razoável ao considerar a modelagem de requisitos não
% funcionais para aplicaçoes orientadas a serviços.

\newText{
Since the goal of SOD-M and $\pi$SOD-M approaches are different, analyze the number of
non-functional requirements modelled is not
comparable, however points to an evolution and improvement of the SOD-M approach. Using SOD-M, it is not possible
to identify any non-functional requirement during the modeling. However, using
$\pi$SOD-M, it is possible to model different non-functional requirements and
refine them. From this point of view, both approaches are not comparable because
they have different focus of development, otherwise the proposal presented in
$\pi$SOD-M adds reasonable value considering the modeling of non-functional
requirements for service-oriented applications.
}

% Considerando a especificação das aplicações descritas em cada na lingagem
% piPEWS, fazemos uma análise da diferença do número de linhas geradas em relação
% as ambas abordagens. Como a modelagem em SOD-M considera apenas a especificação
% dos requisitos funcionais, e piSOD-M consideram ambos aspectos, funcionais e não
% funcionais, a análise compara a especificação piPEWS de cada exemplo,
% considerando a geração dos contratos (piSOD-M) e a não geração dos contratos
% (SOD-M) que representam a verificação de aspectios não-funcionais. O número de
% linha e código cresce em piSOD-M devido ao fato da espeficicação gerar contratos
% que definem pre e pos-condições sobre os serviços do workflow. O crescimento
% depende da quantidade de restrições descritas durante a modelagem dos exemplos.
% Não existe nenhum padrão com relação a diferença de número de linhas de código.
% Por exemplo,o exemplo 1 a diferença entre as abordagens é de 9 linhas de código,
% enquanto o exemplo 2 a diferença é de 37 linhas. No exemplo 3 a diferença é de
% 20 linhas de código, sendo 20 linhas geradas usando a abordagem SOD-M e 40
% usando a abordagem piSOD-M. É importante notar que o número de requisitos não
% funcionais modelados tem como consequencia um maior número de linhas de código,
% perceba-se os exemplos 2 e 3, pois ambos modelam 3 requisitos não-funcionais e
% isso implica em um número maior de linhas de na especificação em piPEWS.

\newText{
Regarding the specification code for each example, we analyzed the difference of
lines of code ($\pi$-PEWS language) generated in both approaches. SOD-M
only considers the specification of functional aspects, and $\pi$SOD-M consider both aspects, functional and non-functional,
thereby we compare the $\pi$-PEWS specification for each example, considering the
generation of the workflow (SOD-M)  and the generation of contracts
($\pi$SOD-M). The number of lines of specification grows in $\pi$SOD-M due to
the fact that this approach generates contracts which defines pre- and post-conditions
for the services. The growth depends on the number of constraints identified
during the modeling. There is no standard concerning the difference in terms of the number of lines
generated. For example, in the first example the difference between the
approaches is 9 lines of code, while in the second example the difference is 37
lines. In the example 3 the difference is 20 lines of code, where 20 lines were
generated using the SOD-M approach  and 40 lines using $\pi$SOD-M. We can
conclude is that the number of non-functional requirements modeled implies a greater number of lines of specification code, to realize the examples
2 and 3, both has three non-functional requirements and it implied a larger
number of lines in $\pi$-PEWS specification, 56 e 40 respectively.
}

% Outro fator a ser analisado é o número de transformações em cada abordagem. Na
% modelagem de todos os exemplos em SOD-M foram realisadas 2 transformações no
% nível PIM, enquanto que na modelagem dos exemplos usando piSOD-M foram
% realizadas 3 trasnformações, 2 no nível PIM e 1 no nível PSM. O número maior de
% transformações implica em um maior refinamento dos conceitos, e consequentemente
% na geração de código para uma plataforma específica.

\newText{
Another factor to be considered is the number of transformations in each
approach. In the modeling of all the examples in SOD-M were performed two 
transformations at PIM level, whereas in the modeling of the examples using
$\pi$SOD-M were performed three transformation, 2 in the PIM level and 1 on PSM
level. The larger number of transformations in $\pi$SOD-M represents a step of
refinement and the generation of code for a specific platform.      
}

% 
% Apesar dos números apresentados na tabela X mostrarem que existe um esforço
% maior no uso da metodologia piSOD-M, isso se deve ao fato de que restrições
% sobre requisitos e serviços saão feitas desde os níveis mais abstratos de
% modelagem. Isso favorece a identificação prévia e revinamento adequado de
% aspectos transversais e restrições da aplicação. A metodologia piSOD-M agrega
% valor e conceitos novos em relação a SOD-M melhorando e facilitando a descrição
% de requisitos de qualidade das aplicações orientadas a serviços.

\newText{
Despite the numbers presented in table \ref{tab6:comparison} show that there is a
larger effort in the use of the $\pi$SOD-M methodology, compared with SOD-M,
this is due to the fact that non-functional requirements and restrictions on
services are made from the most abstract levels of modeling. This favors the
early identification of system restrictions and improves the refinement of
transversal aspects and constraints of the application. The 
$\pi$SOD-M methodology adds reasonable value and new concepts regarding SOD-M,
improving and facilitating the description of quality requirements for
service-oriented applications. }

\section{Conclusions}
\label{sec:conclusion_validation}

This chapter presented the use of the $\pi$SOD-M methodology, as a form of
empirical validation. We described, modeled and implemented three
\correctingText{example of application development} to use concepts, model
properties and detail the process development. The methodology and its concepts supported the development of applications that use service
compositions, mainly in the identification and refinement of quality requirements. The use of $\pi$SOD-M
and its environment can help to identify the non-functional requirements and
improve the system modeling and specification process.

We conducted, in this chapter, two types of evaluations, the first followed the
proposal based on Basili~\cite{basili:1985} for data evaluation. And the second
was a comparison between $\pi$SOD-M and SOD-M, considering different aspects.

\newText{The development of applications by using SOD-M favors and improves
the modelling of services in more abstract levels of development and help to
identify non-functional requirements related with each of these services.}

The transformations of models enables the semi-automatic
development of applications. Owing to the generation of the
system specification code from the service composition model, the designer and
developer can focus on the description and detail of functional and non-functional requirements. The
generated code contains the expression of the interaction with services and
their associated namespaces. This specification can be executed after the model
transformation.

As the process is iterative and incremental, after the generation of each model
it is possible to make adjustments to the generated model. For example, after
the $\pi$-UseCase model design and the generation of $\pi$-ServiceProcess 
model, the designer can make adjustments and improve the model so that it
becomes more expressive.

### 2013-09-30 WESOA2013/source/instructions.tex
inicial --martin

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}
% 
\begin{document}
\markboth{\LaTeXe{} Class for Lecture Notes in Computer
Science}{\LaTeXe{} Class for Lecture Notes in Computer Science}
\thispagestyle{empty}
\begin{flushleft}
\LARGE\bfseries Instructions for Authors\\
Coding with \LaTeX\\[2cm]
\end{flushleft}
\rule{\textwidth}{1pt}
\vspace{2pt}
\begin{flushright}
\Huge
\begin{tabular}{@{}l}
\LaTeXe{} Class\\
for Lecture Notes\\
in Computer Science\\[6pt] 
{\Large Version 2.4}
\end{tabular}
\end{flushright}
\rule{\textwidth}{1pt}
\vfill
%\begin{flushleft}
%\large\itshape 
%\begin{tabular}{@{}l}
%{\Large\upshape\bfseries Springer}\\[8pt]
%Berlin\enspace Heidelberg\enspace New\kern0.1em York\\[5pt]
%Barcelona\enspace Budapest\enspace Hong\kern0.2em Kong\\[5pt]
%London\enspace Milan\enspace Paris\enspace\\[5pt]
%Santa\kern0.2em Clara\enspace Singapore\enspace Tokyo
%\end{tabular}
%\end{flushleft}
\newpage
%
\section*{For further information please contact us:}
%
\begin{flushleft}
\begin{tabular}{l@{\quad}l@{\hspace{3mm}}l@{\qquad}l}
$\bullet$&\multicolumn{3}{@{}l}{\bfseries LNCS Editorial Office}\\[1mm]
&\multicolumn{3}{@{}l}{Springer-Verlag}\\
&\multicolumn{3}{@{}l}{Computer Science Editorial}\\
&\multicolumn{3}{@{}l}{Tiergartenstra▀e 17}\\
&\multicolumn{3}{@{}l}{69121 Heidelberg}\\
&\multicolumn{3}{@{}l}{Germany}\\[0.5mm]
 & Tel:       & +49-6221-487-8706\\
 & Fax:       & +49-6221-487-8588\\
 & e-mail:    & \tt lncs@springer.com    & for editorial questions\\
 &            & \tt texhelp@springer.de & for \TeX{} problems\\[2mm]
\noalign{\rule{\textwidth}{1pt}}
\noalign{\vskip2mm}
%
%{\tt svserv@vax.ntp.springer.de}\hfil first try the \verb|help|
%command.
%
$\bullet$&\multicolumn{3}{@{}l}{\bfseries We are also reachable through the world wide web:}\\[1mm]
         &\multicolumn{2}{@{}l}{\texttt{http://www.springer.com}}&Springer Global Website\\
         &\multicolumn{2}{@{}l}{\texttt{http://www.springer.com/lncs}}&LNCS home page\\
         &\multicolumn{2}{@{}l}{\texttt{http://www.springerlink.com}}&data repository\\
         &\multicolumn{2}{@{}l}{\texttt{ftp://ftp.springer.de}}&FTP server
\end{tabular}
\end{flushleft}

%
\newpage
\tableofcontents
\newpage
%
\section{Introduction}
%
Authors wishing to code their contribution
with \LaTeX{}, as well as those who have already coded with \LaTeX{},
will be provided with a document class that will give the text the
desired layout. Authors are requested to
adhere strictly to these instructions; {\em the class
file must not be changed}.

The text output area is automatically set within an area of
12.2\,cm horizontally  and 19.3\,cm vertically.

If you are already familiar with \LaTeX{}, then the
LLNCS class should not give you any major difficulties.
It will change the layout to the required LLNCS style
(it will for instance define the layout of \verb|\section|).
We had to invent some extra commands,
which are not provided by \LaTeX{} (e.g.\
\verb|\institute|, see also Sect.\,\ref{contbegin})

For the main body of the paper (the text) you
should use the commands of the standard \LaTeX{} ``article'' class.
Even if you are familiar with those commands, we urge you to read
this entire documentation thoroughly. It contains many suggestions on
how to use our commands properly; thus your paper
will be formatted exactly to LLNCS standard.
For the input of the references at the end of your contribution,
please follow our instructions given in Sect.\,\ref{refer} References.

The majority of these hints are not specific for LLNCS; they may improve
your use of \LaTeX{} in general.
Furthermore, the documentation provides suggestions about the proper
editing and use
of the input files (capitalization, abbreviation etc.) (see
Sect.\,\ref{refedit} How to Edit Your Input File).
%
\section{How to Proceed}
%
The package consists of the following files:
\begin{flushleft}
\begin{tabular}{@{}p{2.5cm}l}
{\tt history.txt}& the version history of the package\\[2pt]
{\tt llncs.cls}  & class file for \LaTeX{}\\[2pt]
{\tt llncs.dem}  & an example showing how to code the text\\[2pt]
{\tt llncs.doc}  & general instructions (source of this document),\\
        & {\tt llncs.doc} means {\itshape l\/}atex {\itshape doc\/}umentation for\\
        & {\itshape L\/}ecture {\itshape N}otes in {\itshape C\/}omputer {\itshape S\/}cience\\
{\tt llncsdoc.pdf}& the documentation of the class (PDF version),\\
{\tt llncs.doc}  & general instructions (source of this document),\\
{\tt llncsdoc.sty}  & class modifications to help for the instructions\\
{\tt llncs.ind} & an external (faked) author index file\\
{\tt subjidx.ind} & subject index demo from the Springer book package\\
{\tt llncs.dvi}    & the resultig DVI file (remember to use binary transfer!)\\[2pt]
{\tt sprmindx.sty} & supplementary style file for MakeIndex\\
                   & (usage: {\tt makeindex -s sprmindx.sty <yourfile.idx>})
\end{tabular}
\end{flushleft}
%
\subsection{How to Invoke the LLNCS Document Class}
%
The LLNCS class is an extension of the standard \LaTeX{} ``article''
document class. Therefore you may use all ``article'' commands for the
body of your contribution to prepare your manuscript.
LLNCS class is invoked by replacing ``article'' by ``llncs'' in the
first line of your document:
\begin{verbatim}
\documentclass{llncs}
%
\begin{document}
  <Your contribution>
\end{document}
\end{verbatim}
%
\subsection{Contributions Already Coded with \protect\LaTeX{} without
the LLNCS document class}
%
If your file is already coded with \LaTeX{} you can easily
adapt it a posteriori to the LLNCS document class.

Please refrain from using any \LaTeX{} or \TeX{} commands
that affect the layout or formatting of your document (i.e. commands
like \verb|\textheight|, \verb|\vspace|, \verb|\headsep| etc.).
There may nevertheless be exceptional occasions on which to
use some of them.

The LLNCS document class has been carefully designed to produce the
right layout from your \LaTeX{} input. If there is anything specific you
would like to do and for which the style file does not provide a
command, {\em please contact us}. Same holds for any error and bug you
discover (there is however no reward for this -- sorry).
%
\section{General Rules for Coding Formulas}
%
With mathematical formulas you may proceed as described
in Sect.\,3.3 of the {\em \LaTeX{} User's Guide \& Reference
Manual\/} by Leslie Lamport (2nd~ed. 1994), Addison-Wesley Publishing
Company, Inc.

Equations are automatically numbered sequentially throughout your
contribution using arabic numerals in parentheses on the right-hand
side.

When you are working in math mode everything is typeset in italics.
Sometimes you need to insert non-mathematical elements (e.g.\
words or phrases). Such insertions should be coded in roman
(with \verb|\mbox|) as illustrated in the following example:
\begin{flushleft}
{\itshape Sample Input}
\end{flushleft}
\begin{verbatim}
\begin{equation}
  \left(\frac{a^{2} + b^{2}}{c^{3}} \right) = 1 \quad
  \mbox{ if } c\neq 0 \mbox{ and if } a,b,c\in \bbbr \enspace .
\end{equation}
\end{verbatim}
{\itshape Sample Output}
\begin{equation}
  \left(\frac{a^{2} + b^{2}}{c^{3}} \right) = 1 \quad
  \mbox{ if } c\neq 0 \mbox{ and if } a,b,c\in \bbbr \enspace .
\end{equation}

If you wish to start a new paragraph immediately after a displayed
equation, insert a blank line so as to produce the required
indentation. If there is no new paragraph either do not insert
a blank line or code \verb|\noindent| immediately before
continuing the text.

Please punctuate a displayed equation in the same way as other
ordinary text but with an \verb|\enspace| before end punctuation.

Note that the sizes of the parentheses or other delimiter
symbols used in equations should ideally match the height of the
formulas being enclosed. This is automatically taken care of by
the following \LaTeX{} commands:\\[2mm]
\verb|\left(| or \verb|\left[| and
\verb|\right)| or \verb|\right]|.
%
\subsection{Italic and Roman Type in Math Mode}
%
\begin{alpherate}
\item
In math mode \LaTeX{} treats all letters as though they
were mathematical or physical variables, hence they are typeset as
characters of their own in
italics. However, for certain components of formulas, like short texts,
this would be incorrect and therefore coding in roman is required.
Roman should also be used for
subscripts and superscripts {\em in formulas\/} where these are
merely labels and not in themselves variables,
e.g. $T_{\mathrm{eff}}$ \emph{not} $T_{eff}$,
$T_{\mathrm K}$ \emph{not} $T_K$ (K = Kelvin),
$m_{\mathrm e}$ \emph{not} $m_e$ (e = electron).
However, do not code for roman
if the sub/superscripts represent variables,
e.g.\ $\sum_{i=1}^{n} a_{i}$.
\item
Please ensure that {\em physical units\/} (e.g.\ pc, erg s$^{-1}$
K, cm$^{-3}$, W m$^{-2}$ Hz$^{-1}$, m kg s$^{-2}$ A$^{-2}$) and
{\em abbreviations\/} such as Ord, Var, GL, SL, sgn, const.\
are always set in roman type. To ensure
this use the \verb|\mathrm| command: \verb|\mathrm{Hz}|.
On p.\ 44 of the {\em \LaTeX{} User's Guide \& Reference
Manual\/} by Leslie Lamport you will find the names of
common mathe\-matical functions, such as log, sin, exp, max and sup.
These should be coded as \verb|\log|,
\verb|\sin|, \verb|\exp|, \verb|\max|, \verb|\sup|
and will appear in roman automatically.
\item
Chemical symbols and formulas should be coded for roman,
e.g.\ Fe not $Fe$, H$_2$O not {\em H$_2$O}.
\item
Familiar foreign words and phrases, e.g.\ et al.,
a priori, in situ, brems\-strah\-lung, eigenvalues should not be
italicized.
\end{alpherate}
%
\section{How to Edit Your Input (Source) File}
\label{refedit}
%
\subsection{Headings}\label{headings}
%
All words in headings should be capitalized except for conjunctions,
prepositions (e.g.\ on, of, by, and, or, but, from, with, without,
under) and definite and indefinite articles (the, a, an) unless they
appear at the beginning. Formula letters must be typeset as in the text.
%
\subsection{Capitalization and Non-capitalization}
%
\begin{alpherate}
\item
The following should always be capitalized:
\begin{itemize}
\item
Headings (see preceding Sect.\,\ref{headings})
\item
Abbreviations and expressions
in the text such as  Fig(s)., Table(s), Sect(s)., Chap(s).,
Theorem, Corollary, Definition etc. when used with numbers, e.g.\
Fig.\,3, Table\,1, Theorem 2.
\end{itemize}
Please follow the special rules in Sect.\,\ref{abbrev} for referring to
equations.
\item
The following should {\em not\/} be capitalized:
\begin{itemize}
\item
The words figure(s), table(s), equation(s), theorem(s) in the text when
used without an accompanying number.
\item
Figure legends and table captions except for names and abbreviations.
\end{itemize}
\end{alpherate}
%
\subsection{Abbreviation of Words}\label{abbrev}
%
\begin{alpherate}
\item
The following {\em should} be abbreviated when they appear in running
text {\em unless\/} they come at the beginning of a sentence: Chap.,
Sect., Fig.; e.g.\ The results are depicted in Fig.\,5. Figure 9 reveals
that \dots .\\
{\em Please note\/}: Equations should usually be referred to solely by
their number in parentheses: e.g.\ (14). However, when the reference
comes at the beginning of a sentence, the unabbreviated word
``Equation'' should be used: e.g.\ Equation (14) is very important.
However, (15) makes it clear that \dots .
\item
If abbreviations of names or concepts are used
throughout the text, they should be defined at first occurrence,
e.g.\ Plurisubharmonic (PSH) Functions, Strong Optimization (SOPT)
Problem.
\end{alpherate}
%
\section{How to Code the Beginning of Your Contribution}
\label{contbegin}
%
The title of a single contribution (it is mandatory) should be coded as
follows:
\begin{verbatim}
\title{<Your contribution title>}
\end{verbatim}
All words in titles should be capitalized except for conjunctions,
prepositions (e.g.\ on, of, by, and, or, but, from, with, without,
under) and definite and indefinite articles (the, a, an) unless they
appear at the beginning. Formula letters must be typeset as in the text.
Titles have no end punctuation.

If a long \verb|\title| must be divided please use the code \verb|\\|
(for new line).

If you are to produce running heads for a specific volume the standard
(of no such running heads) is overwritten with the \verb|[runningheads]|
option in the \verb|\documentclass| line. For long titles that do not
fit in the single line of the running head a warning is generated.
You can specify an abbreviated title for the running head on odd pages
with the command
\begin{verbatim}
\titlerunning{<Your abbreviated contribution title>}
\end{verbatim}

There is also a possibility to change the text of the title that goes
into the table of contents (that's for volume editors only -- there is
no table of contents for a single contribution). For this use the
command
\begin{verbatim}
\toctitle{<Your changed title for the table of contents>}
\end{verbatim}

An optional subtitle may follow then:
\begin{verbatim}
\subtitle{<subtitle of your contribution>}
\end{verbatim}

Now the name(s) of the author(s) must be given:
\begin{verbatim}
\author{<author(s) name(s)>}
\end{verbatim}
Numbers referring to different addresses or affiliations are
to be attached to each author with the \verb|\inst{<no>}| command.
If there is more than one author, the order is up to you;
the \verb|\and| command provides for the separation.

If you have done this correctly, this entry now reads, for example:
\begin{verbatim}
\author{Ivar Ekeland\inst{1} \and Roger Temam\inst{2}}
\end{verbatim}
The first name\footnote{Other initials are optional
and may be inserted if this is the usual
way of writing your name, e.g.\ Alfred J.~Holmes, E.~Henry Green.}
is followed by the surname.

As for the title there exist two additional commands (again for volume
editors only) for a different author list. One for the running head
(on odd pages) -- if there is any:
\begin{verbatim}
\authorrunning{<abbreviated author list>}
\end{verbatim}
And one for the table of contents where the
affiliation of each author is simply added in braces.
\begin{verbatim}
\tocauthor{<enhanced author list for the table of contents>}
\end{verbatim}

Next the address(es) of institute(s), company etc. is (are) required.
If there is more than one address, the entries are numbered
automatically with \verb|\and|, in the order in which you type them.
Please make sure that the numbers match those placed next to
to the authors' names to reflect the affiliation.
\begin{verbatim}
\institute{<name of an institute>
\and <name of the next institute>
\and <name of the next institute>}
\end{verbatim}

In addition, you can use
\begin{verbatim}
\email{<email address>}
\end{verbatim}
to provide your email address within \verb|\institute|. If you need to
typeset the tilde character -- e.g. for your web page in your unix
system's home directory -- the \verb|\homedir| command will happily do
this. Please note that, if your email address is given in your paper, 
it will also be included in the meta data of the online version.
\enlargethispage*{6mm}

\medskip
If footnote like things are needed anywhere in the contribution heading
please code
(immediately after the word where the footnote indicator should be
placed):
\begin{verbatim}
\thanks{<text>}
\end{verbatim}
\verb|\thanks| may only appear in \verb|\title|, \verb|\author|
and \verb|\institute| to footnote anything. If there are two or more
footnotes or affiliation marks to a specific item separate them with
\verb|\fnmsep| (i.e. {\itshape f}oot\emph note \emph mark
\emph{sep}arator).

\medskip\noindent
The command
\begin{verbatim}
\maketitle
\end{verbatim}
then formats the complete heading of your article. If you leave
it out the work done so far will produce \emph{no} text.

Then the abstract should follow. Simply code
\begin{verbatim}
\begin{abstract}
<Text of the summary of your article>
\end{abstract}
\end{verbatim}
or refer to the demonstration file {\tt llncs.dem} for an example or
to the {\em Sample Input\/} on p.~\pageref{samppage}.

\subsubsection{Remark to Running Heads and the Table of Contents}
\leavevmode\\[\medskipamount]
If you are the author of a single contribution you normally have no
running heads and no table of contents. Both are done only by the editor
of the volume or at the printers.
%
\section{Special Commands for the Volume Editor}
The volume editor can produce a complete camera ready output including
running heads, a table of contents, preliminary text (frontmatter), and
index or glossary. For activating the running heads there is the class
option \verb|[runningheads]|.

The table of contents of the volume is printed wherever
\verb|\tableofcontents| is placed. A simple compilation of all
contributions (fields \verb|\title| and \verb|\author|) is done. If you
wish to change this automatically produced list use the commands
\begin{verbatim}
\titlerunning  \toctitle
\authorrunning \tocauthor
\end{verbatim}
to enhance the information in the specific contributions. See the
demonstration file \verb|llncs.dem| for examples.

An additional structure can be added to the table of contents with the
\verb|\addtocmark{<text>}| command. It has an optional numerical
argument, a digit from 1 through 3. 3 (the default) makes an unnumbered
chapter like entry in the table of contents. If you code
\verb|\addtocmark[2]{text}| the corresponding page number is listed
also, \verb|\addtocmark[1]{text}| even introduces a chapter number
beyond it.
%
\section{How to Code Your Text}
%
The contribution title and all headings should be capitalized
except for conjunctions, prepositions (e.g.\ on, of, by, and, or, but,
from, with, without, under) and definite and indefinite articles (the,
a, an) unless they appear at the beginning. Formula letters must be
typeset as in the text.

Headings will be automatically numbered by the following codes.\\[2mm]
{\itshape Sample Input}
\begin{verbatim}
\section{This is a First-Order Title}
\subsection{This is a Second-Order Title}
\subsubsection{This is a Third-Order Title.}
\paragraph{This is a Fourth-Order Title.}
\end{verbatim}
\verb|\section| and \verb|\subsection| have no end punctuation.\\
\verb|\subsubsection| and \verb|\paragraph|
need to be punctuated at the end.

In addition to the above-mentioned headings your text may be structured
by subsections indicated by run-in headings (theorem-like environments).
All the theorem-like environments are numbered automatically
throughout the sections of your document -- each with its own counter.
If you want the theorem-like environments to use the same counter
just specify the documentclass option \verb|envcountsame|:
\begin{verbatim}
\documentclass[envcountsame]{llncs}
\end{verbatim}
If your first call for a theorem-like environment then is e.g.
\verb|\begin{lemma}|, it will be numbered 1; if corollary follows,
this will be numbered 2; if you then call lemma again, this will be
numbered 3.

But in case you want to reset such counters to 1 in each section,
please specify the documentclass option \verb|envcountreset|:
\begin{verbatim}
\documentclass[envcountreset]{llncs}
\end{verbatim}

Even a numbering on section level (including the section counter) is
possible with the documentclass option \verb|envcountsect|.

\section{Predefined Theorem like Environments}\label{builtintheo}
The following variety of run-in headings are at your disposal:
\begin{alpherate}
\item
{\bfseries Bold} run-in headings with italicized text
as built-in environments:
\begin{verbatim}
\begin{corollary} <text> \end{corollary}
\begin{lemma} <text> \end{lemma}
\begin{proposition} <text> \end{proposition}
\begin{theorem} <text> \end{theorem}
\end{verbatim}
\item
The following generally appears as {\itshape italic} run-in heading:
\begin{verbatim}
\begin{proof} <text>    \qed    \end{proof}
\end{verbatim}
It is unnumbered and may contain an eye catching square (call for that
with \verb|\qed|) before the environment ends.
\item
Further {\itshape italic} or {\bfseries bold} run-in headings with roman
environment body may also occur:
\begin{verbatim}
\begin{definition} <text> \end{definition}
\begin{example} <text> \end{example}
\begin{exercise} <text> \end{exercise}
\begin{note} <text> \end{note}
\begin{problem} <text> \end{problem}
\begin{question} <text> \end{question}
\begin{remark} <text> \end{remark}
\begin{solution} <text> \end{solution}
\end{verbatim}
\end{alpherate}

\section{Defining your Own Theorem like Environments}
We have enhanced the standard \verb|\newtheorem| command and slightly
changed its syntax to get two new commands \verb|\spnewtheorem| and
\verb|\spnewtheorem*| that now can be used to define additional
environments. They require two additional arguments namely the type
style in which the keyword of the environment appears and second the
style for the text of your new environment.

\verb|\spnewtheorem| can be used in two ways.
\subsection{Method 1 {\itshape (preferred)}}
You may want to create an environment that shares its counter
with another environment, say {\em main theorem\/} to be numbered like
the predefined {\em theorem\/}. In this case, use the syntax
\begin{verbatim}
\spnewtheorem{<env_nam>}[<num_like>]{<caption>}
{<cap_font>}{<body_font>}
\end{verbatim}

\noindent
Here the environment with which the new environment should share its
counter is specified with the optional argument \verb|[<num_like>]|.

\paragraph{Sample Input}
\begin{verbatim}
\spnewtheorem{mainth}[theorem]{Main Theorem}{\bfseries}{\itshape}
\begin{theorem} The early bird gets the worm. \end{theorem}
\begin{mainth} The early worm gets eaten. \end{mainth}
\end{verbatim}
\medskip\noindent
{\em Sample Output}

\medskip\noindent
{\bfseries Theorem 3.}\enspace {\em The early bird gets the worm.}

\medskip\noindent
{\bfseries Main Theorem 4.} The early worm gets eaten.

\bigskip
The sharing of the default counter (\verb|[theorem]|) is desired. If you
omit the optional second argument of \verb|\spnewtheorem| a separate
counter for your new environment is used throughout your document.

\subsection[Method 2]{Method 2 {\itshape (assumes {\tt[envcountsect]}
documentstyle option)}}
\begin{verbatim}
\spnewtheorem{<env_nam>}{<caption>}[<within>]
{<cap_font>}{<body_font>}
\end{verbatim}

\noindent
This defines a new environment \verb|<env_nam>| which prints the caption
\verb|<caption>| in the font \verb|<cap_font>| and the text itself in
the font \verb|<body_font>|. The environment is numbered beginning anew
with every new sectioning element you specify with the optional
parameter \verb|<within>|.

\medskip\noindent
\paragraph{Example} \leavevmode

\medskip\noindent
\verb|\spnewtheorem{joke}{Joke}[subsection]{\bfseries}{\rmfamily}|

\medskip
\noindent defines a new environment called \verb|joke| which prints the
caption {\bfseries Joke} in boldface and the text in roman. The jokes are
numbered starting from 1 at the beginning of every subsection with the
number of the subsection preceding the number of the joke e.g. 7.2.1 for
the first joke in subsection 7.2.

\subsection{Unnumbered Environments}
If you wish to have an unnumbered environment, please
use the syntax
\begin{verbatim}
\spnewtheorem*{<env_nam>}{<caption>}{<cap_font>}{<body_font>}
\end{verbatim}

\section{Program Codes}
In case you want to show pieces of program code, just use the
\verb|verbatim| environment or the \verb|verbatim| package of \LaTeX.
(There also exist various pretty printers for some programming
languages.)
%
\noindent
\subsection*{Sample Input {\rmfamily(of a simple
contribution)}}\label{samppage}
\begin{verbatim}
\title{Hamiltonian Mechanics}

\author{Ivar Ekeland\inst{1} \and Roger Temam\inst{2}}

\institute{Princeton University, Princeton NJ 08544, USA
\and
Universit\'{e} de Paris-Sud,
Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
F-91405 Orsay Cedex, France}

\maketitle
%
\begin{abstract}
This paragraph shall summarize the contents of the paper
in short terms.
\end{abstract}
%
\section{Fixed-Period Problems: The Sublinear Case}
%
With this chapter, the preliminaries are over, and we begin the
search for periodic solutions \dots
%
\subsection{Autonomous Systems}
%
In this section we will consider the case when the Hamiltonian
$H(x)$ \dots
%
\subsubsection*{The General Case: Nontriviality.}
%
We assume that $H$ is
$\left(A_{\infty}, B_{\infty}\right)$-subqua\-dra\-tic
at infinity, for some constant \dots
%
\paragraph{Notes and Comments.}
The first results on subharmonics were \dots
%
\begin{proposition}
Assume $H'(0)=0$ and $ H(0)=0$. Set \dots
\end{proposition}
\begin{proof}[of proposition]
Condition (8) means that, for every $\delta'>\delta$, there is
some $\varepsilon>0$ such that \dots \qed
\end{proof}
%
\begin{example}[\rmfamily (External forcing)]
Consider the system \dots
\end{example}
\begin{corollary}
Assume $H$ is $C^{2}$ and
$\left(a_{\infty}, b_{\infty}\right)$-subquadratic
at infinity. Let \dots
\end{corollary}
\begin{lemma}
Assume that $H$ is $C^{2}$ on $\bbbr^{2n}\backslash \{0\}$
and that $H''(x)$ is \dots
\end{lemma}
\begin{theorem}[(Ghoussoub-Preiss)]
Let $X$ be a Banach Space and $\Phi:X\to\bbbr$ \dots
\end{theorem}
\begin{definition}
We shall say that a $C^{1}$ function $\Phi:X\to\bbbr$
satisfies \dots
\end{definition}
\end{verbatim}
{\itshape Sample Output\/} (follows on the next page together with
examples of the above run-in headings)
\newcounter{save}\setcounter{save}{\value{section}}
{\def\addtocontents#1#2{}%
\def\addcontentsline#1#2#3{}%
\def\markboth#1#2{}%
%
\title{Hamiltonian Mechanics}

\author{Ivar Ekeland\inst{1} \and Roger Temam\inst{2}}

\institute{Princeton University, Princeton NJ 08544, USA
\and
Universit\'{e} de Paris-Sud,
Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
F-91405 Orsay Cedex, France}

\maketitle
%
\begin{abstract}
This paragraph shall summarize the contents of the paper
in short terms.
\end{abstract}
%
\section{Fixed-Period Problems: The Sublinear Case}
%
With this chapter, the preliminaries are over, and we begin the search
for periodic solutions \dots
%
\subsection{Autonomous Systems}
%
In this section we will consider the case when the Hamiltonian
$H(x)$ \dots
%
\subsubsection{The General Case: Nontriviality.}
%
We assume that $H$ is
$\left(A_{\infty}, B_{\infty}\right)$-subqua\-dra\-tic at
infinity, for some constant \dots
%
\paragraph{Notes and Comments.}
The first results on subharmonics were \dots
%
\begin{proposition}
Assume $H'(0)=0$ and $ H(0)=0$. Set \dots
\end{proposition}
\begin{proof}[of proposition]
Condition (8) means that, for every $\delta'>\delta$, there is
some $\varepsilon>0$ such that \dots \qed
\end{proof}
%
\begin{example}[{{\rmfamily External forcing}}]
Consider the system \dots
\end{example}
\begin{corollary}
Assume $H$ is $C^{2}$ and
$\left(a_{\infty}, b_{\infty}\right)$-subquadratic
at infinity. Let \dots
\end{corollary}
\begin{lemma}
Assume that $H$ is $C^{2}$ on $\bbbr^{2n}\backslash \{0\}$
and that $H''(x)$ is \dots
\end{lemma}
\begin{theorem}[Ghoussoub-Preiss]
Let $X$ be a Banach Space and $\Phi:X\to\bbbr$ \dots
\end{theorem}
\begin{definition}
We shall say that a $C^{1}$ function $\Phi:X\to\bbbr$ satisfies \dots
\end{definition}
%
}\setcounter{section}{\value{save}}
\section{Fine Tuning of the Text}
%
The following should be used to improve the readability of the text:
\begin{flushleft}
\begin{tabular}{@{}p{.19\textwidth}p{.79\textwidth}}
\verb|\,|   & a thin space, e.g.\ between numbers or between units
              and num\-bers; a line division will not be made
              following this space\\
\verb|--|   & en dash; two strokes, without a space at either end\\
\verb*| -- |& en dash; two strokes, with  a space at either end\\
\verb|-|    & hyphen; one stroke, no space at either end\\
\verb|$-$|  & minus, in the text {\em only} \\[8mm]
{\em Input} & \verb|21\,$^{\circ}$C etc.,|\\
            &  \verb|Dr h.\,c.\,Rockefellar-Smith \dots|\\
            & \verb|20,000\,km and  Prof.\,Dr Mallory \dots|\\
            & \verb|1950--1985 \dots|\\
            & \verb|this -- written on a computer -- is now printed|\\
            & \verb|$-30$\,K \dots|\\[3mm]
{\em Output}& 21\,$^{\circ}$C etc., Dr h.\,c.\,Rockefellar-Smith \dots\\
            & 20,000\,km and  Prof.\,Dr Mallory \dots\\
            & 1950--1985 \dots\\
            & this -- written on a computer -- is now printed\\
            & $-30$\,K \dots
\end{tabular}
\end{flushleft}
%
\section {Special Typefaces}
%
Normal type (roman text) need not be coded. {\itshape Italic}
(\verb|{\em <text>}| better still \verb|\emph{<text>}|) or, if
necessary, {\bfseries boldface} should be used for emphasis.\\[6pt]
\begin{minipage}[t]{\textwidth}
\begin{flushleft}
\begin{tabular}{@{}p{.25\textwidth}@{\hskip6pt}p{.73\textwidth}@{}}
\verb|{\itshape Text}|   & {\itshape Italicized Text}\\[2pt]
\verb|{\em Text}|   & {\em Emphasized Text --
   if you would like to emphasize a {\em definition} within an
   italicized text (e.g.\ of a {\em theorem)} you should code the
   expression to be emphasized by} \verb|\em|.\\[2pt]
\verb|{\bfseries Text}|& {\bfseries Important Text}\\[2pt]
\verb|\vec{Symbol}| & Vectors may only appear in math mode. The default
   \LaTeX{} vector symbol has been adapted\footnotemark\
 to LLNCS conventions.\\[2pt]
 & \verb|$\vec{A \times B\cdot C}| yields $\vec{A\times B\cdot C}$\\
 & \verb|$\vec{A}^{T} \otimes \vec{B} \otimes|\\
 & \verb|\vec{\hat{D}}$|yields $\vec{A}^{T} \otimes \vec{B} \otimes
\vec{\hat{D}}$
\end{tabular}
\end{flushleft}
\end{minipage}

\footnotetext{If you absolutely must revive the original \LaTeX{}
design of the vector symbol (as an arrow accent), please specify the
option \texttt{[orivec]} in the \texttt{documentclass} line.}
\newpage
%
\section {Footnotes}
%
Footnotes within the text should be coded:
\begin{verbatim}
\footnote{Text}
\end{verbatim}
{\itshape Sample Input}
\begin{flushleft}
Text with a footnote\verb|\footnote{The |{\tt footnote is automatically
numbered.}\verb|}| and text continues \dots
\end{flushleft}
{\itshape Sample Output}
\begin{flushleft}
Text with a footnote\footnote{The footnote is automatically numbered.}
and text continues \dots
\end{flushleft}
%
\section {Lists}
%
Please code lists as described below:\\[2mm]
{\itshape Sample  Input}
\begin{verbatim}
\begin{enumerate}
  \item First item
  \item Second item
  \begin{enumerate}
    \item First nested item
    \item Second nested item
  \end{enumerate}
  \item Third item
\end{enumerate}
\end{verbatim}
{\itshape Sample Output}
 \begin{enumerate}
\item First item
\item Second item
  \begin{enumerate}
    \item First nested item
    \item Second nested item
  \end{enumerate}
\item Third item
\end{enumerate}
%
\section {Figures}
%
Figure environments should be inserted after (not in)
the  paragraph in which the figure is first mentioned.
They will be numbered automatically.

Preferably the images should be enclosed as PostScript files -- best as
EPS data using the epsfig package.

If you cannot include them into your output this way and use other
techniques for a separate production,
the figures (line drawings and those containing halftone inserts
as well as halftone figures) {\em should not be pasted into your
laserprinter output}. They should be enclosed separately in camera-ready
form (original artwork, glossy prints, photographs and/or slides). The
lettering should be suitable for reproduction, and after a
probably necessary reduction the height of capital letters should be at
least 1.8\,mm and not more than 2.5\,mm.
Check that lines and other details are uniformly black and
that the lettering on figures is clearly legible.

To leave the desired amount of space for the height of
your figures, please use the coding described below.
As can be seen in the output, we will automatically
provide 1\,cm space above and below the figure,
so that you should only leave the space equivalent to the size of the
figure itself. Please note that ``\verb|x|'' in the following
coding stands for the actual height of the figure:
\begin{verbatim}
\begin{figure}
\vspace{x cm}
\caption[ ]{...text of caption...}          (Do type [ ])
\end{figure}
\end{verbatim}
\begin{flushleft}
{\itshape Sample Input}
\end{flushleft}
\begin{verbatim}
\begin{figure}
\vspace{2.5cm}
\caption{This is the caption of the figure displaying a white
eagle and a white horse on a snow field}
\end{figure}
\end{verbatim}
\begin{flushleft}
{\itshape Sample Output}
\end{flushleft}
\begin{figure}
\vspace{2.5cm}
\caption{This is the caption of the figure displaying a white eagle and
a white horse on a snow field}
\end{figure}
%
\section{Tables}
%
Table captions should be treated
in the same way as figure legends, except that
the table captions appear {\itshape above} the tables. The tables
will be numbered automatically.
%
\subsection{Tables Coded with \protect\LaTeX{}}
%
Please use the following coding:\\[2mm]
{\itshape Sample Input}
\begin{verbatim}
\begin{table}
\caption{Critical $N$ values}
\begin{tabular}{llllll}
\hline\noalign{\smallskip}
${\mathrm M}_\odot$ & $\beta_{0}$ & $T_{\mathrm c6}$ & $\gamma$
  & $N_{\mathrm{crit}}^{\mathrm L}$
  & $N_{\mathrm{crit}}^{\mathrm{Te}}$\\
\noalign{\smallskip}
\hline
\noalign{\smallskip}
 30 & 0.82 & 38.4 & 35.7 & 154 & 320 \\
 60 & 0.67 & 42.1 & 34.7 & 138 & 340 \\
120 & 0.52 & 45.1 & 34.0 & 124 & 370 \\
\hline
\end{tabular}
\end{table}
\end{verbatim}

\medskip\noindent{\itshape Sample Output}
\begin{table}
\caption{Critical $N$ values}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{llllll}
\hline\noalign{\smallskip}
${\mathrm M}_\odot$ & $\beta_{0}$ & $T_{\mathrm c6}$ & $\gamma$
  & $N_{\mathrm{crit}}^{\mathrm L}$
  & $N_{\mathrm{crit}}^{\mathrm{Te}}$\\
\noalign{\smallskip}
\hline
\noalign{\smallskip}
 30 & 0.82 & 38.4 & 35.7 & 154 & 320 \\
 60 & 0.67 & 42.1 & 34.7 & 138 & 340 \\
120 & 0.52 & 45.1 & 34.0 & 124 & 370 \\
\hline
\end{tabular}
\end{center}
\end{table}

Before continuing your text you need an empty line. \dots

\vspace{3mm}
For further information you will find a complete description of
the tabular environment
on p.~62~ff. and p.~204 of the {\em \LaTeX{} User's Guide \& Reference
Manual\/} by Leslie Lamport.
%
\subsection{Tables Not Coded with \protect\LaTeX{}}
%
If you do not wish to code your table using \LaTeX{}
but prefer to have it reproduced separately,
proceed as for figures and use the following coding:\\[2mm]
{\itshape Sample Input}
\begin{verbatim}
\begin{table}
\caption{text of your caption}
\vspace{x cm}     % the actual height needed for your table
\end{table}
\end{verbatim}
%
\subsection{Signs and Characters}
%
\subsubsection*{Special Signs.}
%
You may need to use special signs.  The available ones are listed in the
{\em \LaTeX{} User's Guide \& Reference Manual\/} by Leslie Lamport,
pp.~41\,ff.
We have created further symbols for math mode (enclosed in \$):
\begin{center}
\begin{tabular}{l@{\hspace{1em}yields\hspace{1em}}
c@{\hspace{3em}}l@{\hspace{1em}yields\hspace{1em}}c}
\verb|\grole| & $\grole$ & \verb|\getsto| & $\getsto$\\
\verb|\lid|   & $\lid$   & \verb|\gid|    & $\gid$
\end{tabular}
\end{center}
%
\subsubsection*{Gothic (Fraktur).}
%
If gothic letters are {\itshape necessary}, please use those of the
relevant \AmSTeX{} alphabet which are available using the amstex
package of the American Mathematical Society.

In \LaTeX{} only the following gothic letters are available:
\verb|$\Re$| yields $\Re$ and \verb|$\Im$| yields $\Im$. These should
{\itshape not\/} be used when you need gothic letters for your contribution.
Use \AmSTeX{} gothic as explained above. For the real and the imaginary
parts of a complex number within math mode you should use instead:
\verb|$\mathrm{Re}$| (which yields Re) or \verb|$\mathrm{Im}$| (which
yields Im).
%
\subsubsection*{Script.}
%
For script capitals use the coding
\begin{center}
\begin{tabular}{l@{\hspace{1em}which yields\hspace{1em}}c}
\verb|$\mathcal{AB}$| & $\mathcal{AB}$
\end{tabular}
\end{center}
(see p.~42 of  the \LaTeX{} book).
%
\subsubsection*{Special Roman.}
%
If you need other symbols than those below, you could use
the blackboard bold characters of \AmSTeX{},  but there might arise
capacity problems
in loading additional \AmSTeX{} fonts. Therefore  we created
the blackboard bold characters listed below.
Some of them are not esthetically
satisfactory. This need not deter you from using them:
in the final printed form they will be
replaced by the well-designed MT (monotype) characters of
the phototypesetting machine.
\begin{flushleft}
\begin{tabular}{@{}ll@{ yields }
c@{\hspace{1.em}}ll@{ yields }c}
\verb|\bbbc| & (complex numbers)   & $\bbbc$
  & \verb|\bbbf| & (blackboard bold F) & $\bbbf$\\
\verb|\bbbh| & (blackboard bold H) & $\bbbh$
  & \verb|\bbbk| & (blackboard bold K) & $\bbbk$\\
\verb|\bbbm| & (blackboard bold M) & $\bbbm$
  & \verb|\bbbn| & (natural numbers N) & $\bbbn$\\
\verb|\bbbp| & (blackboard bold P) & $\bbbp$
  & \verb|\bbbq| & (rational numbers)  & $\bbbq$\\
\verb|\bbbr| & (real numbers)      & $\bbbr$
  & \verb|\bbbs| & (blackboard bold S) & $\bbbs$\\
\verb|\bbbt| & (blackboard bold T) & $\bbbt$
  & \verb|\bbbz| & (whole numbers)     & $\bbbz$\\
\verb|\bbbone| & (symbol one)      & $\bbbone$
\end{tabular}
\end{flushleft}
\begin{displaymath}
\begin{array}{c}
\bbbc^{\bbbc^{\bbbc}} \otimes
\bbbf_{\bbbf_{\bbbf}} \otimes
\bbbh_{\bbbh_{\bbbh}} \otimes
\bbbk_{\bbbk_{\bbbk}} \otimes
\bbbm^{\bbbm^{\bbbm}} \otimes
\bbbn_{\bbbn_{\bbbn}} \otimes
\bbbp^{\bbbp^{\bbbp}}\\[2mm]
\otimes
\bbbq_{\bbbq_{\bbbq}} \otimes
\bbbr^{\bbbr^{\bbbr}} \otimes
\bbbs^{\bbbs_{\bbbs}} \otimes
\bbbt^{\bbbt^{\bbbt}} \otimes
\bbbz \otimes
\bbbone^{\bbbone_{\bbbone}}
\end{array}
\end{displaymath}
%
\section{References}
\label{refer}
%
There are three reference systems available; only one, of course,
should be used for your contribution. With each system (by
number only, by letter-number or by author-year) a reference list
containing all citations in the
text, should be included at the end of your contribution placing the
\LaTeX{} environment \verb|thebibliography| there.
For an overall information on that environment
see the {\em \LaTeX{} User's Guide \& Reference
Manual\/} by Leslie Lamport, p.~71.

There is a special {\sc Bib}\TeX{} style for LLNCS that works along
with the class: \verb|splncs.bst|
-- call for it with a line \verb|\bibliographystyle{splncs}|.
If you plan to use another {\sc Bib}\TeX{} style you are customed to,
please specify the option \verb|[oribibl]| in the
\verb|documentclass| line, like:
\begin{verbatim}
\documentclass[oribibl]{llncs}
\end{verbatim}
This will retain the original \LaTeX{} code for the bibliographic
environment and the \verb|\cite| mechanism that many {\sc Bib}\TeX{}
applications rely on.
%
\subsection{References by Letter-Number or by Number Only}
%
References are cited in the text -- using the \verb|\cite|
command of \LaTeX{} -- by number or by letter-number in square
brackets, e.g.\ [1] or [E1, S2], [P1], according to your use of the
\verb|\bibitem| command in the \verb|thebibliography| environment. The
coding is as follows: if you choose your own label for the sources by
giving an optional argument to the \verb|\bibitem| command the citations
in the text are marked with the label you supplied. Otherwise a simple
numbering is done, which is preferred.
\begin{verbatim}
The results in this section are a refined version
of \cite{clar:eke}; the minimality result of Proposition~14
was the first of its kind.
\end{verbatim}
The above input produces the citation: ``\dots\ refined version of
[CE1]; the min\-i\-mality\dots''. Then the \verb|\bibitem| entry of
the \verb|thebibliography| environment should read:
\begin{verbatim}
\begin{thebibliography}{[MT1]}
.
.
\bibitem[CE1]{clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and boundary-value problems for
Hamiltonian systems.
Arch. Rat. Mech. Anal. {\bfseries 78} (1982) 315--333
.
.
\end{thebibliography}
\end{verbatim}
The complete bibliography looks like this:
%
\begin{thebibliography}{[MT1]}
%
\bibitem[CE1]{clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and
boundary-value problems for Hamiltonian systems.
Arch. Rat. Mech. Anal. {\bfseries 78} (1982) 315--333
%
\bibitem[CE2]{clar:eke:2}
Clarke, F., Ekeland, I.:
Solutions p\'{e}riodiques, du
p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
Note CRAS Paris {\bfseries 287} (1978) 1013--1015
%
\bibitem[MT1]{mich:tar}
Michalek, R., Tarantello, G.:
Subharmonic solutions with prescribed minimal
period for nonautonomous Hamiltonian systems.
J. Diff. Eq. {\bfseries 72} (1988) 28--55
%
\bibitem[Ta1]{tar}
Tarantello, G.:
Subharmonic solutions for Hamiltonian
systems via a $\bbbz_{p}$ pseudoindex theory.
Annali di Matematica Pura (to appear)
%
\bibitem[Ra1]{rab}
Rabinowitz, P.:
On subharmonic solutions of a Hamiltonian system.
Comm. Pure Appl. Math. {\bfseries 33} (1980) 609--633
\end{thebibliography}
%
\subsubsection*{Number-Only System.}
%
For this preferred system do not use the optional argument
in the \verb|\bibitem| command: then, only numbers will
appear for the citations in the text (enclosed in square brackets)
as well as for the marks in your
bibliography (here the number is only end-punctuated without
square brackets).

Subsequent citation numbers in the text are collapsed to ranges.
Non-numeric and undefined labels are handled correctly but no sorting is
done.

E.g., \verb|\cite{n1,n3,n2,n3,n4,n5,foo,n1,n2,n3,?,n4,n5}| -- where
\verb|n|$x$ is the key of the $x^{\mathrm{th}}$ \verb|\bibitem|
command in sequence, \verb|foo| is the key of a \verb|\bibitem| with an
optional argument, and \verb|?| is an undefined reference -- gives
1,3,2-5,foo,1-3,?,4,5 as the citation reference.

\begin{verbatim}
\begin{thebibliography}{1}
\bibitem {clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and boundary-value problems for
Hamiltonian systems.
Arch. Rat. Mech. Anal. {\bfseries 78} (1982) 315--333
\end{thebibliography}
\end{verbatim}
%
\subsection{Author-Year System}
%
References are cited in the text by name and year in parentheses
and should look as follows:
(Smith 1970, 1980), (Ekeland et al. 1985, Theorem 2), (Jones and Jaffe
1986; Farrow 1988, Chap.\,2). If the name is part of the sentence
only the year may appear in parentheses,
e.g.\ Ekeland et al. (1985, Sect.\,2.1)
The reference list should contain all citations occurring in the text,
ordered alphabetically by surname (with initials following). If there
are several works by the same author(s) the references should be listed
in the appropriate order indicated below:
\begin{alpherate}
\setlength{\hfuzz}{5pt}
\item
One author: list works chronologically;
\item
Author and same co-author(s): list works chronologically;
\item
Author and different co-authors: list works alphabetically
according to co-authors.
\end{alpherate}
If there are several works by the same author(s) and in the same year,
but which are cited separately, they should be distinguished by the use
of ``a'', ``b'' etc., e.g.\ (Smith 1982a), (Ekeland et al. 1982b).
%
\subsubsection*{How to Code Author-Year System.}
%
If you want to use this system you have to specify the option
\verb|[citeauthoryear]| in the \verb|documentclass|, like:
\begin{verbatim}
\documentclass[citeauthoryear]{llncs}
\end{verbatim}
Write your citations in the text explicitly except for the year, leaving
that up to \LaTeX{} with the \verb|\cite| command. Then give only the
appropriate year as the optional argument (i.e. the label in square
brackets) with the \verb|\bibitem| command(s).\\[2mm]
{\itshape Sample Input}
\begin{verbatim}
The results in this section are a refined version
of Clarke and Ekeland (\cite{clar:eke}); the minimality result of
Proposition~14 was the first of its kind.
\end{verbatim}
The above input produces the citation: ``\dots\ refined version of
Clarke and Ekeland (1982); the minimality\dots''. Then the
\verb|\bibitem| entry of \verb|clar:eke| in the \verb|thebibliography|
environment should read:
\begin{verbatim}
\begin{thebibliography}{}  % (do not forget {})
.
.
\bibitem[1982]{clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and boundary-value problems for
Hamiltonian systems.
Arch. Rat. Mech. Anal. {\bfseries 78} (1982) 315--333
.
.
\end{thebibliography}
\end{verbatim}
{\itshape Sample Output}
\bibauthoryear
%
\end{document}

### 2013-09-30 WESOA2013/source/paper.tex
inicial --martin

--- 2013-09-30 WESOA2013/source/auxiliar/validation.tex
@@
-\chapter{Evaluation}
-\label{chapter:validation}
+%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is the LaTeX source for the instructions to authors using
+% the LaTeX document class 'llncs.cls' for contributions to
+% the Lecture Notes in Computer Sciences series.
+% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
+%
+% It may be used as a template for your own input - copy it
+% to a new file with a new name and use it as the basis
+% for your article.
+%
+% NB: the document class 'llncs' has its own and detailed documentation, see
+% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\epigraph{``\textit{Let us change our traditional attitude to the construction
-of programs: Instead of imagining that our main task is to instruct a computer
-what to do, let us concentrate rather on explaining to human beings what we want
-a computer to do.}''}{Donald E. Knuth}
+\documentclass[runningheads,a4paper]{llncs}
 
-This chapter describes the experimentation we conducted for
-\correctingText{evaluating} the $\pi$SOD-M methodology. The chapter presents
-three \correctingText{examples}. The first two \correctingText{examples} concern simplified
-service-based applications that focus on the application's logic and explicit business rules. They show
-how $\pi$SOD-M can be used for developing applications and exemplify the
-consideration of non-functional aspects along the whole
-development process. The last \correctingText{example}
-(\textit{GesIMED}~\cite{valeriaThesis} application) \correctingText{aims to perform an
-evaluation of our approach, analyzing the main differences and particularities
-between SOD-M and $\pi$SOD-M.}
+\usepackage{amssymb}
+\setcounter{tocdepth}{3}
+\usepackage{graphicx}
 
-\newText{The objective of the experimentation was to evaluate the features and
-concepts proposed by $\pi$SOD-M, considering the implementation of some
-examples, and perform a comparison with SOD-M. This comparison focuses on
-verifying which of those proposed aspects favored the development of
-service-oriented applications in the presence of non-functional requirements. 
-In this direction, our intention is: (i) analyze if it is possible to model
-non-functional requirements from initial levels of abstraction in the
-service-oriented development, (ii) and verify what are the advantages of using
-the proposed approach, compared with the original proposal of SOD-M.}
+\usepackage{url}
+\urldef{\mailsa}\path|{\ldots,|
+\urldef{\mailsb}\path|\ldots,|
+\urldef{\mailsc}\path|\ldots}@\ldots.com|    
+\newcommand{\keywords}[1]{\par\addvspace\baselineskip
+\noindent\keywordname\enspace\ignorespaces#1}
 
-% O objetivo principal deste capítulo é avaliar as características e conceitos
-% adicionados em piSOD-M e fazer uma comparação com SOD-M, com a intenção de
-% destacar que é possível modelar requisitos não funcionais desde níveis iniciais
-% de abstração no desenvolvimento orientado a serviços.
+\begin{document}
 
-%  used for
-% validating the specification and generation of the models proposed in
-% $\pi$SOD-M. The validation split the thesis into parts, each one can be
-% validated individually. This parts are: (i) the modelling of each $\pi$SOD-M
-% model, (ii) the model transformations and (iii) the results.
+\mainmatter  % start of an individual contribution
+
+% first the title is needed
+\title{Web Services Composition in the presence of Non-Functional Properties: A
+MDD Approach}
+
+% a short form should be given in case it is too long for the running head
+%\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}
+
+% the name(s) of the author(s) follow(s) next
 %
+% NB: Chinese authors should write their first names(s) in front of
+% their surnames. This ensures that the names appear correctly in
+% the running heads and the author index.
 %
-% The validation is performed through three case studies. The first case study
-% aims to perform a qualitative evaluation of our approach, analyzing the main
-% differences and particularities between SOD and $\pi$SOD-M. The others case
-% studies aim to validate and verify the use of $\pi$SOD-M and its
-% environment, analyzing the positive results and the benefits to 
-% model non-functional requirements from early stage of development.
+\author{Pl\'acido A. Souza Neto \and Martin A. Musicante \and\\%
+%
+ Genoveva Vargas-Solar\and Umberto Costa\and Val\'eria de Castro\\}
+%
+%\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
+% (feature abused for this document to repeat the title also on left hand pages)
 
-%The thesis' validation of the  follows from the sub-validations.
+% the affiliations are given next; don't give your e-mail address
+% unless you accept that it will be published
+\institute{\ldots\\
+\mailsa\\
+\mailsb\\
+\mailsc}
 
-%\bigskip
+%
+% NB: a more complex sample for affiliations and the mapping to the
+% corresponding authors can be found in the file "llncs.dem"
+% (search for the string "\mainmatter" where a contribution starts).
+% "llncs.dem" accompanies the document class "llncs.cls".
+%
 
-% The \textit{IEEE Standard Computer Dictionary} \cite{ieeeDict} defines
-% validation as \textit{the process of evaluating a system or component during or
-% at the end of the development process to determine whether it satisfies specified
-% requirements}. If we apply this definition to a methodology: \textit{Validation
-% is the process of evaluating a method during or at the end of the development process to determine whether it
-% satisfies specified requirements and eases the development of a system}. This
-% chapter validates the methodology proposed in this thesis in order to verify
-% whether it promotes a better development of reliable
-% services composition applications.
+%\toctitle{Lecture Notes in Computer Science}
+%\tocauthor{Authors' Instructions}
+\maketitle
 
-% The steps during the thesis were analyzed by application of laboratory case
-% studies with an iterative and incremental approach. Thus, while the $\pi$SOD-M
-% concepts have been defined, the concepts application have been made in each case
-% studies to verify its validity.
+\begin{abstract}
+\ldots.
+\keywords{\ldots}
+\end{abstract}
 
-% The validation process has been performed in a continuous and
-% progressive approach along the definition of $\pi$SOD-M.
+\section{Introduction}
+\ldots
 
-This chapter is organized as follows: section
-\ref{sec:toPublishMusic_inpractice} shows the \textit{To Listen Music}
-\correctingText{example}, section \ref{sec:crimes_inpractice} introduces the
-\correctingText{example} \textit{Crime Map} and section \ref{sec:siw_gesimed}
-introduces the \textit{GesIMED} \correctingText{example} that we used for
-validating the methodology. Section \ref{sec:leassonsLearned} presents the lessons learned
-with the validation process. Finally, Section \ref{sec:conclusion_validation} concludes
-the chapter and discusses on validation results.
+\section{A methodology for web services com\-po\-si\-tions}
 
-% From strategy for
-% validation, the process was carried out with the development resolution step as
-% part of an iterative process and feedback cyclic, allowing solve problems of
-% concepts and its application while the methodology was being defined. Thus, in
-% each step of methodology the validation was applied iteratively
-% on two different case studies.
-% 
-% The process defined in the research method validation was therefore not
-% a linear process, but iterative and incremental, thus providing the opportunity
-% to go strengthen the results and learning from each research step.  
-% 
-% Following, each developed case study is described, indicating both the needs
-% identified, and new contributions emerged from them, which are also shown
-% graphically, using the methodology models. The case studies resolution were
-% performed using the proposed approach.
+In this section we present $\pi$SOD-M. $\pi$SOD-M is a MDA (Model Driven
+Architecture) based methodology. It provides a environment for building service compositions considering
+their non-functional requirements. $\pi$SOD-M extends the SOD-M \cite{valeriaThesis} method by adding
+the concept of \textit{Policy} \cite{Espinosa-OviedoVZC09,Espinosa-Oviedo2011a}
+for representing NFR associated to service-based applications. $\pi$SOD-M also
+proposes the generation of a set of models at different abstraction levels, as
+well as transformations between these models.
 
-% \section{$\pi$SOD-M in Practice}
-% \label{sec:execution}
+$\pi$SOD-M's models represent both the cross-cutting aspects of the application
+being modelled, as well as the constraints associated to services. The systems
+cross-cutting concerns affect functional concerns, such as availability;
+recovery; and persistence aspects. Constraints are restrictions that must be
+respected during the execution of the application, for example the fact that a
+service requires an authentication for executing system functions.
 
-% This section presents, in detail, the design and implementation of two case
-% studies conducted to validate the method proposed in this
-% thesis. Both case studies were developed using the models proposed by
-% $\pi$SOD-M.
+$\pi$-SOD-M supports the construction of service-oriented applications that implement business processes.
+Therefore, it proposes a development process based on the definition of models
+(instances of the meta-modes) and transformations for semi-automatically
+generating different levels of abstraction that describe a service-oriented
+application from abstract platform independent views (CIM and PIM level) to
+platform dependent views and the PSM and implementation levels. We extended the
+Business and Service views of the SOD-M method \cite{CastroMV11}. The Business
+view defines  the concepts for modeling a business process, while the Service
+view defines the concepts  for designing services compositions. Our methodology
+introduces concepts (e.g. NF-requirement, constraint, assertion, contract,
+policy) in the Policy view for describing constraints associated to services and
+non-functional properties associated to service processes.
 
-% Following all models will be developed for this case and then
-% describe the main aspects of the implementation of each of them.
-
-\section{Example 1: To Publish Music}
-\label{sec:toPublishMusic_inpractice}
-
-An organization wants to provide the service-based
- application that monitors the music a person is listening during some periods
- of time and sends the song title to this person's Twitter or Facebook accounts.
- Thus, this social network user will have her status synchronized in Twitter or
- Facebook. The scenario begins by get through the music service Spotify for
- fetching the user's musical status. The social network services are then
- contacted to modify the user's status with the music information. The user can
- also download a specific music, he can proceed with the download process, which
- includes the payment process.
-
-The user can download a music, after having proceed with the
- payment via PayPal or credit card. All services are available via Spotify and
- this application needs to interact with Spotify users, so they can listen
- music, publish them on Facebook and Twitter, or buy music online. The scenario
- to monitor and publish songs has the User and Spotify actors.
-
-The functionality of the application are: \textit{Search music;
-Choose music; Download music; Listen Music; Buy music} and \textit{Publish music}.
-
- \subsection{$\pi$-UseCase Model}
-\label{sec:usecase_topublishMusic}
-
-Figure \ref{fig:publishMusic_usecase} shows the main part of the use case model
-for the \textit{To Publish Music} application\footnote{The complete
-\textit{$\pi$-UseCase} model can be found in appendix \ref{append:diagrams}.}.
-The $\pi$-UseCase model describes the system functions and constraints. Figures
-\ref{fig:toolpublishMusic_usecase} and \ref{fig:toolpublishMusic_usecase2}
-detail the model and configuration properties for the described example
-(actual view taken from the $\pi$SOD-M environment).
-Three restrictions are specified on the use cases described by the
-model, two on the \textit{buy music} use case and a restriction on the
-\textit{pay} use case. These restrictions are value and business constraints.
-\textit{Buy music} restrictions are related to secure connection and to
-ensure that the user needs to have an account in the music network. They are
-marked using \textit{\#\ldots\#}, figures \ref{fig:publishMusic_usecase},
-\ref{fig:toolpublishMusic_usecase} and \ref{fig:toolpublishMusic_usecase2}.
-Notice that those NFRs and constraints will be the base to define assertions and
-contracts by the transformations of models.
-
-\begin{figure}[ht!]    
-\centering
-\includegraphics[width=.99\textwidth]{chapters/methodology/figs/piusecase/application/packages2.pdf}
-\caption{\textit{To Publish Music} $\pi$-UseCase.}
-\label{fig:publishMusic_usecase}
-\end{figure}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create_model-UseCase.pdf}
-\caption{\textit{To Publish Music} $\pi$-UseCase Environment Detail.}
-\label{fig:toolpublishMusic_usecase}
-\end{figure}
-
-Figures \ref{fig:toolpublishMusic_usecase} and
-\ref{fig:toolpublishMusic_usecase2} present the update music model and show how
-the properties are defined in the $\pi$SOD-M environment. This example describes
-the properties over the \textit{update music} (item 4, figure
-\ref{fig:toolpublishMusic_usecase}) use case and the restriction over the
-authentication process (item 3, figure \ref{fig:toolpublishMusic_usecase}). To
-update a music in a social network, the user must be authenticated.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-UseCase.pdf}
-\caption{\textit{To Publish Music}$\pi$-UseCase Environment Detail.}
-\label{fig:toolpublishMusic_usecase2}
-\end{figure}
-
-The modeling of the application properties using the environment $\pi$SOD-M
-are the same presented in the $\pi$-UseCase model, as described in Figure
-\ref{fig:toolpublishMusic_usecase2}. For example, the authentication restriction
- details the type of constraint (\textit{VALUE}), a \textit{description},
- \textit{name} and a \textit{non-functional attribute}
- (\textit{authentication}). These properties are described in all constraints
- (item 2, figure \ref{fig:toolpublishMusic_usecase2}).
-
-Figure \ref{fig:servicesToPublishMusic} shows the service components that must
-interact with the \textit{application} and the interaction among them. There are
-four services that are invoked to produce results for the \textit{Publish Music} application:
-\textit{Spotify, Facebook, Twitter} and \textit{Bank}.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.8\textwidth]{chapters/validation/figs/toPublishMusic}
-\caption{To Publish Music Services.}
-\label{fig:servicesToPublishMusic}
-\end{figure}
-
- \subsection{$\pi$-ServiceProcess Model}
-\label{sec:serviceProcess_topublishMusic}
-
-Figure \ref{fig:publishmusic_piserviceProcessToolModelProperties} shows the
-\textit{$\pi$-ServiceProcess} diagram for the execution of activities based on
-contracts. We present here the part of the diagram produced from the concepts of the figure
-\ref{fig:toolpublishMusic_usecase2}. The complete \textit{$\pi$-ServiceProcess}
-diagram of the application is shown in figure \ref{fig:serviceprocessContract}.
-Notice that it is necessary to represent each {\sc Constraint} modelled in a {\sc Contract} based service activity diagram. Each constraint
-described in the previous model is transformed into a contract in the
-$\pi$-ServiceProcess model.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/implementation/figs/model-ServiceProcess.pdf}
-\caption{\textit{To Publish Music} $\pi$-ServiceProcess.}
-\label{fig:publishmusic_piserviceProcessToolModelProperties}
-\end{figure}
-
-\textit{Publish Twitter} and \textit{publish Facebook} (figure
-\ref{fig:publishmusic_piserviceProcessToolModelProperties}) have pre- and
-post-conditions assertions that are grouped into a contract for each service:
-(i) verify if the user data are correct; (ii) if the user is already logged in
-Spotify; As post-condition, it ensures the complete transaction and
-verifies whether the payment authorization notification was sent to the user and to Spotify.
-There may be different restrictions depending on the services participating in
-the publish process. The user must be logged in Spotify in order to access
-Facebook, and Facebook must send a ``successful login'' acknowledge which is
-verified by the post-condition. For publishing a message, Twitter imposes
-pre-conditions on the length of the message and user data login. In contrast,
-Facebook only sends a confirmation acknowledge. Items 4 and 5 in figure
-\ref{fig:publishmusic_piserviceProcessToolModelProperties} show the
-pre-conditions for ``\textit{publish Twitter}'' concerning \textit{(i)} the format
-of the music file and (\textit{ii}) access authorization credentials, namely
-login and password.
-
-\correctingText{The development of the $\pi$-ServiceProcess refine
-the constraints into contracts for the services as shown
-in Figure \ref{fig:publishmusic_piserviceProcessToolModelProperties}.}
-
-% , about the payment authorization. To \textit{update} services, depending of each
-% service, there may be different restrictions.
-
-%  The Facebook service requires that the user is already logged on
-%  Spotify and these data are the same as Facebook. As post-condition, it ensures
-%  that the Facebook service send a notification of success. The Twitter service
-%   requires pre-conditions over the message length and user privacy data, while
-%   to update the Facebook is a necessary precondition and a confirmation notice
-%   as post-condition. As a precondition for ``\textit{publish twitter}'' (itens
-%   4 and 5, figure \ref{fig:publishmusic_piserviceProcessToolModelProperties}) it is necessary that
-%   \textit{(i)} the music format is correct and (\textit{ii})  the twitter login and password is correct for the update.
-
- \subsection{$\pi$-ServiceComposition Model}
-\label{sec:serviceComposition_topublishMusic}
-
- There are four external {\sc Business
-Collaborators} in this application: \textit{Bank, Spotify, Twitter} and
-\textit{Facebook} (items 3, figure
-\ref{fig:topublishMusic_piserviceCompositionToolModel}). Figure \ref{fig:topublishMusic_piserviceCompositionToolModel} shows the application
-services that consist of a set of functions: \textit{search music, select
-music, buy music, download music, listen music} and \textit{publish music}. The
-\textit{publish music} action calls two service collaborators (Facebook and
-Twitter), and the \textit{buy music} action calls two functions of the Bank
-service collaborator.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/implementation/figs/create-model-Serviceconposition.pdf}
-\caption{\textit{To Publish Music} $\pi$-ServiceComposition Model.}
-\label{fig:topublishMusic_piserviceCompositionToolModel}
-\end{figure}
-
-The \textit{To Publish Music} $\pi$-ServiceComposition model defines three
-policies: \textit{transactionPolicy} (item 4, figure
-\ref{fig:topublishMusic_piserviceCompositionToolModel}), \textit{httpAuthPolicy}
-(item 2) and \textit{authPolicy} (item 6). The \textit{trasactionPolicy}
-verifies the Bank data restrictions over the payment and confirmation process.
-The \textit{httpAuthPolicy} policy specifies rules over the
-Facebook function for the authentication and notification processes. These rules
-are executed during the publish massage function request. The
-\textit{authPolicy} policy implements the open authentication protocol
-required for accessing the Twitter service.
-
-From the \textit{To Publish Music} rules and policies it is possible to model
-and associate non-functional properties to services' compositions,
-\textit{e.g.} the integration of information retrieved from different
-social network services, automatic generation of an integrated view of the
-operations executed in different social networks or for providing security in
-the communication channel when the payment service is called.
-
-From this stage, the \textit{$\pi$-PEWS} model (PSM) depicted in figures
-\ref{fig:pewsRepresentation} and \ref{fig:contractRepresentation} can be
-generated.
-
-%  \subsubsection{$\pi$-PEWS Model}
-% \label{sec:pews_topublishMusic}
-
-% \begin{figure}[ht!]
-% \centering
-% \includegraphics[width=.90\textwidth]{chapters/implementation/figs/create_model-PEWSProperties.pdf}
-% \caption{$\pi$-PEWS Model Properties in $\pi$SOD-M Eclipse Plugin.} 
-% \label{fig:piPEWSToolModelProperties}
-% \end{figure}
-
-% \section{Results}
-% \label{sec:results}
-
-\section{Example 2: Crime Map}
-\label{sec:crimes_inpractice}
-
-The \textit{Crime Map} \correctingText{example} addresses the design and
-implementation of a system that deals with statistical data on criminal activities. The system works
-in conjunction with third-party services, in order to process a search supplied
-by a user. The system must provide a public utility service to inform
-citizens about the real crime situation in a particular location or region.
-Making use of third party services, the application must provide a statistics crime portal. The result depends on user queries and upon a user
-demand results can be presented on a map. The system also interacts with
-micro-blogging, for looking up messages containing specific crime
-information. The presentation of the data concerning criminal activities is
-provided by the \textit{Police service}. The presentation of a locality map
-crime is made by the \textit{Google Maps service} and messages presentation on
-the micro-blogging is done by the \textit{Twitter service}. The system also has
-a \textit{Post service}, for searching an address location that can be then
-presented in a map by Google maps. The scenarios to query data and crime
-statistics have the following actors: \textit{(i) User, (ii) Map service, (iii) Post service, (iv) Micro-blogging
-service} and \textit{(v) Police service}, to query the criminal activities
-(figure \ref{fig:usecaseCrimeMap_total}).
-
-Considering the application functions, the crimes may be searched by types,
-\textit{e.g.} \textit{theft, murder, kidnapping}, etc. The search can also be
-made by city' district or region, for example, \textit{center, west, south,
-north} or \textit{east}, and it can process search temporal criteria, for
-example, crimes of the previous day of the last weeks or months. Finally, the
-service can also process conjunctive queries, \textit{e.g.}, \textit{the number of kidnapping of a specific district in the last month}.
-
-Whenever a new crime is published by the police, the system should post a
- message on Twitter stating the new crime, with a specific hashtag,
- \textit{e.g.} \textit{\#crime043\_kidnapping\_center\_05-08-2012}. The hashtag
- structure contains the crime number, type, local and the day it happens. From
- this, users can comment about the crime and these comment will appear in the
- application interface. All users must have an account to use the application.
-
- The application functionalities are: \textit{Choose search type of crime; see
- crime information; Receive information on the chosen query; See map; List crimes;}
- and \textit{Comment about a crime.}
-
- Non-functional requirements for this application include: \textit{queries
- response performance, crime information validation}, \textit{crime
- information presented in a map} and \textit{ensuring crimes data security.}
-
-\subsection{$\pi$-UseCase Model}
-\label{sec:usecase_crimesMap}
-
-% The modeling of the system starts with the $\pi$-UseCase model. This involves
-% determining the services identified in the list of requirements, which final
-% consumers identified in the list of business services, and which of the will be
-% end consumers or service users. Thus, it is possible to identify the modeling
-% elements that are part of the use case model.
-
- Figure \ref{fig:usecaseCrimeMap_total} presents the Crime Map $\pi$-UseCase
-model. This model includes examples of \textit{business, value}
-and \textit{exceptional behaviours} constraints. The application accesses
-data from third party services to provide the user an easy way to visualize new
-information, these services are: \textit{Map, Police, Postal} and
-\textit{Microb-logging Services}.
-
-The user interacts with the application that will perform the services
-for obtain and publish details about the crimes. The services are modeled as
-actors (they are external applications). Each service has a set of functions
-that can be performed. The $\pi$-UseCase model describes both the external services
-and the functionality of the application (see figure
-\ref{fig:usecaseCrimeMap_total}).
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase}
-\caption{Crime Map $\pi$-UseCase.}
-\label{fig:usecaseCrimeMap_total}
-\end{figure}
-
-The user can search for crimes and view its occurrences in a map. The search may be performed by \textit{type of
-crime, region} or \textit{period}. The user can also perform a search that
-combines these criteria. If the application's user does not choose any
-option, the search will retrieve up to 30 recent occurrences. If the location
-maps service is unavailable or there is no response from it after 5 seconds,
-only the crimes information will be presented, without a map view.
-
-Figure \ref{fig:usecaseCrimeMap_total} also
-presents the $\pi$-UseCase model tree representation generated by the $\pi$SOD-M
-environment model .Figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2} detail the
-use cases and constraints presented in Figure
-\ref{fig:usecaseCrimeMap_total}.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase1}
-\caption{\textit{Search crime} $\pi$-UseCase Detail.}
-\label{fig:usecaseCrimeMap1}
-\end{figure}
-
-The \textit{search crimes} use case (Figure \ref{fig:usecaseCrimeMap1})
-requests the \textit{Police} service to verify the crimes according to the
-parameters supplied by the user. The restriction of this use case depends on
-the parameters selected by the user. If the user chooses one or several types of
-specific crime's type, the system will present those crimes that have more than
-five occurrences. For \textit{search by address}, the address must comply with
-the restrictions of the police service, the research can be done by street,
-neighborhood, city or district. \textit{Search by period} will provide only
-occurrences within the last six months.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/crimes_useCase2}
-\caption{\textit{See crime information} $\pi$-UseCase Detail.}
-\label{fig:usecaseCrimeMap2}
-\end{figure}
-
-Figure \ref{fig:usecaseCrimeMap2} describes the design of the use cases:
-\textit{see crime information, see statistics, show map} and \textit{share
-information}. To share information of a crime, it is necessary to access the
-Twitter micro-blogging service to proceed with the authentication and posting.
-The service requires the user id and password, however the
-application must ensure the User privacy information while performing
-authentication. To see statistics of crimes, the application should process a
-larger volume of data to generate statistical results. Thus, it is required to
-provide resources to ensure the data processing performance. The restriction on
-map presentation is over response time and accessibility of the \textit{Postal}
-and \textit{Google maps} services. The response time limit is 5 seconds. If the
-map is not processed, only the information of the crime are presented. Figures
-\ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2} also show the
-properties used in the $\pi$SOD-M environment and the equivalent concepts in the
-model.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.8\textwidth]{chapters/validation/figs/crimesMap}
-\caption{Crime Map Services.}
-\label{fig:servicesCrimeMap}
-\end{figure}
-
-After the identification the various functions that are required by the system
-to perform the business services, the $\pi$-ServiceProcess model is used to
-represent the workflow necessary to perform a service within the system.
-
-\subsection{$\pi$-ServiceProcess Model}
-\label{sec:serviceprocess_crimesMap}
-
-The $\pi$-ServiceProcess model is used to represent the workflow necessary to
-perform the system service. The service processes execution is described in two
-steps. Figures \ref{fig:serviceProcessCrimeMap} and \ref{fig:serviceProcessCrimeMap2} show the
-flow of functions execution described in the $\pi$-UseCase models shown
-in Figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}. Each
-diagram can represent different business services (see figures
-\ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}).
-
-The diagrams of figures \ref{fig:serviceProcessCrimeMap} and
-\ref{fig:serviceProcessCrimeMap2} were obtained by applying the
-\textit{$\pi$-UseCase2$\pi$-ServiceProcess} transformation to the diagrams of
-figures \ref{fig:usecaseCrimeMap1} and \ref{fig:usecaseCrimeMap2}, respectively.
-
-%  Thus, it makes easier to represent the performance of services. It is also possible to describe the
-% process in a large workflow, but for applications running multiple
-% services, it is better to have smaller processes to better represent the
-% service execution. Thus we present the services business through the different
-% processes as presented in Figures \ref{fig:usecaseCrimeMap1} and
-% \ref{fig:usecaseCrimeMap2}.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceProcess}
-\caption{\textit{Crime information} $\pi$-ServiceProcess.}
-\label{fig:serviceProcessCrimeMap}
-\end{figure}
-
-\begin{figure}[ht!]    
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceProcess_crime1}
-\caption{\textit{Crime information}
-$\pi$-ServiceProcess Environment.}
-\label{fig:toolserviceProcessCrimeMap}
-\end{figure}
-
-Each constraint of figures \ref{fig:serviceProcessCrimeMap} and
-\ref{fig:serviceProcessCrimeMap2} are transformed into assertions. For example,
-the assertions of address format,  crime types and time are restrictions over 
-the action \textit{search crime}, being one pre-condition and two
-post-conditions. These restrictions form the contract on the \textit{search
-crime} function. Similarly, for viewing maps, if the answer to the request
-takes longer than 5 seconds (5000 milliseconds), the map display is suspended and only 
-the information is presented. Each assertion in this $\pi$-ServiceProcess model
-stems from the $\pi$-UseCase model constraints. These restrictions are shown
-in Figure \ref{fig:serviceProcessCrimeMap}. Figure
-\ref{fig:toolserviceProcessCrimeMap} presents the equivalent model
-generated by the methodology transformation in the $\pi$SOD-M environment for
-the process detailed in Figure \ref{fig:serviceProcessCrimeMap}. Each node
-defines the process element and its properties.
-
-In Figure \ref{fig:toolserviceProcessCrimeMap}, \textit{i\_node} corresponds
-to the initial process node. {\sc Actions} are grouped into  {\sc Service
-Activities}, such as \textit{choose crimes property} action that is part of
-the \textit{search crime} service activity. A {\sc Control Flow} is an edge that
-links two nodes, for example \textit{cf1} connect the initial node
-\textit{i\_node} (source) with the \textit{choose crimes property} action
-(target). Figure \ref{fig:toolserviceProcessCrimeMap} also presents the
-assertions described in Figure \ref{fig:serviceProcessCrimeMap} grouped into 
-{\sc Contracts}. The contracts are: \textit{searchCrimeContract} and
-\textit{seeCrimeInformationContract}. These contracts are related with the
-actions \textit{search crime} and \textit{see crimes information}.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceProcess2}
-\caption{\textit{See crime statistic and share information} $\pi$-ServiceProcess
-Detail.}
-\label{fig:serviceProcessCrimeMap2}
-\end{figure}
-
-Figure \ref{fig:serviceProcessCrimeMap2} presents the assertions over the
- \textit{see statistics} and \textit{share information} actions. Both have 2 
-pre-conditions. To \textit{share information} on Twitter it is necessary
-to verify the format of crime information (140 characters) and the Twitter id
-and password, for authentication. Regarding the \textit{see statistics} action, there is a
-business restriction over data volume, and a value restriction over the crime
-information format. Thus, the presentation of the crimes statistics must be done
-after the contract verification. Figure
-\ref{fig:toolserviceProcessCrimeMap2} presents the equivalent model
-generated by the methodology transformation in the $\pi$SOD-M environment for
-the process detailed in Figure \ref{fig:serviceProcessCrimeMap2}. Each node
-defines the process element and its properties. Figure
-\ref{fig:toolserviceProcessCrimeMap2} also presents the assertions grouped
-into three contracts that are related with its specific actions. The contracts
-are: \textit{seeStatisticsContract}, \textit{seeCrimeInformationContract} and
-\textit{shareInformationContract}. The contracts are related with the action
-\textit{see statistics}, \textit{see crimes information} and \textit{share
-information}, respectively.
-
-\begin{figure}[ht!]    
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceProcess_crime2}
-\caption{\textit{Crime statistic and share information}
-$\pi$-ServiceComposition Environment.}
-\label{fig:toolserviceProcessCrimeMap2}
-\end{figure}
-
-These models provide an overview of the business processes from the requirements
-described in the $\pi$-UseCase model. These models also provide a more detailed
-view of the execution of business processes and the possible
-interaction with external services. The description of this interaction is the
-result of the refinement of these models ($\pi$-ServiceProcess models) through
-the $\pi$-ServiceComposition model.
-
-\subsection{$\pi$-ServiceComposition Model}
-\label{sec:servicecomposition_crimesMap}
-
-Figures \ref{fig:serviceCompositionCrimeMap} and
-\ref{fig:serviceCompositionCrimeMap2} show the $\pi$-ServiceComposition models,
-which represent the service composition processes of each
-business service and additional indication of which are members of the business
-that execute the action. Actions are derived from service activities,
-identifying the set of actions that are necessary for the completion of each service.
-
-The main partition expresses the \textit{Application} execution
-(\textit{External false}) that represents the general process. Actions have the equivalent
-service relation, which realize the external service, such as \textit{Police}, \textit{Google Maps} or
-\textit{Twitter} services. They are expressed by a {\sc Business Collaborator}.
-Both, Figure \ref{fig:serviceCompositionCrimeMap} and
-\ref{fig:serviceCompositionCrimeMap2} present two external {\sc Business
-Collaborators} describing the application services that are invoked. These
-$\pi$-ServiceComposition models refine the $\pi$-ServiceProcess models, matching
-action with real service functions that may be executed by the system
-application.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceComposition}
-\caption{\textit{Crime information}
-$\pi$-ServiceComposition.}
-\label{fig:serviceCompositionCrimeMap} 
-\end{figure}
-
-\begin{figure}[ht!]    
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceCOmposition_crime1}
-\caption{\textit{Crime information}
-$\pi$-ServiceComposition Environment.}
-\label{fig:toolserviceCompositionCrimeMap}
-\end{figure}
-
-For each $\pi$-ServiceComposition model, the contracts described in the
-$\pi$-ServiceProcess model are grouped in policies. Figure
-\ref{fig:serviceCompositionCrimeMap} describes the \textit{performancePolicy}
-and \textit{reliabilityPolicy} policies, both associated with a service activity
-and its actions. It is important to highlight that all policies are applied to the
-entire actions associated with a service activity, for example, the
-\textit{performancePolicy} verifies the request time for presenting the Google
-maps. If the time exceeds 5000 milliseconds, the maps call is ignored and only
-the crime information is presented. The \textit{reliabilityPolicy} is
-associated with crime data verification. There are three rules over crime
-address format, types of crime and the period of time in which it happens. The
-crime presentation may obey the pre- and post-condition described in this policy. Figure
-\ref{fig:toolserviceCompositionCrimeMap} presents the $\pi$SOD-M tool
-description for the model described in Figure
-\ref{fig:serviceCompositionCrimeMap}.
-
-\begin{figure}[ht!]   
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/model_piServiceComposition2}
-\caption{\textit{Crime statistic and share information}
-$\pi$-ServiceComposition.}
-\label{fig:serviceCompositionCrimeMap2}
-\end{figure}
-
-\begin{figure}[ht!]    
-\centering
-\includegraphics[width=0.99\textwidth]{chapters/validation/figs/toolServiceCOmposition_crime2}
-\caption{\textit{See crime statistic and share information}
-$\pi$-ServiceComposition Environment Detail.}
-\label{fig:toolserviceCompositionCrimeMap2}
-\end{figure}
-
-Figure \ref{fig:serviceCompositionCrimeMap2} refines the $\pi$-ServiceProcess
-model described in Figure \ref{fig:serviceProcessCrimeMap2} and Figure
-\ref{fig:toolserviceCompositionCrimeMap2} presents the $\pi$SOD-M tool
-description for the model described in Figure
-\ref{fig:serviceCompositionCrimeMap2}. This model details the policies for
-\textit{crime statistic} and \textit{share information} services. The
-policies are \textit{statisticPolicy} and \textit{authenticationPolicy},
-respectively.
-
-These models provide an overview of the external services and its composition
-from the actions described in the $\pi$-ServiceProcess model. These models also
-provide a more detailed view of the system restrictions, applying policies
-over external services. This model refines the $\pi$-ServiceProcess models
-detailing the Business Collaborators that are expressed as external services,
-and grouping contract into policies.
-
-% From these composition models, next
-% step is generate the specification system code in $\pi$-PEWS.
-
-% \subsubsection{$\pi$-PEWS Model}
-% \label{sec:pews_crimesMap}
-
-\begin{figure}[ht!]
-\tiny
-\centering
-
-\begin{lstlisting}[label=list:crimeInformationPEWS,caption=pi-PEWS
-Specification: Crime information. ]
-
-1  ns police =  "http:\\www.pm.rn.gov.br/service"
-2  ns googleMaps =  "http://maps.googleapis.com/maps/api/"
-
-4  alias chooseCrimesProperties = portType/chooseCrimesProperties in police
-5  alias searchCrimes = portType/searchCrimes in police
-6  alias seeMap = portType/showMap in googleMaps
-7  alias seeCrimesInformation = portType/getCrimeDetail in police
-
-9 (chooseCrimesProperties . searchCrimes)* . seeMap . seeCrimesInformation
-
-11 def contract searchCrimesContract{
-12  isAppliedTo: searchCrimes;    
-13  requires: addressFormat == ?? 
-14     (onFailureDo: call(chooseCrimesProperties) );  
-15    ensures : !(getCrimes(types).length > 5) ==> false &&
-16              result[i].period < 1800
-17       (onFailureDo: show(getCrimesInformation())); 
-18 }
-
-20 def contract showMapContract{
-21  isAppliedTo: seeMap;    
-22    ensures : serviceResponse.getTime() < 5000
-23       (onFailureDo: skip); 
-24 }
-\end{lstlisting}
-\label{fig:crimeInformationPEWS}  
-\end{figure}
-
-The result of the transformation of this model to the \textit{$\pi$-PEWS} model
-is gives in the listing \ref{list:crimeInformationPEWS} and
-\ref{list:crimeStatisticsPEWS}.
-
-\begin{figure}[ht!]
-\tiny
-\centering
-\begin{lstlisting}[label=list:crimeStatisticsPEWS,caption=pi-PEWS Specification: Crime
-statistic and share information. ]
-
-1  ns police =  "http:\\www.pm.rn.gov.br/service"
-2  ns twitter =  "https://dev.twitter.com/docs/api/"
-
-5  alias searchCrimes = portType/searchCrimes in police
-7  alias seeCrimesInformation = portType/getCrimeDetail in police
-4  alias seeStatistics = portType/getCrimeStatistics in police
-6  alias shareInformation = portType/publishMessage in twitter
-
-9  searchCrimes . ((seeCrimesInformation . shareInformation) |
-10					seeStatistics)
-
-12 def contract searchCrimesContract{
-13  isAppliedTo: searchCrimes;    
-14  requires: addressFormat == ?? 
-15     (onFailureDo: call(chooseCrimesProperties) );  
-16    ensures : !(getCrimes(types).length > 5) ==> false &&
-17              result[i].period < 1800
-18       (onFailureDo: show(getCrimesInformation())); 
-19 }
-
-21 def contract shareInformationContract{
-22  isAppliedTo: shareInformation;    
-23  requires: name == ?? AND password == ?? &&
-24              crimes[i].Format == ?? 
-25     (onFailureDo: call(shareInformation) );  
-26 }
-
-28 def contract seeStatisticsContract{
-29  isAppliedTo: seeStatistics;    
-30  requires: crimes[i].Format == ?? 
-31     (onFailureDo: call(seeStatistics) );  
-32 }
-\end{lstlisting}
-\label{fig:crimeStatisticsPEWS}  
-\end{figure}
-
-\section{Example 3: GesIMED Application}
-\label{sec:siw_gesimed}
-
-Aiming to perform a qualitative analysis and the validation of $\pi$SOD-M
-methodology, we present the \textit{GesIMED Application}\footnote{If necessary, a detailed description of this case
- study can be found in \cite{valeriaThesis}.}
- case study.
-
-% This case
-% study was described in~\cite{valeriaThesis} with the goal of validating the
-% SOD-M methodology. In
-% this section we present the different produced by using the SOD-M and
-% $\pi$SOD-M, as well as the contributions generated by of our proposal. We will
-% model the same business case for both methods and see the difference between them.
-
-This \correctingText{example} was originally developed in \cite{valeriaThesis} for the SOD-M
-methodology. Here , we adapt the requirements of the system to include NFR. This
-system is a management Web system that processes medical images through the Web \cite{valeriaThesis}. The objective is to
-manage information about the creation and maintenance of scientific studies
-for neuroscience research. The application is designed to be used primarily by
-researchers in neuroscience, such as neurologists, neuropsychologists and
-neuroradiologists who conducting research in this area.
-
-The application requirements are the following: (i) a
-database of medical images that can be accessed by the user; (ii) it has an
-interface for querying the database; (iii)  implement standard  procedures for analyzing
-and processing stored images; and (iv) the images analysis and processing
-results must be also stored, so they can be used in future studies.
-
-The  Medical Image Analysis Laboratory from the Universidad Rey Juan Carlos
-(LAIM) offers three specific services to researchers in neuroscience. The access
-to services has a cost and financial revenues are assigned to LAIM. The services
-are:
-
-\begin{itemize}
-  \item Storage and retrieval medical imaging service (SACim);
-  \item Image processing service (SPim);
-  \item Image visualization service (SVim);
-\end{itemize}
-
-From the functions offered by these services we will detail the application
-characteristics. The application uses these services offered by LAIM for
-modeling the application.
-
-% From this case study description we present the modeling performed in SOD-M, and
-% soon after perform the modeling in $\pi$SOD-M. From the business case
-% \textit{perform image processing}, we will differentiate the characteristics of
-% each methodology.
-
-% \subsection{$\pi$SOD-M Modeling}
-% \label{sec:GesIMED-pisodm}
-\bigskip
-\bigskip
-
-Given the description of the business requirements and the expected
-functionality of the system, $\pi$SOD-M is used for developing this application.
-
-% The modeling of \textit{GesIMED} application in $\pi$SOD-M has
-% the same input considered in SOD-M method, the description of business
-% requirements and system functionality. Thereafter, we follow the development of
-% each model according to specification of the $\pi$SOD-M  methodology. Our goal
-% in this section is to highlight the particularities of $\pi$SOD-M regarding
-% SOD-M.
-
-We will detail the \textit{perform image processing} business service. In $\pi$SOD-M, the \textit{$\pi$-UseCase} model
-describes the application functions, restrictions and its non-functional
-attributes. This business service (\textit{perform image processing}) is 
-sufficiently rich to describe the peculiarities of $\pi$SOD-M
-compared to the original model of SOD-M.
-
-\subsection{\textit{$\pi$-UseCase} Model}
-
-Figure \ref{fig:piUseCase-imageProcessing} shows the \textit{$\pi$-UseCase}
-model. This model defines four constraints: \textit{authenticate, response time,
-data format} and \textit{validate payment card}. These constraints are restrictions over the functions of the
-\textit{GesIMED} application. In the description of each constraint it is
-necessary to describe the constraint type (\textit{business} or \textit{value}).
-The model has one business constraint and three value constraints. It is
-possible to describe constraints for each use case application.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-piuc-01.pdf}
-\caption{$\pi$-UseCase: Perform Image Processing.}
-\label{fig:piUseCase-imageProcessing}
-\end{figure}
-
-% Notice that the other use cases are the same of those that were
-% modeled in SOD-M. The difference is that they are now systematized in packages.
-% Our approach guides the systematic functions through packages, because these
-% packages represent business collaborators.
-
-Besides the constraints, the model details the non-functional attributes
-related to each constraint, for example, the \textit{response time} constraint.
-For value constraints, it is necessary identify candidates variables to be
-verified, through the use of \textit{@}. For example, the
-authentication constraint, has the \textit{@id} and \textit{@password}
-identifiers that must be verified.
-
-% It is possible to identify a real difference between the use case models in
-% SOD-M and $pi$SOD-M, especially with regard of constraints, in a way to describe
-% details of quality requirements.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-pisp-01.pdf}
-\caption{$\pi$-ServiceProcess: Perform Image Processing.}
-\label{fig:piServiceProcess-imageProcessing}
-\end{figure}
-
-The main feature of our approach is the association of non-functional
-requirements with use cases. For example, the user can identify that the
-quality requirement related to a payment process is
-\textit{transactional} or \textit{data privacy}. So, through the process, it is
-possible to refine this information, considering other details provided.
-
-The result of the \textit{$\pi$-UseCase} model is a detailed list of
-constraints, its respective use case, system requirement information and
-non-functional attributes required for each service.
-
-\subsection{\textit{$\pi$-ServiceProcess} Model}
-
-Figure \ref{fig:piServiceProcess-imageProcessing} shows the
-\textit{$\pi$-ServiceProcess} model. This model transforms the use cases into actions,
-and constraints into assertions. The resulting workflow describes pre or
-post-conditions on workflow execution. The assertions are wrappers around the
-action. All the assertions related to a specific action become a
-contract. For example, the \textit{proceed payments service} action has a
-pre- and a post-condition. The pre-condition is a restriction over the payment
-information (\textit{card number, name} and \textit{value}), while the
-post-condition defines the final state after the payment. The other assertions
-of this model are the result of the transformation of each constraint described
-in the \textit{$\pi$-UseCase} model.
-
-%  Notice that the workflow (figure \ref{fig:piServiceProcess-imageProcessing})
-%  is the same as described in SOD-M, with the difference of having assertions
-%  during the process, so that the constraints are verified.
-
- \begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-pisc-01.pdf}
-\caption{$\pi$-ServiceComposition: Perform Image Processing.}
-\label{fig:piServiceComposition-imageProcessing}
-\end{figure}
-
-\subsection{\textit{$\pi$-ServiceComposition} Model}
-
-The \textit{$\pi$-ServiceComposition} model presented in figure
-\ref{fig:piServiceComposition-imageProcessing} refines the 
-\textit{$\pi$-ServiceProcess} model. This model
-describes the real services and their associated collaborators. The information
-of the business collaborator comes from the package description. Contracts with
-the same non-functional requirements become service policies. The policies
-define rules, which are a direct transformation of the assertions in the
-previous model.
-
-% Another difference
-% of our approach to the original approach of SOD-M, is that we do not treat each
-% collaborator as a business process.
-
-\begin{figure}
-\tiny
-\centering
-
-\begin{lstlisting}[label=list:pewsNeuro,caption=pi-PEWS Specification: Perform
-Image Processing. ]
-
-1  ns app =  "http:\\www.neuro.laim.org"
-2  ns LAIM =  "http:\\www.neuro.laim.org/laim-service.wsdl"
-3  ns SACim = "http:\\www.neuro.laim.org/sacim-service.wsdl"
-4  ns SPim =  "http:\\www.neuro.laim.org/spim-service.wsdl"
-
-6  alias authenticate = portType/verifyUserandPassword in LAIM
-7  alias proceedServicePayment = portType/payment in LAIM
-8 alias provideImage = portType/provideImage in app
-9  alias receiveImage = portType/receiveImage in SACim
-10 alias processData = portType/processData in SPim
-11 alias verifyImageFormatAndResult = portType/verifyImageFormatAndResult in
-12 SACim alias getResults = portType/verifyExecutionTipe in LAIM
-13 alias downloadResults = portType/downloadFiles in SACim
-
-15 service performProcessing = 
-16            receiveImage . processData . verifyImageFormatAndResult
-
-18 authenticate . proceedServicePayment . provideImage . 
-19   performProcessing . [act(getResults).time > 5]getResults .
-20   {downloadResults}
-
-21 def contract authenticateContract{
-22  isAppliedTo: authenticate;    
-23  requires: id == ?? && password = ??;
-24     (onFailureDo: call(authenticate) )  
-25 }
-
-27 def contract paymentContract{
-28  isAppliedTo: proceedServicePayment;    
-29  requires: card_name =?? && 
-30  			user_name == ?? && 
-31  			value == ??;  
-32  ensures : isPaid == true ;
-33 }
-
-35 def contract performProcessingContract{
-36  isAppliedTo: performProcessing;    
-37  requires: imageExtension == '.png' || imageExtension == '.pdf'
-38    (onFailureDo: skip );    
-39  ensures : resultValue <= 2 ;
-40 }
-\end{lstlisting}
-\label{fig:pewsNeuro}  
-\end{figure}
-
-Listing \ref{list:pewsNeuro} shows the \textit{$\pi$-PEWS} specification that
-was generated from the \textit{$\pi$-ServiceComposition} model. This
-specification considers the operations (lines 6-16), its business collaborators
-(namespaces in the lines 1-4), the main workflow (lines 18-20) and services
-restrictions (described as contracts in the lines 21-40). The final result of
-our approach is the specification that describes the structure and sequence of
-execution to be performed by an orchestrator. \textit{$\pi$-PEWS}
-specification defines a set of contracts to represent the application restrictions. Thus,
-using $\pi$SOD-M it is possible to refine the models until the code generation.
-
-\section{Lessons Learned}
-\label{sec:leassonsLearned}
-
-This section summarizes the experience we obtained after the use of $\pi$SOD-M
-for the three \correctingText{examples}. \newText{In this section we also present some
-results collected from the evaluation made ​​between the two approaches.}
-
-\subsection{Example 1: To Publish Music}
-
-In this case study we can see that the restrictions on web services can be
-modeled from the early stages of development, not leaving it to the programming
-language, or to the programmer to resolve this type of problem. More
-specifically, this case study had restrictions \textit{performance} and \textit{security}. Since
-the \textit{$\pi$-UseCase} model to the \textit{$\pi$-ServiceComposition} model,
-non-functional requirements and their attributes might be described without
-serious impact on the modeling of other application requirements.
-
-The restrictions on the functions provided by
-the \textit{Spotify} service was basically \textit{performance}. This type of
-restriction is not easy to represent in the implementation, because it relies on
-the network available to the user. We described the
-constraints and refined them into a \textit{ performance} policy for updating
-the music. The security restrictions have been described for the service that provides banking information. The security policy generated was described
-for two functions of the Bank service, \textit{pay} and \textit{send
-confirmation} functions. This type of specification can be refined incrementally
-by applying the modelling steps defined by the $\pi$SOD-M methodology.
-
-\correctingText{It} does not cover all the possible restrictions in
-an application but it shows that it is possible to guarantee
-quality requirements refining information at each iteration of the modeling,
-such as constraints of \textit{performance} and \textit{security}.
-
-The difficulty in modeling this \correctingText{example} was the definition of policies for
-each type of non-functional requirement. For example, the \textit{httpAuthPolicy} policy for
-the authentication on Facebook service has restrictions from
-both types, \textit{performance} and \textit{security}.
-
-\subsection{Example 2: Crime Map}
-
-The most important process when designing a system is the requirement analysis
-process, especially identifying the non-functional requirements and the system
-restrictions. In order to ease the coding process and let the developer focus on
-technical issues, the system architecture must be automatically generated, from
-the system design.
-
-Using $\pi$SOD-M has enabled the explicit specification of functional and
-non-functional requirements and their refinement along the different phases of
-the development process.
-
-% its structure
-% may be automatically generated from the system design. If we carry out with some technical problem while coding, you can refer the document for the
-% programming language you are using and solve it quickly, however problems with
-% architecture or requirements analysis does not take longer to resolve. So, a
-% good methodology is more important than is good language, and we could see this
-% on the validation of $\pi$SOD-M.
-
-In this \correctingText{example} we can see the restrictions on four web services:
-\textit{Police, Google Maps, Twitter} and \textit{Post} service. As these
-services are independent, the constraints of the application being developed
-must comply with the service APIs. If a restriction is inconsistent with a
-function, there is no guarantee that it will be respected. For example, if the police offers crimes information on a
-particular region by type or day, it is not possible to present the restrictions
-of a particular time of day, for example, crimes that happen in the afternoon.
-
-This \correctingText{example} has two \textit{reliability} policies and one \textit{security}
-policy. Policies on data are common, because they are restrictions on
-the service interface. We notice in the development of this \correctingText{example} that
-function restrictions, especially the execution time and time
-response are more difficult to ensure.
-
-\subsection{Example 3: GesIMED}
-
-The $\pi$SOD-M methodology helps to define a detailed description of the
-application, its restrictions and non-functional requirements associated with each service.
-The refinement of the quality requirements of each level helps to improve the
-development detail in order to produce a more reliable application. The NFRs are
-not only described in a general way, but the restrictions are designed to each
-service individually.
-
-Considering the modeling of this \correctingText{example} in SOD-M, the first step of the is
-to list the global requirements (business case). Figure \ref{fig:sodm_usecase}
-shows the use case model for the application. This model is very simple and describes the main application requirements. The actor is represented as the ``neuroscience
-researcher'', which is the final consumer of the services to be implemented.
-Furthermore, the model is represented as use cases, which describes the main
-system business services: ``perform image processing\footnote{During our
-analysis, we will detail the \textit{perform image processing} requirement, as a
-way to better understand the particularities of both methods.}'', ``perform
-image view'' and ``perform query''.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.6\textwidth]{chapters/validation/figs/neuro-useCase.pdf}
-\caption{Use Case Model~\cite{valeriaThesis}.}
-\label{fig:sodm_usecase}
-\end{figure}
-
-The next model is the the extended use cases model. This model is used to model
-the functionality required by the system to perform each business services.
-Business services described in the previous model, become more detailed in the
-extended use case model. In this model are described both the application's
-features such as the possible services to be performed. Figure
-\ref{fig:sodm_extendedusecase} presents the extended use case model for
-\textit{perform image processing} business service. This model is detailed in 6
-use cases that are necessary for the image processing. The use cases are: (i)
-perform image processing; (ii) proceed with the service payment; (iii) provide
-images; (iv) authenticate; (v) get results; and (vi) download results.
-
-\begin{figure}[ht!] 
-\centering
-\includegraphics[width=.89\textwidth]{chapters/validation/figs/neuro-extendedUseCase.pdf}
-\caption{Extended Use Case Model~\cite{valeriaThesis}.}
-\label{fig:sodm_extendedusecase}
-\end{figure}
-
-Once identified the different functionalities that are required by the
-system to perform business services, the service process model is used to
-represent the workflow necessary to perform a business service. SOD-M represents
-each business service in different diagrams, which are shown in the following
-figures. Figure \ref{fig:sp-imageProcessing}, for the business service ``
-\textit{perform image processing}''. The activities represented in these figures
-are derived from the basic use cases represented in the model extended use case.
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sp01.pdf}
-\caption{Image Processing - Service Process Diagram~\cite{valeriaThesis}.}
-\label{fig:sp-imageProcessing}
-\end{figure}
-
-% \begin{figure}[ht!]
-% \centering
-% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-03.pdf}
-% \caption{Image Visualization - Service Process Diagram~\cite{valeriaThesis}.}
-% \label{fig:sp-imageVisualization}
-% \end{figure}
-% 
-% \begin{figure}[ht!]
-% \centering
-% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-02.pdf}
-% \caption{Perform Query - Service Process Diagram~\cite{valeriaThesis}.}
-% \label{fig:sp-query}
-% \end{figure}
-
-The service composition models represent the processes of composition of the
-various actions that are necessary to perform each business service, detailing
-which business collaborator performing each action.
-
- \begin{figure}[ht!]
-\centering
-\includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sc-01.pdf}
-\caption{Image Processing - Service Composition Diagram~\cite{valeriaThesis}.}
-\label{fig:sc-imageProcessing}
-\end{figure}
-
-% \begin{figure}[ht!]
-% \centering
-% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sc-02.pdf}
-% \caption{Image Visualization - Service Composition
-% Diagram~\cite{valeriaThesis}.}
-% \label{fig:sc_example02}
-% \end{figure}
-% 
-% \begin{figure}[ht!]
-% \centering
-% \includegraphics[width=.99\textwidth]{chapters/validation/figs/neuro-sc-03.pdf}
-% \caption{Perform Query - Service Composition Diagra~\cite{valeriaThesis}.}
-% \label{fig:sc_example03}  
-% \end{figure}
-
-Figure \ref{fig:sc-imageProcessing} present the details of the services composition
-for the business service we are modelling. It is important to realize that the
-service processes described previously, are refined by the execution of these service.
-Notice that SOD-M creates a workflow for each business service, independently.
-This workflow is a sequence of services that can be run
-considering the system features. Figure
-\ref{fig:sc-imageProcessing} presents that the workflow
-uses the services offered by LAIM to perform each function, according the SACim, SPim and SVim
-services. It is also important to note that the external
-application services are marked with the stereotype \textit{<<WS>>}.
-
-The result of applying the SOD-M method is a detailed service composition model
-from the business use case, in our specific case the ``\textit{perform image
-processing}'' business service.
-
- %\bigskip
-
-% Notice that the non-functional properties is not modelled by SOD-M, in contract
-% to $\pi$SOD-M models.
+\section{Modeling web services com\-po\-si\-tions with
+$\pi$SOD-M}
 
@@
 to evaluate the benefits, advantages and drawbacks of a research project.
-
-The data collection process itself requires a basic paradigm that traces the
-goals of the collected process, \textit{i.e.} the reasons the data are being
-collected. \cite{basili:1985,Kontio:1996} describe that the need for information
-must be quantified whenever possible and the quantification analyzed as to whether or not it
-satisfies the needs. This quantification of the goals should then be mapped into
-a set of data that can be collected.
-
-Each evaluation \textit{goal} generates a set of questions that attempt to
-define and quantify the specific goal which is at the root of its goal tree. The
-\textit{goal} is only as well defined as the \textit{question} that it
-generates. Each \textit{question} generates a set of \textit{metrics} or
-\textit{distributions of data}.
-
-The paradigm proposed by \cite{basili:1985} defines a set of six steps to the
-data collection:
-
-\begin{enumerate}
-  \item Generate a set of goals;
-  \item Derive a set of questions of interest or hypotheses which quantify those
-  goals;
-  \item Develop a set of data metrics and distributions which provide the
-  information needed to answer the question of interest;
-  \item Define a mechanism for collecting the data;
-  \item Perform a evaluation of the data; 
-  \item Analyze the data collected to answer the question posed.
-\end{enumerate}
-
-\bigskip
 
@@
 
-\hline 1. Verify if the identification of non-functional & 1. What are the
+\hline 1. Verify if the definition of constraints at initial& 1. What are the
 levels of modeling  \\
 
-requirements  at initial levels of development   & presented in the
+ the identification levels of development improves  & presented in the
 methodology?  \\
 
- improves the modeling of reliable web services. &    And how the NFRs are
- modelled?  \\
+   of non-functional requirements. &    And how the
+  constraints are modelled?  \\
 
@@
 
- into specification code, and if it improves &  is used in the methodology? \\
+ into contracts assertions, and if it facilitates &  is used in the
+ methodology?
+ \\
 
-the software quality. & \\
+the identification of non-functional requirements. & \\
 
-& 2. How are modeled the non-functional\\
-& requirements?\\
+& 2. How are modeled the contract assertions?\\
 
@@
 
-3. Verify if it is possible to reuse code  &1. Which level of development the code \\
-generated for requirements non-functional. &  is generated?  and what kind of\\ 
- &code is generated?\\
+3. Verify if it is possible to classify the contracts  &1. Which level of
+development the contracts \\ 
+generated from the constraints. &  are
+generated? and Does the code\\ 
+&is generated?\\
 &\\
-& 3. The code is generated automatically \\
+& 3. The code can be generated automatically \\
 & or manual?\\
@@
 4. Verify how the non-functional requirements & 1. How are classified the
-non-functional \\ 
-are specified in the  methodology. & requirements?  \\
+contracts? \\ 
+could be classified from the contract definition. &  \\
 
@@
 \end{table}
-% 
-% Seguindo a abordagem proposta em Basile, as perguntas referentes a cada objetivo
-% são.
-
-% OBJ1 - QUESTIONS
-% 1 - Quais os níveis de modelagem que a metodologia piSOD-M apresenta? E de que
-% forma os NFR são representados?
-% 2 - Qual o objetivo de cada nível de modelagem
-% da metodologia piSOD-M? 
-% 3 - Que tipo de artefato é gerado em cada nível de desenvolvimento?
-% 
-% OBJ2 - QUESTIONS
-% 1 - Qual abordagem de desenvolvimento é utilizada na metodologia?
-% 2 - Como são modelados os requisitos não-funcionais?
-% 
-% OBJ3 - QUESTIONS
-% 1 - Em qual nível de desenvolvimento o código é gerado? E que tipo de código é
-% gerado? 
-% 2 - O código é gerado de maneira automática ou manual?
-% 
-% OBJ4 - QUESTIONS
-% 1 - Como são classificados os requisitos não-funcionais?
-% 2 - Existe algum tipo de classificação dos RNF?
-
-% Ainda seguindo a abordagem proposta por Basili, identificamos métricas com o
-% objetivo de quantificar as perguntas, de forma a gerar dados que possam ser
-% analisado e avaliados com o objetivo de verificar se os objetivos da avaliação
-% foi satisfeito. As métricas são por listadas por objetivos/questões (por
-% exemplo, #1.2. Esta métrica está relacionada ao objetivo 1, e a questão 2
-% identificados anteriormente), dessa forma as medidas a serem quantificadas são
-% apresentadas na tabela X.
 
@@
 &  of each case study? \\
-\hline \#2.2 & 1. How many RNFs were modeled using the approach?  \\
+\hline \#2.2 & 1. How many contracts were modeled using the approach?  \\
 \hline \#3.1 & 1. How many lines of code were generated for \\
-& specify the non-functional requirement identified? \\
+& specify the contracts identified? \\
 
@@
 
-\hline \#4.1 &  1. How many types of non-functional requirements \\
-& were identified?\\
+\hline \#4.1 &  1. Have been any non-functional requirement identified?\\
 
- &  2. Which non-functional requirements \\
-& were identified?\\
+ &  2. Which non-functional requirements were identified?\\
 
@@
 
-% Metricas
-% 
-% 1.1.1 Quantos níveis a metodologia define?
-% 1.1.2 Quantos modelos são gerados em cada nível?
-% 
-% 1.2.1 Quantas restrições são modelada?
-% 
-% 1.3.1 Quantos artefatos podem ser gerados?
-% 
-% 2.1.1 Qual o tempo médio para o desenvolvimento de cada estudo de caso?
-% 2.2.1 Quantos RNF foram modelados?
-% 
-% 3.1.1 Quantas línhas de código foram geradas para cada requisito não funcional
-% identificado?
-% 3.2.1 Qual a soma total de linha de código geradas?
-% 
-% 4.1.1 e 4.1.2 Quantos tipos de requisitos não-funcionais foram identificados?
-%
-
 \paragraph{Collecting the data:} After the development of metrics and the types
@@
 methodology and yielded the following results.
-
-% As tabelas 9 e 10 apresentam as perguntas e os dados gerados a partir da
-% avaliação de cada exemplo. Todos os exemplos foram modelados de acordo com a
-% metodologia e gerou os seguintes resultados.
 
@@
 
-% A metodologia define 4 níveis de modelagem, no entanto os estudos de caso
-% geraram um número diferente na quantidade de modelos ($\pi$-UseCase,
-% $\pi$-ServiceProcess, $\pi$-ServiceComposition e $\pi$-PEWS). No exemplo 1 foram
-% gerados 1 modelo de cada, No exemplo 2 foram gerados 2 modelos de cada tipo, e
-% por fim, o exemplo 3 gerou 3 modelo de caso de uso e 1 referente aos demais.
-% verificamos que essa variação não tem nenhuma implicação com a identificação dos
-% requisitos.
-
 Regarding the types of restrictions modeled at each level, we find that
@@
 (\textit{\#1.2.1}, \textit{\#3.1.1} and \textit{\#3.2.1}).
-
-% Em relação aos tipos de resquições modeladas em cada nível, verificamos que
-% quanto mais restrições são identificadas implica em um aumento no número de
-% lineas de código. O exemplo defini 5 constraints no modelo de $\pi$-UseCase, 8
-% assertions no modelo de $\pi$-ServiceProcess, 6 regras no modelo de
-% $\pi$-ServiceComposition e 2 contratos no modelo de $\pi$-PEWS
-% (\textit{5c-8a-6r-2co}\footnote{5c = 5 constraints; 8a = 8 assertions; 6r = 6
-% regras; 2co = 2 contratos}). Esse exemplo tem 29 linhas de código de
-% especificação. Enquanto o exemplo 2 apresenta o seguinte dado, 7c-9a-7r-5co,
-% gerando 56 linhas de código de especificação. Percebe-se que o número de
-% restrições aumenta o número de linhas de especificação. O exemplo 3 tem as
-% seguintes resquições, 4c-6a-6r-3co, gerando 40 linhas de código de
-% especificação.
 
@@
 
-% Segundo a análise, verificamos também que o número de linha de especificação em
-% apenas para a definiçào dos contratos, cresce de acordo com o números de regras
-% identificadas. A quantidade de dias gastos para a modelagem também variou de
-% acordo com o número de restrições da aplicação. Dessa forma, o tempo gasto com a
-% identificação e modelagem das restrições do exemplo teve uma implicação direta
-% com o número de dias gastos. O exemplo 2 foi modelago em 10 dias e este
-% apresentou um número maior de restrições, e requisitos não funcionais. Podemos
-% supor que se essas restrições não fossem modeladas desde as etapas iniciais, o
-% tempo gasto com manutenção poderia ser maior. O exemplo 3 teve o menor número de
-% restrições, e consequentemente levou menos tempo para ser modelado. Dessa forma
-% o número de requisitos não-funcionais de uma aplicação implica no tempo gasto
-% no desenvolvimento do projeto.
+\section{Final remarks}
 
-\paragraph{Conclusions:}
-
-We conclude that the identification of NFRs in initial levels of modelling helps
-the development of services, mainly because the restrictions can be
-refined and transformed into code specification. This favors the
-prior identification of errors, and help in the classification of the
-non-functional requirements.
-
-Classify restrictions as constraints, then refine this concept into assertions,
-and then group it into policy improves identifying types of non-functional
-requirements, providing reusability of services and better use of the code
-specification.
-
-We also found that the use of the method provides identification of different
-types of NFRs. These requirements are translated into contracts and can be
-checked at runtime. We can see that the number of lines of code generated
-for each type of non-functional requirement does not have a regular variation,
-and it does not generates apparent side effects.
-
-\subsection{Comparison of the SOD-M and $\pi$SOD-M approaches}
-
-% A tabela X apresenta dados em relacao aos 3 exemplos utilizados na avaliacao. O
-% resultado reflete as diferencaos entre as duas abordagem com a intencao de
-% verificar como a extensao de SOD-M pode melhorar o processo de modelagem de
-% aplicacoes orientadas a servicos na presenta de requisitos nao funcionais. 
-\newText{
-We also perform a comparative analysis between $\pi$SOD-M and SOD-M,
-considering other aspects. Table \ref{tab6:comparison} presents the data related with the
-three examples used for the $\pi$SOD-M evaluation: (i) To Publish Music example (\textit{Ex. 1}),  (ii) Crime Map example (\textit{Ex. 2}), and  (iii) GesIMED
-example (\textit{Ex. 3}). The result reflects the difference between the two
-approaches (SOD-M and $\pi$SOD-M) with the intention of verifying how $\pi$SOD-M
-can improve the modeling process of service-oriented applications in the presence of
-non-functional requirements. }
-
-% Os aspectos analisados foram: (i) a quantidade de use cases necessarios para a
-% modelagem de mais alto nivel tanto dos requisitos como das restricoes da
-% aplicacao; (ii) quantas actions sao necessarias para a representacao do processo
-% de negocio; (iii) quando business collaborators sao necessarios para a
-% execucao completa do workflow; (iv) quantos requisitos nao funcionais foram
-% identificados durante a modelagem da aplicacao; (v) quantas linhas de codigo
-% podem ser geradas a partir do modelo de composicao de servicos; e (vi) quantas
-% transformacoes de modelos sao realizadas durante todo o processo de
-% desenvolvimento.
-\newText{
-The aspects analyzed and presented in table \ref{tab6:comparison} are: (i) the
-number of use cases required for modeling both, the
-restrictions and requirements of the application, (ii) how many actions are required for the
-representation of the business process, (iii) how many business collaborators
-are needed to the complete execution of the workflow, (iv) the number of
-non-functional requirements that were identified during the modeling of the
-application, (v) how many lines of code were generated from the service
-composition model of each example, and (vi) how many model transformations held
-throughout the development process.
-}
-
-% Considerando o numero de casos de uso necessarios para a modelagem dos exemplos
-% apresentados neste capitulo, os exemplo modelados com piSOD-M apresentam mais
-% casos de uso pois e possivel modelar as restricoes da aplicacao atraves de casos
-% de uso esteriotipados com o esteriotipo costraint. Devido a este motivo o
-% exemplo 1 apresenta 9 casos de uso usando SOD-M, enquanto que utilizando a
-% abordagem piSOD-M se tem 13 casos de uso. A diferenca (4 casos de uso)
-% representa a quantidade de restricoes que foi identificada neste nivel atraves
-% do modelo de pi-UseCase. No exemplo 2 temos 8 casos de uso utilizando SOD-M
-% enquanto temos 13 utilizando piSOD-M. Uma diferenca de 5 casos de uso. E por fim
-% o exemplo 3 temos 5 casos de uso utilizando SOD-M, e 9 casos de uso atraves da
-% modelagem com piSOD-M.  piSOD-M apresenta uma quantidade maior de casos de uso
-% pelo fato deste modelo tambem representar as restricoes da aplicacao, mesmo no
-% nivel mais abstrado. Verificamos que apesar, de um esforco relativamente maior
-% de modelagem, a representatividade do modelo no que diz respeito aos requisitos
-% funcionais e nao-funcionais e maior em piSOD-M que em SOD-M.
-
-\newText{
-Considering the number of use cases for modeling the examples presented in this
-chapter, the models from the $\pi$SOD-M approach present more use cases then
-the models in SOD-M, because using $\pi$SOD-M it is possible to model the
-restrictions of the application through stereotyped use cases
-(\textit{constraint}). Due to this reason, example 1 has 9
-use cases using SOD-M, while using $\pi$SOD-M, the model has 13 use cases. The difference (4 use cases) between
-both approaches represents the amount of restrictions that have been identified
-at this level through the $\pi$-UseCase model. In example 2, it has 8 use cases
-using SOD-M while using $\pi$SOD-M it has 13 use cases. A difference of 5 use
-cases. In the third example, it presents 5 use cases using SOD-M, and 9 cases
-through $\pi$SOD-M. $\pi$SOD-M presents a greater number of use cases because
-this model also represent the constraints of the application, even in the most
-abstract level. We found that in spite of a relatively greater effort modeling,
-$\pi$SOD-M describes functional and non-functional aspects, while SOD-M models
-only functional aspects.
-}
-
-\begin{table}
-\centering
-\caption{Comparison: SOD-M and $\pi$SOD-M}
-\label{tab6:comparison}
-
-\begin{tabular}{r|c|c|c||c|c|c|}
-
-\cline{2-7}
-& \multicolumn{3}{ |c|| }{SOD-M} & \multicolumn{3}{ |c| }{$\pi$SOD-M} \\
-
-\cline{2-7} & Ex. 1 & Ex. 2 & Ex. 3 & Ex.
-1 & Ex. 2 & Ex. 3
-\\
-\hline 
-\hline Use Case &9 &8 & 5& 13 & 14 &9\\
-\hline Action &7 & 7 &5 &15 &12 &11\\
-\hline Business  & 4&3 &3 &4 &3 &3\\
-       Collaborator & & & & & &\\
-\hline NFRs & - & - & - &2 &3 &3\\
-	   modelled &  &  &  & & &\\
-\hline Lines & 20 & 19 &20 & 29 & 56&40\\
-	   of Code & & & & & &\\
-\hline Transformations  & 2 & 2 & 2 & 3 & 3 & 3\\
-
-\hline
-
-\end{tabular}
-
-\end{table}
-
-% Em relacao ao numero de actions, a nossa abordagem apresenta um numero maior
-% pois e necessário descrever pre- e pos-condições para a execuçao do workflow. O
-% numero maior de actions em piSOD-M favorece que a execução do workflow obedeça a
-% restrições do sistema. Estas actions são resultado das trasnformações dos casos
-% de uso. Os casos de uso marcados com esteriotipos constraint, sao transformados
-% em actions com esteriotipos assertion. É importante perceber na tabela, que no
-% primeiro exemplo, o numero de action varia de 7 para 15, no segundo de 7 para 12
-% e no terceiro exemplo varia de 5 para 11. A verificação de restrições atravéd de
-% assertions favorece a representacao de requisitos nao-funcionais necessários
-% para a execução do workflow.
-
-\newText{
-In relation to the number of actions, the $\pi$SOD-M approach provides a larger
-number, because it needs to describe pre- and post-conditions for the
-services. The larger number of actions in $\pi$SOD-M favors the verification of
-restrictions during the execution of the workflow. These
-actions are the result of a model transformation. The use cases marked
-with \textit{constraint} stereotype are transformed into actions with
-the \textit{assertion} stereotype. It is important to realize in the table that
-in the first example, the number of action ranges from 7 to 15,  the second 7
-to 12 and in the third example ranges from 5 to 11. The constraint checking
-through assertions favors the representation of non-functional requirements
-necessary for the workflow execution.
-}
-
-% Considerando o número de Business Collaborator, esse valor nao sofre nenhuma
-% variação entre as abordagens, pelo fado dos serviços a serem executados nos 3
-% exemplos e nas 2 abordagens serem os mesmos. Apesar do número de casos de uso e
-% actions variarem isso não implica na variação da quantidade de serviços externos
-% (business collaborator) necessários para a execução da aplicação.
-
-\newText{Considering the number of \textit{Business Collaborator}, this value does not
-suffer any variation between both approaches, because of the services to be
-performed in the 3 examples in both approaches are the same. Although the number
-of use cases and actions vary, it does not imply the variation of the amount of external
-services (\textit{Business Collaborator}) required for execute the application.  
-}
-
-% Como o objetivo de ambas as abordagens são destintos, verificar a quantidade de
-% requisitos não-funcionais não é comparável, no entanto aponta para uma melhoria
-% e evolução da abordagem SOD-M. Usando SOD-M não é possível identificar nenhum
-% requisito não funcional durante a modelagem. No entanto em piSOD-M é possível
-% modelar distintos requisitos não-funcionais e refina-los durante a modelagem,
-% fazendo com que em um nível mais baixo sejam representados como propriedades não
-% funcionais \cite{MylopoulosBook99}. Apesar de não ser comparável pois ambas as
-% abordagens tem focos de desenvolvimento diferentes, a proposta apresentada em
-% piSOD-M abraga um valor razoável ao considerar a modelagem de requisitos não
-% funcionais para aplicaçoes orientadas a serviços.
-
-\newText{
-Since the goal of SOD-M and $\pi$SOD-M approaches are different, analyze the number of
-non-functional requirements modelled is not
-comparable, however points to an evolution and improvement of the SOD-M approach. Using SOD-M, it is not possible
-to identify any non-functional requirement during the modeling. However, using
-$\pi$SOD-M, it is possible to model different non-functional requirements and
-refine them. From this point of view, both approaches are not comparable because
-they have different focus of development, otherwise the proposal presented in
-$\pi$SOD-M adds reasonable value considering the modeling of non-functional
-requirements for service-oriented applications.
-}
-
-% Considerando a especificação das aplicações descritas em cada na lingagem
-% piPEWS, fazemos uma análise da diferença do número de linhas geradas em relação
-% as ambas abordagens. Como a modelagem em SOD-M considera apenas a especificação
-% dos requisitos funcionais, e piSOD-M consideram ambos aspectos, funcionais e não
-% funcionais, a análise compara a especificação piPEWS de cada exemplo,
-% considerando a geração dos contratos (piSOD-M) e a não geração dos contratos
-% (SOD-M) que representam a verificação de aspectios não-funcionais. O número de
-% linha e código cresce em piSOD-M devido ao fato da espeficicação gerar contratos
-% que definem pre e pos-condições sobre os serviços do workflow. O crescimento
-% depende da quantidade de restrições descritas durante a modelagem dos exemplos.
-% Não existe nenhum padrão com relação a diferença de número de linhas de código.
-% Por exemplo,o exemplo 1 a diferença entre as abordagens é de 9 linhas de código,
-% enquanto o exemplo 2 a diferença é de 37 linhas. No exemplo 3 a diferença é de
-% 20 linhas de código, sendo 20 linhas geradas usando a abordagem SOD-M e 40
-% usando a abordagem piSOD-M. É importante notar que o número de requisitos não
-% funcionais modelados tem como consequencia um maior número de linhas de código,
-% perceba-se os exemplos 2 e 3, pois ambos modelam 3 requisitos não-funcionais e
-% isso implica em um número maior de linhas de na especificação em piPEWS.
-
-\newText{
-Regarding the specification code for each example, we analyzed the difference of
-lines of code ($\pi$-PEWS language) generated in both approaches. SOD-M
-only considers the specification of functional aspects, and $\pi$SOD-M consider both aspects, functional and non-functional,
-thereby we compare the $\pi$-PEWS specification for each example, considering the
-generation of the workflow (SOD-M)  and the generation of contracts
-($\pi$SOD-M). The number of lines of specification grows in $\pi$SOD-M due to
-the fact that this approach generates contracts which defines pre- and post-conditions
-for the services. The growth depends on the number of constraints identified
-during the modeling. There is no standard concerning the difference in terms of the number of lines
-generated. For example, in the first example the difference between the
-approaches is 9 lines of code, while in the second example the difference is 37
-lines. In the example 3 the difference is 20 lines of code, where 20 lines were
-generated using the SOD-M approach  and 40 lines using $\pi$SOD-M. We can
-conclude is that the number of non-functional requirements modeled implies a greater number of lines of specification code, to realize the examples
-2 and 3, both has three non-functional requirements and it implied a larger
-number of lines in $\pi$-PEWS specification, 56 e 40 respectively.
-}
-
-% Outro fator a ser analisado é o número de transformações em cada abordagem. Na
-% modelagem de todos os exemplos em SOD-M foram realisadas 2 transformações no
-% nível PIM, enquanto que na modelagem dos exemplos usando piSOD-M foram
-% realizadas 3 trasnformações, 2 no nível PIM e 1 no nível PSM. O número maior de
-% transformações implica em um maior refinamento dos conceitos, e consequentemente
-% na geração de código para uma plataforma específica.
-
-\newText{
-Another factor to be considered is the number of transformations in each
-approach. In the modeling of all the examples in SOD-M were performed two 
-transformations at PIM level, whereas in the modeling of the examples using
-$\pi$SOD-M were performed three transformation, 2 in the PIM level and 1 on PSM
-level. The larger number of transformations in $\pi$SOD-M represents a step of
-refinement and the generation of code for a specific platform.      
-}
-
-% 
-% Apesar dos números apresentados na tabela X mostrarem que existe um esforço
-% maior no uso da metodologia piSOD-M, isso se deve ao fato de que restrições
-% sobre requisitos e serviços saão feitas desde os níveis mais abstratos de
-% modelagem. Isso favorece a identificação prévia e revinamento adequado de
-% aspectos transversais e restrições da aplicação. A metodologia piSOD-M agrega
-% valor e conceitos novos em relação a SOD-M melhorando e facilitando a descrição
-% de requisitos de qualidade das aplicações orientadas a serviços.
-
-\newText{
-Despite the numbers presented in table \ref{tab6:comparison} show that there is a
-larger effort in the use of the $\pi$SOD-M methodology, compared with SOD-M,
-this is due to the fact that non-functional requirements and restrictions on
-services are made from the most abstract levels of modeling. This favors the
-early identification of system restrictions and improves the refinement of
-transversal aspects and constraints of the application. The 
-$\pi$SOD-M methodology adds reasonable value and new concepts regarding SOD-M,
-improving and facilitating the description of quality requirements for
-service-oriented applications. }
-
-\section{Conclusions}
-\label{sec:conclusion_validation}
-
-This chapter presented the use of the $\pi$SOD-M methodology, as a form of
-empirical validation. We described, modeled and implemented three
-\correctingText{example of application development} to use concepts, model
-properties and detail the process development. The methodology and its concepts supported the development of applications that use service
-compositions, mainly in the identification and refinement of quality requirements. The use of $\pi$SOD-M
-and its environment can help to identify the non-functional requirements and
-improve the system modeling and specification process.
-
-We conducted, in this chapter, two types of evaluations, the first followed the
-proposal based on Basili~\cite{basili:1985} for data evaluation. And the second
-was a comparison between $\pi$SOD-M and SOD-M, considering different aspects.
-
-\newText{The development of applications by using SOD-M favors and improves
-the modelling of services in more abstract levels of development and help to
-identify non-functional requirements related with each of these services.}
-
-The transformations of models enables the semi-automatic
-development of applications. Owing to the generation of the
-system specification code from the service composition model, the designer and
-developer can focus on the description and detail of functional and non-functional requirements. The
-generated code contains the expression of the interaction with services and
-their associated namespaces. This specification can be executed after the model
-transformation.
-
-As the process is iterative and incremental, after the generation of each model
-it is possible to make adjustments to the generated model. For example, after
-the $\pi$-UseCase model design and the generation of $\pi$-ServiceProcess 
-model, the designer can make adjustments and improve the model so that it
-becomes more expressive.
+\bibliography{biblio} 
+\bibliographystyle{plain}
+\end{document}

### 2013-09-30 WESOA2013/source/paperSOAP_tosubmit.tex
inicial --martin

--- 2013-09-30 SOAP_Track2013/src/paperSOAP_tosubmit.tex
@@
 
-\numberofauthors{1} %  in this sample file, there are a *total*
+\numberofauthors{8} %  in this sample file, there are a *total*
 % of EIGHT authors. SIX appear on the 'first-page' (for formatting
@@
 and reference architectures to build services' based applications mainly address 
-functional aspects \cite{1,2,decastro1,PapazoglouH06}.  Non-functional aspects
-concerning services' and application's "semantics", often expressed as
+functional aspects \cite{1,2,decastro1,PapazoglouH06,QueirozB12}. 
+Non-functional aspects concerning services' and application's ``semantics'', often expressed as
 requirements and constraints in general purpose methodologies, are not fully

### 2013-10-04 JournalToSubmit2013/1-intro.tex
no message

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
+\[ \mathbb{Intro\ SAC} \]
 
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
+Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.
+Functional properties are not the only crucial aspect in the software development process. 
+Other properties need to be addressed to fit in the application with its context.
+These other aspects are called Non-Functional Properties.
 
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
+Non-Functional Requirements (NFRs) specify those properties that are not addressed by the functional  specification.
+They are often called \textit{qualities} of the software system.
+%They are also referred as ``constraints'', ``quality attributes'', ``quality goals'', ``quality of
+%service requirements'' and ``non-behavioural requirements'' \cite{Stellman2005}.
+Non-Functional Requirements may specify response time, security constraints or quality of the solution, among others.
 
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
+Service-Oriented Computing~\cite{Papazoglou2007}, is a software development paradigm where pre-existing services are combined to produce more complex applications. 
+%The selection of services is usually guided by the functional requirements of the application. 
+%In this scenario, there exists the need to provide support for the specification of non-functional requirements, such as
+%security, reliability, and efficiency. 
+The development of service-based applications can benefit from the inclusion of NFRs to the software process from its early stages.
+Failure to comply with this inclusion means that the final application is obtained from a partial specification, making the deployment a difficult task.
+%Ideally, non-functional requirements
+%would be considered along with all the stages of the software development. 
+The adoption of non-functional specifications from the early states of development
+can help the developer to produce applications that are capable of dealing with
+their context.
+Non-functional properties of service oriented applications have been
+addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
+Different proposals~\cite{Babamir2010,AgarwalLS09,CholletL09,GutierrezRF10,XiaoCZBOLH08,JeongCL09,TsadimasNA12}
+support non-functional requirements in the context of web service development.
 
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
+Most software development methods define software processes that use the notion of refinement.
+Software process begins with the formulation of an abstract specification, which is successively refined to yield the implementation of the system.
+Methods for the development of web service applications are no exception to this rule.
+At least two levels of abstraction can be distinguished: a \textit{Business Level}, including the abstract specification, and a \textit{System Level}, including actual computer programs that implements the system.
 
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
+In the case of web service applications we will distinguish two separate layers of the implementation.
+The \textit{Composition Layer} is the upper layer of the implementation. 
+It defines the workflow of the system, in terms of individual service calls.
+The \textit{Service Layer} defines those services that are called by the composition.
 
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
-\begin{document}
-
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
-
-\begin{abstract}
-This paper presents\ldots
-\end{abstract}
-
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
+\[ \mathbb{Original} \]
 Service oriented computing is at the origin of an evolution in the field of software development. 
@@
 Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
-%*********************************************************************************************************
-Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-\begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export a method for  updating the status of a given user:
-\begin{itemize} 
-%\item{\sf\small get-Status ( usedid ): String ; 
-\item {\sf\small update-Status ( userid, new-status ): String}; 
-\end{itemize}
-\end{itemize}
-%
-%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-%the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
-starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
-
-Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
-
-\begin{itemize}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions.
-They are graphically presented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
-
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
-\centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
-
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-%
-\begin{figure}
-\centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
-
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
-For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
-
-%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-SOD-M Development Environment}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture shown in Figure \ref{fig:sodm}.
-Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[htpb]
-	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
-	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
-   \label{fig:policymanager}
-\end{figure}
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-
-\item	 We used  ATL \footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
-
-\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
-We coded the pews.mt program  that implements the model to text transformation for generating executable code.
-It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
-{\em A-policy} based services' composition execution environment.
-
-%\item Finally, we created a chain execution  to execute the model to text transformation.
-\end{itemize}
-%
-%
-%
-
-%
-As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
-The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-
-The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
-Each activity is seen as the process where the service method call is executed.
-The execution of an activity has four states: prepared, started, terminated, and failure.
-The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
-
-At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
-If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
-%The order of policies has implications on the global order of the rules to be executed.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Validation}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-We validated our approach  by implementing the "To Publish Music" application.
-It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.
-Once the services' composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose.
-Our implementation includes a  code generator that generates executable code for a reliable services' composition.
-The model represents an services' composition expression and its associated policies.
-
-We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
-Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available.
-We also implemented a reference coordination without policies.
-We dealt with the availability exceptions by hand and the authentication protocols were embedded within the activities code.
-During our validation Twitter changed the authentication protocol and the maintenance of the application this implied re-programming the reference application.
-Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.
-
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
-Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
-They also include methods and approaches for modeling non-functional constraints.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
-introduced in \cite{schuldt-etal-TODS} is extended to web services
-for addressing atomicity.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-There are few methodologies and approaches that address the explicit modeling  of non functional
-properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
-
- \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
-A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
-Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
- The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several
- \cite{AbrialLNSS91} B-machines refinements that describe the services and their
-compositions.
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
-%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-%specification, realization, implementation and deployment monitoring and management.
-
- %\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-%represents a composition of services. It classifies services   as public utilities, business services or
-%composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
-
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
-We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
-We extended the SOD-M method, particularly the $\pi$-SCM (services' composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code.
-We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.
-
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
-We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications.
-Having such business rules expressed and then translated and associated to the services' composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.
-
-Programming non-functional properties is not an easy task, so we are defining a set of predefined {\em A-policy} types with the associated use rules for guiding the programmer when she associates them to a concrete application.
-{\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
-\end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-10-04 JournalToSubmit2013/2-modeling.tex
no message

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
-\begin{document}
-
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
-
-\begin{abstract}
-This paper presents\ldots
-\end{abstract}
-
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
-%*********************************************************************************************************
 Consider for instance the following scenario.
@@
 Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
 \begin{figure} [htpb]
@@
 Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
-
-\begin{itemize}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions.
-They are graphically presented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
-
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
-\centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
-
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-%
-\begin{figure}
-\centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
-
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
-For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
-
-%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-SOD-M Development Environment}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture shown in Figure \ref{fig:sodm}.
-Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[htpb]
-	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
-	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
-   \label{fig:policymanager}
-\end{figure}
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-
-\item	 We used  ATL \footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
-
-\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
-We coded the pews.mt program  that implements the model to text transformation for generating executable code.
-It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
-{\em A-policy} based services' composition execution environment.
-
-%\item Finally, we created a chain execution  to execute the model to text transformation.
-\end{itemize}
-%
-%
-%
-
-%
-As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
-The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-
-The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
-Each activity is seen as the process where the service method call is executed.
-The execution of an activity has four states: prepared, started, terminated, and failure.
-The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
-
-At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
-If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
-%The order of policies has implications on the global order of the rules to be executed.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Validation}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-We validated our approach  by implementing the "To Publish Music" application.
-It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.
-Once the services' composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose.
-Our implementation includes a  code generator that generates executable code for a reliable services' composition.
-The model represents an services' composition expression and its associated policies.
-
-We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
-Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available.
-We also implemented a reference coordination without policies.
-We dealt with the availability exceptions by hand and the authentication protocols were embedded within the activities code.
-During our validation Twitter changed the authentication protocol and the maintenance of the application this implied re-programming the reference application.
-Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.
-
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
-Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
-They also include methods and approaches for modeling non-functional constraints.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
-introduced in \cite{schuldt-etal-TODS} is extended to web services
-for addressing atomicity.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-There are few methodologies and approaches that address the explicit modeling  of non functional
-properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
-
- \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
-A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
-Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
- The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several
- \cite{AbrialLNSS91} B-machines refinements that describe the services and their
-compositions.
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
-%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-%specification, realization, implementation and deployment monitoring and management.
-
- %\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-%represents a composition of services. It classifies services   as public utilities, business services or
-%composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
-
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
-We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
-We extended the SOD-M method, particularly the $\pi$-SCM (services' composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code.
-We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.
-
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
-We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications.
-Having such business rules expressed and then translated and associated to the services' composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.
-
-Programming non-functional properties is not an easy task, so we are defining a set of predefined {\em A-policy} types with the associated use rules for guiding the programmer when she associates them to a concrete application.
-{\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
-\end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-10-04 JournalToSubmit2013/3-metamodel.tex
no message

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
-\begin{document}
-
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
-
-\begin{abstract}
-This paper presents\ldots
-\end{abstract}
-
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
-%*********************************************************************************************************
-Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-\begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export a method for  updating the status of a given user:
-\begin{itemize} 
-%\item{\sf\small get-Status ( usedid ): String ; 
-\item {\sf\small update-Status ( userid, new-status ): String}; 
-\end{itemize}
-\end{itemize}
-%
-%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-%the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
-starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
-
-Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
 The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
@@
 %An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
+%An {\em A-policy}  defines a set of constraints over the execution states of the servicesÕ composition and defines strategies to enforce this property at the servicesÕ composition execution time.
 An {\em A-policy} groups a set of rules.
@@
 %\end{figure}
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
-For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
-
-%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-SOD-M Development Environment}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture shown in Figure \ref{fig:sodm}.
-Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[htpb]
-	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
-	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
-   \label{fig:policymanager}
-\end{figure}
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-
-\item	 We used  ATL \footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
-
-\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
-We coded the pews.mt program  that implements the model to text transformation for generating executable code.
-It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
-{\em A-policy} based services' composition execution environment.
-
-%\item Finally, we created a chain execution  to execute the model to text transformation.
-\end{itemize}
-%
-%
-%
-
-%
-As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
-The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-
-The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
-Each activity is seen as the process where the service method call is executed.
-The execution of an activity has four states: prepared, started, terminated, and failure.
-The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
-
-At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
-If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
-%The order of policies has implications on the global order of the rules to be executed.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Validation}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-We validated our approach  by implementing the "To Publish Music" application.
-It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.
-Once the services' composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose.
-Our implementation includes a  code generator that generates executable code for a reliable services' composition.
-The model represents an services' composition expression and its associated policies.
-
-We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
-Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available.
-We also implemented a reference coordination without policies.
-We dealt with the availability exceptions by hand and the authentication protocols were embedded within the activities code.
-During our validation Twitter changed the authentication protocol and the maintenance of the application this implied re-programming the reference application.
-Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.
-
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
-Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
-They also include methods and approaches for modeling non-functional constraints.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
-introduced in \cite{schuldt-etal-TODS} is extended to web services
-for addressing atomicity.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-There are few methodologies and approaches that address the explicit modeling  of non functional
-properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
-
- \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
-A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
-Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
- The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several
- \cite{AbrialLNSS91} B-machines refinements that describe the services and their
-compositions.
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
-%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-%specification, realization, implementation and deployment monitoring and management.
-
- %\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-%represents a composition of services. It classifies services   as public utilities, business services or
-%composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
-
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
-We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
-We extended the SOD-M method, particularly the $\pi$-SCM (services' composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code.
-We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.
-
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
-We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications.
-Having such business rules expressed and then translated and associated to the services' composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.
-
-Programming non-functional properties is not an easy task, so we are defining a set of predefined {\em A-policy} types with the associated use rules for guiding the programmer when she associates them to a concrete application.
-{\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
-\end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-10-04 JournalToSubmit2013/4-implementation.tex
no message

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
-\begin{document}
-
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
-
-\begin{abstract}
-This paper presents\ldots
-\end{abstract}
-
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
-%*********************************************************************************************************
-Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-\begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export a method for  updating the status of a given user:
-\begin{itemize} 
-%\item{\sf\small get-Status ( usedid ): String ; 
-\item {\sf\small update-Status ( userid, new-status ): String}; 
-\end{itemize}
-\end{itemize}
-%
-%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-%the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
-starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
-
-Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
-
-\begin{itemize}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions.
-They are graphically presented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
-
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
-\centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
-
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-%
-\begin{figure}
-\centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
-
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
 This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
@@
 Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.
-
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
-Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
-They also include methods and approaches for modeling non-functional constraints.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
-introduced in \cite{schuldt-etal-TODS} is extended to web services
-for addressing atomicity.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-There are few methodologies and approaches that address the explicit modeling  of non functional
-properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
-
- \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
-A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
-Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
- The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several
- \cite{AbrialLNSS91} B-machines refinements that describe the services and their
-compositions.
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
-%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-%specification, realization, implementation and deployment monitoring and management.
-
- %\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-%represents a composition of services. It classifies services   as public utilities, business services or
-%composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
-
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
-We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
-We extended the SOD-M method, particularly the $\pi$-SCM (services' composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code.
-We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.
-
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
-We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications.
-Having such business rules expressed and then translated and associated to the services' composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.
-
-Programming non-functional properties is not an easy task, so we are defining a set of predefined {\em A-policy} types with the associated use rules for guiding the programmer when she associates them to a concrete application.
-{\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
-\end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-10-04 JournalToSubmit2013/5-relworks.tex
no message

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
-\begin{document}
-
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
-
-\begin{abstract}
-This paper presents\ldots
-\end{abstract}
-
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
-%*********************************************************************************************************
-Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-\begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export a method for  updating the status of a given user:
-\begin{itemize} 
-%\item{\sf\small get-Status ( usedid ): String ; 
-\item {\sf\small update-Status ( userid, new-status ): String}; 
-\end{itemize}
-\end{itemize}
-%
-%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-%the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
-starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
-
-Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
-
-\begin{itemize}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions.
-They are graphically presented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
-
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
-\centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
-
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-%
-\begin{figure}
-\centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
-
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
-For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
-
-%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-SOD-M Development Environment}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture shown in Figure \ref{fig:sodm}.
-Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[htpb]
-	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
-	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
-   \label{fig:policymanager}
-\end{figure}
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-
-\item	 We used  ATL \footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
-
-\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
-We coded the pews.mt program  that implements the model to text transformation for generating executable code.
-It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
-{\em A-policy} based services' composition execution environment.
-
-%\item Finally, we created a chain execution  to execute the model to text transformation.
-\end{itemize}
-%
-%
-%
-
-%
-As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
-The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-
-The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
-Each activity is seen as the process where the service method call is executed.
-The execution of an activity has four states: prepared, started, terminated, and failure.
-The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
-
-At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
-If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
-%The order of policies has implications on the global order of the rules to be executed.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Validation}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-We validated our approach  by implementing the "To Publish Music" application.
-It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.
-Once the services' composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose.
-Our implementation includes a  code generator that generates executable code for a reliable services' composition.
-The model represents an services' composition expression and its associated policies.
-
-We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
-Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available.
-We also implemented a reference coordination without policies.
-We dealt with the availability exceptions by hand and the authentication protocols were embedded within the activities code.
-During our validation Twitter changed the authentication protocol and the maintenance of the application this implied re-programming the reference application.
-Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.
-
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
 Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
@@
 Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
-
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
-We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
-We extended the SOD-M method, particularly the $\pi$-SCM (services' composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code.
-We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.
-
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
-We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications.
-Having such business rules expressed and then translated and associated to the services' composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.
-
-Programming non-functional properties is not an easy task, so we are defining a set of predefined {\em A-policy} types with the associated use rules for guiding the programmer when she associates them to a concrete application.
-{\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
-\end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-10-04 JournalToSubmit2013/6-conclusions.tex
no message

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
-\begin{document}
-
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
-
-\begin{abstract}
-This paper presents\ldots
-\end{abstract}
-
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
-%*********************************************************************************************************
-Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-\begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export a method for  updating the status of a given user:
-\begin{itemize} 
-%\item{\sf\small get-Status ( usedid ): String ; 
-\item {\sf\small update-Status ( userid, new-status ): String}; 
-\end{itemize}
-\end{itemize}
-%
-%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-%the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
-starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
-
-Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
-
-\begin{itemize}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions.
-They are graphically presented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
-
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
-\centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
-
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-%
-\begin{figure}
-\centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
-
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
-For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
-
-%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-SOD-M Development Environment}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture shown in Figure \ref{fig:sodm}.
-Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[htpb]
-	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
-	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
-   \label{fig:policymanager}
-\end{figure}
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-
-\item	 We used  ATL \footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
-
-\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
-We coded the pews.mt program  that implements the model to text transformation for generating executable code.
-It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
-{\em A-policy} based services' composition execution environment.
-
-%\item Finally, we created a chain execution  to execute the model to text transformation.
-\end{itemize}
-%
-%
-%
-
-%
-As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
-The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-
-The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
-Each activity is seen as the process where the service method call is executed.
-The execution of an activity has four states: prepared, started, terminated, and failure.
-The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
-
-At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
-If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
-%The order of policies has implications on the global order of the rules to be executed.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Validation}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-We validated our approach  by implementing the "To Publish Music" application.
-It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.
-Once the services' composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose.
-Our implementation includes a  code generator that generates executable code for a reliable services' composition.
-The model represents an services' composition expression and its associated policies.
-
-We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
-Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available.
-We also implemented a reference coordination without policies.
-We dealt with the availability exceptions by hand and the authentication protocols were embedded within the activities code.
-During our validation Twitter changed the authentication protocol and the maintenance of the application this implied re-programming the reference application.
-Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.
-
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
-Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
-They also include methods and approaches for modeling non-functional constraints.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
-introduced in \cite{schuldt-etal-TODS} is extended to web services
-for addressing atomicity.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-There are few methodologies and approaches that address the explicit modeling  of non functional
-properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
-
- \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
-A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
-Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
- The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several
- \cite{AbrialLNSS91} B-machines refinements that describe the services and their
-compositions.
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
-%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-%specification, realization, implementation and deployment monitoring and management.
-
- %\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-%represents a composition of services. It classifies services   as public utilities, business services or
-%composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
-
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
 This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
@@
 {\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
-
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
-
-\end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-10-04 JournalToSubmit2013/1-intro.tex
no message
n
intro...

@@
-\[ \mathbb{Intro\ SAC} \]
+Model Driven Development (MDD) is a top-down approach for the development of software systems. 
+The main ideas of MDD were originaly proposed by the Object Management Group (OMG)~\cite{mda} and provides a set of guidelines for the structuring of specifications.
+The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
+Abstraction levels are called \textit{viewpoints}. 
+MDD defines three viewpoints:
+
+\begin{trivlist}
+\item \textbf{Computation Independent Models (CIM):} This level focusses on the
+environment of the system, as well as on its business and requirement specifications. 
+This viewpoint represents the software system at its highest level of abstraction. 
+At this moment of the development, the structure and system processing details are still unknown or undetermined.
+
+\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
+The specification defines those parts of the system that do not change from one platform to another.
+
+\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
+Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
+\end{trivlist}
+
+Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} within and between levels.
+Intra-level transformations are used to provide a unified representation of concepts of a given level.
+Inter-level transformation implement the refinement process between levels.
+Transformations may be automatic or semi-automatic.
+
+Service oriented computing is at the origin of an evolution in the field of software development. 
+An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually developed.
+(Moreover, IT systems need to evolve according to the business needs.)
+Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs~\cite{bell}. 
+The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications~\cite{5}.
+In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals~\cite{watson}.
+
+Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects~\cite{1,2,decastro1,PapazoglouH06}.
+Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
+This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
+
+\bigskip
+
+In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
+combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
+Some methodologies and techniques have been proposed to help the software
+developer in the specification of functional requirements of the business logic,
+such as the Service Oriented Development Method
+(SOD-M)~\cite{decastro1}. 
+SOD-M is based on MDD and proposes
+models, practices and techniques to aid in software development. SOD-M does not
+provide suport to the specification of non-functional requirements, such as
+security, reliability, and efficiency. Ideally, non-functional requirements
+would be considered along with all the stages of the software development. The
+adoption of non-functional specifications from the early states of development
+can help the developer to produce applications that are capable of dealing with
+the application context.
+
+\bigskip
+
+Non-functional properties of service oriented applications have been
+addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
+Dealing with these kind of properties involves the use of specific technologies
+in different layers of the SOC architecture, for instance during the description
+of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
+service coordinations (like WS-BPEL~\cite{bpel03}).
+
+%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
+%\begin{trivlist}
+%\item[-] Those coming from the Business Process Domain and from the Web
+%service standards that propose ad-hoc protocols. 
+%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
+%\item[-] Those adopting a classic middleware approach were non-functional
+%properties are provided as middleware services, like the ones presented
+%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
+%\item[-] Those providing languages and enabling the specification of protocols
+%used for expressing service
+%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
+%\item[-] Those adopting separation of concerns,
+%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
+%\end{trivlist}
+
+Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
+They also assume that each service exports its interface.
+So, the challenge of supporting non-functional properties is related to
+\textit{(i)} The specification of the business rules of the application; and 
+\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
+
+%In this context, there is a need of 
+%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
+%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
+%
+%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
+%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
+%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
+%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
+
+The main goals of our work are:
+\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
+\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
+\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
+
+\[ \mathrm{Intro\ SAC} \]
 
@@
 
-\[ \mathbb{Original} \]
+\[ \mathrm{Original} \]
 Service oriented computing is at the origin of an evolution in the field of software development. 

### 2013-10-07 JournalToSubmit2013/1-intro.tex
intro...

@@
 Model Driven Development (MDD) is a top-down approach for the development of software systems. 
-The main ideas of MDD were originaly proposed by the Object Management Group (OMG)~\cite{mda} and provides a set of guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
-Abstraction levels are called \textit{viewpoints}. 
-MDD defines three viewpoints:
+The main ideas of MDD were originally proposed by the Object Management Group (OMG)~\cite{mda}, as a set of guidelines for the structuring of specifications.
+The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
 
@@
 Intra-level transformations are used to provide a unified representation of concepts of a given level.
-Inter-level transformation implement the refinement process between levels.
+Inter-level transformation implement a refinement process between levels.
 Transformations may be automatic or semi-automatic.
 
-Service oriented computing is at the origin of an evolution in the field of software development. 
+MDD techniques has been successfully used for the development of hardware and software systems~\cite{MDDvariosAqui}. 
+In particular, we are interested in the application of MDD to the design and implementation of web service applications.
+
+Service oriented~\cite{Papazoglou2007} computing is at the origin of an evolution in the field of software development~\cite{??}. 
 An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually developed.
 (Moreover, IT systems need to evolve according to the business needs.)
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs~\cite{bell}. 
+Thus, organizations are seeking for mechanisms to bridge the gap between the systems developed and business needs~\cite{bell}. 
 The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications~\cite{5}.
-In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals~\cite{watson}.
+In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
 
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects~\cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-\bigskip
-
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
-Some methodologies and techniques have been proposed to help the software
-developer in the specification of functional requirements of the business logic,
-such as the Service Oriented Development Method
-(SOD-M)~\cite{decastro1}. 
-SOD-M is based on MDD and proposes
-models, practices and techniques to aid in software development. SOD-M does not
-provide suport to the specification of non-functional requirements, such as
-security, reliability, and efficiency. Ideally, non-functional requirements
-would be considered along with all the stages of the software development. The
-adoption of non-functional specifications from the early states of development
+In Service-Oriented Computing, pre-existing services are
+combined to produce applications and provide the business logic. 
+The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}. 
+Non-functional aspects of the services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the beginning of the (service) software process.
+Most methods considers them only after the application has been implemented, in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). 
+This leads to service based applications that are partly specified and, thereby, partly compliant with the requirements of the application.
+Ideally, non-functional requirements should be considered along with all the stages of the software development. 
+The adoption of non-functional specifications from the early states of development
 can help the developer to produce applications that are capable of dealing with
@@
 
-\bigskip
+In this work, we are interested in the extension of the \textit{Service Oriented Development Method} (SOD-M)~\cite{decastro1}, to support non-functional aspects, from the early stages of software development.
+SOD-M is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications.
+SOD-M does not provide support for the specification of non-functional requirements, such as
+security, reliability, and efficiency.
+
+The main goals of our work are:
+\begin{trivlist}
+\item \textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
+\item \textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
+\item \textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
+\end{trivlist}
+
+The remainder of the paper is organized as follows. 
+Section\dots
+
+\subsection*{The rest of this section is just for draft\dots}
 
@@
 %\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-The main goals of our work are:
-\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
 

### 2013-10-07 JournalToSubmit2013/1-intro.tex
no message

@@
 The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}. 
+(Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.)
+Functional properties are not the only crucial aspect in the software development process. 
+Other properties need to be addressed to fit in the application with its context.
+These other aspects are called Non-Functional Properties.
+
 Non-functional aspects of the services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the beginning of the (service) software process.
-Most methods considers them only after the application has been implemented, in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). 
+Most methods consider them only after the application has been implemented, in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). 
 This leads to service based applications that are partly specified and, thereby, partly compliant with the requirements of the application.
@@
 
-The remainder of the paper is organized as follows. 
+The rest of the paper is organized as follows. 
 Section\dots
 
+\COMMENT{
 \subsection*{The rest of this section is just for draft\dots}
@@
 Section \ref{sec:conclusions} concludes the paper and discusses future work.
+
+}

### 2013-10-07 JournalToSubmit2013/2-modeling.tex
Genoveva 1452 section 2

@@
-Consider for instance the following scenario.
+%Consider for instance the following scenario.
 An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
 Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-\begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export a method for  updating the status of a given user:
-\begin{itemize} 
-%\item{\sf\small get-Status ( usedid ): String ; 
-\item {\sf\small update-Status ( userid, new-status ): String}; 
-\end{itemize}
-\end{itemize}
+%For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
+%\begin{itemize}
+%\item The music service   Spotify exports a method for obtaining information  %about the music a given user is listening:
+%%\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; %%\end{itemize}
+%\item Facebook and Twitter services export a method for  updating the status of a given user:
+%\begin{itemize} 
+%%\item{\sf\small get-Status ( usedid ): String ; 
+%\item {\sf\small update-Status ( userid, new-status ): String}; 
+%\end{itemize}
+%\end{itemize}
 %
@@
 
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
-starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
+%Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
+%The "To Publish Music" scenario 
+%%(showed by doted line) 
+%starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
+%Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
+%\begin{figure}
+%\centering
+%\includegraphics[width=0.55\textwidth]{figs/SC}
+%\caption{BPMN model of the "To Publish Music" scenario}
+%\label{fig:E3valuemodel}
+%\end{figure}
 
@@
 
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
+%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
 
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
+Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call a service in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
+% song title disseminated in the walls of the user's Facebook and Twitter accounts.
 

### 2013-10-07 JournalToSubmit2013/1-intro.tex
no message

@@
 
-Service oriented~\cite{Papazoglou2007} computing is at the origin of an evolution in the field of software development~\cite{??}. 
+Service oriented computing~\cite{Papazoglou2007} is at the origin of an evolution in the field of software development~\cite{??}. 
 An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually developed.
@@
 Thus, organizations are seeking for mechanisms to bridge the gap between the systems developed and business needs~\cite{bell}. 
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications~\cite{5}.
+The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful service based applications~\cite{5}.
 In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
@@
 }
+
+\bigskip
+\textsc{The next two sections need to be relocated --Martin}
+\bigskip
+
+\section{SOD-M}\label{sec:sodm}
+The Service-Oriented Development Method (SOD-M)~\cite{decastro1} is a MDD approach for service-based applications.
+SOD-M provides a framework with models and standards to express functionalities of applications at a high-level of abstraction. 
+SOD-M meta-models are organized into three levels: CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+
+Two models are defined at the CIM level: \textit{value model} 
+and \textit{BPMN model}. 
+The PIM level models the entire structure of the application flow,
+while, the PSM level provides transformations towards more specific platforms.
+The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
+\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
+These three levels have no support for describing non-functional requirements.
+
+The SOD-M approach includes transformations between models:
+\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
+an abstract model at the CIM level, it is possible to apply transformations for
+generating a model of the PSM level. In this context, it is necessary to
+follow the process activities described by the methodology.
+
+SOD-M considers two points of view:
+\textit{(i)} \textit{business}, focusing on the characteristics and requirements
+of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
+features and processes to be implemented in order application requirements. In
+this way, SOD-M aims to simplify the design of service-oriented applications, as
+well as its implementation using current technologies.
+
+\section{$\pi$SOD-M}\label{sec:pisodm}
+
+%In this section we present $\pi$SOD-M, an MDD based methodology. 
+$\pi$SOD-M provides an environment for building service compositions considering
+their non-functional requirements. 
+$\pi$SOD-M extends the SOD-M meta-models by adding
+the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
+to represent non-functional requirements.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/piSODM}
+\caption{$\pi$SOD-M.}
+\label{fig:piSOD-M}
+\end{figure}
+
+$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
+well as transformations between these models.
+$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
+Constraints are restrictions that must be verified during the execution of the application. 
+An example of this is the requirement of the user's authentication for executing some system functions.
+
+Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
+$\pi$SOD-M proposes a development process based on the definition of models
+(instances of the meta-modes) and transformations between models.
+There are two kinds of transformations:
+Model-to-model transformations are used during the software process to refine the specification.
+Model-to-text transformations are the last step of the process and generate code.
+
+We extended SOD-M to include non-functional specifications.
+Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
+The former three are extensions of SOD-M meta-models and belong to the PIM level.
+The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
+
+The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
+Non-functional requirements are defined as \textit{constraints} over processing and data. 
+The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
+The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
+described in the \textit{$\pi$-UseCase} model into contracts that are associated
+with services. 
+The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
+which put together contracts with similar non-functional requirements. 
+For instance, security and privacy restrictions may be grouped into a security policy.
+\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
+
+At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
+The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
+\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
+Although PEWS is our language of choice, other composition languages can be used as target.
+This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
+
+In the next section we develop an example, to serve as a proof-of-concept.
+The example will show the actual notation used for models.

### 2013-10-07 JournalToSubmit2013/1-intro.tex
intro --mam

@@
 The Service-Oriented Development Method (SOD-M)~\cite{decastro1} is a MDD approach for service-based applications.
-SOD-M provides a framework with models and standards to express functionalities of applications at a high-level of abstraction. 
-SOD-M meta-models are organized into three levels: CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+SOD-M provides models and notation to specify service applications at different levels of abstraction. 
+As usual for MDD, SOD-M meta-models are organized into three levels: CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
 
@@
 and \textit{BPMN model}. 
-The PIM level models the entire structure of the application flow,
+\textsc{Explain these two here --Martin.}
+
+Models at the PIM level describe the structure of the application flow,
 while, the PSM level provides transformations towards more specific platforms.
-The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
-\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
+The PIM-level models proposed by SOD-M are: 
+\begin{trivlist}
+\item \textit{use case}: 
+\item \textit{extended use case}:
+\item \textit{service process}:
+\item \textit{service composition}. 
+\end{trivlist}
+
+SOD-M defines three meta-models at the PSM level: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
 These three levels have no support for describing non-functional requirements.

### 2013-10-07 JournalToSubmit2013/2-modeling.tex
Genoveva 1701

@@
 %This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
+In order to do, our work organizes non-functional constraints into 
+three layers representing: application modeling, services composition and services.
+The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements.
+At the application layer NFP can refer to business rules and values.
+A value NFR expresses constraints about the way data and functions can be accessed and executed.
+For example accessing methods under security protocols.
+
+Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
+For example, response time, storage capacity (e.g., Dropbox service provides 5Giga free storage).
+Value constraints concern more on the conditions in which services can be used.
+For example, accessing to a function within an authentica-tion protocol.
+
+Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
+For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
+
+As a first step in our approach,  we started modeling non-functional constraints at the PSM level.
 Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).

### 2013-10-11 JournalToSubmit2013/intro.tex
reorg - martin

--- 2013-10-07 JournalToSubmit2013/1-intro.tex
@@
 \begin{trivlist}
-\item \textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\item \textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\item \textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
+\item \textit{(i)} To define a NFR model including a set of concepts need for the modeling of NFR in service-oriented applications.
+\item \textit{(ii)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
+\item \textit{(iii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
+\item \textit{(iv)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
 \end{trivlist}
@@
 Section\dots
-
-\COMMENT{
-\subsection*{The rest of this section is just for draft\dots}
-
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-\[ \mathrm{Intro\ SAC} \]
-
-Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.
-Functional properties are not the only crucial aspect in the software development process. 
-Other properties need to be addressed to fit in the application with its context.
-These other aspects are called Non-Functional Properties.
-
-Non-Functional Requirements (NFRs) specify those properties that are not addressed by the functional  specification.
-They are often called \textit{qualities} of the software system.
-%They are also referred as ``constraints'', ``quality attributes'', ``quality goals'', ``quality of
-%service requirements'' and ``non-behavioural requirements'' \cite{Stellman2005}.
-Non-Functional Requirements may specify response time, security constraints or quality of the solution, among others.
-
-Service-Oriented Computing~\cite{Papazoglou2007}, is a software development paradigm where pre-existing services are combined to produce more complex applications. 
-%The selection of services is usually guided by the functional requirements of the application. 
-%In this scenario, there exists the need to provide support for the specification of non-functional requirements, such as
-%security, reliability, and efficiency. 
-The development of service-based applications can benefit from the inclusion of NFRs to the software process from its early stages.
-Failure to comply with this inclusion means that the final application is obtained from a partial specification, making the deployment a difficult task.
-%Ideally, non-functional requirements
-%would be considered along with all the stages of the software development. 
-The adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-their context.
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Different proposals~\cite{Babamir2010,AgarwalLS09,CholletL09,GutierrezRF10,XiaoCZBOLH08,JeongCL09,TsadimasNA12}
-support non-functional requirements in the context of web service development.
-
-Most software development methods define software processes that use the notion of refinement.
-Software process begins with the formulation of an abstract specification, which is successively refined to yield the implementation of the system.
-Methods for the development of web service applications are no exception to this rule.
-At least two levels of abstraction can be distinguished: a \textit{Business Level}, including the abstract specification, and a \textit{System Level}, including actual computer programs that implements the system.
-
-In the case of web service applications we will distinguish two separate layers of the implementation.
-The \textit{Composition Layer} is the upper layer of the implementation. 
-It defines the workflow of the system, in terms of individual service calls.
-The \textit{Service Layer} defines those services that are called by the composition.
-
-\[ \mathrm{Original} \]
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-}

### 2013-10-11 JournalToSubmit2013/metamodels.tex
reorg - martin

--- 2013-10-04 JournalToSubmit2013/3-metamodel.tex
@@
+%3.1. pi-SoD-M metamodels (presentar los metamodels)
+
 The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
+\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
@@
 The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}

### 2013-10-11 JournalToSubmit2013/methodologiesForNFR.tex
reorg - martin

%(aqui propuestas relacionadas, analizar sus caracteristicas, si son MDD, si parte de alto nivel, etc.)

### 2013-10-11 JournalToSubmit2013/modelingWithPISODM.tex
reorg - martin

%(Aquí lo dejaría con la estructura que teníamos para Caise, presentamos el framework y posteriormente comentamos modelos y transformaciones)

### 2013-10-11 JournalToSubmit2013/sodmAndPisodm.tex
reorg - martin

--- 2013-10-07 JournalToSubmit2013/1-intro.tex
@@
-Model Driven Development (MDD) is a top-down approach for the development of software systems. 
-The main ideas of MDD were originally proposed by the Object Management Group (OMG)~\cite{mda}, as a set of guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
-
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} within and between levels.
-Intra-level transformations are used to provide a unified representation of concepts of a given level.
-Inter-level transformation implement a refinement process between levels.
-Transformations may be automatic or semi-automatic.
-
-MDD techniques has been successfully used for the development of hardware and software systems~\cite{MDDvariosAqui}. 
-In particular, we are interested in the application of MDD to the design and implementation of web service applications.
-
-Service oriented computing~\cite{Papazoglou2007} is at the origin of an evolution in the field of software development~\cite{??}. 
-An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually developed.
-(Moreover, IT systems need to evolve according to the business needs.)
-Thus, organizations are seeking for mechanisms to bridge the gap between the systems developed and business needs~\cite{bell}. 
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful service based applications~\cite{5}.
-In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
-
-In Service-Oriented Computing, pre-existing services are
-combined to produce applications and provide the business logic. 
-The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}. 
-(Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.)
-Functional properties are not the only crucial aspect in the software development process. 
-Other properties need to be addressed to fit in the application with its context.
-These other aspects are called Non-Functional Properties.
-
-Non-functional aspects of the services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the beginning of the (service) software process.
-Most methods consider them only after the application has been implemented, in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). 
-This leads to service based applications that are partly specified and, thereby, partly compliant with the requirements of the application.
-Ideally, non-functional requirements should be considered along with all the stages of the software development. 
-The adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-the application context.
-
-In this work, we are interested in the extension of the \textit{Service Oriented Development Method} (SOD-M)~\cite{decastro1}, to support non-functional aspects, from the early stages of software development.
-SOD-M is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications.
-SOD-M does not provide support for the specification of non-functional requirements, such as
-security, reliability, and efficiency.
-
-The main goals of our work are:
-\begin{trivlist}
-\item \textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\item \textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\item \textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
-\end{trivlist}
-
-The rest of the paper is organized as follows. 
-Section\dots
-
-\COMMENT{
-\subsection*{The rest of this section is just for draft\dots}
-
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-\[ \mathrm{Intro\ SAC} \]
-
-Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.
-Functional properties are not the only crucial aspect in the software development process. 
-Other properties need to be addressed to fit in the application with its context.
-These other aspects are called Non-Functional Properties.
-
-Non-Functional Requirements (NFRs) specify those properties that are not addressed by the functional  specification.
-They are often called \textit{qualities} of the software system.
-%They are also referred as ``constraints'', ``quality attributes'', ``quality goals'', ``quality of
-%service requirements'' and ``non-behavioural requirements'' \cite{Stellman2005}.
-Non-Functional Requirements may specify response time, security constraints or quality of the solution, among others.
-
-Service-Oriented Computing~\cite{Papazoglou2007}, is a software development paradigm where pre-existing services are combined to produce more complex applications. 
-%The selection of services is usually guided by the functional requirements of the application. 
-%In this scenario, there exists the need to provide support for the specification of non-functional requirements, such as
-%security, reliability, and efficiency. 
-The development of service-based applications can benefit from the inclusion of NFRs to the software process from its early stages.
-Failure to comply with this inclusion means that the final application is obtained from a partial specification, making the deployment a difficult task.
-%Ideally, non-functional requirements
-%would be considered along with all the stages of the software development. 
-The adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-their context.
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Different proposals~\cite{Babamir2010,AgarwalLS09,CholletL09,GutierrezRF10,XiaoCZBOLH08,JeongCL09,TsadimasNA12}
-support non-functional requirements in the context of web service development.
-
-Most software development methods define software processes that use the notion of refinement.
-Software process begins with the formulation of an abstract specification, which is successively refined to yield the implementation of the system.
-Methods for the development of web service applications are no exception to this rule.
-At least two levels of abstraction can be distinguished: a \textit{Business Level}, including the abstract specification, and a \textit{System Level}, including actual computer programs that implements the system.
-
-In the case of web service applications we will distinguish two separate layers of the implementation.
-The \textit{Composition Layer} is the upper layer of the implementation. 
-It defines the workflow of the system, in terms of individual service calls.
-The \textit{Service Layer} defines those services that are called by the composition.
-
-\[ \mathrm{Original} \]
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-}
-
 \bigskip

### 2013-10-11 JournalToSubmit2013/transformations.tex
reorg - martin

--- 2013-10-04 JournalToSubmit2013/3-metamodel.tex
@@
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
-
-\begin{itemize}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions.
-They are graphically presented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
-
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
-\centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the servicesÕ composition and defines strategies to enforce this property at the servicesÕ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
-
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-%
-\begin{figure}
-\centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
-
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
 Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.

### 2013-10-11 JournalToSubmit2013/modelingWithPISODM.tex
martin

@@
 %(Aquí lo dejaría con la estructura que teníamos para Caise, presentamos el framework y posteriormente comentamos modelos y transformaciones)
+
+In this section we present $\pi$SOD-M, an MDD based methodology. 
+$\pi$SOD-M provides an environment for building service compositions considering
+their non-functional requirements. 
+$\pi$SOD-M extends the SOD-M meta-models by adding
+the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
+to represent non-functional requirements.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/piSODM}
+\caption{$\pi$SOD-M.}
+\label{fig:piSOD-M}
+\end{figure}
+
+$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
+well as transformations between these models.
+$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
+Constraints are restrictions that must be verified during the execution of the application. 
+An example of this is the requirement of the user's authentication for executing some system functions.
+
+Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
+$\pi$SOD-M proposes a development process based on the definition of models
+(instances of the meta-modes) and transformations between models.
+There are two kinds of transformations:
+Model-to-model transformations are used during the software process to refine the specification.
+Model-to-text transformations are the last step of the process and generate code.
+
+We extended SOD-M to include non-functional specifications.
+Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
+The former three are extensions of SOD-M meta-models and belong to the PIM level.
+The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
+
+The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
+Non-functional requirements are defined as \textit{constraints} over processing and data. 
+The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
+The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
+described in the \textit{$\pi$-UseCase} model into contracts that are associated
+with services. 
+The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
+which put together contracts with similar non-functional requirements. 
+For instance, security and privacy restrictions may be grouped into a security policy.
+\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
+
+At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
+The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
+\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
+Although PEWS is our language of choice, other composition languages can be used as target.
+This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

### 2013-10-11 JournalToSubmit2013/relworks.tex
martin

--- 2013-10-04 JournalToSubmit2013/5-relworks.tex
@@
+In this section we present some methods and techniques that are related to our work.
+We begin presenting SOD-M~\cite{decastro1}, a method for the development of web applications.
+SOD-M will be extended in the next sections to deal with NFRs.
+We also present here some other frameworks that are related to our proposal.
+
+\subsection{SOD-M}\label{sec:sodm}
+The Service-Oriented Development Method (SOD-M)~\cite{decastro1} is a MDD approach for service-based applications.
+SOD-M provides models and notation to specify service applications at different levels of abstraction. 
+As usual for MDD, SOD-M meta-models are organized into three levels: CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+
+Two models are defined at the CIM level: \textit{value model} 
+and \textit{BPMN model}. 
+\textsc{Explain these two here --Martin.}
+
+Models at the PIM level describe the structure of the application flow,
+while, the PSM level provides transformations towards more specific platforms.
+The PIM-level models proposed by SOD-M are: 
+\begin{trivlist}
+\item \textit{use case}: 
+\item \textit{extended use case}:
+\item \textit{service process}:
+\item \textit{service composition}. 
+\end{trivlist}
+
+SOD-M defines three meta-models at the PSM level: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
+These three levels have no support for describing non-functional requirements.
+
+The SOD-M approach includes transformations between models:
+\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
+an abstract model at the CIM level, it is possible to apply transformations for
+generating a model of the PSM level. In this context, it is necessary to
+follow the process activities described by the methodology.
+
+SOD-M considers two points of view:
+\textit{(i)} \textit{business}, focusing on the characteristics and requirements
+of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
+features and processes to be implemented in order application requirements. In
+this way, SOD-M aims to simplify the design of service-oriented applications, as
+well as its implementation using current technologies.
+
 Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.

### 2013-10-12 SOFSEM2014/Submitted/main.tex
Ready to work on SOFSEM final version

\documentclass{llncs}
\usepackage{tikz}
\usepackage{amsfonts}
\usepackage{graphicx,url}
\usepackage[T1]{fontenc}
%\usepackage{psfig}
\usepackage{graphics}
\usepackage{amssymb,amsmath,latexsym}
\usepackage{listings} 
\usepackage{subfig}

%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{example}[theorem]{Example}
%\newtheorem{algorithm}[theorem]{Algorithm}
%\newtheorem{definition}[theorem]{Definition}

\usepackage[thmmarks]{ntheorem}

\usepackage{listings}
\theoremstyle{plain}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{.}
\theoremindent0cm
\theoremnumbering{arabic}
\theoremsymbol{\ensuremath{\Box}} 
\newtheorem{defi}{Definition}%[section]

\theoremheaderfont{\itshape}
\newtheorem{examp}{Example}%[section]

\theoremstyle{plain}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{}
\theoremindent0cm
\theoremnumbering{arabic}
%\theoremsymbol{\ensuremath{\Box}} 
\newtheorem{algo}{Algorithm}%[section]

\theoremstyle{plain} 
\theoremsymbol{\ensuremath{\Box}} 
\theoremseparator{.} 
\newtheorem{prop}{Property}

\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
        true,false,rise,exception,extend,resume,to,return,function}}

\def\B{\hfill{$\Box$}}
\def\A{\hbox{${\cal A}$}}
\def\P{\hbox{${\cal P}$}}
\def\T{\hbox{${\cal T}$}}
\def\O{\hbox{${\cal O}$}}
\def\W{\hbox{${\cal W}$}}
\def\useOpt{\hbox{${has\_opt}$}}

\def\bigS{\hbox{${\cal S}$}}
\def\bigSd{\hbox{${\cal S}_D$}}
\def\bigSsm{\hbox{${\cal S}_D^{score}$}}

\newcommand{\ie}{\ifmmode \mathit{i.e.}\else \textit{i.e.}\fi}
\newcommand{\wrt}{\ifmmode \mathit{w.r.t.}\else \textit{w.r.t.}\fi}
\newcommand{\eg}{\ifmmode \mathit{e.g.}\else \textit{e.g.}\fi}
\newcommand{\tuple}[1]{\ifmmode \langle #1 \rangle\else $\langle #1 \rangle$\fi}

\usepackage{authblk}

\begin{document}
%
% --- Author Metadata here ---

% --- End of Author Metadata ---

\title{Supporting Non-Functional Requirements for Services Software Process: An MDD Approach}

\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}

\institute{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\  \email{Valeria.deCastro@urjc.es} 
\and Federal University of Rio Grande do Norte (UFRN) -- Natal-RN, Brazil\\
\email{\{mam,umberto\}@dimap.ufrn.br}
\and Federal Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
\and French Council of Scientific Research (CNRS) -- Grenoble, France\\
\email{Genoveva.Vargas-Solar@imag.fr}
}

\maketitle
\begin{abstract}
This paper presents an extension to the Service Oriented Development Method (SOD-M) to support the representation of some non-functional requirements. 
Specifically, we propose to extend SOD-M with: 
\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
\textit{(ii)} model to model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
We apply our proposal to develop a proof-of-concept example.
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]

%\terms{Delphi theory}

\keywords{MDD, Service Oriented Applications, Non-functional Properties}

\section{Introduction}

Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
The main ideas of MDD were originaly proposed by the Object Management Group
(OMG)~\cite{mda}. 
MDD provides a set of
guidelines for the structuring of specifications.
The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
Abstraction levels are called \textit{viewpoints}. 
MDD defines three viewpoints:

\begin{trivlist}
\item \textbf{Computation Independent Models (CIM):} This level focusses on the
environment of the system, as well as on its business and requirement specifications. 
This viewpoint represents the software system at its highest level of abstraction. 
At this moment of the development, the structure and system processing details are still unknown or undetermined.

\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
The specification defines those parts of the system that do not change from one platform to another.

\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
\end{trivlist}

Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
These transformations may be automatic or semi-automatic and implement the refinement process between levels.

\bigskip

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
Some methodologies and techniques have been proposed to help the software
developer in the specification of functional requirements of the business logic,
such as the Service Oriented Development Method
(SOD-M)~\cite{decastro1}. 
SOD-M is based on MDD and proposes
models, practices and techniques to aid in software development. SOD-M does not
provide suport to the specification of non-functional requirements, such as
security, reliability, and efficiency. Ideally, non-functional requirements
would be considered along with all the stages of the software development. The
adoption of non-functional specifications from the early states of development
can help the developer to produce applications that are capable of dealing with
the application context.

\bigskip

Non-functional properties of service oriented applications have been
addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
Dealing with these kind of properties involves the use of specific technologies
in different layers of the SOC architecture, for instance during the description
of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
service coordinations (like WS-BPEL~\cite{bpel03}).

%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
%\begin{trivlist}
%\item[-] Those coming from the Business Process Domain and from the Web
%service standards that propose ad-hoc protocols. 
%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
%\item[-] Those adopting a classic middleware approach were non-functional
%properties are provided as middleware services, like the ones presented
%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
%\item[-] Those providing languages and enabling the specification of protocols
%used for expressing service
%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
%\item[-] Those adopting separation of concerns,
%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
%\end{trivlist}

Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
They also assume that each service exports its interface.
So, the challenge of supporting non-functional properties is related to
\textit{(i)} The specification of the business rules of the application; and 
\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.

%In this context, there is a need of 
%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
%
%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.

The main goals of our work are:
\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;

\bigskip
This paper is organized as follows:
Sections~\ref{sec:sodm} and~\ref{sec:pisodm} presents, respectively, the SOD-M method of service software process and our proposed extension to deal with non-functional requirements.
A proof of concept example is developed in Section~\ref{sec:poc}.
We conclude the paper by presenting some related work and final remarks.

\section{SOD-M}\label{sec:sodm}
The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
proposes the usage of the MDD approach in the context of service-based applications.
SOD-M provides a framework with models and standards to express functionalities
of applications at a high-level of abstraction. SOD-M meta-models are organized 
into three levels: CIM (\textit{Computational Independent Models}), 
PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).

Two models are defined at the CIM level: \textit{value model} 
and \textit{BPMN model}. 
The PIM level models the entire structure of the application flow,
while, the PSM level provides transformations towards more specific platforms.
The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
These three levels have no support for describing non-functional requirements.

The SOD-M approach includes transformations between models:
\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
an abstract model at the CIM level, it is possible to apply transformations for
generating a model of the PSM level. In this context, it is necessary to
follow the process activities described by the methodology.

SOD-M considers two points of view:
\textit{(i)} \textit{business}, focusing on the characteristics and requirements
of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
features and processes to be implemented in order application requirements. In
this way, SOD-M aims to simplify the design of service-oriented applications, as
well as its implementation using current technologies.

\section{$\pi$SOD-M}\label{sec:pisodm}

%In this section we present $\pi$SOD-M, an MDD based methodology. 
$\pi$SOD-M provides an environment for building service compositions considering
their non-functional requirements. 
$\pi$SOD-M extends the SOD-M meta-models by adding
the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
to represent non-functional requirements.

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{figs/piSODM}
\caption{$\pi$SOD-M.}
\label{fig:piSOD-M}
\end{figure}

$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
well as transformations between these models.
$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
Constraints are restrictions that must be verified during the execution of the application. 
An example of this is the requirement of the user's authentication for executing some system functions.

Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
$\pi$SOD-M proposes a development process based on the definition of models
(instances of the meta-modes) and transformations between models.
There are two kinds of transformations:
Model-to-model transformations are used during the software process to refine the specification.
Model-to-text transformations are the last step of the process and generate code.

We extended SOD-M to include non-functional specifications.
Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
The former three are extensions of SOD-M meta-models and belong to the PIM level.
The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).

The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
Non-functional requirements are defined as \textit{constraints} over processing and data. 
The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
described in the \textit{$\pi$-UseCase} model into contracts that are associated
with services. 
The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
which put together contracts with similar non-functional requirements. 
For instance, security and privacy restrictions may be grouped into a security policy.
\textit{$\pi$-ServiceComposition} models can be refined into PSMs.

At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
Although PEWS is our language of choice, other composition languages can be used as target.
This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

In the next section we develop an example, to serve as a proof-of-concept.
The example will show the actual notation used for models.

\section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}

Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
Civilian users signal crimes using Twitter.
Police officers can notify crimes, as well as update information about solving cases.
Some of these information are confidential while other can be shared
to the community of users using this application.

Users can track crimes in given zones. 
Crime information stored by the system may be visualized on a map. 
Some users have different access rights than others.
For example, police officers have more access rights than civilians.

In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize the information.
The business process defines the logic of the application and is specified in terms of tasks. 
Tasks can be performed by persons or computers.

The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
Busines process is represented as a graph while requirements are given as text boxes.
\begin{figure}[ht!]
\centering
%\subfloat[\textit{Tracking Crimes} - Illustration of the Problem.]
%{\label{fig:problem}\includegraphics[width=0.95\textwidth]{figs/problem}}
%~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
%\\
%\subfloat[Contraints.]
{\label{fig:trans06}\includegraphics[width=0.95\textwidth]{figs/constraints}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\caption{Business process for the tracking crime example.}
\label{fig:constraints}
\end{figure}

In our example, crime processing can start with one of two tasks: \textit{(i) notify a crime}, or \textit{(ii) track a crime}.
Notified crimes are stored in a database. 
Tracked crimes are visualized in a map.
The used can ask for detailed information. 
The application is built upon four services: \textsf{twitter} and an \textsf{ad-hoc police service} for notifying crimes, \textsf{Amazon} used as  persistence service and  \textsf{Google Maps} for locating and displaying crimes on a map.

Non-functional requirements are specified by rules and conditions to be verified during the excecution of tasks.
In our example we have the following non-functional requirements:
\begin{trivlist}
\item[1.] Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
\item[2.] Crime notification needs privileged access.
\item[3.] Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
\item[4.] If \textsf{Google Maps} is unavailable, the results are delivered as text. 
\item[5.] Querying about crimes without having proper clearance yields an empty map.
\item[6.] Access rights to detailed information depends on user clearance and zone assignment for police officers. 
\item[7.] The application maintains a detailed log. 
\end{trivlist}

Considering the example of tracking crimes, all the system restrictions are
model as constraints in the methodology. We have three types of constraints:
\textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
Each use case can be associated to one or more constraints.

\paragraph*{\textit{$\pi$-UseCase} model:} 
In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
system functions (tasks) and constraints. 
We will not detail the functional part of the specification, due to lack of space.
The constrains defined for our tracking crime example are:

\begin{trivlist}
  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
  In this case, it represents the maximmum number of allowed login attempts;
\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
This is an example of \textit{business constraint}.
Additionally the requesting user must be logged in.
\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text. This is an example of \textit{exceptional behaviour constraint}. 
The availability of the Google Maps service is verified by a \textit{business constraint}.
 \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
\end{trivlist}
\begin{figure}[ht!]
\centering
\includegraphics[width=0.8\textwidth]{figs/piUseCase}
\caption{$\pi$-UseCase Model}
\label{fig:piUC}
\end{figure}
\paragraph*{\textit{$\pi$-ServiceProcess} model:}
The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
The new model describes the application's activities and defines contracts for each activity or for parts of the application.

A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
This (semi-automatic) transformation process is supported by a tool (described in~\cite{SouzaNeto:2012}).

The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
\begin{enumerate}
\item Tasks of the previous model are transformed into \textit{actions};
\item Actions are grouped into \textit{activities} (in accordance to the business logic).
\item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
\end{enumerate}

\begin{figure}[ht!]
\centering
\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
\caption{$\pi$-\textit{ServiceProcess} Model}
\label{fig:piSP}
\end{figure}

\paragraph*{\textit{$\pi$-ServiceComposition} model:} 
This model refines the previous model by using the activities to produce the workflow of the application.
The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
This model identifies the services and functions that correspond to each action in the business process.

\begin{figure}[t]
\centering
\subfloat[$\pi$-ServiceComposition Model]
{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\\
\subfloat[$\pi$-ServiceComposition Policies]
{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\caption{Service Composition and Policies.}
\label{fig:policies}
\end{figure}
This model describes the service compositions and their related business
collaborators (external services), policies, rules associated with a policy and
the whole application process and functionalities.
The workflow execution will consider the composition of business services and
their specific Business Collaborators. 
Each action will trigger the execution of one or more external services. 
Before and after each service call, policy rules are verified.

In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{busines collaborators} (external services and entities).
The interaction occurs by means of function calls (denoted by dotted lines in the figure).
Figure~\ref{fig:crimePolicies} shows the definition of three \textit{policies}, which define rules for service execution.
In our case we have policies for \textit{Security}, \textit{Performance} and \textit{Persistence}.

\paragraph*{$\pi$\textit{-PEWS} Model:}
These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
This code is a service composition program that can be compiled into executable code.
$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.

\begin{figure}[t]
\footnotesize
%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
%\begin{lstlisting}[label=list:rulePre]
\begin{verbatim}
//Namespaces specify service URI
1 namespace twitter = www.twitter.com/service.wsdl
2 namespace googlemaps = maps.googleapis.com/maps/api/service
3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
4 namespace police = www.police.fr/service.wsdl

//Operations 
5 alias publishTwitter = portType/publishTwitter in twitter
6 alias searchCrime = portType/searchCrime in amazondynamodb 
7 alias showMap = portType/showMap in googlemaps 
...
//Services
8  service notifyCrime = publishCrime  . publishTwitter 
9  service trackCrime= searchCrime . verifyService
10 Service visualizeCrime =  showMap . getCrimeDetail
...
//Path
11 ( notifyCrime . storeiCrime ) || 
12 ( trackCrime . visualizeCrime . getCrimeDetail )
...
//Contracts
13 defContract notifyCrimeContract{
14	isAppliedTo: notifyCrime
15	requires: userId == ?? && passw == ?? && 
16	                               req(notifyCrime)  < 3
17 	  (OnFailureDo: NOT(action_publish(crime));
18	ensures: publishTwitter(crime) == true
19 	  (OnFailureDo: skip);
20 }
...
\end{verbatim}
% \end{lstlisting}
\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
\end{figure}

The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
The figure shows a simplified program code, produced in accordance to the following guidelines:
\begin{trivlist}
\item[1.] Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
\item[2.] Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
\item[3.] The workflow in Figure~\ref{fig:piSC} is translated into a textual representation (lines 11 and 12 of the program).
\item[4.] \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed in the case thar a condition is not verified. 
Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
\end{trivlist}

For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.

\section{Related Work and Conclusions}
\label{sec:conclusions}

Over the last years, a number of approaches have been proposed for the
development of web services. These approaches range from the proposal of new 
languages for web service descriptions~\cite{bpel03,SBS04} 
to techniques to support phases of the development cycle of this kind of
software~\cite{lipari2007,BianculliGSBG07}. In general, these
approaches concentrate on specific problems, like supporting transactions or
QoS, in order to improve the security and reliability of service-based
applications. Some proposals address service composition: workflow
definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
services~\cite{BHM06}. The proposed solutions come from many communities, 
including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
Software Engineering~\cite{Aal03,MendesPDB09}, Programming
Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.

Despite the variety of tecniques proposed, there is not yet a consensus on a software
process methodology for web services. Some methodologies address the service-based 
development towards a standard or a new way to develop reliable applications. 
SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
S-Cube \cite{scube2010book} is focused on the representation of business processes and 
service-based development; SOMA \cite{soma} is a methodology described by IBM
for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
methodology for building service-based infrastructure for collaborative
enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
that acts as a bridge between business modeling and design models, and traditional 
approaches for software engineering \cite{sommerville08} applied to SOC.

\bigskip

This paper presented the $\pi$SOD-M software process for specifying and designing service based applications in the presence of some non-functional constraints. 
Our proposal harness the SOD-M method with constraints, policies and contracts in order to consider non-functional constraints.
We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.

$\pi$SOD-M is being used in an academic environment.
So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).

\bibliographystyle{splncs_srt}
\bibliography{biblio2} 
\end{document}

### 2013-10-14 SOFSEM2014/FinalVersion/conclusion.tex
final - chop chop

--- 2013-10-12 SOFSEM2014/Submitted/main.tex
@@
-\documentclass{llncs}
-\usepackage{tikz}
-\usepackage{amsfonts}
-\usepackage{graphicx,url}
-\usepackage[T1]{fontenc}
-%\usepackage{psfig}
-\usepackage{graphics}
-\usepackage{amssymb,amsmath,latexsym}
-\usepackage{listings} 
-\usepackage{subfig}
-
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{example}[theorem]{Example}
-%\newtheorem{algorithm}[theorem]{Algorithm}
-%\newtheorem{definition}[theorem]{Definition}
-
-\usepackage[thmmarks]{ntheorem}
-
-\usepackage{listings}
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{.}
-\theoremindent0cm
-\theoremnumbering{arabic}
-\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{defi}{Definition}%[section]
-
-\theoremheaderfont{\itshape}
-\newtheorem{examp}{Example}%[section]
-
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{}
-\theoremindent0cm
-\theoremnumbering{arabic}
-%\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{algo}{Algorithm}%[section]
-
-\theoremstyle{plain} 
-\theoremsymbol{\ensuremath{\Box}} 
-\theoremseparator{.} 
-\newtheorem{prop}{Property}
-
-\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
-        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
-        true,false,rise,exception,extend,resume,to,return,function}}
-
-\def\B{\hfill{$\Box$}}
-\def\A{\hbox{${\cal A}$}}
-\def\P{\hbox{${\cal P}$}}
-\def\T{\hbox{${\cal T}$}}
-\def\O{\hbox{${\cal O}$}}
-\def\W{\hbox{${\cal W}$}}
-\def\useOpt{\hbox{${has\_opt}$}}
-
-\def\bigS{\hbox{${\cal S}$}}
-\def\bigSd{\hbox{${\cal S}_D$}}
-\def\bigSsm{\hbox{${\cal S}_D^{score}$}}
-
-\newcommand{\ie}{\ifmmode \mathit{i.e.}\else \textit{i.e.}\fi}
-\newcommand{\wrt}{\ifmmode \mathit{w.r.t.}\else \textit{w.r.t.}\fi}
-\newcommand{\eg}{\ifmmode \mathit{e.g.}\else \textit{e.g.}\fi}
-\newcommand{\tuple}[1]{\ifmmode \langle #1 \rangle\else $\langle #1 \rangle$\fi}
-
-\usepackage{authblk}
-
-\begin{document}
-%
-% --- Author Metadata here ---
-
-% --- End of Author Metadata ---
-
-\title{Supporting Non-Functional Requirements for Services Software Process: An MDD Approach}
-
-\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}
-
-\institute{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\  \email{Valeria.deCastro@urjc.es} 
-\and Federal University of Rio Grande do Norte (UFRN) -- Natal-RN, Brazil\\
-\email{\{mam,umberto\}@dimap.ufrn.br}
-\and Federal Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
-\and French Council of Scientific Research (CNRS) -- Grenoble, France\\
-\email{Genoveva.Vargas-Solar@imag.fr}
-}
-
-\maketitle
-\begin{abstract}
-This paper presents an extension to the Service Oriented Development Method (SOD-M) to support the representation of some non-functional requirements. 
-Specifically, we propose to extend SOD-M with: 
-\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
-\textit{(ii)} model to model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
-\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
-We apply our proposal to develop a proof-of-concept example.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
-%performance measures]
-
-%\terms{Delphi theory}
-
-\keywords{MDD, Service Oriented Applications, Non-functional Properties}
-
-\section{Introduction}
-
-Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
-The main ideas of MDD were originaly proposed by the Object Management Group
-(OMG)~\cite{mda}. 
-MDD provides a set of
-guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
-Abstraction levels are called \textit{viewpoints}. 
-MDD defines three viewpoints:
-
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
-These transformations may be automatic or semi-automatic and implement the refinement process between levels.
-
-\bigskip
-
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
-Some methodologies and techniques have been proposed to help the software
-developer in the specification of functional requirements of the business logic,
-such as the Service Oriented Development Method
-(SOD-M)~\cite{decastro1}. 
-SOD-M is based on MDD and proposes
-models, practices and techniques to aid in software development. SOD-M does not
-provide suport to the specification of non-functional requirements, such as
-security, reliability, and efficiency. Ideally, non-functional requirements
-would be considered along with all the stages of the software development. The
-adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-the application context.
-
-\bigskip
-
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-The main goals of our work are:
-\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
-
-\bigskip
-This paper is organized as follows:
-Sections~\ref{sec:sodm} and~\ref{sec:pisodm} presents, respectively, the SOD-M method of service software process and our proposed extension to deal with non-functional requirements.
-A proof of concept example is developed in Section~\ref{sec:poc}.
-We conclude the paper by presenting some related work and final remarks.
-
-\section{SOD-M}\label{sec:sodm}
-The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
-proposes the usage of the MDD approach in the context of service-based applications.
-SOD-M provides a framework with models and standards to express functionalities
-of applications at a high-level of abstraction. SOD-M meta-models are organized 
-into three levels: CIM (\textit{Computational Independent Models}), 
-PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-
-Two models are defined at the CIM level: \textit{value model} 
-and \textit{BPMN model}. 
-The PIM level models the entire structure of the application flow,
-while, the PSM level provides transformations towards more specific platforms.
-The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
-\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
-These three levels have no support for describing non-functional requirements.
-
-The SOD-M approach includes transformations between models:
-\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
-an abstract model at the CIM level, it is possible to apply transformations for
-generating a model of the PSM level. In this context, it is necessary to
-follow the process activities described by the methodology.
-
-SOD-M considers two points of view:
-\textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
-this way, SOD-M aims to simplify the design of service-oriented applications, as
-well as its implementation using current technologies.
-
-\section{$\pi$SOD-M}\label{sec:pisodm}
-
-%In this section we present $\pi$SOD-M, an MDD based methodology. 
-$\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. 
-$\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
-\caption{$\pi$SOD-M.}
-\label{fig:piSOD-M}
-\end{figure}
-
-$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
-well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
-Constraints are restrictions that must be verified during the execution of the application. 
-An example of this is the requirement of the user's authentication for executing some system functions.
-
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-modes) and transformations between models.
-There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
-
-We extended SOD-M to include non-functional specifications.
-Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
-
-The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data. 
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
-The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services. 
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
-which put together contracts with similar non-functional requirements. 
-For instance, security and privacy restrictions may be grouped into a security policy.
-\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
-
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
-Although PEWS is our language of choice, other composition languages can be used as target.
-This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
-
-In the next section we develop an example, to serve as a proof-of-concept.
-The example will show the actual notation used for models.
-
-\section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
-
-Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
-Civilian users signal crimes using Twitter.
-Police officers can notify crimes, as well as update information about solving cases.
-Some of these information are confidential while other can be shared
-to the community of users using this application.
-
-Users can track crimes in given zones. 
-Crime information stored by the system may be visualized on a map. 
-Some users have different access rights than others.
-For example, police officers have more access rights than civilians.
-
-In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize the information.
-The business process defines the logic of the application and is specified in terms of tasks. 
-Tasks can be performed by persons or computers.
-
-The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
-Busines process is represented as a graph while requirements are given as text boxes.
-\begin{figure}[ht!]
-\centering
-%\subfloat[\textit{Tracking Crimes} - Illustration of the Problem.]
-%{\label{fig:problem}\includegraphics[width=0.95\textwidth]{figs/problem}}
-%~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-%\\
-%\subfloat[Contraints.]
-{\label{fig:trans06}\includegraphics[width=0.95\textwidth]{figs/constraints}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Business process for the tracking crime example.}
-\label{fig:constraints}
-\end{figure}
-
-In our example, crime processing can start with one of two tasks: \textit{(i) notify a crime}, or \textit{(ii) track a crime}.
-Notified crimes are stored in a database. 
-Tracked crimes are visualized in a map.
-The used can ask for detailed information. 
-The application is built upon four services: \textsf{twitter} and an \textsf{ad-hoc police service} for notifying crimes, \textsf{Amazon} used as  persistence service and  \textsf{Google Maps} for locating and displaying crimes on a map.
-
-Non-functional requirements are specified by rules and conditions to be verified during the excecution of tasks.
-In our example we have the following non-functional requirements:
-\begin{trivlist}
-\item[1.] Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
-\item[2.] Crime notification needs privileged access.
-\item[3.] Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
-\item[4.] If \textsf{Google Maps} is unavailable, the results are delivered as text. 
-\item[5.] Querying about crimes without having proper clearance yields an empty map.
-\item[6.] Access rights to detailed information depends on user clearance and zone assignment for police officers. 
-\item[7.] The application maintains a detailed log. 
-\end{trivlist}
-
-Considering the example of tracking crimes, all the system restrictions are
-model as constraints in the methodology. We have three types of constraints:
-\textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case can be associated to one or more constraints.
-
-\paragraph*{\textit{$\pi$-UseCase} model:} 
-In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
-system functions (tasks) and constraints. 
-We will not detail the functional part of the specification, due to lack of space.
-The constrains defined for our tracking crime example are:
-
-\begin{trivlist}
-  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
-  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
-  In this case, it represents the maximmum number of allowed login attempts;
-\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
-\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
-This is an example of \textit{business constraint}.
-Additionally the requesting user must be logged in.
-\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text.
-This is an example of \textit{exceptional behaviour constraint}.
-The availability of the Google Maps service is verified by a \textit{business constraint}.
- \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
-\end{trivlist}
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.8\textwidth]{figs/piUseCase}
-\caption{$\pi$-UseCase Model}
-\label{fig:piUC}
-\end{figure}
-\paragraph*{\textit{$\pi$-ServiceProcess} model:}
-The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
-The new model describes the application's activities and defines contracts for each activity or for parts of the application.
-
-A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
-This (semi-automatic) transformation process is supported by a tool (described in~\cite{SouzaNeto:2012}).
-
-The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
-\begin{enumerate}
-\item Tasks of the previous model are transformed into \textit{actions};
-\item Actions are grouped into \textit{activities} (in accordance to the business logic).
-\item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
-\end{enumerate}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
-\caption{$\pi$-\textit{ServiceProcess} Model}
-\label{fig:piSP}
-\end{figure}
-
-\paragraph*{\textit{$\pi$-ServiceComposition} model:} 
-This model refines the previous model by using the activities to produce the workflow of the application.
-The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
-This model identifies the services and functions that correspond to each action in the business process.
-
-\begin{figure}[t]
-\centering
-\subfloat[$\pi$-ServiceComposition Model]
-{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\\
-\subfloat[$\pi$-ServiceComposition Policies]
-{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Service Composition and Policies.}
-\label{fig:policies}
-\end{figure}
-This model describes the service compositions and their related business
-collaborators (external services), policies, rules associated with a policy and
-the whole application process and functionalities.
-The workflow execution will consider the composition of business services and
-their specific Business Collaborators. 
-Each action will trigger the execution of one or more external services. 
-Before and after each service call, policy rules are verified.
-
-In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
-Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{busines collaborators} (external services and entities).
-The interaction occurs by means of function calls (denoted by dotted lines in the figure).
-Figure~\ref{fig:crimePolicies} shows the definition of three \textit{policies}, which define rules for service execution.
-In our case we have policies for \textit{Security}, \textit{Performance} and \textit{Persistence}.
-
-\paragraph*{$\pi$\textit{-PEWS} Model:}
-These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
-This code is a service composition program that can be compiled into executable code.
-$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
-
-\begin{figure}[t]
-\footnotesize
-%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
-%\begin{lstlisting}[label=list:rulePre]
-\begin{verbatim}
-//Namespaces specify service URI
-1 namespace twitter = www.twitter.com/service.wsdl
-2 namespace googlemaps = maps.googleapis.com/maps/api/service
-3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
-4 namespace police = www.police.fr/service.wsdl
-
-//Operations 
-5 alias publishTwitter = portType/publishTwitter in twitter
-6 alias searchCrime = portType/searchCrime in amazondynamodb 
-7 alias showMap = portType/showMap in googlemaps 
-...
-//Services
-8  service notifyCrime = publishCrime  . publishTwitter 
-9  service trackCrime= searchCrime . verifyService
-10 Service visualizeCrime =  showMap . getCrimeDetail
-...
-//Path
-11 ( notifyCrime . storeiCrime ) || 
-12 ( trackCrime . visualizeCrime . getCrimeDetail )
-...
-//Contracts
-13 defContract notifyCrimeContract{
-14	isAppliedTo: notifyCrime
-15	requires: userId == ?? && passw == ?? && 
-16	                               req(notifyCrime)  < 3
-17 	  (OnFailureDo: NOT(action_publish(crime));
-18	ensures: publishTwitter(crime) == true
-19 	  (OnFailureDo: skip);
-20 }
-...
-\end{verbatim}
-% \end{lstlisting}
-\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
-\end{figure}
-
-The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
-The figure shows a simplified program code, produced in accordance to the following guidelines:
-\begin{trivlist}
-\item[1.] Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
-We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
-\item[2.] Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
-\item[3.] The workflow in Figure~\ref{fig:piSC} is translated into a textual representation (lines 11 and 12 of the program).
-\item[4.] \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed in the case thar a condition is not verified.
-Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
-\end{trivlist}
-
-For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.
-
-\section{Related Work and Conclusions}
-\label{sec:conclusions}
-
 Over the last years, a number of approaches have been proposed for the
@@
 We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
-
-\bibliographystyle{splncs_srt}
-\bibliography{biblio2} 
-\end{document}

### 2013-10-14 SOFSEM2014/FinalVersion/introduction.tex
final - chop chop

--- 2013-10-12 SOFSEM2014/Submitted/main.tex
@@
-\documentclass{llncs}
-\usepackage{tikz}
-\usepackage{amsfonts}
-\usepackage{graphicx,url}
-\usepackage[T1]{fontenc}
-%\usepackage{psfig}
-\usepackage{graphics}
-\usepackage{amssymb,amsmath,latexsym}
-\usepackage{listings} 
-\usepackage{subfig}
-
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{example}[theorem]{Example}
-%\newtheorem{algorithm}[theorem]{Algorithm}
-%\newtheorem{definition}[theorem]{Definition}
-
-\usepackage[thmmarks]{ntheorem}
-
-\usepackage{listings}
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{.}
-\theoremindent0cm
-\theoremnumbering{arabic}
-\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{defi}{Definition}%[section]
-
-\theoremheaderfont{\itshape}
-\newtheorem{examp}{Example}%[section]
-
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{}
-\theoremindent0cm
-\theoremnumbering{arabic}
-%\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{algo}{Algorithm}%[section]
-
-\theoremstyle{plain} 
-\theoremsymbol{\ensuremath{\Box}} 
-\theoremseparator{.} 
-\newtheorem{prop}{Property}
-
-\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
-        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
-        true,false,rise,exception,extend,resume,to,return,function}}
-
-\def\B{\hfill{$\Box$}}
-\def\A{\hbox{${\cal A}$}}
-\def\P{\hbox{${\cal P}$}}
-\def\T{\hbox{${\cal T}$}}
-\def\O{\hbox{${\cal O}$}}
-\def\W{\hbox{${\cal W}$}}
-\def\useOpt{\hbox{${has\_opt}$}}
-
-\def\bigS{\hbox{${\cal S}$}}
-\def\bigSd{\hbox{${\cal S}_D$}}
-\def\bigSsm{\hbox{${\cal S}_D^{score}$}}
-
-\newcommand{\ie}{\ifmmode \mathit{i.e.}\else \textit{i.e.}\fi}
-\newcommand{\wrt}{\ifmmode \mathit{w.r.t.}\else \textit{w.r.t.}\fi}
-\newcommand{\eg}{\ifmmode \mathit{e.g.}\else \textit{e.g.}\fi}
-\newcommand{\tuple}[1]{\ifmmode \langle #1 \rangle\else $\langle #1 \rangle$\fi}
-
-\usepackage{authblk}
-
-\begin{document}
-%
-% --- Author Metadata here ---
-
-% --- End of Author Metadata ---
-
-\title{Supporting Non-Functional Requirements for Services Software Process: An MDD Approach}
-
-\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}
-
-\institute{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\  \email{Valeria.deCastro@urjc.es} 
-\and Federal University of Rio Grande do Norte (UFRN) -- Natal-RN, Brazil\\
-\email{\{mam,umberto\}@dimap.ufrn.br}
-\and Federal Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
-\and French Council of Scientific Research (CNRS) -- Grenoble, France\\
-\email{Genoveva.Vargas-Solar@imag.fr}
-}
-
-\maketitle
-\begin{abstract}
-This paper presents an extension to the Service Oriented Development Method (SOD-M) to support the representation of some non-functional requirements. 
-Specifically, we propose to extend SOD-M with: 
-\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
-\textit{(ii)} model to model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
-\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
-We apply our proposal to develop a proof-of-concept example.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
-%performance measures]
-
-%\terms{Delphi theory}
-
-\keywords{MDD, Service Oriented Applications, Non-functional Properties}
-
-\section{Introduction}
-
 Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
@@
 We conclude the paper by presenting some related work and final remarks.
-
-\section{SOD-M}\label{sec:sodm}
-The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
-proposes the usage of the MDD approach in the context of service-based applications.
-SOD-M provides a framework with models and standards to express functionalities
-of applications at a high-level of abstraction. SOD-M meta-models are organized 
-into three levels: CIM (\textit{Computational Independent Models}), 
-PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-
-Two models are defined at the CIM level: \textit{value model} 
-and \textit{BPMN model}. 
-The PIM level models the entire structure of the application flow,
-while, the PSM level provides transformations towards more specific platforms.
-The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
-\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
-These three levels have no support for describing non-functional requirements.
-
-The SOD-M approach includes transformations between models:
-\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
-an abstract model at the CIM level, it is possible to apply transformations for
-generating a model of the PSM level. In this context, it is necessary to
-follow the process activities described by the methodology.
-
-SOD-M considers two points of view:
-\textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
-this way, SOD-M aims to simplify the design of service-oriented applications, as
-well as its implementation using current technologies.
-
-\section{$\pi$SOD-M}\label{sec:pisodm}
-
-%In this section we present $\pi$SOD-M, an MDD based methodology. 
-$\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. 
-$\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
-\caption{$\pi$SOD-M.}
-\label{fig:piSOD-M}
-\end{figure}
-
-$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
-well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
-Constraints are restrictions that must be verified during the execution of the application. 
-An example of this is the requirement of the user's authentication for executing some system functions.
-
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-modes) and transformations between models.
-There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
-
-We extended SOD-M to include non-functional specifications.
-Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
-
-The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data. 
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
-The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services. 
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
-which put together contracts with similar non-functional requirements. 
-For instance, security and privacy restrictions may be grouped into a security policy.
-\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
-
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
-Although PEWS is our language of choice, other composition languages can be used as target.
-This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
-
-In the next section we develop an example, to serve as a proof-of-concept.
-The example will show the actual notation used for models.
-
-\section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
-
-Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
-Civilian users signal crimes using Twitter.
-Police officers can notify crimes, as well as update information about solving cases.
-Some of these information are confidential while other can be shared
-to the community of users using this application.
-
-Users can track crimes in given zones. 
-Crime information stored by the system may be visualized on a map. 
-Some users have different access rights than others.
-For example, police officers have more access rights than civilians.
-
-In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize the information.
-The business process defines the logic of the application and is specified in terms of tasks. 
-Tasks can be performed by persons or computers.
-
-The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
-Busines process is represented as a graph while requirements are given as text boxes.
-\begin{figure}[ht!]
-\centering
-%\subfloat[\textit{Tracking Crimes} - Illustration of the Problem.]
-%{\label{fig:problem}\includegraphics[width=0.95\textwidth]{figs/problem}}
-%~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-%\\
-%\subfloat[Contraints.]
-{\label{fig:trans06}\includegraphics[width=0.95\textwidth]{figs/constraints}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Business process for the tracking crime example.}
-\label{fig:constraints}
-\end{figure}
-
-In our example, crime processing can start with one of two tasks: \textit{(i) notify a crime}, or \textit{(ii) track a crime}.
-Notified crimes are stored in a database. 
-Tracked crimes are visualized in a map.
-The used can ask for detailed information. 
-The application is built upon four services: \textsf{twitter} and an \textsf{ad-hoc police service} for notifying crimes, \textsf{Amazon} used as  persistence service and  \textsf{Google Maps} for locating and displaying crimes on a map.
-
-Non-functional requirements are specified by rules and conditions to be verified during the excecution of tasks.
-In our example we have the following non-functional requirements:
-\begin{trivlist}
-\item[1.] Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
-\item[2.] Crime notification needs privileged access.
-\item[3.] Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
-\item[4.] If \textsf{Google Maps} is unavailable, the results are delivered as text. 
-\item[5.] Querying about crimes without having proper clearance yields an empty map.
-\item[6.] Access rights to detailed information depends on user clearance and zone assignment for police officers. 
-\item[7.] The application maintains a detailed log. 
-\end{trivlist}
-
-Considering the example of tracking crimes, all the system restrictions are
-model as constraints in the methodology. We have three types of constraints:
-\textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case can be associated to one or more constraints.
-
-\paragraph*{\textit{$\pi$-UseCase} model:} 
-In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
-system functions (tasks) and constraints. 
-We will not detail the functional part of the specification, due to lack of space.
-The constrains defined for our tracking crime example are:
-
-\begin{trivlist}
-  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
-  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
-  In this case, it represents the maximmum number of allowed login attempts;
-\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
-\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
-This is an example of \textit{business constraint}.
-Additionally the requesting user must be logged in.
-\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text.
-This is an example of \textit{exceptional behaviour constraint}.
-The availability of the Google Maps service is verified by a \textit{business constraint}.
- \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
-\end{trivlist}
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.8\textwidth]{figs/piUseCase}
-\caption{$\pi$-UseCase Model}
-\label{fig:piUC}
-\end{figure}
-\paragraph*{\textit{$\pi$-ServiceProcess} model:}
-The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
-The new model describes the application's activities and defines contracts for each activity or for parts of the application.
-
-A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
-This (semi-automatic) transformation process is supported by a tool (described in~\cite{SouzaNeto:2012}).
-
-The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
-\begin{enumerate}
-\item Tasks of the previous model are transformed into \textit{actions};
-\item Actions are grouped into \textit{activities} (in accordance to the business logic).
-\item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
-\end{enumerate}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
-\caption{$\pi$-\textit{ServiceProcess} Model}
-\label{fig:piSP}
-\end{figure}
-
-\paragraph*{\textit{$\pi$-ServiceComposition} model:} 
-This model refines the previous model by using the activities to produce the workflow of the application.
-The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
-This model identifies the services and functions that correspond to each action in the business process.
-
-\begin{figure}[t]
-\centering
-\subfloat[$\pi$-ServiceComposition Model]
-{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\\
-\subfloat[$\pi$-ServiceComposition Policies]
-{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Service Composition and Policies.}
-\label{fig:policies}
-\end{figure}
-This model describes the service compositions and their related business
-collaborators (external services), policies, rules associated with a policy and
-the whole application process and functionalities.
-The workflow execution will consider the composition of business services and
-their specific Business Collaborators. 
-Each action will trigger the execution of one or more external services. 
-Before and after each service call, policy rules are verified.
-
-In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
-Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{busines collaborators} (external services and entities).
-The interaction occurs by means of function calls (denoted by dotted lines in the figure).
-Figure~\ref{fig:crimePolicies} shows the definition of three \textit{policies}, which define rules for service execution.
-In our case we have policies for \textit{Security}, \textit{Performance} and \textit{Persistence}.
-
-\paragraph*{$\pi$\textit{-PEWS} Model:}
-These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
-This code is a service composition program that can be compiled into executable code.
-$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
-
-\begin{figure}[t]
-\footnotesize
-%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
-%\begin{lstlisting}[label=list:rulePre]
-\begin{verbatim}
-//Namespaces specify service URI
-1 namespace twitter = www.twitter.com/service.wsdl
-2 namespace googlemaps = maps.googleapis.com/maps/api/service
-3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
-4 namespace police = www.police.fr/service.wsdl
-
-//Operations 
-5 alias publishTwitter = portType/publishTwitter in twitter
-6 alias searchCrime = portType/searchCrime in amazondynamodb 
-7 alias showMap = portType/showMap in googlemaps 
-...
-//Services
-8  service notifyCrime = publishCrime  . publishTwitter 
-9  service trackCrime= searchCrime . verifyService
-10 Service visualizeCrime =  showMap . getCrimeDetail
-...
-//Path
-11 ( notifyCrime . storeiCrime ) || 
-12 ( trackCrime . visualizeCrime . getCrimeDetail )
-...
-//Contracts
-13 defContract notifyCrimeContract{
-14	isAppliedTo: notifyCrime
-15	requires: userId == ?? && passw == ?? && 
-16	                               req(notifyCrime)  < 3
-17 	  (OnFailureDo: NOT(action_publish(crime));
-18	ensures: publishTwitter(crime) == true
-19 	  (OnFailureDo: skip);
-20 }
-...
-\end{verbatim}
-% \end{lstlisting}
-\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
-\end{figure}
-
-The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
-The figure shows a simplified program code, produced in accordance to the following guidelines:
-\begin{trivlist}
-\item[1.] Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
-We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
-\item[2.] Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
-\item[3.] The workflow in Figure~\ref{fig:piSC} is translated into a textual representation (lines 11 and 12 of the program).
-\item[4.] \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed in the case thar a condition is not verified.
-Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
-\end{trivlist}
-
-For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.
-
-\section{Related Work and Conclusions}
-\label{sec:conclusions}
-
-Over the last years, a number of approaches have been proposed for the
-development of web services. These approaches range from the proposal of new 
-languages for web service descriptions~\cite{bpel03,SBS04} 
-to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007,BianculliGSBG07}. In general, these
-approaches concentrate on specific problems, like supporting transactions or
-QoS, in order to improve the security and reliability of service-based
-applications. Some proposals address service composition: workflow
-definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
-services~\cite{BHM06}. The proposed solutions come from many communities, 
-including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
-Software Engineering~\cite{Aal03,MendesPDB09}, Programming
-Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
-
-Despite the variety of tecniques proposed, there is not yet a consensus on a software
-process methodology for web services. Some methodologies address the service-based 
-development towards a standard or a new way to develop reliable applications. 
-SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
-S-Cube \cite{scube2010book} is focused on the representation of business processes and 
-service-based development; SOMA \cite{soma} is a methodology described by IBM
-for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
-BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
-methodology for building service-based infrastructure for collaborative
-enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering \cite{sommerville08} applied to SOC.
-
-\bigskip
-
-This paper presented the $\pi$SOD-M software process for specifying and designing service based applications in the presence of some non-functional constraints. 
-Our proposal harness the SOD-M method with constraints, policies and contracts in order to consider non-functional constraints.
-We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.
-
-$\pi$SOD-M is being used in an academic environment.
-So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
-We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
-
-\bibliographystyle{splncs_srt}
-\bibliography{biblio2} 
-\end{document}

### 2013-10-14 SOFSEM2014/FinalVersion/main.tex
final - chop chop

--- 2013-10-12 SOFSEM2014/Submitted/main.tex
@@
 \section{Introduction}
-
-Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
-The main ideas of MDD were originaly proposed by the Object Management Group
-(OMG)~\cite{mda}. 
-MDD provides a set of
-guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
-Abstraction levels are called \textit{viewpoints}. 
-MDD defines three viewpoints:
-
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
-These transformations may be automatic or semi-automatic and implement the refinement process between levels.
-
-\bigskip
-
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
-Some methodologies and techniques have been proposed to help the software
-developer in the specification of functional requirements of the business logic,
-such as the Service Oriented Development Method
-(SOD-M)~\cite{decastro1}. 
-SOD-M is based on MDD and proposes
-models, practices and techniques to aid in software development. SOD-M does not
-provide suport to the specification of non-functional requirements, such as
-security, reliability, and efficiency. Ideally, non-functional requirements
-would be considered along with all the stages of the software development. The
-adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-the application context.
-
-\bigskip
-
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-The main goals of our work are:
-\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
-
-\bigskip
-This paper is organized as follows:
-Sections~\ref{sec:sodm} and~\ref{sec:pisodm} presents, respectively, the SOD-M method of service software process and our proposed extension to deal with non-functional requirements.
-A proof of concept example is developed in Section~\ref{sec:poc}.
-We conclude the paper by presenting some related work and final remarks.
+\label{sec:intro}
+\include{introduction}
 
 \section{SOD-M}\label{sec:sodm}
-The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
-proposes the usage of the MDD approach in the context of service-based applications.
-SOD-M provides a framework with models and standards to express functionalities
-of applications at a high-level of abstraction. SOD-M meta-models are organized 
-into three levels: CIM (\textit{Computational Independent Models}), 
-PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-
-Two models are defined at the CIM level: \textit{value model} 
-and \textit{BPMN model}. 
-The PIM level models the entire structure of the application flow,
-while, the PSM level provides transformations towards more specific platforms.
-The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
-\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
-These three levels have no support for describing non-functional requirements.
-
-The SOD-M approach includes transformations between models:
-\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
-an abstract model at the CIM level, it is possible to apply transformations for
-generating a model of the PSM level. In this context, it is necessary to
-follow the process activities described by the methodology.
-
-SOD-M considers two points of view:
-\textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
-this way, SOD-M aims to simplify the design of service-oriented applications, as
-well as its implementation using current technologies.
+\include{sodm}
 
 \section{$\pi$SOD-M}\label{sec:pisodm}
-
-%In this section we present $\pi$SOD-M, an MDD based methodology. 
-$\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. 
-$\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
-\caption{$\pi$SOD-M.}
-\label{fig:piSOD-M}
-\end{figure}
-
-$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
-well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
-Constraints are restrictions that must be verified during the execution of the application. 
-An example of this is the requirement of the user's authentication for executing some system functions.
-
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-modes) and transformations between models.
-There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
-
-We extended SOD-M to include non-functional specifications.
-Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
-
-The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data. 
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
-The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services. 
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
-which put together contracts with similar non-functional requirements. 
-For instance, security and privacy restrictions may be grouped into a security policy.
-\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
-
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
-Although PEWS is our language of choice, other composition languages can be used as target.
-This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
-
-In the next section we develop an example, to serve as a proof-of-concept.
-The example will show the actual notation used for models.
+\include{pisodm}
 
 \section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
-
-Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
-Civilian users signal crimes using Twitter.
-Police officers can notify crimes, as well as update information about solving cases.
-Some of these information are confidential while other can be shared
-to the community of users using this application.
-
-Users can track crimes in given zones. 
-Crime information stored by the system may be visualized on a map. 
-Some users have different access rights than others.
-For example, police officers have more access rights than civilians.
-
-In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize the information.
-The business process defines the logic of the application and is specified in terms of tasks. 
-Tasks can be performed by persons or computers.
-
-The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
-Busines process is represented as a graph while requirements are given as text boxes.
-\begin{figure}[ht!]
-\centering
-%\subfloat[\textit{Tracking Crimes} - Illustration of the Problem.]
-%{\label{fig:problem}\includegraphics[width=0.95\textwidth]{figs/problem}}
-%~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-%\\
-%\subfloat[Contraints.]
-{\label{fig:trans06}\includegraphics[width=0.95\textwidth]{figs/constraints}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Business process for the tracking crime example.}
-\label{fig:constraints}
-\end{figure}
-
-In our example, crime processing can start with one of two tasks: \textit{(i) notify a crime}, or \textit{(ii) track a crime}.
-Notified crimes are stored in a database. 
-Tracked crimes are visualized in a map.
-The used can ask for detailed information. 
-The application is built upon four services: \textsf{twitter} and an \textsf{ad-hoc police service} for notifying crimes, \textsf{Amazon} used as  persistence service and  \textsf{Google Maps} for locating and displaying crimes on a map.
-
-Non-functional requirements are specified by rules and conditions to be verified during the excecution of tasks.
-In our example we have the following non-functional requirements:
-\begin{trivlist}
-\item[1.] Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
-\item[2.] Crime notification needs privileged access.
-\item[3.] Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
-\item[4.] If \textsf{Google Maps} is unavailable, the results are delivered as text. 
-\item[5.] Querying about crimes without having proper clearance yields an empty map.
-\item[6.] Access rights to detailed information depends on user clearance and zone assignment for police officers. 
-\item[7.] The application maintains a detailed log. 
-\end{trivlist}
-
-Considering the example of tracking crimes, all the system restrictions are
-model as constraints in the methodology. We have three types of constraints:
-\textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case can be associated to one or more constraints.
-
-\paragraph*{\textit{$\pi$-UseCase} model:} 
-In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
-system functions (tasks) and constraints. 
-We will not detail the functional part of the specification, due to lack of space.
-The constrains defined for our tracking crime example are:
-
-\begin{trivlist}
-  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
-  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
-  In this case, it represents the maximmum number of allowed login attempts;
-\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
-\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
-This is an example of \textit{business constraint}.
-Additionally the requesting user must be logged in.
-\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text.
-This is an example of \textit{exceptional behaviour constraint}.
-The availability of the Google Maps service is verified by a \textit{business constraint}.
- \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
-\end{trivlist}
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.8\textwidth]{figs/piUseCase}
-\caption{$\pi$-UseCase Model}
-\label{fig:piUC}
-\end{figure}
-\paragraph*{\textit{$\pi$-ServiceProcess} model:}
-The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
-The new model describes the application's activities and defines contracts for each activity or for parts of the application.
-
-A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
-This (semi-automatic) transformation process is supported by a tool (described in~\cite{SouzaNeto:2012}).
-
-The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
-\begin{enumerate}
-\item Tasks of the previous model are transformed into \textit{actions};
-\item Actions are grouped into \textit{activities} (in accordance to the business logic).
-\item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
-\end{enumerate}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
-\caption{$\pi$-\textit{ServiceProcess} Model}
-\label{fig:piSP}
-\end{figure}
-
-\paragraph*{\textit{$\pi$-ServiceComposition} model:} 
-This model refines the previous model by using the activities to produce the workflow of the application.
-The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
-This model identifies the services and functions that correspond to each action in the business process.
-
-\begin{figure}[t]
-\centering
-\subfloat[$\pi$-ServiceComposition Model]
-{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\\
-\subfloat[$\pi$-ServiceComposition Policies]
-{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Service Composition and Policies.}
-\label{fig:policies}
-\end{figure}
-This model describes the service compositions and their related business
-collaborators (external services), policies, rules associated with a policy and
-the whole application process and functionalities.
-The workflow execution will consider the composition of business services and
-their specific Business Collaborators. 
-Each action will trigger the execution of one or more external services. 
-Before and after each service call, policy rules are verified.
-
-In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
-Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{busines collaborators} (external services and entities).
-The interaction occurs by means of function calls (denoted by dotted lines in the figure).
-Figure~\ref{fig:crimePolicies} shows the definition of three \textit{policies}, which define rules for service execution.
-In our case we have policies for \textit{Security}, \textit{Performance} and \textit{Persistence}.
-
-\paragraph*{$\pi$\textit{-PEWS} Model:}
-These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
-This code is a service composition program that can be compiled into executable code.
-$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
-
-\begin{figure}[t]
-\footnotesize
-%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
-%\begin{lstlisting}[label=list:rulePre]
-\begin{verbatim}
-//Namespaces specify service URI
-1 namespace twitter = www.twitter.com/service.wsdl
-2 namespace googlemaps = maps.googleapis.com/maps/api/service
-3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
-4 namespace police = www.police.fr/service.wsdl
-
-//Operations 
-5 alias publishTwitter = portType/publishTwitter in twitter
-6 alias searchCrime = portType/searchCrime in amazondynamodb 
-7 alias showMap = portType/showMap in googlemaps 
-...
-//Services
-8  service notifyCrime = publishCrime  . publishTwitter 
-9  service trackCrime= searchCrime . verifyService
-10 Service visualizeCrime =  showMap . getCrimeDetail
-...
-//Path
-11 ( notifyCrime . storeiCrime ) || 
-12 ( trackCrime . visualizeCrime . getCrimeDetail )
-...
-//Contracts
-13 defContract notifyCrimeContract{
-14	isAppliedTo: notifyCrime
-15	requires: userId == ?? && passw == ?? && 
-16	                               req(notifyCrime)  < 3
-17 	  (OnFailureDo: NOT(action_publish(crime));
-18	ensures: publishTwitter(crime) == true
-19 	  (OnFailureDo: skip);
-20 }
-...
-\end{verbatim}
-% \end{lstlisting}
-\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
-\end{figure}
-
-The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
-The figure shows a simplified program code, produced in accordance to the following guidelines:
-\begin{trivlist}
-\item[1.] Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
-We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
-\item[2.] Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
-\item[3.] The workflow in Figure~\ref{fig:piSC} is translated into a textual representation (lines 11 and 12 of the program).
-\item[4.] \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed in the case thar a condition is not verified.
-Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
-\end{trivlist}
-
-For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.
+\include{proofOfConcept}
 
@@
 \label{sec:conclusions}
-
-Over the last years, a number of approaches have been proposed for the
-development of web services. These approaches range from the proposal of new 
-languages for web service descriptions~\cite{bpel03,SBS04} 
-to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007,BianculliGSBG07}. In general, these
-approaches concentrate on specific problems, like supporting transactions or
-QoS, in order to improve the security and reliability of service-based
-applications. Some proposals address service composition: workflow
-definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
-services~\cite{BHM06}. The proposed solutions come from many communities, 
-including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
-Software Engineering~\cite{Aal03,MendesPDB09}, Programming
-Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
-
-Despite the variety of tecniques proposed, there is not yet a consensus on a software
-process methodology for web services. Some methodologies address the service-based 
-development towards a standard or a new way to develop reliable applications. 
-SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
-S-Cube \cite{scube2010book} is focused on the representation of business processes and 
-service-based development; SOMA \cite{soma} is a methodology described by IBM
-for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
-BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
-methodology for building service-based infrastructure for collaborative
-enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering \cite{sommerville08} applied to SOC.
-
-\bigskip
-
-This paper presented the $\pi$SOD-M software process for specifying and designing service based applications in the presence of some non-functional constraints. 
-Our proposal harness the SOD-M method with constraints, policies and contracts in order to consider non-functional constraints.
-We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.
-
-$\pi$SOD-M is being used in an academic environment.
-So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
-We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
+\include{conclusion}
 

### 2013-10-14 SOFSEM2014/FinalVersion/pisodm.tex
final - chop chop

--- 2013-10-12 SOFSEM2014/Submitted/main.tex
@@
-\documentclass{llncs}
-\usepackage{tikz}
-\usepackage{amsfonts}
-\usepackage{graphicx,url}
-\usepackage[T1]{fontenc}
-%\usepackage{psfig}
-\usepackage{graphics}
-\usepackage{amssymb,amsmath,latexsym}
-\usepackage{listings} 
-\usepackage{subfig}
-
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{example}[theorem]{Example}
-%\newtheorem{algorithm}[theorem]{Algorithm}
-%\newtheorem{definition}[theorem]{Definition}
-
-\usepackage[thmmarks]{ntheorem}
-
-\usepackage{listings}
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{.}
-\theoremindent0cm
-\theoremnumbering{arabic}
-\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{defi}{Definition}%[section]
-
-\theoremheaderfont{\itshape}
-\newtheorem{examp}{Example}%[section]
-
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{}
-\theoremindent0cm
-\theoremnumbering{arabic}
-%\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{algo}{Algorithm}%[section]
-
-\theoremstyle{plain} 
-\theoremsymbol{\ensuremath{\Box}} 
-\theoremseparator{.} 
-\newtheorem{prop}{Property}
-
-\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
-        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
-        true,false,rise,exception,extend,resume,to,return,function}}
-
-\def\B{\hfill{$\Box$}}
-\def\A{\hbox{${\cal A}$}}
-\def\P{\hbox{${\cal P}$}}
-\def\T{\hbox{${\cal T}$}}
-\def\O{\hbox{${\cal O}$}}
-\def\W{\hbox{${\cal W}$}}
-\def\useOpt{\hbox{${has\_opt}$}}
-
-\def\bigS{\hbox{${\cal S}$}}
-\def\bigSd{\hbox{${\cal S}_D$}}
-\def\bigSsm{\hbox{${\cal S}_D^{score}$}}
-
-\newcommand{\ie}{\ifmmode \mathit{i.e.}\else \textit{i.e.}\fi}
-\newcommand{\wrt}{\ifmmode \mathit{w.r.t.}\else \textit{w.r.t.}\fi}
-\newcommand{\eg}{\ifmmode \mathit{e.g.}\else \textit{e.g.}\fi}
-\newcommand{\tuple}[1]{\ifmmode \langle #1 \rangle\else $\langle #1 \rangle$\fi}
-
-\usepackage{authblk}
-
-\begin{document}
-%
-% --- Author Metadata here ---
-
-% --- End of Author Metadata ---
-
-\title{Supporting Non-Functional Requirements for Services Software Process: An MDD Approach}
-
-\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}
-
-\institute{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\  \email{Valeria.deCastro@urjc.es} 
-\and Federal University of Rio Grande do Norte (UFRN) -- Natal-RN, Brazil\\
-\email{\{mam,umberto\}@dimap.ufrn.br}
-\and Federal Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
-\and French Council of Scientific Research (CNRS) -- Grenoble, France\\
-\email{Genoveva.Vargas-Solar@imag.fr}
-}
-
-\maketitle
-\begin{abstract}
-This paper presents an extension to the Service Oriented Development Method (SOD-M) to support the representation of some non-functional requirements. 
-Specifically, we propose to extend SOD-M with: 
-\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
-\textit{(ii)} model to model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
-\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
-We apply our proposal to develop a proof-of-concept example.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
-%performance measures]
-
-%\terms{Delphi theory}
-
-\keywords{MDD, Service Oriented Applications, Non-functional Properties}
-
-\section{Introduction}
-
-Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
-The main ideas of MDD were originaly proposed by the Object Management Group
-(OMG)~\cite{mda}. 
-MDD provides a set of
-guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
-Abstraction levels are called \textit{viewpoints}. 
-MDD defines three viewpoints:
-
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
-These transformations may be automatic or semi-automatic and implement the refinement process between levels.
-
-\bigskip
-
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
-Some methodologies and techniques have been proposed to help the software
-developer in the specification of functional requirements of the business logic,
-such as the Service Oriented Development Method
-(SOD-M)~\cite{decastro1}. 
-SOD-M is based on MDD and proposes
-models, practices and techniques to aid in software development. SOD-M does not
-provide suport to the specification of non-functional requirements, such as
-security, reliability, and efficiency. Ideally, non-functional requirements
-would be considered along with all the stages of the software development. The
-adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-the application context.
-
-\bigskip
-
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-The main goals of our work are:
-\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
-
-\bigskip
-This paper is organized as follows:
-Sections~\ref{sec:sodm} and~\ref{sec:pisodm} presents, respectively, the SOD-M method of service software process and our proposed extension to deal with non-functional requirements.
-A proof of concept example is developed in Section~\ref{sec:poc}.
-We conclude the paper by presenting some related work and final remarks.
-
-\section{SOD-M}\label{sec:sodm}
-The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
-proposes the usage of the MDD approach in the context of service-based applications.
-SOD-M provides a framework with models and standards to express functionalities
-of applications at a high-level of abstraction. SOD-M meta-models are organized 
-into three levels: CIM (\textit{Computational Independent Models}), 
-PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-
-Two models are defined at the CIM level: \textit{value model} 
-and \textit{BPMN model}. 
-The PIM level models the entire structure of the application flow,
-while, the PSM level provides transformations towards more specific platforms.
-The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
-\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
-These three levels have no support for describing non-functional requirements.
-
-The SOD-M approach includes transformations between models:
-\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
-an abstract model at the CIM level, it is possible to apply transformations for
-generating a model of the PSM level. In this context, it is necessary to
-follow the process activities described by the methodology.
-
-SOD-M considers two points of view:
-\textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
-this way, SOD-M aims to simplify the design of service-oriented applications, as
-well as its implementation using current technologies.
-
-\section{$\pi$SOD-M}\label{sec:pisodm}
-
 %In this section we present $\pi$SOD-M, an MDD based methodology. 
@@
 The example will show the actual notation used for models.
-
-\section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
-
-Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
-Civilian users signal crimes using Twitter.
-Police officers can notify crimes, as well as update information about solving cases.
-Some of these information are confidential while other can be shared
-to the community of users using this application.
-
-Users can track crimes in given zones. 
-Crime information stored by the system may be visualized on a map. 
-Some users have different access rights than others.
-For example, police officers have more access rights than civilians.
-
-In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize the information.
-The business process defines the logic of the application and is specified in terms of tasks. 
-Tasks can be performed by persons or computers.
-
-The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
-Busines process is represented as a graph while requirements are given as text boxes.
-\begin{figure}[ht!]
-\centering
-%\subfloat[\textit{Tracking Crimes} - Illustration of the Problem.]
-%{\label{fig:problem}\includegraphics[width=0.95\textwidth]{figs/problem}}
-%~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-%\\
-%\subfloat[Contraints.]
-{\label{fig:trans06}\includegraphics[width=0.95\textwidth]{figs/constraints}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Business process for the tracking crime example.}
-\label{fig:constraints}
-\end{figure}
-
-In our example, crime processing can start with one of two tasks: \textit{(i) notify a crime}, or \textit{(ii) track a crime}.
-Notified crimes are stored in a database. 
-Tracked crimes are visualized in a map.
-The used can ask for detailed information. 
-The application is built upon four services: \textsf{twitter} and an \textsf{ad-hoc police service} for notifying crimes, \textsf{Amazon} used as  persistence service and  \textsf{Google Maps} for locating and displaying crimes on a map.
-
-Non-functional requirements are specified by rules and conditions to be verified during the excecution of tasks.
-In our example we have the following non-functional requirements:
-\begin{trivlist}
-\item[1.] Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
-\item[2.] Crime notification needs privileged access.
-\item[3.] Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
-\item[4.] If \textsf{Google Maps} is unavailable, the results are delivered as text. 
-\item[5.] Querying about crimes without having proper clearance yields an empty map.
-\item[6.] Access rights to detailed information depends on user clearance and zone assignment for police officers. 
-\item[7.] The application maintains a detailed log. 
-\end{trivlist}
-
-Considering the example of tracking crimes, all the system restrictions are
-model as constraints in the methodology. We have three types of constraints:
-\textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case can be associated to one or more constraints.
-
-\paragraph*{\textit{$\pi$-UseCase} model:} 
-In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
-system functions (tasks) and constraints. 
-We will not detail the functional part of the specification, due to lack of space.
-The constrains defined for our tracking crime example are:
-
-\begin{trivlist}
-  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
-  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
-  In this case, it represents the maximmum number of allowed login attempts;
-\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
-\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
-This is an example of \textit{business constraint}.
-Additionally the requesting user must be logged in.
-\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text.
-This is an example of \textit{exceptional behaviour constraint}.
-The availability of the Google Maps service is verified by a \textit{business constraint}.
- \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
-\end{trivlist}
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.8\textwidth]{figs/piUseCase}
-\caption{$\pi$-UseCase Model}
-\label{fig:piUC}
-\end{figure}
-\paragraph*{\textit{$\pi$-ServiceProcess} model:}
-The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
-The new model describes the application's activities and defines contracts for each activity or for parts of the application.
-
-A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
-This (semi-automatic) transformation process is supported by a tool (described in~\cite{SouzaNeto:2012}).
-
-The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
-\begin{enumerate}
-\item Tasks of the previous model are transformed into \textit{actions};
-\item Actions are grouped into \textit{activities} (in accordance to the business logic).
-\item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
-\end{enumerate}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
-\caption{$\pi$-\textit{ServiceProcess} Model}
-\label{fig:piSP}
-\end{figure}
-
-\paragraph*{\textit{$\pi$-ServiceComposition} model:} 
-This model refines the previous model by using the activities to produce the workflow of the application.
-The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
-This model identifies the services and functions that correspond to each action in the business process.
-
-\begin{figure}[t]
-\centering
-\subfloat[$\pi$-ServiceComposition Model]
-{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\\
-\subfloat[$\pi$-ServiceComposition Policies]
-{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Service Composition and Policies.}
-\label{fig:policies}
-\end{figure}
-This model describes the service compositions and their related business
-collaborators (external services), policies, rules associated with a policy and
-the whole application process and functionalities.
-The workflow execution will consider the composition of business services and
-their specific Business Collaborators. 
-Each action will trigger the execution of one or more external services. 
-Before and after each service call, policy rules are verified.
-
-In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
-Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{busines collaborators} (external services and entities).
-The interaction occurs by means of function calls (denoted by dotted lines in the figure).
-Figure~\ref{fig:crimePolicies} shows the definition of three \textit{policies}, which define rules for service execution.
-In our case we have policies for \textit{Security}, \textit{Performance} and \textit{Persistence}.
-
-\paragraph*{$\pi$\textit{-PEWS} Model:}
-These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
-This code is a service composition program that can be compiled into executable code.
-$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
-
-\begin{figure}[t]
-\footnotesize
-%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
-%\begin{lstlisting}[label=list:rulePre]
-\begin{verbatim}
-//Namespaces specify service URI
-1 namespace twitter = www.twitter.com/service.wsdl
-2 namespace googlemaps = maps.googleapis.com/maps/api/service
-3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
-4 namespace police = www.police.fr/service.wsdl
-
-//Operations 
-5 alias publishTwitter = portType/publishTwitter in twitter
-6 alias searchCrime = portType/searchCrime in amazondynamodb 
-7 alias showMap = portType/showMap in googlemaps 
-...
-//Services
-8  service notifyCrime = publishCrime  . publishTwitter 
-9  service trackCrime= searchCrime . verifyService
-10 Service visualizeCrime =  showMap . getCrimeDetail
-...
-//Path
-11 ( notifyCrime . storeiCrime ) || 
-12 ( trackCrime . visualizeCrime . getCrimeDetail )
-...
-//Contracts
-13 defContract notifyCrimeContract{
-14	isAppliedTo: notifyCrime
-15	requires: userId == ?? && passw == ?? && 
-16	                               req(notifyCrime)  < 3
-17 	  (OnFailureDo: NOT(action_publish(crime));
-18	ensures: publishTwitter(crime) == true
-19 	  (OnFailureDo: skip);
-20 }
-...
-\end{verbatim}
-% \end{lstlisting}
-\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
-\end{figure}
-
-The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
-The figure shows a simplified program code, produced in accordance to the following guidelines:
-\begin{trivlist}
-\item[1.] Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
-We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
-\item[2.] Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
-\item[3.] The workflow in Figure~\ref{fig:piSC} is translated into a textual representation (lines 11 and 12 of the program).
-\item[4.] \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed in the case thar a condition is not verified.
-Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
-\end{trivlist}
-
-For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.
-
-\section{Related Work and Conclusions}
-\label{sec:conclusions}
-
-Over the last years, a number of approaches have been proposed for the
-development of web services. These approaches range from the proposal of new 
-languages for web service descriptions~\cite{bpel03,SBS04} 
-to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007,BianculliGSBG07}. In general, these
-approaches concentrate on specific problems, like supporting transactions or
-QoS, in order to improve the security and reliability of service-based
-applications. Some proposals address service composition: workflow
-definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
-services~\cite{BHM06}. The proposed solutions come from many communities, 
-including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
-Software Engineering~\cite{Aal03,MendesPDB09}, Programming
-Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
-
-Despite the variety of tecniques proposed, there is not yet a consensus on a software
-process methodology for web services. Some methodologies address the service-based 
-development towards a standard or a new way to develop reliable applications. 
-SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
-S-Cube \cite{scube2010book} is focused on the representation of business processes and 
-service-based development; SOMA \cite{soma} is a methodology described by IBM
-for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
-BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
-methodology for building service-based infrastructure for collaborative
-enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering \cite{sommerville08} applied to SOC.
-
-\bigskip
-
-This paper presented the $\pi$SOD-M software process for specifying and designing service based applications in the presence of some non-functional constraints. 
-Our proposal harness the SOD-M method with constraints, policies and contracts in order to consider non-functional constraints.
-We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.
-
-$\pi$SOD-M is being used in an academic environment.
-So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
-We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
-
-\bibliographystyle{splncs_srt}
-\bibliography{biblio2} 
-\end{document}

### 2013-10-14 SOFSEM2014/FinalVersion/proofOfConcept.tex
final - chop chop

--- 2013-10-12 SOFSEM2014/Submitted/main.tex
@@
-\documentclass{llncs}
-\usepackage{tikz}
-\usepackage{amsfonts}
-\usepackage{graphicx,url}
-\usepackage[T1]{fontenc}
-%\usepackage{psfig}
-\usepackage{graphics}
-\usepackage{amssymb,amsmath,latexsym}
-\usepackage{listings} 
-\usepackage{subfig}
-
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{example}[theorem]{Example}
-%\newtheorem{algorithm}[theorem]{Algorithm}
-%\newtheorem{definition}[theorem]{Definition}
-
-\usepackage[thmmarks]{ntheorem}
-
-\usepackage{listings}
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{.}
-\theoremindent0cm
-\theoremnumbering{arabic}
-\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{defi}{Definition}%[section]
-
-\theoremheaderfont{\itshape}
-\newtheorem{examp}{Example}%[section]
-
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{}
-\theoremindent0cm
-\theoremnumbering{arabic}
-%\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{algo}{Algorithm}%[section]
-
-\theoremstyle{plain} 
-\theoremsymbol{\ensuremath{\Box}} 
-\theoremseparator{.} 
-\newtheorem{prop}{Property}
-
-\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
-        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
-        true,false,rise,exception,extend,resume,to,return,function}}
-
-\def\B{\hfill{$\Box$}}
-\def\A{\hbox{${\cal A}$}}
-\def\P{\hbox{${\cal P}$}}
-\def\T{\hbox{${\cal T}$}}
-\def\O{\hbox{${\cal O}$}}
-\def\W{\hbox{${\cal W}$}}
-\def\useOpt{\hbox{${has\_opt}$}}
-
-\def\bigS{\hbox{${\cal S}$}}
-\def\bigSd{\hbox{${\cal S}_D$}}
-\def\bigSsm{\hbox{${\cal S}_D^{score}$}}
-
-\newcommand{\ie}{\ifmmode \mathit{i.e.}\else \textit{i.e.}\fi}
-\newcommand{\wrt}{\ifmmode \mathit{w.r.t.}\else \textit{w.r.t.}\fi}
-\newcommand{\eg}{\ifmmode \mathit{e.g.}\else \textit{e.g.}\fi}
-\newcommand{\tuple}[1]{\ifmmode \langle #1 \rangle\else $\langle #1 \rangle$\fi}
-
-\usepackage{authblk}
-
-\begin{document}
-%
-% --- Author Metadata here ---
-
-% --- End of Author Metadata ---
-
-\title{Supporting Non-Functional Requirements for Services Software Process: An MDD Approach}
-
-\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}
-
-\institute{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\  \email{Valeria.deCastro@urjc.es} 
-\and Federal University of Rio Grande do Norte (UFRN) -- Natal-RN, Brazil\\
-\email{\{mam,umberto\}@dimap.ufrn.br}
-\and Federal Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
-\and French Council of Scientific Research (CNRS) -- Grenoble, France\\
-\email{Genoveva.Vargas-Solar@imag.fr}
-}
-
-\maketitle
-\begin{abstract}
-This paper presents an extension to the Service Oriented Development Method (SOD-M) to support the representation of some non-functional requirements. 
-Specifically, we propose to extend SOD-M with: 
-\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
-\textit{(ii)} model to model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
-\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
-We apply our proposal to develop a proof-of-concept example.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
-%performance measures]
-
-%\terms{Delphi theory}
-
-\keywords{MDD, Service Oriented Applications, Non-functional Properties}
-
-\section{Introduction}
-
-Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
-The main ideas of MDD were originaly proposed by the Object Management Group
-(OMG)~\cite{mda}. 
-MDD provides a set of
-guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
-Abstraction levels are called \textit{viewpoints}. 
-MDD defines three viewpoints:
-
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
-These transformations may be automatic or semi-automatic and implement the refinement process between levels.
-
-\bigskip
-
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
-Some methodologies and techniques have been proposed to help the software
-developer in the specification of functional requirements of the business logic,
-such as the Service Oriented Development Method
-(SOD-M)~\cite{decastro1}. 
-SOD-M is based on MDD and proposes
-models, practices and techniques to aid in software development. SOD-M does not
-provide suport to the specification of non-functional requirements, such as
-security, reliability, and efficiency. Ideally, non-functional requirements
-would be considered along with all the stages of the software development. The
-adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-the application context.
-
-\bigskip
-
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-The main goals of our work are:
-\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
-
-\bigskip
-This paper is organized as follows:
-Sections~\ref{sec:sodm} and~\ref{sec:pisodm} presents, respectively, the SOD-M method of service software process and our proposed extension to deal with non-functional requirements.
-A proof of concept example is developed in Section~\ref{sec:poc}.
-We conclude the paper by presenting some related work and final remarks.
-
-\section{SOD-M}\label{sec:sodm}
-The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
-proposes the usage of the MDD approach in the context of service-based applications.
-SOD-M provides a framework with models and standards to express functionalities
-of applications at a high-level of abstraction. SOD-M meta-models are organized 
-into three levels: CIM (\textit{Computational Independent Models}), 
-PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-
-Two models are defined at the CIM level: \textit{value model} 
-and \textit{BPMN model}. 
-The PIM level models the entire structure of the application flow,
-while, the PSM level provides transformations towards more specific platforms.
-The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
-\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}. 
-These three levels have no support for describing non-functional requirements.
-
-The SOD-M approach includes transformations between models:
-\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
-an abstract model at the CIM level, it is possible to apply transformations for
-generating a model of the PSM level. In this context, it is necessary to
-follow the process activities described by the methodology.
-
-SOD-M considers two points of view:
-\textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
-this way, SOD-M aims to simplify the design of service-oriented applications, as
-well as its implementation using current technologies.
-
-\section{$\pi$SOD-M}\label{sec:pisodm}
-
-%In this section we present $\pi$SOD-M, an MDD based methodology. 
-$\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. 
-$\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
-\caption{$\pi$SOD-M.}
-\label{fig:piSOD-M}
-\end{figure}
-
-$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
-well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
-Constraints are restrictions that must be verified during the execution of the application. 
-An example of this is the requirement of the user's authentication for executing some system functions.
-
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-modes) and transformations between models.
-There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
-
-We extended SOD-M to include non-functional specifications.
-Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
-
-The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data. 
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
-The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services. 
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
-which put together contracts with similar non-functional requirements. 
-For instance, security and privacy restrictions may be grouped into a security policy.
-\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
-
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
-Although PEWS is our language of choice, other composition languages can be used as target.
-This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
-
-In the next section we develop an example, to serve as a proof-of-concept.
-The example will show the actual notation used for models.
-
-\section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
-
 Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
@@
 For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.
-
-\section{Related Work and Conclusions}
-\label{sec:conclusions}
-
-Over the last years, a number of approaches have been proposed for the
-development of web services. These approaches range from the proposal of new 
-languages for web service descriptions~\cite{bpel03,SBS04} 
-to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007,BianculliGSBG07}. In general, these
-approaches concentrate on specific problems, like supporting transactions or
-QoS, in order to improve the security and reliability of service-based
-applications. Some proposals address service composition: workflow
-definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
-services~\cite{BHM06}. The proposed solutions come from many communities, 
-including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
-Software Engineering~\cite{Aal03,MendesPDB09}, Programming
-Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
-
-Despite the variety of tecniques proposed, there is not yet a consensus on a software
-process methodology for web services. Some methodologies address the service-based 
-development towards a standard or a new way to develop reliable applications. 
-SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
-S-Cube \cite{scube2010book} is focused on the representation of business processes and 
-service-based development; SOMA \cite{soma} is a methodology described by IBM
-for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
-BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
-methodology for building service-based infrastructure for collaborative
-enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering \cite{sommerville08} applied to SOC.
-
-\bigskip
-
-This paper presented the $\pi$SOD-M software process for specifying and designing service based applications in the presence of some non-functional constraints. 
-Our proposal harness the SOD-M method with constraints, policies and contracts in order to consider non-functional constraints.
-We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.
-
-$\pi$SOD-M is being used in an academic environment.
-So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
-We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
-
-\bibliographystyle{splncs_srt}
-\bibliography{biblio2} 
-\end{document}

### 2013-10-14 SOFSEM2014/FinalVersion/sodm.tex
final - chop chop

--- 2013-10-12 SOFSEM2014/Submitted/main.tex
@@
-\documentclass{llncs}
-\usepackage{tikz}
-\usepackage{amsfonts}
-\usepackage{graphicx,url}
-\usepackage[T1]{fontenc}
-%\usepackage{psfig}
-\usepackage{graphics}
-\usepackage{amssymb,amsmath,latexsym}
-\usepackage{listings} 
-\usepackage{subfig}
-
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{example}[theorem]{Example}
-%\newtheorem{algorithm}[theorem]{Algorithm}
-%\newtheorem{definition}[theorem]{Definition}
-
-\usepackage[thmmarks]{ntheorem}
-
-\usepackage{listings}
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{.}
-\theoremindent0cm
-\theoremnumbering{arabic}
-\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{defi}{Definition}%[section]
-
-\theoremheaderfont{\itshape}
-\newtheorem{examp}{Example}%[section]
-
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{}
-\theoremindent0cm
-\theoremnumbering{arabic}
-%\theoremsymbol{\ensuremath{\Box}} 
-\newtheorem{algo}{Algorithm}%[section]
-
-\theoremstyle{plain} 
-\theoremsymbol{\ensuremath{\Box}} 
-\theoremseparator{.} 
-\newtheorem{prop}{Property}
-
-\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
-        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
-        true,false,rise,exception,extend,resume,to,return,function}}
-
-\def\B{\hfill{$\Box$}}
-\def\A{\hbox{${\cal A}$}}
-\def\P{\hbox{${\cal P}$}}
-\def\T{\hbox{${\cal T}$}}
-\def\O{\hbox{${\cal O}$}}
-\def\W{\hbox{${\cal W}$}}
-\def\useOpt{\hbox{${has\_opt}$}}
-
-\def\bigS{\hbox{${\cal S}$}}
-\def\bigSd{\hbox{${\cal S}_D$}}
-\def\bigSsm{\hbox{${\cal S}_D^{score}$}}
-
-\newcommand{\ie}{\ifmmode \mathit{i.e.}\else \textit{i.e.}\fi}
-\newcommand{\wrt}{\ifmmode \mathit{w.r.t.}\else \textit{w.r.t.}\fi}
-\newcommand{\eg}{\ifmmode \mathit{e.g.}\else \textit{e.g.}\fi}
-\newcommand{\tuple}[1]{\ifmmode \langle #1 \rangle\else $\langle #1 \rangle$\fi}
-
-\usepackage{authblk}
-
-\begin{document}
-%
-% --- Author Metadata here ---
-
-% --- End of Author Metadata ---
-
-\title{Supporting Non-Functional Requirements for Services Software Process: An MDD Approach}
-
-\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}
-
-\institute{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\  \email{Valeria.deCastro@urjc.es} 
-\and Federal University of Rio Grande do Norte (UFRN) -- Natal-RN, Brazil\\
-\email{\{mam,umberto\}@dimap.ufrn.br}
-\and Federal Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
-\and French Council of Scientific Research (CNRS) -- Grenoble, France\\
-\email{Genoveva.Vargas-Solar@imag.fr}
-}
-
-\maketitle
-\begin{abstract}
-This paper presents an extension to the Service Oriented Development Method (SOD-M) to support the representation of some non-functional requirements. 
-Specifically, we propose to extend SOD-M with: 
-\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
-\textit{(ii)} model to model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
-\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
-We apply our proposal to develop a proof-of-concept example.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
-%performance measures]
-
-%\terms{Delphi theory}
-
-\keywords{MDD, Service Oriented Applications, Non-functional Properties}
-
-\section{Introduction}
-
-Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
-The main ideas of MDD were originaly proposed by the Object Management Group
-(OMG)~\cite{mda}. 
-MDD provides a set of
-guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
-Abstraction levels are called \textit{viewpoints}. 
-MDD defines three viewpoints:
-
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
-These transformations may be automatic or semi-automatic and implement the refinement process between levels.
-
-\bigskip
-
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
-Some methodologies and techniques have been proposed to help the software
-developer in the specification of functional requirements of the business logic,
-such as the Service Oriented Development Method
-(SOD-M)~\cite{decastro1}. 
-SOD-M is based on MDD and proposes
-models, practices and techniques to aid in software development. SOD-M does not
-provide suport to the specification of non-functional requirements, such as
-security, reliability, and efficiency. Ideally, non-functional requirements
-would be considered along with all the stages of the software development. The
-adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-the application context.
-
-\bigskip
-
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-The main goals of our work are:
-\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
-
-\bigskip
-This paper is organized as follows:
-Sections~\ref{sec:sodm} and~\ref{sec:pisodm} presents, respectively, the SOD-M method of service software process and our proposed extension to deal with non-functional requirements.
-A proof of concept example is developed in Section~\ref{sec:poc}.
-We conclude the paper by presenting some related work and final remarks.
-
-\section{SOD-M}\label{sec:sodm}
 The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
@@
 well as its implementation using current technologies.
-
-\section{$\pi$SOD-M}\label{sec:pisodm}
-
-%In this section we present $\pi$SOD-M, an MDD based methodology. 
-$\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. 
-$\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
-\caption{$\pi$SOD-M.}
-\label{fig:piSOD-M}
-\end{figure}
-
-$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
-well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
-Constraints are restrictions that must be verified during the execution of the application. 
-An example of this is the requirement of the user's authentication for executing some system functions.
-
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-modes) and transformations between models.
-There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
-
-We extended SOD-M to include non-functional specifications.
-Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
-
-The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data. 
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
-The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services. 
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
-which put together contracts with similar non-functional requirements. 
-For instance, security and privacy restrictions may be grouped into a security policy.
-\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
-
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
-Although PEWS is our language of choice, other composition languages can be used as target.
-This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
-
-In the next section we develop an example, to serve as a proof-of-concept.
-The example will show the actual notation used for models.
-
-\section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
-
-Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
-Civilian users signal crimes using Twitter.
-Police officers can notify crimes, as well as update information about solving cases.
-Some of these information are confidential while other can be shared
-to the community of users using this application.
-
-Users can track crimes in given zones. 
-Crime information stored by the system may be visualized on a map. 
-Some users have different access rights than others.
-For example, police officers have more access rights than civilians.
-
-In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize the information.
-The business process defines the logic of the application and is specified in terms of tasks. 
-Tasks can be performed by persons or computers.
-
-The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
-Busines process is represented as a graph while requirements are given as text boxes.
-\begin{figure}[ht!]
-\centering
-%\subfloat[\textit{Tracking Crimes} - Illustration of the Problem.]
-%{\label{fig:problem}\includegraphics[width=0.95\textwidth]{figs/problem}}
-%~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-%\\
-%\subfloat[Contraints.]
-{\label{fig:trans06}\includegraphics[width=0.95\textwidth]{figs/constraints}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Business process for the tracking crime example.}
-\label{fig:constraints}
-\end{figure}
-
-In our example, crime processing can start with one of two tasks: \textit{(i) notify a crime}, or \textit{(ii) track a crime}.
-Notified crimes are stored in a database. 
-Tracked crimes are visualized in a map.
-The used can ask for detailed information. 
-The application is built upon four services: \textsf{twitter} and an \textsf{ad-hoc police service} for notifying crimes, \textsf{Amazon} used as  persistence service and  \textsf{Google Maps} for locating and displaying crimes on a map.
-
-Non-functional requirements are specified by rules and conditions to be verified during the excecution of tasks.
-In our example we have the following non-functional requirements:
-\begin{trivlist}
-\item[1.] Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
-\item[2.] Crime notification needs privileged access.
-\item[3.] Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
-\item[4.] If \textsf{Google Maps} is unavailable, the results are delivered as text. 
-\item[5.] Querying about crimes without having proper clearance yields an empty map.
-\item[6.] Access rights to detailed information depends on user clearance and zone assignment for police officers. 
-\item[7.] The application maintains a detailed log. 
-\end{trivlist}
-
-Considering the example of tracking crimes, all the system restrictions are
-model as constraints in the methodology. We have three types of constraints:
-\textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case can be associated to one or more constraints.
-
-\paragraph*{\textit{$\pi$-UseCase} model:} 
-In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
-system functions (tasks) and constraints. 
-We will not detail the functional part of the specification, due to lack of space.
-The constrains defined for our tracking crime example are:
-
-\begin{trivlist}
-  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
-  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
-  In this case, it represents the maximmum number of allowed login attempts;
-\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
-\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
-This is an example of \textit{business constraint}.
-Additionally the requesting user must be logged in.
-\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text.
-This is an example of \textit{exceptional behaviour constraint}.
-The availability of the Google Maps service is verified by a \textit{business constraint}.
- \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
-\end{trivlist}
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.8\textwidth]{figs/piUseCase}
-\caption{$\pi$-UseCase Model}
-\label{fig:piUC}
-\end{figure}
-\paragraph*{\textit{$\pi$-ServiceProcess} model:}
-The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
-The new model describes the application's activities and defines contracts for each activity or for parts of the application.
-
-A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
-This (semi-automatic) transformation process is supported by a tool (described in~\cite{SouzaNeto:2012}).
-
-The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
-\begin{enumerate}
-\item Tasks of the previous model are transformed into \textit{actions};
-\item Actions are grouped into \textit{activities} (in accordance to the business logic).
-\item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
-\end{enumerate}
-
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
-\caption{$\pi$-\textit{ServiceProcess} Model}
-\label{fig:piSP}
-\end{figure}
-
-\paragraph*{\textit{$\pi$-ServiceComposition} model:} 
-This model refines the previous model by using the activities to produce the workflow of the application.
-The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
-This model identifies the services and functions that correspond to each action in the business process.
-
-\begin{figure}[t]
-\centering
-\subfloat[$\pi$-ServiceComposition Model]
-{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\\
-\subfloat[$\pi$-ServiceComposition Policies]
-{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
-~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\caption{Service Composition and Policies.}
-\label{fig:policies}
-\end{figure}
-This model describes the service compositions and their related business
-collaborators (external services), policies, rules associated with a policy and
-the whole application process and functionalities.
-The workflow execution will consider the composition of business services and
-their specific Business Collaborators. 
-Each action will trigger the execution of one or more external services. 
-Before and after each service call, policy rules are verified.
-
-In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
-Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{busines collaborators} (external services and entities).
-The interaction occurs by means of function calls (denoted by dotted lines in the figure).
-Figure~\ref{fig:crimePolicies} shows the definition of three \textit{policies}, which define rules for service execution.
-In our case we have policies for \textit{Security}, \textit{Performance} and \textit{Persistence}.
-
-\paragraph*{$\pi$\textit{-PEWS} Model:}
-These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
-This code is a service composition program that can be compiled into executable code.
-$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
-
-\begin{figure}[t]
-\footnotesize
-%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
-%\begin{lstlisting}[label=list:rulePre]
-\begin{verbatim}
-//Namespaces specify service URI
-1 namespace twitter = www.twitter.com/service.wsdl
-2 namespace googlemaps = maps.googleapis.com/maps/api/service
-3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
-4 namespace police = www.police.fr/service.wsdl
-
-//Operations 
-5 alias publishTwitter = portType/publishTwitter in twitter
-6 alias searchCrime = portType/searchCrime in amazondynamodb 
-7 alias showMap = portType/showMap in googlemaps 
-...
-//Services
-8  service notifyCrime = publishCrime  . publishTwitter 
-9  service trackCrime= searchCrime . verifyService
-10 Service visualizeCrime =  showMap . getCrimeDetail
-...
-//Path
-11 ( notifyCrime . storeiCrime ) || 
-12 ( trackCrime . visualizeCrime . getCrimeDetail )
-...
-//Contracts
-13 defContract notifyCrimeContract{
-14	isAppliedTo: notifyCrime
-15	requires: userId == ?? && passw == ?? && 
-16	                               req(notifyCrime)  < 3
-17 	  (OnFailureDo: NOT(action_publish(crime));
-18	ensures: publishTwitter(crime) == true
-19 	  (OnFailureDo: skip);
-20 }
-...
-\end{verbatim}
-% \end{lstlisting}
-\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
-\end{figure}
-
-The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
-The figure shows a simplified program code, produced in accordance to the following guidelines:
-\begin{trivlist}
-\item[1.] Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
-We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
-\item[2.] Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
-\item[3.] The workflow in Figure~\ref{fig:piSC} is translated into a textual representation (lines 11 and 12 of the program).
-\item[4.] \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed in the case thar a condition is not verified.
-Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
-\end{trivlist}
-
-For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.
-
-\section{Related Work and Conclusions}
-\label{sec:conclusions}
-
-Over the last years, a number of approaches have been proposed for the
-development of web services. These approaches range from the proposal of new 
-languages for web service descriptions~\cite{bpel03,SBS04} 
-to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007,BianculliGSBG07}. In general, these
-approaches concentrate on specific problems, like supporting transactions or
-QoS, in order to improve the security and reliability of service-based
-applications. Some proposals address service composition: workflow
-definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
-services~\cite{BHM06}. The proposed solutions come from many communities, 
-including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
-Software Engineering~\cite{Aal03,MendesPDB09}, Programming
-Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
-
-Despite the variety of tecniques proposed, there is not yet a consensus on a software
-process methodology for web services. Some methodologies address the service-based 
-development towards a standard or a new way to develop reliable applications. 
-SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
-S-Cube \cite{scube2010book} is focused on the representation of business processes and 
-service-based development; SOMA \cite{soma} is a methodology described by IBM
-for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
-BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
-methodology for building service-based infrastructure for collaborative
-enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering \cite{sommerville08} applied to SOC.
-
-\bigskip
-
-This paper presented the $\pi$SOD-M software process for specifying and designing service based applications in the presence of some non-functional constraints. 
-Our proposal harness the SOD-M method with constraints, policies and contracts in order to consider non-functional constraints.
-We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.
-
-$\pi$SOD-M is being used in an academic environment.
-So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
-We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
-
-\bibliographystyle{splncs_srt}
-\bibliography{biblio2} 
-\end{document}

### 2013-10-14 SOFSEM2014/FinalVersion/pisodm.tex
begin corrections --martin

@@
 to represent non-functional requirements.
+
+Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
+Constraints are restrictions that must be verified during the execution of the application. 
+An example of this is the requirement of authentication for executing some of the system functions. 
+Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
+
+In this context, we provide concepts at different levels of abstraction for expressing business rules of a process as policies (to implement existing NFP protocols).
 
@@
 well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints. 
-Constraints are restrictions that must be verified during the execution of the application. 
-An example of this is the requirement of the user's authentication for executing some system functions.
+$\pi$SOD-M models represent both the functional aspects of the application as well as their non-functional behaviour.
 

### 2013-10-14 SOFSEM2014/FinalVersion/main.tex
begin corrections --martin
Ack -- martin

@@
 \label{sec:intro}
-\include{introduction}
+\input{introduction}
 
 \section{SOD-M}\label{sec:sodm}
-\include{sodm}
+\input{sodm}
 
 \section{$\pi$SOD-M}\label{sec:pisodm}
-\include{pisodm}
+\input{pisodm}
 
 \section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
-\include{proofOfConcept}
+\input{proofOfConcept}
 
@@
 \label{sec:conclusions}
-\include{conclusion}
+\input{conclusion}
+
+\section*{Acknowledgements}
+This work was partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grant 573964/2008-4; CAPES/UdelaR (Brazil/Uruguay) grant 021/2010; CAPES/STIC-AmSud (Brazil) grant 020/2010).
 

### 2013-10-15 SOFSEM2014/FinalVersion/main.tex
Ack -- martin

--- 2013-10-14 SOFSEM2014/FinalVersion/main.tex
@@
 \label{sec:conclusions}
-\include{conclusion}
+\input{conclusion}
+
+\section*{Acknowledgements}
+This work was partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/2010; CAPES/STIC-AmSud (Brazil) grant 020/2010).
 

### 2013-10-17 SOFSEM2014/FinalVersion/main.tex
Ack -- martin

--- 2013-10-14 SOFSEM2014/FinalVersion/main.tex
@@
 \label{sec:conclusions}
-\include{conclusion}
+\input{conclusion}
+
+\section*{Acknowledgements}
+This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financied by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy.
 

### 2013-10-17 SOFSEM2014/FinalVersion/conclusion.tex
minor changes (and 12 pages!) --Martin

@@
 to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007,BianculliGSBG07}. In general, these
+software~\cite{lipari2007}. In general, these
 approaches concentrate on specific problems, like supporting transactions or
@@
 
-Despite the variety of tecniques proposed, there is not yet a consensus on a software
+Despite the variety of techniques proposed, there is not yet a consensus on a software
 process methodology for web services. Some methodologies address the service-based 

### 2013-10-17 SOFSEM2014/FinalVersion/proofOfConcept.tex
minor changes (and 12 pages!) --Martin

@@
 In our example we have the following non-functional requirements:
-\begin{trivlist}
-\item[1.] Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
-\item[2.] Crime notification needs privileged access.
-\item[3.] Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
-\item[4.] If \textsf{Google Maps} is unavailable, the results are delivered as text. 
-\item[5.] Querying about crimes without having proper clearance yields an empty map.
-\item[6.] Access rights to detailed information depends on user clearance and zone assignment for police officers. 
-\item[7.] The application maintains a detailed log. 
-\end{trivlist}
+\begin{numtrivlist}
+\item Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
+\item Crime notification needs privileged access.
+\item Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
+\item If \textsf{Google Maps} is unavailable, the results are delivered as text. 
+\item Querying about crimes without having proper clearance yields an empty map.
+\item Access rights to detailed information depends on user clearance and zone assignment for police officers. 
+\item The application maintains a detailed log. 
+\end{numtrivlist}
 
 Considering the example of tracking crimes, all the system restrictions are
-model as constraints in the methodology. We have three types of constraints:
+modelled as constraints. 
+We have three types of constraints:
 \textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case can be associated to one or more constraints.
+Each use case (model) can be associated to one or more constraints.
 
@@
 A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
-This (semi-automatic) transformation process is supported by a tool (described in~\cite{SouzaNeto:2012}).
+This (semi-automatic) transformation process is supported by a tool (described in~\cite{PlacidoThesis2012}).
 
 The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
-\begin{enumerate}
+\begin{numtrivlist}
 \item Tasks of the previous model are transformed into \textit{actions};
@@
 \item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
-\end{enumerate}
+\end{numtrivlist}
 
@@
 \begin{figure}[t]
-\footnotesize
+\begin{footnotesize}
 %\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
@@
 4 namespace police = www.police.fr/service.wsdl
-
 //Operations 
@@
 //Path
-11 ( notifyCrime . storeiCrime ) || 
-12 ( trackCrime . visualizeCrime . getCrimeDetail )
+11 (notifyCrime.storeiCrime) || (trackCrime.visualizeCrime.getCrimeDetail)
 ...
 //Contracts
-13 defContract notifyCrimeContract{
-14	isAppliedTo: notifyCrime
-15	requires: userId == ?? && passw == ?? && 
-16	                               req(notifyCrime)  < 3
-17 	  (OnFailureDo: NOT(action_publish(crime));
-18	ensures: publishTwitter(crime) == true
-19 	  (OnFailureDo: skip);
-20 }
-...
+12 defContract notifyCrimeContract{ 
+13    isAppliedTo: notifyCrime
+14	  requires: userId == ?? && passw == ?? && req(notifyCrime)  < 3
+15 	      (OnFailureDo: NOT(action_publish(crime));
+16	  ensures: publishTwitter(crime) == true
+17 	      (OnFailureDo: skip);
+28 }
 \end{verbatim}
 % \end{lstlisting}
+\end{footnotesize}
 \caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
@@
 The figure shows a simplified program code, produced in accordance to the following guidelines:
-\begin{trivlist}
-\item[1.] Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
+\begin{numtrivlist}
+\item Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
 We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
-\item[2.] Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
-\item[3.] The workflow in Figure~\ref{fig:piSC} is translated into a textual representation (lines 11 and 12 of the program).
-\item[4.] \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed in the case thar a condition is not verified.
+\item Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
+\item The workflow in Figure~\ref{fig:piSC} is translated into the text in line 11.
+\item \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed case a condition is not verified.
 Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
-\end{trivlist}
+\end{numtrivlist}
 
-For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{SouzaNeto:2012}.
+For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{PlacidoThesis2012}.

### 2013-10-19 SOFSEM2014/FinalVersion/introduction.tex
minor changes everywere -- Martin

@@
 In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provides the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
+combined to produce applications and provide the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
 Some methodologies and techniques have been proposed to help the software
 developer in the specification of functional requirements of the business logic,
-such as the Service Oriented Development Method
+such as the Service-Oriented Development Method
 (SOD-M)~\cite{decastro1}. 
@@
 
-Non-functional properties of service oriented applications have been
+Non-functional properties of service-oriented applications have been
 addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
@@
 in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
+of service APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
 service coordinations (like WS-BPEL~\cite{bpel03}).
@@
 So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
+\textit{(i)} the specification of the business rules of the application; and 
+\textit{(ii)} dealing with the technical characteristics of the infrastructure where the application is executed.
 
@@
 The main goals of our work are:
-\textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
+\textit{(i)} to propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
+\textit{(ii)} to improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
+\textit{(iii)} to reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
 

### 2013-10-19 SOFSEM2014/FinalVersion/main.tex
minor changes (and 12 pages!) --Martin
minor changes everywere -- Martin

@@
 
+\newcounter{numberInTrivlist}
+
+\newenvironment{numtrivlist}{\begin{list}{\arabic{numberInTrivlist}.} 
+                                         {\usecounter{numberInTrivlist}
+                                          \setlength{\leftmargin}{0pt}
+                                          \setlength{\rightmargin}{0pt}
+                                          \setlength{\itemindent}{0pt}
+                                          \setlength{\listparindent}{0pt}}}
+                            {\end{list}}
+
 \def\B{\hfill{$\Box$}}
@@
 \email{\{mam,umberto\}@dimap.ufrn.br}
-\and Federal Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
+\and Federal Technological Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
 \and French Council of Scientific Research (CNRS) -- Grenoble, France\\
@@
 \textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
-\textit{(ii)} model to model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
+\textit{(ii)} model-to-model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
 \textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 

### 2013-10-19 SOFSEM2014/FinalVersion/pisodm.tex
working… --martin
minor changes everywere -- Martin

@@
 Constraints are restrictions that must be verified during the execution of the application. 
-An example of this is the requirement of authentication for executing some of the system functions. 
 Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
+An example of policy is the requirement of authentication for executing some of the system functions. 
+The action associated to this policy may perform the authentication of the user.
 
@@
 $\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-modes) and transformations between models.
+(instances of the meta-models) and transformations between models.
 There are two kinds of transformations:
@@
 The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
+\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BaCAM05,Placido2010LTPD} or BPEL~\cite{bpel03} code.
 Although PEWS is our language of choice, other composition languages can be used as target.

### 2013-10-19 SOFSEM2014/FinalVersion/proofOfConcept.tex
minor changes everywere -- Martin

@@
 The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
-Busines process is represented as a graph while requirements are given as text boxes.
+Business process is represented as a graph while requirements are given as text boxes.
 \begin{figure}[ht!]
@@
 
-Non-functional requirements are specified by rules and conditions to be verified during the excecution of tasks.
+Non-functional requirements are specified by rules and conditions to be verified during the execution of tasks.
 In our example we have the following non-functional requirements:
@@
   This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
-  In this case, it represents the maximmum number of allowed login attempts;
+  In this case, it represents the maximum number of allowed login attempts;
 \item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
@@
 In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
-Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{busines collaborators} (external services and entities).
+Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{business collaborators} (external services and entities).
 The interaction occurs by means of function calls (denoted by dotted lines in the figure).

### 2013-10-19 SOFSEM2014/FinalVersion/introduction.tex
minor changes everywere -- Martin

@@
-Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
+Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach for the design and development of software systems. 
 The main ideas of MDD were originaly proposed by the Object Management Group
-(OMG)~\cite{mda}. 
+(OMG)\footnote{\texttt{http://www.omg.org/mda}.}. 
 MDD provides a set of

### 2013-10-19 SOFSEM2014/FinalVersion/conclusion.tex
no message

@@
-Over the last years, a number of approaches have been proposed for the
-development of web services. These approaches range from the proposal of new 
-languages for web service descriptions~\cite{bpel03,SBS04} 
-to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007}. In general, these
-approaches concentrate on specific problems, like supporting transactions or
-QoS, in order to improve the security and reliability of service-based
-applications. Some proposals address service composition: workflow
-definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
-services~\cite{BHM06}. The proposed solutions come from many communities, 
-including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
-Software Engineering~\cite{Aal03,MendesPDB09}, Programming
-Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
-
-Despite the variety of techniques proposed, there is not yet a consensus on a software
-process methodology for web services. Some methodologies address the service-based 
-development towards a standard or a new way to develop reliable applications. 
-SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
-S-Cube \cite{scube2010book} is focused on the representation of business processes and 
-service-based development; SOMA \cite{soma} is a methodology described by IBM
-for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
-BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
-methodology for building service-based infrastructure for collaborative
-enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering \cite{sommerville08} applied to SOC.
-
-\bigskip
-
-This paper presented the $\pi$SOD-M software process for specifying and designing service based applications in the presence of some non-functional constraints. 
-Our proposal harness the SOD-M method with constraints, policies and contracts in order to consider non-functional constraints.
+This paper presented the $\pi$SOD-M software method for specifying and designing service based applications in the presence of  non-functional constraints. 
+Our proposal enhances the SOD-M method with constraints, policies and contracts  to consider non-functional constraints of applications.
 We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.

### 2013-10-19 SOFSEM2014/FinalVersion/main.tex
no message

@@
 
-\section{Related Work and Conclusions}
+\section{Related work}
+\label{sec:related}
+Over the last years, a number of approaches have been proposed for the
+development of web services. These approaches range from the proposal of new 
+languages for web service descriptions~\cite{bpel03,SBS04} 
+to techniques to support phases of the development cycle of this kind of
+software~\cite{lipari2007}. In general, these
+approaches concentrate on specific problems, like supporting transactions or
+QoS, in order to improve the security and reliability of service-based
+applications. Some proposals address service composition: workflow
+definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
+services~\cite{BHM06}. The proposed solutions come from many communities, 
+including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
+Software Engineering~\cite{Aal03,MendesPDB09}, Programming
+Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
+
+Despite the variety of techniques proposed, there is not yet a consensus on a software
+process methodology for web services. Some methodologies address the service-based 
+development towards a standard or a new way to develop reliable applications. 
+SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
+S-Cube \cite{scube2010book} is focused on the representation of business processes and 
+service-based development; SOMA \cite{soma} is a methodology described by IBM
+for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
+BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
+methodology for building service-based infrastructure for collaborative
+enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
+that acts as a bridge between business modeling and design models, and traditional 
+approaches for software engineering \cite{sommerville08} applied to SOC.
+
+\section{Conclusions}
 \label{sec:conclusions}

### 2013-10-19 SOFSEM2014/FinalVersion/introduction.tex
no message

--- 2013-10-19 SOFSEM2014/FinalVersion/introduction.tex
@@
-Model Driven Development (MDD) is top-down approach for the design and development of software systems. 
-The main ideas of MDD were originaly proposed by the Object Management Group
-(OMG)~\cite{mda}. 
+Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach for  designing and developing software systems 
+%The main ideas of MDD were originaly 
+proposed by the Object Management Group
+(OMG)\footnote{\texttt{http://www.omg.org/mda}.}. 
 MDD provides a set of
-guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction.
-Abstraction levels are called \textit{viewpoints}. 
-MDD defines three viewpoints:
+guidelines for  structuring  specifications using
+ \textit{models} that specify a software system at different levels of abstraction or \textit{viewpoints}: 
+%MDD defines three viewpoints:
 
 \begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
+\item \textit{Computation Independent Models (CIM):} This viewpoint represents the software system at its highest level of abstraction. It focusses on the system
+environment, and on its business and requirement specifications. 
 At this moment of the development, the structure and system processing details are still unknown or undetermined.
 
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
+\item \textit{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
+%The specification defines those parts of the system that are independent of the development platform.
 
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
+\item \textit{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
 Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
@@
 
-\bigskip
-
+%\bigskip 
+MDD has been applied for developing service oriented applications.
 In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
@@
 models, practices and techniques to aid in software development. SOD-M does not
-provide suport to the specification of non-functional requirements, such as
+ suport  the specification of non-functional requirements, such as
 security, reliability, and efficiency. Ideally, non-functional requirements
@@
 
-\bigskip
+%\bigskip
 
@@
 
-The main goals of our work are:
-\textit{(i)} to propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\textit{(ii)} to improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(iii)} to reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
+This paper presents $\pi$SOD-M  a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements.
+The goal of the methdology are to:
+\textit{(i)} improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
+\textit{(ii)} reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
 
-\bigskip
-This paper is organized as follows:
-Sections~\ref{sec:sodm} and~\ref{sec:pisodm} presents, respectively, the SOD-M method of service software process and our proposed extension to deal with non-functional requirements.
-A proof of concept example is developed in Section~\ref{sec:poc}.
-We conclude the paper by presenting some related work and final remarks.
+%\bigskip
+Accordingly, the remainder of the paper is organized as follows:
+Sections~\ref{sec:sodm} and~\ref{sec:pisodm} present, respectively, the SOD-M method of service software process and $\pi$SOD-M our proposed extension to deal with non-functional requirements.
+A proof of concept  is developed in Section~\ref{sec:poc}. Section~\ref{sec:related} describes related works.
+Section~\ref{sec:conclusions} concludes the paper  and gives final remarks.

### 2013-10-19 SOFSEM2014/FinalVersion/pisodm.tex
no message

@@
 $\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. 
-$\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements.
-
-Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
-Constraints are restrictions that must be verified during the execution of the application. 
-Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
-An example of policy is the requirement of authentication for executing some of the system functions. 
-The action associated to this policy may perform the authentication of the user.
-
-In this context, we provide concepts at different levels of abstraction for expressing business rules of a process as policies (to implement existing NFP protocols).
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
-\caption{$\pi$SOD-M.}
-\label{fig:piSOD-M}
-\end{figure}
-
-$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
-well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as their non-functional behaviour.
-
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-models) and transformations between models.
-There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
-
-We extended SOD-M to include non-functional specifications.
-Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
+their non-functional requirements. $\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as well as transformations between these models.
+$\pi$SOD-M includes non-functional specifications through
+ four meta-models that extend PIM SOD-M meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
 
@@
 
+The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
 At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
@@
 Although PEWS is our language of choice, other composition languages can be used as target.
-This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
+%This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
+
+Note that $\pi$SOD-M extends the SOD-M meta-models by adding
+the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
+to represent non-functional requirements. $\pi$SOD-M models represent both the functional aspects of the application as well as their non-functional behaviour. 
+Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
+Constraints are restrictions that must be verified during the execution of the application. 
+Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
+An example of policy is the requirement of authentication for executing some of the system functions. 
+The action associated to this policy may perform the authentication of the user.
+
+%In this context, we provide concepts at different levels of abstraction for expressing business rules of a process as policies (to implement existing NFP protocols).
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/piSODM}
+\caption{$\pi$SOD-M.}
+\label{fig:piSOD-M}
+\end{figure}
+
+%
+%Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
+Thus, $\pi$SOD-M proposes a development process based on the definition of models
+(instances of the meta-models) and transformations between models.
+There are two kinds of transformations:
+Model-to-model transformations are used during the software process to refine the specification.
+Model-to-text transformations are the last step of the process and generate code.
 

### 2013-10-19 SOFSEM2014/FinalVersion/sodm.tex
no message

@@
 The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
-proposes the usage of the MDD approach in the context of service-based applications.
+adopts  the MDD approach to build service-based applications. SOD-M considers two points of view:
+\textit{(i)} \textit{business}, focusing on the characteristics and requirements
+of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
+features and processes to be implemented in order application requirements. In
+this way, SOD-M  simplifies the design of service-oriented applications, as
+well as their implementation using current technologies.
+
 SOD-M provides a framework with models and standards to express functionalities
@@
 PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-
 Two models are defined at the CIM level: \textit{value model} 
@@
 \textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}.
-These three levels have no support for describing non-functional requirements.
 
@@
 
-SOD-M considers two points of view:
-\textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
-this way, SOD-M aims to simplify the design of service-oriented applications, as
-well as its implementation using current technologies.
+These three SOD-M levels have no support for describing non-functional requirements.The following section introduces $\pi$-SODM the extension proposed for considering these requirements.

### 2013-10-19 SOFSEM2014/FinalVersion/main.tex
no message

@@
 
-\title{Supporting Non-Functional Requirements for Services Software Process: An MDD Approach}
+\title{Supporting Non-Functional Requirements in Services Software Development Process: An MDD Approach}
 
@@
 \section*{Acknowledgements}
+\small
 This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financied by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy.

### 2013-10-19 SOFSEM2014/FinalVersion/pisodm.tex
no message

@@
 $\pi$SOD-M includes non-functional specifications through
- four meta-models that extend PIM SOD-M meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-
-The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data. 
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
-The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services. 
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
-which put together contracts with similar non-functional requirements. 
-For instance, security and privacy restrictions may be grouped into a security policy.
-\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
-
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BaCAM05,Placido2010LTPD} or BPEL~\cite{bpel03} code.
-Although PEWS is our language of choice, other composition languages can be used as target.
-%This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
-
-Note that $\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements. $\pi$SOD-M models represent both the functional aspects of the application as well as their non-functional behaviour. 
-Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
-Constraints are restrictions that must be verified during the execution of the application. 
-Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
-An example of policy is the requirement of authentication for executing some of the system functions. 
-The action associated to this policy may perform the authentication of the user.
-
-%In this context, we provide concepts at different levels of abstraction for expressing business rules of a process as policies (to implement existing NFP protocols).
+ four meta-models that extend PIM SOD-M meta-models (see Figure~\ref{fig:piSOD-M}): \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
 
@@
 
+The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
+Non-functional requirements are defined as \textit{constraints} over processing and data.
+
+The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions.
+
+The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
+described in the \textit{$\pi$-UseCase} model into contracts that are associated
+with services.
+
+The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
+which put together contracts with similar non-functional requirements. 
+For instance, security and privacy restrictions may be grouped into a security policy.
+\textit{$\pi$-ServiceComposition} models can be refined into PSMs. Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
+Constraints are restrictions that must be verified during the execution of the application. 
+Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
+An example of policy is the requirement of authentication for executing some of the system functions. 
+The action associated to this policy may perform the authentication of the user.
+
+The \textit{$\pi$-PEWS} meta-model is a PSM (see Figure~\ref{fig:piSOD-M}).
+At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
+The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
+\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BaCAM05,Placido2010LTPD} or BPEL~\cite{bpel03} code.
+Although PEWS is our language of choice, other composition languages can be used as target.
+%This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
+
+%Note that $\pi$SOD-M extends the SOD-M meta-models by adding
+%the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
+%to represent non-functional requirements. 
+%$\pi$SOD-M models represent both the functional aspects of the application as well as their non-functional behaviour.
+
+%In this context, we provide concepts at different levels of abstraction for expressing business rules of a process as policies (to implement existing NFP protocols).
+
 %
@@
 Model-to-text transformations are the last step of the process and generate code.
-
 In the next section we develop an example, to serve as a proof-of-concept.

### 2013-10-19 SOFSEM2014/FinalVersion/proofOfConcept.tex
no message

@@
 to the community of users using this application. 
-
 Users can track crimes in given zones. 
@@
 The business process defines the logic of the application and is specified in terms of tasks. 
-Tasks can be performed by persons or computers.
+Tasks can be performed by people or computers.
 
 The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
-Business process is represented as a graph while requirements are given as text boxes.
+The business process is represented as a graph while requirements are given as text boxes.
 \begin{figure}[ht!]
@@
 \begin{numtrivlist}
-\item Twitter requires to be accessed through an authentication protocol that allows three login failures before blocking. 
+\item Twitter requires  an authentication protocol that allows three login failures before blocking. 
 \item Crime notification needs privileged access.
@@
 \end{numtrivlist}
+The idea about these requirements is to leave the application logic expressed by functional requirements as independent as possible from exceptional situations like the unavailability of a service and the conditions in which services are called for example through an authentication protocol.
+These requirements can be  weaved as activities and branches of the composition or kept independent.
+The interest of the second option is that the maintenance and the evolution of the application logic can be easier.
+For example, the services called by the application are not hard coded (Twitter and Google Maps in the example), neither the actions to deal with exceptions (replacing anohter Map service or doing nothing).
 
@@
 modelled as constraints. 
-We have three types of constraints:
+$\pi$-SODM provides three types of constraints:
 \textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case (model) can be associated to one or more constraints.
+Each use case (model) can be associated to one or more constraints \footnote{For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{PlacidoThesis2012}.}.
 
@@
 \end{numtrivlist}
-
-For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{PlacidoThesis2012}.

### 2013-10-19 SOFSEM2014/FinalVersion/main.tex
All sections reviewed, and modified.

@@
 \begin{abstract}
-This paper presents an extension to the Service Oriented Development Method (SOD-M) to support the representation of some non-functional requirements. 
-Specifically, we propose to extend SOD-M with: 
+This paper presents the $\pi$-SODM method an extension to the Service Oriented Development Method (SOD-M) to support the developement of services software considering their functional and    non-functional requirements.
+Specifically, $\pi$-SODM proposes: 
 \textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
@@
 \textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
-We apply our proposal to develop a proof-of-concept example.
+In order to illustrate the use of this methodology the paper describes how its use to develop a proof-of-concept.
 \end{abstract}

### 2013-10-19 SOFSEM2014/FinalVersion/introduction.tex
no message
New figure and more modifications maintaining 12 pages

@@
 %MDD defines three viewpoints:
-
+%
 \begin{trivlist}
@@
 At this moment of the development, the structure and system processing details are still unknown or undetermined. 
-
-\item \textit{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
+\item \textit{Platform Independent Models (PIM):} This viewpoint focusses on the system functionality, hiding the details of any particular platform. 
 %The specification defines those parts of the system that are independent of the development platform. 
-
-\item \textit{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
+\item \textit{Platform Specific Models (PSM):} This viewpoint focusses on the functionality, in the context of a particular implementation platform.
 Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
 \end{trivlist}
-
+%
 Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
@@
 (SOD-M)~\cite{decastro1}. 
-SOD-M is based on MDD and proposes
-models, practices and techniques to aid in software development. SOD-M does not
- suport  the specification of non-functional requirements, such as
-security, reliability, and efficiency. Ideally, non-functional requirements
+%SOD-M is based on MDD and proposes
+%models, practices and techniques to aid in software development. 
+%SOD-M does not
+ %suport  the specification of non-functional requirements, such as
+%security, reliability, and efficiency.
+
+Ideally, non-functional requirements such as
+security, reliability, and efficiency
 would be considered along with all the stages of the software development. The
@@
 the application context.
-
+%
 %\bigskip
-
+%
 Non-functional properties of service-oriented applications  have been
@@
 \textit{(ii)} reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
-
+%
 %\bigskip

### 2013-10-21 SOFSEM2014/FinalVersion/pisodm.tex
New figure and more modifications maintaining 12 pages
Changes in the sodm.tex and pisodm.tex, The included text are marked with comments.

@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
+\includegraphics[width=0.9\textwidth]{figs/piSOD-M.pdf}
 \caption{$\pi$SOD-M.}
@@
 Model-to-text transformations are the last step of the process and generate code.
+
+%%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
+$\pi$SOD-M environment is built on the top of Eclipse framework, which is a
+framework to build Integrated Development Environments (IDEs). We also use the
+Eclipse Modelling Framework (EMF), a meta-modelling framework that was devised to be
+extended and provides with the utilities needed to define, edit and handle
+(meta)-models. To automate the transformation models we use the ATL Language\footnote{ATL uses EMF to handle models (that is to serialize, to navigate and to modify
+them). Once the ATL transformation is executed, the ATL engine matches the
+source pattern and the source model and instantiates each pattern into the
+target model.} \cite{atl_manual}, a model transformation language framed in
+Eclipse. Another language used for model transformation is Acceleo \cite{acceleo}, a
+model to text engine. $\pi$SOD-M uses this environment~\footnote{The environmental requirements are the Eclipse Modelling (Galilleo or higher)
+and all proposed plugins, \textit{$\pi$-UseCase, $\pi$-ServiceProcess,
+$\pi$-SerciveComposition} and \textit{$\pi$-PEWS} editors; and also the
+transformation plugins. Because the environment is proposed in the
+context of Eclipse, the environmental requirements are also the same needed to
+utilize the Eclipse tool.} to generate service
+composition specification code in \textit{$\pi$-PEWS} and semi-automatize the
+ implementation of reliable service-based applications.
+%%%END{Text Included by Placido Neto for Camera-Ready}%%%%%%
+
 In the next section we develop an example, to serve as a proof-of-concept.

### 2013-10-21 SOFSEM2014/FinalVersion/sodm.tex
New figure and more modifications maintaining 12 pages
Changes in the sodm.tex and pisodm.tex, The included text are marked with comments.

@@
 
+ %%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
+The \textit{value model} is a business model that describes a business
+case as a set of values and value activities shared by business actors. The
+\textit{BPMN model} (business process model) is used to describe the business
+process related to the environment which the system will run. These two models
+represent the independent aspects of computing. The \textit{use case model} is used to represent
+the business services to be implemented by the system, while the \textit{extended use case model} is a
+behavioral model, to represent the system features as a way to implement
+the business services. The \textit{service process model} describes the set of
+activities that must be performed on the system to implement a business service.
+Finally, the \textit{service composition model} represents the full flow of
+business system. This model is an extension of the service process model,
+however, in more detail. These four models represent the platform independent
+aspects.
+%%%END{Text Included by Placido Neto for Camera-Ready}%%%%%%
+
 The SOD-M approach includes transformations between models:
@@
 follow the process activities described by the methodology. 
-
+%
 These three SOD-M levels have no support for describing non-functional requirements.The following section introduces $\pi$-SODM the extension proposed for considering these requirements.

### 2013-10-21 SOFSEM2014/FinalVersion/introduction.tex
12 pages!!

@@
 %MDD defines three viewpoints:
-%
-\begin{trivlist}
-\item \textit{Computation Independent Models (CIM):} This viewpoint represents the software system at its highest level of abstraction. It focusses on the system
+
+\noindent \textit{Computation Independent Models (CIM):} This viewpoint represents the software system at its highest level of abstraction. It focusses on the system
 environment, and on its business and requirement specifications. 
 At this moment of the development, the structure and system processing details are still unknown or undetermined.
-\item \textit{Platform Independent Models (PIM):} This viewpoint focusses on the system functionality, hiding the details of any particular platform. 
+
+\noindent \textit{Platform Independent Models (PIM):} This viewpoint focusses on the system functionality, hiding the details of any particular platform. 
 %The specification defines those parts of the system that are independent of the development platform.
-\item \textit{Platform Specific Models (PSM):} This viewpoint focusses on the functionality, in the context of a particular implementation platform.
+
+\noindent \textit{Platform Specific Models (PSM):} This viewpoint focusses on the functionality, in the context of a particular implementation platform.
 Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
-\end{trivlist}
-%
+
 Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
@@
 MDD has been applied for developing service oriented applications.
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
+In Service-Oriented Computing~\cite{scube2010book}, pre-existing services are
 combined to produce applications and provide the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
@@
 Non-functional properties of service-oriented applications  have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
+addressed in academic works and standards.%~\cite{ws-co,ws-tra,wsci}.
 Dealing with these kind of properties involves the use of specific technologies

### 2013-10-21 SOFSEM2014/FinalVersion/main.tex
12 pages!!

@@
 
-\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}
+\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Pl\'acido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}
 
@@
 \begin{abstract}
-This paper presents the $\pi$-SODM method an extension to the Service Oriented Development Method (SOD-M) to support the developement of services software considering their functional and    non-functional requirements.
+This paper presents the $\pi$-SODM method an extension to the Service Oriented Development Method (SOD-M) to support the development of services software considering their functional and    non-functional requirements.
 Specifically, $\pi$-SODM proposes: 
@@
 Over the last years, a number of approaches have been proposed for the
-development of web services. These approaches range from the proposal of new 
+development of web services. 
+These approaches range from the proposal of new 
 languages for web service descriptions~\cite{bpel03,SBS04} 
 to techniques to support phases of the development cycle of this kind of
-software~\cite{lipari2007}. In general, these
-approaches concentrate on specific problems, like supporting transactions or
+software~\cite{lipari2007}. 
+In general, these approaches concentrate on specific problems, like supporting transactions or
 QoS, in order to improve the security and reliability of service-based
-applications. Some proposals address service composition: workflow
+applications. 
+Some proposals address service composition: workflow
 definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
-services~\cite{BHM06}. The proposed solutions come from many communities, 
-including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
-Software Engineering~\cite{Aal03,MendesPDB09}, Programming
-Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
+services~\cite{BHM06}. 
+%The proposed solutions come from many communities, 
+%including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
+%Software Engineering~\cite{Aal03,MendesPDB09}, Programming
+%Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.
 
-Despite the variety of techniques proposed, there is not yet a consensus on a software
-process methodology for web services. Some methodologies address the service-based 
+Works dealing with non-functional properties in service-oriented 
+development can be organized in two main groups: those working on the \textit{modeling 
+of particular non-functional properties or QoS attributes} and those 
+proposing \textit{architectures or frameworks to manage and 
+validate QoS attributes in web service composition processes}. 
+The first group considers specific non-functional concerns (e.g. security)
+which is modelled and then associated to functional models of the application. 
+The work of Chollet et al.~\cite{CholletL09} defines a proposal to associate non-functional 
+quality properties (security properties) to functional activities in a web service composition model. Schmeling et al.~\cite{SchmelingCM11} present 
+an approach and also a toolset for specifying and implementing non-functional 
+concerns in web service compositions.
+Non-functional concerns are modelled and then related to a service composition 
+represented in a BPMN diagram. 
+Ovaska et al.~\cite{OvaskaEHPA10} present an approach to 
+support quality management at design time. 
+Quality requirements are modelled in a first phase and then
+represented in an architectural model where quality requirements are associated to some
+components of the model.
+
+%We propose to model non-functional properties together with the modelling of functional concerns.
+%We claim what this approach simplifies the web service development process. 
+%Non-functional properties are represented in our work as constraints and policies 
+%but more general quality model such as the proposed by~\cite{Goeb2011,Klass2009} 
+%could be taken in consideration in further works.
+
+The second group of works dealing with non-functional requirements for services 
+propose specific architectures or frameworks to manage and validate 
+QoS attributes in service composition processes~\cite{XiaoCZBOLH08,Babamir2010,Karunamurthy2012787}. 
+%Although our proposal focus on the design of service oriented application with non-functional 
+%properties and not on the validation of QoS attributes at runtime.
+
+Despite the variety of techniques proposed, there is not yet a consensus on a software methodology 
+for web services. 
+Some methodologies address the service-based 
 development towards a standard or a new way to develop reliable applications. 
-SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
-S-Cube \cite{scube2010book} is focused on the representation of business processes and 
-service-based development; SOMA \cite{soma} is a methodology described by IBM
-for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
-BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
+SOD-M and SOMF~\cite{somf} are MDD approaches for web services; 
+S-Cube~\cite{scube2010book} is focused on the representation of business processes and 
+service-based development; SOMA~\cite{soma} is a methodology 
+for SOA solutions; 
+%Extended SOA~\cite{PapazoglouH06} merges RUP and
+%BPM~\cite{bpm} concepts for service modelling; 
+DEVISE~\cite{DEVISE} is a
 methodology for building service-based infrastructure for collaborative
-enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
+enterprises. 
+Other proposals include, the WIED model~\cite{TongrungrojanaL04}, 
 that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering \cite{sommerville08} applied to SOC.
+approaches for software engineering applied to SOC.
 

### 2013-10-21 SOFSEM2014/FinalVersion/pisodm.tex
12 pages!!

--- 2013-10-19 SOFSEM2014/FinalVersion/pisodm.tex
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 \centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
+\includegraphics[width=1\textwidth]{figs/piSOD-M.png}
 \caption{$\pi$SOD-M.}
@@
 \end{figure}
-
+%
 The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
 Non-functional requirements are defined as \textit{constraints} over processing and data. 
-
+%
 The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
-
+%
 The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
@@
 with services. 
-
+%
 The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
@@
 The action associated to this policy may perform the authentication of the user.
-
+%
 The \textit{$\pi$-PEWS} meta-model is a PSM (see Figure~\ref{fig:piSOD-M}).
@@
 The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BaCAM05,Placido2010LTPD} or BPEL~\cite{bpel03} code.
+\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BHM06} %~\cite{BaCAM05,Placido2010LTPD} 
+or BPEL~\cite{bpel03} code.
 Although PEWS is our language of choice, other composition languages can be used as target.
@@
 Model-to-text transformations are the last step of the process and generate code.
+
+%%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
+$\pi$SOD-M environment is built on the top of Eclipse. 
+We also used the Eclipse Modelling Framework (EMF) to define, edit and handle
+(meta)-models. 
+To automate the transformation models we use ATL~\cite{atl_manual} and Acceleo~\cite{acceleo}.
+
 In the next section we develop an example, to serve as a proof-of-concept.

### 2013-10-21 SOFSEM2014/FinalVersion/proofOfConcept.tex
12 pages!!

@@
 \begin{numtrivlist}
-\item Twitter requires  an authentication protocol that allows three login failures before blocking. 
+\item Twitter requires authentication and allows three login failures before blocking. 
 \item Crime notification needs privileged access.
@@
 The constrains defined for our tracking crime example are: 
-
 \begin{trivlist}
@@
 \end{trivlist}
-\begin{figure}[ht!]
+\begin{figure}[t]
 \centering
@@
 
-\begin{figure}[ht!]
+\begin{figure}[t]
 \centering
-\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
+\includegraphics[width=1.05\textwidth]{figs/piServiceProcess}
 \caption{$\pi$-\textit{ServiceProcess} Model}
@@
 ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
-\\
+\\[-1mm]
 \subfloat[$\pi$-ServiceComposition Policies]
@@
 \end{figure}
-This model describes the service compositions and their related business
-collaborators (external services), policies, rules associated with a policy and
-the whole application process and functionalities.
-The workflow execution will consider the composition of business services and
-their specific Business Collaborators. 
-Each action will trigger the execution of one or more external services. 
-Before and after each service call, policy rules are verified.
+%This model describes the service compositions and their related business
+%collaborators (external services), policies, rules associated with a policy and
+%the whole application process and functionalities.
+%The workflow execution will consider the composition of business services and
+%their specific Business Collaborators. 
+%Each action will trigger the execution of one or more external services. 
+%Before and after each service call, policy rules are verified.
 
@@
 $\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
-
 \begin{figure}[t]
-\begin{footnotesize}
+\begin{small}
 %\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
@@
 7 alias showMap = portType/showMap in googlemaps 
-...
 //Services
@@
 10 Service visualizeCrime =  showMap . getCrimeDetail
-...
 //Path
 11 (notifyCrime.storeiCrime) || (trackCrime.visualizeCrime.getCrimeDetail)
-...
 //Contracts
-12 defContract notifyCrimeContract{ 
-13    isAppliedTo: notifyCrime
-14	  requires: userId == ?? && passw == ?? && req(notifyCrime)  < 3
-15 	      (OnFailureDo: NOT(action_publish(crime));
-16	  ensures: publishTwitter(crime) == true
-17 	      (OnFailureDo: skip);
-28 }
+12 defContract notifyCrimeContract{  isAppliedTo: notifyCrime
+13	  requires: userId == ?? && passw == ?? && req(notifyCrime)  < 3
+14 	      (OnFailureDo: NOT(action_publish(crime));
+15	  ensures: publishTwitter(crime) == true  (OnFailureDo: skip); }
 \end{verbatim}
 % \end{lstlisting}
-\end{footnotesize}
+\end{small}
 \caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
 \end{figure}
-
 The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 

### 2013-10-22 SOFSEM2014/FinalVersion/conclusion.tex
Proof-reading completed! -- Umberto

@@
-This paper presented the $\pi$SOD-M software method for specifying and designing service based applications in the presence of  non-functional constraints. 
+This paper presented the $\pi$SOD-M software method for specifying and designing service-based applications in the presence of  non-functional constraints. 
 Our proposal enhances the SOD-M method with constraints, policies and contracts  to consider non-functional constraints of applications.
@@
 So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
-We are now working on the definition of a PCM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
+We are now working on the definition of a PSM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).

### 2013-10-22 SOFSEM2014/FinalVersion/introduction.tex
Proof-reading completed! -- Umberto

@@
-Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach for  designing and developing software systems 
-%The main ideas of MDD were originaly 
-proposed by the Object Management Group
-(OMG)\footnote{\texttt{http://www.omg.org/mda}.}. 
-MDD provides a set of
-guidelines for  structuring  specifications using
- \textit{models} that specify a software system at different levels of abstraction or \textit{viewpoints}: 
-%MDD defines three viewpoints:
+Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach proposed by the Object Management Group
+(OMG)\footnote{\texttt{http://www.omg.org/mda}.} for designing and developing software systems. 
+MDD provides a set of guidelines for  structuring  specifications by using \textit{models} to specify software systems at different levels of abstraction or \textit{viewpoints}:
 
-\noindent \textit{Computation Independent Models (CIM):} This viewpoint represents the software system at its highest level of abstraction. It focusses on the system
-environment, and on its business and requirement specifications. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
+\begin{itemize}
 
-\noindent \textit{Platform Independent Models (PIM):} This viewpoint focusses on the system functionality, hiding the details of any particular platform. 
+\item
+%\noindent 
+\textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
+environment and on  business and requirement specifications. 
+At this level, the structure of the system  and its processing details are still unknown or undetermined. 
+\item
+%\noindent 
+\textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
 %The specification defines those parts of the system that are independent of the development platform. 
+\item
+%\noindent 
+\textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
+Models at this level combine the platform-independent view with specific aspects of the target platform in order to implement the system.  
+\end{itemize}
 
-\noindent \textit{Platform Specific Models (PSM):} This viewpoint focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
+Besides the notion of models at each level of abstraction, MDD requires  \textit{model transformations} between levels.
 These transformations may be automatic or semi-automatic and implement the refinement process between levels.
@@
 %\bigskip 
-MDD has been applied for developing service oriented applications.
+MDD has been applied for developing service-oriented applications.
 In Service-Oriented Computing~\cite{scube2010book}, pre-existing services are
@@
 Non-functional properties of service-oriented applications  have been
-addressed in academic works and standards.%~\cite{ws-co,ws-tra,wsci}.
+addressed in academic works and standards. %~\cite{ws-co,ws-tra,wsci} 
 Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
+at different layers of the SOC architecture, for instance during the description
 of service APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
@@
 \textit{(i)} the specification of the business rules of the application; and 
-\textit{(ii)} dealing with the technical characteristics of the infrastructure where the application is executed.
+\textit{(ii)} dealing with the technical characteristics of the infrastructure where the application is  going to be executed.
 
@@
 
-This paper presents $\pi$SOD-M  a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements.
-The goal of the methdology are to:
-\textit{(i)} improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\textit{(ii)} reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
+This paper presents $\pi$SOD-M,  a methodology for supporting the development of service-oriented applications by taking into account both functional and non-functional requirements.
+This methodology aims to:
+\textit{(i)} improve the development process by providing an abstract view of the application and ensuring its conformance to the business requirements;
+\textit{(ii)} reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high-abstraction models.
 %
@@
 Accordingly, the remainder of the paper is organized as follows:
-Sections~\ref{sec:sodm} and~\ref{sec:pisodm} present, respectively, the SOD-M method of service software process and $\pi$SOD-M our proposed extension to deal with non-functional requirements.
-A proof of concept  is developed in Section~\ref{sec:poc}. Section~\ref{sec:related} describes related works.
-Section~\ref{sec:conclusions} concludes the paper  and gives final remarks.
+Section~\ref{sec:sodm}  gives an overview of the SOD-M approach;
+Section~\ref{sec:pisodm} presents $\pi$SOD-M, the methodology that we propose to extend SOD-M;
+Section~\ref{sec:poc} shows the development of a proof of concept;
+Section~\ref{sec:related} describes related works; and
+Section~\ref{sec:conclusions} concludes the paper. % and gives final remarks.

### 2013-10-22 SOFSEM2014/FinalVersion/main.tex
Proof-reading completed! -- Umberto

@@
 
-\title{Supporting Non-Functional Requirements in Services Software Development Process: An MDD Approach}
+\title{Supporting Non-Functional Requirements in Services Software Development Process: an MDD Approach}
 
@@
 \begin{abstract}
-This paper presents the $\pi$-SODM method an extension to the Service Oriented Development Method (SOD-M) to support the development of services software considering their functional and    non-functional requirements.
-Specifically, $\pi$-SODM proposes: 
-\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
+This paper presents the $\pi$SOD-M method, an extension to the Service-Oriented Development Method (SOD-M) to support the development of services software by considering their functional and    non-functional requirements.
+Specifically, $\pi$SOD-M proposes: 
+\textit{(i)} meta-models for representing non-functional requirements at different abstraction levels; 
 \textit{(ii)} model-to-model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
 \textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
-In order to illustrate the use of this methodology the paper describes how its use to develop a proof-of-concept.
+In order to illustrate our proposal, the paper also describes how to apply the methodology to develop a proof of concept.
 \end{abstract}
@@
 validate QoS attributes in web service composition processes}. 
-The first group considers specific non-functional concerns (e.g. security)
+The first group considers specific non-functional concerns (e.g., security)
 which is modelled and then associated to functional models of the application. 
@@
 enterprises. 
-Other proposals include, the WIED model~\cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models, and traditional 
-approaches for software engineering applied to SOC.
+Other proposals include the WIED model~\cite{TongrungrojanaL04}, 
+that acts as a bridge between business modeling and design models on one hand, and traditional 
+approaches for software engineering applied to SOC on the other hand.
 

### 2013-10-22 SOFSEM2014/FinalVersion/pisodm.tex
Proof-reading completed! -- Umberto

@@
-%In this section we present $\pi$SOD-M, an MDD based methodology. 
-$\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. $\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as well as transformations between these models.
+$\pi$SOD-M provides an environment for building service compositions by considering their non-functional requirements.
+$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as well as transformations between these models.
 $\pi$SOD-M includes non-functional specifications through
- four meta-models that extend PIM SOD-M meta-models (see Figure~\ref{fig:piSOD-M}): \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
+ four meta-models that extend the SOD-M meta-models at the PIM and PSM levels (see Figure~\ref{fig:piSOD-M}): \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
 
@@
 The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data. 
+Non-functional requirements are defined as \textit{constraints} over processing steps and data. 
 %
@@
 The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services. 
+described in the \textit{$\pi$-UseCase} model into contracts that are associated with services. 
 %
@@
 Constraints are restrictions that must be verified during the execution of the application. 
-Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
+Failure to verify some constraint will trigger an exceptional behavior to execute the corresponding recovery action.
 An example of policy is the requirement of authentication for executing some of the system functions. 
@@
 %
-The \textit{$\pi$-PEWS} meta-model is a PSM (see Figure~\ref{fig:piSOD-M}).
+%The \textit{$\pi$-PEWS} meta-model is a PSM (see Figure~\ref{fig:piSOD-M}).
 At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
+The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work~(see Figure~\ref{fig:piSOD-M}).
 \textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BHM06} %~\cite{BaCAM05,Placido2010LTPD} 
-or BPEL~\cite{bpel03} code.
-Although PEWS is our language of choice, other composition languages can be used as target.
+code.
 %This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
@@
 There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
+model-to-model transformations, used to refine the specification during the software process; and
+model-to-text transformations, used   to generate code in the last step of the software process.
+Notice that other composition languages, such as BPEL~\cite{bpel03}, can be used as target of the proposed software process.
+Another target language can be supported by defining the corresponding PIM-to-PSM and PSM-to-text transformations.
 
@@
 $\pi$SOD-M environment is built on the top of Eclipse. 
-We also used the Eclipse Modelling Framework (EMF) to define, edit and handle
+We also used the Eclipse Modeling Framework (EMF) to define, edit and handle
 (meta)-models. 
@@
 
-In the next section we develop an example, to serve as a proof-of-concept.
+In the next section we develop an example, in order to illustrate our proposal.
 The example will show the actual notation used for models.

### 2013-10-22 SOFSEM2014/FinalVersion/proofOfConcept.tex
Proof-reading completed! -- Umberto

@@
 
-In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize the information.
+In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize  information.
 The business process defines the logic of the application and is specified in terms of tasks. 
@@
 Notified crimes are stored in a database. 
-Tracked crimes are visualized in a map.
+Tracked crimes are visualized on a map.
 The used can ask for detailed information. 
-The application is built upon four services: \textsf{twitter} and an \textsf{ad-hoc police service} for notifying crimes, \textsf{Amazon} used as  persistence service and  \textsf{Google Maps} for locating and displaying crimes on a map.
+The application is built upon four services: \textsf{Twitter} and an \textsf{ad-hoc police service}, for notifying crimes; \textsf{Amazon}, used as  persistence service; and  \textsf{Google Maps}, for locating and displaying crimes on a map.
 
@@
 In our example we have the following non-functional requirements:
-\begin{numtrivlist}
+\begin{itemize}
 \item Twitter requires authentication and allows three login failures before blocking. 
 \item Crime notification needs privileged access.
-\item Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
+\item Civilian users can only track crimes for which they have clearance: civilian population cannot track all the crimes notified by the police. 
 \item If \textsf{Google Maps} is unavailable, the results are delivered as text. 
@@
 \item The application maintains a detailed log. 
-\end{numtrivlist}
-The idea about these requirements is to leave the application logic expressed by functional requirements as independent as possible from exceptional situations like the unavailability of a service and the conditions in which services are called for example through an authentication protocol.
-These requirements can be  weaved as activities and branches of the composition or kept independent.
-The interest of the second option is that the maintenance and the evolution of the application logic can be easier.
-For example, the services called by the application are not hard coded (Twitter and Google Maps in the example), neither the actions to deal with exceptions (replacing anohter Map service or doing nothing).
+\end{itemize}
+The idea about these requirements is to leave the application logic expressed by functional requirements as independent as possible from exceptional situations (like the unavailability of a service) and from conditions under which services are called (for instance, through an authentication protocol).
+These requirements can be  weaved as activities and branches of the composition or implemented apart.
+The second option is better because makes the maintenance and evolution of applications easier.
+For instance, the services called by the application are not hard coded (\textsf{Twitter} and \textsf{Google Maps} in the example), neither the actions to deal with exceptions (replacing another map service or doing nothing).
 
-Considering the example of tracking crimes, all the system restrictions are
-modelled as constraints. 
-$\pi$-SODM provides three types of constraints:
+All the system restrictions are modelled as constraints in this example. 
+$\pi$SOD-M provides three types of constraints:
 \textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
-Each use case (model) can be associated to one or more constraints \footnote{For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{PlacidoThesis2012}.}.
-
+Each use case (model) can be associated to one or more constraints\footnote{For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{PlacidoThesis2012}.}.
 \paragraph*{\textit{$\pi$-UseCase} model:} 
-In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
-system functions (tasks) and constraints. 
-We will not detail the functional part of the specification, due to lack of space.
-The constrains defined for our tracking crime example are: 
-\begin{trivlist}
-  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
+our example has five use cases which represent the
+ functions (tasks) and constraints of the system (Figure~\ref{fig:piUC}). 
+We do not detail the functional part of the specification, due to lack of space.
+The constraints defined are: 
+\begin{itemize}
+  \item The \textsf{Notify crime} task requires that the user is logged in. 
   This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
   In this case, it represents the maximum number of allowed login attempts;
-\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
-\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
+\item The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
+\item In order to perform the \textsf{Track crimes} task,
+the contact list of the requesting user must include the user that notified the crime.
 This is an example of \textit{business constraint}.
 Additionally the requesting user must be logged in.
-\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text.
+\item  For the \textsf{View crimes' map} task, the specification defines that if the service \textsf{Google Maps}  is not available, the result is presented as text.
 This is an example of \textit{exceptional behaviour constraint}.
-The availability of the Google Maps service is verified by a \textit{business constraint}.
- \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
-\end{trivlist}
+The availability of the \textsf{Google Maps} service is verified by a \textit{business constraint}.
+ \item The \textsf{Show crimes' details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; a \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; another \textit{value constraint} defines that the log is to be maintained.
+\end{itemize}
 \begin{figure}[t]
@@
 \end{figure}
+
+A model-to-model transformation is defined to refine the \textit{$\pi$-UseCase} model into the
+corresponding \textit{$\pi$-ServiceProcess} model, a more detailed representation. 
+This (semi-automatic) transformation  is supported by a tool (described in~\cite{PlacidoThesis2012}).
+
 \paragraph*{\textit{$\pi$-ServiceProcess} model:}
-The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
-The new model describes the application's activities and defines contracts for each activity or for parts of the application.
-
-A model-to model transformation is defined in order refine the \textit{$\pi$-UseCase} model of the application into the more detailed model. 
-This (semi-automatic) transformation process is supported by a tool (described in~\cite{PlacidoThesis2012}).
-
-The \textit{$\pi$-ServiceProcess} model defined for our tracking crime application is presented in Figure~\ref{fig:piSP}, where: 
-\begin{numtrivlist}
+in this model task nodes of the $\pi$-UseCase model are better detailed, by refining the control and data flows, and its constraints are transformed into \textit{contracts} (pre- and post-conditions).
+The $\pi$-ServiceProcess model describes the activities of the application and defines contracts for each activity or parts of the application (Figure~\ref{fig:piSP}). The main transformations are: 
+\begin{itemize}
 \item Tasks of the previous model are transformed into \textit{actions};
@@
 \item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
-\end{numtrivlist}
+\end{itemize}
 
@@
 \paragraph*{\textit{$\pi$-ServiceComposition} model:} 
-This model refines the previous model by using the activities to produce the workflow of the application.
+this model refines the previous model by using the activities to produce the workflow of the application.
 The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
@@
 \paragraph*{$\pi$\textit{-PEWS} Model:}
-These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
-This code is a service composition program that can be compiled into executable code.
+this model is produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
+This code is a service composition program that can be compiled into an executable composition.
 $\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
@@
 The figure shows a simplified program code, produced in accordance to the following guidelines:
-\begin{numtrivlist}
-\item Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
+\begin{itemize}
+\item
+Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
 We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
-\item Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
-\item The workflow in Figure~\ref{fig:piSC} is translated into the text in line 11.
-\item \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed case a condition is not verified.
-Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
-\end{numtrivlist}
+\item
+Specific operations exported by each business collaborator are identified as an \textit{operation} of the program (to each operation is given an \texttt{alias}).
+\item
+The workflow in Figure~\ref{fig:piSC} is translated into the text at line 11.
+\item
+\textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed whenever a condition is not verified.
+Contracts are generated from Policies (Figure~\ref{fig:crimePolicies}).
+\end{itemize}

### 2013-10-22 SOFSEM2014/FinalVersion/sodm.tex
Proof-reading completed! -- Umberto

--- 2013-10-19 SOFSEM2014/FinalVersion/sodm.tex
@@
 \textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
+of the organization; and 
+\textit{(ii)} \textit{system requirements}, focusing on
+features and processes to be implemented in order to build service-based applications in accordance to the business requirements. In
 this way, SOD-M  simplifies the design of service-oriented applications, as
@@
 SOD-M provides a framework with models and standards to express functionalities
-of applications at a high-level of abstraction. SOD-M meta-models are organized 
+of applications at a high-level of abstraction. The SOD-M meta-models are organized 
 into three levels: CIM (\textit{Computational Independent Models}), 
@@
 and \textit{BPMN model}. 
-The PIM level models the entire structure of the application flow,
-while, the PSM level provides transformations towards more specific platforms.
+The PIM level models the entire structure of the application, as
+the PSM level provides transformations towards more specific platforms.
 The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
-\textit{service composition}. The PSM level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}.
+\textit{service composition}. The PSM-level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}.
+
+ %%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
+At the CIM level, the \textit{value model} describes a business
+scenario as a set of values and activities shared by business actors. 
+The \textit{BPMN model} describes a business process and the corresponding environment. 
+%These two models represent the computing-independent aspects of the application logic. 
+At the PIM level, the \textit{use case model} represents
+a business service, as the \textit{extended use case model} contains 
+behavioral descriptions of features to be implemented. 
+The \textit{service process model} describes a set of
+activities to be performed in order to implement a business service.
+Finally, the \textit{service composition model} represents the complete flow of a business system. 
+This model is an extension of the service process model. 
+%These four models represent the platform independent aspects.
+%%%END{Text Included by Placido Neto for Camera-Ready}%%%%%%
 
@@
 follow the process activities described by the methodology. 
-
-These three SOD-M levels have no support for describing non-functional requirements.The following section introduces $\pi$-SODM the extension proposed for considering these requirements.
+%
+These three SOD-M levels have no support for describing non-functional requirements. The following section introduces $\pi$SOD-M, the extension that we propose for suporting these requirements.

### 2013-10-22 SOFSEM2014/FinalVersion/proofOfConcept.tex
minor changes everywere -- Martin

@@
 \includegraphics[width=0.8\textwidth]{figs/piUseCase}
-\caption{$\pi$-UseCase Model}
+\caption{$\pi$-UseCase Model.}
 \label{fig:piUC}
@@
 \includegraphics[width=1.05\textwidth]{figs/piServiceProcess}
-\caption{$\pi$-\textit{ServiceProcess} Model}
+\caption{$\pi$-\textit{ServiceProcess} Model.}
 \label{fig:piSP}
@@
 \centering
-\subfloat[$\pi$-ServiceComposition Model]
+\subfloat[$\pi$-ServiceComposition Model.]
 {\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
@@
 \\[-1mm]
-\subfloat[$\pi$-ServiceComposition Policies]
+\subfloat[$\pi$-ServiceComposition Policies.]
 {\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}

### 2013-10-22 SOFSEM2014/FinalVersion/introduction.tex
12 pages, Valeria's reccomndations.

@@
 (OMG)\footnote{\texttt{http://www.omg.org/mda}.} for designing and developing software systems. 
-MDD provides a set of guidelines for  structuring  specifications by using \textit{models} to specify software systems at different levels of abstraction or \textit{viewpoints}:
-
-\begin{itemize}
-
-\item
-%\noindent 
-\textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
+MDD provides a set of guidelines for  structuring  specifications by using \textit{models} to specify software systems at different levels of abstraction or \textit{viewpoints}:\\[1mm]
+\noindent 
+-\textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
 environment and on  business and requirement specifications. 
 At this level, the structure of the system  and its processing details are still unknown or undetermined.
-\item
-%\noindent 
-\textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
+
+\noindent 
+-\textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
 %The specification defines those parts of the system that are independent of the development platform.
-\item
-%\noindent 
-\textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
+
+\noindent 
+-\textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
 Models at this level combine the platform-independent view with specific aspects of the target platform in order to implement the system.
-\end{itemize}
 

### 2013-10-22 SOFSEM2014/FinalVersion/main.tex
minor changes everywere -- Martin
12 pages, Valeria's reccomndations.

@@
 components of the model. 
-
-%We propose to model non-functional properties together with the modelling of functional concerns.
-%We claim what this approach simplifies the web service development process. 
-%Non-functional properties are represented in our work as constraints and policies 
-%but more general quality model such as the proposed by~\cite{Goeb2011,Klass2009} 
-%could be taken in consideration in further works.
+We propose to model functional and non-functional properties at the same time during the software process.
+We claim what this approach simplifies the web service development task. 
+Non-functional properties are represented in our work as constraints and policies 
+but more general quality model such as the proposed by~\cite{Goeb2011,Klass2009} 
+could be taken in consideration in further works.
 
@@
 QoS attributes in service composition processes~\cite{XiaoCZBOLH08,Babamir2010,Karunamurthy2012787}. 
-%Although our proposal focus on the design of service oriented application with non-functional 
-%properties and not on the validation of QoS attributes at runtime.
+Our proposal is similar to these, but focusses on more general non-functional properties.
 
@@
 This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financied by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy.
+\normalsize
 

### 2013-10-22 SOFSEM2014/FinalVersion/FilesSentToThePublisher/main.tex
Final version to SOFSEM!

\documentclass{llncs}
\usepackage{tikz}
\usepackage{amsfonts}
\usepackage{graphicx,url}
\usepackage[T1]{fontenc}
%\usepackage{psfig}
\usepackage{graphics}
\usepackage{amssymb,amsmath,latexsym}
\usepackage{listings} 
\usepackage{subfig}

%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{example}[theorem]{Example}
%\newtheorem{algorithm}[theorem]{Algorithm}
%\newtheorem{definition}[theorem]{Definition}

\usepackage[thmmarks]{ntheorem}

\usepackage{listings}
\theoremstyle{plain}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{.}
\theoremindent0cm
\theoremnumbering{arabic}
\theoremsymbol{\ensuremath{\Box}} 
\newtheorem{defi}{Definition}%[section]

\theoremheaderfont{\itshape}
\newtheorem{examp}{Example}%[section]

\theoremstyle{plain}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{}
\theoremindent0cm
\theoremnumbering{arabic}
%\theoremsymbol{\ensuremath{\Box}} 
\newtheorem{algo}{Algorithm}%[section]

\theoremstyle{plain} 
\theoremsymbol{\ensuremath{\Box}} 
\theoremseparator{.} 
\newtheorem{prop}{Property}

\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
        true,false,rise,exception,extend,resume,to,return,function}}

\newcounter{numberInTrivlist}

\newenvironment{numtrivlist}{\begin{list}{\arabic{numberInTrivlist}.} 
                                         {\usecounter{numberInTrivlist}
                                          \setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{0pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\def\B{\hfill{$\Box$}}
\def\A{\hbox{${\cal A}$}}
\def\P{\hbox{${\cal P}$}}
\def\T{\hbox{${\cal T}$}}
\def\O{\hbox{${\cal O}$}}
\def\W{\hbox{${\cal W}$}}
\def\useOpt{\hbox{${has\_opt}$}}

\def\bigS{\hbox{${\cal S}$}}
\def\bigSd{\hbox{${\cal S}_D$}}
\def\bigSsm{\hbox{${\cal S}_D^{score}$}}

\newcommand{\ie}{\ifmmode \mathit{i.e.}\else \textit{i.e.}\fi}
\newcommand{\wrt}{\ifmmode \mathit{w.r.t.}\else \textit{w.r.t.}\fi}
\newcommand{\eg}{\ifmmode \mathit{e.g.}\else \textit{e.g.}\fi}
\newcommand{\tuple}[1]{\ifmmode \langle #1 \rangle\else $\langle #1 \rangle$\fi}

\usepackage{authblk}

\begin{document}
%
% --- Author Metadata here ---

% --- End of Author Metadata ---

\title{Supporting Non-Functional Requirements in Services Software Development Process: an MDD Approach}

\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Pl\'acido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}

\institute{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\  \email{Valeria.deCastro@urjc.es} 
\and Federal University of Rio Grande do Norte (UFRN) -- Natal-RN, Brazil\\
\email{\{mam,umberto\}@dimap.ufrn.br}
\and Federal Technological Institute of Rio Grande do Norte (IFRN) -- Natal-RN, Brazil\\ \email{placido.neto@ifrn.edu.br}
\and French Council of Scientific Research (CNRS) -- Grenoble, France\\
\email{Genoveva.Vargas-Solar@imag.fr}
}

\maketitle
\begin{abstract}
This paper presents the $\pi$SOD-M method, an extension to the Service-Oriented Development Method (SOD-M) to support the development of services software by considering their functional and    non-functional requirements. 
Specifically, $\pi$SOD-M proposes: 
\textit{(i)} meta-models for representing non-functional requirements at different abstraction levels; 
\textit{(ii)} model-to-model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
In order to illustrate our proposal, the paper also describes how to apply the methodology to develop a proof of concept.
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]

%\terms{Delphi theory}

\keywords{MDD, Service Oriented Applications, Non-functional Properties}

\section{Introduction}
\label{sec:intro}
%\input{introduction}
Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach proposed by the Object Management Group
(OMG)\footnote{\texttt{http://www.omg.org/mda}.} for designing and developing software systems. 
MDD provides a set of guidelines for  structuring  specifications by using \textit{models} to specify software systems at different levels of abstraction or \textit{viewpoints}:\\[1mm]
\noindent 
- \textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
environment and on  business and requirement specifications. 
At this level, the structure of the system  and its processing details are still unknown or undetermined.

\noindent 
- \textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
%The specification defines those parts of the system that are independent of the development platform.

\noindent 
- \textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with specific aspects of the target platform in order to implement the system.

Besides the notion of models at each level of abstraction, MDD requires  \textit{model transformations} between levels.
These transformations may be automatic or semi-automatic and implement the refinement process between levels.

%\bigskip 
MDD has been applied for developing service-oriented applications.
In Service-Oriented Computing~\cite{scube2010book}, pre-existing services are
combined to produce applications and provide the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
Some methodologies and techniques have been proposed to help the software
developer in the specification of functional requirements of the business logic,
such as the Service-Oriented Development Method
(SOD-M)~\cite{decastro1}. 
%SOD-M is based on MDD and proposes
%models, practices and techniques to aid in software development. 
%SOD-M does not
 %suport  the specification of non-functional requirements, such as
%security, reliability, and efficiency.

Ideally, non-functional requirements such as
security, reliability, and efficiency
would be considered along with all the stages of the software development. The
adoption of non-functional specifications from the early states of development
can help the developer to produce applications that are capable of dealing with
the application context.
%
%\bigskip
%
Non-functional properties of service-oriented applications  have been
addressed in academic works and standards. %~\cite{ws-co,ws-tra,wsci} 
Dealing with these kind of properties involves the use of specific technologies
at different layers of the SOC architecture, for instance during the description
of service APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
service coordinations (like WS-BPEL~\cite{bpel03}).

%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
%\begin{trivlist}
%\item[-] Those coming from the Business Process Domain and from the Web
%service standards that propose ad-hoc protocols. 
%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
%\item[-] Those adopting a classic middleware approach were non-functional
%properties are provided as middleware services, like the ones presented
%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
%\item[-] Those providing languages and enabling the specification of protocols
%used for expressing service
%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
%\item[-] Those adopting separation of concerns,
%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
%\end{trivlist}

Protocols and models implementing non-functional properties assume the existence of a global control of the artefacts implementing the application.
They also assume that each service exports its interface.
So, the challenge of supporting non-functional properties is related to
\textit{(i)} the specification of the business rules of the application; and 
\textit{(ii)} dealing with the technical characteristics of the infrastructure where the application is  going to be executed.

%In this context, there is a need of 
%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
%
%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.

This paper presents $\pi$SOD-M,  a methodology for supporting the development of service-oriented applications by taking into account both functional and non-functional requirements. This methodology aims to:
\textit{(i)} improve the development process by providing an abstract view of the application and ensuring its conformance to the business requirements;
\textit{(ii)} reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high-abstraction models.
%
%\bigskip
Accordingly, the remainder of the paper is organized as follows:
Section~\ref{sec:sodm}  gives an overview of the SOD-M approach;
Section~\ref{sec:pisodm} presents $\pi$SOD-M, the methodology that we propose to extend SOD-M;
Section~\ref{sec:poc} shows the development of a proof of concept;
Section~\ref{sec:related} describes related works; and
Section~\ref{sec:conclusions} concludes the paper. % and gives final remarks.

\section{SOD-M}\label{sec:sodm}
%\input{sodm}

The Service-Oriented Development Method (SOD-M)~\cite{decastro1}
adopts  the MDD approach to build service-based applications. SOD-M considers two points of view:
\textit{(i)} \textit{business}, focusing on the characteristics and requirements
of the organization; and 
\textit{(ii)} \textit{system requirements}, focusing on
features and processes to be implemented in order to build service-based applications in accordance to the business requirements. In
this way, SOD-M  simplifies the design of service-oriented applications, as
well as their implementation using current technologies.

SOD-M provides a framework with models and standards to express functionalities
of applications at a high-level of abstraction. The SOD-M meta-models are organized 
into three levels: CIM (\textit{Computational Independent Models}), 
PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
Two models are defined at the CIM level: \textit{value model} 
and \textit{BPMN model}. 
The PIM level models the entire structure of the application, as
the PSM level provides transformations towards more specific platforms.
The PIM-level models are: \textit{use case}, \textit{extended use case}, \textit{service process} and
\textit{service composition}. The PSM-level models are: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}.

 %%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
At the CIM level, the \textit{value model} describes a business
scenario as a set of values and activities shared by business actors. 
The \textit{BPMN model} describes a business process and the corresponding environment. 
%These two models represent the computing-independent aspects of the application logic. 
At the PIM level, the \textit{use case model} represents
a business service, as the \textit{extended use case model} contains 
behavioral descriptions of features to be implemented. 
The \textit{service process model} describes a set of
activities to be performed in order to implement a business service.
Finally, the \textit{service composition model} represents the complete flow of a business system. 
This model is an extension of the service process model. 
%These four models represent the platform independent aspects.
%%%END{Text Included by Placido Neto for Camera-Ready}%%%%%%

The SOD-M approach includes transformations between models:
\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
an abstract model at the CIM level, it is possible to apply transformations for
generating a model of the PSM level. In this context, it is necessary to
follow the process activities described by the methodology. 
%
These three SOD-M levels have no support for describing non-functional requirements. The following section introduces $\pi$SOD-M, the extension that we propose for supporting these requirements.

\section{$\pi$SOD-M}\label{sec:pisodm}
%\input{pisodm}

$\pi$SOD-M provides an environment for building service compositions by considering their non-functional requirements. $\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as well as transformations between these models.
$\pi$SOD-M includes non-functional specifications through
 four meta-models that extend the SOD-M meta-models at the PIM and PSM levels (see Figure~\ref{fig:piSOD-M}): \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth]{piSOD-M.png}
\caption{$\pi$SOD-M.}
\label{fig:piSOD-M}
\end{figure}
%
The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
Non-functional requirements are defined as \textit{constraints} over processing steps and data. 
%
The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
%
The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
described in the \textit{$\pi$-UseCase} model into contracts that are associated with services. 
%
The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
which put together contracts with similar non-functional requirements. 
For instance, security and privacy restrictions may be grouped into a security policy.
\textit{$\pi$-ServiceComposition} models can be refined into PSMs. Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
Constraints are restrictions that must be verified during the execution of the application. 
Failure to verify some constraint will trigger an exceptional behavior to execute the corresponding recovery action.
An example of policy is the requirement of authentication for executing some of the system functions. 
The action associated to this policy may perform the authentication of the user.
%
%The \textit{$\pi$-PEWS} meta-model is a PSM (see Figure~\ref{fig:piSOD-M}).
At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work~(see Figure~\ref{fig:piSOD-M}).
\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BHM06} %~\cite{BaCAM05,Placido2010LTPD} 
code.
%This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

%Note that $\pi$SOD-M extends the SOD-M meta-models by adding
%the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
%to represent non-functional requirements. 
%$\pi$SOD-M models represent both the functional aspects of the application as well as their non-functional behaviour.

%In this context, we provide concepts at different levels of abstraction for expressing business rules of a process as policies (to implement existing NFP protocols).

%
%Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
Thus, $\pi$SOD-M proposes a development process based on the definition of models
(instances of the meta-models) and transformations between models.
There are two kinds of transformations:
model-to-model transformations, used to refine the specification during the software process; and
model-to-text transformations, used   to generate code in the last step of the software process.
Notice that other composition languages, such as BPEL~\cite{bpel03}, can be used as target of the proposed software process. Another target language can be supported by defining the corresponding PIM-to-PSM and PSM-to-text transformations.

%%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
$\pi$SOD-M environment is built on the top of Eclipse. 
We also used the Eclipse Modelling Framework (EMF) to define, edit and handle
(meta)-models. 
To automate the transformation models we use ATL~\cite{atl_manual} and Acceleo.

In the next section we develop an example, in order to illustrate our proposal.
The example will show the actual notation used for models.

\section{Proof of Concept: \textit{Tracking Crimes}}\label{sec:poc}
%\input{proofOfConcept}

Consider a tracking crime application where civilians and police share information about criminality in given zones of a city. 
Civilian users signal crimes using Twitter.
Police officers can notify crimes, as well as update information about solving cases.
Some of these information are confidential while other can be shared
to the community of users using this application. 
Users can track crimes in given zones. 
Crime information stored by the system may be visualized on a map. 
Some users have different access rights than others.
For example, police officers have more access rights than civilians.

In order to provide these functionalities, the application uses pre-existing services to provide, store and visualize  information.
The business process defines the logic of the application and is specified in terms of tasks. 
Tasks can be performed by people or computers.

The business process and requirements specifications presented in Figure~\ref{fig:constraints} are instances of the Computation-Independent models of Figure~\ref{fig:piSOD-M}.
The business process is represented as a graph while requirements are given as text boxes.
\begin{figure}[ht!]
\centering
%\subfloat[\textit{Tracking Crimes} - Illustration of the Problem.]
%{\label{fig:problem}\includegraphics[width=0.95\textwidth]{figs/problem}}
%~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
%\\
%\subfloat[Contraints.]
{\label{fig:trans06}\includegraphics[width=0.95\textwidth]{constraints.png}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\caption{Business process for the tracking crime example.}
\label{fig:constraints}
\end{figure}

In our example, crime processing can start with one of two tasks: \textit{(i) notify a crime}, or \textit{(ii) track a crime}.
Notified crimes are stored in a database. 
Tracked crimes are visualized on a map.
The used can ask for detailed information. 
The application is built upon four services: \textsf{Twitter} and an \textsf{ad-hoc police service}, for notifying crimes; \textsf{Amazon}, used as  persistence service; and  \textsf{Google Maps}, for locating and displaying crimes on a map.

Non-functional requirements are specified by rules and conditions to be verified during the execution of tasks.
In our example we have the following non-functional requirements:
\begin{itemize}
\item Twitter requires authentication and allows three login failures before blocking. 
\item Crime notification needs privileged access.
\item Civilian users can only track crimes for which they have clearance: civilian population cannot track all the crimes notified by the police. 
\item If \textsf{Google Maps} is unavailable, the results are delivered as text. 
\item Querying about crimes without having proper clearance yields an empty map.
\item Access rights to detailed information depends on user clearance and zone assignment for police officers. 
\item The application maintains a detailed log. 
\end{itemize}
The idea about these requirements is to leave the application logic expressed by functional requirements as independent as possible from exceptional situations (like the unavailability of a service) and from conditions under which services are called (for instance, through an authentication protocol). These requirements can be  weaved as activities and branches of the composition or implemented apart. The second option is better because it makes the maintenance and evolution of applications easier. For instance, the services called by the application are not hard coded (\textsf{Twitter} and \textsf{Google Maps} in the example), neither the actions to deal with exceptions (replacing another map service or doing nothing).

All the system restrictions are modelled as constraints in this example. 
$\pi$SOD-M provides three types of constraints:
\textit{value}, \textit{business} and \textit{exceptions behaviour} constraints.
Each use case (model) can be associated to one or more constraints\footnote{For a more comprehensive account of $\pi$SOD-M the reader can refer to~\cite{PlacidoThesis2012}.}.
\paragraph*{\textit{$\pi$-UseCase} model:} 
our example has five use cases which represent the
 functions (tasks) and constraints of the system (Figure~\ref{fig:piUC}). 
We do not detail the functional part of the specification, due to lack of space.
The constraints defined are: 
\begin{itemize}
  \item The \textsf{Notify crime} task requires that the user is logged in. 
  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
  In this case, it represents the maximum number of allowed login attempts;
\item The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
\item In order to perform the \textsf{Track crimes} task,
the contact list of the requesting user must include the user that notified the crime.
This is an example of \textit{business constraint}.
Additionally the requesting user must be logged in.
\item  For the \textsf{View crimes' map} task, the specification defines that if the service \textsf{Google Maps}  is not available, the result is presented as text. This is an example of \textit{exceptional behaviour constraint}. 
The availability of the \textsf{Google Maps} service is verified by a \textit{business constraint}.
 \item The \textsf{Show crimes' details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; a \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; another \textit{value constraint} defines that the log is to be maintained.
\end{itemize}
\begin{figure}[t]
\centering
\includegraphics[width=0.8\textwidth]{piUseCase.png}
\caption{$\pi$-UseCase Model.}
\label{fig:piUC}
\end{figure}

A model-to-model transformation is defined to refine the \textit{$\pi$-UseCase} model into the
corresponding \textit{$\pi$-ServiceProcess} model, a more detailed representation. 
This (semi-automatic) transformation  is supported by a tool (described in~\cite{PlacidoThesis2012}).

\paragraph*{\textit{$\pi$-ServiceProcess} model:}
in this model task nodes of the $\pi$-UseCase model are better detailed, by refining the control and data flows, and its constraints are transformed into \textit{contracts} (pre- and post-conditions).
The $\pi$-ServiceProcess model describes the activities of the application and defines contracts for each activity or parts of the application (Figure~\ref{fig:piSP}). The main transformations are: 
\begin{itemize}
\item Tasks of the previous model are transformed into \textit{actions};
\item Actions are grouped into \textit{activities} (in accordance to the business logic).
\item Constraints of the \textit{$\pi$-UseCase} model are transformed into assertions.
\end{itemize}

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth]{piServiceProcess.png}
\caption{$\pi$-\textit{ServiceProcess} Model.}
\label{fig:piSP}
\end{figure}

\paragraph*{\textit{$\pi$-ServiceComposition} model:} 
this model refines the previous model by using the activities to produce the workflow of the application.
The model serves to identify those entities that collaborate with the service process by providing services to execute actions. 
This model identifies the services and functions that correspond to each action in the business process.

\begin{figure}[t]
\centering
\subfloat[$\pi$-ServiceComposition Model.]
{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{piServiceComposition.png}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\\[-1mm]
\subfloat[$\pi$-ServiceComposition Policies.]
{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{policies.png}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\caption{Service Composition and Policies.}
\label{fig:policies}
\end{figure}
%This model describes the service compositions and their related business
%collaborators (external services), policies, rules associated with a policy and
%the whole application process and functionalities.
%The workflow execution will consider the composition of business services and
%their specific Business Collaborators. 
%Each action will trigger the execution of one or more external services. 
%Before and after each service call, policy rules are verified.

In the case of our crime tracking example, the model produced from the $\pi$-\textit{ServiceProcess} model of Figure~\ref{fig:piSP} is given in Figures~\ref{fig:piSC} and~\ref{fig:crimePolicies}.
Figure~\ref{fig:piSC} shows how the crime application interacts with its \textit{business collaborators} (external services and entities).
The interaction occurs by means of function calls (denoted by dotted lines in the figure).
Figure~\ref{fig:crimePolicies} shows the definition of three \textit{policies}, which define rules for service execution.
In our case we have policies for \textit{Security}, \textit{Performance} and \textit{Persistence}.

\paragraph*{$\pi$\textit{-PEWS} Model:}
this model is produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
This code is a service composition program that can be compiled into an executable composition.
$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
\begin{figure}[t]
\begin{small}
%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
%\begin{lstlisting}[label=list:rulePre]
\begin{verbatim}
//Namespaces specify service URI
1 namespace twitter = www.twitter.com/service.wsdl
2 namespace googlemaps = maps.googleapis.com/maps/api/service
3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
4 namespace police = www.police.fr/service.wsdl
//Operations 
5 alias publishTwitter = portType/publishTwitter in twitter
6 alias searchCrime = portType/searchCrime in amazondynamodb 
7 alias showMap = portType/showMap in googlemaps 
//Services
8  service notifyCrime = publishCrime  . publishTwitter 
9  service trackCrime= searchCrime . verifyService
10 Service visualizeCrime =  showMap . getCrimeDetail
//Path
11 (notifyCrime.storeiCrime) || (trackCrime.visualizeCrime.getCrimeDetail)
//Contracts
12 defContract notifyCrimeContract{  isAppliedTo: notifyCrime
13	  requires: userId == ?? && passw == ?? && req(notifyCrime)  < 3
14 	      (OnFailureDo: NOT(action_publish(crime));
15	  ensures: publishTwitter(crime) == true  (OnFailureDo: skip); }
\end{verbatim}
% \end{lstlisting}
\end{small}
\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
\end{figure}
The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
The figure shows a simplified program code, produced in accordance to the following guidelines:
\begin{itemize}
\item
Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
\item
Specific operations exported by each business collaborator are identified as an \textit{operation} of the program (to each operation is given an \texttt{alias}).
\item
The workflow in Figure~\ref{fig:piSC} is translated into the text at line 11.
\item
\textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed whenever a condition is not verified. 
Contracts are generated from Policies (Figure~\ref{fig:crimePolicies}).
\end{itemize}

\section{Related work}
\label{sec:related}
Over the last years, a number of approaches have been proposed for the
development of web services. 
These approaches range from the proposal of new 
languages for web service descriptions~\cite{bpel03,SBS04} 
to techniques to support phases of the development cycle of this kind of
software~\cite{lipari2007}. 
In general, these approaches concentrate on specific problems, like supporting transactions or
QoS, in order to improve the security and reliability of service-based
applications. 
Some proposals address service composition: workflow
definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
services~\cite{BHM06}. 
%The proposed solutions come from many communities, 
%including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
%Software Engineering~\cite{Aal03,MendesPDB09}, Programming
%Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.

Works dealing with non-functional properties in service-oriented 
development can be organized in two main groups: those working on the \textit{modelling 
of particular non-functional properties or QoS attributes} and those 
proposing \textit{architectures or frameworks to manage and 
validate QoS attributes in web service composition processes}. 
The first group considers specific non-functional concerns (e.g., security)
which is modelled and then associated to functional models of the application. 
The work of Chollet et al.~\cite{CholletL09} defines a proposal to associate non-functional 
quality properties (security properties) to functional activities in a web service composition model. Schmeling et al.~\cite{SchmelingCM11} present 
an approach and also a toolset for specifying and implementing non-functional 
concerns in web service compositions.
Non-functional concerns are modelled and then related to a service composition 
represented in a BPMN diagram. 
Ovaska et al.~\cite{OvaskaEHPA10} present an approach to 
support quality management at design time. 
Quality requirements are modelled in a first phase and then
represented in an architectural model where quality requirements are associated to some
components of the model. 
We propose to model functional and non-functional properties at the same time during the software process.
We claim what this approach simplifies the web service development task. 
Non-functional properties are represented in our work as constraints and policies 
but more general quality model such as the proposed by~\cite{Goeb2011,Klass2009} 
could be taken in consideration in further works.

The second group of works dealing with non-functional requirements for services 
propose specific architectures or frameworks to manage and validate 
QoS attributes in service composition processes~\cite{XiaoCZBOLH08,Babamir2010,Karunamurthy2012787}. 
Our proposal is similar to these, but focusses on more general non-functional properties.

Despite the variety of techniques proposed, there is not yet a consensus on a software methodology 
for web services. 
Some methodologies address the service-based 
development towards a standard or a new way to develop reliable applications. 
SOD-M and SOMF~\cite{somf} are MDD approaches for web services; 
S-Cube~\cite{scube2010book} is focused on the representation of business processes and 
service-based development; SOMA~\cite{soma} is a methodology 
for SOA solutions; 
%Extended SOA~\cite{PapazoglouH06} merges RUP and
%BPM~\cite{bpm} concepts for service modelling; 
DEVISE~\cite{DEVISE} is a
methodology for building service-based infrastructure for collaborative
enterprises. 
Other proposals include the WIED model~\cite{TongrungrojanaL04}, 
that acts as a bridge between business modelling and design models on one hand, and traditional 
approaches for software engineering applied to SOC on the other hand.

\section{Conclusions}
\label{sec:conclusions}
%\input{conclusion}

This paper presented the $\pi$SOD-M software method for specifying and designing service-based applications in the presence of  non-functional constraints. 
Our proposal enhances the SOD-M method with constraints, policies and contracts  to consider non-functional constraints of applications.
We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developing a simple application.

$\pi$SOD-M is being used in an academic environment.
So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
We are now working on the definition of a PSM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS) in order to be able to generate code for the de-facto service coordination standard.

\section*{Acknowledgements}
\small
This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financed by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy.

\normalsize  
\bibliographystyle{splncs_srt}
\bibliography{biblio2} 
\end{document}

### 2013-10-22 SOFSEM2014/FinalVersion/conclusion.tex
Final version to SOFSEM!

@@
 Our proposal enhances the SOD-M method with constraints, policies and contracts  to consider non-functional constraints of applications.
-We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developping a simple application.
+We implemented the proposed meta-models on the Eclipse platform and we illustrated the approach by developing a simple application.
 
@@
 So far, the preliminary results indicate that $\pi$SOD-M approach is useful for the development of complex web service applications. 
-We are now working on the definition of a PSM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS).
+We are now working on the definition of a PSM-level meta-model to generate BPEL programs (instead of $\pi$-PEWS) in order to be able to generate code for the de-facto service coordination standard.

### 2013-10-22 SOFSEM2014/FinalVersion/introduction.tex
Final version to SOFSEM!

@@
 \noindent 
--\textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
+- \textit{Computation Independent Models (CIM)}: this viewpoint represents the software system at its highest level of abstraction. It focusses on the system
 environment and on  business and requirement specifications. 
@@
 \noindent 
--\textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
+- \textit{Platform Independent Models (PIM)}: this viewpoint focusses on the system functionality, hiding the details of any particular platform. 
 %The specification defines those parts of the system that are independent of the development platform.
@@
 \noindent 
--\textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
+- \textit{Platform Specific Models (PSM)}: this viewpoint focusses on the functionality, in the context of a particular implementation platform.
 Models at this level combine the platform-independent view with specific aspects of the target platform in order to implement the system.
@@
 
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
+Protocols and models implementing non-functional properties assume the existence of a global control of the artefacts implementing the application.
 They also assume that each service exports its interface.

### 2013-10-22 SOFSEM2014/FinalVersion/main.tex
Final version to SOFSEM!

@@
 Works dealing with non-functional properties in service-oriented 
-development can be organized in two main groups: those working on the \textit{modeling 
+development can be organized in two main groups: those working on the \textit{modelling 
 of particular non-functional properties or QoS attributes} and those 
@@
 Other proposals include the WIED model~\cite{TongrungrojanaL04}, 
-that acts as a bridge between business modeling and design models on one hand, and traditional 
+that acts as a bridge between business modelling and design models on one hand, and traditional 
 approaches for software engineering applied to SOC on the other hand.
@@
 \small
-This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financied by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy.
+This research is partly supported by the National Institute of Science and Technology for Software Engineering (INES\footnote{www.ines.org.br}), funded by CNPq (Brazil), grants 573964/2008-4 and 305619/2012-8; CAPES/UdelaR (Brazil/Uruguay) grant 021/ 2010; CAPES/STIC-AmSud (Brazil) grant 020/2010); MASAI project (TIN-2011-22617) financed by the Spanish Ministry of Science and Innovation and the Spanish Network on Service Science (TIN2011-15497-E) financed by the Spanish Ministry of Competitiveness and Economy.
+
 \normalsize  
-
 \bibliographystyle{splncs_srt}

### 2013-10-22 SOFSEM2014/FinalVersion/pisodm.tex
12 pages, Valeria's reccomndations.
Final version to SOFSEM!

@@
 $\pi$SOD-M environment is built on the top of Eclipse. 
-We also used the Eclipse Modeling Framework (EMF) to define, edit and handle
+We also used the Eclipse Modelling Framework (EMF) to define, edit and handle
 (meta)-models. 
-To automate the transformation models we use ATL~\cite{atl_manual} and Acceleo~\cite{acceleo}.
+To automate the transformation models we use ATL~\cite{atl_manual} and Acceleo.
 

### 2013-10-29 JournalToSubmit2013/modelingFlyingPig.tex

--- 2013-10-07 JournalToSubmit2013/2-modeling.tex
@@
 %\end{figure}
+
+In order to introduce the context of our work consider the FlyingPig scenario concerning a risk assesment system, where finanacial companies need to be assesed about the risks implied by their business practice.
+For instance, the condition in which confidential transactions are performed, the physical security for accessing reserved areas (i.e.
+servers storing).
+The system provides a set questionaires that provide a thorough description of the practice of the company.
+This information is used to determine whether there are risky actions within the business processes pof the company that can be amended.
+It is also used to determine a degree of compliance of security management of financial companies norms.
+When things have to be amended it is necessary to verify that the amendment reduces the risk.
+
+ In order to provide these functions the application benefits from existing services that provide information,  storage and data visualization functions.
+Thus, the application is a service based application that implements a business process.
+ The business process is specified in terms of ordered tasks and define the application logic of a Service oriented Application.
+Tasks can be performed by a person or an entity.
+In our context, tasks are implemented by Services.
+Service is an application implemented by a provider that exports an API through a network (e.g., Internet)
+API is defined using an interface definition language (IDL, e.g., WSDL for Web services).
+ In our example the business procces can start with one of two tasks: Notify a crime, or track a crime.
+A notified crime can be then stored in a database.
+Tracked crimes are visualized in a map and then the used can ask for detailed information.
+They are implemented by four services: twitter and an adhoc police service for notifying crimes, Amazon as persistence service and google maps for visualizing and locating crimes in a map.
+Note that a task can be implemented by a service composition.
+This is the case of the task Notify crime in our example that enables to notify crimes through twitter or through the police notification service.
+
+Business processes have also associated rules and constraints that define their non functional requirements.  
+NFR represents the “semantics” and the conditions in which the tasks must be done. 
+In our example we have some constraints. 
+\begin{itemize}
+\item Twtter requires to be accessed through an authentication protocol, and the police crime notification service has a control  access protocol and allows only 3 tries to access it. 
+\item Only users with enough privileges can store crimes’ notifications. 
+\item Users can only track crimes notified by providers t that hey have authorization to contact; For example civil population cannot track all the crimes notified by the police. 
+\item If the google map is unavailable the results of a track request are delivered on text. 
+\item If a user tracks crimes that are provided by a service for which she does not have authorization then show an empty map. 
+\item Detailed information about a crime depends on the user privileges, on the police assignment regions. 
+\item The tracking crime app keeps track of the accesses to detailed information about crimes. 
+\end{itemize}
+
+Through the example we underlined that every application implements functional aspects that describe its application logic.
+Recall that an application logic refers to routines that perform the activities to reach the application objective.
+Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
+These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
+
+The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
+API does not necessarily export information about methods dependency (e.g., in the REST protocol);
+they do not share their state (stateless).
 

### 2013-10-29 JournalToSubmit2013/modelingFlyingPig.tex
no message

@@
-%Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-%For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-%\begin{itemize}
-%\item The music service   Spotify exports a method for obtaining information  %about the music a given user is listening:
-%%\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; %%\end{itemize}
-%\item Facebook and Twitter services export a method for  updating the status of a given user:
-%\begin{itemize} 
-%%\item{\sf\small get-Status ( usedid ): String ; 
-%\item {\sf\small update-Status ( userid, new-status ): String}; 
-%\end{itemize}
-%\end{itemize}
-%
-%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-%the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-%Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-%The "To Publish Music" scenario 
-%%(showed by doted line) 
-%starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-%Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-%\begin{figure}
-%\centering
-%\includegraphics[width=0.55\textwidth]{figs/SC}
-%\caption{BPMN model of the "To Publish Music" scenario}
-%\label{fig:E3valuemodel}
-%\end{figure}
-
 In order to introduce the context of our work consider the FlyingPig scenario concerning a risk assesment system, where finanacial companies need to be assesed about the risks implied by their business practice.
 For instance, the condition in which confidential transactions are performed, the physical security for accessing reserved areas (i.e.
-servers storing).
+access to the critical servers room should be secured).
 The system provides a set questionaires that provide a thorough description of the practice of the company.
-This information is used to determine whether there are risky actions within the business processes pof the company that can be amended.
+This information is used to determine whether there are risky actions within the business processes of the company that can be amended.
 It is also used to determine a degree of compliance of security management of financial companies norms.
-When things have to be amended it is necessary to verify that the amendment reduces the risk.
+When things have to be amended it is necessary to verify that evidence that the amendment was done and whether it  reduces the risk.
 
- In order to provide these functions the application benefits from existing services that provide information,  storage and data visualization functions.
-Thus, the application is a service based application that implements a business process.
+ In order to provide these functions the application FlyingFish benefits from existing services that provide information,  storage, assesment and data visualization functions.
+Thus, FlyingFish is a service based application that implements a business process.
  The business process is specified in terms of ordered tasks and define the application logic of a Service oriented Application.
@@
 API is defined using an interface definition language (IDL, e.g., WSDL for Web services).
- In our example the business procces can start with one of two tasks: Notify a crime, or track a crime.
-A notified crime can be then stored in a database.
-Tracked crimes are visualized in a map and then the used can ask for detailed information.
-They are implemented by four services: twitter and an adhoc police service for notifying crimes, Amazon as persistence service and google maps for visualizing and locating crimes in a map.
-Note that a task can be implemented by a service composition.
-This is the case of the task Notify crime in our example that enables to notify crimes through twitter or through the police notification service.
+ In our example the business procces  starts by a request from a company branch asking for an assesment.
+The request leads to the declaration of a group of users that will answer quetionaires that will be used as input for evaluating risk.
+Questionaires are considered tasks that users will have to perform.
+Given a list of tasks a user decides whether she takes on the task (i.e., answering a questionaire) or delegate it to another user.
+
+During the process of answering a questionnaire, the answer of a question can trigger the generation of other questonaires, that become new tasks to be executed.
+Once a questionnaire has been completely answered it is stored and processed to generate a set of uncompliant situations associated to their corresponding amendments if there are any or a report specifying a compliance level, incidents and a risk map.
+
+Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
+the scenario.
+The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
+The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
+The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
+Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
+
+Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
+
+%The "To Publish Music" scenario 
+%%(showed by doted line) 
+%starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
+%Twitter and Facebook services are then contacted in parallel for %updating the user's status with the corresponding song title %(activities {\sf Update Twitter} and {\sf Update Facebook}).
+
+\begin{figure}
+\centering
+\includegraphics[width=0.55\textwidth]{figs/SC}
+\caption{BPMN model of the "To Publish Music" scenario}
+\label{fig:E3valuemodel}
+\end{figure}
 

### 2013-10-30 JournalToSubmit2013/modelingFlyingPig.tex
no message

@@
 
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{figs/ValueModel_GCP}
+\caption{E3value model of the FlyingPig scenario}
+\label{fig:E3valuemodel}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{figs/BPMN_GCP.png}
+\caption{BPMN model of the FlyingPig scenario}
+\label{fig:bpmn}
+\end{figure}
+
+Figure \ref{fig:bpmn} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
 
@@
 %Twitter and Facebook services are then contacted in parallel for %updating the user's status with the corresponding song title %(activities {\sf Update Twitter} and {\sf Update Facebook}).
-
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
 

### 2013-10-31 CAISE2011/submitted/body.tex
CAISE2011 y algo mas

%*********************************************************************************************************
\section{Introduction}
%*********************************************************************************************************
Along with the emergence of Web 2.0 and services based computing,  the Web has become a software as services development platform for applications. Lambda users have thereby access to different services some exported by their favorite social networks, music and leisure providers.  As in other platforms, the applications' integration problem emerges again  and calls for simple coordination solutions that can give users' a global view of their Web services within a whole, unique and, dynamic space. For example, having an integrated view of the mood of a user according to the status of her favorite social networks. The integration of data services  can be done today by coordinating services accessible on the  network. For example, synchronizing music lists listened in LastFM  with a user account in Twitter and Facebook ({\em status: "Desert Rose - Sting"}) and observe and use this information to determine the degree of music preference convergence of a community of users. Yet, such services produce and update some of these data dynamically (e.g. the status in messenger when the user listens music) and integrated views must be continuously and automatically updated according to specific  requirements: data freshness, periodic and atomic update of data shared among several services despite eventual exceptions.

Current standards in services' coordination implement functional, non-functional properties (NFP) and communication aspects by combining different languages and protocols. WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services. For adding a transactional behaviour to a services' coordination it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction. The selection of the adequate protocols for adding a specific NFP to a services' coordination (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer. As a consequence, the development of an application based on a services' coordination is a complex and a time-consuming process. This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. Therefore, in contrast to these approaches, our work is inspired in the philosophy of \textit{separation of concerns} adopted in the construction of middleware for specifying NFP's to a services' coordination in orthogonal way.
%(ii) generating mechanisms for ensuring these properties at services' coordination execution time.

This paper presents a methodology   for ...  to ensure NFP's policies specified by the coordination designer. For example, update the status of all user Facebook, Twitter and Yahoo accounts or maintain the current status if one of them is not accessible.  Thanks to policies associated to a services based application running in a Web dynamic environment, it is possible to associate a personalized behaviour: atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks.

The remainder of the paper is organized as follows. 
%Section \ref{sec:policycoord} introduces our approach for defining policies and associating them to service coordination for adding and ensuring NFP's at execution time. Section \ref{sec:reaction} describes observation and reaction strategies used for monitoring the execution of the coordination and reacting on it according to the policies associated to the coordination. Section \ref{sec:musicmood} describes implementation issues and a validation experience integrating well-known social networks services namely FaceBook, Twitter and, LastFM. 
%The section insists on the synchronization of the messages expressed automatically or explicitly by social network users and then shows how the services are coordinated observed and then synchronized thanks to policies. 
Section \ref{sec:related} analyses related work concerning policy/contract based programming and, service coordination platforms. Section \ref{sec:conclusions} concludes the paper and discusses future work.

%*********************************************************************************************************
\section{Framework overview }\label{sec:approach}
%*********************************************************************************************************
Figure \ref{fig:framework} shows the general overview to the framework we propose for specifying and designing reliable service based applications. The objective   is to model and associate non functional properties to  service based applications that represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints stemming from the services used for implementing them (e.g., the fact that a service requires imposes an authentication protocol for executing a method)  Thus, the framework proposes a meta-modelling approach that  (i) extends the  CIM and PIM ??? layers of the SODM methodology \ref{decastro-etal10} that  provides concepts for modeling services' compositions with the notion of {\sf\small Policy} for representing nonfunctional properties and, (ii) defines  transformation rules for generating concrete reliable service compositions implementations. Service compositions are implemented in a intermediate composition language named PEWS \ref{pews} that provides constructs for expressing the composition and the policies. PEWS expressions are then transformed into Java programs using the PEWS metamodel and transformation rules into Java that we have defined.

\begin{figure}[htpb]
\begin{center}
%\includegraphics[width=100.0mm]{figs/SC}
\end{center}
		\caption{General overview of the framework}
   \label{fig:framework}

\end{figure}

%*********************************************************************************************************
%\section{Running example}\label{sec:usecase}
%*********************************************************************************************************

Consider for instance the following scenario. An imaginary music provider...

A programmer wants to develop the application Status Broadcaster that estimates the "mood" of people  based on the music they are listening during some periods of time and sends the song title and the mood to Twitter and Facebook. For example, if we consider that people listening to pop-music are "happy" and most songs listened by a person are pop-music, then the application estimates that the person is "happy". Social network users will have their status synchronized in  Twitter and Facebook  with the title of the music they are listening in LastFM (http://www.lastfm.fr/) and the associated mood.
For developing this application the programmer has to coordinate the calls to the methods offered by the following services:
\begin{itemize}
\item Music Services like  LastFM export methods for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item Facebook and Twitter services export methods for getting and updating the status of a given user: 
\begin{itemize} \item{\sf\small get-Status ( usedid ): String ; \item update-Status ( userid, new-status ): String}; \end{itemize}
\item Mood Service exports a method for retrieving the mood of a  user: 
\begin{itemize} \item{\sf\small get-Mood ( userid ): String} ;\end{itemize}
It is a complex process that uses a log of the music listened by the user in LastFM during some period of time, and then according to the classification of each song it will determine her mood.
\end{itemize}
The corresponding services' coordination is represented by the workflow depicted in Figure \ref{fig:musicmoodwk} using BPMN \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}.
\begin{figure}[htpb]
\begin{center}
\includegraphics[width=100.0mm]{figs/SC}
\end{center}
		\caption{Status Broadcaster process expressed as a workflow}
   \label{fig:musicmoodwk}

\end{figure}
As shown in the Figure, the process Status Broadcaster starts by contacting the music service LastFM for retrieving the user's  musical status. This information is sent to the mood service in order to calculate the mood of the user. Therefore, the Twitter and Facebook services are called for retrieving the status of the user which is compared with other analysis data in order to determine the mood.  This process is not explicitly expressed in the Figure, it is implemented by the complex activity Compute Mood  (see the C. Activity representation in the Figure  \ref{fig:musicmoodwk}).  Finally, Twitter and Facebook services are contacted in parallel for updating the user's status with the corresponding song and mood.

Now consider that the Facebook and Twitter services require authentication protocols in order to execute methods that will read and update the users' space. When these services are called by activities of a coordination the call must be part of the authentication protocol required by these services.  Some services use the http authentication protocol that consists in providing the login and password of the user within the communication protocol. Once the access granted it is possible to read and write the status content  of the user's wall.   Twitter and Facebook use the open authentication protocol that consists in making participate a confidence tier that provides a token for granting access to a user's account. In the following we focus the notion of policy used for specifying the authentication protocols for  services' coordinations.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{E3value model}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

\begin{figure}        
\centering  
%\fbox{
%\epsfig{file=soa.png, height=1.5in, width=2.1in}
\includegraphics[width=0.60\textwidth]{figs/e3value}
%height=2in, width=2in
%}
\caption{E3value model of the Music-shop scenario.} 
\label{fig:E3valuemodel}
\end{figure}

\begin{figure}        
\centering  
%\fbox{
%\epsfig{file=soa.png, height=1.5in, width=2.1in}
\includegraphics[width=0.60\textwidth]{figs/BS-Model}
%height=2in, width=2in
%}
\caption{Use case diagram of the scenario} 
\label{fig:usecase}
\end{figure}

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Service composition model}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

\begin{figure}        
\centering  
%\fbox{
%\epsfig{file=soa.png, height=1.5in, width=2.1in}
\includegraphics[width=0.80\textwidth]{figs/Composition-Model}
%height=2in, width=2in
%}
\caption{Service composition model} 
\label{fig:servicecompositionmodel}
\end{figure}

%*********************************************************************************************************
\section{Reliable SODM}\label{sec:sodm}
%*********************************************************************************************************
Explanation of the extension of the SODM

%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._
\subsection{Stereotyped SODM}
%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._

%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._
\subsection{Policy}
%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._
Figure \ref{fig:policymodel} shows the UML class diagram of our A-Policy model. As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Activity}, an {\sc Operator}, a {\sc Workflow} or a {\sc Policy}. 
It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of tye E occurs if  condition C is verified then execute the action A}.  Thus, an A-Policy represents a set of re-actions to be possibly executed if one or several triggering events of its rules are notified.

\begin{figure}[htpb]
	\begin{center}
		\includegraphics[width=110.0mm]{Figures/AP}
	\end{center}
		\caption{UML class diagram of the A-Policy Model}
   \label{fig:policymodel}
\end{figure}

\begin{itemize}
\item The class {\sc Scope} represents any element of a services' coordination (i.e., activity, operator, workflow).
%a process as an ordered set of execution states  as a set of values of the variables of the execution state of a services' coordination at instant t (or timestamp). The states of a scope are ordered with respect to their timestamps.

\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}. A policy has variables that represent the view of the execution state of its associated scope, that is required for executing the rules. The value of a variable is represented using the type Variable.
\end{itemize}

%*********************************************************************************************************
\section{PEWS metamodel}\label{sec:pewsmetamodel}
%*********************************************************************************************************

Figure \ref{fig:metamodel} presents the PEWS-CT metamodel. Each class refers to
the language terms that can be expressed with it. The structure of the metamodel
has its beginning in the representation of the object \textit{PEWS-CT
Specification}. This class is directly linked to three other classes, they are:
\textit{Namespace, Variable} and \textit{Path Expression}. That classes, in
figure \ref{fig:pews} it are represented in the lines (\textit{1-3, 18-19} and
\textit{32}), respectively.

A question could be made. How about the operations and services also expressed
in the specification, why they are not directly related to the class
\textit{PEWS-CT Specification}? The answer is because every operation and
service, is directly connected to a namespace. The operations and their
compositions do not exist without the \textit{namespaces}.

\begin{figure}        
\centering  
%\fbox{
%\epsfig{file=soa.png, height=1.5in, width=2.1in}
\includegraphics[width=0.99\textwidth]{figs/PEWSMetamodel.png}
%height=2in, width=2in
%}
\caption{PEWS-CT Metamodel.} 
\label{fig:metamodel}
\end{figure}

The \textit{Service} class abstracts both, a single \textit{Operation} and a
composition (\textit{Operation Composition}) represented by the service clause
in PEWS-CT. A \textit{Operation Composition} is a aggregation of one or more
\textit{Operations}. All services have one \textit{Alias} and are composed using
different \textit{Composition Operator}, such as \textit{sequence, parallel,
choice} and so on.

Each service has zero or more \textit{Contract}s. A contract is a composition of
\textit{Constraint Condition}, that can be as pre-condition, post-condition or
time restriction. A contract condition, if necessary, can performs an
\textit{Action}. An action can be a call to another service; throws an
exception; or stop the process.

The \textit{Constraint Set} is the role constraint in a requires or ensures
clause (figure \ref{fig:pewscontract} ,lines 3-5 and 7), while a \textit{Constraint} is
a single \textit{Expression}, that can be unary or not.

%*********************************************************************************************************
\section{Implementation issues}\label{sec:musicmood}
%*********************************************************************************************************
We conducted an implementation of the Status Broadcaster application for synchronizing the status of users in different social networks. We used the Windows Workflow Foundation runtime environment and its underlying notification service that enables us to couple the policies execution.

The workflow is a simplified version of our example that retrieves the music listened by a user on LastFM and then it atomically updates the status of the user's account in Facebook and Twitter  and it stores the status in a log. This means that if one of the services is not available, the call is compensated. For doing so, the status is updated with the previous one and then the coordination starts again. Figure \ref{fig:musicmoodpolicies}  shows the policies that implement these aspects. They are expressed using our policy language that includes constructs for expressing event types (simple and compound) in the ON part and simple programs in the DO part of the policy rules. Policies are classes of type {\sc a-policy} and they have a scope of type {\sc Activity}, {\sc Workflow} and, {\sc Policy}.

\begin{figure*}
\begin{center}
\includegraphics[scale=.50]{figs/Policies}
\end{center}
		\caption{Status broadcaster policies}
   \label{fig:musicmoodpolicies}
\end{figure*}

The policy {\sc\small Compensation} in Figure \ref{fig:musicmoodpolicies} is associated to the parallel operator and handles an exception of any of the activities {\sc\small Update Facebook/Twitter}   compensating it. Therefore it uses the backup status   retrieved by the {\sf\small Status Backup } policy for updating the status in the user wall. The policy {\sc\small Status-Backup} logs the status used for updating the FaceBook and Twitter.

%Given a services' coordination and a set of policies, we generate the code for evaluating every policy as well as the synchronization interface to interact with the services' coordination execution-engine. 
%(see Figure  \ref{fig:9}). 
%A policy-compiler validates whether type declarations and policy expressions are well formed and implements transformation rules for generating classes. The components of a policy-engine import these classes and implement evaluation strategies for executing the policies. The classes implementing the event-types are used for defining the synchronization interface since event detection and event notification are specified on this interface. The classes implementing the actions are used by the policy-engine for synchronizing the execution of services' coordination through the synchronization interface.
%The policy engine consists of a set of components (coordinated by the N-FP Evaluator component) that can be specialized for evaluating policies. The policy evaluation process must be synchronized with a services’ coordination. Thus, the policy engine interacts with an coordination engine through a synchronization interface. This interface is a complex component that implements messages based asynchronous communication, where an event type represents a message.

%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._
\subsection{EMF SODM/PEWS}
%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._

%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._
\subsection{Transformation M-M}
%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._

%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._
\subsection{Transformation PIM - PCM}
%._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._._

%*********************************************************************************************************
\section{Related works}\label{sec:related}
%*********************************************************************************************************

%*********************************************************************************************************
\section{Conclusions and future work}\label{sec:conclusions}
%*********************************************************************************************************

### 2013-10-31 CAISE2011/submitted/caise-etal old-old.tex
CAISE2011 y algo mas

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
-%% This is file `elsarticle-template-1a-num.tex',
-%%
-%% Copyright 2009 Elsevier Ltd
-%%
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-%%
-%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
-%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
-%%
-\documentclass[preprint,12pt]{elsarticle}
+\documentclass{llncs}
 \usepackage{geometry}
@@
 \usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
 \begin{document}
 
-\begin{frontmatter}
+\title{Generating reliable services' composition  using policies: a model-driven approach\thanks{This work is partially financed by the projects {\sc Orchestra} ECOS-ANUIES, {\sc Clever} STIC-AMSUD, and {\sc Masai}.}}
 
-%% Title, authors and addresses
+\author{Genoveva Vargas-Solar\inst{1}
+\and Valeria de Castro\inst{4}
+\and Pl\'{a}cido Antonio de Souza Neto\thanks{Financed by CAPES/STIC-AMSUD Brasil, BEX 4112/11-3} \inst{5,6}
+\and Javier A. Espinosa-Oviedo\inst{3}
+\and Esperanza Marcos\inst{4}
+\and Martin A. Musicante\inst{5}
+\and Jos\'{e}-Luis Zechinelli-Martini\inst{2,1}
+\and Christine Collet \inst{3}}
 
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
+\institute{
+	French Council of Scientific Research, LIG-LAFMIA	\\
+	681 Rue de la Passerelle, BP. 72, 38402	Saint Martin d'H\`{e}res Cedex, France		\\
+		\email{Genoveva.Vargas-Solar@imag.fr}
+\and
+	Fundaci\'{o}n Universidad de las Am\'{e}ricas, Puebla	\\
+	Exhacienda Sta. Catarina M\'{a}rtir s/n 72820, San Andr\'{e}s Cholula, Puebla, M\'{e}xico			\\
+	\email{joseluis.zechinelli@udlap.mx}
+\and
+	Grenoble Institute of Technology	\\
+	46 Avenue F'{e}lix Viallet, 38031 Grenoble Cedex 1		\\
+	\email{Christine.Collet@grenoble-inp.fr}, \email{Javier.Espinosa@imag.fr} \\
+\and
+	Universidad Rey Juan Carlos	\\
+	Av Tulip\'{a}n, M\'{o}stoles, Spain 		\\
+	\email{Valeria.deCastro@urjc.es, esperanza.marcos@urjc.es} \\
+\and
+	DIMAp - UFRN, ForAll - Formal Methods and Language Research Laboratory	\\
+	Campus Universitário - Lagoa Nova, Natal - RN, Brasil		\\
+	\email{mam@dimap.ufrn.br} \\
+	\and
+	Instituto Federal do Rio Grande do Norte	\\
+	Av. Senador Salgador Filho, 1559 - Tirol, Natal - RN, Brasil		\\
+	\email{placido.neto@ifrn.edu.br}
+}
 
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
+\maketitle
 
 \begin{abstract}
-This paper presents\ldots
+This paper presents an approach for modeling and associating policies to  services' based applications.
+%and thereby representing both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available, recovery, persistence) and  constraints associated to   services, that must be respected for using them (e.g., the fact that a service requires  an authentication protocol for executing a method).
+It proposes to extend the SOD-M model driven method with   (i) the $\pi$-{\sc SCM} a {\em A-Policy} services' composition meta-model for representing non-functional constraints associated to services' based applications;  (ii) the $\pi$-{\sc Pews} meta-model providing guidelines for expressing the composition and the policies; and, (iii)  model to model and model to text transformation rules for semi-automatizing the implementation of reliable services' compositions
+%for generating  $\pi$-{\sc Pews} models, and model to text rules for generating executable code implementing  .
+% (iv) defines model to text transformation rules for generating the  program that implements both the services' composition and the associated policies and that is executed by an adapted engine.
+As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
 \end{abstract}
 
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
+%*********************************************************************************************************
+\section{Introduction}
+%*********************************************************************************************************
 Service oriented computing is at the origin of an evolution in the field of software development. 
@@
 The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
+In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users’ goals \cite{watson}.
 
 Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
+Non-functional aspects concerning services and application "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
 This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
@@
 The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
+Section \ref{sec:motivation} describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
+Section \ref{sec:sodm} describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable.
+Section \ref{sec:implementation} describes implementation issues.
 Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
@@
 %*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
+\section{Motivation example}\label{sec:motivation}
 %*********************************************************************************************************
+
 Consider for instance the following scenario.
@@
 
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
 Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
 
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
@@
 
+Model-Driven Engineering (MDE) \cite{schmidt} and particularly Model Driven Architecture (MDA) \cite{miller} \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} provide
+%is an evolving approach to software development that deals with the provision of 
+models, transformations between them and code generators to address software development. 
+%One of the main advantages of model-driven approaches is the provision of 
+It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
+
+This paper proposes the $\pi$-SOD-M, a model-driven  method \cite{decastro1,decastro2} for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition.
+This method is described in the following sections.
+
+%*********************************************************************************************************
+\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:sodm}
+%*********************************************************************************************************
+ Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach that  provides  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
+Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a MDA-based approach.
+Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+\begin{figure} [htpb]
+\centering
+\includegraphics[width=0.65\textwidth]{figs/SODM}
+\caption{SOD-M development process}
+\label{fig:sodm}
+\end{figure} 
+%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views.
+
+As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
+Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
+
+As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
+Thus, in this paper we  focus on $\pi$-SCM the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{subsec:piscm}).
+$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{subsec:mmrules}).
+Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
+
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
+\subsection{$\pi$ services' composition meta-model}\label{subsec:piscm}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
+The {\em A-policy} based services' composition meta-model ($\pi$-SCM in  Figure \ref{fig:sodm})
 represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
 This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
+Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivities} and {\sc Actions} (see the  elements  highlighted in blue in Figure \ref{fig:servicecompositionmodel}).
 
@@
 
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
+\item {\sc Actions}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
 Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
@@
 We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
 
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
+A {\sc Rule} represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
+A {\sc Rule} can be associated to activity boxes, to the control flow, message passing and to the service activities of the $\pi$-SCM.
 %An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
 %An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
+An {\em A-policy} groups a set of rules. It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event and and condition parts.
 \begin{figure}[htpb]
@@
 Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
+For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks.
+Or for providing security in the communication channel when the payment service is called.
 
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
+\subsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
@@
 
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
+Given a $\pi$-SCM model of a specific services' based application, it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
 The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
+\subsection{Transformation rules}\label{subsec:mmrules}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
+Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM used for representing the services' composition into the elements of the $\pi$-{\sc Pews}.
+There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} elements; and those that transform rules grouped by policies into {\em A-policy} types.
 
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
+\noindent
 
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
+{\bf\em Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements:}
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
@@
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
+\noindent
 
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
+{\bf\em Transformation of rules grouped by a-policies   in the $\pi$-SCM to A-Policies of the $\pi$-{\sc Pews}:}
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
@@
 %*********************************************************************************************************
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
+This section  depicts the $\pi$-SOD-M development environment that implements the generation of reliable services' compositions.
 For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
@@
 Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[htpb]
-	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
-	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
-   \label{fig:policymanager}
-\end{figure}
+
 \begin{itemize}
@@
 %
-
+\begin{figure}[htpb]
+	\begin{center}
+		\includegraphics[width=0.60\textwidth]{figs/architecture}
+	\end{center}
+		\caption{$\pi$-SOD-M Development Environment}
+   \label{fig:policymanager}
+\end{figure}
 %
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
+\noindent{\bf\em Programming non-functional properties for services}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
@@
 that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
+to compose transactional services. In {\cite{Vid04} the model
 introduced in \cite{schuldt-etal-TODS} is extended to web services
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
+\noindent{\bf\em Modeling non-functional properties}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
+\noindent{\bf\em Discussion}
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
@@
 
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
+\bibliographystyle{plain}
+\bibliography{caise,S-Coord,Bib/e-commerce,Bib/trans,Bib/Component,Bib/XML,Bib/Workflow,Bib/SODM,Bib/G,Bib/docbi}
 
-%% \section{}
-%% \label{}
+%*********************************************************************************************************
+%\section*{Appendix}\label{sec:appendix}
+%*********************************************************************************************************
+%Figure \ref{fig:mappingrules-1} shows the transformation rules between the $\pi$-SCM meta-model to the $\pi$-{\sc Pews}  meta-model.
+%\begin{figure}
+%\includegraphics[width=1.0\textwidth]{figs/Mapping-1}
+%\caption{Mapping rules between $\pi$-SCM  and $\pi$-{\sc Pews} meta-models}
+%\label{fig:mappingrules-1}
+%\end{figure}
 
-%% References
-%%
-%% Following citation commands can be used in the body text:
-%% Usage of \cite is as follows:
-%%   \cite{key}          ==>>  [#]
-%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
-%%   \citet{key}         ==>>  Author [#]
+%\begin{figure}
+%\includegraphics[width=0.70\textwidth]{figs/Mapping-3}
+%\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} policy elements}
+%\label{fig:mappingrules-3}
+%\end{figure}
 
-%% References with bibTeX database:
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%% Authors are advised to submit their bibtex database files. They are
-%% requested to list a bibtex style file in the manuscript if they do
-%% not want to use model1a-num-names.bst.
-
-%% References without bibTeX database:
-
-% \begin{thebibliography}{00}
-
-%% \bibitem must have the following form:
-%%   \bibitem{key}...
-%%
-
-% \bibitem{}
-
-% \end{thebibliography}
+%\begin{figure}
+%\includegraphics[width=0.70\textwidth]{figs/Mapping-2}
+%\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} rule elements}
+%\label{fig:mappingrules-2}
+%\end{figure}
 
 \end{document}
-
-%%
-%% End of file `elsarticle-template-1a-num.tex'.

### 2013-10-31 CAISE2011/submitted/caise-etal-old.tex
CAISE2011 y algo mas

\documentclass{llncs}
\usepackage{geometry}                
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\begin{document}

\title{Generating reliable service composition  using policies: a model-driven approach\thanks{This work is partially financed by the projects {\sc Orchestra} ECOS-ANUIES, {\sc e-CLOUDSS} Microsoft,  {\sc Clever} STIC-AMSUD, and {\sc Masai}.}}

\author{Genoveva Vargas-Solar\inst{1} 
\and Valeria de Castro\inst{4}
\and Pl\'{a}cido A. de Souza Neto\thanks{Financed by CAPES/SticAmbud Brasil, BEX 4112/11-3} \inst{5,6} 
\and Javier A. Espinosa-Oviedo\inst{3} 
\and Esperanza Marcos\inst{4} 
\and Mart\'{i}n A. Musicante\inst{5} 
\and Jos\'{e}-Luis Zechinelli-Martini\inst{2,1} 
\and Christine Collet \inst{3}}

\institute{
	French Council of Scientific Research, LIG-LAFMIA	\\
	681 Rue de la Passerelle, BP. 72, 38402	Saint Martin d'H\`{e}res Cedex, France		\\
		\email{Genoveva.Vargas-Solar@imag.fr}
\and
	Fundaci\'{o}n Universidad de las Am\'{e}ricas Puebla	\\
	Exhacienda Sta. Catarina M\'{a}rtir s/n 72820, San Andr\'{e}s Cholula, Puebla, M\'{e}xico			\\
	\email{joseluis.zechinelli@udlap.mx}
\and
	Grenoble Institute of Technology	\\
	46 Avenue F'{e}lix Viallet, 38031 Grenoble Cedex 1		\\
	\email{Christine.Collet@grenoble-inp.fr}, \email{Javier.Espinosa@imag.fr} \\
\and
	Universidad Rey Juan Carlos	\\
	Av Tulip\'{a}n, M\'{o}stoles, Spain 		\\
	\email{Valeria.deCastro@urjc.es, esperanza.marcos@urjc.es} \\
\and
	DIMAp - UFRN, ForAll - Formal Methods and Language Research Laboratory	\\
	Campus Universitário - Lagoa Nova, Natal - RN, Brasil		\\
	\email{mam@dimap.ufrn.br} \\
	\and
	Instituto Federal do Rio Grande do Norte	\\
	Av. Senador Salgador Filho, 1559 - Tirol, Natal - RN, Brasil		\\
	\email{placido.neto@ifrn.edu.br}
}

\maketitle

\begin{abstract}
This paper presents an approach for modeling and associating policies to  service based applications for representing both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available, recovery, persistence) and  constraints stemming from the services used for implementing them (e.g., the fact that a service requires imposes an authentication protocol for executing a method).  Our work proposes a model driven  methodology that  (i) extends     PIM layer's composition model of the SOD-M method \cite{decastro1} with the notion of {\sf\small A-Policy} \cite{Espinosa-Oviedo2011a} for representing constraints associated to service based applications and, (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the policies (iii) defines   Model to model transformation rules for generating  the implementation of reliable services' compositions.
% (iv) defines model to text transformation rules for generating the  program that implements both the services' composition and the associated policies and that is executed by an adapted engine.

\end{abstract}

%*********************************************************************************************************
\section{Introduction}
%*********************************************************************************************************
%Along with the emergence of Web 2.0 and services based computing,  the Web has become a software as services development platform for applications. Lambda users \footnote{Lambda user is a user with few computing knowledge.} have thereby access to different services some exported by their favorite social networks, music and leisure providers.  As in other platforms, the applications' integration problem emerges again  and calls for simple  solutions that can give users' a global view of their Web services within a whole, unique and, dynamic space.

%For example, having an integrated view of the mood of a user according to the status of her favorite social networks. The integration of data services  can be done today by coordinating services accessible on the  network. For example, synchronizing music posts of a user accounts in Twitter and Facebook ({\em status: "Desert Rose - Sting"}) according to the music listened in Spotify    and observe and use this information to determine the degree of music preference convergence of a community of users. Yet, such services produce and update some of these data dynamically (e.g. the status in Facebook when the user listens music) and integrated views must be continuously and automatically updated according to specific  requirements: data freshness, periodic and atomic update of data shared among several services despite eventual exceptions.

Service-Oriented Computing has provoked an evolution in the field of software development. The literature stresses the need for methodologies and techniques for service-oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful service-oriented applications \cite{papazoglou}. Service-oriented development methodologies providing models, best practices, and reference architectures to built service-based applications mainly address  functional aspects of service based applications \cite{1,2,decastro1,4}.  Non-functional aspects concerning services and application "semantics", often expressed as requirements, and constraints in general purpose methodologies are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). This leads to service based applications that are partially specified and that are thereby partially compliant with application requirements.

The objective of this work   is to model non-functional constraints and associate them to  service based applications  early during the service composition model phase. Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1}  with the notion of A-policy.    
SOD-M defines a model-driven method for  developing  service-oriented information systems. It defines a process  starting with the  identification of business services through business modeling, and, by means of models transformations it allows to obtain a service composition model \cite{decastro1}  extended with policies. Policies are used to express constraints which can be applied to all the service composition process or to a particular service used for implementing it. They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services used for implementing them (e.g., the fact that a service requires imposes an authentication protocol for executing a method).

%Figure \ref{fig:framework} shows the general overview of the approach we propose for specifying and designing reliable service based applications. 
Our work proposes a model driven  methodology that  (i) extends     PIM layer's composition model of the SOD-M method \cite{decastro1} with the notion of {\sf\small A-Policy} \cite{Espinosa-Oviedo2011a} for representing constraints associated to service based applications and, (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the policies (iii) defines   Model to model transformation rules for generating  the implementation of reliable services' compositions.

%\begin{figure}[htpb]
%
The remainder of the paper is organized as follows. Section \ref{sec:motivation} describes a motivation example that integrates and synchronizes well-known social networks services namely FaceBook, Twitter and, Spotify. Section \ref{sec:sodm} describes $\pi$-SOD-M method that enables the representation and association of policies to services composition and thereby making them reliable. Section \ref{sec:implementation} describes implementation issues.
Section \ref{sec:related} analyses related work concerning policy/contract based programming and, service composition platforms. Section \ref{sec:conclusions} concludes the paper and discusses future work.

%*********************************************************************************************************
\section{Motivation example}\label{sec:motivation}
%*********************************************************************************************************
\begin{figure}        
\centering  
\includegraphics[width=0.70\textwidth]{figs/e3value}
\caption{E3value model of the "To Publish Music" scenario} 
\label{fig:E3valuemodel}
\end{figure}

Consider for instance the following scenario. A programmer wants to develop the application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to Twitter and Facebook.  Social network users will have their status synchronized in  Twitter and Facebook  with the title of the music they are listening in Spotify.
For developing this application the programmer has to coordinate the calls to the methods offered by the following services:
\begin{itemize}
\item Music Services like  Spotify export methods for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item Facebook and Twitter services export methods for getting and updating the status of a given user: 
\begin{itemize} \item{\sf\small get-Status ( usedid ): String ; \item update-Status ( userid, new-status ): String}; \end{itemize}
%\item Mood Service exports a method for retrieving the mood of a  user: 
%\begin{itemize} \item{\sf\small get-Mood ( userid ): String} ;\end{itemize}
%It is a complex process that uses a log of the music listened by the user in Spotify during some period of time, and then according to the classification of each song it will determine her mood.
\end{itemize}

%

Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of 
the scenario. The model shows  Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publish in information about music being listened by users. The private application   interacts with Spotify  for obtaining free information about the flow of music being listened by a user. Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).

The "To Publish Music" scenario 
starts by contacting the music service Spotify for retrieving the user's  musical status. This information is sent to the mood service in order to calculate the mood of the user. Therefore, the Twitter and Facebook services are called for retrieving the status of the user.
%which is compared with other analysis data in order to determine the mood.  
Finally, Twitter and Facebook services are contacted in parallel for updating the user's status with the corresponding song.

Besides the service composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model the other constraints that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated. This paper proposes the $\pi$-SOD-M method \cite{decastro1,decastro2}  for building applications by   modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. This method is described in the following sections.

%*********************************************************************************************************
\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:sodm}
%*********************************************************************************************************
\begin{figure} [htpb]       
\centering  
\includegraphics[width=0.65\textwidth]{figs/SODM}
\caption{SOD-M development process} 
\label{fig:sodm}
\end{figure}
SOD-M (see Figure \ref{fig:servicecompositionmodel}) defines a service-oriented approach that  provides  a set of guidelines to build IS based exclusively on services  \cite{decastro1,decastro2}.  Therefore, SOD-M proposes to use services as first-class objects for the whole process of the IS development and it  follows a MDA-based approach.
SOD-M proposes a set of models, extending from the highest level of abstraction of the MDA:  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro3}. For the sake of simplicity, in this work we  focus on the service composition model of SOD-M (highlighted in Figure  \ref{fig:sodm}) that we extended for considering non-functional properties.

Our work  extends the  composition model of the PIM layer of the SOD-M method \cite{decastro1} with the notion of {\sf\small A-Policy} for representing service composition constraints \cite{Espinosa-Oviedo2011a}.  It  defines the $\pi$-{\sc Pews} \cite{Placido2010LTPD} meta-model providing guidelines for expressing the services composition and the policies, and also defines  Model to model transformation rules for generating  the implementation of reliable services' compositions implementations. Finally, our work defines model to text transformation rules for generating the  program that implements both the services' composition and the associated policies and that is executed by an adapted engine.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{$\pi$ service composition meta-model}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

The policy  service composition meta-model ($\pi$-SCM)
represents the workflow needed to implement a business service by identifying {\sc Activities} and entities collaborating in the business processes (called {\sc Business Collaborators}) and performing them. An activity identified in the meta-model describes a fundamental behaviour unit which represents some transformation or processing in the system being modeled. There are two types of activities in this model: i) a Web Service (stereotyped as {\sc WS}); and ii) a simple operation not supported by a Web Service, called an Activity Operation (stereotyped as {\sc AOP}). Business collaborators are displayed in this model as a partition in the activity diagram. 
 \begin{figure}[htpb]       
\centering  
\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
\caption{Policy based service composition meta-model ($\pi$-SCM)} 
\label{fig:e-scomposition-metamodel}
\end{figure}

Instead of programming different  protocols within the application logic, we propose to include the modeling of non-functional constraints at the early stages of the service composition engineering process.  We model non-functional constraints of services compositions using the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09b,CIC:eovszmc09c}, a kind of  pattern for specifying policy types. In order to represent constraints associated to services compositions, we  extended  the SOD-M service composition model with two concepts : {\sc Rule} and {\sc A-Policy} (see $\pi$-service composition model in Figure \ref{fig:e-scomposition-metamodel}).

A {\sc Rule} represents the moment in which a constraint that must be evaluated (Event - Action) and the {\sc Action} to be executed for reinforcing  it. A {\sc Rule} can be associated to  activity boxes, to the control flow, message passing and to the compartments of the $\pi$-SCM. An A-policy type describes a specific constraint types that can correspond to non-functional constraints like   transactional behaviour, security and adaptability. An A-policy  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.

An a-policy groups a set of rules that implement a non-functional constraint The a-policy expresses global variables and operations that can be shared by the rules and that can be used for expressing the condition and the condition parts.

\begin{figure}[htpb]         
\centering  
\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
\caption{Service Composition Model for the business service To publish music} 
\label{fig:servicecompositionmodel}
\end{figure}

Figure \ref{fig:servicecompositionmodel} shows the $\pi$-SCM  that represents the composition process of the different activities needed to carry out the business service “To Publish Music”. There are three external business collaborators (Spotify, Twitter and Facebook). It also shows business process  of the  "To Publish Music" application that consists of three service activities. Note that in the service activity {\em PublishMusic} the service activity calls the activities of two service collaborators namely {\em Facebook} and {\em Twitter}.

Given that Facebook and Twitter services require authentication protocols in order to execute methods that will read and update the users' space. When these services are called  the call must be part of the authentication protocol required by these services.  
In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sc Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}). In the same way, the  class {\sc Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}. 
 {\sf\small OAuth}  implements the open authentication protocol. 
As shown in Figure \ref{fig:servicecompositionmodel}, the a-policy  has a variable {\sf\small Token} that will be used to store the authentication token provided by the service. This variable type is imported through the library {\sf\small OAuth.Token}. The policy  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}). Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.

{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.  As shown in Figure  \ref{fig:servicecompositionmodel}, the policy imports an HTTP protocol library and it has two variables {\sf\small username} and {\sf\small password}.  The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. On the notification of an event of that type, a credential is obtained using the username and password values. The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.

Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code. For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks. Or for providing security in the communication channel when the payment service is called.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
The idea of the $\pi$-{\sc Pews} meta-model is based on the service composition approach provided by the language PEWS\cite{BaAM06} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
implement each operation of the service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model  
consisting of  classes representing :
\begin{itemize}
\item a service composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a service composition and {\sc PewsExpression} representing a services' composition.
An expression can be an{\sc Operation} or a {\sc Compound Operation}, 
denoted by an identifier, the sequential ($\ . \ $) or parallel ($\ \| \ $) composition of services, 
the choice ($\ + \ $) among services, 
the sequential ($*$) or parallel ($\{\dots\}$) repetition of operation or the conditional execution of an operation ($[C]S$).

\item policies that can be associated to a service composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
\end{itemize}

\begin{figure}        
\centering  
\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Metamodel} 
\label{fig:metamodel}
\end{figure}

As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation},  an {\sc Operator}, and a {\sc Path}.  It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.  Thus, an A-Policy represents a set of re-actions to be possibly executed if one or several triggering events of its rules are notified. 
\begin{itemize}
\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).

\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}. A policy has variables that represent the view of the execution state of its associated scope, that is required for executing the rules. The value of a variable is represented using the type Variable.
\end{itemize}
The class {\sc A-Policy} is specialized for defining specific constraints and  associating them a  to a concrete services’ composition.

%
\begin{figure}[htpb]        
\includegraphics[width=0.80\textwidth]{figs/modeloPEWS}
\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application} 
\label{fig:p-scim}
\end{figure}
%
Figure \ref{fig:p-scim} shows the generated $\pi$-{\sc Pews} model for our example generated by the Eclipse plug-in that we implemented for this purpose. 
An authentication A-Policy represents the situation where an invocation in
an activity occurs until its sender and/or its recipient have been
identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Transformation rules}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM used for representing the service composition into the elements of the $\pi$-{\sc Pews} (see details of transformation rules in the appendix). There are two groups of rules: those that transform service composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} elements; and those that transform rules grouped by policies into A-policy types.

\begin{figure}    
\includegraphics[width=0.95\textwidth]{figs/PI-SC-PI-P}
\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation} 
\label{fig:transformations}
\end{figure}

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\paragraph{First group of transformation rules:}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a class {\sc Operation} with a corresponding attribute name {\sc Operation:name}. A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc Operation} and {\sc Operation:name}. When more than one actions are called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} the transformations are (see Figure \ref{fig:mappingrules-2}) :
\begin{itemize}
\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
\item $op_1 \parallel op_2 . op_3$ if control nodes of type {\sc\em fork, join} are combined 
 \item $op_1 + op_2 . op_3$ if control nodes of type {\sc\em decision, merge} are combined 
\end{itemize}

In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}. Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form {\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}.

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\paragraph{Second group of transformation rules:}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
The a-policies defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model. The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.   The variables associated to a policy expressed in the$\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
As shown in Figure \ref{fig:mappingrules-1} of the Appendix, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}. The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type. As shown in Figure \ref{fig:mappingrules-2} of the Appendix, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}. The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.

%
In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into policies of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario. Thus in both cases the events are of type {\sf ActivityPrepared}. These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}. In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.

%*********************************************************************************************************
\section{Implementation issues}\label{sec:implementation}
%*********************************************************************************************************
\begin{figure}[htpb]
	\begin{center}
		\includegraphics[width=0.68\textwidth]{figs/Architecture.png}
	\end{center}
		\caption{General overview of the reliable service composition process}
   \label{fig:policymanager}
\end{figure}
This section  depicts the proposed process for generating reliable service compositions. For a given service based application, the process  consist in generating the  code starting from a $\pi$-SCM modeling an application. Note that the service composition model is not modeled from scratch at this level, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.

To achieve the whole model transformation process, the Eclipse Modelling Framework (EMF) has been used. The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model. From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.

In order to automate the transformation we specify the transformation rules using ATL model transformation language \cite{11}. An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.
Finally, in order to generate code we have used Acceleo [http://www.acceleo.org/pages/home/en].

Figure \ref{fig:policymanager} depicts a general overview of the proposed process showing the set of plug-ins which have been developed in order to support the proposed framework:
\begin{itemize}
\item 	First, we  implemented using EMF, the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}. After that, starting form these meta-models, we have developed the models plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} model ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins). 
\item	Then, we  developed using ATL, the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in). 
\item 	Finally, we  created the code generation plug-in using Acceleo. To this end, the pews.mt program was codified including the entire model to text transformation needed to generate a  code starting form a $\pi$-PEWS model; and then a chain execution was created in order to execute the model to text transformation. 
\end{itemize}

As  shown in Figure \ref{fig:policymanager} , once an instance of a PEWS code is obtained starting form a particular $\pi$-service composition model it can be executed over policy based services' composition execution environment  consisting of a composition engine and a policy manager.  The  policy manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions. The A-Policy Manager interacts with a composition engine thanks to a  message communication layer (MOM).

%Figure \ref{fig:policymanager} gives the global architecture of a policy based services' composition execution environment  consisting of a composition engine and a policy manager.
The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow. 
Each activity is seen as the process where the service method call is executed. 
The execution of an activity has four states: prepared, started, terminated, and failure. 
The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure. 
%

At execution time, the evaluation of policies done by the policy manager must be synchronized with the execution of the service composition (i.e., the execution of an activity or a control flow).  Policies associated to a scope are activated when the execution of its scope starts. A policy will have to be executed only if one or several of its rules is triggered. If several rules are triggered the policy manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section. Once rules have been executed, the policy finishes its execution and returns to a sleeping state.
If rules belonging to several policies are triggered then policies are also ordered according to an execution plan. The order of policies has implications on the global order of the rules to be executed.

We conducted an experimental validation of our approach  by implementing the "To Publish Music" application. Once the service composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose. Our implementation includes a Java code generator that generates executable code for a reliable service composition. 
The model represents an services' composition expression and its associated policies. Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process. 
\begin{figure}        
\includegraphics[width=0.85\textwidth]{figs/pews-expression}
\caption{Pews program implementing the "To Publish Music" application} 
\label{fig:pewsexpression}
\end{figure}
%*********************************************************************************************************
\section{Related works}\label{sec:related}
%*********************************************************************************************************
Current standards in services' composition implement functional, non-functional properties (NFP) and communication aspects by combining different languages and protocols. WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services. For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction. The selection of the adequate protocols for adding a specific NFP to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer. As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process. This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
Therefore, in contrast to these approaches, our work is inspired in the philosophy of \textit{separation of concerns} adopted in the construction of middleware for specifying policies to a services' composition in orthogonal way.

There are few methodologies and approaches that address the explicit modeling  of non functional
properties for service based applications.   Software process methodologies for
building  services based applications have been proposed in\cite{PapazoglouH06,Papazoglou03,cdl2006,MilanovicM06,FeuerlichtM05,Ramollari_asurvey,somet2005}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.

 \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed. A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called. Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02} .

 The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
description language, for defining contracts for services. There are an associated architecture framework, design standards and a methodology \cite{MilanovicM05,Milanovic05,Milanovic06,MilanovicM06}, for developing applications using services.  A services' based application  specification is generated after several 
B \cite{AbrialLNSS91} machines refinements that describe the services and their
compositions.

\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
oriented business process development methodology with phases for business process development. The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.

IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}. SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
specification, realization, implementation and deployment monitoring and management.

 \cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes. Activities in business process can be performed by services so that the model of business process
represents a composition of services. It classifies services   as public utilities, business services or
composition services. Software development that uses services involves creating programs for composing and configuring services to create new composite services. The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.

Unlike the works presented, the main contribution of our proposal is the methodology description together with model representations in three levels (CIM, PIM and PSM)  for the design and development of distributed applications that can be reused and that are reliable.

%*********************************************************************************************************
\section{Conclusions and future work}\label{sec:conclusions}
%*********************************************************************************************************
This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications. We model and associate policies to  service based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.  We extended the SOD-M methodology, particularly the $\pi$-SCM (service composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code. We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.

Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services. We are currently working on the definition of a methodology for explicitly expressing such properties in the early stages of the specification of services based applications. Having such business rules expressed and then translated and associated to the service composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.

Programming non-functional properties is not an easy task, so we are defining a set of predefined policy types with the associated use rules for guiding the programmer when she associates them to a concrete application. Policy type  that can also serve as patterns for programming or specializing the way non-functional properties are programmed.

\bibliographystyle{plain}
\bibliography{caise,S-Coord,Bib/e-commerce,Bib/trans,Bib/Component,Bib/XML,Bib/Workflow,Bib/SODM,Bib/G}

%*********************************************************************************************************
\section*{Appendix}\label{sec:appendix}
%*********************************************************************************************************
Figures \ref{fig:mappingrules-1},  \ref{fig:mappingrules-3}, \ref{fig:mappingrules-2} show the transformation rules between the $\pi$-SCM meta-model to the $\pi$-{\sc Pews}  meta-model. 
\begin{figure}        
\includegraphics[width=0.75\textwidth]{figs/Mapping-1}
\caption{Mapping rules between service composition elements} 
\label{fig:mappingrules-1}
\end{figure}

\begin{figure}        
\includegraphics[width=0.75\textwidth]{figs/Mapping-3}
\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} policy elements} 
\label{fig:mappingrules-3}
\end{figure}

\begin{figure}        
\includegraphics[width=0.75\textwidth]{figs/Mapping-2}
\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} rule elements} 
\label{fig:mappingrules-2}
\end{figure}

\end{document}

### 2013-10-31 CAISE2011/submitted/caise-etal.tex
CAISE2011 y algo mas

--- 2013-09-30 JournalToSubmit2013/index.tex
@@
 \usepackage{epstopdf}
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
-
-%% if you use PostScript figures in your article
-%% use the graphics package for simple commands
-%% \usepackage{graphics}
-%% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers after \end{frontmatter}.
-%% \usepackage{lineno}
-
-%% natbib.sty is loaded by default. However, natbib options can be
-%% provided with \biboptions{...} command. Following options are
-%% valid:
-
-%%   round  -  round parentheses are used (default)
-%%   square -  square brackets are used   [option]
-%%   curly  -  curly braces are used      {option}
-%%   angle  -  angle brackets are used    <option>
-%%   semicolon  -  multiple citations separated by semi-colon
-%%   colon  - same as semicolon, an earlier confusion
-%%   comma  -  separated by comma
-%%   numbers-  selects numerical citations
-%%   super  -  numerical citations as superscripts
-%%   sort   -  sorts multiple citations according to order in ref. list
-%%   sort&compress   -  like sort, but also compresses numerical citations
-%%   compress - compresses without sorting
-%%
-%% \biboptions{comma,round}
-
-% \biboptions{}
-
-\journal{Journal of Systems and Software}
-
 \begin{document}
 
-\begin{frontmatter}
+\title{Generating reliable services' composition  using A-policies: a model-driven approach\thanks{This work is partially financed by the projects {\sc Orchestra} ECOS-ANUIES, {\sc Clever} STIC-AMSUD, and {\sc Masai}.}}
 
-%% Title, authors and addresses
+\author{Genoveva Vargas-Solar\inst{1}
+\and Valeria de Castro\inst{4}
+\and Pl\'{a}cido Antonio de Souza Neto\thanks{Financed by CAPES/STIC-AMSUD Brasil, BEX 4112/11-3} \inst{5,6}
+\and Javier A. Espinosa-Oviedo\inst{3}
+\and Esperanza Marcos\inst{4}
+\and Martin A. Musicante\inst{5}
+\and Jos\'{e}-Luis Zechinelli-Martini\inst{2,1}
+\and Christine Collet \inst{3}}
 
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
+\institute{
+	French Council of Scientific Research, LIG-LAFMIA	\\
+	681 Rue de la Passerelle, BP. 72, 38402	Saint Martin d'H\`{e}res Cedex, France		\\
+		\email{Genoveva.Vargas-Solar@imag.fr}
+\and
+	Fundaci\'{o}n Universidad de las Am\'{e}ricas, Puebla	\\
+	Exhacienda Sta. Catarina M\'{a}rtir s/n 72820, San Andr\'{e}s Cholula, Puebla, M\'{e}xico			\\
+	\email{joseluis.zechinelli@udlap.mx}
+\and
+	Grenoble Institute of Technology	\\
+	46 Avenue F'{e}lix Viallet, 38031 Grenoble Cedex 1		\\
+	\email{Christine.Collet@grenoble-inp.fr}, \email{Javier.Espinosa@imag.fr} \\
+\and
+	Universidad Rey Juan Carlos	\\
+	Av Tulip\'{a}n, M\'{o}stoles, Spain 		\\
+	\email{Valeria.deCastro@urjc.es, esperanza.marcos@urjc.es} \\
+\and
+	DIMAp - UFRN, ForAll - Formal Methods and Language Research Laboratory	\\
+	Campus Universitário - Lagoa Nova, Natal - RN, Brasil		\\
+	\email{mam@dimap.ufrn.br} \\
+	\and
+	Instituto Federal do Rio Grande do Norte	\\
+	Av. Senador Salgador Filho, 1559 - Tirol, Natal - RN, Brasil		\\
+	\email{placido.neto@ifrn.edu.br}
+}
 
-\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
-the Presence of Non-Functional Properties}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[inst1]{Pl\'acido A. Souza Neto}
-\author[inst2]{Genoveva Vargas-Solar}
-\author[inst3]{Martin A. Musicante}
-\author[inst4]{Valeria~de~Castro}
-\author[inst3]{Umberto Costa}
-
-\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}
-
-\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
-Brazil}
-
-\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}
+\maketitle
 
 \begin{abstract}
-This paper presents\ldots
+This paper presents an approach for modeling and associating {\em A-Policies} to  services' based applications.
+It proposes to extend the SOD-M model driven method with   (i) the $\pi$-{\sc SCM} an {\em A-Policy} services' composition meta-model for representing non-functional constraints associated to services' based applications;  (ii) the $\pi$-{\sc Pews} meta-model providing guidelines for expressing the composition and the policies; and, (iii)  model to model and model to text transformation rules for semi-automatizing the implementation of reliable services' compositions.
+As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
 \end{abstract}
 
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
-
-%% main text
-
+%*********************************************************************************************************
+\section{Introduction}
+%*********************************************************************************************************
 Service oriented computing is at the origin of an evolution in the field of software development. 

### 2013-10-31 CAISE2011/submitted/support/CaiSE/src/caise.tex
CAISE2011 y algo mas

--- 2013-10-31 CAISE2011/submitted/caise-etal-old.tex
@@
-\documentclass{llncs}
-\usepackage{geometry}                
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
+%%This is a very basic article template.
+%%There is just one section and two subsections.
+\documentclass{article}
+
 \begin{document}
 
-\title{Generating reliable service composition  using policies: a model-driven approach\thanks{This work is partially financed by the projects {\sc Orchestra} ECOS-ANUIES, {\sc e-CLOUDSS} Microsoft,  {\sc Clever} STIC-AMSUD, and {\sc Masai}.}}
+\section{Related Works}
 
-\author{Genoveva Vargas-Solar\inst{1} 
-\and Valeria de Castro\inst{4}
-\and Pl\'{a}cido A. de Souza Neto\thanks{Financed by CAPES/SticAmbud Brasil, BEX 4112/11-3} \inst{5,6} 
-\and Javier A. Espinosa-Oviedo\inst{3} 
-\and Esperanza Marcos\inst{4} 
-\and Mart\'{i}n A. Musicante\inst{5} 
-\and Jos\'{e}-Luis Zechinelli-Martini\inst{2,1} 
-\and Christine Collet \inst{3}}
+%Some works related with service-oriented
+%development have presented that the impact of the service-oriented
+%computing paradigm, the quality garantees on software development and the
+%way in which systems can be constructed have been growing up, however
 
-\institute{
-	French Council of Scientific Research, LIG-LAFMIA	\\
-	681 Rue de la Passerelle, BP. 72, 38402	Saint Martin d'H\`{e}res Cedex, France		\\
-		\email{Genoveva.Vargas-Solar@imag.fr}
-\and
-	Fundaci\'{o}n Universidad de las Am\'{e}ricas Puebla	\\
-	Exhacienda Sta. Catarina M\'{a}rtir s/n 72820, San Andr\'{e}s Cholula, Puebla, M\'{e}xico			\\
-	\email{joseluis.zechinelli@udlap.mx}
-\and
-	Grenoble Institute of Technology	\\
-	46 Avenue F'{e}lix Viallet, 38031 Grenoble Cedex 1		\\
-	\email{Christine.Collet@grenoble-inp.fr}, \email{Javier.Espinosa@imag.fr} \\
-\and
-	Universidad Rey Juan Carlos	\\
-	Av Tulip\'{a}n, M\'{o}stoles, Spain 		\\
-	\email{Valeria.deCastro@urjc.es, esperanza.marcos@urjc.es} \\
-\and
-	DIMAp - UFRN, ForAll - Formal Methods and Language Research Laboratory	\\
-	Campus Universitário - Lagoa Nova, Natal - RN, Brasil		\\
-	\email{mam@dimap.ufrn.br} \\
-	\and
-	Instituto Federal do Rio Grande do Norte	\\
-	Av. Senador Salgador Filho, 1559 - Tirol, Natal - RN, Brasil		\\
-	\email{placido.neto@ifrn.edu.br}
-}
-
-\maketitle
-
-\begin{abstract}
-This paper presents an approach for modeling and associating policies to  service based applications for representing both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available, recovery, persistence) and  constraints stemming from the services used for implementing them (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work proposes a model driven  methodology that  (i) extends     PIM layer's composition model of the SOD-M method \cite{decastro1} with the notion of {\sf\small A-Policy} \cite{Espinosa-Oviedo2011a} for representing constraints associated to service based applications and, (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the policies (iii) defines   Model to model transformation rules for generating  the implementation of reliable services' compositions.
-% (iv) defines model to text transformation rules for generating the  program that implements both the services' composition and the associated policies and that is executed by an adapted engine.
-
-\end{abstract}
-
-%*********************************************************************************************************
-\section{Introduction}
-%*********************************************************************************************************
-%Along with the emergence of Web 2.0 and services based computing,  the Web has become a software as services development platform for applications.
-Lambda users \footnote{Lambda user is a user with few computing knowledge.} have thereby access to different services some exported by their favorite social networks, music and leisure providers.
-As in other platforms, the applications' integration problem emerges again  and calls for simple  solutions that can give users' a global view of their Web services within a whole, unique and, dynamic space.
-
-%For example, having an integrated view of the mood of a user according to the status of her favorite social networks.
-The integration of data services  can be done today by coordinating services accessible on the  network.
-For example, synchronizing music posts of a user accounts in Twitter and Facebook ({\em status: "Desert Rose - Sting"}) according to the music listened in Spotify    and observe and use this information to determine the degree of music preference convergence of a community of users.
-Yet, such services produce and update some of these data dynamically (e.g.
-the status in Facebook when the user listens music) and integrated views must be continuously and automatically updated according to specific  requirements: data freshness, periodic and atomic update of data shared among several services despite eventual exceptions.
-
-Service-Oriented Computing has provoked an evolution in the field of software development.
-The literature stresses the need for methodologies and techniques for service-oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful service-oriented applications \cite{papazoglou}.
-Service-oriented development methodologies providing models, best practices, and reference architectures to built service-based applications mainly address  functional aspects of service based applications \cite{1,2,decastro1,4}.
-Non-functional aspects concerning services and application "semantics", often expressed as requirements, and constraints in general purpose methodologies are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to service based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  service based applications  early during the service composition model phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1}  with the notion of A-policy.
-SOD-M defines a model-driven method for  developing  service-oriented information systems.
-It defines a process  starting with the  identification of business services through business modeling, and, by means of models transformations it allows to obtain a service composition model \cite{decastro1}  extended with policies.
-Policies are used to express constraints which can be applied to all the service composition process or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services used for implementing them (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-
-%Figure \ref{fig:framework} shows the general overview of the approach we propose for specifying and designing reliable service based applications. 
-Our work proposes a model driven  methodology that  (i) extends     PIM layer's composition model of the SOD-M method \cite{decastro1} with the notion of {\sf\small A-Policy} \cite{Espinosa-Oviedo2011a} for representing constraints associated to service based applications and, (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the policies (iii) defines   Model to model transformation rules for generating  the implementation of reliable services' compositions.
-
-%\begin{figure}[htpb]
-%
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} describes a motivation example that integrates and synchronizes well-known social networks services namely FaceBook, Twitter and, Spotify.
-Section \ref{sec:sodm} describes $\pi$-SOD-M method that enables the representation and association of policies to services composition and thereby making them reliable.
-Section \ref{sec:implementation} describes implementation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, service composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Motivation example}\label{sec:motivation}
-%*********************************************************************************************************
-\begin{figure}        
-\centering  
-\includegraphics[width=0.70\textwidth]{figs/e3value}
-\caption{E3value model of the "To Publish Music" scenario} 
-\label{fig:E3valuemodel}
-\end{figure}
-
-Consider for instance the following scenario.
-A programmer wants to develop the application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to Twitter and Facebook.
-Social network users will have their status synchronized in  Twitter and Facebook  with the title of the music they are listening in Spotify.
-For developing this application the programmer has to coordinate the calls to the methods offered by the following services:
-\begin{itemize}
-\item Music Services like  Spotify export methods for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export methods for getting and updating the status of a given user: 
-\begin{itemize} \item{\sf\small get-Status ( usedid ): String ; \item update-Status ( userid, new-status ): String}; \end{itemize}
-%\item Mood Service exports a method for retrieving the mood of a  user: 
-%\begin{itemize} \item{\sf\small get-Mood ( userid ): String} ;\end{itemize}
-%It is a complex process that uses a log of the music listened by the user in Spotify during some period of time, and then according to the classification of each song it will determine her mood.
-\end{itemize}
-
-%
-
-Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of 
-the scenario.
-The model shows  Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publish in information about music being listened by users.
-The private application   interacts with Spotify  for obtaining free information about the flow of music being listened by a user.
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-The "To Publish Music" scenario 
-starts by contacting the music service Spotify for retrieving the user's  musical status.
-This information is sent to the mood service in order to calculate the mood of the user.
-Therefore, the Twitter and Facebook services are called for retrieving the status of the user.
-%which is compared with other analysis data in order to determine the mood.  
-Finally, Twitter and Facebook services are contacted in parallel for updating the user's status with the corresponding song.
-
-Besides the service composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model the other constraints that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-This paper proposes the $\pi$-SOD-M method \cite{decastro1,decastro2}  for building applications by   modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition.
-This method is described in the following sections.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:sodm}
-%*********************************************************************************************************
-\begin{figure} [htpb]       
-\centering  
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process} 
-\label{fig:sodm}
-\end{figure}
-SOD-M (see Figure \ref{fig:servicecompositionmodel}) defines a service-oriented approach that  provides  a set of guidelines to build IS based exclusively on services  \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the IS development and it  follows a MDA-based approach.
-SOD-M proposes a set of models, extending from the highest level of abstraction of the MDA:  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views.
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro3}.
-For the sake of simplicity, in this work we  focus on the service composition model of SOD-M (highlighted in Figure  \ref{fig:sodm}) that we extended for considering non-functional properties.
-
-Our work  extends the  composition model of the PIM layer of the SOD-M method \cite{decastro1} with the notion of {\sf\small A-Policy} for representing service composition constraints \cite{Espinosa-Oviedo2011a}.
-It  defines the $\pi$-{\sc Pews} \cite{Placido2010LTPD} meta-model providing guidelines for expressing the services composition and the policies, and also defines  Model to model transformation rules for generating  the implementation of reliable services' compositions implementations.
-Finally, our work defines model to text transformation rules for generating the  program that implements both the services' composition and the associated policies and that is executed by an adapted engine.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$ service composition meta-model}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-The policy  service composition meta-model ($\pi$-SCM)
-represents the workflow needed to implement a business service by identifying {\sc Activities} and entities collaborating in the business processes (called {\sc Business Collaborators}) and performing them.
-An activity identified in the meta-model describes a fundamental behaviour unit which represents some transformation or processing in the system being modeled.
-There are two types of activities in this model: i) a Web Service (stereotyped as {\sc WS}); and ii) a simple operation not supported by a Web Service, called an Activity Operation (stereotyped as {\sc AOP}).
-Business collaborators are displayed in this model as a partition in the activity diagram.
- \begin{figure}[htpb]       
-\centering  
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{Policy based service composition meta-model ($\pi$-SCM)} 
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-Instead of programming different  protocols within the application logic, we propose to include the modeling of non-functional constraints at the early stages of the service composition engineering process.
-We model non-functional constraints of services compositions using the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09b,CIC:eovszmc09c}, a kind of  pattern for specifying policy types.
-In order to represent constraints associated to services compositions, we  extended  the SOD-M service composition model with two concepts : {\sc Rule} and {\sc A-Policy} (see $\pi$-service composition model in Figure \ref{fig:e-scomposition-metamodel}).
-
-A {\sc Rule} represents the moment in which a constraint that must be evaluated (Event - Action) and the {\sc Action} to be executed for reinforcing  it.
-A {\sc Rule} can be associated to  activity boxes, to the control flow, message passing and to the compartments of the $\pi$-SCM.
-An A-policy type describes a specific constraint types that can correspond to non-functional constraints like   transactional behaviour, security and adaptability.
-An A-policy  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-
-An a-policy groups a set of rules that implement a non-functional constraint The a-policy expresses global variables and operations that can be shared by the rules and that can be used for expressing the condition and the condition parts.
-
-\begin{figure}[htpb]         
-\centering  
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Service Composition Model for the business service To publish music} 
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Figure \ref{fig:servicecompositionmodel} shows the $\pi$-SCM  that represents the composition process of the different activities needed to carry out the business service “To Publish Music”.
-There are three external business collaborators (Spotify, Twitter and Facebook).
-It also shows business process  of the  "To Publish Music" application that consists of three service activities.
-Note that in the service activity {\em PublishMusic} the service activity calls the activities of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Given that Facebook and Twitter services require authentication protocols in order to execute methods that will read and update the users' space.
-When these services are called  the call must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sc Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sc Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol. 
-As shown in Figure \ref{fig:servicecompositionmodel}, the a-policy  has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The policy  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the policy imports an HTTP protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks.
-Or for providing security in the communication channel when the payment service is called.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the service composition approach provided by the language PEWS\cite{BaAM06} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of the service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model  
-consisting of  classes representing :
-\begin{itemize}
-\item a service composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a service composition and {\sc PewsExpression} representing a services' composition.
-An expression can be an{\sc Operation} or a {\sc Compound Operation}, 
-denoted by an identifier, the sequential ($\ . \ $) or parallel ($\ \| \ $) composition of services, 
-the choice ($\ + \ $) among services, 
-the sequential ($*$) or parallel ($\{\dots\}$) repetition of operation or the conditional execution of an operation ($[C]S$).
-
-\item policies that can be associated to a service composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-
-\begin{figure}        
-\centering  
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel} 
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation},  an {\sc Operator}, and a {\sc Path}.
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an A-Policy represents a set of re-actions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A policy has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type Variable.
-\end{itemize}
-The class {\sc A-Policy} is specialized for defining specific constraints and  associating them a  to a concrete services’ composition.
-
-%
-\begin{figure}[htpb]        
-\includegraphics[width=0.80\textwidth]{figs/modeloPEWS}
-\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application} 
-\label{fig:p-scim}
-\end{figure}
-%
-Figure \ref{fig:p-scim} shows the generated $\pi$-{\sc Pews} model for our example generated by the Eclipse plug-in that we implemented for this purpose. 
-An authentication A-Policy represents the situation where an invocation in
-an activity occurs until its sender and/or its recipient have been
-identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Transformation rules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM used for representing the service composition into the elements of the $\pi$-{\sc Pews} (see details of transformation rules in the appendix).
-There are two groups of rules: those that transform service composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} elements; and those that transform rules grouped by policies into A-policy types.
-
-\begin{figure}    
-\includegraphics[width=0.95\textwidth]{figs/PI-SC-PI-P}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation} 
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\paragraph{First group of transformation rules:}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc Operation} and {\sc Operation:name}.
-When more than one actions are called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} the transformations are (see Figure \ref{fig:mappingrules-2}) :
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item $op_1 \parallel op_2 . op_3$ if control nodes of type {\sc\em fork, join} are combined 
- \item $op_1 + op_2 . op_3$ if control nodes of type {\sc\em decision, merge} are combined 
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form {\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\paragraph{Second group of transformation rules:}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The a-policies defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to a policy expressed in the$\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-As shown in Figure \ref{fig:mappingrules-1} of the Appendix, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-As shown in Figure \ref{fig:mappingrules-2} of the Appendix, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into policies of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
-\begin{figure}[htpb]
-	\begin{center}
-		\includegraphics[width=0.68\textwidth]{figs/Architecture.png}
-	\end{center}
-		\caption{General overview of the reliable service composition process}
-   \label{fig:policymanager}
-\end{figure}
-This section  depicts the proposed process for generating reliable service compositions.
-For a given service based application, the process  consist in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the service composition model is not modeled from scratch at this level, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
-
-To achieve the whole model transformation process, the Eclipse Modelling Framework (EMF) has been used.
-The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-
-In order to automate the transformation we specify the transformation rules using ATL model transformation language \cite{11}.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.
-Finally, in order to generate code we have used Acceleo [http://www.acceleo.org/pages/home/en].
-
-Figure \ref{fig:policymanager} depicts a general overview of the proposed process showing the set of plug-ins which have been developed in order to support the proposed framework:
-\begin{itemize}
-\item 	First, we  implemented using EMF, the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-After that, starting form these meta-models, we have developed the models plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} model ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-\item	Then, we  developed using ATL, the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in). 
-\item 	Finally, we  created the code generation plug-in using Acceleo.
-To this end, the pews.mt program was codified including the entire model to text transformation needed to generate a  code starting form a $\pi$-PEWS model; and then a chain execution was created in order to execute the model to text transformation.
-\end{itemize}
-
-As  shown in Figure \ref{fig:policymanager} , once an instance of a PEWS code is obtained starting form a particular $\pi$-service composition model it can be executed over policy based services' composition execution environment  consisting of a composition engine and a policy manager.
-The  policy manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The A-Policy Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-
-%Figure \ref{fig:policymanager} gives the global architecture of a policy based services' composition execution environment  consisting of a composition engine and a policy manager.
-The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow. 
-Each activity is seen as the process where the service method call is executed. 
-The execution of an activity has four states: prepared, started, terminated, and failure. 
-The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure. 
-%
-
-At execution time, the evaluation of policies done by the policy manager must be synchronized with the execution of the service composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A policy will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the policy manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-Once rules have been executed, the policy finishes its execution and returns to a sleeping state.
-If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The order of policies has implications on the global order of the rules to be executed.
-
-We conducted an experimental validation of our approach  by implementing the "To Publish Music" application.
-Once the service composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose.
-Our implementation includes a Java code generator that generates executable code for a reliable service composition.
-The model represents an services' composition expression and its associated policies.
-Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-\begin{figure}        
-\includegraphics[width=0.85\textwidth]{figs/pews-expression}
-\caption{Pews program implementing the "To Publish Music" application} 
-\label{fig:pewsexpression}
-\end{figure}
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
-Current standards in services' composition implement functional, non-functional properties (NFP) and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific NFP to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Therefore, in contrast to these approaches, our work is inspired in the philosophy of \textit{separation of concerns} adopted in the construction of middleware for specifying policies to a services' composition in orthogonal way.
-
-There are few methodologies and approaches that address the explicit modeling  of non functional
+There are
+ few methodologies and approaches that address the explicit modeling  of non functional
 properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,Papazoglou03,cdl2006,MilanovicM06,FeuerlichtM05,Ramollari_asurvey,somet2005}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
+building  services based applications have been proposed in \cite{PapazoglouH06,Papazoglou03,cdl2006,MilanovicM06,FeuerlichtM05,Ramollari_asurvey,somet2005}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
 
@@
  The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology \cite{MilanovicM05,Milanovic05,Milanovic06,MilanovicM06}, for developing applications using services.
+description language, for defining contracts for services. There are an associated architecture framework, design standards and a
+methodology \cite{MilanovicM05,Milanovic05,Milanovic06,MilanovicM06}, for developing applications using services. 
 A services' based application  specification is generated after several 
@@
 
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
+\cite{PapazoglouH06} proposes a methodology 
+based on a SOA extension. This work defines a service
+oriented business process development methodology with phases for business process
+development. The whole life-cycle is based on six phases: planning, analysis and design,
+construction and testing, provisioning, deployment, and execution
+and monitoring.
 
-IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-specification, realization, implementation and deployment monitoring and management.
+IBM proposes a methodology for the development of SOA
+solutions, called SOMA \cite{soma}. SOMA defines a life-cycle with seven
+phases: business modelling and transformation, solution management, identification,
+specification, realization, implementation and deployment monitoring
+and management.
 
- \cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-represents a composition of services. It classifies services   as public utilities, business services or
-composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
+Sommerville \cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information
+exchanged in a business processes. Activities in business process
+can be performed by services so that the model of business process
+represents a composition of services. It classifies services as  as public utilities, business services or
+coordination services. Software development that uses services involves creating
+programs for composing and configuring services to create new
+composite services. The service engineering process involves identifying services
+candidates, service interface and
+implementation definition, testing and deployment of each service.
 
-Unlike the works presented, the main contribution of our proposal is the methodology description together with model representations in three levels (CIM, PIM and PSM)  for the design and development of distributed applications that can be reused and that are reliable.
+Unlike the works presented, the main contribution of our proposal is the
+methodology description together with model representations in three levels (CIM, PIM and PSM) 
+ for the design and development of distributed applications that can be reused
+ and that are reliable.
 
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
-We model and associate policies to  service based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
-We extended the SOD-M methodology, particularly the $\pi$-SCM (service composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code.
-We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.
+\section{Statements}  
+\label{sec:challenges}
 
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
-We are currently working on the definition of a methodology for explicitly expressing such properties in the early stages of the specification of services based applications.
-Having such business rules expressed and then translated and associated to the service composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.
+The research questions that guide our research are
+intimately related with the spirit of classic software engineering, specifically
+the definition of a method for non functional properties-based distributed
+applications development. We also have in mind the application of the concept related with
+model driven development. There is a need to develop a proper model for these
+emerging technologies to reduce developing costs and to produce flexible and adaptable
+services based on quality properties. Therefore, we must ask the following:
 
-Programming non-functional properties is not an easy task, so we are defining a set of predefined policy types with the associated use rules for guiding the programmer when she associates them to a concrete application.
-Policy type  that can also serve as patterns for programming or specializing the way non-functional properties are programmed.
+\begin{enumerate}
+  \item What the elements that make the development of web
+  service applications different from traditional software development?
+  \item Is it possible to define a methodology for web service application
+  development that is based on non functional properties requirements?
+  \item Is it possible to define a language to specify composition constraints
+  and contract restriction for web service?
+  \item How to verify web services contract restrictions at runtime?
+  \item It is possible to summarize all the development through models?
+\end{enumerate}
 
+Therefore, we will defend the following concepts:
+
+``\textit{Developing applications that use compositions of services from a
+methodology for this purpose and which is based on non functional properties
+assurance requirements, can provide a better development and outcome. Allied to this, a
+method based on MDD / MDA can provide a better reuse of applications and models.
+Finally, a specification language that expresses web service composition and its
+constraints can help further the development.}''
+
+\bibliography{caise}  
 \bibliographystyle{plain}
-\bibliography{caise,S-Coord,Bib/e-commerce,Bib/trans,Bib/Component,Bib/XML,Bib/Workflow,Bib/SODM,Bib/G}
-
-%*********************************************************************************************************
-\section*{Appendix}\label{sec:appendix}
-%*********************************************************************************************************
-Figures \ref{fig:mappingrules-1},  \ref{fig:mappingrules-3}, \ref{fig:mappingrules-2} show the transformation rules between the $\pi$-SCM meta-model to the $\pi$-{\sc Pews}  meta-model. 
-\begin{figure}        
-\includegraphics[width=0.75\textwidth]{figs/Mapping-1}
-\caption{Mapping rules between service composition elements} 
-\label{fig:mappingrules-1}
-\end{figure}
-
-\begin{figure}        
-\includegraphics[width=0.75\textwidth]{figs/Mapping-3}
-\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} policy elements} 
-\label{fig:mappingrules-3}
-\end{figure}
-
-\begin{figure}        
-\includegraphics[width=0.75\textwidth]{figs/Mapping-2}
-\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} rule elements} 
-\label{fig:mappingrules-2}
-\end{figure}
 

### 2013-10-31 CAISE2011/submitted/Bib/Services/images.pl

@@ -1,0 +1,12 @@
+# LaTeX2HTML 99.2beta6 (1.42)
+# Associate images original text with physical files.
+
+
+$key = q/mbox{OSAK}^*;MSF=1.6;AAT/;
+$cached_env_img{$key} = q|<IMG
+ WIDTH="57" HEIGHT="16" ALIGN="BOTTOM" BORDER="0"
+ SRC="|."$dir".q|img1.gif"
+ ALT="$\mbox{OSAK}^*$">|; 
+
+1;
+

### 2013-10-31 CAISE2011/submitted/Bib/WF/images.pl

@@ -1,0 +1,18 @@
+# LaTeX2HTML 99.2beta6 (1.42)
+# Associate images original text with physical files.
+
+
+$key = q/^{eme};MSF=1.6;AAT/;
+$cached_env_img{$key} = q|<IMG
+ WIDTH="28" HEIGHT="15" ALIGN="BOTTOM" BORDER="0"
+ SRC="|."$dir".q|img2.gif"
+ ALT="$^{eme}$">|; 
+
+$key = q/nomath_inline}oenomath_inline};MSF=1.6;AAT/;
+$cached_env_img{$key} = q|<IMG
+ WIDTH="13" HEIGHT="9" ALIGN="BOTTOM" BORDER="0"
+ SRC="|."$dir".q|img1.gif"
+ ALT="\oe">|; 
+
+1;
+

### 2013-11-01 JournalToSubmit2013/intro.tex

@@
 Section\dots
+
+\begin{center}
+\textsc{\underline{Version from CAISE submission}}
+\end{center}
+
+Service oriented computing is at the origin of an evolution in the field of software development. 
+An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
+%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
+Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
+The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
+In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
+
+Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
+Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
+This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
+
+The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
+Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
+%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
+
+Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
+%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
+They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
+Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
+Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
+As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
+
+The remainder of the paper is organized as follows.
+Section \ref{sec:motivation} gives an overview of our approach.
+It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
+Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
+%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
+%
+Section \ref{sec:implementation} describes implementation and validation issues.
+Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
+Section \ref{sec:conclusions} concludes the paper and discusses future work.

### 2013-11-04 JournalToSubmit2013/intro.tex
intro... first version

--- 2013-10-11 JournalToSubmit2013/intro.tex
@@
-Model Driven Development (MDD) is a top-down approach for the development of software systems. 
-The main ideas of MDD were originally proposed by the Object Management Group (OMG)~\cite{mda}, as a set of guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
+Service oriented computing~\cite{Papazoglou2007} is at the origin of an evolution in the field of software development.
+Service oriented methods advocates for the construction of software systems formed by the composition of heterogeneous, loosely coupled modules.
+These modules (or services) communicate in order to achieve a common purpose.
 
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications. 
-This viewpoint represents the software system at its highest level of abstraction. 
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} within and between levels.
-Intra-level transformations are used to provide a unified representation of concepts of a given level.
-Inter-level transformation implement a refinement process between levels.
-Transformations may be automatic or semi-automatic.
-
-MDD techniques has been successfully used for the development of hardware and software systems~\cite{MDDvariosAqui}. 
-In particular, we are interested in the application of MDD to the design and implementation of web service applications.
-
-Service oriented computing~\cite{Papazoglou2007} is at the origin of an evolution in the field of software development~\cite{??}. 
-An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually developed.
+An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually being developed.
 (Moreover, IT systems need to evolve according to the business needs.)
-Thus, organizations are seeking for mechanisms to bridge the gap between the systems developed and business needs~\cite{bell}. 
+Thus, organizations are seeking for mechanisms to bridge the gap between the actually developed systems and their business needs~\cite{bell}. 
 The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone in the development of meaningful service based applications~\cite{5}.
-In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
 
@@
 
-Non-functional aspects of the services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the beginning of the (service) software process.
+Non-functional aspects of the services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
 Most methods consider them only after the application has been implemented, in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). 
@@
 the application context.
+
+\bigskip
+
+Model Driven Development (MDD) is a top-down approach for the development of software systems. 
+The main ideas of MDD were originally proposed by the Object Management Group (OMG)~\cite{mda}, as a set of guidelines for the structuring of specifications.
+The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}). 
+These models are successively refined from abstract specifications into actual computer programs.
+
+In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
 

### 2013-11-04 JournalToSubmit2013/methodologiesForNFR.tex
intro... first version

@@
 %(aqui propuestas relacionadas, analizar sus caracteristicas, si son MDD, si parte de alto nivel, etc.)
+
+Model Driven Development (MDD)~\cite{mda} is a top-down approach for the development of software systems. 
+MDD uses the notion of \textit{model} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
+
+\begin{trivlist}
+\item \textbf{Computation Independent Models (CIM):} This level focusses on the
+environment of the system, as well as on its business and requirement specifications. 
+This viewpoint represents the software system at its highest level of abstraction. 
+At this moment of the development, the structure and system processing details are still unknown or undetermined.
+
+\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform. 
+The specification defines those parts of the system that do not change from one platform to another.
+
+\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
+Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
+\end{trivlist}
+
+Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} within and between levels.
+Intra-level transformations are used to provide a unified representation of concepts of a given level.
+Inter-level transformation implement a refinement process between levels.
+Transformations may be automatic or semi-automatic.

### 2013-11-06 JournalToSubmit2013/SOD-M.tex
Changes: -Complete RelatedWorks section -Changes on Introdution (improving descriptions of MDD and MDA) -Add a section of background (including the description of concepts of MDA and SOD-M) -New references. -Minor comments on section methodologies, modelingNFR and modelingWithPISODM

--- 2013-10-07 JournalToSubmit2013/1-intro.tex
@@
-Model Driven Development (MDD) is a top-down approach for the development of software systems. 
-The main ideas of MDD were originally proposed by the Object Management Group (OMG)~\cite{mda}, as a set of guidelines for the structuring of specifications.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
+The methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements presented in this work extends the \textit{Service Oriented Development Method} (SOD-M) proposed by De Castro et al.\cite{decastro1}.
+
+SOD-M define a model-driven approach for development service-based applications.
+The model-driven method provided by SOD-M is based on the Model Driven Architecture (MDA) proposed by the OMG \cite{miller}.
+MDA defines a top-down approach for the development of software systems that uses the notion of \textit{model} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
 
@@
 
-Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} within and between levels.
+Besides the notion of model at each level of abstraction, MDA propose the use of \textit{model transformations} within and between levels.
 Intra-level transformations are used to provide a unified representation of concepts of a given level.
@@
 
-MDD techniques has been successfully used for the development of hardware and software systems~\cite{MDDvariosAqui}. 
-In particular, we are interested in the application of MDD to the design and implementation of web service applications.
+According to MDA guidelines, SOD-M meta-models are organized into three levels: CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
 
-Service oriented~\cite{Papazoglou2007} computing is at the origin of an evolution in the field of software development~\cite{??}. 
-An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually developed.
-(Moreover, IT systems need to evolve according to the business needs.)
-Thus, organizations are seeking for mechanisms to bridge the gap between the systems developed and business needs~\cite{bell}. 
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications~\cite{5}.
-In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
+Two models are defined at the CIM level: \textit{value model}
+and \textit{BPMN model}.
+\textsc{Explain these two here --Martin.}
 
-In Service-Oriented Computing, pre-existing services are
-combined to produce applications and provide the business logic. 
-The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}. 
-(Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.)
-Functional properties are not the only crucial aspect in the software development process. 
-Other properties need to be addressed to fit in the application with its context.
-These other aspects are called Non-Functional Properties.
-
-Non-functional aspects of the services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the beginning of the (service) software process.
-Most methods consider them only after the application has been implemented, in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). 
-This leads to service based applications that are partly specified and, thereby, partly compliant with the requirements of the application.
-Ideally, non-functional requirements should be considered along with all the stages of the software development. 
-The adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-the application context.
-
-In this work, we are interested in the extension of the \textit{Service Oriented Development Method} (SOD-M)~\cite{decastro1}, to support non-functional aspects, from the early stages of software development.
-SOD-M is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications.
-SOD-M does not provide support for the specification of non-functional requirements, such as
-security, reliability, and efficiency.
-
-The main goals of our work are:
+Models at the PIM level describe the structure of the application flow,
+while, the PSM level provides transformations towards more specific platforms.
+The PIM-level models proposed by SOD-M are:
 \begin{trivlist}
-\item \textit{(i)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\item \textit{(ii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\item \textit{(iii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
+\item \textit{use case}:
+\item \textit{extended use case}:
+\item \textit{service process}:
+\item \textit{service composition}.
 \end{trivlist}
 
-The rest of the paper is organized as follows. 
-Section\dots
+SOD-M defines three meta-models at the PSM level: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}.
+These three levels have no support for describing non-functional requirements.
 
-\COMMENT{
-\subsection*{The rest of this section is just for draft\dots}
+The SOD-M approach includes transformations between models:
+\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
+an abstract model at the CIM level, it is possible to apply transformations for
+generating a model of the PSM level. In this context, it is necessary to
+follow the process activities described by the methodology.
 
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Dealing with these kind of properties involves the use of specific technologies
-in different layers of the SOC architecture, for instance during the description
-of services APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
-service coordinations (like WS-BPEL~\cite{bpel03}).
-
-%Existing work addressing non-functional properties for service-oriented applications can be classified under four approaches:
-%\begin{trivlist}
-%\item[-] Those coming from the Business Process Domain and from the Web
-%service standards that propose ad-hoc protocols. 
-%Examples of these include the WS*-Family of the W3C\cite{ws-co,ws-tra,btp}.
-%\item[-] Those adopting a classic middleware approach were non-functional
-%properties are provided as middleware services, like the ones presented
-%in~\cite{BeVaC00,RohmBSS02,NepalFGJKS05,Bonita}.
-%\item[-] Those providing languages and enabling the specification of protocols
-%used for expressing service
-%compositions~\cite{LakhalKY05,LakhalKY05b,RouachedGABG06,FauvetDDB05}.
-%\item[-] Those adopting separation of concerns,
-%like~\cite{Milanovic06,Espinosa-OviedoVZC09,SchmelingCM11,PastranaPK11}.
-%\end{trivlist}
-
-Protocols and models implementing non-functional properties assume the existence of a global control of the artifacts implementing the application.
-They also assume that each service exports its interface.
-So, the challenge of supporting non-functional properties is related to
-\textit{(i)} The specification of the business rules of the application; and 
-\textit{(ii)} Dealing with the technical characteristics of the infrastructure where the application is executed.
-
-%In this context, there is a need of 
-%\textit{(i)} flexible protocols and models accepting a best effort approach for ensuring non-functional properties.
-%\textit{(ii)} A methodology for specifying the application logic and its associated non-functional properties, starting at the early phases of the development process.
-%
-%In this work, we proppose to address the structured engineering of service-oriented applications in the presence of non-functional properties where: 
-%\textit{(i)} The designer must make the diference between requirements that concern the application logic and the non-functional requirements;
-%\textit{(ii)} General concepts must be provided to support the representation of  different non-functional requirements;
-%\textit{(iii)} The non-functional requirements, specified in an abstract way, must provide enough information to be translated into the specification of technical aspects implementing concrete non-functional properties to be verified at runtime.
-
-\[ \mathrm{Intro\ SAC} \]
-
-Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.
-Functional properties are not the only crucial aspect in the software development process. 
-Other properties need to be addressed to fit in the application with its context.
-These other aspects are called Non-Functional Properties.
-
-Non-Functional Requirements (NFRs) specify those properties that are not addressed by the functional  specification.
-They are often called \textit{qualities} of the software system.
-%They are also referred as ``constraints'', ``quality attributes'', ``quality goals'', ``quality of
-%service requirements'' and ``non-behavioural requirements'' \cite{Stellman2005}.
-Non-Functional Requirements may specify response time, security constraints or quality of the solution, among others.
-
-Service-Oriented Computing~\cite{Papazoglou2007}, is a software development paradigm where pre-existing services are combined to produce more complex applications. 
-%The selection of services is usually guided by the functional requirements of the application. 
-%In this scenario, there exists the need to provide support for the specification of non-functional requirements, such as
-%security, reliability, and efficiency. 
-The development of service-based applications can benefit from the inclusion of NFRs to the software process from its early stages.
-Failure to comply with this inclusion means that the final application is obtained from a partial specification, making the deployment a difficult task.
-%Ideally, non-functional requirements
-%would be considered along with all the stages of the software development. 
-The adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
-their context.
-Non-functional properties of service oriented applications have been
-addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
-Different proposals~\cite{Babamir2010,AgarwalLS09,CholletL09,GutierrezRF10,XiaoCZBOLH08,JeongCL09,TsadimasNA12}
-support non-functional requirements in the context of web service development.
-
-Most software development methods define software processes that use the notion of refinement.
-Software process begins with the formulation of an abstract specification, which is successively refined to yield the implementation of the system.
-Methods for the development of web service applications are no exception to this rule.
-At least two levels of abstraction can be distinguished: a \textit{Business Level}, including the abstract specification, and a \textit{System Level}, including actual computer programs that implements the system.
-
-In the case of web service applications we will distinguish two separate layers of the implementation.
-The \textit{Composition Layer} is the upper layer of the implementation. 
-It defines the workflow of the system, in terms of individual service calls.
-The \textit{Service Layer} defines those services that are called by the composition.
-
-\[ \mathrm{Original} \]
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-}
+SOD-M considers two points of view:
+\textit{(i)} \textit{business}, focusing on the characteristics and requirements
+of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
+features and processes to be implemented in order application requirements. In
+this way, SOD-M aims to simplify the design of service-oriented applications, as
+well as its implementation using current technologies.

### 2013-11-06 JournalToSubmit2013/intro.tex
Changes: -Complete RelatedWorks section -Changes on Introdution (improving descriptions of MDD and MDA) -Add a section of background (including the description of concepts of MDA and SOD-M) -New references. -Minor comments on section methodologies, modelingNFR and modelingWithPISODM

@@
 
-Model Driven Development (MDD) is a top-down approach for the development of software systems. 
-The main ideas of MDD were originally proposed by the Object Management Group (OMG)~\cite{mda}, as a set of guidelines for the structuring of specifications.
+Model Driven Development (MDD) is approach for the development of software systems mainly characterised by the use of models as a product \cite{Selic03}.
+The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}.
+MDA states a set of guidelines for structuring the specifications of software systems.
 The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}).

### 2013-11-06 JournalToSubmit2013/relworks.tex

--- 2013-10-04 JournalToSubmit2013/5-relworks.tex
@@
-Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
-They also include methods and approaches for modeling non-functional constraints.
+Related works to our approach includes existing proposal dealing with the inclusion of non-functional properties in service-oriented development.
+Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase.
+Finally exiting methodology for the development of service-oriented applications including non-functional properties are here depicted and compared with our approach.
+%As a result of our study of related works, in this section we present a categorization of main concepts related the treatment if non-functional requirement in the field of service-oriented development.
 
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+\subsection{Non-functional properties in service-oriented applications}
+
+In previous work we have carried out a systematic review \cite{placidoPhDThesis2012} in order to analyze how non-functional properties are taking into account in existing approaches for service-oriented development.
+Existing proposals in this field may be classified in three types of works which are described in the followed.
+
+The first types of works are thus considering a particular non-functional concern (e.g.
+security) which is modeled and then "glued" or associated to functional models of an application.
+The work of Chollet et al.
+\cite{CholletL09} defines a proposal to associate non-functional quality properties (security properties in that case) to functional activities which are modeled in a web service composition model.
+Schmeling et al.
+\cite{SchmelingCM11} present an approach and also a toolset for specifying and implementing non-functional concerns in web service compositions; non-functional concerns are modeled first in non-functional model and then related to a service composition represented in a BPMN diagram.
+Ovaska et al.
+\cite{OvaskaEHPA10} present an approach to support quality management at design time.
+Quality requirements are modeled in a first phase and represented then in an architectural model where quality requirements are associated to some components in the model.
+In our case we propose to model non-functional properties joint to the modeling of functional concerns what simplifies the web service development process.
+Non-functional properties are represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.
+
+Second type of works related to our proposal proposes specifics architecture or frameworks to manage and validate QoS attributes in service composition processes.
+The work of Xiao et al.
+\cite{XiaoCZBOLH08} present a framework to verify SLA compliance in composed services.
+Task in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time.
+A similar work is presented in \cite{SchmelingCM11} by using a formal method that allows to choice best suited services instances according to a service composition pattern that include some QoS attributes.
+Babamir et al.
+present in \cite{Babamir2010} a framework for evaluate quality attributes of web services.
+By executing an algorithm, the framework allow to choice best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime.
+In \cite{Karunamurthy2012787} authors propose architecture in which web services may be described with non-functional characteristics that are validated in the service composition process.
+
+Third type of works include proposal working on the evaluation and matching of non-functional properties in the process of search or invocation of particular web services.\cite{AgarwalLS09} define a policy language that provides a way to specific constraints on functional as well as non-functional properties of Web services with the aim of facilitating automated discovery and selection of services.\cite{JeongCL09,Kamalabad2012} defines also proposals to the validate quality attributes of web services when discover and compose.\cite{MohantyRP10} works also on the selection of web service for use considering non-functional properties.
+The paper present web services quality prediction models, which take non-functional properties into account.
+Similarly, \cite{Yeom2006} define a quality driven selection of web services providing a QoS model.
+
+Our proposal by the contrary to the second and third type of proposals is focusing on the design of service-oriented application with non-functional properties and not on the validation of QoS attributes at runtime.
+
+Other relevant works such as \cite{DAmbrogio06,Liu20121080} propose extensions to WSDL and BPEL languages respectively for the inclusion of QoS attributes that can be evaluated at runtime.
+
+\subsection{Standards for programming non-functional properties in service-oriented applications}
+
 Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
@@
 This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In \cite{Vid04} the model
-introduced in \cite{schuldt-etal-TODS} is extended to web services
-for addressing atomicity.
+Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services.
+In \cite{Vidyasankar:2004} the model introduced in \cite{SchuldtABS02} is extended to web services for addressing atomicity.
 
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-There are few methodologies and approaches that address the explicit modeling  of non functional
-properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
+As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
+In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
+Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
+In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
 
- \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
+\subsection{Methodologies for Service-oriented applications development including non-functional properties}
+
+Beside the works presented above, there are few methodologies for service-oriented development that address the explicit modeling of non functional properties for service-based applications.
+Important methodologies for building service-based applications have been proposed in the last years \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008}, but all of they focus mainly on the modeling and construction process of service-based business processes that represent the functionally of information systems.
+
+Some methodologies dealing with particular non-functional properties in service orientation have also appeared.
+Tran et al.
+\cite{Tran2012531} present a model-driven approach for addressing problems related to compliance concerns.
+Authors define extend elements of a business process models with elements of a compliance model in which QoS policies are represented.\cite{GutierrezRF10} present PWSSec process, a methodology that provide developers with all the activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.
+
+Unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that allows to model service-based applications starting from a high-level business modelling, and defining meta-models at different abstraction levels (CIM, PIM and PSM), it enables the design and development of services-based applications that can be reused and that are reliable.
+The definition of a model-driven approach that allows to model non-functional properties in the early stages of the software development process, and the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefits of our proposal.
+
+%\textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
 A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
 Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
- The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
+The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based description language, for defining contracts for services.
 There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several
- \cite{AbrialLNSS91} B-machines refinements that describe the services and their
-compositions.
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
+A services' based application  specification is generated after several \cite{AbrialLNSS91} B-machines refinements that describe the services and their compositions.
+
+%\cite{PapazoglouH06} proposes a methodology based on a SOA extension.
+This work defines a service oriented business process development methodology with phases for business process development.
 The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
@@
 The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.

### 2013-12-04 JournalToSubmit2013/metamodels.tex
to publish music

@@
 
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
+The {\em A-policy} based service composition meta-model (Figure \ref{fig:e-scomposition-metamodel})
+provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.}.
+The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model classes.} and the {\sc Actions} they perform. 
+Instances of this meta-model are UML activity diagrams. 
+Figure~\ref{fig:e-scomposition-metamodel} show the modeling elements.
+Those classes pictured as coloured boxes deal with non-functional properties.
+The classes pictured with a white background are those of SOD-M.
 
-\begin{itemize}
+In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
+\begin{itemizedTrivlist}
 \item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions. 
 They are graphically presented as a partition in the activity diagram. 
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
+A collaborator can be either internal or external to the system. 
+When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
 
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
+\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as an activity. 
+Each action represents some type of transformation or processing. 
+There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
+\begin{figure}[t]
 \centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
+\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
+\caption{{\em A-policy} based service composition meta-model ($\pi$-SCM)}
 \label{fig:e-scomposition-metamodel}
@@
 
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
+\item The {\sc ServiceActivity} element is a composite activity that must be carried out as part of a business service and is composed by one or more executable nodes.
 
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+\item In order to represent constraints associated to services compositions, we extended the SOD-M service composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
+An {\em A-policy} is formed by attributes and rules. 
+Intuitively, the conditions of each rule will be checked.
+In case of no compliance, the actions defined by the rule will be performed.
+The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
+An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
+\end{itemizedTrivlist}
 
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the servicesÕ composition and defines strategies to enforce this property at the servicesÕ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
+\begin{figure}[t]%[htpb]
+\centering
+\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
 
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
+{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
+
+\caption{Service composition model for the ``To publish music'' business service.}
 \label{fig:servicecompositionmodel}
@@
 
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
+\begin{example}[To Publish Music]
+To illustrate the use of the $\pi$-SCM meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
+In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
+The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}. 
+Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
+Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
+%A call to such services must be part of the authentication protocol required by these services.
+In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
+In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
 
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
+{\sf\small OAuthPolicy} will implement the open authentication protocol.
+The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
+This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
+The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
+Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
+
+{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
+The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
 The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
+On the notification of an event of that type, a credential is obtained using the username and password. 
+\hfill\openbox
+\end{example}
+
+{\color{magenta} Aqui\dots  }
 

### 2013-12-04 JournalToSubmit2013/modelingWithPISODM.tex
Changes: -Complete RelatedWorks section -Changes on Introdution (improving descriptions of MDD and MDA) -Add a section of background (including the description of concepts of MDA and SOD-M) -New references. -Minor comments on section methodologies, modelingNFR and modelingWithPISODM
to publish music

@@
 In this section we present $\pi$SOD-M, an MDD based methodology.
-$\pi$SOD-M provides an environment for building service compositions considering
-their non-functional requirements. 
-$\pi$SOD-M extends the SOD-M meta-models by adding
-the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
-to represent non-functional requirements.
+$\pi$SOD-M provides an environment for building service compositions.
+$\pi$SOD-M includes the modelling of non-functional requirements at the early stages of the development.
+The methof extends the SOD-M meta-models by adding the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a} to represent non-functional requirements.
 
@@
 \includegraphics[width=1.0\textwidth]{figs/piSODM}
-\caption{$\pi$SOD-M.}
+\caption{$\pi$SOD-M Overview.}
 \label{fig:piSOD-M}
@@
 
-$\pi$SOD-M proposes the generation of a set of models at different abstraction levels, as
-well as transformations between these models.
+$\pi$SOD-M (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
 $\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints.
@@
 Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models
-(instances of the meta-modes) and transformations between models.
-There are two kinds of transformations:
-Model-to-model transformations are used during the software process to refine the specification.
+$\pi$SOD-M proposes a development process based on the definition of models (instances of the meta-modes) and transformations between models.
+There are two kinds of transformations: Model-to-model transformations are used during the software process to refine the specification.
 Model-to-text transformations are the last step of the process and generate code.
 
-We extended SOD-M to include non-functional specifications.
+We extend SOD-M to include non-functional specifications.
 Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
 The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM (Figure~\ref{fig:piSOD-M}).
+The \textit{$\pi$-PEWS} meta-model is a PSM.
 

### 2013-12-04 JournalToSubmit2013/relworks.tex
to publish music

@@
 
-\subsection{Standards for programming non-functional properties in service-oriented applications}
+\subsection{Standards for programming non-func\-tion\-al properties in service-oriented applications}
 
@@
 
-\subsection{Methodologies for Service-oriented applications development including non-functional properties}
+\subsection{Methodologies for Service-oriented applications development including non-func\-tion\-al properties}
 

### 2013-12-06 JournalToSubmit2013/metamodels.tex
section 4

@@
 
-The {\em A-policy} based service composition meta-model (Figure \ref{fig:e-scomposition-metamodel})
+The {\em A-policy} based service composition ($\pi$-SCM) meta-model, described in Figure \ref{fig:e-scomposition-metamodel},
 provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.}.
@@
 Instances of this meta-model are UML activity diagrams. 
-Figure~\ref{fig:e-scomposition-metamodel} show the modeling elements.
-Those classes pictured as coloured boxes deal with non-functional properties.
+Figure~\ref{fig:e-scomposition-metamodel} show the modelling elements.
+Classes pictured as coloured boxes deal with non-functional properties.
 The classes pictured with a white background are those of SOD-M.
@@
 \includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based service composition meta-model ($\pi$-SCM)}
+\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
 \label{fig:e-scomposition-metamodel}
@@
 
-\caption{Service composition model for the ``To publish music'' business service.}
+\caption{$\pi$-SCM for the ``To publish music'' business service.}
 \label{fig:servicecompositionmodel}
@@
 
-\begin{example}[To Publish Music]
-To illustrate the use of the $\pi$-SCM meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
-In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
+\begin{example}[To Publish Music]\label{ex:toPublicMusic}
+To illustrate the use of the $\pi$-Service Composition Meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
+In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
+% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
 The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}. 
@@
 
-{\color{magenta} Aqui\dots  }
+We propose the use of rules and policies to model and associate non-functional properties to service compositions.
+These artifacts will be used to generate the actual programs that will implement the application:
+Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
+The $\pi$-PEWS  meta-model is described in the next section.
 
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
+\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
 
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
+PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
+The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as B	PEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
+Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
+\begin{itemizedTrivlist}
+\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
+A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
+A {\sc Compound Operation} is defined using an {\sc Operator}.
+The language defines operators to denote guarded operations ($[C]S$); sequential ($\ .
+\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.
 
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
+\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
+A brie description of these classes is given next.
+\end{itemizedTrivlist}
 %
-\begin{figure}
+\begin{figure}[t]
 \centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
+\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
 \caption{$\pi$-{\sc Pews} Metamodel}
@@
 
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
+Figure~\ref{fig:metamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
+Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.  
+In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.
+%In this way,
+%\begin{itemizedTrivlist}
+%\item The class {\sc Scope} represents any element of a service composition (i.e., operation, operator, path).
+%\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}. 
+%An {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
 The value of a variable is represented using the type {\sc Variable}.
 The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
+%\end{itemizedTrivlist}
+%
 %An authentication {\em A-policy} represents the situation where an invocation in
@@
 %identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
 
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
+Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
+The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.

### 2013-12-06 JournalToSubmit2013/modelingWithPISODM.tex
section 4

@@
 The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS code~\cite{BaCAM05,Placido2010LTPD}.
-Although PEWS is our language of choice, other composition languages can be used as target.
+\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
 This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

### 2013-12-06 JournalToSubmit2013/transformations.tex
section 4

@@
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
+Figures~\ref{fig:transformationsA}, \ref{fig:transformationsB} and \ref{fig:transformationsC} show the transformation rules from the elements of a $\pi$-SCM meta-model into elements of the $\pi$-{\sc Pews} meta-model.
+There are two groups of rules: those that transform service composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
 \begin{figure}
 \centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
+\includegraphics[width=0.96\textwidth]{figs/Mapping-1a.png}
+}
+\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformations (Services).}
+\label{fig:transformationsA}
 \end{figure}
 
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
+\subsection{Transformation of $\pi$-SCM service composition elements into $\pi$-{\sc Pews} elements}
 
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
+The transformation rules concerning service composition elements of $\pi$-SCM are shown in figure~\ref{fig:transformationsA}.
+Named actions of $\pi$-SCM (represented by {\sc\em Action} and {\sc\em Action:name}) are transformed into a named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}. 
+Named service activities represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM are transformed into named operations of the $\pi$-{\sc Pews} model, represented by the elements {\sc CompositeOperation} and {\sc CompositeOperation:name}.
+Workflow definitions are translated to the usual control flow structures, using the sequential, parallel and choice combinators of PEWS.
 
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
+\begin{example}
+In the scenario ``To Publish Music'' of Example~\ref{ex:toPublicMusic}, the service activity {\sf PublishMusic} of the $\pi$-SC model calls to two {\sf Activities} of type {\em UpdateMusic}, respectively concerning the {\em Facebook} and {\em Twitter} business services.
+The {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model is obtained.
+This operation uses the parallel constructor and it is written: {\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}.
+\end{example}
 
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
+\begin{figure}[t]
+\centering{
+\includegraphics[width=0.96\textwidth]{figs/Mapping-1a.png}
+}
+\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformations.}
+\label{fig:transformationsA}
+\end{figure}
 
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
+\subsection{Transformation of A-policies and rules into $\pi$-{\sc Pews}}
 
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
+The {\em A-policies} defined for the elements of $\pi$-SCM are transformed into $\pi$-{\sc Pews} {\sc A-Policy} classes.
+These classes keep their names, as expressed in the source model. 
+The transformation of \textit{rules} is guided by the event types associated to these rules.
+The $\pi$-SCM {\em A-policy} variables, given as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} of the $\pi$-SCM model.
+
+As shown in Figures~\ref{fig:transformationsA} and~\ref{fig:transformationsB}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
+The condition inside a $\pi$-SCM rule ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
+
+\begin{figure}
+\centering{
+\includegraphics[width=0.96\textwidth]{figs/Mapping-1b.png}
+}
+\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformations (\textit{A-Policies}).}
+\label{fig:transformationsB}
+\end{figure}
+
+\begin{example}
+In the ``To Publish Music'' scenario of Example~\ref{ex:toPublicMusic}, the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model.
+In both cases the events are of type {\sf ActivityPrepared}. 
 These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}. 
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
+Their transformed counterparts are associated to operations {\em PublishFacebook} and {\em PublishTwitter}.
+\end{example}
 
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}
+\begin{figure}
+\centering{
+\includegraphics[width=0.96\textwidth]{figs/Mapping-1c.png}
+}
+\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformations (Rules).}
+\label{fig:transformationsC}
+\end{figure}

### 2013-12-09 JournalToSubmit2013/SOD-M.tex
no message

@@
 well as its implementation using current technologies.
+
+{\LARGE \color{red} Relocate the next subsections or delete them --Genoveva and Martin}
+
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+\subsection{Modeling a services' based application}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+ Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
+Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
+Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+\begin{figure} [htpb]
+\centering
+\includegraphics[width=0.65\textwidth]{figs/SODM}
+\caption{SOD-M development process}
+\label{fig:sodm}
+\end{figure} 
+As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
+Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
+and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
+%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
+%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
+%provide
+%is an evolving approach to software development that deals with the provision of 
+%models, transformations between them and code generators to address software development. 
+%One of the main advantages of model-driven approaches is the provision of 
+%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
+
+%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
+
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+\subsection{Modeling non-functional constraints of services' based applications}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call a service in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
+% song title disseminated in the walls of the user's Facebook and Twitter accounts.
+
+Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
+Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
+Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
+Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
+In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
+
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+%\subsection{$\pi$-SOD-M}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+
+%
+%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
+%
+Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
+%This method is described in the following sections.
+In order to do, our work organizes non-functional constraints into 
+three layers representing: application modeling, services composition and services.
+The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements.
+At the application layer NFP can refer to business rules and values.
+A value NFR expresses constraints about the way data and functions can be accessed and executed.
+For example accessing methods under security protocols.
+
+Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
+For example, response time, storage capacity (e.g., Dropbox service provides 5Giga free storage).
+Value constraints concern more on the conditions in which services can be used.
+For example, accessing to a function within an authentica-tion protocol.
+
+Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
+For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
+
+As a first step in our approach,  we started modeling non-functional constraints at the PSM level.
+Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
+$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
+Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).

### 2013-12-09 JournalToSubmit2013/implementation.tex
no message

--- 2013-10-31 CAISE2011/submitted/caise-etal.tex
@@
-\documentclass{llncs}
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-\begin{document}
-
-\title{Generating reliable services' composition  using A-policies: a model-driven approach\thanks{This work is partially financed by the projects {\sc Orchestra} ECOS-ANUIES, {\sc Clever} STIC-AMSUD, and {\sc Masai}.}}
-
-\author{Genoveva Vargas-Solar\inst{1}
-\and Valeria de Castro\inst{4}
-\and Pl\'{a}cido Antonio de Souza Neto\thanks{Financed by CAPES/STIC-AMSUD Brasil, BEX 4112/11-3} \inst{5,6}
-\and Javier A. Espinosa-Oviedo\inst{3}
-\and Esperanza Marcos\inst{4}
-\and Martin A. Musicante\inst{5}
-\and Jos\'{e}-Luis Zechinelli-Martini\inst{2,1}
-\and Christine Collet \inst{3}}
-
-\institute{
-	French Council of Scientific Research, LIG-LAFMIA	\\
-	681 Rue de la Passerelle, BP. 72, 38402	Saint Martin d'H\`{e}res Cedex, France		\\
-		\email{Genoveva.Vargas-Solar@imag.fr}
-\and
-	Fundaci\'{o}n Universidad de las Am\'{e}ricas, Puebla	\\
-	Exhacienda Sta. Catarina M\'{a}rtir s/n 72820, San Andr\'{e}s Cholula, Puebla, M\'{e}xico			\\
-	\email{joseluis.zechinelli@udlap.mx}
-\and
-	Grenoble Institute of Technology	\\
-	46 Avenue F'{e}lix Viallet, 38031 Grenoble Cedex 1		\\
-	\email{Christine.Collet@grenoble-inp.fr}, \email{Javier.Espinosa@imag.fr} \\
-\and
-	Universidad Rey Juan Carlos	\\
-	Av Tulip\'{a}n, M\'{o}stoles, Spain 		\\
-	\email{Valeria.deCastro@urjc.es, esperanza.marcos@urjc.es} \\
-\and
-	DIMAp - UFRN, ForAll - Formal Methods and Language Research Laboratory	\\
-	Campus Universitário - Lagoa Nova, Natal - RN, Brasil		\\
-	\email{mam@dimap.ufrn.br} \\
-	\and
-	Instituto Federal do Rio Grande do Norte	\\
-	Av. Senador Salgador Filho, 1559 - Tirol, Natal - RN, Brasil		\\
-	\email{placido.neto@ifrn.edu.br}
-}
-
-\maketitle
-
-\begin{abstract}
-This paper presents an approach for modeling and associating {\em A-Policies} to  services' based applications.
-It proposes to extend the SOD-M model driven method with   (i) the $\pi$-{\sc SCM} an {\em A-Policy} services' composition meta-model for representing non-functional constraints associated to services' based applications;  (ii) the $\pi$-{\sc Pews} meta-model providing guidelines for expressing the composition and the policies; and, (iii)  model to model and model to text transformation rules for semi-automatizing the implementation of reliable services' compositions.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-\end{abstract}
-
-%*********************************************************************************************************
-\section{Introduction}
-%*********************************************************************************************************
-Service oriented computing is at the origin of an evolution in the field of software development. 
-An important challenge of service oriented development is  to ensure the alignment between IT systems and the business logic.
-%dealing thereby with the promise that IT systems can  evolve  according to the business needs. 
-Thus, organizations are  seeking for mechanisms to deal with the gap between the systems developed and business needs \cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone  in the development of meaningful services' based applications \cite{5}.
-In this context, some authors argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are the key to make real the idea of rapid, accurate development of software that serves, rather than dictates, software users' goals \cite{watson}.
-
-Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.
-Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.
-
-The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase.
-Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS).
-%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.
-
-Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.
-%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it.
-They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method).
-Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}.
-Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation.
-As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).
-
-The remainder of the paper is organized as follows.
-Section \ref{sec:motivation} gives an overview of our approach.
-It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify.
-Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
-%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
-%
-Section \ref{sec:implementation} describes implementation and validation issues.
-Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms.
-Section \ref{sec:conclusions} concludes the paper and discusses future work.
-
-%*********************************************************************************************************
-\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
-%*********************************************************************************************************
-Consider for instance the following scenario.
-An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
-\begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export a method for  updating the status of a given user:
-\begin{itemize} 
-%\item{\sf\small get-Status ( usedid ): String ; 
-\item {\sf\small update-Status ( userid, new-status ): String}; 
-\end{itemize}
-\end{itemize}
-%
-%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-%the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
-The "To Publish Music" scenario 
-%(showed by doted line) 
-starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\begin{figure}
-\centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
-\end{figure}
-
-Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
-Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).
-
-%*********************************************************************************************************
-\section{$\pi$ services' composition meta-model}\label{sec:piscm}
-%*********************************************************************************************************
-
-The {\em A-policy} based services' composition meta-model (see  in Figure \ref{fig:servicecompositionmodel})
-represents  a workflow needed to implement a services' composition, identifying those entities that collaborate in the business processes (called {\sc Business Collaborators} \footnote{We use {\sc capitals} for referring to meta-models' classes.}) and the {\sc Actions} that  they perform.
-This model is represented by means of a UML activity diagram.
-Thus, as shown in Figure \ref{fig:e-scomposition-metamodel}, the meta-model includes typical modeling elements of the activity diagram such as {\sc ActivityNodes}, {\sc InitialNodes} and {\sc FinalNodes}, {\sc DecisionNodes}, etc., along with new elements defined by SOD-M such as {\sc Business Collaborators}, {\sc ServiceActivity} and {\sc Action} (see the white  elements   in Figure \ref{fig:servicecompositionmodel}).
-
-\begin{itemize}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions.
-They are graphically presented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system being modelled.
-When the collaborator of the business is external to the system, the attribute {\sf\small IsExternal} \footnote{We use the {\sf sans serif} font for referring to models' classes defined using a meta-model.} of the collaborator is set to true.
-
-\item {\sc Action}, a kind of {\sc ExecutableNode}, are represented in the model as an activity.
-Each action identified in the model describes a fundamental behaviour unit which represents some type of transformation or processing in the system being modelled.
-There are two types of actions: i) a WebService (attribute Type is {\sf\small WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\item The {\sc ServiceActivity} element is a composed activity that  must be carried out as part of a business service and is composed of one or more executable nodes.
-\end{itemize}
-
- \begin{figure}[htpb]
-\centering
-\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
-\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
-Figure \ref{fig:servicecompositionmodel}).
-There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}.
-Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-
-Instead of programming different protocols within the application logic, we propose to include the modeling of non-functional constraints like transactional behaviour, security and adaptability  at the early stages of the services' composition engineering process.
-We model non-functional constraints of services' compositions using the notion of {\em A-policy} \cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}, a kind of pattern for specifying {\em A-policy} types.
-In order to represent constraints associated to services compositions, we extended the SOD-M services' composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  can be evaluated according to a condition represented by the {\sc Condition} part and the  action  to be executed for reinforcing  it represented by the {\sc Action} part.
-%An {\em A-policy}  describes  specific constraint types that can correspond to non-functional constraints like  transactional behaviour, security and adaptability. 
-%An {\em A-policy}  defines a set of constraints over the execution states of the services’ composition and defines strategies to enforce this property at the services’ composition execution time.
-An {\em A-policy} groups a set of rules.
-It describes global variables and operations that can be shared by the rules and that can be used for expressing their Event  and Condition parts.
-An {\em A-Policy} is associated to the elements {\sc BusinessCollaborator}, {\sc ServiceActivity} and, {\sc Action}  of the $\pi$-SCM meta-model (see Figure \ref{fig:e-scomposition-metamodel}) .
-
-\begin{figure}[htpb]
-\centering
-\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
-\caption{Services' composition model for the business service "To publish music"}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-Given that {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the users' space.
-A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small Twitter OAuthPolicy} that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the  class {\sf\small Facebook HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
- {\sf\small OAuth}  implements the open authentication protocol.
-As shown in Figure \ref{fig:servicecompositionmodel}, the {\em A-policy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable type is imported through the library {\sf\small OAuth.Token}.
-The {\em A-policy}  defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (i) if no token has been associated to the variable {\sf\small token}, stated in by the condition of rule {\sf\small R$_1$}, then a token is obtained (action part of {\sf\small R$_1$}); (ii) if the token has expired, stated in the condition of rule {\sf\small R$_2$}, then it is renewed (action part of {\sf\small R$_2$}).
-Note that the code in the actions profits from the imported  {\sf\small OAuth.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTP-Auth} implements the HTTP-Auth protocol.
-As shown in Figure  \ref{fig:servicecompositionmodel}, the {\em A-policy} imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password values.
-The object storing the credential is associated to the scope, i.e., the activity that will then use it for executing the method call.
-
-Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code.
-For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.
-
-Back to the  definition process of a SIS, once the {\em A-policy} based services' composition model has been defined, then it can be transformed into a model (i.e., $\pi$-PEWS model) that can support then executable code generation.
-The following Section describes the $\pi$-PEWS  meta-model that supports this representation.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-The idea of the $\pi$-{\sc Pews} meta-model is based on the services' composition approach provided by the language PEWS\cite{BaAM06,Placido2010LTPD} (\textit{Path Expressions for Web Services}), a programming language that lets the service designer  combine the methods or subprograms that
-implement each operation of a service, in order to achieve the desired application logic. Figure \ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model
-consisting of  classes representing:
-\begin{itemize}
-\item A services' composition: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, and  {\sc Operator} for representing a services' composition and {\sc Path} representing a services' composition.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}
-denoted by an identifier. A {\sc Compound Operation} is defined using an  {\sc Operator}  that can be represent  sequential ($\ . \ $) and parallel ($\ \| \ $) composition of services,
- choice ($\ + \ $) among services,
-the sequential ($*$) and parallel ($\{\dots\}$) repetition of an operation or the conditional execution of an operation ($[C]S$).
-
-\item {\em A-Policies} that can be associated to a services' composition:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-\end{itemize}
-%
-\begin{figure}
-\centering
-\includegraphics[width=0.80\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-As shown in the diagram an {\sc A-Policy} is applied to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators, the authorized delay between reading a song title in Spotify and updating the walls must be less then 30 seconds), and a {\sc Path} (e.g., executing the walls' update under a strict atomicity protocol -- all or noting).
-It groups a set of ECA rules, each rule having a classic semantics, i.e, {\em when an event of type E occurs if  condition C is verified then execute the action A}.
-Thus, an {\em A-policy} represents a set of reactions to be possibly executed if one or several triggering events of its rules are notified.
-\begin{itemize}
-\item The class {\sc Scope} represents any element of a services' composition (i.e., operation, operator, path).
-\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}.
-A {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-\end{itemize}
-
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-%
-
-Given a $\pi$-SCM model of a specific services' based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model thanks to transformation rules.
-The following Section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models of our method.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\section{Transformation rules}\label{sec:mmrules}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
-There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
-A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
-
-In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
-Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
-\begin{figure}
-\centering{
-%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
-\label{fig:transformations}
-\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%\noindent
-
-%{\bf\em 
-\subsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
-
-As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
-
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
-
-%
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
-%\end{figure}
-
-%*********************************************************************************************************
-\section{Implementation issues}\label{sec:implementation}
-%*********************************************************************************************************
 This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
@@
 
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{$\pi$-SOD-M Development Environment}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+%\subsection{$\pi$-SOD-M Development Environment}
+%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 
@@
 \end{itemize}
-%
-%
-%
 
@@
 %The order of policies has implications on the global order of the rules to be executed.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Validation}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-We validated our approach  by implementing the "To Publish Music" application.
-It consists in atomically synchronizing  the status of a user's accounts according to the music she listens in Spotify and using authentication protocols for updating the walls' status in her Facebook and Twitter accounts.
-Once the services' composition model has been annotated with the corresponding policies, then the reliable composition code can be generated using the rules defined for this purpose.
-Our implementation includes a  code generator that generates executable code for a reliable services' composition.
-The model represents an services' composition expression and its associated policies.
-
-We tested the application  in order to see whether the coordination tolerated services unavailability keeping the status synchronized.
-Particularly, the token of the activity associated to Twitter was consistently obtained when the service was available.
-We also implemented a reference coordination without policies.
-We dealt with the availability exceptions by hand and the authentication protocols were embedded within the activities code.
-During our validation Twitter changed the authentication protocol and the maintenance of the application this implied re-programming the reference application.
-Instead, with the policies approach we only had to desactivate the corresponding policy and associate the appropriate one to the activity calling the service Twitter (one code line), and run the application again.
-
-%*********************************************************************************************************
-\section{Related works}\label{sec:related}
-%*********************************************************************************************************
-Related works to our approach include standards devoted for expressing non-functional constraints for services and services' compositions.
-They also include methods and approaches for modeling non-functional constraints.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Programming non-functional properties for services}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
-Other works, like \cite{helga2}
-introduce a model for transactional services composition based on
-an advanced transactional model. \cite{samy} proposes an approach
-that consists of a set of algorithms and rules to assist designers
-to compose transactional services. In {\cite{Vid04} the model
-introduced in \cite{schuldt-etal-TODS} is extended to web services
-for addressing atomicity.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Modeling non-functional properties}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-There are few methodologies and approaches that address the explicit modeling  of non functional
-properties for service based applications.   Software process methodologies for
-building  services based applications have been proposed in\cite{PapazoglouH06,cdl2006,FeuerlichtM05,Ramollari_asurvey}, and they focus mainly on the modeling and construction process of services based business processes that represent the application logic of information systems.
-
- \textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
-A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
-Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
- The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based
-description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several
- \cite{AbrialLNSS91} B-machines refinements that describe the services and their
-compositions.
-\cite{PapazoglouH06} proposes a methodology based on a SOA extension. This work defines a service
-oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
-%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-%specification, realization, implementation and deployment monitoring and management.
-
- %\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-%represents a composition of services. It classifies services   as public utilities, business services or
-%composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\noindent{\bf\em 
-\subsection{Discussion}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-As WS-*  and similar approaches, our work enables the specification and programing of  crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional requirements are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined {\em A-Policies} for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-Furthermore, unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that proposes meta-models at different levels (CIM, PIM and PSM) and extending the PSM meta-models, it enables  the design and development of services' based applications that can be reused and that are reliable.
-
-%*********************************************************************************************************
-\section{Conclusions and future work}\label{sec:conclusions}
-%*********************************************************************************************************
-This paper presented $\pi$-SOD-M for specifying and designing reliable service based applications.
-We model and associate policies to  services' based applications that represent both systems' cross-cutting aspects and use constraints stemming from the services used for implementing them.
-We extended the SOD-M method, particularly the $\pi$-SCM (services' composition meta-model) and $\pi$-{\sc Pews} meta-models for representing both the application logic and its associated non-functional constraints and then generating its executable code.
-We implemented the meta-models on the Eclipse platform and we validated the approach using a use case that uses authentication  policies.
-
-Non-functional constraints are related to business rules associated to the general "semantics" of the application and in the case of services' based applications, they also concern the use constraints imposed by the services.
-We are currently working on the definition of a method for explicitly expressing such properties in the early stages of the specification of services based applications.
-Having such business rules expressed and then translated and associated to the services' composition can help to ensure that the resulting application is compliant to the user requirements and also to the characteristics of the services it uses.
-
-Programming non-functional properties is not an easy task, so we are defining a set of predefined {\em A-policy} types with the associated use rules for guiding the programmer when she associates them to a concrete application.
-{\em A-policy} type  that can also serve as patterns for programming or specializing the way non-functional constraints are programmed.
-
-\bibliographystyle{plain}
-\bibliography{caise,S-Coord,Bib/e-commerce,Bib/trans,Bib/Component,Bib/XML,Bib/Workflow,Bib/SODM,Bib/G,Bib/docbi}
-
-%*********************************************************************************************************
-%\section*{Appendix}\label{sec:appendix}
-%*********************************************************************************************************
-%Figure \ref{fig:mappingrules-1} shows the transformation rules between the $\pi$-SCM meta-model to the $\pi$-{\sc Pews}  meta-model.
-%\begin{figure}
-%\includegraphics[width=1.0\textwidth]{figs/Mapping-1}
-%\caption{Mapping rules between $\pi$-SCM  and $\pi$-{\sc Pews} meta-models}
-%\label{fig:mappingrules-1}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[width=0.70\textwidth]{figs/Mapping-3}
-%\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} policy elements}
-%\label{fig:mappingrules-3}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[width=0.70\textwidth]{figs/Mapping-2}
-%\caption{Mapping rules between $\pi$-SCM meta-model and $\pi$-{\sc Pews} rule elements}
-%\label{fig:mappingrules-2}
-%\end{figure}
-
-\end{document}

### 2013-12-09 JournalToSubmit2013/modelingFlyingPig.tex
no message

--- 2013-10-29 JournalToSubmit2013/modelingFlyingPig.tex
@@
-In order to introduce the context of our work consider the FlyingPig scenario concerning a risk assesment system, where finanacial companies need to be assesed about the risks implied by their business practice.
-For instance, the condition in which confidential transactions are performed, the physical security for accessing reserved areas (i.e.
-access to the critical servers room should be secured).
-The system provides a set questionaires that provide a thorough description of the practice of the company.
+We validated our methodology by in a use-case concerning risk assessment for financial companies, inspired on the ORCA System\footnote{The ORCA System is a trademark of GCP Global www.gcpglobal.com.}.
+Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied the client's business practice. 
+For instance, the condition in which confidential transactions are performed, the physical security for accessing reserved areas (i.e. access to the critical servers room should be secured). 
 This information is used to determine whether there are risky actions within the business processes of the company that can be amended. 
-It is also used to determine a degree of compliance of security management of financial companies norms.
-When things have to be amended it is necessary to verify that evidence that the amendment was done and whether it  reduces the risk.
+It is also used to determine a degree of compliance to existing norms. 
+By analysing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
 
- In order to provide these functions the application FlyingFish benefits from existing services that provide information,  storage, assesment and data visualization functions.
-Thus, FlyingFish is a service based application that implements a business process.
- The business process is specified in terms of ordered tasks and define the application logic of a Service oriented Application.
-Tasks can be performed by a person or an entity.
-In our context, tasks are implemented by Services.
-Service is an application implemented by a provider that exports an API through a network (e.g., Internet)
-API is defined using an interface definition language (IDL, e.g., WSDL for Web services).
- In our example the business procces  starts by a request from a company branch asking for an assesment.
-The request leads to the declaration of a group of users that will answer quetionaires that will be used as input for evaluating risk.
-Questionaires are considered tasks that users will have to perform.
-Given a list of tasks a user decides whether she takes on the task (i.e., answering a questionaire) or delegate it to another user.
+Our goal is to model a service based application (called \textsl{FlyingPig}), for providing risk assessment as a service.
+In order to provide this functionality \textsl{FlyingPig} would benefit from ORCA's legacy services: storage, assessment and data visualization functions.
 
-During the process of answering a questionnaire, the answer of a question can trigger the generation of other questonaires, that become new tasks to be executed.
-Once a questionnaire has been completely answered it is stored and processed to generate a set of uncompliant situations associated to their corresponding amendments if there are any or a report specifying a compliance level, incidents and a risk map.
-
-Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
-the scenario.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
-The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
-
-Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
-
-%The "To Publish Music" scenario 
-%%(showed by doted line) 
-%starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
-%Twitter and Facebook services are then contacted in parallel for %updating the user's status with the corresponding song title %(activities {\sf Update Twitter} and {\sf Update Facebook}).
+In the following, we describe the results of applying $\Pi$-SODM to develop the \textsl{FlyingPig} risk assessment system.
+The models presented next are the final results of the application of $\Pi$-SODM.
+They were generated as a result of interactions with GCP Global software developers.
 
@@
 \centering
-\includegraphics[width=0.55\textwidth]{figs/SC}
-\caption{BPMN model of the "To Publish Music" scenario}
-\label{fig:E3valuemodel}
+\includegraphics[width=0.7\textwidth]{figs/E3-Value.pdf}
+\caption{E3value model of the FlyingPig.\label{fig:E3valuemodel}}
 \end{figure}
 
+\subsection{CIM}
+
+$\Pi$-SODM uses two models, to be built at the CIM level (See Section~\ref{sec:modelingWithPISODM}).
+
+Figure~\ref{fig:E3valuemodel} shows the E3 value model~\cite{e3value} for \textsl{FlyingPig}.
+The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\bigtriangleup\ \bigtriangledown$) and value activities (rounded boxes) performed by business actors (squared boxes).
+
+In our use-case, we identify two business actors: ORCA and the client company that interact to produce value.
+Value exchanges are determined by a service (provided by ORCA) and information and economic values (provided by the client).
+
+{\color{magenta} Valeria: Could you please complete these ideas? --G \& M.}
+
+Figure~\ref{fig:bpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} for the scenario. 
+There are three business actors: ORCA, the Client company and a User.
+The two latter business actors belong to a channel, which is a third-party broker that manages the communication between GCP Global and the Client company. 
+The User is a contact member of the Company, who will coordinate the assessment process.
+This process will involve other members of the company as well.
+
+The process  starts by a request from a company branch asking for an assessment. 
+The request leads to the definition of a group of users that will answer questionnaires for evaluating risk. 
+Questionnaires are considered \textit{tasks} that users will have to perform. 
+Given a list of tasks a user decides whether she takes on the task (i.e., answering a questionnaire) or delegate it to another user.
+
+During the process of answering a questionnaire, the answer of a question can trigger the generation of other questionnaires, that become new tasks to be executed.
+Once a questionnaire has been completely answered it is stored and processed to generate a set of un-compliant situations associated to their corresponding amendments if there are any or a report specifying a compliance level, incidents and a risk map.
+
+{\color{magenta} Valeria and Khalid: Could you please complete these ideas? --G \& M.}
+
 Business processes have also associated rules and constraints that define their non functional requirements.  
-NFR represents the “semantics” and the conditions in which the tasks must be done. 
+NFR represents the ``semantics'' and the conditions in which the tasks must be done. 
 In our example we have some constraints. 
 \begin{itemize}
-\item Twtter requires to be accessed through an authentication protocol, and the police crime notification service has a control  access protocol and allows only 3 tries to access it. 
-\item Only users with enough privileges can store crimes’ notifications. 
-\item Users can only track crimes notified by providers t that hey have authorization to contact; For example civil population cannot track all the crimes notified by the police. 
-\item If the google map is unavailable the results of a track request are delivered on text. 
-\item If a user tracks crimes that are provided by a service for which she does not have authorization then show an empty map. 
-\item Detailed information about a crime depends on the user privileges, on the police assignment regions. 
-\item The tracking crime app keeps track of the accesses to detailed information about crimes. 
+\item {\color{magenta} Placido and Umberto: Could you please complete these items? --G \& M.}
 \end{itemize}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
+\caption{BPMN model for FlyingPig.\label{fig:BPMNmodel}}
+\label{fig:bpmn}
+\end{figure}
+
+\subsection{PIM}
+
+{\color{magenta} Placido and Umberto: Could you please comment models here? --G \& M.}
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{figs/UseCaseGeneral.png}
+\caption{$\pi$-UseCase model for FlyingPig.\label{fig:piUseCaseModel}}
+\label{fig:bpmn}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneral.png}
+\caption{$\pi$-ServiceProcess model for FlyingPig.\label{fig:PiServiceProcessModel}}
+\label{fig:bpmn}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneral.png}
+\caption{$\pi$-ServiceComposition model for FlyingPig.\label{fig:PiServiceCompositionModel}}
+\label{fig:bpmn}
+\end{figure}
+
+\subsection{PSM}
+
+{\color{magenta} Placido and Umberto: Could you please insert the $\pi$-PEWS model here? --G \& M.}
+
+\subsection{Lessons Learned}
 
@@
 Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling a services' based application}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
- Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
-Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
-Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-\begin{figure} [htpb]
-\centering
-\includegraphics[width=0.65\textwidth]{figs/SODM}
-\caption{SOD-M development process}
-\label{fig:sodm}
-\end{figure} 
-As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
-Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
-%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
-%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
-%provide
-%is an evolving approach to software development that deals with the provision of 
-%models, transformations between them and code generators to address software development. 
-%One of the main advantages of model-driven approaches is the provision of 
-%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.
-
-%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{Modeling non-functional constraints of services' based applications}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call a service in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
-% song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-%
-%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints.
-%
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-In order to do, our work organizes non-functional constraints into 
-three layers representing: application modeling, services composition and services.
-The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements.
-At the application layer NFP can refer to business rules and values.
-A value NFR expresses constraints about the way data and functions can be accessed and executed.
-For example accessing methods under security protocols.
-
-Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
-For example, response time, storage capacity (e.g., Dropbox service provides 5Giga free storage).
-Value constraints concern more on the conditions in which services can be used.
-For example, accessing to a function within an authentica-tion protocol.
-
-Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
-For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
-
-As a first step in our approach,  we started modeling non-functional constraints at the PSM level.
-Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).
-$\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}).
-Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).

### 2013-12-12 JournalToSubmit2013/modelingFlyingPig.tex
Value model and BPMN

@@
 
-Figure~\ref{fig:E3valuemodel} shows the E3 value model~\cite{e3value} for \textsl{FlyingPig}.
-The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\bigtriangleup\ \bigtriangledown$) and value activities (rounded boxes) performed by business actors (squared boxes).
+Figure~\ref{fig:E3valuemodel} shows the value model~\cite{e3value} for the case of \textsl{FlyingPig}.
+The value model is a business model that represents a business case graphically as a set of value exchanges ($\bigtriangleup\ \bigtriangledown$) and value activities (rounded boxes) performed by business actors (squared boxes).
 
-In our use-case, we identify two business actors: ORCA and the client company that interact to produce value.
-Value exchanges are determined by a service (provided by ORCA) and information and economic values (provided by the client).
+In our use-case, we identify two business actors: \textsl{ORCA} and \textsl{Broker}.
+Brokers are the responsible for channeling requests for risk assessment of one or several companies.
+ORCA have two value activities which are services that provide it an economical benefit: to \textsl{identify amendments} and the possibility to \textsl{asses risk situation}.
+The value exchanged between ORCA and Brokers are: \textsl{amendments} and \textsl{evaluation reports} which are value objects ([...]) for the any companies that need to have a risk assessment, and the \textsl{questionnaire and evidences} and the risk assessment \textsl{fee} which are value objects for ORCA System.
 
-{\color{magenta} Valeria: Could you please complete these ideas? --G \& M.}
+Beside the static representation of a business case, the value model allows for the representation of dependency paths which enhance the understanding of a business idea by showing all value exchanges triggered by the occurrence of an end-consumer need.
+A dependency path has a direction and consists of dependency nodes linked by mean of connections.
+A dependency node is a start stimulus, an AND-fork or AND-join, an OR-fork or OR-join, or an end node (see Legend on Figure~\ref{fig:E3valuemodel}).
+An start stimulus represents a trigger for the exchange of economic value objects while an end node represents a model boundary.
 
+Figure~\ref{fig:E3valuemodel} shows the dependency paths for \textsl{FlyingPig}.
+The dependency path identified in that case is the need of assessment of a particular company.
+It denotes a stimulus caused by the client company who ask for risk assessment.
+Once this need occurs the value exchanges between ORCA and Broker are triggered.
+The value model represent that ORCA will provide amendments and an evaluation report in return for the questionnaire and evidences and the fees provided by the Broker.
+
+{\color{blue} Javier: Please modify the value model to include fee as a value object in return for evaluation report. Modified figure is on the git}
+
+A better understanding of the process in which these value exchanges occurs may be modeled using a BPMN model.
 Figure~\ref{fig:bpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} for the scenario.
-There are three business actors: ORCA, the Client company and a User.
-The two latter business actors belong to a channel, which is a third-party broker that manages the communication between GCP Global and the Client company. 
-The User is a contact member of the Company, who will coordinate the assessment process.
+It includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
+Brokers have two lanes, the client \textsl{Company} and a \textsl{User}.
+The user is a contact member of the company, who will coordinate the assessment process.
 This process will involve other members of the company as well.
 
-The process  starts by a request from a company branch asking for an assessment. 
+The process starts by a request from a company asking for a risk assessment.
+Note that it is in agreement with the explanation of the start stimulus on the value model above.
 The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
-Questionnaires are considered \textit{tasks} that users will have to perform. 
-Given a list of tasks a user decides whether she takes on the task (i.e., answering a questionnaire) or delegate it to another user.
+Questionnaires are considered tasks that users will have to perform.
+Given a list of tasks, the company assign a responsible and delegate the task.
+A task may be to answer a questionnaire or add an evidence but also could implies to solve some "risky situation" (e.g.
+to repair a door {\color{blue} ...¿Genoveva mira por favor si esto puede ser o hay algun ejemplo mas bueno/serio???} ).
 
-During the process of answering a questionnaire, the answer of a question can trigger the generation of other questionnaires, that become new tasks to be executed.
-Once a questionnaire has been completely answered it is stored and processed to generate a set of un-compliant situations associated to their corresponding amendments if there are any or a report specifying a compliance level, incidents and a risk map.
-
-{\color{magenta} Valeria and Khalid: Could you please complete these ideas? --G \& M.}
+Once a tasks are completed they are stored and analyzed to generate a set of un-compliant situations associated to their corresponding amendments if there are any or a report specifying a compliance level, incidents and a risk map.
+During the process of analysing a questionnaire, the answers of some questions can trigger the generation of other questionnaires or amendments, that become new tasks to be executed.
 

### 2013-12-13 JournalToSubmit2013/intro.tex
Introduction new cleaned version ...

@@
-Service oriented computing~\cite{Papazoglou2007} is at the origin of an evolution in the field of software development.
-Service oriented methods advocates for the construction of software systems formed by the composition of heterogeneous, loosely coupled modules.
-These modules (or services) communicate in order to achieve a common purpose.
+%Service oriented computing is at the origin of an evolution in the field of software development.
+%Service oriented methods advocates for the construction of software systems formed by the composition of heterogeneous, loosely coupled modules.
+%These modules (or services) communicate in order to achieve a common purpose.
 
+%(Moreover, IT systems need to evolve according to the business needs.)
+%Thus, organizations are seeking for mechanisms to bridge the gap between the actually developed systems and their business needs~\cite{bell}.
+
+In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
+combined to produce applications and provide a business logic.
+The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
 An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually being developed.
-(Moreover, IT systems need to evolve according to the business needs.)
-Thus, organizations are seeking for mechanisms to bridge the gap between the actually developed systems and their business needs~\cite{bell}.
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, claiming that they are the cornerstone in the development of meaningful service based applications~\cite{5}.
 
-In Service-Oriented Computing, pre-existing services are
-combined to produce applications and provide the business logic.
-The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}.
-(Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.)
-Functional properties are not the only crucial aspect in the software development process.
-Other properties need to be addressed to fit in the application with its context.
-These other aspects are called Non-Functional Properties.
+Functional properties are not the only  aspect in the software development process.
+Other non-functional properties need to be addressed to fit in the application with its context (e.g., data privacy, exception handling, atomicity, data persistence).
+Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
+Most methods integrate them only after the application has been implemented.
+This leads to service based applications that are partly specified and, thereby, partialy compliant with the requirements of the application.
 
-Non-functional aspects of the services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
-Most methods consider them only after the application has been implemented, in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence).
-This leads to service based applications that are partly specified and, thereby, partly compliant with the requirements of the application.
 Ideally, non-functional requirements should be considered along with all the stages of the software development.
 The adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that are capable of dealing with
+can help the developer to produce applications that  can deal with
 the application context.
 
-\bigskip
-
-Model Driven Development (MDD) is approach for the development of software systems mainly characterised by the use of models as a product \cite{Selic03}.
+The literature stresses the need for methodologies and techniques for service oriented analysis and design, since they are the cornerstone in the development of meaningful service based applications~\cite{5}.
+Existing approaches argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
+Model Driven Development (MDD)  for the development of software systems mainly characterised by the use of models as a product \cite{Selic03}.
 The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}.
 MDA states a set of guidelines for structuring the specifications of software systems.
-The technique advocates for the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}).
+The technique promotes  the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}).
 These models are successively refined from abstract specifications into actual computer programs.
 
-In this context, we argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
-
-In this work, we are interested in the extension of the \textit{Service Oriented Development Method} (SOD-M)~\cite{decastro1}, to support non-functional aspects, from the early stages of software development.
-SOD-M is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications.
-SOD-M does not provide support for the specification of non-functional requirements, such as
-security, reliability, and efficiency.
-
-The main goals of our work are:
+This paper proposes  extension of the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
+%~\cite{decastro1}, 
+to support non-functional aspects of service-oriented applications, from the early stages of software development.
+$\pi$-SODM is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications. It is intended:
+%SOD-M does not provide support for the specification of non-functional requirements, such as
+%security, reliability, and efficiency.
 \begin{trivlist}
-\item \textit{(i)} To define a NFR model including a set of concepts need for the modeling of NFR in service-oriented applications.
-\item \textit{(ii)} To propose a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\item \textit{(iii)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\item \textit{(iv)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models;
+\item \textit{(i)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
+\item \textit{(ii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
 \end{trivlist}
 
-The rest of the paper is organized as follows.
+The main contributions of our work are:
+\begin{trivlist}
+\item \textit{(i)} An NFR model including a set of concepts need for the modeling of NFR in service-oriented applications.
+\item \textit{(ii)} A methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
+\end{trivlist}
+Accordingly, the remainder of the paper is organized as follows.
 Section\dots

### 2013-12-13 JournalToSubmit2013/SOD-M.tex
Intro mas clean que antes ... y related y Pi-SODM en chantier ..

@@
-The methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements presented in this work extends the \textit{Service Oriented Development Method} (SOD-M) proposed by De Castro et al.\cite{decastro1}.
+%\textsc{Explain these two here --Martin.}
 
-SOD-M define a model-driven approach for development service-based applications.
-The model-driven method provided by SOD-M is based on the Model Driven Architecture (MDA) proposed by the OMG \cite{miller}.
-MDA defines a top-down approach for the development of software systems that uses the notion of \textit{model} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
+%Models at the PIM level describe the structure of the application flow,
+%while, the PSM level provides transformations towards more specific platforms.
+%The PIM-level models proposed by SOD-M are:
+%\begin{trivlist}
+%\item \textit{use case}:
+%\item \textit{extended use case}:
+%\item \textit{service process}:
+%\item \textit{service composition}.
+%\end{trivlist}
 
-\begin{trivlist}
-\item \textbf{Computation Independent Models (CIM):} This level focusses on the
-environment of the system, as well as on its business and requirement specifications.
-This viewpoint represents the software system at its highest level of abstraction.
-At this moment of the development, the structure and system processing details are still unknown or undetermined.
-
-\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform.
-The specification defines those parts of the system that do not change from one platform to another.
-
-\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
-\end{trivlist}
-
-Besides the notion of model at each level of abstraction, MDA propose the use of \textit{model transformations} within and between levels.
-Intra-level transformations are used to provide a unified representation of concepts of a given level.
-Inter-level transformation implement a refinement process between levels.
-Transformations may be automatic or semi-automatic.
-
-According to MDA guidelines, SOD-M meta-models are organized into three levels: CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-
-Two models are defined at the CIM level: \textit{value model}
-and \textit{BPMN model}.
-\textsc{Explain these two here --Martin.}
-
-Models at the PIM level describe the structure of the application flow,
-while, the PSM level provides transformations towards more specific platforms.
-The PIM-level models proposed by SOD-M are:
-\begin{trivlist}
-\item \textit{use case}:
-\item \textit{extended use case}:
-\item \textit{service process}:
-\item \textit{service composition}.
-\end{trivlist}
-
-SOD-M defines three meta-models at the PSM level: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}.
-These three levels have no support for describing non-functional requirements.
-
-The SOD-M approach includes transformations between models:
-\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
-an abstract model at the CIM level, it is possible to apply transformations for
-generating a model of the PSM level. In this context, it is necessary to
-follow the process activities described by the methodology.
-
-SOD-M considers two points of view:
-\textit{(i)} \textit{business}, focusing on the characteristics and requirements
-of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
-features and processes to be implemented in order application requirements. In
-this way, SOD-M aims to simplify the design of service-oriented applications, as
-well as its implementation using current technologies.
-
-{\LARGE \color{red} Relocate the next subsections or delete them --Genoveva and Martin}
+%SOD-M defines three meta-models at the PSM level: \textit{web service interface}, \textit{extended composition service} and \textit{business logic}.
+%These three levels have no support for describing non-functional requirements.
+%
+%The SOD-M approach includes transformations between models:
+%\textit{CIM-to-PIM, PIM-to-PIM} and \textit{PIM-to-PSM} transformations. Given
+%an abstract model at the CIM level, it is possible to apply transformations for
+%generating a model of the PSM level. In this context, it is necessary to
+%follow the process activities described by the methodology.
+%
+%SOD-M considers two points of view:
+%\textit{(i)} \textit{business}, focusing on the characteristics and requirements
+%of the organization, and \textit{(ii)} \textit{system requirements}, focusing on
+%features and processes to be implemented in order application requirements. In
+%this way, SOD-M aims to simplify the design of service-oriented applications, as
+%well as its implementation using current technologies. 
+%
+%{\LARGE \color{red} Relocate the next subsections or delete them --Genoveva and Martin}
 
@@
 %
-Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
-%This method is described in the following sections.
-In order to do, our work organizes non-functional constraints into 
-three layers representing: application modeling, services composition and services.
+%Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition.
+%%This method is described in the following sections.
+%In order to do, our work organizes non-functional constraints into 
+%three layers representing: application modeling, services composition and services.
 The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements.
-At the application layer NFP can refer to business rules and values.
+%At the application layer NFP can refer to business rules and values.
 A value NFR expresses constraints about the way data and functions can be accessed and executed.
 For example accessing methods under security protocols.
-
-Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
+%
+%Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
 For example, response time, storage capacity (e.g., Dropbox service provides 5Giga free storage).
@@
 For example, accessing to a function within an authentica-tion protocol.
-
-Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
+%
+%Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
 For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
-
-As a first step in our approach,  we started modeling non-functional constraints at the PSM level.
+%
+%As a first step in our approach,  we started modeling non-functional constraints at the PSM level.
 Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).

### 2013-12-13 JournalToSubmit2013/modelingWithPISODM.tex
Intro mas clean que antes ... y related y Pi-SODM en chantier ..

@@
 In this section we present $\pi$SOD-M, an MDD based methodology.
+According to MDA guidelines, $\pi$-SODM meta-models are organized into three levels: CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+
+Two models are defined at the CIM level: \textit{value model}
+and \textit{BPMN model}.
+
 $\pi$SOD-M provides an environment for building service compositions.
@@
 
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
+Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.
+Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach.
+Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+\begin{figure} [htpb]
+\centering
+\includegraphics[width=0.65\textwidth]{figs/SODM}
+\caption{SOD-M development process}
+\label{fig:sodm}
+\end{figure} 
+As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}.
+Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
+and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
+
+%Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
 $\pi$SOD-M proposes a development process based on the definition of models (instances of the meta-modes) and transformations between models.
@@
 
-We extend SOD-M to include non-functional specifications.
+%We extend SOD-M to include non-functional specifications.
 Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
@@
 This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
+
+%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.

### 2013-12-13 JournalToSubmit2013/relworks.tex
Intro mas clean que antes ... y related y Pi-SODM en chantier ..

@@
 The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
+
+SOD-M define a model-driven approach for development service-based applications.
+The model-driven method provided by SOD-M is based on the Model Driven Architecture (MDA) proposed by the OMG \cite{miller}.
+MDA defines a top-down approach for the development of software systems that uses the notion of \textit{model} to specify a software system at different levels of abstraction (called \textit{viewpoints}):
+
+\begin{trivlist}
+\item \textbf{Computation Independent Models (CIM):} This level focusses on the
+environment of the system, as well as on its business and requirement specifications.
+This viewpoint represents the software system at its highest level of abstraction.
+At this moment of the development, the structure and system processing details are still unknown or undetermined.
+
+\item \textbf{Platform Independent Models (PIM):} This level focusses on the system functionality, hiding the details of any particular platform.
+The specification defines those parts of the system that do not change from one platform to another.
+
+\item \textbf{Platform Specific Models (PSM):} This level focusses on the functionality, in the context of a particular implementation platform.
+Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
+\end{trivlist}
+
+Besides the notion of model at each level of abstraction, MDA propose the use of \textit{model transformations} within and between levels.
+Intra-level transformations are used to provide a unified representation of concepts of a given level.
+Inter-level transformation implement a refinement process between levels.
+Transformations may be automatic or semi-automatic.
+
+The methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements presented in this work extends the \textit{Service Oriented Development Method} (SOD-M) proposed by De Castro et al.\cite{decastro1}.

### 2013-12-13 JournalToSubmit2013/intro.tex

@@
 This leads to service based applications that are partly specified and, thereby, partialy compliant with the requirements of the application.
+
+Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call a service in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
+% song title disseminated in the walls of the user's Facebook and Twitter accounts.
+
+Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
+Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
+Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
+Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
+In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
 
@@
 
-The main contributions of our work are:
-\begin{trivlist}
-\item \textit{(i)} An NFR model including a set of concepts need for the modeling of NFR in service-oriented applications.
-\item \textit{(ii)} A methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements;
-\end{trivlist}
+The main contribution of our work is a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements.
+The methodology is based on  an NFR model including a set of concepts need for the modeling of NFR in service-oriented applications.
+
 Accordingly, the remainder of the paper is organized as follows.

### 2013-12-13 JournalToSubmit2013/modelingFlyingPig.tex

@@
-We validated our methodology by in a use-case concerning risk assessment for financial companies, inspired on the ORCA System\footnote{The ORCA System is a trademark of GCP Global www.gcpglobal.com.}.
+We validated our methodology by developing a use-case concerning risk assessment for financial companies.
+Our application is inspired on the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
 Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied the client's business practice.
-For instance, the condition in which confidential transactions are performed, the physical security for accessing reserved areas (i.e. access to the critical servers room should be secured).
-This information is used to determine whether there are risky actions within the business processes of the company that can be amended.
-It is also used to determine a degree of compliance to existing norms.
+For instance, the conditions and protocols used to perform confidential transactions, the physical security for accessing reserved areas such as computing server installations.
+The information gathered by the questionnaires is used to determine whether there are risky practices within the business processes of the company, as well as to propose amends to these practices.
+The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
 By analysing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
 
-Our goal is to model a service based application (called \textsl{FlyingPig}), for providing risk assessment as a service.
-In order to provide this functionality \textsl{FlyingPig} would benefit from ORCA's legacy services: storage, assessment and data visualization functions.
+Our goal is to model a service based application (called \FlyingPig), for providing risk assessment as a service.
+In order to provide this functionality, \FlyingPig\ would benefit from ORCA's legacy services: storage, assessment and data visualization functions.
 
-In the following, we describe the results of applying $\Pi$-SODM to develop the \textsl{FlyingPig} risk assessment system.
-The models presented next are the final results of the application of $\Pi$-SODM.
-They were generated as a result of interactions with GCP Global software developers.
+In the following, we describe the results of applying $\Pi$-SODM to develop the \FlyingPig\ risk assessment system.
+The models presented next were generated as a result of interacting with software developers at GCP Global.
 
@@
 \centering
-\includegraphics[width=0.7\textwidth]{figs/E3-Value.pdf}
-\caption{E3value model of the FlyingPig.\label{fig:E3valuemodel}}
+\includegraphics[width=0.7\textwidth]{figs/3ValueModel.pdf}
+\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
+\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
 \end{figure}
 
-\subsection{CIM}
+\subsection{Computation-Independent Models (CIM)}
 
-$\Pi$-SODM uses two models, to be built at the CIM level (See Section~\ref{sec:modelingWithPISODM}).
+$\Pi$-SODM uses two models, to be built at the CIM level (See Section~\ref{sec:modelingWithPISODM}): The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
 
-Figure~\ref{fig:E3valuemodel} shows the value model~\cite{e3value} for the case of \textsl{FlyingPig}.
-The value model is a business model that represents a business case graphically as a set of value exchanges ($\bigtriangleup\ \bigtriangledown$) and value activities (rounded boxes) performed by business actors (squared boxes).
+Figure~\ref{fig:E3valuemodel} shows the value model for the case of \textsl{FlyingPig}.
+The value model is a business model that represents a business case graphically as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
 
 In our use-case, we identify two business actors: \textsl{ORCA} and \textsl{Broker}. 
-Brokers are the responsible for channeling requests for risk assessment of one or several companies.
-ORCA have two value activities which are services that provide it an economical benefit: to \textsl{identify amendments} and the possibility to \textsl{asses risk situation}.
-The value exchanged between ORCA and Brokers are: \textsl{amendments} and \textsl{evaluation reports} which are value objects ([...]) for the any companies that need to have a risk assessment, and the \textsl{questionnaire and evidences} and the risk assessment \textsl{fee} which are value objects for ORCA System.
+Brokers are responsible for channelling requests for risk assessment of one or several companies. 
+ORCA have two value activities which are services that provide an economical benefit: to \textsl{Identify Amendments} and the possibility to \textsl{Asses Risk Situation}. 
+The values exchanged between ORCA and the brokers are: \textsl{Amendments} and \textsl{Evaluation Reports} which are value objects ([ \!\dots]) for the companies that need to have a risk assessment, as well as  \textsl{Questionnaire and Evidences} and the risk assessment \textsl{Fee} which are value objects for ORCA System.
 
-Beside the static representation of a business case, the value model allows for the representation of dependency paths which enhance the understanding of a business idea by showing all value exchanges triggered by the occurrence of an end-consumer need.
+The e3value model is a static representation of a business case.
+The model defines \textit{dependency paths}, which show that the value exchanges are triggered by the occurrence of an end-consumer need. 
+A dependency path has a direction and consists of a sequence of linked nodes. 
+A dependency node is a start stimulus, an AND-fork or AND-join, an OR-fork or OR-join, or an end node (see Legend on Figure~\ref{fig:E3valuemodel}).
+An start stimulus represents a trigger for the exchange of economic value objects while an end node represents a model boundary.
+
+Besides the static representation of a business case, the value model allows for the representation of dependency paths which enhance the understanding of a business idea by showing all value exchanges triggered by the occurrence of an end-consumer need.
 A dependency path has a direction and consists of dependency nodes linked by mean of connections.

### 2013-12-13 JournalToSubmit2013/modelingFlyingPig.tex
no message

@@
 The e3value model is a static representation of a business case.
-The model defines \textit{dependency paths}, which show that the value exchanges are triggered by the occurrence of an end-consumer need. 
-A dependency path has a direction and consists of a sequence of linked nodes. 
-A dependency node is a start stimulus, an AND-fork or AND-join, an OR-fork or OR-join, or an end node (see Legend on Figure~\ref{fig:E3valuemodel}).
-An start stimulus represents a trigger for the exchange of economic value objects while an end node represents a model boundary.
+The model defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
+A dependency path has a direction and consists of a sequence of linked dependency nodes.
+A dependency path startr with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Legend on Figure~\ref{fig:E3valuemodel}).
 
-Besides the static representation of a business case, the value model allows for the representation of dependency paths which enhance the understanding of a business idea by showing all value exchanges triggered by the occurrence of an end-consumer need.
-A dependency path has a direction and consists of dependency nodes linked by mean of connections.
-A dependency node is a start stimulus, an AND-fork or AND-join, an OR-fork or OR-join, or an end node (see Legend on Figure~\ref{fig:E3valuemodel}).
-An start stimulus represents a trigger for the exchange of economic value objects while an end node represents a model boundary.
+AQUI!!!
 

### 2013-12-13 JournalToSubmit2013/modelingFlyingPig.tex
no message

@@
 
-$\Pi$-SODM uses two models, to be built at the CIM level (See Section~\ref{sec:modelingWithPISODM}): The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
+$\Pi$-SODM uses two models at the CIM level (see Section~\ref{sec:modelingWithPISODM}): The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
+The former is a simple model just to identify the transference of value information between components of the system.
+The BPMN model establishes which are the actors and main tasks of the application.
 
-Figure~\ref{fig:E3valuemodel} shows the value model for the case of \textsl{FlyingPig}.
-The value model is a business model that represents a business case graphically as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
+Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
+It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
 
@@
 Brokers are responsible for channelling requests for risk assessment of one or several companies. 
-ORCA have two value activities which are services that provide an economical benefit: to \textsl{Identify Amendments} and the possibility to \textsl{Asses Risk Situation}. 
+ORCA have two value activities which are services that provide an economical benefit: to \textsl{Identify Amendments} and the possibility to \textsl{Assess Risk Situation}. 
 The values exchanged between ORCA and the brokers are: \textsl{Amendments} and \textsl{Evaluation Reports} which are value objects ([ \!\dots]) for the companies that need to have a risk assessment, as well as  \textsl{Questionnaire and Evidences} and the risk assessment \textsl{Fee} which are value objects for ORCA System.
 
-The e3value model is a static representation of a business case.
-The model defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
+The e3value defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
 A dependency path has a direction and consists of a sequence of linked dependency nodes.
-A dependency path startr with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Legend on Figure~\ref{fig:E3valuemodel}).
+A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Legend on Figure~\ref{fig:E3valuemodel}). 
+Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
 
-AQUI!!!
+The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment by a particular company. 
+Once this need occurs, the value exchanges between ORCA and Broker are triggered. 
+The client company will provide orca wit information (answers to a questionary), evidence (to support the information) and a fee (monetary value).
+ORCA will provide amendments (recommendations to change practices) and an evaluation report.
 
-Figure~\ref{fig:E3valuemodel} shows the dependency paths for \textsl{FlyingPig}.
-The dependency path identified in that case is the need of assessment of a particular company.
-It denotes a stimulus caused by the client company who ask for risk assessment.
-Once this need occurs the value exchanges between ORCA and Broker are triggered.
-The value model represent that ORCA will provide amendments and an evaluation report in return for the questionnaire and evidences and the fees provided by the Broker.
-
-{\color{blue} Javier: Please modify the value model to include fee as a value object in return for evaluation report. Modified figure is on the git}
-
-A better understanding of the process in which these value exchanges occurs may be modeled using a BPMN model.
-Figure~\ref{fig:bpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} for the scenario.
-It includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
+The BPMN model is devised to better understand the process in which the value exchanges occur.
+Figure~\ref{fig:BPMNmodel} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.} for the \FlyingPig\ scenario. 
+The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}. 
 Brokers have two lanes, the client \textsl{Company} and a \textsl{User}. 
@@
 
-The process starts by a request from a company asking for a risk assessment.
-Note that it is in agreement with the explanation of the start stimulus on the value model above.
+The risk assessment process starts after a request from a company.
+This agrees with the value model, in which the start stimulus triggers the whole process. 
 The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
 Questionnaires are considered tasks that users will have to perform. 
-Given a list of tasks, the company assign a responsible and delegate the task.
-A task may be to answer a questionnaire or add an evidence but also could implies to solve some "risky situation" (e.g.
+Given a list of tasks, the company delegates the task to a user.
+A task may be to answer a questionnaire, add an evidence as well as to solve a ``risky situation'' (e.g.
 to repair a door {\color{blue} ...¿Genoveva mira por favor si esto puede ser o hay algun ejemplo mas bueno/serio???} ).
@@
 \includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
-\caption{BPMN model for FlyingPig.\label{fig:BPMNmodel}}
-\label{fig:bpmn}
+\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{figs/UseCaseGeneral.png}
-\caption{$\pi$-UseCase model for FlyingPig.\label{fig:piUseCaseModel}}
-\label{fig:bpmn}
+\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneral.png}
-\caption{$\pi$-ServiceProcess model for FlyingPig.\label{fig:PiServiceProcessModel}}
-\label{fig:bpmn}
+\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
 \end{figure}
@@
 \includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneral.png}
-\caption{$\pi$-ServiceComposition model for FlyingPig.\label{fig:PiServiceCompositionModel}}
-\label{fig:bpmn}
+\caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
 \end{figure}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+%% By Placido...
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Pi-UseCase Model description for FlyingPig
+
+The piUseCase model shown in Figure  X models the use cases (features) and constrainsts for the FlyingPig application.
+3 actors are identified in this model, the are: (I) Company, (II) User and the (III) Channel-Broker.
+The Company is the whom ask for risk evaluation, the (II) Channel-Broker is the responsible for managing a defining the evaluation flow, beyond create responsible for evaluating and delegate the tasks.
+This model also represents the (III) User actor which answers questionnaires for the assessment, including evidence and proceeding with necessary corrections when asked for this type of task.
+
+Each actor is associated with use cases.
+Use cases represent the actions that the actors are responsible.
+Altogether, this model defines 6 (six) use cases.
+The use cases related to the Company's actor is (i) request assessment.
+The use cases related to Broker-Channel's actor are: (ii) create a responsable and (iii) delegate task; and the use cases related User's actor are: (iv) answer questionnaire and add evidences, (v) register task, and (vi) perform amendment.
+
+In this model, each use case may be related with one or more constraints.
+The constraints can be of three types: value , business or exceptional behavior .
+This model as a whole has 7 constraints .
+The use case (i) request assessment has 2 constraints related with, the are from the value type.
+One is related to ( i.a) response time (confirmation) that should be up to 30 seconds , and the other ( i.b ) describes that the system should support a large number of requests from different companies .
+Moreover ( i.c ) if the number of requests exceeds 200 , FlyingPig should make a load balance of the requests.
+The use case create a responsable has a value constraint ( ii.d ) which defines that the Channel-Broker must have privileges verified before the execution to perform this action.
+This use case also ( ii.c ) is also related with the load balancing verification.
+The use case ( iii ) delegate task has no associated constraint .
+The use case answer questionnaire and add evidences have 2 constraints , they are: the User which performs this action ( iv.e ) must have privileges to answer the questionnaires and ( iv.f ) the time limit must be less than 15 days.
+The use case register task has 3 constraints associated .
+Two of them are the same constraints related with the request assessment use case, where (v.a) need a response time confirmation that should be up to 30 seconds, and the other ( v.b ) describes what the system should support a large number of requests for this action, once Users from various companies will be answering the questionnaire and performing tasks.
+The other constraint requires that ( v.g ) throughout the survey all the questions must be properly answered .
+Finally , the use case ( vi.f ) perform amendment must be executed in a set time limit in days, the same constraint related with the use case questionnaire answer and add evidences.
+Figure A presents the relation between the use case and its constraints.
+
+TABLE A - RELATION BETWEEN USE CASE AND CONSTRAINT
+use cases / constraints	(i) request assessment	(ii) create e responsable 	(iii) delegate task	(iv) answer questionnaire and add evidences	(v) register task	(vi) perform amendment
+(a) confirmation must be received in up to 30sec	X				X	
+(b) requires concurrency for more than 200 users	X				X	
+(c) load balance if there is more than 200 requests	X	X				
+(d) requires Broker’ privileges		X				
+(e) requires User' privileges				X		
+(f) all tasks must be done within a given time limit (e.g. maximum of 15 days)				X		X
+(g) requires questionnaire completion					X
+
+Pi-ServiceProcess Model description for FlyingPig
+
+The pi-ServiceProcess model (Figure Y) represents the application execution flow of FlyingPig.
+In this model the action are the result of the use case transformation identified in pi-UseCase model according to their respective actors.
+The Company, Broker-Channel and User actors are transformed into rays that represent the business collaborators.
+
+The restrictions associated with each use case are transformed into assertions.
+The set of assertions related to a particular action form the contract for this action.
+The process of FlyingPig executing begins with the action ( i ) request assessment performed by the Company.
+This action has one pre-condition regarding the response time and a post-condition concerning the load balance control for the server when it reaches 200 requests.
+Once the request is made by the Company for risk analysis, the Broker ( ii ) creates responsible for its analysis and ( iii ) delegate the necessary risk assessment.
+These actions are executed in sequence and has a pre-condition concerning the Broker's privilege.
+This can only create responsible and delegate the tasks if their authentication datas are correct and if and there was less than 3 requests.
+The tasks to be performed by the User can be ( iv ) answer questionnaire and add evidences; or ( v) perform amendments.
+Both the assertion has one pre-condition concerning the time limit for form submission.
+In this example, we use the limit of 15 days for form submission.
+If the User is performing the task of answer questionnaire and add evidences, this will have to enter with their authentication data.
+As postcondition, it is checked if all the answers were answered .
+The action of perform amendments has no additional verification.
+Finally, the user registers the tasks performed.
+This action has response time verification.
+
+This model refines the concepts defined in the pi-UseCase model in order to better describe the assertions and groups them into contracts.
+
+Pi-ServiceComposition Model description for FlyingPig
+
+This model (presented in Figure Z) describes the interactions of the actions described in pi-ServiceProcess model with the external services.
+Services are provided through the ORCA system and the access is accomplished by FlyingPig interface.
+FlyingPig is the point of interaction between the actions described in the precess with the services offered by ORCA.
+FlyingPig provides an interface with five actions, they are: receive a request, generate new interface, create questionnaire, notify responsable e receive answers..
+This interface makes the direct calls for the ORCA' services.
+The services offered by ORCA are: create space company, generate the questionnaire, analyze answer, data store, call for amendment and create report.
+These services are called according to the process execution.
+
+With respect to non-functional requirements, the assertions grouped into contract in the pi-ServiceProcess model are transformed into policies in the pi-ServiceComposition model.
+The policies identified in this case study are: security, performance and conformability policies.
+Each policy is associated with specific the services and it is verified at the service execution time regarding the specific rule which the sercive is associated with..
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Lessons Learned}

### 2013-12-13 JournalToSubmit2013/modelingWithPISODM.tex
pi sodm mas limpio pero undergoing

--- 2013-12-06 JournalToSubmit2013/modelingWithPISODM.tex
@@
 
-In this section we present $\pi$SOD-M, an MDD based methodology.
-$\pi$SOD-M provides an environment for building service compositions.
-$\pi$SOD-M includes the modelling of non-functional requirements at the early stages of the development.
-The methof extends the SOD-M meta-models by adding the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a} to represent non-functional requirements.
+In this section we present $\pi$SOD-M, an MDD based methodology for building service compositions with non-functional requirements.
+$\pi$SOD-M provides concepts for  modelling  non-functional requirements at the early stages of the development.
+%The method uses the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a} to represent non-functional requirements.
 
@@
 
+As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM meta-models are organized into three levels: 
+\begin{itemize}
+\item CIM (\textit{Computational Independent Models}). Two models are defined at the CIM level: \textit{value model}
+and \textit{BPMN model}.
+Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
+and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
+
+ \item PIM (\textit{Platform Independent Models}). Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, and \textit{$\pi$-ServiceCom\-po\-si\-tion}.
+%and \textit{$\pi$-PEWS}.
+
+ The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.Non-functional requirements are defined as \textit{constraints} over actions and data.
+
+The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
+%that must be performed upon certain conditions.
+%The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
+described in the \textit{$\pi$-UseCase}.
+%model into contracts that are associated
+%with services.
+
+The \textit{$\pi$-ServiceComposition} meta-model provides the concept  \textit{Policy}
+to integrate contracts with similar non-functional requirements.
+For instance, security and privacy restrictions may be grouped into a security policy.
+%\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
+
+  \item PSM (\textit{Platform Specific Models}).
+At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
+The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions.
+
+\end{itemize}
+
 $\pi$SOD-M (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
 $\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints.
-Constraints are restrictions that must be verified during the execution of the application.
-An example of this is the requirement of the user's authentication for executing some system functions.
-
-Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
-$\pi$SOD-M proposes a development process based on the definition of models (instances of the meta-modes) and transformations between models.
-There are two kinds of transformations: Model-to-model transformations are used during the software process to refine the specification.
-Model-to-text transformations are the last step of the process and generate code.
-
-We extend SOD-M to include non-functional specifications.
-Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, \textit{$\pi$-ServiceCom\-po\-si\-tion} and \textit{$\pi$-PEWS}.
-The former three are extensions of SOD-M meta-models and belong to the PIM level.
-The \textit{$\pi$-PEWS} meta-model is a PSM.
-
-The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over processing and data.
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions.
-The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase} model into contracts that are associated
-with services.
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}
-which put together contracts with similar non-functional requirements.
-For instance, security and privacy restrictions may be grouped into a security policy.
-\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
-
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
 \textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
 This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
+
+Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+
+%Constraints are restrictions that must be verified during the execution of the application.
+%An example of this is the requirement of the user's authentication for executing some system functions.
+%
+
+%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.

### 2013-12-16 JournalToSubmit2013/modelingFlyingPig.tex

@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
+
+{\color{red} Javier: Please change \underline{Asses} by \underline{Assess}. --M}
+\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
+\end{figure}
+
 The BPMN model is devised to better understand the process in which the value exchanges occur.
@@
 The risk assessment process starts after a request from a company.
-This agrees with the value model, in which the start stimulus triggers the whole process. 
+(This agrees with the value model, in which the start stimulus triggers the whole process.)
 The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
 Questionnaires are considered tasks that users will have to perform. 
-Given a list of tasks, the company delegates the task to a user.
-A task may be to answer a questionnaire, add an evidence as well as to solve a ``risky situation'' (e.g.
-to repair a door {\color{blue} ...¿Genoveva mira por favor si esto puede ser o hay algun ejemplo mas bueno/serio???} ).
+Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include from physical facts such as not having easy access to handicapped persons or having an unsecured access to the premises of the company, to more intangible ones, such as the use of an less-than-optimal protocol to access data on the company's computer server.}.
 
-Once a tasks are completed they are stored and analyzed to generate a set of un-compliant situations associated to their corresponding amendments if there are any or a report specifying a compliance level, incidents and a risk map.
-During the process of analysing a questionnaire, the answers of some questions can trigger the generation of other questionnaires or amendments, that become new tasks to be executed.
+Once tasks are completed, they are stored and analysed to generate a list of un-compliant situations, associated to their corresponding \textit{calls for amendment}, in case there are any, or a report specifying a compliance level, incidents and a risk map.
+During the process of analysing a questionnaire, the answers of some questions can trigger the generation of other questionnaires or amendments, that will become new tasks.
 
@@
 In our example we have some constraints.
+
+{\color{red}
+Placido, Valeria: What can we say about non-functional requirements at the CIM level?
+}
+
 \begin{itemize}
@@
 
+\subsection{Platform-Independent Models (PIM)}
+
+In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
+These models are built next, for the \FlyingPig\ scenario.
+
 \begin{figure}[t]
 \centering
-\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
-\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
+\includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
+
+{\color{red} 
+``Create a responsible'' should be changed for ``Designate user in charge''
+
+``requires concurrency for more than 200 users'' should be changed by ``Maximum number of users should be greater than 200''
+}
+\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
 \end{figure}
 
-\subsection{PIM}
+\paragraph{\underline{$\pi$-UseCase Model for \FlyingPig}}~
 
-{\color{magenta} Placido and Umberto: Could you please comment models here? --G \& M.}
+The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application. 
+In this model, three actors are identified: \textit{Company}, \textit{User} and \textit{Channel-Broker}\footnote{\color{red} Can we change this for Broker ? --M.}. 
+They are represented as stick figures.
+In the context of \FlyingPig, Company is the actor asking for risk evaluation.
+A Channel-Broker is the responsible for channelling the the evaluation process, assigning users to be in charge of tasks as well as delegating tasks. 
+A User, in this model is an actor who answers questionnaires (with base on the actual facts about the Company).
+The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
 
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/UseCaseGeneral.png}
-\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
-\end{figure}
+Each actor is associated to one or more use cases (depicted as white ovals in Figure~\ref{fig:piUseCaseModel}). 
+Use cases describe the main functionalities of the system.
+The $\pi$-UseCase model for \FlyingPig\ defines six use cases.
+
+In our model, each use case may be associated to one or more (non-functional) constraints (depicted as coloured ovals in Figure~\ref{fig:piUseCaseModel}). 
+The model defines three types of constraints: \textit{value} , \textit{business} or \textit{exceptional behavior}. 
+Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
+
+In the case of \FlyingPig, the model counts seven constraints:
+The use case \textsf{request assessment} has two value constraints.
+One of them requires a response time of less than 30 seconds while the other requires that the system's infrastructure should be prepared to deal with, at least 200 users.
+
+PARE AQUI --Martin
+
+Moreover, an associated constraint requires that if the number of requests exceeds 200 , FlyingPig should make a load balance of the requests.
+The use case create a responsible has a value constraint ( ii.d ) which defines that the Channel-Broker must have privileges verified before the execution to perform this action.
+This use case also ( ii.c ) is also related with the load balancing verification.
+The use case ( iii ) delegate task has no associated constraint .
+The use case answer questionnaire and add evidences have 2 constraints , they are: the User which performs this action ( iv.e ) must have privileges to answer the questionnaires and ( iv.f ) the time limit must be less than 15 days.
+The use case register task has 3 constraints associated .
+Two of them are the same constraints related with the request assessment use case, where (v.a) need a response time confirmation that should be up to 30 seconds, and the other ( v.b ) describes what the system should support a large number of requests for this action, once Users from various companies will be answering the questionnaire and performing tasks.
+The other constraint requires that ( v.g ) throughout the survey all the questions must be properly answered .
+Finally , the use case ( vi.f ) perform amendment must be executed in a set time limit in days, the same constraint related with the use case questionnaire answer and add evidences.
+Figure A presents the relation between the use case and its constraints.
 

### 2013-12-16 JournalToSubmit2013/modelingFlyingPig.tex
no message

@@
 
-{\color{red} 
-``Create a responsible'' should be changed for ``Designate user in charge''
+{\color{red} \raggedright
+$\star$ ``Create a responsible'' should be changed for ``Designate user in charge''.
 
-``requires concurrency for more than 200 users'' should be changed by ``Maximum number of users should be greater than 200''
+$\star$ ``requires concurrency for more than 200 users'' should be changed by ``Maximum number of users should be greater than 200''.
 }
@@
 In the case of \FlyingPig, the model counts seven constraints:
-The use case \textsf{request assessment} has two value constraints.
-One of them requires a response time of less than 30 seconds while the other requires that the system's infrastructure should be prepared to deal with, at least 200 users.
+\begin{numtrivlist}
+\item An acknowledgement is due less than 30 seconds after registering a task or demand for assessment. 
+\item The system's infrastructure should be prepared to deal with, at least, 200 users. 
+\item If the number of requests exceeds 200, \FlyingPig\ should make a load balance of the requests. 
+\item The Channel-Broker must have their privileges verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} use case.
+\item Users must have their privileges verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} use case.
+\item All questionnaires need to be fully answered, in order to consider that a task is completed.
+\item There is a time limit (in days) for each amendment required by the system.
+\end{numtrivlist}
 
-PARE AQUI --Martin
-
-Moreover, an associated constraint requires that if the number of requests exceeds 200 , FlyingPig should make a load balance of the requests.
-The use case create a responsible has a value constraint ( ii.d ) which defines that the Channel-Broker must have privileges verified before the execution to perform this action.
-This use case also ( ii.c ) is also related with the load balancing verification.
-The use case ( iii ) delegate task has no associated constraint .
-The use case answer questionnaire and add evidences have 2 constraints , they are: the User which performs this action ( iv.e ) must have privileges to answer the questionnaires and ( iv.f ) the time limit must be less than 15 days.
-The use case register task has 3 constraints associated .
-Two of them are the same constraints related with the request assessment use case, where (v.a) need a response time confirmation that should be up to 30 seconds, and the other ( v.b ) describes what the system should support a large number of requests for this action, once Users from various companies will be answering the questionnaire and performing tasks.
-The other constraint requires that ( v.g ) throughout the survey all the questions must be properly answered .
-Finally , the use case ( vi.f ) perform amendment must be executed in a set time limit in days, the same constraint related with the use case questionnaire answer and add evidences.
-Figure A presents the relation between the use case and its constraints.
+{\color{red} Anything else here? --M}
 
@@
 
+\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
+
+The $\pi$-ServiceProcess model (Figure~\ref{fig:PiServiceProcessModel}) presents the workflow for \FlyingPig.
+The actions in this model were obtained by applying the use case transformation rules described in Section~\ref{sec:pewsmetamodel}.
+
+The Company, Broker-Channel and User actors are transformed into lanes that represent the business collaborators.
+Use cases are transformed into \textit{actions} and are represented by white boxes.
+The restrictions associated to each use case are transformed into \textit{assertions} (represented by coloured boxes) and may be decorated with pre- and post-conditions. 
+The set of assertions related to a particular action form a \textit{contract} for this action.
+
+This model refines the concepts defined in the $\pi$-UseCase model and it is more suited to better describe the assertions, by grouping them into contracts.
+
 \begin{figure}
@@
 \end{figure}
+
+\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
+
+This model presented in Figure~\ref{fig:PiServiceCompositionModel} describes the interactions of the actions described in pi-ServiceProcess model with the external services.
+Services are provided through the ORCA system and the access is accomplished by FlyingPig interface.
+FlyingPig is the point of interaction between the actions described in the precess with the services offered by ORCA.
+FlyingPig provides an interface with five actions, they are: receive a request, generate new interface, create questionnaire, notify responsable e receive answers..
+This interface makes the direct calls for the ORCA' services.
+The services offered by ORCA are: create space company, generate the questionnaire, analyze answer, data store, call for amendment and create report.
+These services are called according to the process execution.
+
+With respect to non-functional requirements, the assertions grouped into contract in the pi-ServiceProcess model are transformed into policies in the pi-ServiceComposition model.
+The policies identified in this case study are: security, performance and conformability policies.
+Each policy is associated with specific the services and it is verified at the service execution time regarding the specific rule which the sercive is associated with..
 

### 2013-12-16 JournalToSubmit2013/relworks.tex
main y related works par genoveva

@@
-Related works to our approach includes existing proposal dealing with the inclusion of non-functional properties in service-oriented development.
-Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase.
+We conducted a systematic review \cite{placidoPhDThesis2012} in order to analyze approaches that consider non-functional properties for developing service-oriented applications.
+These  approaches address the problem from three perspectives: NFP expressed as quality of service properties (QoS attributes) that can be used for looking up services that can participate in a service composition, programming NFP as protocols that are asserted to service composition code, and methodologies that provide tools for building service compositions with associated NFP.
+
+%  Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase.
 Finally exiting methodology for the development of service-oriented applications including non-functional properties are here depicted and compared with our approach.
@@
 
-In previous work we have carried out a systematic review \cite{placidoPhDThesis2012} in order to analyze how non-functional properties are taking into account in existing approaches for service-oriented development.
-Existing proposals in this field may be classified in three types of works which are described in the followed.
+The first types of works are thus considering a particular non-functional concern (e.g. security) which is modeled and then "glued" or associated to functional models of an application.
 
-The first types of works are thus considering a particular non-functional concern (e.g.
-security) which is modeled and then "glued" or associated to functional models of an application.
-The work of Chollet et al.
 \cite{CholletL09} defines a proposal to associate non-functional quality properties (security properties in that case) to functional activities which are modeled in a web service composition model.
-Schmeling et al.
+
 \cite{SchmelingCM11} present an approach and also a toolset for specifying and implementing non-functional concerns in web service compositions; non-functional concerns are modeled first in non-functional model and then related to a service composition represented in a BPMN diagram.
-Ovaska et al.
+
  \cite{OvaskaEHPA10} present an approach to support quality management at design time.
 Quality requirements are modeled in a first phase and represented then in an architectural model where quality requirements are associated to some components in the model.
-In our case we propose to model non-functional properties joint to the modeling of functional concerns what simplifies the web service development process.
-Non-functional properties are represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.
 
 Second type of works related to our proposal proposes specifics architecture or frameworks to manage and validate QoS attributes in service composition processes.
-The work of Xiao et al.
+
 \cite{XiaoCZBOLH08} present a framework to verify SLA compliance in composed services.
@@
 Other relevant works such as \cite{DAmbrogio06,Liu20121080} propose extensions to WSDL and BPEL languages respectively for the inclusion of QoS attributes that can be evaluated at runtime.
+
+In our case we propose to model non-functional properties joint to the modeling of functional concerns what simplifies the web service development process.
+Non-functional properties are represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.
 

### 2013-12-16 JournalToSubmit2013/modelingFlyingPig.tex
no message

@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneral.png}
+\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneralCut.png}
 \caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
@@
 
-The Company, Broker-Channel and User actors are transformed into lanes that represent the business collaborators.
+The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
 Use cases are transformed into \textit{actions} and are represented by white boxes.
 The restrictions associated to each use case are transformed into \textit{assertions} (represented by coloured boxes) and may be decorated with pre- and post-conditions. 
-The set of assertions related to a particular action form a \textit{contract} for this action.
+We can see that this model refines the concepts defined in the $\pi$-UseCase model.
+The assertions specify those non-functional requirements, as they are seen by the actors. 
+The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.
 
-This model refines the concepts defined in the $\pi$-UseCase model and it is more suited to better describe the assertions, by grouping them into contracts.
-
-\begin{figure}
+\begin{figure}[t]
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneral.png}
+\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut.png}
+{\color{red} \raggedright
+$\star$ Change FlyingFish by \FlyingPig.
+}
 \caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
@@
 
-This model presented in Figure~\ref{fig:PiServiceCompositionModel} describes the interactions of the actions described in pi-ServiceProcess model with the external services.
-Services are provided through the ORCA system and the access is accomplished by FlyingPig interface.
-FlyingPig is the point of interaction between the actions described in the precess with the services offered by ORCA.
-FlyingPig provides an interface with five actions, they are: receive a request, generate new interface, create questionnaire, notify responsable e receive answers..
-This interface makes the direct calls for the ORCA' services.
-The services offered by ORCA are: create space company, generate the questionnaire, analyze answer, data store, call for amendment and create report.
-These services are called according to the process execution.
+The model presented in Figure~\ref{fig:PiServiceCompositionModel} offers a system-centric view of the application.
+The previous model is now enhanced with a view of the \FlyingPig\ application, including those services upon  it is build.
+The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
+These policies contain the pre- and post-conditions of the previous model.
+They are associated to actions of the system.
 
-With respect to non-functional requirements, the assertions grouped into contract in the pi-ServiceProcess model are transformed into policies in the pi-ServiceComposition model.
-The policies identified in this case study are: security, performance and conformability policies.
-Each policy is associated with specific the services and it is verified at the service execution time regarding the specific rule which the sercive is associated with..
+\begin{figure}[t]
+\centering
+%\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut.png}
+\fbox{\LARGE $\pi$-PEWS Model.}
 
-\subsection{PSM}
+{\color{red} \raggedright
+$\star$ Placido: Please, place the model here.
+}
+\caption{$\pi$-PEWS model for \FlyingPig.\label{fig:PiPEWSModel}}
+\end{figure}
 
-{\color{magenta} Placido and Umberto: Could you please insert the $\pi$-PEWS model here? --G \& M.}
+\subsection{Platform-Specific Model (PSM)}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-%% By Placido...
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The model presented in Figure~\ref{fig:PiPEWSModel} is the result of translating the previous model into service composition code.
+Notice that the workflow for \FlyingPig\ is implemented by using the constructors present in BPEL~\cite{BPEL}, with the addition of the notion of \textit{contract}.
 
-Pi-UseCase Model description for FlyingPig
-
-The piUseCase model shown in Figure  X models the use cases (features) and constrainsts for the FlyingPig application.
-3 actors are identified in this model, the are: (I) Company, (II) User and the (III) Channel-Broker.
-The Company is the whom ask for risk evaluation, the (II) Channel-Broker is the responsible for managing a defining the evaluation flow, beyond create responsible for evaluating and delegate the tasks.
-This model also represents the (III) User actor which answers questionnaires for the assessment, including evidence and proceeding with necessary corrections when asked for this type of task.
-
-Each actor is associated with use cases.
-Use cases represent the actions that the actors are responsible.
-Altogether, this model defines 6 (six) use cases.
-The use cases related to the Company's actor is (i) request assessment.
-The use cases related to Broker-Channel's actor are: (ii) create a responsable and (iii) delegate task; and the use cases related User's actor are: (iv) answer questionnaire and add evidences, (v) register task, and (vi) perform amendment.
-
-In this model, each use case may be related with one or more constraints.
-The constraints can be of three types: value , business or exceptional behavior .
-This model as a whole has 7 constraints .
-The use case (i) request assessment has 2 constraints related with, the are from the value type.
-One is related to ( i.a) response time (confirmation) that should be up to 30 seconds , and the other ( i.b ) describes that the system should support a large number of requests from different companies .
-Moreover ( i.c ) if the number of requests exceeds 200 , FlyingPig should make a load balance of the requests.
-The use case create a responsable has a value constraint ( ii.d ) which defines that the Channel-Broker must have privileges verified before the execution to perform this action.
-This use case also ( ii.c ) is also related with the load balancing verification.
-The use case ( iii ) delegate task has no associated constraint .
-The use case answer questionnaire and add evidences have 2 constraints , they are: the User which performs this action ( iv.e ) must have privileges to answer the questionnaires and ( iv.f ) the time limit must be less than 15 days.
-The use case register task has 3 constraints associated .
-Two of them are the same constraints related with the request assessment use case, where (v.a) need a response time confirmation that should be up to 30 seconds, and the other ( v.b ) describes what the system should support a large number of requests for this action, once Users from various companies will be answering the questionnaire and performing tasks.
-The other constraint requires that ( v.g ) throughout the survey all the questions must be properly answered .
-Finally , the use case ( vi.f ) perform amendment must be executed in a set time limit in days, the same constraint related with the use case questionnaire answer and add evidences.
-Figure A presents the relation between the use case and its constraints.
-
-TABLE A - RELATION BETWEEN USE CASE AND CONSTRAINT
-use cases / constraints	(i) request assessment	(ii) create e responsable 	(iii) delegate task	(iv) answer questionnaire and add evidences	(v) register task	(vi) perform amendment
-(a) confirmation must be received in up to 30sec	X				X	
-(b) requires concurrency for more than 200 users	X				X	
-(c) load balance if there is more than 200 requests	X	X				
-(d) requires Broker’ privileges		X				
-(e) requires User' privileges				X		
-(f) all tasks must be done within a given time limit (e.g. maximum of 15 days)				X		X
-(g) requires questionnaire completion					X
-
-Pi-ServiceProcess Model description for FlyingPig
-
-The pi-ServiceProcess model (Figure Y) represents the application execution flow of FlyingPig.
-In this model the action are the result of the use case transformation identified in pi-UseCase model according to their respective actors.
-The Company, Broker-Channel and User actors are transformed into rays that represent the business collaborators.
-
-The restrictions associated with each use case are transformed into assertions.
-The set of assertions related to a particular action form the contract for this action.
-The process of FlyingPig executing begins with the action ( i ) request assessment performed by the Company.
-This action has one pre-condition regarding the response time and a post-condition concerning the load balance control for the server when it reaches 200 requests.
-Once the request is made by the Company for risk analysis, the Broker ( ii ) creates responsible for its analysis and ( iii ) delegate the necessary risk assessment.
-These actions are executed in sequence and has a pre-condition concerning the Broker's privilege.
-This can only create responsible and delegate the tasks if their authentication datas are correct and if and there was less than 3 requests.
-The tasks to be performed by the User can be ( iv ) answer questionnaire and add evidences; or ( v) perform amendments.
-Both the assertion has one pre-condition concerning the time limit for form submission.
-In this example, we use the limit of 15 days for form submission.
-If the User is performing the task of answer questionnaire and add evidences, this will have to enter with their authentication data.
-As postcondition, it is checked if all the answers were answered .
-The action of perform amendments has no additional verification.
-Finally, the user registers the tasks performed.
-This action has response time verification.
-
-This model refines the concepts defined in the pi-UseCase model in order to better describe the assertions and groups them into contracts.
-
-Pi-ServiceComposition Model description for FlyingPig
-
-This model (presented in Figure Z) describes the interactions of the actions described in pi-ServiceProcess model with the external services.
-Services are provided through the ORCA system and the access is accomplished by FlyingPig interface.
-FlyingPig is the point of interaction between the actions described in the precess with the services offered by ORCA.
-FlyingPig provides an interface with five actions, they are: receive a request, generate new interface, create questionnaire, notify responsable e receive answers..
-This interface makes the direct calls for the ORCA' services.
-The services offered by ORCA are: create space company, generate the questionnaire, analyze answer, data store, call for amendment and create report.
-These services are called according to the process execution.
-
-With respect to non-functional requirements, the assertions grouped into contract in the pi-ServiceProcess model are transformed into policies in the pi-ServiceComposition model.
-The policies identified in this case study are: security, performance and conformability policies.
-Each policy is associated with specific the services and it is verified at the service execution time regarding the specific rule which the sercive is associated with..
+Contracts may be implemented as transversal concerns~\cite{aspects}.
+Their conditions may be verified at specific point of the composition and their actions are executed if needed.
 

### 2013-12-16 JournalToSubmit2013/metamodels.tex
Cleaned related works and section 3 ...

@@
 
-The {\em A-policy} based service composition ($\pi$-SCM) meta-model, described in Figure \ref{fig:e-scomposition-metamodel},
-provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.}.
+The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
+provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
 The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model classes.} and the {\sc Actions} they perform. 
-Instances of this meta-model are UML activity diagrams. 
-Figure~\ref{fig:e-scomposition-metamodel} show the modelling elements.
-Classes pictured as coloured boxes deal with non-functional properties.
-The classes pictured with a white background are those of SOD-M.
+Instances of this meta-model are represented as UML activity diagrams. 
+In Figure~\ref{fig:e-scomposition-metamodel} 
+ Coloured boxes illustrate classes  modelling  non-functional properties and 
+ white boxes represent classes modelling functional ones.
 
-In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
+%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
 \begin{itemizedTrivlist}
@@
 
-\item In order to represent constraints associated to services compositions, we extended the SOD-M service composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+\item In order to represent constraints associated to services compositions, we extended introduced the concepts {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
 We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
-An {\em A-policy} is formed by attributes and rules. 
+An {\em A-policy} is defined by attributes and rules. 
 Intuitively, the conditions of each rule will be checked.

### 2013-12-16 JournalToSubmit2013/modelingWithPISODM.tex
Cleaned related works and section 3 ...

@@
 
+$\pi$-SODM meta-models represent both the functional aspects of the application as well as its non-functional constraints.
 As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM meta-models are organized into three levels: 
 \begin{itemize}
-\item CIM (\textit{Computational Independent Models}). Two models are defined at the CIM level: \textit{value model}
-and \textit{BPMN model}.
-Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
+\item CIM (\textit{Computational Independent Models}). This level focusses on the
+environment of the system, as well as on its business and requirement specifications.
+This viewpoint represents the software system at its highest level of abstraction.
+At this stage of the development, the structure and system processing details are still unknown or undetermined.  $\pi$-SODM defines the \textit{value}
+and \textit{BPMN} meta-moels for this purpose. 
+%Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
+%and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
 
- \item PIM (\textit{Platform Independent Models}). Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, and \textit{$\pi$-ServiceCom\-po\-si\-tion}.
-%and \textit{$\pi$-PEWS}.
+ \item PIM (\textit{Platform Independent Models}). This level focusses on the system functionality, hiding the details of any particular platform.
+The specification defines those parts of the system that do not change from one platform to another. Our method defines four meta-models:
 
@@
 
-  \item PSM (\textit{Platform Specific Models}).
+  \item PSM (\textit{Platform Specific Models}). This level focusses on the functionality, in the context of a particular implementation platform.
+Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
 At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
 The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions.
+\textit{$\pi$-PEWS} models instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
 
@@
 
-$\pi$SOD-M (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
-$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints.
-\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
+$\pi$-SODM (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
 This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
@@
 Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+The following sections describe the $\pi$-SODM meta-models.
 

### 2013-12-16 JournalToSubmit2013/relworks.tex
Cleaned related works and section 3 ...

--- 2013-12-04 JournalToSubmit2013/relworks.tex
@@
-Related works to our approach includes existing proposal dealing with the inclusion of non-functional properties in service-oriented development.
-Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase.
+A non-functional property (NFP) specifies criteria about the behavior of a system.
+These criteria are related to the conditions in which they are executed and to its performance.
+Non-functional properties are also referred as “constraints”, “quality attributes”, “quality goals”, “quality of service requirements” and “non-behavioral requirements” \cite{StGr05}.
+In the case of service-based applications, non-functional requirements concern the application itself as well as its component services.
+The majority of the work on NFPs uses a product-oriented approach, which is concerned with measuring how much a software system is in accordance with the set of NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
+
+These  approaches address the problem from four perspectives: (i) NFP expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for associating certain non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.
+
+\subsection{Classifying Non-functional properties}
+We conducted a systematic review \cite{placidoPhDThesis2012} in order to analyze approaches that consider non-functional properties for developing service-oriented applications.
+As a result this analysis proposes a classification of NFP as a result of a study concerning software methodologies for the construction of service-oriented systems.
+The classification is organized in three layers representing: application modeling, services composition and services.
+The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP).
+At the application layer NFP can refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an “@” and a “.”.
+A value NFR expresses constraints about the way data and functions can be ac-cessed and executed.
+Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
+Value constraints concern more on the conditions in which services can be used.
+For example, accessing to a function within an authentication protocol.
+Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
+For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
+
+\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing exist-ing approaches addressing NFP for service coordinations. Therefore they propose:
+\begin{itemize}
+\item	A meta-model that introduces for characterizing NFP according to the entity to which they are associated: attribute, concern, action, and activity. 
+\item	Six requirements for studying NFP definition approaches for service coordinations: NFP specifications, NFP actions specification, Web service subjects specification, non functional attributes execution order specification, composite Web service subjects specification, stateful non functional constraint specification.
+\item	Seven requirements for studying NFP enforcement approaches: separation of con-cerns, transparent integration of functional and NFP, quantification, superimposi-tion, integration of NFP with distributed Web service, programming language in-dependence, Web service composition support.
+\end{itemize}
+
+NFP concepts are close to those defined by \cite{Souz12} but they are not organized into layers.
+They correspond essentially to the service and service composition layers.
+The enforcement requirements describe the way NFP are weaved to service coordinations.
+Existing works respect all or a subset of these requirements.
+The chosen requirements have an impact in the way NFP are enforced at execution time.
+
+%  Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase.
 Finally exiting methodology for the development of service-oriented applications including non-functional properties are here depicted and compared with our approach.
@@
 
-\subsection{Non-functional properties in service-oriented applications}
+\subsection{Specifying Non-functional properties}
 
-In previous work we have carried out a systematic review \cite{placidoPhDThesis2012} in order to analyze how non-functional properties are taking into account in existing approaches for service-oriented development.
-Existing proposals in this field may be classified in three types of works which are described in the followed.
+Extensions to UML are proposed to express NFPs.
+NFPs can be integrated into the Class, Sequence, and Collaboration Diagrams \cite{CyPr04}.
+The performance engineering [DiSD02] annotates quantitative performance constraints to UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram.
+The UML Profile for Modeling Quality of Service and Fault Tolerance Characteris-tics and Mechanisms \cite{Omg04} describes a set of QoS characteristics such as throughput, latency, integrity, dependability, among others.
+The characteristics can also be parameter-ized with for example the type of keys or types of encryption algorithms in case of the secu-rity characteristics.
+For all UML profiles based approaches it is generally possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g.
+Class or Opera-tion.
+The ordering of stereotypes cannot be expressed but composition- related require-ments can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams.
 
-The first types of works are thus considering a particular non-functional concern (e.g.
-security) which is modeled and then "glued" or associated to functional models of an application.
-The work of Chollet et al.
-\cite{CholletL09} defines a proposal to associate non-functional quality properties (security properties in that case) to functional activities which are modeled in a web service composition model.
-Schmeling et al.
-\cite{SchmelingCM11} present an approach and also a toolset for specifying and implementing non-functional concerns in web service compositions; non-functional concerns are modeled first in non-functional model and then related to a service composition represented in a BPMN diagram.
-Ovaska et al.
-\cite{OvaskaEHPA10} present an approach to support quality management at design time.
-Quality requirements are modeled in a first phase and represented then in an architectural model where quality requirements are associated to some components in the model.
-In our case we propose to model non-functional properties joint to the modeling of functional concerns what simplifies the web service development process.
-Non-functional properties are represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.
+\cite{OrHe07} incorporate extra-functional properties into UML models by the use of UML stereotypes.
+For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services.
+Design patterns help in documenting and communicating proven design solutions to recurring problems.
+Design patterns not only describe how to solve design problems but also document the designers’ intentions - why the solution is chosen over others and what tradeoffs are made.
 
-Second type of works related to our proposal proposes specifics architecture or frameworks to manage and validate QoS attributes in service composition processes.
-The work of Xiao et al.
-\cite{XiaoCZBOLH08} present a framework to verify SLA compliance in composed services.
-Task in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time.
-A similar work is presented in \cite{SchmelingCM11} by using a formal method that allows to choice best suited services instances according to a service composition pattern that include some QoS attributes.
-Babamir et al.
-present in \cite{Babamir2010} a framework for evaluate quality attributes of web services.
-By executing an algorithm, the framework allow to choice best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime.
-In \cite{Karunamurthy2012787} authors propose architecture in which web services may be described with non-functional characteristics that are validated in the service composition process.
+\cite{GrYu01} observe that non-functional properties  (NFPs) are pervasive in the descriptions of design patterns and propose their systematic treatment in descrip-tions of patterns and when applying patterns during design.
 
-Third type of works include proposal working on the evaluation and matching of non-functional properties in the process of search or invocation of particular web services.\cite{AgarwalLS09} define a policy language that provides a way to specific constraints on functional as well as non-functional properties of Web services with the aim of facilitating automated discovery and selection of services.\cite{JeongCL09,Kamalabad2012} defines also proposals to the validate quality attributes of web services when discover and compose.\cite{MohantyRP10} works also on the selection of web service for use considering non-functional properties.
-The paper present web services quality prediction models, which take non-functional properties into account.
-Similarly, \cite{Yeom2006} define a quality driven selection of web services providing a QoS model.
+Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application.
+According to the Busi-ness Rules Group \cite{Hay00}, a business rule is a statement that defines or constrains some aspect of the business.
+It is intended to assert business structure or to control the behavior of the business \cite{Hay00}.
+Business rules are usually expressed either as constraints or in the form if conditions then action.
+The conditions are also called rule premises.
+The business rule approach encompasses a collection of terms (definitions), facts (connection between terms) and rules (computation, constraints and conditional logic) \cite{Hall01}.
+Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making.
+A business rule system is a system in which the rules are separated logically and perhaps physically from the other parts \cite{ChMe04}.
+\cite{ChMe04} investigate a hybrid approach, which combines business processes as known e.g., from BPEL with business rules \cite{Hay00}.
 
-Our proposal by the contrary to the second and third type of proposals is focusing on the design of service-oriented application with non-functional properties and not on the validation of QoS attributes at runtime.
+\cite{ChLa09} proposed a generative framework to address NFPs for web service composi-tions.
+As part of the framework they separate the orchestration meta-model from other non-functional meta-models.
+They propose a non-functional meta-model for security where concepts exist for requirements and how these requirements are ensured.
+The NFPs are then added to the composition model by adding annotations.
+KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic that requirements engineer can use to precisely specify NFPs, such as performance, as part of attribute specification of functional requirements.
 
-Other relevant works such as \cite{DAmbrogio06,Liu20121080} propose extensions to WSDL and BPEL languages respectively for the inclusion of QoS attributes that can be evaluated at runtime.
+The AO4BPEL \cite{CSHM06} deployment descriptor (DD) is used for specifying non-functional requirements for BPEL processes.
+First the service (one of logging, security, transactions and reliable messaging) is selected.
+Then the class (e.g.
+authentication or con-fidentiality for security) can be specified.
+The type (e.g.
+encryption or decryption for confi-dentiality) is then the concrete action that must be executed in order to enforce the re-quirement.
+Then a selector can be defined such as for a certain BPEL activity.
+Each selector can be mapped to a requirement.
 
-\subsection{Standards for programming non-func\-tion\-al properties in service-oriented applications}
-
-Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
+The W3C proposes a set of languages and protocols for specifying the NFP of a ser-vices’ coordination.
+A classification of NFP protocols, provided by the WS-* family protocols proposed by the World Wide Web Consortium (W3C).
+%These protocols suppose an underlying communication protocol. 
+%
+%The following sections describe representative examples of the WS-* protocol-based approaches.
+%
+ Standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
+WSDL and SOAP are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
 For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
@@
 As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications.
+ % This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
  Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services.
@@
 
+\subsection{Non-functional properties in service compositions}
+
+Several approaches adding non-functional properties to service compositions extend the underlying composition model.
+Non-functional properties can concern for example securty, atomicity, availability, response time.
+\cite{CholletL09,SchmelingCM11}  associate non-functional  properties  to functional activities  modeled by a web service composition model. In \cite{chmelingCM11} 
+non-functional propoerties are  are modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
+In \cite{OvaskaEHPA10} non-functional properties concern  quality management and they are specified at design time.
+Quality requirements  are then associated to the components of an architecture that implements a service composition.
+
+Other approaches specify architectures or frameworks to manage and validate QoS attributes in service composition processes. 
+\cite{XiaoCZBOLH08} proposes a framework to verify NFP compliance in composed services.
+Tasks in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time.
+\cite{SchmelingCM11} uses a formal method  to choose best suited services instances according to a service composition pattern that includes some QoS attributes. 
+ \cite{Babamir2010} proposes a framework for evaluating quality attributes of web services.
+By executing an algorithm, the framework allow to choose the best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime.
+\cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for includin  QoS attributes that can be evaluated at runtime.
+ \cite{Karunamurthy2012787} proposes an architecture in which web services are described with non-functional characteristics that are validated in the service composition process.
+
+Other approaches \cite{JeongCL09,Kamalabad2012,MohantyRP10} use matching techniques for choosing services that can cope to non-functional properties required by a given application.
+\cite{AgarwalLS09} defines a policy language to specificy constraints on functional  properties of Web services for facilitating automated discovery and selection of services. 
+\cite{Yeom2006} define a quality driven selection of web services providing a QoS model.
+
+%\cite{} works also on the selection of web service for use considering non-functional properties.
+The paper present web services quality prediction models, which take non-functional properties into account.
+
+Our proposal in contrast, focuses on the design of service-oriented application with non-functional properties modelled  for services and service compositions and then ensured at execution thanks to appropriate environments.
+
+%We propose to model non-functional properties  represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.
+
 \subsection{Methodologies for Service-oriented applications development  including non-func\-tion\-al properties}
 
-Beside the works presented above, there are few methodologies for service-oriented development that address the explicit modeling of non functional properties for service-based applications.
-Important methodologies for building service-based applications have been proposed in the last years \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008}, but all of they focus mainly on the modeling and construction process of service-based business processes that represent the functionally of information systems.
-
-Some methodologies dealing with particular non-functional properties in service orientation have also appeared.
-Tran et al.
-\cite{Tran2012531} present a model-driven approach for addressing problems related to compliance concerns.
-Authors define extend elements of a business process models with elements of a compliance model in which QoS policies are represented.\cite{GutierrezRF10} present PWSSec process, a methodology that provide developers with all the activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.
+ \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008} propose methodologies for building service-based applications that  mainly focus on modeling functional aspectos of service-based business processes.
+SOD-M \cite{decastro1} defines a model-driven approach for development service-based applications.
+Few service-oriented development methodologies    explicitly address  non functional properties modeling \cite{Tran2012531,GutierrezRF10}.
+\cite{GutierrezRF10} presents PWSSec process, a methodology that provides  activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.
 
@@
 The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
+
+%Besides the notion of model at each level of abstraction, MDA propose the use of \textit{model transformations} within and between levels.
+Intra-level transformations are used to provide a unified representation of concepts of a given level.
+Inter-level transformation implement a refinement process between levels.
+Transformations may be automatic or semi-automatic.
+
+%The methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements presented in this work extends the \textit{Service Oriented Development Method} (SOD-M) proposed by De Castro et al.\cite{decastro1}.

### 2013-12-30 JournalToSubmit2013/intro.tex
another version ...still  needs examples, use case finalization, and paper organization

@@
 In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to produce applications and provide a business logic.
+combined to build applications that provide a business logic.
 The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
@@
 Most methods integrate them only after the application has been implemented.
-This leads to service based applications that are partly specified and, thereby, partialy compliant with the requirements of the application.
+This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.
 
@@
 
-Service oriented computing promotes ease of information systems' construction thanks, for instance, to services' reuse.
-Yet, this is not applied to non-functional constraints as the ones described previously, because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
+Service oriented computing promotes ease of information systems construction thanks to services reuse.
+Yet, this is not applied to non-functional constraints because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
 Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
@@
 The literature stresses the need for methodologies and techniques for service oriented analysis and design, since they are the cornerstone in the development of meaningful service based applications~\cite{5}.
-Existing approaches argue that the convergence of model-driven software development, service orientation and better techniques for documenting and improving business processes are key to make real the idea of rapid, accurate development of software that serves, rather than dictates the needs of its users~\cite{watson}.
+Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software that serves, rather than dictates the needs of its users~\cite{watson}.
 Model Driven Development (MDD)  for the development of software systems mainly characterised by the use of models as a product \cite{Selic03}.
-The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}.
-MDA states a set of guidelines for structuring the specifications of software systems.
+%The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}. 
+Model Driven Architecture (MDA)~\cite{miller} states a set of guidelines for structuring the specifications of software systems.
 The technique promotes  the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}).
@@
 %~\cite{decastro1}, 
-to support non-functional aspects of service-oriented applications, from the early stages of software development.
+to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
 $\pi$-SODM is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications. It is intended:
@@
 
-The main contribution of our work is a methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements.
-The methodology is based on  an NFR model including a set of concepts need for the modeling of NFR in service-oriented applications.
+%The main contribution of our work is a methodology for supporting the construction of service-oriented applications,
 
+ The methodology is based on  an NFR model including a set of concepts  for  modeling  NFR in service-oriented applications.
 Accordingly, the remainder of the paper is organized as follows.

### 2013-12-30 JournalToSubmit2013/metamodels.tex
another version ...still  needs examples, use case finalization, and paper organization

@@
 The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
-provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
-The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model classes.} and the {\sc Actions} they perform. 
+provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
+The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
 Instances of this meta-model are represented as UML activity diagrams. 
-In Figure~\ref{fig:e-scomposition-metamodel} 
- Coloured boxes illustrate classes  modelling  non-functional properties and 
+In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
  white boxes represent classes modelling functional ones.
@@
 \begin{itemizedTrivlist}
-\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions. 
-They are graphically presented as a partition in the activity diagram. 
+\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
+An instance of this meta-class is graphically represented as a partition in the activity diagram. 
 A collaborator can be either internal or external to the system. 
@@
 
-\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as an activity. 
-Each action represents some type of transformation or processing. 
-There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation that is not supported by a Web Service, called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
+\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
+A class action represents some type of transformation or processing. 
+There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
 \begin{figure}[t]
@@
 
-\item The {\sc ServiceActivity} element is a composite activity that must be carried out as part of a business service and is composed by one or more executable nodes.
+\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
 
-\item In order to represent constraints associated to services compositions, we extended introduced the concepts {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
 We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
@@
 In case of no compliance, the actions defined by the rule will be performed.
-The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
+The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
 An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
@@
 PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
-The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as B	PEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
+The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
 Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
@@
 \item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-A brie description of these classes is given next.
+A brief description of these classes is given next.
 \end{itemizedTrivlist}

### 2013-12-30 JournalToSubmit2013/modelingFlyingPig.tex
another version ...still  needs examples, use case finalization, and paper organization

@@
 
-$\Pi$-SODM uses two models at the CIM level (see Section~\ref{sec:modelingWithPISODM}): The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
-The former is a simple model just to identify the transference of value information between components of the system.
+$\Pi$-SODM uses two models at the CIM level: The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
+The former  identifies the transference of value information between components of the system.
 The BPMN model establishes which are the actors and main tasks of the application.
@@
 }
-
-\begin{itemize}
-\item {\color{magenta} Placido and Umberto: Could you please complete these items? --G \& M.}
-\end{itemize}
 
@@
 Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
+
+As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
+In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way.
+Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
+In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.

### 2013-12-30 JournalToSubmit2013/modelingWithPISODM.tex
another version ...still  needs examples, use case finalization, and paper organization

@@
 
-$\pi$-SODM meta-models represent both the functional aspects of the application as well as its non-functional constraints.
-As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM meta-models are organized into three levels: 
+%$\pi$-SODM meta-models represent both the functional aspects of the application as well as its non-functional constraints.
+As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM proposes meta-models  organized into three levels: 
 \begin{itemize}
@@
 environment of the system, as well as on its business and requirement specifications.
-This viewpoint represents the software system at its highest level of abstraction.
+This level represents the software system at its highest level of abstraction.
 At this stage of the development, the structure and system processing details are still unknown or undetermined.  $\pi$-SODM defines the \textit{value}
@@
 
- The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.Non-functional requirements are defined as \textit{constraints} over actions and data.
+ The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements.
+Non-functional requirements are defined as \textit{constraints} over actions and data.
 
@@
 
-Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
+Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMs); next,  starting from such models build platform independent models (PIMs) specifying the system details; next transform such models into platform specific models (PSMs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
 The following sections describe the $\pi$-SODM meta-models.

### 2014-02-07 JournalToSubmit2013/modelingWithPISODM.tex
working on the metamodels... not finished yet.

@@
 This level represents the software system at its highest level of abstraction.
-At this stage of the development, the structure and system processing details are still unknown or undetermined.  $\pi$-SODM defines the \textit{value}
-and \textit{BPMN} meta-moels for this purpose. 
+At this stage of the development, the structure and system processing details are still unknown or undetermined.  
+$\pi$-SODM uses the \textit{E3value}~\cite{e3value} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose. 
 %Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
@@
 \item PIM (\textit{Platform Independent Models}). This level focusses on the system functionality, hiding the details of any particular platform.
-The specification defines those parts of the system that do not change from one platform to another. Our method defines four meta-models:
-
- The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements.
-Non-functional requirements are defined as \textit{constraints} over actions and data.
-
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
-%that must be performed upon certain conditions.
-%The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
-described in the \textit{$\pi$-UseCase}.
-%model into contracts that are associated
-%with services.
-
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept  \textit{Policy}
-to integrate contracts with similar non-functional requirements.
-For instance, security and privacy restrictions may be grouped into a security policy.
-%\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
+The specification defines those parts of the system that do not change from one platform to another. 
+Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess} and \textit{$\pi$-ServiceComposition}.
 
@@
 At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
-The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions.
-\textit{$\pi$-PEWS} models instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
-
+We have defined one meta-model at this level.
 \end{itemize}
@@
 
-Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMs); next,  starting from such models build platform independent models (PIMs) specifying the system details; next transform such models into platform specific models (PSMs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.
-The following sections describe the $\pi$-SODM meta-models.
+The following sections describe the $\pi$-SODM meta-models, as well as the transformations between models.
+The models and transformations for an industrial case are presented in section~\ref{sec:flyingPig}.
 

### 2014-02-12 JournalToSubmit2013/metamodels.tex

@@
 %3.1. pi-SoD-M metamodels (presentar los metamodels)
+
+Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMs); next,  starting from such models build platform independent models (PIMs) specifying the system details; next transform such models into platform specific models (PSMs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into code that implements the system.
+
+\subsubsection{Computation Independent Models}
+$\Pi$-SODM uses two models at the CIM level: The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
+The former  identifies the transference of value information between components of the system.
+The BPMN model establishes which are the actors and main tasks of the application.
+
+The e3value defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
+A dependency path has a direction and consists of a sequence of linked dependency nodes.
+A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node\footnote{See Legend on Figure~\ref{fig:E3valuemodel}}. 
+Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
+
+\subsubsection{Platform Independent Models}
+ The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements.
+Non-functional requirements are defined as \textit{constraints} over actions and data.
+
+The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
+%that must be performed upon certain conditions.
+%The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
+described in the \textit{$\pi$-UseCase}.
+%model into contracts that are associated
+%with services.
+
+The \textit{$\pi$-ServiceComposition} meta-model provides the concept  \textit{Policy}
+to integrate contracts with similar non-functional requirements.
+For instance, security and privacy restrictions may be grouped into a security policy.
+%\textit{$\pi$-ServiceComposition} models can be refined into PSMs.
+
+\subsubsection{Platform Specific Models}
+
+The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions. 
+Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
+
+PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
+The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
+Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
+\begin{itemizedTrivlist}
+\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
+A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
+A {\sc Compound Operation} is defined using an {\sc Operator}.
+The language defines operators to denote guarded operations ($[C]S$); sequential ($\ .
+\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.
+
+\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
+A brief description of these classes is given next.
+\end{itemizedTrivlist}
+%
+\begin{figure}[t]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
+\caption{$\pi$-{\sc Pews} Metamodel}
+\label{fig:metamodel}
+\end{figure}
+
+Figure~\ref{fig:metamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
+Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.  
+In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.
+%In this way,
+%\begin{itemizedTrivlist}
+%\item The class {\sc Scope} represents any element of a service composition (i.e., operation, operator, path).
+%\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}. 
+%An {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
+The value of a variable is represented using the type {\sc Variable}.
+The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
+%\end{itemizedTrivlist}
+%
+%An authentication {\em A-policy} represents the situation where an invocation in
+%an activity occurs until its sender and/or its recipient have been
+%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
+
+Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
+The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.
+
+===================================
 
@@
 %..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-
-PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
-The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
-Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
-\begin{itemizedTrivlist}
-\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
-A {\sc Compound Operation} is defined using an {\sc Operator}.
-The language defines operators to denote guarded operations ($[C]S$); sequential ($\ .
-\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.
-
-\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-A brief description of these classes is given next.
-\end{itemizedTrivlist}
-%
-\begin{figure}[t]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
-\end{figure}
-
-Figure~\ref{fig:metamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
-Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.  
-In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.
-%In this way,
-%\begin{itemizedTrivlist}
-%\item The class {\sc Scope} represents any element of a service composition (i.e., operation, operator, path).
-%\item The class {\sc A-Policy} represents a recovery strategy implemented by ECA rules of the form {\sc Event} - {\sc Condition} - {\sc Action}. 
-%An {\em A-policy} has variables that represent the view of the execution state of its associated scope, that is required for executing the rules.
-The value of a variable is represented using the type {\sc Variable}.
-The class {\sc A-Policy} is specialized for defining specific constraints, for instance authentication {\em A-policies}.
-%\end{itemizedTrivlist}
-%
-%An authentication {\em A-policy} represents the situation where an invocation in
-%an activity occurs until its sender and/or its recipient have been
-%identified. Typically, authentication A-Policies ensure that the invocation of the activity will be done within an authentication protocol.
-
-Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
-The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.

### 2014-02-12 JournalToSubmit2013/modelingFlyingPig.tex

@@
 
-$\Pi$-SODM uses two models at the CIM level: The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
-The former  identifies the transference of value information between components of the system.
-The BPMN model establishes which are the actors and main tasks of the application.
-
 Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
@@
 The values exchanged between ORCA and the brokers are: \textsl{Amendments} and \textsl{Evaluation Reports} which are value objects ([ \!\dots]) for the companies that need to have a risk assessment, as well as  \textsl{Questionnaire and Evidences} and the risk assessment \textsl{Fee} which are value objects for ORCA System.
-
-The e3value defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
-A dependency path has a direction and consists of a sequence of linked dependency nodes.
-A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Legend on Figure~\ref{fig:E3valuemodel}). 
-Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
 

### 2014-02-12 JournalToSubmit2013/relworks.tex

@@
 In the case of service-based applications, non-functional requirements concern the application itself as well as its component services.
-The majority of the work on NFPs uses a product-oriented approach, which is concerned with measuring how much a software system is in accordance with the set of NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
+The majority of the work on NFPs uses a product-oriented approach concerned with measuring how much a software system is in accordance with the set of NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
 
@@
 At the application layer NFP can refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an “@” and a “.”.
-A value NFR expresses constraints about the way data and functions can be ac-cessed and executed.
+A value NFR expresses constraints about the way data and functions can be accessed and executed.
 Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
@@
 
-\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing exist-ing approaches addressing NFP for service coordinations. Therefore they propose:
-\begin{itemize}
-\item	A meta-model that introduces for characterizing NFP according to the entity to which they are associated: attribute, concern, action, and activity. 
-\item	Six requirements for studying NFP definition approaches for service coordinations: NFP specifications, NFP actions specification, Web service subjects specification, non functional attributes execution order specification, composite Web service subjects specification, stateful non functional constraint specification.
-\item	Seven requirements for studying NFP enforcement approaches: separation of con-cerns, transparent integration of functional and NFP, quantification, superimposi-tion, integration of NFP with distributed Web service, programming language in-dependence, Web service composition support.
-\end{itemize}
-
+\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing exist-ing approaches addressing NFP for service coordinations. 
+%
+%Therefore they propose:
+%\begin{itemize}
+%\item	A meta-model that introduces for characterizing NFP according to the entity to which they are associated: attribute, concern, action, and activity. 
+%\item	Six requirements for studying NFP definition approaches for service coordinations: NFP specifications, NFP actions specification, Web service subjects specification, non functional attributes execution order specification, composite Web service subjects specification, stateful non functional constraint specification.
+%\item	Seven requirements for studying NFP enforcement approaches: separation of con-cerns, transparent integration of functional and NFP, quantification, superimposi-tion, integration of NFP with distributed Web service, programming language in-dependence, Web service composition support.
+%\end{itemize}
+%
 NFP concepts are close to those defined by \cite{Souz12} but they are not organized into layers.
@@
 
-Extensions to UML are proposed to express NFPs.
-NFPs can be integrated into the Class, Sequence, and Collaboration Diagrams \cite{CyPr04}.
-The performance engineering [DiSD02] annotates quantitative performance constraints to UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram.
-The UML Profile for Modeling Quality of Service and Fault Tolerance Characteris-tics and Mechanisms \cite{Omg04} describes a set of QoS characteristics such as throughput, latency, integrity, dependability, among others.
-The characteristics can also be parameter-ized with for example the type of keys or types of encryption algorithms in case of the secu-rity characteristics.
+Extensions to UML are proposed to express NFPs that can be integrated into the Class, Sequence, and Collaboration Diagrams \cite{CyPr04}.
+UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram are annotated with quantitative performance constraints.
+The UML Profile for Modeling Quality of Service and Fault Tolerance Characteristics and Mechanisms \cite{Omg04} describe a set of QoS characteristics such as throughput, latency, integrity, and dependability.
+The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms in case of the security characteristics.
 For all UML profiles based approaches it is generally possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g.
-Class or Opera-tion.
-The ordering of stereotypes cannot be expressed but composition- related require-ments can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams.
+Class or Operation.
+For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services.
+The ordering of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.
 
-\cite{OrHe07} incorporate extra-functional properties into UML models by the use of UML stereotypes.
-For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services.
-Design patterns help in documenting and communicating proven design solutions to recurring problems.
+ %incorporate extra-functional properties into UML models by the use of UML stereotypes. 
+%Design patterns help in documenting and communicating proven design solutions to recurring problems.
 Design patterns not only describe how to solve design problems but also document the designers’ intentions - why the solution is chosen over others and what tradeoffs are made.
 
-\cite{GrYu01} observe that non-functional properties  (NFPs) are pervasive in the descriptions of design patterns and propose their systematic treatment in descrip-tions of patterns and when applying patterns during design.
+%\cite{GrYu01} observe that non-functional properties  (NFPs) are pervasive in the descriptions of design patterns and propose their systematic treatment in descrip-tions of patterns and when applying patterns during design.
 
-Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application.
-According to the Busi-ness Rules Group \cite{Hay00}, a business rule is a statement that defines or constrains some aspect of the business.
-It is intended to assert business structure or to control the behavior of the business \cite{Hay00}.
+%Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application.
+According to the Business Rules Group \cite{Hay00}, a business rule  is intended to assert business structure or to control the behavior of the business \cite{Hay00}.
 Business rules are usually expressed either as constraints or in the form if conditions then action.
@@
 Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making.
-A business rule system is a system in which the rules are separated logically and perhaps physically from the other parts \cite{ChMe04}.
+In a business rule system  rules are logically separated  and perhaps physically from the other parts \cite{ChMe04}.
 \cite{ChMe04} investigate a hybrid approach, which combines business processes as known e.g., from BPEL with business rules \cite{Hay00}.
@@
 Then the class (e.g.
-authentication or con-fidentiality for security) can be specified.
+authentication or confidentiality for security) can be specified.
 The type (e.g.
-encryption or decryption for confi-dentiality) is then the concrete action that must be executed in order to enforce the re-quirement.
-Then a selector can be defined such as for a certain BPEL activity.
+encryption or decryption for confidentiality) is then the concrete action that must be executed in order to enforce the requirement.
+A selector can be defined such as for a certain BPEL activity.
 Each selector can be mapped to a requirement.
@@
 For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
-As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
- % This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
- Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services.
+The selection of the adequate protocols for adding a specific non-functional constraints to a services composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
+
+% As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
+% % This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
+% Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services.
 In \cite{Vidyasankar:2004} the model introduced in \cite{SchuldtABS02} is extended to web services for addressing atomicity.
-
-As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way.
-Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
 
@@
 \cite{CholletL09,SchmelingCM11}  associate non-functional  properties  to functional activities  modeled by a web service composition model. In \cite{chmelingCM11} 
-non-functional propoerties are  are modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
+non-functional propoerties are   modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
 In \cite{OvaskaEHPA10} non-functional properties concern  quality management and they are specified at design time.
@@
  \cite{Babamir2010} proposes a framework for evaluating quality attributes of web services.
-By executing an algorithm, the framework allow to choose the best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime.
-\cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for includin  QoS attributes that can be evaluated at runtime.
+By executing an algorithm, the framework  chooses the best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime.
+\cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for including  QoS attributes that can be evaluated at runtime.
  \cite{Karunamurthy2012787} proposes an architecture in which web services are described with non-functional characteristics that are validated in the service composition process.
@@
 
-\subsection{Methodologies for Service-oriented applications development  including non-func\-tion\-al properties}
+\subsection{Towards methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}
 
@@
 Unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that allows to model service-based applications starting from a high-level business modelling, and defining meta-models at different abstraction levels (CIM, PIM and PSM), it enables the design and development of services-based applications that can be reused and that are reliable.
-The definition of a model-driven approach that allows to model non-functional properties in the early stages of the software development process, and the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefits of our proposal.
+The definition of a model-driven approach  allows to model non-functional properties in the early stages of the software development process, and the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefits of our proposal.
 

### 2014-02-12 JournalToSubmit2013/metamodels.tex
section 3

@@
 
-Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMs); next,  starting from such models build platform independent models (PIMs) specifying the system details; next transform such models into platform specific models (PSMs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into code that implements the system.
+%Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMs); next,  starting from such models build platform independent models (PIMs) specifying the system details; next transform such models into platform specific models (PSMs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into code that implements the system.
 
-\subsubsection{Computation Independent Models}
-$\Pi$-SODM uses two models at the CIM level: The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
-The former  identifies the transference of value information between components of the system.
+\subsection{Computation Independent Models}
+$\Pi$-SODM uses two models at the CIM level: The e$^3$value~\cite{e3value} model and BPMN~\cite{BPMN}.
+The former identifies the value/information exchange between components of the system.
 The BPMN model establishes which are the actors and main tasks of the application.
 
-The e3value defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
+The e$^3$value model defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
 A dependency path has a direction and consists of a sequence of linked dependency nodes.
@@
 
-\subsubsection{Platform Independent Models}
+The next example ilustrates the use of the models.
+
+\begin{example}[To Publish Music]\label{ex:toPublicMusic}
+To illustrate our proposal, we build the models for a ``To Publish Music'' problem. 
+This problem involves three actors ({\em Spotify, Twitter} and {\em Facebook}).
+We model a business process that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
+The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
+Both {\em Facebook} and {\em Twitter} require authentication. 
+Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
+
+THe e$^3$value and BPMN models for this applications are shown in Figure~\ref{fig:CIM:tpm}.
+\hfill\openbox
+\end{example}
+
+\begin{figure}
+\framebox{\color{red}\Huge Valeria, please add the figures?}
+\caption{\label{fig:CIM:tpm} CIM models for ``To Publish Music''.}
+\end{figure}
+
+In the next sections, we present each meta-model of $\pi$SOD-M, the models for the ``To Publish Music'' example, as well as the transformations between models.
+Notice that \textit{ad-hoc} transformations are used to obtain PIM models from the CIM models proposed here.
+
+\subsection{Platform Independent Models}
  The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements.
@@
 % \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
-The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}. 
+The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
 Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.

### 2014-02-12 JournalToSubmit2013/modelingFlyingPig.tex
section 3

@@
 
-\begin{figure}[t]
-\centering
-%\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut.png}
-\fbox{\LARGE $\pi$-PEWS Model.}
+%\begin{figure}[t]
+%\centering
+\begin{lstlisting}[label=list:flyingPigPiPEWS,caption=pi-PEWS
+Specification: FlyingPig. ]!\label{fig:PiPEWSModel}!
 
-{\color{red} \raggedright
-$\star$ Placido: Please, place the model here.
-}
-\caption{$\pi$-PEWS model for \FlyingPig.\label{fig:PiPEWSModel}}
-\end{figure}
+//Namespaces specify service URI
+namespace orca = www.orca.mx/service.wsdl
+
+//Operations
+alias createCompanySpace = portType/createCompanySpace in orca
+alias generateQuestionnaire = portType/generateQuestionnaire in orca
+alias analizeAnswer = portType/analizeAnswer in orca
+alias storeData = portType/storeData in orca
+alias createReport = portType/createReport in orca
+alias callForAmendments = portType/callForAmendments in orca
+
+//Services
+service receiveRequest(R, Id) = createCompanySpace(R, Id)
+
+service generateNewInterface(Id, NULL) = ...
+
+service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
+
+service notifyResponsable((Id, Q); NULL) = ...
+
+service receiveAnswers((Id, T); P) =
+			analizeAnswer((Id, T), NULL) 
+			. storeData((Id, T), NULL)
+			. (
+			  (createReport(Id, P) . return(P))
+			  + (callForAmendments(Id,T) . return(NULL))
+			  )
+
+//Path
+
+receiveRequest(R, Id)
+||
+( generateNewInterface(Id, NULL)
+  . createQuestionnaire(Id, Q)
+  .  notifyResponsable((Id, Q); NULL
+ )
+||
+ ( receiveAnswers((Id, T); P)
+ . [P != NULL] STOP
+ )*
+
+//Policies
+
+\end{lstlisting}
+%\fbox{\LARGE $\pi$-PEWS Model.}
+%
+%{\color{red} \raggedright
+%$\star$ Placido: Please, place the model here.
+%}
+%\caption{$\pi$-PEWS model for \FlyingPig.\label{fig:PiPEWSModel}}
+%\end{figure}
 

### 2014-02-12 JournalToSubmit2013/modelingWithPISODM.tex
section 3

@@
 
-In this section we present $\pi$SOD-M, an MDD based methodology for building service compositions with non-functional requirements.
-$\pi$SOD-M provides concepts for  modelling  non-functional requirements at the early stages of the development.
+In this section we present $\pi$SOD-M, an MDD-based methodology for building service compositions with non-functional requirements. 
+$\pi$SOD-M extends SOD-M with concepts for modelling non-functional requirements.
+This is done at all levels of MDA, starting from the early stages of the development.
 %The method uses the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a} to represent non-functional requirements.
+%$\pi$-SOD-M meta-models represent both the functional aspects of the application as well as its non-functional constraints.
 
@@
 
-%$\pi$-SODM meta-models represent both the functional aspects of the application as well as its non-functional constraints.
-As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM proposes meta-models  organized into three levels: 
+$\pi$-SODM proposes meta-models organized into three levels (Figure~\ref{fig:piSOD-M}): 
 \begin{itemize}
@@
 At this stage of the development, the structure and system processing details are still unknown or undetermined.  
-$\pi$-SODM uses the \textit{E3value}~\cite{e3value} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose. 
+$\pi$-SODM uses the \textit{e$^3$value}~\cite{e3value} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose. 
 %Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)

### 2014-02-25 JournalToSubmit2013/modelingFlyingPig.tex
Section 3.3

@@
+\subsection{Platform-Specific Models}
+
+This level focusses on the functionality, in the context of a particular implementation infrastructure.
+Models at this level combine the platform-independent view with specific aspects of the application to be deployed. 
+At the PSM level we have lower-level models that can be automatically translated into any service composition language. 
+We have defined one meta-model at this level, called \textit{$\pi$-PEWS}.
+
+The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions, including non-functional features.
+\textit{$\pi$-PEWS} models specify: (i) The workflow of the application, and (ii) a mechanism to implement non-functional restrictions.
+
+The workflow constructors of \textit{$\pi$-PEWS} are similar to those of  most service orchestration languages, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
+Non-functional behaviour is specified as Event-Condition-Action (ECA) rules~\cite{ECA}.
+
+\textit{$\pi$-PEWS} models have two main parts:
+\begin{itemizedTrivlist}
+\item \textit{Service compositions (workflows)} formed by individual service operations and composite operations defined by the use of workflow operators, such as guarded operations ($[C]S$); sequential ($\ .
+\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetitions.
+
+\item \textit{ECA Rules}, with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}. 
+\end{itemizedTrivlist}
+%%
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
+%\caption{$\pi$-{\sc Pews} Metamodel}
+%\label{fig:metamodel}
+%\end{figure}
+%%
+
+The following example presents the \textit{$\pi$-PEWS} model for the ``To Publish Music'' application.
+
+\begin{example}[To Publish Music (end)]\label{Colocar-o-label-correto-aqui}
+The next code presents the \textit{$\pi$-PEWS}  model for our running example.\footnote{\color{red} Placido: Please change the pdf file as needed!}
+
+\begin{center}
+\includegraphics[scale=0.7]{figs/pewsEspecificationContract.pdf}
+\end{center}
+
+The upper part of the model consists of a workflow expression.
+After the execution of searchSong, followed by selectSong,
+the user may either buy, download and listen a song or just listen to a song (conditional workflow).
+The last part of the wokflow publish the title of the song either on tweeter or facebook.
+
+ECA rules are specified as ``contracts''.
+In our case, the \texttt{buySongContract} specifies conditions to be verified when the \texttt{buySong} service is executed.
+The contract specifies pre and post-conditions, as well as time constraints.\footnote{\color{red} Placido: Please verify the pre-condition assertions!}
+The precondition\dots
+
+In case of failure in verifying the precondition, the service \texttt{buySong} is retried.
+
+The postcondition ensures that if the service completes, the payment is processed.
+
+The time constraint indicates that the \texttt{sendConfirmation} service is invoked immediately after \texttt{buySong}.
+\end{example}
+
+===================================
+
+Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
+The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.
+
+The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
+provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
+The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
+Instances of this meta-model are represented as UML activity diagrams. 
+In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
+ white boxes represent classes modelling functional ones.
+
+%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
+\begin{itemizedTrivlist}
+\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
+An instance of this meta-class is graphically represented as a partition in the activity diagram. 
+A collaborator can be either internal or external to the system. 
+When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
+
+\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
+A class action represents some type of transformation or processing. 
+There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
+\begin{figure}[t]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
+\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
+\label{fig:e-scomposition-metamodel}
+\end{figure}
+
+\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
+
+\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
+An {\em A-policy} is defined by attributes and rules. 
+Intuitively, the conditions of each rule will be checked.
+In case of no compliance, the actions defined by the rule will be performed.
+The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
+An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
+\end{itemizedTrivlist}
+
+\begin{figure}[t]%[htpb]
+\centering
+\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
+
+{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
+
+\caption{$\pi$-SCM for the ``To publish music'' business service.}
+\label{fig:servicecompositionmodel}
+\end{figure}
+
+\begin{example}[To Publish Music]\label{ex:toPublicMusic}
+To illustrate the use of the $\pi$-Service Composition Meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
+In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
+% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
+The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
+Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
+Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
+%A call to such services must be part of the authentication protocol required by these services.
+In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
+In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
+
+{\sf\small OAuthPolicy} will implement the open authentication protocol.
+The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
+This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
+The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
+Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
+
+{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
+The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
+The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
+On the notification of an event of that type, a credential is obtained using the username and password. 
+\hfill\openbox
+\end{example}
+
+We propose the use of rules and policies to model and associate non-functional properties to service compositions.
+These artifacts will be used to generate the actual programs that will implement the application:
+Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
+The $\pi$-PEWS  meta-model is described in the next section.
+
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
+\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
+
+This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
+For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
+Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
+
+%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
+From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
+%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
+
+%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+%\subsection{$\pi$-SOD-M Development Environment}
+%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+
+Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
+The environment implements the abstract architecture.
+Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
+\begin{figure}[htpb]
+	\begin{center}
+		\includegraphics[width=0.60\textwidth]{figs/architecture}
+	\end{center}
+		\caption{$\pi$-SOD-M Development Environment}
+   \label{fig:policymanager}
+\end{figure}
+\begin{itemize}
+\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
+Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
+
+\item	 We used  ATL \footnote{http://eclipse.org/atl/.
+An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
+for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
+
+\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
+We coded the pews.mt program  that implements the model to text transformation for generating executable code.
+It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
+{\em A-policy} based services' composition execution environment.
+
+%\item Finally, we created a chain execution  to execute the model to text transformation.
+\end{itemize}
+
+%
+As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
+The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
+The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
+
+The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
+Each activity is seen as the process where the service method call is executed.
+The execution of an activity has four states: prepared, started, terminated, and failure.
+The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
+
+At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
+Policies associated to a scope are activated when the execution of its scope starts.
+A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
+If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
+%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
+If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
+The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
+%The order of policies has implications on the global order of the rules to be executed.
+
+=============================================================================
+
 We validated our methodology by developing a use-case concerning risk assessment for financial companies.

### 2014-02-27 JournalToSubmit2013/modelingFlyingPig.tex
Section 3.3 and 4.1

@@
 The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions, including non-functional features.
-\textit{$\pi$-PEWS} models specify: (i) The workflow of the application, and (ii) a mechanism to implement non-functional restrictions.
-
-The workflow constructors of \textit{$\pi$-PEWS} are similar to those of  most service orchestration languages, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
-Non-functional behaviour is specified as Event-Condition-Action (ECA) rules~\cite{ECA}.
-
 \textit{$\pi$-PEWS} models have two main parts:
@@
 \item \textit{ECA Rules}, with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}. 
+These rules are used to specify non-functional behaviour.
 \end{itemizedTrivlist}
+
+The workflow constructors of \textit{$\pi$-PEWS} are similar to those of  most service orchestration languages, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
+
 %%
@@
 
+\bigskip
 The following example presents the \textit{$\pi$-PEWS} model for the ``To Publish Music'' application.
-
 \begin{example}[To Publish Music (end)]\label{Colocar-o-label-correto-aqui}
@@
 
-The upper part of the model consists of a workflow expression.
-After the execution of searchSong, followed by selectSong,
-the user may either buy, download and listen a song or just listen to a song (conditional workflow).
-The last part of the wokflow publish the title of the song either on tweeter or facebook.
+The initial part of the model consists of a workflow expression.
+In the given code, the execution of \texttt{searchSong} is followed by \texttt{selectSong}.
+At this point, the user may either buy, download and listen a song or just listen to a song (conditional workflow).
+The last part of the workflow publishes the title of the song either on tweeter or facebook.
 
-ECA rules are specified as ``contracts''.
-In our case, the \texttt{buySongContract} specifies conditions to be verified when the \texttt{buySong} service is executed.
-The contract specifies pre and post-conditions, as well as time constraints.\footnote{\color{red} Placido: Please verify the pre-condition assertions!}
-The precondition\dots
+ECA rules are specified as contracts~\cite{Meyer92}.
+In our case, the \texttt{buySongContract} specifies conditions to be verified upon the execution of \texttt{buySong}.
+This contract specifies pre and post-conditions, as well as time constraints.\footnote{\color{red} Placido: Please verify the pre-condition assertions!}
+The precondition (EXPLAIN!)\dots
 
@@
 
-The postcondition ensures that if the service completes, the payment is processed.
+The postcondition ensures that the payment succeeds before .
 
@@
 \end{example}
-
-===================================
 
@@
 Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied the client's business practice.
-For instance, the conditions and protocols used to perform confidential transactions, the physical security for accessing reserved areas such as computing server installations.
-The information gathered by the questionnaires is used to determine whether there are risky practices within the business processes of the company, as well as to propose amends to these practices.
+Examples of business practices are the conditions and protocols used to perform confidential transactions or the physical security for accessing reserved areas (such as computing server installations).
+The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amends to these practices.
 The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
@@
 Our goal is to model a service based application (called \FlyingPig), for providing risk assessment as a service.
-In order to provide this functionality, \FlyingPig\ would benefit from ORCA's legacy services: storage, assessment and data visualization functions.
+In order to provide this functionality, \FlyingPig\ would benefit from ORCA's legacy services for storage, assessment and data visualization functions.
 
@@
 Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
-It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
-
-In our use-case, we identify two business actors: \textsl{ORCA} and \textsl{Broker}. 
-Brokers are responsible for channelling requests for risk assessment of one or several companies. 
-ORCA have two value activities which are services that provide an economical benefit: to \textsl{Identify Amendments} and the possibility to \textsl{Assess Risk Situation}. 
-The values exchanged between ORCA and the brokers are: \textsl{Amendments} and \textsl{Evaluation Reports} which are value objects ([ \!\dots]) for the companies that need to have a risk assessment, as well as  \textsl{Questionnaire and Evidences} and the risk assessment \textsl{Fee} which are value objects for ORCA System.
+It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).\footnote{\color{red} This sentence need to move to section 3.1!!}
+We identify two business actors: \textsl{ORCA} and \textsl{Broker}. 
+Brokers channel requests for risk assessment for one or more companies. 
+ORCA have two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}. 
+The values exchanged between ORCA and the brokers are: 
+(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
+(ii) \textsl{Amendments} which are ORCA's recommendations, based on the data provided by the answers to questionnaires; 
+(iii) \textsl{Evaluation Reports} for the client companies; 
+and 
+(iv) the risk assessment \textsl{Fee}.
 
@@
 Once this need occurs, the value exchanges between ORCA and Broker are triggered. 
-The client company will provide orca wit information (answers to a questionary), evidence (to support the information) and a fee (monetary value).
+The client company will provide ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
 ORCA will provide amendments (recommendations to change practices) and an evaluation report.
@@
 
-The BPMN model is devised to better understand the process in which the value exchanges occur.
-Figure~\ref{fig:BPMNmodel} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.} for the \FlyingPig\ scenario. 
+The BPMN model for \FlyingPig\ is devised to better understand the process in which the value exchanges occur.
+Figure~\ref{fig:BPMNmodel} shows the BPMN model\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.} for the \FlyingPig\ scenario. 
 The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}. 
@@
 
-Once tasks are completed, they are stored and analysed to generate a list of un-compliant situations, associated to their corresponding \textit{calls for amendment}, in case there are any, or a report specifying a compliance level, incidents and a risk map.
-During the process of analysing a questionnaire, the answers of some questions can trigger the generation of other questionnaires or amendments, that will become new tasks.
+Once tasks are completed, they are stored and analysed to generate a list of noncompliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
+During the process of analysing a questionnaire, the answers of some questions can trigger the generation of additional questionnaires or amendments, that will become new tasks.
 
-Business processes have also associated rules and constraints that define their non functional requirements.
-NFR represents the ``semantics'' and the conditions in which the tasks must be done.
-In our example we have some constraints.
-
+Business processes also have associated rules and constraints to define their non-functional requirements (NFR).
 {\color{red}

### 2014-02-28 JournalToSubmit2013/modelingWithPISODM.tex
section 3

@@
 $\pi$SOD-M extends SOD-M with concepts for modelling non-functional requirements.
-This is done at all levels of MDA, starting from the early stages of the development.
-%The method uses the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a} to represent non-functional requirements.
-%$\pi$-SOD-M meta-models represent both the functional aspects of the application as well as its non-functional constraints.
-
+$\pi$-SODM (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
 \begin{figure}[h]
 \centering
-\includegraphics[width=1.0\textwidth]{figs/piSODM}
+\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
 \caption{$\pi$SOD-M Overview.}
@@
 
-$\pi$-SODM proposes meta-models organized into three levels (Figure~\ref{fig:piSOD-M}): 
-\begin{itemize}
-\item CIM (\textit{Computational Independent Models}). This level focusses on the
-environment of the system, as well as on its business and requirement specifications.
-This level represents the software system at its highest level of abstraction.
+$\pi$-SODM proposes meta-models, organized into three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+These models are explained next.
+
+\subsection{Computation Independent Models}
+
+This level focusses on the highest-level view of the system, including its business and requirement specifications.
 At this stage of the development, the structure and system processing details are still unknown or undetermined.  
 $\pi$-SODM uses the \textit{e$^3$value}~\cite{e3value} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.
-%Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes)
-%and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
 
-\item PIM (\textit{Platform Independent Models}). This level focusses on the system functionality, hiding the details of any particular platform.
+The e$^3$value model identifies the value/information exchange between components of the system. 
+It defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need.
+A dependency path has a direction and consists of a sequence of linked dependency nodes.
+A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node\footnote{See Legend on Figure~\ref{fig:CIM:tpme3v}}. 
+Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
+
+The BPMN model is a graphical representation that establishes the high-level workflow of the application.
+
+The next example illustrates the use of these models.
+\begin{example}[To Publish Music]\label{ex:toPublicMusic}
+To illustrate our proposal, we build the models for a ``To Publish Music'' problem. 
+This problem involves three external actors ({\em Spotify, Twitter} and {\em Facebook}), as well as the user and the application itself.
+
+The e$^3$value and BPMN models for this applications are shown in Figure~\ref{fig:CIM:tpm}.
+Notice that the models at this level show\dots 
+{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
+%\hfill\openbox
+\end{example}
+
+\begin{figure}\center
+\includegraphics[width=0.88\textwidth]{figs/e3value.pdf}
+\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
+\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
+\end{figure}
+
+\begin{figure}\center
+\includegraphics[width=0.49\textwidth]{figs/SC.pdf}
+\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
+\end{figure}
+
+The CIM level models are the basis for the development of the application. 
+The information presented at this level are (manually) refined into PIM-level models.
+
+\subsection{Platform Independent Models}
+
+This level focusses on the system functionality, hiding the details of any particular platform.
 The specification defines those parts of the system that do not change from one platform to another. 
@@
 
- \item PSM (\textit{Platform Specific Models}). This level focusses on the functionality, in the context of a particular implementation platform.
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements. 
+Non-functional requirements are defined as \textit{constraints} over actions and data.
+
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
+{\color{red} Talk about this model for the example.}
+\end{example}
+
+The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
+gathers the constraints described in the \textit{$\pi$-UseCase}.
+
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
+{\color{red} Talk about this model for the example.}
+\end{example}
+
+The \textit{$\pi$-ServiceComposition} meta-model provides the concept  of \textit{Policy}
+to integrate contracts with similar non-functional requirements.
+For instance, security and privacy restrictions may be grouped into a security policy.
+
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
+We model a business process that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
+The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
+Both {\em Facebook} and {\em Twitter} require authentication. 
+Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
+{\color{red} To be completed!.}
+\end{example}
+
+\subsection{Platform Specific Models}
+
+This level focusses on the functionality, in the context of a particular implementation platform.
 Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
@@
 We have defined one meta-model at this level.
-\end{itemize}
 
-$\pi$-SODM (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
-This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.
+The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions. 
+Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
 
-The following sections describe the $\pi$-SODM meta-models, as well as the transformations between models.
-The models and transformations for an industrial case are presented in section~\ref{sec:flyingPig}.
+PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
+The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
+Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
+\begin{itemizedTrivlist}
+\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
+A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}. 
+A {\sc Compound Operation} is defined using an {\sc Operator}.
+The language defines operators to denote guarded operations ($[C]S$); sequential ($\ .
+\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.
 
-%Constraints are restrictions that must be verified during the execution of the application.
-%An example of this is the requirement of the user's authentication for executing some system functions.
+\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
+A brief description of these classes is given next.
+\end{itemizedTrivlist}
 %
+\begin{figure}[t]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
+\caption{$\pi$-{\sc Pews} Metamodel}
+\label{fig:metamodel}
+\end{figure}
 
-%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
+Figure~\ref{fig:metamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
+Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.  
+In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.
+
+Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
+The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.
+
+===================================
+
+The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
+provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
+The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
+Instances of this meta-model are represented as UML activity diagrams. 
+In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
+ white boxes represent classes modelling functional ones.
+
+%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
+\begin{itemizedTrivlist}
+\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
+An instance of this meta-class is graphically represented as a partition in the activity diagram. 
+A collaborator can be either internal or external to the system. 
+When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
+
+\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
+A class action represents some type of transformation or processing. 
+There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
+\begin{figure}[t]
+\centering
+\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
+\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
+\label{fig:e-scomposition-metamodel}
+\end{figure}
+
+\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
+
+\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
+An {\em A-policy} is defined by attributes and rules. 
+Intuitively, the conditions of each rule will be checked.
+In case of no compliance, the actions defined by the rule will be performed.
+The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
+An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
+\end{itemizedTrivlist}
+
+\begin{figure}[t]%[htpb]
+\centering
+\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
+
+{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
+
+\caption{$\pi$-SCM for the ``To publish music'' business service.}
+\label{fig:servicecompositionmodel}
+\end{figure}
+
+\begin{example}[To Publish Music]\label{ex:toPublicMusic}
+To illustrate the use of the $\pi$-Service Composition Meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
+In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
+% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
+The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
+Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
+Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
+%A call to such services must be part of the authentication protocol required by these services.
+In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
+In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
+
+{\sf\small OAuthPolicy} will implement the open authentication protocol.
+The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
+This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
+The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
+Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
+
+{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
+The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
+The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
+On the notification of an event of that type, a credential is obtained using the username and password. 
+\hfill\openbox
+\end{example}
+
+We propose the use of rules and policies to model and associate non-functional properties to service compositions.
+These artifacts will be used to generate the actual programs that will implement the application:
+Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
+The $\pi$-PEWS  meta-model is described in the next section.
+
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
+\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..

### 2014-03-03 JournalToSubmit2013/intro.tex
New version of the Introduction by genoveva 03.04.2014

@@
 In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to build applications that provide a business logic.
+combined to build an application business logic.
 The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
-An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the IT systems actually being developed.
+An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the application actually being developed.
 
 Functional properties are not the only  aspect in the software development process.
-Other non-functional properties need to be addressed to fit in the application with its context (e.g., data privacy, exception handling, atomicity, data persistence).
-Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
-Most methods integrate them only after the application has been implemented.
-This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.
+Other non-functional properties need to be addressed to fit in the application for example  data privacy, exception handling, atomicity  and, data persistence.
+Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
 
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call a service in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
+Service oriented computing promotes ease of information systems construction thanks to services reuse.
+Yet, this is not applied to non-functional constraints because they do not follow in general the same service oriented principle.
+%
+%Furthermore,  often they are  not fully considered in the specification process of existing services' oriented development methods. Rather, 
+%They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed. 
+Ideally, non-functional requirements should be considered along with all the stages of the software development.
+In fact,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
+In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
+
+%Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
+%Most methods integrate them only after the application has been implemented.
+%This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.
+
 % song title disseminated in the walls of the user's Facebook and Twitter accounts.
 
-Service oriented computing promotes ease of information systems construction thanks to services reuse.
-Yet, this is not applied to non-functional constraints because they do not follow in general the same service oriented principle and because they are often not fully considered in the specification process of existing services' oriented development methods.
-Rather, they   are either supposed to be ensured by the underlying execution platform, or they are programmed through ad-hoc protocols.
-Besides,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
-
-Ideally, non-functional requirements should be considered along with all the stages of the software development.
-The adoption of non-functional specifications from the early states of development
-can help the developer to produce applications that  can deal with
-the application context.
+%The adoption of non-functional specifications from the early states of development can help the developer to produce applications that  can deal with the application context.
 
@@
 
-This paper proposes  extension of the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
+This paper proposes  the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
 %~\cite{decastro1}, 

### 2014-03-03 JournalToSubmit2013/relworks.tex
New version of related works ... a lot of bibs are missing 04.04.2014

@@
-A non-functional property (NFP) specifies criteria about the behavior of a system.
-These criteria are related to the conditions in which they are executed and to its performance.
-Non-functional properties are also referred as “constraints”, “quality attributes”, “quality goals”, “quality of service requirements” and “non-behavioral requirements” \cite{StGr05}.
+A non-functional property (NFP) specifies a criterion about the behavior of a system related to the conditions in which it is executed and to its performance.
+Non-functional properties are also referred to as constraints, quality attributesÄù, quality goalsÄù, Äúquality of service requirements and Äúnon-behavioral requirementsÄù \cite{StGr05}.
 In the case of service-based applications, non-functional requirements concern the application itself as well as its component services.
-The majority of the work on NFPs uses a product-oriented approach concerned with measuring how much a software system is in accordance with the set of NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
+The majority of the work on NFPs uses a product-oriented approach concerned with measuring in which extent a software system is fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
 
-These  approaches address the problem from four perspectives: (i) NFP expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for associating certain non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.
+Existing  approaches address NFP from four perspectives: (i) NFP expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for associating certain non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Classifying Non-functional properties}
-We conducted a systematic review \cite{placidoPhDThesis2012} in order to analyze approaches that consider non-functional properties for developing service-oriented applications.
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+We conducted a systematic review \cite{placidoPhDThesis2012} to analyze approaches that consider non-functional properties for developing service-oriented applications.
 As a result this analysis proposes a classification of NFP as a result of a study concerning software methodologies for the construction of service-oriented systems.
-The classification is organized in three layers representing: application modeling, services composition and services.
-The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP).
-At the application layer NFP can refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an “@” and a “.”.
+The classification is organized in three layers: application modeling, services composition and services.
+
+\begin{itemize}
+\item At the application layer NFP  refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an "@"Äù and a "Äú."Äù.
 A value NFR expresses constraints about the way data and functions can be accessed and executed.
-Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).
+
+\item The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP).
+
+\item Business NFP at the service layer concern properties that are associated to services and defines how to call their exported operations (business properties).
 Value constraints concern more on the conditions in which services can be used.
 For example, accessing to a function within an authentication protocol.
-Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
+This layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
 For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
+\end{itemize}
 
-\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing exist-ing approaches addressing NFP for service coordinations. 
+\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing existing approaches addressing NFP for service coordinations. 
 %
@@
 %As a result of our study of related works, in this section we present a categorization of main concepts related the treatment if non-functional requirement in the field of service-oriented development.
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+\subsection{Specifying Non-functional properties}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
-\subsection{Specifying Non-functional properties}
-
-Extensions to UML are proposed to express NFPs that can be integrated into the Class, Sequence, and Collaboration Diagrams \cite{CyPr04}.
+\cite{CyPr04} proposes an extension to UML for expressing NFPs that can be integrated into the Class, Sequence, and Collaboration Diagrams.
 UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram are annotated with quantitative performance constraints.
@@
 The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms in case of the security characteristics.
-For all UML profiles based approaches it is generally possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g.
+For all UML profiles based approaches it is  possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g.
 Class or Operation.
 For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services.
-The ordering of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.
+The ordering or priority of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.
 
@@
 %Design patterns help in documenting and communicating proven design solutions to recurring problems.
-Design patterns not only describe how to solve design problems but also document the designers’ intentions - why the solution is chosen over others and what tradeoffs are made.
+Design patterns not only describe how to solve design problems but also document the designers‚Äô intentions - why the solution is chosen over others and what tradeoffs are made.
 
@@
 Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making.
-In a business rule system  rules are logically separated  and perhaps physically from the other parts \cite{ChMe04}.
-\cite{ChMe04} investigate a hybrid approach, which combines business processes as known e.g., from BPEL with business rules \cite{Hay00}.
+In a business rule system,  rules are logically separated   from the system \cite{ChMe04}.
+\cite{ChMe04} investigate a hybrid approach, that combines business processes as known in BPEL with business rules \cite{Hay00}.
 
-\cite{ChLa09} proposed a generative framework to address NFPs for web service composi-tions.
-As part of the framework they separate the orchestration meta-model from other non-functional meta-models.
-They propose a non-functional meta-model for security where concepts exist for requirements and how these requirements are ensured.
-The NFPs are then added to the composition model by adding annotations.
-KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic that requirements engineer can use to precisely specify NFPs, such as performance, as part of attribute specification of functional requirements.
+\cite{ChLa09} proposed a generative framework to address NFPs for web service compositions.
+They separate the orchestration meta-model from other non-functional meta-models.
+They propose a non-functional meta-model for security  and they specify how these requirements must be ensured.
+The NFPs are  associated to the composition model by adding annotations.
+KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic  to precisely specify NFPs, such as performance, as part of attribute specification of functional requirements.
 
 The AO4BPEL \cite{CSHM06} deployment descriptor (DD) is used for specifying non-functional requirements for BPEL processes.
-First the service (one of logging, security, transactions and reliable messaging) is selected.
-Then the class (e.g.
+First the service (e.g., logging, security, transactions and reliable messaging) is selected.
+Then, the class (e.g.
 authentication or confidentiality for security) can be specified.
 The type (e.g.
-encryption or decryption for confidentiality) is then the concrete action that must be executed in order to enforce the requirement.
+encryption or decryption for confidentiality) is  the concrete action that must be executed in order to enforce the requirement.
 A selector can be defined such as for a certain BPEL activity.
@@
 
-The W3C proposes a set of languages and protocols for specifying the NFP of a ser-vices’ coordination.
+The W3C proposes a set of languages and protocols for specifying the NFP of a servicesÄô coordination.
 A classification of NFP protocols, provided by the WS-* family protocols proposed by the World Wide Web Consortium (W3C).
@@
 %
- Standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
+ Standards in services composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
+WSDL and SOAP are languages used respectively for describing services interfaces and message exchange protocols for calling methods exported by such services.
+For adding a transactional behaviour to a services composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
 The selection of the adequate protocols for adding a specific non-functional constraints to a services composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Non-functional properties in service compositions}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
 Several approaches adding non-functional properties to service compositions extend the underlying composition model.
-Non-functional properties can concern for example securty, atomicity, availability, response time.
+Non-functional properties can concern for example security, atomicity, availability, response time.
 \cite{CholletL09,SchmelingCM11}  associate non-functional  properties  to functional activities  modeled by a web service composition model. In \cite{chmelingCM11} 
-non-functional propoerties are   modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
+non-functional properties are   modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
 In \cite{OvaskaEHPA10} non-functional properties concern  quality management and they are specified at design time.
@@
  \cite{Babamir2010} proposes a framework for evaluating quality attributes of web services.
-By executing an algorithm, the framework  chooses the best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime.
+By executing an algorithm, the framework  chooses the best web services for any task in a process and also candidate web service to replace unavailable ones at runtime.
 \cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for including  QoS attributes that can be evaluated at runtime.
@@
 
-Other approaches \cite{JeongCL09,Kamalabad2012,MohantyRP10} use matching techniques for choosing services that can cope to non-functional properties required by a given application.
-\cite{AgarwalLS09} defines a policy language to specificy constraints on functional  properties of Web services for facilitating automated discovery and selection of services. 
+Other approaches \cite{JeongCL09,Kamalabad2012,MohantyRP10} use matching techniques for choosing services that can cope with non-functional properties required by a given application.
+\cite{AgarwalLS09} defines a policy language to specify constraints on functional  properties of Web services for facilitating automated discovery and selection of services. 
 \cite{Yeom2006} define a quality driven selection of web services providing a QoS model.
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Towards methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
- \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008} propose methodologies for building service-based applications that  mainly focus on modeling functional aspectos of service-based business processes.
+ \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008} propose methodologies for building service-based applications that  mainly focus on modeling functional aspects of service-based business processes.
 SOD-M \cite{decastro1} defines a model-driven approach for development service-based applications.
 Few service-oriented development methodologies    explicitly address  non functional properties modeling \cite{Tran2012531,GutierrezRF10}.
-\cite{GutierrezRF10} presents PWSSec process, a methodology that provides  activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.
+\cite{GutierrezRF10} a methodology that provides  activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.
 
-Unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that allows to model service-based applications starting from a high-level business modelling, and defining meta-models at different abstraction levels (CIM, PIM and PSM), it enables the design and development of services-based applications that can be reused and that are reliable.
-The definition of a model-driven approach  allows to model non-functional properties in the early stages of the software development process, and the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefits of our proposal.
+Unlike methodologies and approaches providing best practices presented above,  our proposal proposes a method for modelling service-based applications starting from a high-level business representation, and defining meta-models at different abstraction levels (CIM, PIM and PSM).
+It supports the design and development of services-based applications that can be reused and that are reliable.
+A model-driven approach  is adopted to model non-functional properties in the early stages of the software development process.
+Our approach also proposes  the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefit of our proposal.
 

### 2014-03-03 JournalToSubmit2013/modelingFlyingPig.tex
New version first modifications on Sections 3 and 4 ... 04.04.2014

--- 2014-02-12 JournalToSubmit2013/modelingFlyingPig.tex
@@
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%\subsection{Platform-Specific Models}
+%
+%
+%This level focusses on the functionality, in the context of a particular implementation infrastructure.
+%Models at this level combine the platform-independent view with specific aspects of the application to be deployed. 
+%At the PSM level we have lower-level models that can be automatically translated into any service composition language. 
+%We have defined one meta-model at this level, called \textit{$\pi$-PEWS}.
+%
+%The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions, including non-functional features.
+%\textit{$\pi$-PEWS} models have two main parts:
+%\begin{itemizedTrivlist}
+%\item \textit{Service compositions (workflows)} formed by individual service operations and composite operations defined by the use of workflow operators, such as guarded operations ($[C]S$); sequential ($\ .
+\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetitions.
+%
+%\item \textit{ECA Rules}, with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}. 
+%These rules are used to specify non-functional behaviour.
+%\end{itemizedTrivlist}
+%
+%The workflow constructors of \textit{$\pi$-PEWS} are similar to those of  most service orchestration languages, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
+%
+%
+%%%
+%%\begin{figure}[t]
+%%\centering
+%%\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
+%%\caption{$\pi$-{\sc Pews} Metamodel}
+%%\label{fig:metamodel}
+%%\end{figure}
+%%%
+%
+%\bigskip
+%The following example presents the \textit{$\pi$-PEWS} model for the ``To Publish Music'' application.
+%\begin{example}[To Publish Music (end)]\label{Colocar-o-label-correto-aqui}
+%The next code presents the \textit{$\pi$-PEWS}  model for our running example.\footnote{\color{red} Placido: Please change the pdf file as needed!}
+%
+%\begin{center}
+%\includegraphics[scale=0.7]{figs/pewsEspecificationContract.pdf}
+%\end{center}
+%
+%The initial part of the model consists of a workflow expression.
+%In the given code, the execution of \texttt{searchSong} is followed by \texttt{selectSong}.
+%At this point, the user may either buy, download and listen a song or just listen to a song (conditional workflow).
+%The last part of the workflow publishes the title of the song either on tweeter or facebook.
+%
+%ECA rules are specified as contracts~\cite{Meyer92}.
+%In our case, the \texttt{buySongContract} specifies conditions to be verified upon the execution of \texttt{buySong}.
+%This contract specifies pre and post-conditions, as well as time constraints.\footnote{\color{red} Placido: Please verify the pre-condition assertions!}
+%The precondition (EXPLAIN!)\dots
+%
+%In case of failure in verifying the precondition, the service \texttt{buySong} is retried.
+%
+%The postcondition ensures that the payment succeeds before .
+%
+%The time constraint indicates that the \texttt{sendConfirmation} service is invoked immediately after \texttt{buySong}.
+%\end{example}
+%
+%
+%
+%
+%Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
+%The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.
+%
+%
+%
+%
+%
+%
+%
+%The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
+%provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
+%The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
+%Instances of this meta-model are represented as UML activity diagrams. 
+%In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
+% white boxes represent classes modelling functional ones. 
+%
+%%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
+%\begin{itemizedTrivlist}
+%\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
+%An instance of this meta-class is graphically represented as a partition in the activity diagram. 
+%A collaborator can be either internal or external to the system. 
+%When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
+%
+%\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
+%A class action represents some type of transformation or processing. 
+%There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
+%\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
+%\label{fig:e-scomposition-metamodel}
+%\end{figure}
+%
+%\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
+%
+%\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+%We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
+%An {\em A-policy} is defined by attributes and rules. 
+%Intuitively, the conditions of each rule will be checked.
+%In case of no compliance, the actions defined by the rule will be performed.
+%The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
+%An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
+%\end{itemizedTrivlist}
+%
+%\begin{figure}[t]%[htpb]
+%\centering
+%\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
+%
+%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
+%
+%\caption{$\pi$-SCM for the ``To publish music'' business service.}
+%\label{fig:servicecompositionmodel}
+%\end{figure}
+%
+%\begin{example}[To Publish Music]\label{ex:toPublicMusic}
+%To illustrate the use of the $\pi$-Service Composition Meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
+%In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
+%% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
+%The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
+%Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
+%Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
+%%A call to such services must be part of the authentication protocol required by these services.
+%In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
+%In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
+%
+%{\sf\small OAuthPolicy} will implement the open authentication protocol.
+%The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
+%This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
+%The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
+%Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
+%
+%{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
+%The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
+%The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
+%On the notification of an event of that type, a credential is obtained using the username and password. 
+%\hfill\openbox
+%\end{example}
+%
+%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
+%These artifacts will be used to generate the actual programs that will implement the application:
+%Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
+%The $\pi$-PEWS  meta-model is described in the next section.
+
+%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
+%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
+%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
+%
+%
+%This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
+For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
+Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
+%
+%%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
+From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
+%%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
+%
+%%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+%%\subsection{$\pi$-SOD-M Development Environment}
+%%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+%
+%Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
+The environment implements the abstract architecture.
+Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
+%\begin{figure}[htpb]
+%	\begin{center}
+%		\includegraphics[width=0.60\textwidth]{figs/architecture}
+%	\end{center}
+%		\caption{$\pi$-SOD-M Development Environment}
+%   \label{fig:policymanager}
+%\end{figure}
+%\begin{itemize}
+%\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
+Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
+%
+%\item	 We used  ATL \footnote{http://eclipse.org/atl/.
+An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
+%for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
+%
+%\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
+We coded the pews.mt program  that implements the model to text transformation for generating executable code.
+It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
+%{\em A-policy} based services' composition execution environment. 
+%
+%%\item Finally, we created a chain execution  to execute the model to text transformation.
+%\end{itemize}
+%
+%%
+%As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
+The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
+The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
+%
+%
+%The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
+%Each activity is seen as the process where the service method call is executed.
+%The execution of an activity has four states: prepared, started, terminated, and failure.
+%The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
+%
+%At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
+Policies associated to a scope are activated when the execution of its scope starts.
+A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
+If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
+%%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
+%If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
+The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
+%%The order of policies has implications on the global order of the rules to be executed.
+%
+%
+%=============================================================================
+
 We validated our methodology by developing a use-case concerning risk assessment for financial companies.

### 2014-03-03 JournalToSubmit2013/modelingWithPISODM.tex
New version first modifications on Sections 3 and 4 ... 04.04.2014

@@
 
-In this section we present $\pi$SOD-M, an MDD-based methodology for building service compositions with non-functional requirements. 
-$\pi$SOD-M extends SOD-M with concepts for modelling non-functional requirements.
-$\pi$-SODM (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
+This section presents $\pi$SOD-M, an MDD-based methodology for building service compositions with non-functional requirements. 
+$\pi$SOD-M extends SOD-M with concepts for modelling non-functional requirements, it proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models (Figure~\ref{fig:piSOD-M}).
 \begin{figure}[h]
@@
 
-$\pi$-SODM proposes meta-models, organized into three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-These models are explained next.
+The $\pi$-SODM  meta-models are organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+These meta-models are explained next.
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Computation Independent Models}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
@@
 
-The BPMN model is a graphical representation that establishes the high-level workflow of the application.
+The BPMN model is a graphical representation that establishes the business process of the application through a high-level workflow. The next example illustrates the use of these models.
 
-The next example illustrates the use of these models.
 \begin{example}[To Publish Music]\label{ex:toPublicMusic}
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Platform Independent Models}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Platform Specific Models}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 

### 2014-03-05 JournalToSubmit2013/modelingWithPISODM.tex
Version undergoing section 3 05.04.2014

@@
-%(Aquí lo dejaría con la estructura que teníamos para Caise, presentamos el framework y posteriormente comentamos modelos y transformaciones)
+%(Aqu√≠ lo dejar√≠a con la estructura que ten√≠amos para Caise, presentamos el framework y posteriormente comentamos modelos y transformaciones)
 
@@
 The e$^3$value model identifies the value/information exchange between components of the system. 
-It defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need.
+It defines \textit{dependency paths}, showing the value exchange between providers and end users once they ask for a service.
+% , which are triggered by the occurrence of an end-consumer need.
 A dependency path has a direction and consists of a sequence of linked dependency nodes.
-A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node\footnote{See Legend on Figure~\ref{fig:CIM:tpme3v}}. 
+A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node\footnote{see Figure~\ref{fig:CIM:tpme3v}}. 
 Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
@@
 \begin{example}[To Publish Music]\label{ex:toPublicMusic}
-To illustrate our proposal, we build the models for a ``To Publish Music'' problem. 
-This problem involves three external actors ({\em Spotify, Twitter} and {\em Facebook}), as well as the user and the application itself.
-
-The e$^3$value and BPMN models for this applications are shown in Figure~\ref{fig:CIM:tpm}.
-Notice that the models at this level show\dots 
-{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
-%\hfill\openbox
+To illustrate our proposal, consider the scenario ``To Publish Music''  based on
+three external actors ({\em Spotify, Twitter} and {\em Facebook}), as well as the user and the application itself. 
+The e$^3$value and BPMN models for this application are shown in Figures~\ref{fig:CIM:tpme3v} and \ref{fig:CIM:tpmbpmn}. 
 \end{example}
 
-\begin{figure}\center
+\begin{figure}[htpb]
+\center
 \includegraphics[width=0.88\textwidth]{figs/e3value.pdf}
@@
 
-\begin{figure}\center
+The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed.
+The E3 value model in Figure \ref{fig:CIM:tpme3v} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
+The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
+Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).
+
+%Note that the models at this level show\dots 
+%{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
+%\hfill\openbox
+
+Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
+It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
+Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
+%
+\begin{figure}[htpb]
+\center
 \includegraphics[width=0.49\textwidth]{figs/SC.pdf}
@@
 Non-functional requirements are defined as \textit{constraints} over actions and data.
+The \textit{$\pi$-UseCase} meta-model describes the functional aspects of the application being developed and the restrictions that applies to each of them.
+These restrictions are modeled using the concepts: (i) constraint type and (ii) non-functional requirement with its specific attribute.
+All constraints are described by stereotyped use cases. Every relationship between an use
+case and a constraint is done through a dependency relationship (from UML).
+A constraint is defined by a stereotype, containing its type and name.
+Constraints are represented by a stereotyped use case (´constraintª).
+Each use case is related with one or more system requirements.
+Moreover, these use cases are related to non-functional requirements necessary for the proper execution of the application function.
+The function described by a use case can be part of a business service (in the case of service remote calls).
+This information is described through comments.
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
-{\color{red} Talk about this model for the example.}
+%Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.
+
+ We use part of our case study to illustrate this meta-model.
+Each use case may have multiple constraints.
+In this example the publish music use case has an authentication constraint (user must authenticate).
+For a song to be published on any social network it is necessary to provide data privacy.
+The constraint type concerns the authentication values and the constraint has an non-functional attributes.
+The use case is also related with requirement, a business service and a non-functional requirement properties.
+%{\color{red} Talk about this model for the example.}
 \end{example}
+
+\begin{figure}[htpb]
+\center
+\includegraphics[width=0.9\textwidth]{figs/UseCase.pdf}
+\caption{\label{fig:CIM:tpmbpmn} $\pi$-UseCase model for ``To Publish Music''.}
+\end{figure}
 

### 2014-03-05 JournalToSubmit2013/modelingFlyingPig.tex
... starting to touch section 4 ... 05.04.2014 22:59

@@
 The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
-By analysing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
+By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
 
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Computation-Independent Models (CIM)}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
@@
 
-Once tasks are completed, they are stored and analysed to generate a list of noncompliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
-During the process of analysing a questionnaire, the answers of some questions can trigger the generation of additional questionnaires or amendments, that will become new tasks.
+Once tasks are completed, they are stored and analyzed to generate a list of noncompliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
+During the process of analyzing a questionnaire, the answers of some questions can trigger the generation of additional questionnaires or amendments, that will become new tasks.
 
@@
 }
-
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Platform-Independent Models (PIM)}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
@@
 Specification: FlyingPig. ]!\label{fig:PiPEWSModel}!
-
  //Namespaces specify service URI
@@
 
+\normalsize
 \end{lstlisting}
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Platform-Specific Model (PSM)}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Lessons Learned}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 

### 2014-03-06 JournalToSubmit2013/implementation.tex
New organization of the paper, new version of section 3 and new section 4. Introduction and related works reviewed 06.04.2014

@@
 %%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%\subsection{$\pi$-SOD-M Development Environment}
+\subsection{General architecture}
 %%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
@@
 Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture shown in Figure \ref{fig:sodm}.
+The environment implements the abstract architecture.
 Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[htpb]
+\begin{figure}[t]
 	\begin{center}
@@
 
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+\subsection{Transformation rules}\label{sec:mmrules}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+
+Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model.
+There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.
+
+% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
+%\noindent
+
+%{\bf\em 
+\subsubsection{Transformation of the services' composition elements of the $\pi$-SCM to the $\pi$-{\sc Pews} elements}
+% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
+A named action of the $\pi$-SCM represented by  {\sc\em Action} and {\sc\em Action:name} is transformed to a  named class {\sc Operation} with a corresponding attribute name {\sc Operation:name}.
+A  named service activity represented by the elements {\sc\em ServiceActivity}  and  {\sc\em ServiceActivity:name} of the $\pi$-SCM, are  transformed into a named operation of the $\pi$-{\sc Pews} represented by the elements  {\sc CompositeOperation} and {\sc CompositeOperation:name}.
+When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-SCM the corresponding transformations, according to the PEWS operators presented above, are (see details in Figure \ref{fig:transformations}):
+\begin{itemize}
+\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
+\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
+ \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
+\end{itemize}
+
+In the scenario "To Publish Music" the service activity {\sf PublishMusic} of the $\pi$-SC model specifies  calls to two {\sf Activitie}s of type {\em UpdateMusic}, respectively concerning the {\sf Business Service}s {\em Facebook} and {\em Twitter}.
+Given that no {\sf ConstrolNode} is specified by the $\pi$-SC model, the corresponding transformation is the expression that defines a {\sf Composite Operation} named {\em PublishSong} of the $\pi$-{\sc Pews} model of the form ({\sf PublishFacebook} $\parallel$ {\sf PublishTwitter}).
+\begin{figure}
+\centering{
+%\includegraphics[width=0.80\textwidth]{figs/PI-SC-PI-P}}
+\includegraphics[width=0.96\textwidth]{figs/Mapping-1}}
+\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformation}
+\label{fig:transformations}
+\end{figure}
+
+% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
+%\noindent
+
+%{\bf\em 
+\subsubsection{Transformation of rules grouped by A-policies   in the $\pi$-SCM to A-Policies of  $\pi$-{\sc Pews}}
+% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
+The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
+The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
+The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
+
+As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
+The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
+
+%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
+
+%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
+The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
+
 %
+%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
+In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
+Thus in both cases the events are of type {\sf ActivityPrepared}.
+These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
+In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
+%\begin{figure}[htpb]
+%\centering{
+%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
+%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
+%\label{fig:p-scim}
+%\end{figure}
+
+%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
+%\begin{figure}
+%\centering{
+%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
+%\caption{Pews program implementing the "To Publish Music" application}
+%\label{fig:pewsexpression}
+%\end{figure}
+
+%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+\subsection{A-policy manager}
+%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
+
 As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.

### 2014-03-06 JournalToSubmit2013/intro.tex
New version of related works ... a lot of bibs are missing 04.04.2014
New organization of the paper, new version of section 3 and new section 4. Introduction and related works reviewed 06.04.2014

@@
 In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-combined to build an application business logic.
+composed to build an application business logic.
 The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
-An important challenge of service oriented development is  to ensure the alignment between the requirements imposed by the business logic and the application actually being developed.
+An important challenge of service oriented development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
 
 Functional properties are not the only  aspect in the software development process.
-Other non-functional properties need to be addressed to fit in the application for example  data privacy, exception handling, atomicity  and, data persistence.
-Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
+Non-functional properties need to be addressed to fit in the application for example  data privacy, exception handling, atomicity  and, data persistence.
+Adding non-functional properties and respecting services constraints while composing services is a complex task that implies programming  protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
 
-Service oriented computing promotes ease of information systems construction thanks to services reuse.
-Yet, this is not applied to non-functional constraints because they do not follow in general the same service oriented principle.
+Even if service oriented computing promotes ease of information systems construction thanks to services reuse.
+This is not applied to non-functional constraints because they do not follow  the same service oriented principle.
 %
@@
 %They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed. 
-Ideally, non-functional requirements should be considered along with all the stages of the software development.
-In fact,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
+Ideally, non-functional requirements should be considered in every phase of the software development.
+Yet,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
 In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
@@
 
-The literature stresses the need for methodologies and techniques for service oriented analysis and design, since they are the cornerstone in the development of meaningful service based applications~\cite{5}.
-Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software that serves, rather than dictates the needs of its users~\cite{watson}.
-Model Driven Development (MDD)  for the development of software systems mainly characterised by the use of models as a product \cite{Selic03}.
+The literature stresses the need for methodologies and techniques for service oriented analysis and design 
+%since they are the cornerstone in the development of meaningful service based applications
+~\cite{5}. Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software 
+%that serves, rather than dictates the needs of its users
+~\cite{watson}. Model Driven Development (MDD)  for the development of software systems is mainly characterized by the use of models as a product \cite{Selic03}.
 %The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}. 
-Model Driven Architecture (MDA)~\cite{miller} states a set of guidelines for structuring the specifications of software systems.
-The technique promotes  the use of \textit{models} to specify a software system at different levels of abstraction (called \textit{viewpoints}).
+%
+% (called \textit{viewpoints}).
 These models are successively refined from abstract specifications into actual computer programs.
 
-This paper proposes  the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
+Our work proposes  the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
 %~\cite{decastro1}, 
 to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
-$\pi$-SODM is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications. It is intended:
+$\pi$-SODM is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications.
+The technique promotes  the use of \textit{models} to specify a software system at different levels of abstraction.
+Models are organized according to the guidelines for structuring the specifications of software systems of the Model Driven Architecture (MDA)~\cite{miller} approach.
+The objective of  $\pi$-SODM are to:
 %SOD-M does not provide support for the specification of non-functional requirements, such as
@@
 \begin{trivlist}
-\item \textit{(i)} To improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\item \textit{(ii)} To reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
+\item \textit{(i)} Improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
+\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
 \end{trivlist}
@@
 
- The methodology is based on  an NFR model including a set of concepts  for  modeling  NFR in service-oriented applications.
+ This paper describes the methodology $\pi$-SODM based on  an NFP model including a set of concepts  for  modeling  NFP in service-oriented applications.
 Accordingly, the remainder of the paper is organized as follows.
-Section\dots
+Section\ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
+Section \ref{sec:motivation} introduces the meta-models of the $\pi$-SODM and transformation rules.
+Section \ref{sec:implementation} describes the $\pi$-SODM environment that provides tools for building service based applications according to the $\pi$-SODM methodology.
+Section\ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned.
+Finally, Section \ref{sec:conclusions} concludes the paper and gives research perspectives.

### 2014-03-06 JournalToSubmit2013/relworks.tex
New organization of the paper, new version of section 3 and new section 4. Introduction and related works reviewed 06.04.2014

@@
 In the case of service-based applications, non-functional requirements concern the application itself as well as its component services.
-The majority of the work on NFPs uses a product-oriented approach concerned with measuring in which extent a software system is fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
+The majority of the works on NFPs focus on measuring in which extent a software system  fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
 
-Existing  approaches address NFP from four perspectives: (i) NFP expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for associating certain non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.
+We consider that existing  approaches address NFP from four perspectives: (i) NFP classification generally expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for satisfying  non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.
 
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Classifying Non-functional properties}
+\subsection{Classifying non-functional properties}
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-We conducted a systematic review \cite{placidoPhDThesis2012} to analyze approaches that consider non-functional properties for developing service-oriented applications.
-As a result this analysis proposes a classification of NFP as a result of a study concerning software methodologies for the construction of service-oriented systems.
-The classification is organized in three layers: application modeling, services composition and services.
+We conducted a systematic review \cite{placidoPhDThesis2012} to analyze approaches that classify non-functional properties for developing service-oriented applications.
+As a result this analysis proposes a classification of NFP.
+%as a result of a study concerning software methodologies for the construction of service-oriented systems. 
+The classification is organized into three layers: application, services composition and services.
 
 \begin{itemize}
-\item At the application layer NFP  refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an "@"Äù and a "Äú."Äù.
-A value NFR expresses constraints about the way data and functions can be accessed and executed.
+\item At the application layer NFP  refer to business rules (e.g., only the user can publish data on her/his Facebook wall) and values for example, the email address is a string containing an "@"Äù and a "Äú."Äù.
+A value NFP expresses constraints about the way data and functions can be accessed and executed.
 
-\item The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP).
+\item The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements.
 
-\item Business NFP at the service layer concern properties that are associated to services and defines how to call their exported operations (business properties).
+\item  At the service layer  business NFP concern properties that are associated to services and defines how to call their exported operations (business properties).
 Value constraints concern more on the conditions in which services can be used.
 For example, accessing to a function within an authentication protocol.
-This layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.
+This layer gives an abstract view of the kind of properties associated to services that can be combined for providing NFP for a composition.
 For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
@@
 The UML Profile for Modeling Quality of Service and Fault Tolerance Characteristics and Mechanisms \cite{Omg04} describe a set of QoS characteristics such as throughput, latency, integrity, and dependability.
-The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms in case of the security characteristics.
+The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms to used for addressing  security.
 For all UML profiles based approaches it is  possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g.
 Class or Operation.
-For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services.
+For instance, there are stereotypes for encryption, decryption and logging which can be added to classes representing operations exported by services.
 The ordering or priority of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.
@@
 Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making.
-In a business rule system,  rules are logically separated   from the system \cite{ChMe04}.
+%In a business rule system,  rules are logically separated   from the system \cite{ChMe04}.
 \cite{ChMe04} investigate a hybrid approach, that combines business processes as known in BPEL with business rules \cite{Hay00}.
 
-\cite{ChLa09} proposed a generative framework to address NFPs for web service compositions.
+\cite{ChLa09} proposes a generative framework to address NFPs for web service compositions.
 They separate the orchestration meta-model from other non-functional meta-models.
@@
 The NFPs are  associated to the composition model by adding annotations.
-KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic  to precisely specify NFPs, such as performance, as part of attribute specification of functional requirements.
+KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic  to precisely specify NFPs, such as performance.
+% as part of attribute specification of functional requirements.
 
@@
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Towards methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}
+\subsection{ Methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

### 2014-03-09 JournalToSubmit2013/intro.tex
Introduction changed 10.04.2014

@@
 Section\ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
-Section \ref{sec:motivation} introduces the meta-models of the $\pi$-SODM and transformation rules.
+Sections \ref{sec:motivation}  and \ref{sec:mmrules} introduce respectively the meta-models of the $\pi$-SODM and transformation rules.
 Section \ref{sec:implementation} describes the $\pi$-SODM environment that provides tools for building service based applications according to the $\pi$-SODM methodology.
-Section\ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned.
+Section \ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned.
 Finally, Section \ref{sec:conclusions} concludes the paper and gives research perspectives.

### 2014-03-09 JournalToSubmit2013/modelingFlyingPig.tex
starting to address section FlyingPig .. 10/04/2014

@@
 \includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
-
-{\color{red} Javier: Please change \underline{Asses} by \underline{Assess}. --M}
 \caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
@@
 The BPMN model for \FlyingPig\ is devised to better understand the process in which the value exchanges occur.
-Figure~\ref{fig:BPMNmodel} shows the BPMN model\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.} for the \FlyingPig\ scenario. 
+Figure~\ref{fig:BPMNmodel} shows the BPMN model
+%\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.} 
+for the \FlyingPig\ scenario. 
 The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}. 
@@
 
-In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
-These models are built next, for the \FlyingPig\ scenario.
+%In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
+The $\pi$-SOD-M models of the PIM level  for the \FlyingPig\ scenario are presented next.
 
@@
 
-{\color{red} Anything else here? --M}
+%{\color{red} Anything else here? --M}
 
@@
 
-The model presented in Figure~\ref{fig:PiPEWSModel} is the result of translating the previous model into service composition code.
+The  PSM models of our methodology result of translating the previous model into service composition code.
 Notice that the workflow for \FlyingPig\ is implemented by using the constructors present in BPEL~\cite{BPEL}, with the addition of the notion of \textit{contract}.

### 2014-03-12 JournalToSubmit2013/intro.tex
sections 1 and 2 -- Martin & Umberto

--- 2014-03-06 JournalToSubmit2013/intro.tex
@@
 In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
-composed to build an application business logic.
+combined to build an application business logic.
 The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
-An important challenge of service oriented development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
+An important challenge of service-oriented development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
 
 Functional properties are not the only  aspect in the software development process.
-Non-functional properties need to be addressed to fit in the application for example  data privacy, exception handling, atomicity  and, data persistence.
-Adding non-functional properties and respecting services constraints while composing services is a complex task that implies programming  protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
+Non-functional properties, such as data privacy, exception handling, atomicity  and, data persistence, need to be addressed  to fit in the application.
+%Adding non-functional properties and respecting services constraints while composing services is a complex task that implies programming  protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
 
-Even if service oriented computing promotes ease of information systems construction thanks to services reuse.
-This is not applied to non-functional constraints because they do not follow  the same service oriented principle.
+Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements.
 %
@@
 Ideally, non-functional requirements should be considered in every phase of the software development.
-Yet,  they are partially or rarely methodologically derived from the application specification, and they are added once the code has been implemented.
-In consequence, the resulting application does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
+Yet,  they are partially or rarely methodologically derived from the specification, being usually added once the code has been implemented. 
+In consequence, the development process will not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
 
@@
 %since they are the cornerstone in the development of meaningful service based applications
-~\cite{5}. Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software 
-%that serves, rather than dictates the needs of its users
-~\cite{watson}. Model Driven Development (MDD)  for the development of software systems is mainly characterized by the use of models as a product \cite{Selic03}.
-%The main ideas of MDD were originally proposed by the Object Management Group (OMG) whit their proposal of a Model Driven Architecture (MDA)~\cite{miller}. 
-%
-% (called \textit{viewpoints}).
+~\cite{5}. 
+Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software~\cite{watson}.
+Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
 These models are successively refined from abstract specifications into actual computer programs.
 
-Our work proposes  the $\pi$-\textit{ Service Oriented Development Method} ($\pi$-SODM)
-%~\cite{decastro1}, 
+Our work proposes  the $\pi$-\textit{Service Oriented Development Method} ($\pi$-SODM)
 to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
-$\pi$-SODM is aligned with the MDD directives and proposes models, practices and techniques for the development of service-based applications.
-The technique promotes  the use of \textit{models} to specify a software system at different levels of abstraction.
+$\pi$-SODM is aligned with the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.
+The technique proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
 Models are organized according to the guidelines for structuring the specifications of software systems of the Model Driven Architecture (MDA)~\cite{miller} approach. 
-The objective of  $\pi$-SODM are to:
+The goals of  $\pi$-SODM are to:
 %SOD-M does not provide support for the specification of non-functional requirements, such as
@@
 \begin{trivlist}
-\item \textit{(i)} Improve the construction process by providing an abstract view of the application and ensure the conformance to its specification;
-\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, to produce concrete implementations from high abstraction models.
+\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification;
+\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
 \end{trivlist}
 
-%The main contribution of our work is a methodology for supporting the construction of service-oriented applications,
-
- This paper describes the methodology $\pi$-SODM based on  an NFP model including a set of concepts  for  modeling  NFP in service-oriented applications.
-Accordingly, the remainder of the paper is organized as follows.
-Section\ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
-Section \ref{sec:motivation} introduces the meta-models of the $\pi$-SODM and transformation rules.
-Section \ref{sec:implementation} describes the $\pi$-SODM environment that provides tools for building service based applications according to the $\pi$-SODM methodology.
-Section\ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned.
-Finally, Section \ref{sec:conclusions} concludes the paper and gives research perspectives.
+This paper is organized as follows.
+Section~\ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
+Sections~\ref{sec:motivation}  and~\ref{sec:mmrules} introduce respectively the meta-models of the $\pi$-SODM and transformation rules. 
+Section~\ref{sec:implementation} describes the $\pi$-SODM environment that provides tools for building service based applications according to the $\pi$-SODM methodology. 
+Section~\ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned. 
+Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.

### 2014-03-12 JournalToSubmit2013/relworks.tex
sections 1 and 2 -- Martin & Umberto

@@
-A non-functional property (NFP) specifies a criterion about the behavior of a system related to the conditions in which it is executed and to its performance.
-Non-functional properties are also referred to as constraints, quality attributesÄù, quality goalsÄù, Äúquality of service requirements and Äúnon-behavioral requirementsÄù \cite{StGr05}.
+While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is preformed.  
+NFRs include aspects such as performance, authentication and quality constraints.
+Non-functional properties are also referred to as constraints, quality attributesÄù, quality goalsÄù, quality of service requirements and Äúnon-behavioural requirementsÄù \cite{StGr05}. 
 In the case of service-based applications, non-functional requirements concern the application itself as well as its component services. 
-The majority of the works on NFPs focus on measuring in which extent a software system  fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
+%The majority of the works on NFPs focus on measuring in which extent a software system  fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
+Most research on NFRs focus on the evaluation of compliance by the software system.
 
-We consider that existing  approaches address NFP from four perspectives: (i) NFP classification generally expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for satisfying  non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.
+In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view, namely,  
+\textit{service level}, \textit{system level} and \textit{business level}.
+In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
+In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
+Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}.
+The work also defines a \textit{web service quality model}, which considers non-functional properties.
 
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Classifying non-functional properties}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-We conducted a systematic review \cite{placidoPhDThesis2012} to analyze approaches that classify non-functional properties for developing service-oriented applications.
-As a result this analysis proposes a classification of NFP.
-%as a result of a study concerning software methodologies for the construction of service-oriented systems. 
-The classification is organized into three layers: application, services composition and services.
+In~\cite{XiaoCZBOLH08} the authors use the terms  
+\textit{non-functional attributes}, \textit{composition mo\-del entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
+The notion of non-functional attribute is used to describe NFRs of the abstract process model. In the lower level, the composition is annotated with non-functional attributes.
 
-\begin{itemize}
-\item At the application layer NFP  refer to business rules (e.g., only the user can publish data on her/his Facebook wall) and values for example, the email address is a string containing an "@"Äù and a "Äú."Äù.
-A value NFP expresses constraints about the way data and functions can be accessed and executed.
+D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}. 
+\textit{Quality dimensions} are used to quantify an individual characteristic.
+For instance, the quality category \textit{performance} groups characteristics such as
+\textit{latency} and \textit{throughput}. 
+The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services.
+A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties as \textit{availability}, \textit{reliability} and \textit{access control}.
 
-\item The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements.
+Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolset for specifying and implementing web service compositions with support to several NFRs.
+The term \textit{non-functional concern} (NFC) is used to denote  NFRs.
+%Two aspects are considered: specification and realization.
+\textit{Non-functional concern} is a general term used to describe non-functional requirements. 
+For instance, \textit{security}, \textit{reliability}, \textit{transactional behavior} are non-functional requirements. 
+A \textit{non-functional action} represents some behavior that implements \textit{non-functional attributes}. 
+An example of \textit{non-functional action} is \textit{encryption}, which provides the implementation of the \textit{non-functional attribute} \textit{confidentiality}. 
+Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.
 
-\item  At the service layer  business NFP concern properties that are associated to services and defines how to call their exported operations (business properties).
-Value constraints concern more on the conditions in which services can be used.
-For example, accessing to a function within an authentication protocol.
-This layer gives an abstract view of the kind of properties associated to services that can be combined for providing NFP for a composition.
-For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.
-\end{itemize}
+Pastrana et al.~\cite{PastranaPK11} use the term \textit{contract} to describe non-functional requirements. 
+\textit{Contracts} may have pre-conditions, post-conditions and invariants. 
+Each contract defines \textit{assertions} associated with \textit{quality properties}. 
+Each service may have as many associated \textit{contracts} as needed.
 
-\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing existing approaches addressing NFP for service coordinations. 
-%
-%Therefore they propose:
-%\begin{itemize}
-%\item	A meta-model that introduces for characterizing NFP according to the entity to which they are associated: attribute, concern, action, and activity. 
-%\item	Six requirements for studying NFP definition approaches for service coordinations: NFP specifications, NFP actions specification, Web service subjects specification, non functional attributes execution order specification, composite Web service subjects specification, stateful non functional constraint specification.
-%\item	Seven requirements for studying NFP enforcement approaches: separation of con-cerns, transparent integration of functional and NFP, quantification, superimposi-tion, integration of NFP with distributed Web service, programming language in-dependence, Web service composition support.
-%\end{itemize}
-%
-NFP concepts are close to those defined by \cite{Souz12} but they are not organized into layers.
-They correspond essentially to the service and service composition layers.
-The enforcement requirements describe the way NFP are weaved to service coordinations.
-Existing works respect all or a subset of these requirements.
-The chosen requirements have an impact in the way NFP are enforced at execution time.
+Chollet et al.~\cite{CholletL09} associate (non-functional) \textit{quality properties} to 
+(functional) activities. The authors present a security meta-model for web service
+composition. The NFRs considered are \textit{authentication}, \textit{integrity} and \textit{confidentiality}. 
+Each NFR is associated with a service activity.
 
-%  Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase.
-Finally exiting methodology for the development of service-oriented applications including non-functional properties are here depicted and compared with our approach.
-%As a result of our study of related works, in this section we present a categorization of main concepts related the treatment if non-functional requirement in the field of service-oriented development.
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Specifying Non-functional properties}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+Ceri et al.\cite{CeriDMF07} uses the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
+Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services. 
+Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}. 
+Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
+Other authors do not define specific terms to refer to NFRs. 
+They use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09}, 
+\textit{property}~\cite{Fabra2011}, 
+\textit{factor}~\cite{MohantyRP10,GutierrezRF10}, 
+\textit{characteristic}~\cite{DiamadopoulouMPS08}, 
+\textit{quality level}~\cite{ModicaTV09}, and
+\textit{value}~\cite{ThissenW06,BasinDL06}.
 
-\cite{CyPr04} proposes an extension to UML for expressing NFPs that can be integrated into the Class, Sequence, and Collaboration Diagrams.
-UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram are annotated with quantitative performance constraints.
-The UML Profile for Modeling Quality of Service and Fault Tolerance Characteristics and Mechanisms \cite{Omg04} describe a set of QoS characteristics such as throughput, latency, integrity, and dependability.
-The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms to used for addressing  security.
-For all UML profiles based approaches it is  possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g.
-Class or Operation.
-For instance, there are stereotypes for encryption, decryption and logging which can be added to classes representing operations exported by services.
-The ordering or priority of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.
+Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
+However, distinct hierarchies and models are proposed for NFRs,  according to different points of view.
+We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.
 
- %incorporate extra-functional properties into UML models by the use of UML stereotypes. 
-%Design patterns help in documenting and communicating proven design solutions to recurring problems.
-Design patterns not only describe how to solve design problems but also document the designers‚Äô intentions - why the solution is chosen over others and what tradeoffs are made.
+Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
+This work  presents a complete development methodology, although this methodology is not centered on NFRs.
+The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account. 
+In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs, 
+whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling (BPM) is used for
+system specification, including NFRs. 
+The majority of the authors concentrate on the  modeling of service compositions, although a significant number of approaches is focused on the definition of NFR models.
 
-%\cite{GrYu01} observe that non-functional properties  (NFPs) are pervasive in the descriptions of design patterns and propose their systematic treatment in descrip-tions of patterns and when applying patterns during design.
+In the method defined in~\cite{XiaoCZBOLH08}, tasks in the process model can be 
+annotated with \textit{non-functional attributes} (NFAs). 
+NFAs are defined apart  and are concerned with data items or tasks. 
+NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.
 
-%Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application.
-According to the Business Rules Group \cite{Hay00}, a business rule  is intended to assert business structure or to control the behavior of the business \cite{Hay00}.
-Business rules are usually expressed either as constraints or in the form if conditions then action.
-The conditions are also called rule premises.
-The business rule approach encompasses a collection of terms (definitions), facts (connection between terms) and rules (computation, constraints and conditional logic) \cite{Hall01}.
-Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making.
-%In a business rule system,  rules are logically separated   from the system \cite{ChMe04}.
-\cite{ChMe04} investigate a hybrid approach, that combines business processes as known in BPEL with business rules \cite{Hay00}.
+The proposal in~\cite{ThissenW06} presents steps for  selecting services 
+by taking QoS information into account. The proposed steps are: 
+\textit{(i)} identification of relevant QoS information; 
+\textit{(ii)} identification of basic composition patterns and 
+QoS aggregation rules for these patterns; and 
+\textit{(iii)} definition of a selection mechanism of services. 
+The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
+\textit{availability}.
 
-\cite{ChLa09} proposes a generative framework to address NFPs for web service compositions.
-They separate the orchestration meta-model from other non-functional meta-models.
-They propose a non-functional meta-model for security  and they specify how these requirements must be ensured.
-The NFPs are  associated to the composition model by adding annotations.
-KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic  to precisely specify NFPs, such as performance.
-% as part of attribute specification of functional requirements.
+Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
+The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.
 
-The AO4BPEL \cite{CSHM06} deployment descriptor (DD) is used for specifying non-functional requirements for BPEL processes.
-First the service (e.g., logging, security, transactions and reliable messaging) is selected.
-Then, the class (e.g.
-authentication or confidentiality for security) can be specified.
-The type (e.g.
-encryption or decryption for confidentiality) is  the concrete action that must be executed in order to enforce the requirement.
-A selector can be defined such as for a certain BPEL activity.
-Each selector can be mapped to a requirement.
+Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
+In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.
 
-The W3C proposes a set of languages and protocols for specifying the NFP of a servicesÄô coordination.
-A classification of NFP protocols, provided by the WS-* family protocols proposed by the World Wide Web Consortium (W3C).
-%These protocols suppose an underlying communication protocol. 
-%
-%The following sections describe representative examples of the WS-* protocol-based approaches.
-%
- Standards in services composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols.
-WSDL and SOAP are languages used respectively for describing services interfaces and message exchange protocols for calling methods exported by such services.
-For adding a transactional behaviour to a services composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction.
-The selection of the adequate protocols for adding a specific non-functional constraints to a services composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.
+Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.  
+For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
+Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.
 
-% As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
-% % This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
-% Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services.
-In \cite{Vidyasankar:2004} the model introduced in \cite{SchuldtABS02} is extended to web services for addressing atomicity.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Non-functional properties in service compositions}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-Several approaches adding non-functional properties to service compositions extend the underlying composition model.
-Non-functional properties can concern for example security, atomicity, availability, response time.
-\cite{CholletL09,SchmelingCM11}  associate non-functional  properties  to functional activities  modeled by a web service composition model. In \cite{chmelingCM11} 
-non-functional properties are   modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
-In \cite{OvaskaEHPA10} non-functional properties concern  quality management and they are specified at design time.
-Quality requirements  are then associated to the components of an architecture that implements a service composition.
-
-Other approaches specify architectures or frameworks to manage and validate QoS attributes in service composition processes. 
-\cite{XiaoCZBOLH08} proposes a framework to verify NFP compliance in composed services.
-Tasks in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time.
-\cite{SchmelingCM11} uses a formal method  to choose best suited services instances according to a service composition pattern that includes some QoS attributes. 
- \cite{Babamir2010} proposes a framework for evaluating quality attributes of web services.
-By executing an algorithm, the framework  chooses the best web services for any task in a process and also candidate web service to replace unavailable ones at runtime.
-\cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for including  QoS attributes that can be evaluated at runtime.
- \cite{Karunamurthy2012787} proposes an architecture in which web services are described with non-functional characteristics that are validated in the service composition process.
-
-Other approaches \cite{JeongCL09,Kamalabad2012,MohantyRP10} use matching techniques for choosing services that can cope with non-functional properties required by a given application.
-\cite{AgarwalLS09} defines a policy language to specify constraints on functional  properties of Web services for facilitating automated discovery and selection of services. 
-\cite{Yeom2006} define a quality driven selection of web services providing a QoS model.
-
-%\cite{} works also on the selection of web service for use considering non-functional properties.
-The paper present web services quality prediction models, which take non-functional properties into account.
-
-Our proposal in contrast, focuses on the design of service-oriented application with non-functional properties modelled  for services and service compositions and then ensured at execution thanks to appropriate environments.
-
-%We propose to model non-functional properties  represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{ Methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
- \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008} propose methodologies for building service-based applications that  mainly focus on modeling functional aspects of service-based business processes.
-SOD-M \cite{decastro1} defines a model-driven approach for development service-based applications.
-Few service-oriented development methodologies    explicitly address  non functional properties modeling \cite{Tran2012531,GutierrezRF10}.
-\cite{GutierrezRF10} a methodology that provides  activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.
-
-Unlike methodologies and approaches providing best practices presented above,  our proposal proposes a method for modelling service-based applications starting from a high-level business representation, and defining meta-models at different abstraction levels (CIM, PIM and PSM).
-It supports the design and development of services-based applications that can be reused and that are reliable.
-A model-driven approach  is adopted to model non-functional properties in the early stages of the software development process.
-Our approach also proposes  the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefit of our proposal.
-
-%\textit{Design by Contract} \cite{HL05TACoS} is an approach for specifying web services and verifying them through runtime checkers before they are deployed.
-A contract adds behavioral information to a service specification, that is, it specifies the conditions in which methods exported by a service can be called.
-Contracts are expressed using the language \textit{jmlrac} \cite{LeavensCCRC02}.
-The \textit{Contract Definition Language} (CDL) \cite{cdl2006} is a XML-based description language, for defining contracts for services.
-There are an associated architecture framework, design standards and a methodology, for developing applications using services.
-A services' based application  specification is generated after several \cite{AbrialLNSS91} B-machines refinements that describe the services and their compositions.
-
-%\cite{PapazoglouH06} proposes a methodology based on a SOA extension.
-This work defines a service oriented business process development methodology with phases for business process development.
-The whole life-cycle is based on six phases: planning, analysis and design, construction and testing, provisioning, deployment, and execution and monitoring.
-%IBM proposes a methodology for the development of SOA solutions, called SOMA \cite{soma}.
-SOMA defines a life-cycle with seven phases: business modeling and transformation, solution management, identification,
-%specification, realization, implementation and deployment monitoring and management.
-
-%\cite{sommerville08} describes some key points for building services' based applications based on business process models that define the activities and information exchanged in a business processes.
-Activities in business process can be performed by services so that the model of business process
-%represents a composition of services. It classifies services   as public utilities, business services or
-%composition services.
-Software development that uses services involves creating programs for composing and configuring services to create new composite services.
-The service engineering process involves identifying services candidates, service interface and implementation definition, testing and deployment of each service.
-
-%Besides the notion of model at each level of abstraction, MDA propose the use of \textit{model transformations} within and between levels.
-Intra-level transformations are used to provide a unified representation of concepts of a given level.
-Inter-level transformation implement a refinement process between levels.
-Transformations may be automatic or semi-automatic.
-
-%The methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements presented in this work extends the \textit{Service Oriented Development Method} (SOD-M) proposed by De Castro et al.\cite{decastro1}.
+\bigskip
+Most works agree on distinguishing three points of view, namely the point of view of the organization (or Business view), of the individual service providers (or Service view) and of the composition designer (or System view).
+The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
+Service and System views are concerned with the implementation of the software solution: The Service level is concerned with the building blocks of the application.
+It may use web services provided by third party sources.
+The System level is concerned with the coordination of services, to implement the business logic.

### 2014-03-25 JournalToSubmit2013/intro.tex
Section 3.3 and 4.1

@@
 combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{1,2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
+The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
 An important challenge of service-oriented development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
@@
 %since they are the cornerstone in the development of meaningful service based applications
-~\cite{5}. 
+~\cite{Papazoglou2007}. 
 Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software~\cite{watson}.
@@
 to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
-$\pi$-SODM is aligned with the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.
+\pisodm is aligned with the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.
 The technique proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
 Models are organized according to the guidelines for structuring the specifications of software systems of the Model Driven Architecture (MDA)~\cite{miller} approach. 
-The goals of  $\pi$-SODM are to:
+The goals of  \pisodm are to:
 %SOD-M does not provide support for the specification of non-functional requirements, such as
@@
 Section~\ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
-Sections~\ref{sec:motivation}  and~\ref{sec:mmrules} introduce respectively the meta-models of the $\pi$-SODM and transformation rules. 
-Section~\ref{sec:implementation} describes the $\pi$-SODM environment that provides tools for building service based applications according to the $\pi$-SODM methodology. 
+Sections~\ref{sec:motivation}  and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules. 
+Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology. 
 Section~\ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned. 

### 2014-03-25 JournalToSubmit2013/relworks.tex
Section 3.3 and 4.1

@@
 NFRs include aspects such as performance, authentication and quality constraints.
-Non-functional properties are also referred to as constraints, quality attributesÄù, quality goalsÄù, quality of service requirements and Äúnon-behavioural requirementsÄù \cite{StGr05}. 
+These requirements are usually specified by conditions, called Non-functional Properties.
+Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioural requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
+Most research on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
 In the case of service-based applications, non-functional requirements concern the application itself as well as its component services.
-%The majority of the works on NFPs focus on measuring in which extent a software system  fulfills the NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
-Most research on NFRs focus on the evaluation of compliance by the software system.
 
@@
 \textit{service level}, \textit{system level} and \textit{business level}.
-In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
-In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
-Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}.
+%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
+%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
+%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}.
 The work also defines a \textit{web service quality model}, which considers non-functional properties.
-
 In~\cite{XiaoCZBOLH08} the authors use the terms  
 \textit{non-functional attributes}, \textit{composition mo\-del entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
-The notion of non-functional attribute is used to describe NFRs of the abstract process model. In the lower level, the composition is annotated with non-functional attributes.
+The notion of non-functional attribute is used to describe NFRs of the abstract process model. 
+In the lower level, the composition is annotated with non-functional attributes.
 
@@
 Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolset for specifying and implementing web service compositions with support to several NFRs. 
-The term \textit{non-functional concern} (NFC) is used to denote  NFRs.
-%Two aspects are considered: specification and realization.
-\textit{Non-functional concern} is a general term used to describe non-functional requirements. 
-For instance, \textit{security}, \textit{reliability}, \textit{transactional behavior} are non-functional requirements. 
-A \textit{non-functional action} represents some behavior that implements \textit{non-functional attributes}. 
-An example of \textit{non-functional action} is \textit{encryption}, which provides the implementation of the \textit{non-functional attribute} \textit{confidentiality}. 
-Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.
+The approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
+Non-functional concern is a general term used to describe non-functional requirements (such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}). 
+Each concern is refined into a set of non-functional attributes.  
+Each non-functional attribute represents some behavior, to be refined into non-functional actions. 
+An example of non-functional action is \textit{encryption}, which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
+%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.
 
-Pastrana et al.~\cite{PastranaPK11} use the term \textit{contract} to describe non-functional requirements. 
-\textit{Contracts} may have pre-conditions, post-conditions and invariants. 
+Pastrana et al.~\cite{PastranaPK11} use a traditional design-by-contract approach~\cite{Meyer97}.
+They use the term \textit{contract} to describe non-functional requirements. 
+Contracts may have pre-conditions, post-conditions and invariants. 
 Each contract defines \textit{assertions} associated with \textit{quality properties}. 
@@
 Chollet et al.~\cite{CholletL09} associate (non-functional) \textit{quality properties} to 
-(functional) activities. The authors present a security meta-model for web service
-composition. The NFRs considered are \textit{authentication}, \textit{integrity} and \textit{confidentiality}. 
+(functional) activities. 
+The authors present a security meta-model that takes into account web service compositions. 
+The NFRs considered are \textit{authentication}, \textit{integrity} and \textit{confidentiality}. 
 Each NFR is associated with a service activity.
 
-Ceri et al.\cite{CeriDMF07} uses the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
+Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
 Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services. 
@@
 Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
-Other authors do not define specific terms to refer to NFRs. 
-They use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09}, 
+Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09}, 
 \textit{property}~\cite{Fabra2011}, 
@@
 
-Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
-This work  presents a complete development methodology, although this methodology is not centered on NFRs.
-The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account. 
+%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications. 
+%The work  presents a development methodology, although this methodology is not centered on NFRs.
+The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development processes that take NFRs into account. 
 In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs, 
-whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling (BPM) is used for
+whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modelling is used for
 system specification, including NFRs. 
-The majority of the authors concentrate on the  modeling of service compositions, although a significant number of approaches is focused on the definition of NFR models.
+Most authors focus their work on modelling service compositions, although a significant number of approaches is focused on the definition of NFR models.
 
-In the method defined in~\cite{XiaoCZBOLH08}, tasks in the process model can be 
-annotated with \textit{non-functional attributes} (NFAs). 
-NFAs are defined apart  and are concerned with data items or tasks. 
+In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be 
+annotated with functional and non-functional attributes (NFAs). 
+Functional and non-functional attributes are independently defined.
+They are attached to specific tasks later in the development of the application. 
 NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.
@@
 \bigskip
-Most works agree on distinguishing three points of view, namely the point of view of the organization (or Business view), of the individual service providers (or Service view) and of the composition designer (or System view).
+Most works agree on distinguishing three points of view, namely the point of view of the organization (or \textit{Business} view), of the individual service providers (or \textit{Service} view) and of the composition designer (or \textit{System} view).
 The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).

### 2014-04-01 JournalToSubmit2013/modelingFlyingPig.tex
Working on FLyingPig

@@
 Our application is inspired on the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied the client's business practice.
-Examples of business practices are the conditions and protocols used to perform confidential transactions or the physical security for accessing reserved areas (such as computing server installations).
-The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amends to these practices.
+Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
+Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security for accessing reserved areas (such as computing server installations).
+The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendmets to these practices.
 The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.

### 2014-04-01 JournalToSubmit2013/modelingFlyingPig.tex
Flying pig revisited ..

@@
 
-%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-%
-%
-%This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
-For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
-%
-%%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-%%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
-%
-%%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%%\subsection{$\pi$-SOD-M Development Environment}
-%%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-%
-%Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture.
-Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-%\begin{figure}[htpb]
-%	\begin{center}
-%		\includegraphics[width=0.60\textwidth]{figs/architecture}
-%	\end{center}
-%		\caption{$\pi$-SOD-M Development Environment}
-%   \label{fig:policymanager}
-%\end{figure}
-%\begin{itemize}
-%\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-%
-%\item	 We used  ATL \footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-%for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
-%
-%\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
-We coded the pews.mt program  that implements the model to text transformation for generating executable code.
-It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
-%{\em A-policy} based services' composition execution environment. 
-%
-%%\item Finally, we created a chain execution  to execute the model to text transformation.
-%\end{itemize}
-%
-%%
-%As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
-The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-%
-%
-%The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
-%Each activity is seen as the process where the service method call is executed.
-%The execution of an activity has four states: prepared, started, terminated, and failure.
-%The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
-%
-%At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-%%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
-%If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
-%%The order of policies has implications on the global order of the rules to be executed.
-%
-%
-%=============================================================================
-
-We validated our methodology by developing a use-case concerning risk assessment for financial companies.
-Our application is inspired on the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
+We validated our methodology by developing a use-case concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
 Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
 Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security for accessing reserved areas (such as computing server installations).
-The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendmets to these practices.
+The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
 The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
@@
 Our goal is to model a service based application (called \FlyingPig), for providing risk assessment as a service.
-In order to provide this functionality, \FlyingPig\ would benefit from ORCA's legacy services for storage, assessment and data visualization functions.
+In order to provide this functionality, \FlyingPig\ would benefit from ORCA's legacy services providing storage, assessment and data visualization functions.
 
-In the following, we describe the results of applying $\Pi$-SODM to develop the \FlyingPig\ risk assessment system.
+In the following sections, we describe the results of applying $\Pi$-SODM to develop the \FlyingPig\ risk assessment system.
 The models presented next were generated as a result of interacting with software developers at GCP Global.
@@
 Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
-It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).\footnote{\color{red} This sentence need to move to section 3.1!!}
+It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
+%\footnote{\color{red} This sentence need to move to section 3.1!!}
 We identify two business actors: \textsl{ORCA} and \textsl{Broker}. 
-Brokers channel requests for risk assessment for one or more companies. 
-ORCA have two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}. 
+Brokers emit requests for risk assessment for one or more companies. 
+ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}. 
 The values exchanged between ORCA and the brokers are: 
@@
 
-The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment by a particular company. 
-Once this need occurs, the value exchanges between ORCA and Broker are triggered. 
+The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emited by a particular company. 
+Once this need has been declared, the value exchanges between ORCA and Broker are triggered. 
 The client company will provide ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
@@
 
-The BPMN model for \FlyingPig\ is devised to better understand the process in which the value exchanges occur.
 Figure~\ref{fig:BPMNmodel} shows the BPMN model
 %\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.} 
-for the \FlyingPig\ scenario. 
+for the \FlyingPig\ scenario. The BPMN model for \FlyingPig\ is partitioned for better understanding the process in which the value exchanges occur.
 The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}. 
 Brokers have two lanes, the client \textsl{Company} and a \textsl{User}. 
-The user is a contact member of the company, who will coordinate the assessment process. 
-This process will involve other members of the company as well.
+The user is a contact member of the company, who  coordinates the assessment process. 
+This process  involves other members of the company as well.
 
 The risk assessment process starts after a request from a company.
-(This agrees with the value model, in which the start stimulus triggers the whole process.)
+This corresponds to the value model, in which the start stimulus triggers the whole process.
 The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
 Questionnaires are considered tasks that users will have to perform. 
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include from physical facts such as not having easy access to handicapped persons or having an unsecured access to the premises of the company, to more intangible ones, such as the use of an less-than-optimal protocol to access data on the company's computer server.}.
+Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecure access to the premises of the company.
+They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
 
-Once tasks are completed, they are stored and analyzed to generate a list of noncompliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
-During the process of analyzing a questionnaire, the answers of some questions can trigger the generation of additional questionnaires or amendments, that will become new tasks.
+Once tasks have been completed, they are stored and analyzed to generate a list of noncompliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
+During the process of analyzing a questionnaire, the answers of some questions can trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
 
-Business processes also have associated rules and constraints to define their non-functional requirements (NFR).
-{\color{red}
-Placido, Valeria: What can we say about non-functional requirements at the CIM level?
-}
+Business processes also have associated rules and constraints to define their non-functional requirements (NFR):
+\begin{numtrivlist}
+\item An acknowledgement is due less than 30 seconds after registering a task or demand for assessment. 
+\item The system should be able to deal with, at least, 200 users. 
+\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests. 
+\item The proviledges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
+\item The privileges of Users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
+\item All questionnaires need to be fully answered, in order to consider that a task is completed.
+\item There is a time limit (in days) for each amendment required by the system.
+\end{numtrivlist}
+
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
@@
 {\color{red} \raggedright
-$\star$ ``Create a responsible'' should be changed for ``Designate user in charge''.
+$\star$  "Assign user in charge".
 
-$\star$ ``requires concurrency for more than 200 users'' should be changed by ``Maximum number of users should be greater than 200''.
+$\star$ "Maximum number of users should be greater than 200".
 }
@@
 The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application. 
-In this model, three actors are identified: \textit{Company}, \textit{User} and \textit{Channel-Broker}\footnote{\color{red} Can we change this for Broker ? --M.}. 
-They are represented as stick figures.
-In the context of \FlyingPig, Company is the actor asking for risk evaluation.
-A Channel-Broker is the responsible for channelling the the evaluation process, assigning users to be in charge of tasks as well as delegating tasks. 
-A User, in this model is an actor who answers questionnaires (with base on the actual facts about the Company).
+In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
+In the context of \FlyingPig, Company is the actor requesting a risk evaluation.
+A Broker is  responsible for coordinating  the evaluation process, assigning users to be in charge of tasks as well as delegating tasks. 
+A User, in this model is an actor who answers questionnaires according to the current situation  of the Company.
 The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
 
-Each actor is associated to one or more use cases (depicted as white ovals in Figure~\ref{fig:piUseCaseModel}). 
-Use cases describe the main functionalities of the system.
-The $\pi$-UseCase model for \FlyingPig\ defines six use cases.
+Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Figure~\ref{fig:piUseCaseModel}). 
+%Use cases 
+that describe the main functionalities of the system.
+The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.
 
-In our model, each use case may be associated to one or more (non-functional) constraints (depicted as coloured ovals in Figure~\ref{fig:piUseCaseModel}). 
-The model defines three types of constraints: \textit{value} , \textit{business} or \textit{exceptional behavior}. 
+In our model, each $\pi$-use case may be associated to one or more (non-functional) constraints (depicted as coloured ovals in Figure~\ref{fig:piUseCaseModel}). 
+The model defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}. 
 Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
 
-In the case of \FlyingPig, the model counts seven constraints:
-\begin{numtrivlist}
-\item An acknowledgement is due less than 30 seconds after registering a task or demand for assessment. 
-\item The system's infrastructure should be prepared to deal with, at least, 200 users. 
-\item If the number of requests exceeds 200, \FlyingPig\ should make a load balance of the requests. 
-\item The Channel-Broker must have their privileges verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} use case.
-\item Users must have their privileges verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} use case.
-\item All questionnaires need to be fully answered, in order to consider that a task is completed.
-\item There is a time limit (in days) for each amendment required by the system.
-\end{numtrivlist}
-
+%In the case of \FlyingPig, as expressed above the model constraints:
 %{\color{red} Anything else here? --M}
@@
 The $\pi$-ServiceProcess model (Figure~\ref{fig:PiServiceProcessModel}) presents the workflow for \FlyingPig.
-The actions in this model were obtained by applying the use case transformation rules described in Section~\ref{sec:pewsmetamodel}.
+The actions in this model were obtained by applying the $\pi$-use case transformation rules.
+% described in Section~\ref{sec:pewsmetamodel}.
 
@@
 Use cases are transformed into \textit{actions} and are represented by white boxes.
-The restrictions associated to each use case are transformed into \textit{assertions} (represented by coloured boxes) and may be decorated with pre- and post-conditions. 
+The restrictions associated to each $\pi$-use case are transformed into \textit{assertions} (represented by coloured boxes) and may be decorated with pre- and post-conditions. 
 We can see that this model refines the concepts defined in the $\pi$-UseCase model.
@@
 
-The model presented in Figure~\ref{fig:PiServiceCompositionModel} offers a system-centric view of the application.
-The previous model is now enhanced with a view of the \FlyingPig\ application, including those services upon  it is build.
+The model presented in Figure~\ref{fig:PiServiceCompositionModel} 
+%offers a system-centric view of the application.
+%The previous model 
+is now  completed with the services that will provide some functions of the application  \FlyingPig\.
 The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
-These policies contain the pre- and post-conditions of the previous model.
-They are associated to actions of the system.
+These policies express the pre- and post-conditions of the previous model.
+They are associated to the actions of the system.
 
@@
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform-Specific Model (PSM)}
+%\subsection{Platform-Specific Model (PSM)}
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
-The  PSM models of our methodology result of translating the previous model into service composition code.
-Notice that the workflow for \FlyingPig\ is implemented by using the constructors present in BPEL~\cite{BPEL}, with the addition of the notion of \textit{contract}.
+The  PSM models of our methodology result by transforming the $\pi$-service composition model into PEWS code.
+Note that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL} adding the notion of \textit{contract}.
 
-Contracts may be implemented as transversal concerns~\cite{aspects}.
-Their conditions may be verified at specific point of the composition and their actions are executed if needed.
+%Contracts may be implemented as transversal concerns~\cite{aspects}.
+%Their conditions may be verified at specific point of the composition and their actions are executed if needed.
 

### 2014-04-02 JournalToSubmit2013/intro.tex
sections 4 and 5

@@
 
-Our work proposes  the $\pi$-\textit{Service Oriented Development Method} ($\pi$-SODM)
+Our work proposes  the $\pi$-\textit{Service Oriented Development Method} (\pisodm)
 to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.

### 2014-04-02 JournalToSubmit2013/transformationRules.tex
sections 4 and 5

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..

\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models. 
These rules are used to build a semi-automatic tool.
The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-UseCase to the $\pi$-ServiceProcess}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .

The refinement of a (composite) $\pi$-Use Case model into a $\pi$-Service Process model is driven by the principle of expressing a set of $\pi$-use cases (i.e., use cases with constraints)   in terms of  a business process.
The resulting model consists of actions related by a control flow and contracts specifying NFPs.

As defined by the $\pi$-ServiceProcess  meta-model (Figure~\ref{fig:CIM:serviceprocessmetamodel}) an {\sc Activity Service} consists of a composition of  entities of type {\sc Action}. 
Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
Every {\sf Extend} relationship identified in a $\pi$-Use Case model is transformed into a  {\sf Fork node}\footnote{Notice that Fork nodes are used both to represent alternative and parallel execution. {\color{red} \sc Placido, please check this!}}.
If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/35}}
%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
%\label{fig:transformationsUseCase-ServiceProcessRules}
%\end{figure}

A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
The set of resulting assertions  are grouped into a {\sf Contract}.
Constraints are transformed according to their type:
{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} attribute set to false are transformed into {\sf Assertion}s;
{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behaviour}s.

In order to transform constraints of type {\sf Value Constraint}, the designer must specify thresholds to be associated to the assertions of a contract.
By default, value constraints are transformed into pre-conditions and business constraints are transformed into post-conditions.

%A {\sf Requirement} and a {\sf Composite use case} in a  $\pi$-UseCase model are transformed respectively into a {\sf Service activity} and an {\sf Action} (see figures  \ref{fig:transformationsUseCase-ServiceProcessRules}-c  and d). 
%
%
%\begin{figure}
%\centering{
%\includegraphics[width=0.80\textwidth]{figs/37}}
%\caption{ Extended Transformation Examples}
%\label{fig:transformaton-examples}
%\end{figure}

%The transformations for {\sc Extend} and {\sc Include} dependencies  are not as simple as the previous transformations (figures \ref{fig:transformaton-examples}-a).

The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  
The generated workflow is composed by {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/36}}
%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
%\label{fig:transformationsUseCase-ServiceProcessRules}
%\end{figure}

{\sf Include} use case entities are transformed into an {\sf Action} sequence.
A {\sc Use case} element is transformed into an {\sf Action}. 
A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements.

The details of these transformations are not included here (due to space restrictions).
They can be found in~\cite{SouzaNeto:2012}.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1}
The rules presented above have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).

The ``listen music'' use case is transformed into a Service Action. 
This Action  represents a Spotify service function that can be invoked to play the music. 
For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}. 
The ``download music'' use case  includes the payment process to buy the music. 
Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
%(as shown in figure \ref{fig:transformation-example-include}). 
The same rule is applied for the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
 \end{example}

%\begin{figure}
%\centering{
%\includegraphics[width=0.76\textwidth]{figs/38}}
%\caption{ Include Transformation Example}
%\label{fig:transformation-example-include}
%\end{figure}

PARAMOS AQUIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIII

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
%The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model. The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.   The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.

%As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}. 
%The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.

%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.

%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}. The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.

%
%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
%In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario. Thus in both cases the events are of type {\sf ActivityPrepared}. These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}. In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
%\begin{figure}[htpb]
%\centering{
%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
%\label{fig:p-scim}
%\end{figure}

%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
%\begin{figure}
%\centering{
%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
%\caption{Pews program implementing the "To Publish Music" application}
%\label{fig:pewsexpression}
%\end{figure}
%Table \ref{fig:ServiceProcess-ServiceComposition}  shows the transformation rules for transforming a .
As illustrated in Figure \ref{fig:ServiceProcess-ServiceComposition-Rules} the  principle of the transformation  of a  $\pi$-Service process model into a   $\pi$-ServiceComposition model is to group respectively  {\sf Contracts}  and {\sf Actions} into {\sf Policies} and {\sf Service activities}.

Each {\sf Assertion} of a {\sf Contract} in a $\pi$-service composition model is transformed into a {\sf Rule} in a $\pi$-service composition model. The set of {\sf Rules}  concerning the same NFP (e.g., safety, performance, and reliability) is grouped into a {\sf Policy} (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-a). Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute. If the {\sf Assertion} has a value type, the name and the attributes are transformed into a {\sf Variable} in the target model.  The {\sf Assertion: Aproperty} attribute can have different transformations, according to  the following rules:
\begin{itemize}
\item a {\sf Post-Condition} is transformed into a {\sf  Post}. 
\item a {\sf Precondition} is transformed into {\sf Pre}.
\item a {\sf Timerestriction} is transformed into {\sf Time}.
\end{itemize}

\begin{figure}
\centering{
\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
\label{fig:ServiceProcess-ServiceComposition-Rules}
\end{figure}

A {\sf Package}  in the $\pi$-use case model is transformed   into a {\sf Business Collaborator}. 
{\sf Actions}  of a $\pi$-service process model are transformed into an {\sf Action} in a $\pi$-service composition model, and every {\sf Service activity} of a $\pi$-service process model is transformed into a {\sf Service activity} in a $\pi$-service composition model (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-b). An {\sf Exceptional behaviour} entity is transformed into an {\sf Action} in a $\pi$-service composition model (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-c ), and every {\sf Non-functional attribute} associated to an element ({\sf Contract} and {\sf Non-functional requirement}) in a $\pi$-service process model becomes a {\sf Non-functional attribute} associated to the corresponding element {\sf Policy} in a $\pi$-service composition model.
 Finally, {\sf Actions} are grouped by  a  {\sf Business collaborator}.

%\begin{figure}
%\centering{
%\includegraphics[width=1\textwidth]{figs/ServiceProcess-ServiceComposition}}
%\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
%\label{fig:ServiceProcess-ServiceComposition}
%\end{figure}

%As the $\pi$-ServiceComposition model refines the $\pi$-ServiceProcess concepts at PIM level, a service previously defined as actions (source model) is refined as composition of those actions (target model) that are necessary to represent a business service, identifying who are the partners involved in the realization ({\sc Business collaborators}). 
%In addition, $\pi$-SOD-M defines a platform specific model based on web services composition. This model is explicitly indicates those actions which are (or will be, if not yet implemented) supported by web services.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
Considering the scenario example, 
%the {\sf Action}s update music {\sc Contract} is transformed is a {\sc Policy} with its {\sc Rules}. All contract {\sc Assertions} are transformed in RULE and its attributes, e.g. the login and password verification. 
the "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
%The {\sc Non- functional requirement} information will be used to the {\sc Policy} generation comes from the initial use case model. Also 
The {\sf Business Collaborator} Facebook and Spotify information come from entity of type {\sc Package}  in the $\pi$-use case model. 
%All {\sc Contracts} of the same {\sc Non- functional requirement} are composed in a {\sc Policy}.
\end{example}

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-ServiceComposition to $\pi$-PEWS}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .

%Table \ref{fig:ServiceComposition-Pews-Rules} 
This section describes the PIM to PSM transformations from a $\pi$-Service composition model to a $\pi$-PEWS program. We propose two groups of rules: those that transform services composition entities of a $\pi$-Service composition model into $\pi$-PEWS model; and those that transform rules grouped by policies into A-policy types.

A named action of the $\pi$-service composition model represented by  {\sf Action} and {\sf Action:name} is transformed to an  {\sf Operation} with a corresponding attribute name {\sf Operation:name}. A  named service activity represented by the elements {\sf ServiceActivity}  and  {\sf ServiceActivity:name} of the $\pi$-service composition model, are  transformed into a named operation of the $\pi$-{Pews} represented by the entities  {\sf CompositeOperation} and {\sf CompositeOperation:name}. When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-service composition model the corresponding transformations, according to the PEWS operators presented above:
\begin{itemize}
\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
 \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
\end{itemize}

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/Table7}}
%\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
%\label{fig:ServiceComposition-Pews-Rules}
%\end{figure}

The A-policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy} entities, named according to the names expressed in the source model. The transformation of the rules expressed in a $\pi$-service composition is guided by the event types associated to these rules. The variables associated to an A-policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into entities  {\sf Variable} with attributes {\sf Name} and {\sf Type} directly specified from the elements {\sf Variable:name} and {\sf Variable:type} of a $\pi$-service composition model.

%As shown in Table \ref{fig:ServiceComposition-Pews-Rules}, 
For an event of type Pre the corresponding transformed rule is a {\sf Precondition}; for an event of type Post the corresponding transformed rule is a {\sf Post- condition}; finally, for an event of type TimeRestriction the corresponding transformed rule is a {\sf Time}. The condition expression of a rule in a $\pi$-service composition ({\sf Rule:condition}) is transformed into a  {\sf Condition:expression} where the attributes of the expression are transformed into  entities {\sf Attribute}.

Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.

\begin{figure}
\centering{
\includegraphics[width=0.96\textwidth]{figs/31-31}}
\caption{$\pi$-PEWS Specific and Contract Representation}
\label{fig:Specific-Contract-Representation}
\end{figure}

\subsection{Implementation}
We have provided tools for aiding the user into transform models for all the levels, except for the CIM into PIM level, which should be manually performed.

This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions. For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application. Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.

%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}. From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.

%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{General architecture}
%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it. The environment implements the abstract architecture. Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
\begin{figure}[t]
	\begin{center}
		\includegraphics[width=0.60\textwidth]{figs/architecture}
	\end{center}
		\caption{$\pi$-SOD-M Development Environment}
   \label{fig:policymanager}
\end{figure}
\begin{itemize}
\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}. Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).

\item	 We used  ATL \footnote{http://eclipse.org/atl/. An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).

\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in. We coded the pews.mt program  that implements the model to text transformation for generating executable code. It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the 
{\em A-policy} based services' composition execution environment.

%\item Finally, we created a chain execution  to execute the model to text transformation.
\end{itemize}

%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{A-policy manager}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--

As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.  The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions. The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).

The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
Each activity is seen as the process where the service method call is executed.
The execution of an activity has four states: prepared, started, terminated, and failure.
The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.

At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).  Policies associated to a scope are activated when the execution of its scope starts. A {\em A-policy} will have to be executed only if one or several of its rules is triggered. If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section. 
%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
If rules belonging to several policies are triggered then policies are also ordered according to an execution plan. The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
%The order of policies has implications on the global order of the rules to be executed.

### 2014-04-04 JournalToSubmit2013/transformationRules.tex
section 4

@@
 
-PARAMOS AQUIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIII
-
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
@@
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-%The {\em A-policies} defined for the elements of the $\pi$-SCM are transformed into {\sc A-Policy} classes, named according to the names expressed in the source model.
-The transformation of the rules expressed in the $\pi$-SCM is guided by the event types associated to these rules.
-The variables associated to an {\em A-policy} expressed in the $\pi$-SCM as {\sc\em $<$Variable:name, Variable:type$>$} are transformed into elements of type {\sc Variable} with attributes {\sc name} and {\sc type} directly specified from the elements {\sc\em  Variable:name} and {\sc\em Variable:type} of the $\pi$-SCM model.
 
-%As shown in Figure \ref{fig:transformations}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
-%The condition expression of a rule in the $\pi$-SCM ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.
+The  principle of the transformation  of a  $\pi$-Service Process model into a $\pi$-Service Composition model is to group {\sf Contracts} into {\sf A-Policies} and {\sf Actions} into {\sf Service Activities}.
 
-%The attribute event of a rule  ({\sc\em Rule:event}) in the $\pi$-SCM is transformed into an {\sc Event Type} according to the rule type.
-
-%As shown in Figure \ref{fig:transformations}, the event type for a rule of type (i) {\sc Precondition} is {\sc ActivityPrepared}; (ii) {\sc Postcondition} is {\sc TermActivity}; (iii) {\sc TimeRestriction} is {\sc Temporal}.
-The {\sc\em Rule:Action} of a rule in the $\pi$-SCM is transformed into an {\sc Action:type}.
+Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule}. 
+{\sf Rules}  concerning the same NFP  are grouped into {\sf A-Policies}. 
+Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute. 
+If the {\sf Assertion} has a value type, the name and the attributes are transformed into {\sf Variables} in the target model.  
+The {\sf Assertion: aProperty} attribute can have different transformations, according to: 
+\textit{(i)} a {\sf Precondition} is transformed into {\sf Pre};
+\textit{(ii)} a {\sf Post-Condition} is transformed into a {\sf  Post};
+\textit{(iii)} a {\sf TimeRestriction} is transformed into {\sf Time}.
 
 %
-%Figure \ref{fig:p-scim} shows the  $\pi$-{\sc Pews} model for our example.
-%In the scenario "To Publish Music" the {\sf Policies} {\em OAuthPolicy} and {\em HTTPAuthPolicy} of the $\pi$-SCM model are transformed into {\em A-policies} of type {\sf Precondition} of the $\pi$-{\sc Pews} model of the scenario.
-Thus in both cases the events are of type {\sf ActivityPrepared}.
-These policies, as stated in the $\pi$-SCM model, are associated to {\sf Activities}.
-In the corresponding transformation they are associated to {\sf Operation}s {\em PublishFacebook} and {\em PublishTwitter}.
-%\begin{figure}[htpb]
-%\centering{
-%\includegraphics[width=0.78\textwidth]{figs/modeloPEWS}}
-%\caption{$\pi$-{\sc Pews} generated model fo the "To Publish Music" application}
-%\label{fig:p-scim}
-%\end{figure}
-
-%Figure \ref{fig:pewsexpression} shows the correspondence between the model and the statements that implement it, with a schematic representation of the business process.
 %\begin{figure}
 %\centering{
-%\includegraphics[width=0.85\textwidth]{figs/pews-expression}}
-%\caption{Pews program implementing the "To Publish Music" application}
-%\label{fig:pewsexpression}
+%\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
+%\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
+%\label{fig:ServiceProcess-ServiceComposition-Rules}
 %\end{figure}
-%Table \ref{fig:ServiceProcess-ServiceComposition}  shows the transformation rules for transforming a .
-As illustrated in Figure \ref{fig:ServiceProcess-ServiceComposition-Rules} the  principle of the transformation  of a  $\pi$-Service process model into a   $\pi$-ServiceComposition model is to group respectively  {\sf Contracts}  and {\sf Actions} into {\sf Policies} and {\sf Service activities}.
+{\color{red}
+A {\sf Package} in the $\pi$-use case model is transformed   into a {\sf Business Collaborator}.
 
-Each {\sf Assertion} of a {\sf Contract} in a $\pi$-service composition model is transformed into a {\sf Rule} in a $\pi$-service composition model.
-The set of {\sf Rules}  concerning the same NFP (e.g., safety, performance, and reliability) is grouped into a {\sf Policy} (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-a).
-Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute.
-If the {\sf Assertion} has a value type, the name and the attributes are transformed into a {\sf Variable} in the target model.
-The {\sf Assertion: Aproperty} attribute can have different transformations, according to  the following rules:
-\begin{itemize}
-\item a {\sf Post-Condition} is transformed into a {\sf  Post}. 
-\item a {\sf Precondition} is transformed into {\sf Pre}.
-\item a {\sf Timerestriction} is transformed into {\sf Time}.
-\end{itemize}
+Placido: Is this correct? Why it is here? We consider the $\pi$-use case in the previous section!
+}
 
-\begin{figure}
-\centering{
-\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
-\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
-\label{fig:ServiceProcess-ServiceComposition-Rules}
-\end{figure}
+{\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their homonym concepts of the $\pi$-Service Composition model.
 
-A {\sf Package}  in the $\pi$-use case model is transformed   into a {\sf Business Collaborator}. 
-{\sf Actions}  of a $\pi$-service process model are transformed into an {\sf Action} in a $\pi$-service composition model, and every {\sf Service activity} of a $\pi$-service process model is transformed into a {\sf Service activity} in a $\pi$-service composition model (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-b).
-An {\sf Exceptional behaviour} entity is transformed into an {\sf Action} in a $\pi$-service composition model (see Figure \ref{fig:ServiceProcess-ServiceComposition-Rules}-c ), and every {\sf Non-functional attribute} associated to an element ({\sf Contract} and {\sf Non-functional requirement}) in a $\pi$-service process model becomes a {\sf Non-functional attribute} associated to the corresponding element {\sf Policy} in a $\pi$-service composition model.
- Finally, {\sf Actions} are grouped by  a  {\sf Business collaborator}.
+An {\sf Exceptional behaviour} entity is transformed into an {\sf Action} in a $\pi$-Service Composition model and every {\sf Non-functional attribute} associated to an element ({\sf Contract} or {\sf Non-functional requirement}) in a $\pi$-service process model becomes a {\sf Non-functional attribute} associated to the corresponding element {\sf A-Policy} in a $\pi$-service composition model.\footnote{\color{red} Not found in the metamodel!
+--Martin and Umberto}
+Finally, {\sf Actions} are grouped by  a {\sf Business collaborator}.\footnote{\color{red} How? Where in the metamodel? --Martin and Umberto}
 
@@
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
-Considering the scenario example, 
-%the {\sf Action}s update music {\sc Contract} is transformed is a {\sc Policy} with its {\sc Rules}.
-All contract {\sc Assertions} are transformed in RULE and its attributes, e.g.
-the login and password verification.
-the "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
-%The {\sc Non- functional requirement} information will be used to the {\sc Policy} generation comes from the initial use case model. Also 
-The {\sf Business Collaborator} Facebook and Spotify information come from entity of type {\sc Package}  in the $\pi$-use case model. 
-%All {\sc Contracts} of the same {\sc Non- functional requirement} are composed in a {\sc Policy}.
+Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.
+
+The "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
+The information about Facebook and Spotify (both of them {\sf Business Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case model.\footnote{\color{red} Como assim?
+- Placido, explain this, please!
+Why have we bypassed one level here?}
 \end{example}
@@
 
-%Table \ref{fig:ServiceComposition-Pews-Rules} 
-This section describes the PIM to PSM transformations from a $\pi$-Service composition model to a $\pi$-PEWS program.
-We propose two groups of rules: those that transform services composition entities of a $\pi$-Service composition model into $\pi$-PEWS model; and those that transform rules grouped by policies into A-policy types.
+This section describes the PIM to PSM transformations from a $\pi$-Service composition model to a $\pi$-PEWS model. 
+We distinguish two groups of rules: \textit{(i)} those transforming service composition entities into workflows; and \textit{(ii)} those that transform  A-Policies into Contracts.
 
-A named action of the $\pi$-service composition model represented by  {\sf Action} and {\sf Action:name} is transformed to an  {\sf Operation} with a corresponding attribute name {\sf Operation:name}.
-A  named service activity represented by the elements {\sf ServiceActivity}  and  {\sf ServiceActivity:name} of the $\pi$-service composition model, are  transformed into a named operation of the $\pi$-{Pews} represented by the entities  {\sf CompositeOperation} and {\sf CompositeOperation:name}.
-When more than one action is called, according to the following  composition patterns expressed using the operators {\sc\em merge, decision, fork and join} in the $\pi$-service composition model the corresponding transformations, according to the PEWS operators presented above:
-\begin{itemize}
-\item   $op_1 . op_2$ if no {\sc\em ControlNode} is specified
-\item ($op_1 \parallel op_2) . op_3$ if control nodes of type {\sc\em fork, join} are combined
- \item ($op_1 + op_2) . op_3$ if control nodes of type {\sc\em decision, merge} are combined
-\end{itemize}
+Single actions (represented by {\sf Action} and {\sf Action:name}\footnote{\color{red} Where is the name defined?}) are transformed into individual service {\sf Operations}. 
+Complex actions (represented by {\sf ServiceActivity}  and  {\sf ServiceActivity:name}\footnote{\color{red} Where is the name defined?}) are transformed into (named) composite operations that defines a (named) workflow of the application.
+Composition patterns expressed using the operators {\sc\em merge, decision, fork and join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
 
@@
 
-The A-policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy} entities, named according to the names expressed in the source model.
+The A-Policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy}
+\footnote{\color{red} Please rename this for CONTRACT on the pi-pews metamodel!} 
+entities, named according to the names expressed in the source model. 
 The transformation of the rules expressed in a $\pi$-service composition is guided by the event types associated to these rules. 
-The variables associated to an A-policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into entities  {\sf Variable} with attributes {\sf Name} and {\sf Type} directly specified from the elements {\sf Variable:name} and {\sf Variable:type} of a $\pi$-service composition model.
+The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into entities  {\sf Variable} with attributes {\sf Name} and {\sf Type} directly specified from the elements {\sf Variable:name} and {\sf Variable:type} of a $\pi$-service composition model.
 
-%As shown in Table \ref{fig:ServiceComposition-Pews-Rules}, 
-For an event of type Pre the corresponding transformed rule is a {\sf Precondition}; for an event of type Post the corresponding transformed rule is a {\sf Post- condition}; finally, for an event of type TimeRestriction the corresponding transformed rule is a {\sf Time}.
-The condition expression of a rule in a $\pi$-service composition ({\sf Rule:condition}) is transformed into a  {\sf Condition:expression} where the attributes of the expression are transformed into  entities {\sf Attribute}.
+Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Preconditions}, {\sf Postcondition} and {\sf TimeRestrictions}.
 
-Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.
-
-\begin{figure}
-\centering{
-\includegraphics[width=0.96\textwidth]{figs/31-31}}
-\caption{$\pi$-PEWS Specific and Contract Representation}
-\label{fig:Specific-Contract-Representation}
-\end{figure}
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT6}
+Figure \ref{fig:piSOD-M} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.
+\end{example}
 
 \subsection{Implementation}
-We have provided tools for aiding the user into transform models for all the levels, except for the CIM into PIM level, which should be manually performed.
+We have provided tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
 
-This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions.
-For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application.
-Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
+Our tool is implemented as a series of Eclipse plug-ins: 
+\begin{itemize}
+\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
+Then, from these meta-models, we  developed plug-ins to support their graphical representation.
 
-%We used the Eclipse Modeling Framework (EMF) to implement the whole model transformation process \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}.
-From a model specification described in XMI, EMF provides tools and runtime support to produce a set of Java classes for the model, along with a set of adapter classes that enable viewing and command-based editing of the model, and a basic editor.
-%In order to automate the transformation we specified  transformation rules using the ATL model transformation language Finally, in order to generate code we  used Acceleo \footnote{http://www.acceleo.org/pages/home/en}.
+\item	 We used  ATL\footnote{http://eclipse.org/atl/.
+An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
+for implementing the  mappings between models.
 
-%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{General architecture}
-%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-Figure \ref{fig:policymanager} depicts a general architecture of the $\pi$-SOD-M Development Environment showing the set of plug-ins  developed in order to implement it.
-The environment implements the abstract architecture.
-Thus, it consists of plug-ins implementing the $\pi$-SCM and $\pi$-{\sc Pews} meta-models used for defining models specifying services' compositions and their associated policies; and ATL rules for transforming  PSM models (model to model transformation) and finally generating code (model to text transformation).
-\begin{figure}[t]
-	\begin{center}
-		\includegraphics[width=0.60\textwidth]{figs/architecture}
-	\end{center}
-		\caption{$\pi$-SOD-M Development Environment}
-   \label{fig:policymanager}
-\end{figure}
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF) \footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the meta-models  $\pi$- SCM and $\pi$-{\sc Pews}.
-Then, starting form these meta-models, we  developed the models' plug-ins needed to support the graphical representation of the $\pi$- SCM and $\pi$-{\sc Pews} models ($\pi$-ServiceCompostion Model and $\pi$-PEWS Model plug-ins).
-
-\item	 We used  ATL \footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for  developing the mapping plug-in implementing the  mappings between models ($\pi$-ServiceComposition2$\pi$-PEWS Plug-in).
-
-\item 	We  used Acceleo \footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in.
-We coded the pews.mt program  that implements the model to text transformation for generating executable code.
-It takes as input a $\pi$-PEWS model implementing a specific services' composition and it generates the code to be executed by the
-{\em A-policy} based services' composition execution environment.
-
-%\item Finally, we created a chain execution  to execute the model to text transformation.
+\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in for generating executable code. 
+It takes a $\pi$-PEWS model and generates the code to be executed by the {\em A-Policy} based service composition execution environment\footnote{\color{red} Placido: Policy or contract??.}. 
 \end{itemize}
-
-%%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-\subsection{A-policy manager}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
-
-As  shown in Figure \ref{fig:policymanager}, once an instance of a PEWS code is obtained starting form a particular $\pi$-services' composition model it can be executed over {\em A-policy} based services' composition execution environment  consisting of a composition engine and a {\em A-policy} manager.
-The  {\em A-policy} manager  consists of three main components Manager, for scheduling the execution of rules, C-Evaluator and A-Executor respectively for evaluating rules' conditions and executing their actions.
-The {\em A-policy} Manager interacts with a composition engine thanks to a  message communication layer (MOM).
-
-The composition engine manages the life cycle of the composition. Once a composition instance is activated, the engine schedules the composition activities according to the composition control flow.
-Each activity is seen as the process where the service method call is executed.
-The execution of an activity has four states: prepared, started, terminated, and failure.
-The execution of the control flow (sequence, and/or split and join) can also be prepared, started, terminated and raise a failure.
-
-At execution time, the evaluation of policies done by the {\em A-policy} manager must be synchronized with the execution of the services' composition (i.e., the execution of an activity or a control flow).
-Policies associated to a scope are activated when the execution of its scope starts.
-A {\em A-policy} will have to be executed only if one or several of its rules is triggered.
-If several rules are triggered the {\em A-policy} manager first builds an execution plan that specifies the order in which such rules will be executed according to the strategies defined in the following section.
-%Once rules have been executed, the {\em A-policy} finishes its execution and returns to a sleeping state.
-If rules belonging to several policies are triggered then policies are also ordered according to an execution plan.
-The execution of policies is out of the scope of this paper, the interested reader can refer to \cite{Espinosa-Oviedo2011a} for further details.
-%The order of policies has implications on the global order of the rules to be executed.

### 2014-04-07 JournalToSubmit2013/modelingFlyingPig.tex
Answers for Martin and Umberto.

@@
 $\star$ "Maximum number of users should be greater than 200".
-}
+} {\color{blue} [[I did not understand what must be done.]]}
 \caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut.png}
+\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
 {\color{red} \raggedright
 $\star$ Change FlyingFish by \FlyingPig.
-}
+} {\color{blue} [[done.]]}
 \caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}

### 2014-04-07 JournalToSubmit2013/transformationRules.tex
Answers for Martin and Umberto.

@@
 Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
-Every {\sf Extend} relationship identified in a $\pi$-Use Case model is transformed into a  {\sf Fork node}\footnote{Notice that Fork nodes are used both to represent alternative and parallel execution.
-{\color{red} \sc Placido, please check this!}}.
+Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
+transformed into a  {\sf Fork node}\footnote{Notice that Fork nodes are used
+both to represent alternative and parallel execution. {\color{red} \sc Placido,
+please check this!} {\color{blue} This is consequence of the use cases' extends
+and include relationships. To transform an use case into an action, it is
+necessary check the dependence relation. The include implies a parallel fork,
+and a extends implies in a alternative fork. I included the references for this definition.
+} }\cite{valeriaThesis,placidoPhDThesis2012}.
 If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
@@
 Placido: Is this correct? Why it is here? We consider the $\pi$-use case in the previous section!
-}
+}{\color{blue} Yes, it is correct. I defined it in the thesis because in the
+previous SOD-M a Business Collaborator comes from 'nothing'. So, as we needed to
+refine most concepts from use case, I decided separete the use cases in
+packages and then tranform it in a Collaborator. As a use case is transformed
+into an action, and all action is from a BC, I related both concepts. There is a
+layer gap ($\pi$-ServiceProcess), because in this layer we define only the
+actions.
+
+We are defining it here because the concept is necessary for the
+transformation. Packages is part of UML in the use case model, and I used it for define the colaborators.}
 
@@
 
-An {\sf Exceptional behaviour} entity is transformed into an {\sf Action} in a $\pi$-Service Composition model and every {\sf Non-functional attribute} associated to an element ({\sf Contract} or {\sf Non-functional requirement}) in a $\pi$-service process model becomes a {\sf Non-functional attribute} associated to the corresponding element {\sf A-Policy} in a $\pi$-service composition model.\footnote{\color{red} Not found in the metamodel!
---Martin and Umberto}
-Finally, {\sf Actions} are grouped by  a {\sf Business collaborator}.\footnote{\color{red} How? Where in the metamodel? --Martin and Umberto}
+{\color{blue} An {\sf Exceptional behaviour} entity is transformed into an {\sf
+Action} in a $\pi$-Serv\-ice\-Com\-po\-si\-tion model and every {\sf
+Non-functional attribute} associated to an {\sf Contract} (it comes from the $\pi$UseCase model) element 
+in a $\pi$-ServiceProcess model becomes a {\sf Non-functional Requirement}
+associated to the corresponding element {\sf A-Policy} in a
+$\pi$-ServiceComposition model}.\footnote{\color{red} Not found in the
+metamodel! --Martin and Umberto \color{blue}  ANSWER:  If you analyze the model $\pi$-UseCase, a constraint may be associated with one or more NFA (Non-functional attribute). The constraints
+associated with the same use case, are grouped into a contract. Each contract is
+related to a actions ($\pi$-ServiceProcess). Thus, the contracts (set of
+constraints) have NFAs associated (from piUseCase). The transformation between
+the first 2 levels (PIM $\pi$-UseCase to $\pi$-ServiceProcess) give us this.
+NFAs associated to an element (Contract) in a $\pi$-ServiceProcess model are
+refined (grouped) into a Non-functional requirement associated to the corresponding
+element A-Policy in the $\pi$-ServiceComposition model. I have to update the
+meta-models (thesis version) because they were outdated.} Finally, {\sf Actions}
+are grouped by  a {\sf Business collaborator}.\footnote{\color{red} How? Where
+in the metamodel? --Martin and Umberto \color{blue} ANSWER: Figure
+\ref{fig:e-scomposition-metamodel} presents the relationship between Action and BusinessCollaborator}
 
@@
 The "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
-The information about Facebook and Spotify (both of them {\sf Business Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case model.\footnote{\color{red} Como assim?
-- Placido, explain this, please!
-Why have we bypassed one level here?}
+The information about Facebook and Spotify (both of them {\sf Business
+Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
+model.\footnote{\color{red} Como assim? - Placido, explain this, please! Why
+have we bypassed one level here? \color{blue} ANSWER: It was explained before.}
 \end{example}
@@
 
-Single actions (represented by {\sf Action} and {\sf Action:name}\footnote{\color{red} Where is the name defined?}) are transformed into individual service {\sf Operations}. 
-Complex actions (represented by {\sf ServiceActivity}  and  {\sf ServiceActivity:name}\footnote{\color{red} Where is the name defined?}) are transformed into (named) composite operations that defines a (named) workflow of the application.
+Single actions (represented by {\sf Action} and {\sf
+Action:name}\footnote{\color{red} Where is the name defined? \color{blue} ANSWER: I updated the metamodel with the name atribute, however
+in general, I ommit the atributes for the metamodel presentation, but in the
+implementation the atributes are there.}) are transformed into individual service {\sf Operations}.
+Complex actions (represented by {\sf ServiceActivity}  and  {\sf
+ServiceActivity:name}\footnote{\color{red} Where is the name defined?
+\color{blue} ANSWER: I updated the metamodel with the name atribute, however
+in general, I ommit the atributes for the metamodel presentation, but in the
+implementation the atributes are there.}) are transformed into (named) composite
+operations that defines a (named) workflow of the application.
 Composition patterns expressed using the operators {\sc\em merge, decision, fork and join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
@@
 The A-Policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy}
-\footnote{\color{red} Please rename this for CONTRACT on the pi-pews metamodel!} 
-entities, named according to the names expressed in the source model. 
+\footnote{\color{red} Please rename this for CONTRACT on the pi-pews metamodel!
+\color{blue} ANSWER: The latest version of the $\pi$-PEWS grammar I defined
+policy.} entities, named according to the names expressed in the source model.
 The transformation of the rules expressed in a $\pi$-service composition is guided by the event types associated to these rules. 
@@
 \item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in for generating executable code. 
-It takes a $\pi$-PEWS model and generates the code to be executed by the {\em A-Policy} based service composition execution environment\footnote{\color{red} Placido: Policy or contract??.}. 
+It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
+A-Policy} based service composition execution environment\footnote{\color{red}
+Placido: Policy or contract??. \color{blue} ANSWER: The latest version of the
+$\pi$-PEWS grammar I defined \textbf{policy}. Particularly I prefer \textbf{policy},
+because for what we're proposing, one same policy can be applied to different
+services and groups a set of contracts.}.
 \end{itemize}

### 2014-04-07 JournalToSubmit2013/transformationRules.tex
small changes everywhere

--- 2014-04-04 JournalToSubmit2013/transformationRules.tex
@@
 Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
-Every {\sf Extend} relationship identified in a $\pi$-Use Case model is transformed into a  {\sf Fork node}\footnote{Notice that Fork nodes are used both to represent alternative and parallel execution.
-{\color{red} \sc Placido, please check this!}}.
+Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
+transformed into a  {\sf Fork node}~\cite{valeriaThesis,placidoPhDThesis2012}.
 If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
@@
 %\end{figure}
-{\color{red}
-A {\sf Package} in the $\pi$-use case model is transformed   into a {\sf Business Collaborator}.
 
-Placido: Is this correct? Why it is here? We consider the $\pi$-use case in the previous section!
-}
+Some elements of the $\pi$-service composition model are obtained from the $pi$-UseCase model: 
+A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
+\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Funtional requirements} of the $\pi$-ServiceComposition model. 
+These requirements will be associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).
 
@@
 
-An {\sf Exceptional behaviour} entity is transformed into an {\sf Action} in a $\pi$-Service Composition model and every {\sf Non-functional attribute} associated to an element ({\sf Contract} or {\sf Non-functional requirement}) in a $\pi$-service process model becomes a {\sf Non-functional attribute} associated to the corresponding element {\sf A-Policy} in a $\pi$-service composition model.\footnote{\color{red} Not found in the metamodel!
---Martin and Umberto}
-Finally, {\sf Actions} are grouped by  a {\sf Business collaborator}.\footnote{\color{red} How? Where in the metamodel? --Martin and Umberto}
-
+Finally, {\sf Actions} are grouped into  a {\sf Business collaborator}.
 %\begin{figure}
@@
 The "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
-The information about Facebook and Spotify (both of them {\sf Business Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case model.\footnote{\color{red} Como assim?
-- Placido, explain this, please!
-Why have we bypassed one level here?}
+The information about Facebook and Spotify (both of them {\sf Business
+Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
+model.
 \end{example}
@@
 
-Single actions (represented by {\sf Action} and {\sf Action:name}\footnote{\color{red} Where is the name defined?}) are transformed into individual service {\sf Operations}. 
-Complex actions (represented by {\sf ServiceActivity}  and  {\sf ServiceActivity:name}\footnote{\color{red} Where is the name defined?}) are transformed into (named) composite operations that defines a (named) workflow of the application.
+Single actions (represented by {\sf Action} and {\sf
+Action:name}) are transformed into individual service {\sf Operations}.
+Complex actions (represented by {\sf ServiceActivity}  and  {\sf
+ServiceActivity:name}) are transformed into (named) composite
+operations that defines a (named) workflow of the application.
 Composition patterns expressed using the operators {\sc\em merge, decision, fork and join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
@@
 
-The A-Policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy}
-\footnote{\color{red} Please rename this for CONTRACT on the pi-pews metamodel!} 
-entities, named according to the names expressed in the source model. 
+The A-Policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy} entities, named according to the names expressed in the source model.
 The transformation of the rules expressed in a $\pi$-service composition is guided by the event types associated to these rules. 
@@
 \item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in for generating executable code. 
-It takes a $\pi$-PEWS model and generates the code to be executed by the {\em A-Policy} based service composition execution environment\footnote{\color{red} Placido: Policy or contract??.}. 
+It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
+A-Policy} based service composition execution environment.
 \end{itemize}

### 2014-04-09 JournalToSubmit2013/modelingFlyingPig.tex
small changes everywhere
Some adjustments in figures and general review.

--- 2014-04-01 JournalToSubmit2013/modelingFlyingPig.tex
@@
 \includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
-
-{\color{red} \raggedright
-$\star$  "Assign user in charge".
-
-$\star$ "Maximum number of users should be greater than 200".
-}
 \caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut.png}
-{\color{red} \raggedright
-$\star$ Change FlyingFish by \FlyingPig.
-}
+\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
 \caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
@@
 The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{1,2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
+Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
 

### 2014-04-28 JournalToSubmit2013/conclusions.tex
Final version?

This paper presented \pisodm for designing and developing reliable service-based applications. 
\pisodm is an MDA methodology that extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
These requirements are taken into account from the initial stages of the software development process.
Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.

Our methodology includes two CIM-level models, three PIM-level models and one PSM-level model. 
We implemented the meta-models on the Eclipse platform and we validated the approach by using an industrially inspired use case.

Our case study demonstrates the applicability of \pisodm.
The case study was developed together with our industrial partner, GCP Global.
The Company is using \pisodm for the development of their product.
The case study presented here is a simplified version of their application.

### 2014-04-28 JournalToSubmit2013/intro.tex
Some adjustments in figures and general review.
Final version?

@@
 \begin{trivlist}
-\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification;
+\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
 \item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
@@
 Sections~\ref{sec:motivation}  and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules. 
-Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology. 
+% Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology. 
 Section~\ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned. 

### 2014-04-28 JournalToSubmit2013/modelingFlyingPig.tex
Final version?

@@
 
-In the following sections, we describe the results of applying $\Pi$-SODM to develop the \FlyingPig\ risk assessment system.
+In the following sections, we describe the results of applying \pisodm to develop the \FlyingPig\ risk assessment system.
 The models presented next were generated as a result of interacting with software developers at GCP Global.
@@
 %In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
-The $\pi$-SOD-M models of the PIM level  for the \FlyingPig\ scenario are presented next.
+The \pisodm models of the PIM level  for the \FlyingPig\ scenario are presented next.
 
@@
 
-\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
+\paragraph{\underline{$\pi$-ServiceComposition Model for \FlyingPig}}~
 
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%\subsection{Platform-Specific Model (PSM)}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+\paragraph{\underline{$\pi$-PEWS Model for \FlyingPig}}~
+
+The  PSM model of our case study is obtained by transforming the $\pi$-service composition model into a workflow.
+Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
+
 %\begin{figure}[t]
 %\centering
-\begin{lstlisting}[label=list:flyingPigPiPEWS,caption=pi-PEWS
+\begin{lstlisting}[basicstyle=\footnotesize,label=list:flyingPigPiPEWS,caption=pi-PEWS
 Specification: FlyingPig. ]!\label{fig:PiPEWSModel}! 
@@
 service receiveRequest(R, Id) = createCompanySpace(R, Id)
+service generateNewInterface(Id, NULL) = ...
+service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
+service notifyResponsable((Id, Q); NULL) = ...
+service receiveAnswers((Id, T); P) =
+	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL)
+	. ((createReport(Id, P) . return(P)) + (callForAmendments(Id,T) . return(NULL)))
 
-service generateNewInterface(Id, NULL) = ...
-
-service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
-
-service notifyResponsable((Id, Q); NULL) = ...
-
-service receiveAnswers((Id, T); P) =
-			analizeAnswer((Id, T), NULL) 
-			. storeData((Id, T), NULL)
-			. (
-			  (createReport(Id, P) . return(P))
-			  + (callForAmendments(Id,T) . return(NULL))
-			  )
-
-//Path
-
+//Workflow
 receiveRequest(R, Id)
 ||
-( generateNewInterface(Id, NULL)
-  . createQuestionnaire(Id, Q)
-  .  notifyResponsable((Id, Q); NULL
- )
+(generateNewInterface(Id, NULL).createQuestionnaire(Id, Q).notifyResponsable((Id, Q); NULL))
 ||
- ( receiveAnswers((Id, T); P)
- . [P != NULL] STOP
- )*
-
-//Policies
-
-\normalsize
+ (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
 \end{lstlisting}
-%\fbox{\LARGE $\pi$-PEWS Model.}
-%
-%{\color{red} \raggedright
-%$\star$ Placido: Please, place the model here.
-%}
 %\caption{$\pi$-PEWS model for \FlyingPig.\label{fig:PiPEWSModel}}
 %\end{figure}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-%\subsection{Platform-Specific Model (PSM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-The  PSM models of our methodology result by transforming the $\pi$-service composition model into PEWS code.
-Note that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL} adding the notion of \textit{contract}.
 
@@
 
-Given a set of services with their exported methods known in advance or provided by a  service directory, building services' based applications can be  a simple task that implies expressing an application logic as a services' composition.
+Given a set of services with their exported methods known in advance or provided by a  service directory, building service-based applications can be  a simple task that implies expressing an application logic as a service composition.
 The challenge being  ensuring the compliance between the specification and the resulting application.

### 2014-05-07 JournalToSubmit2013/modelingWithPISODM.tex
final version (V2)

--- 2014-03-03 JournalToSubmit2013/modelingWithPISODM.tex
@@
-%(Aquí lo dejaría con la estructura que teníamos para Caise, presentamos el framework y posteriormente comentamos modelos y transformaciones)
+This section presents \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
+\pisodm provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
+The refinement process is driven by transformation rules specified between the meta-models.
 
-This section presents $\pi$SOD-M, an MDD-based methodology for building service compositions with non-functional requirements. 
-$\pi$SOD-M extends SOD-M with concepts for modelling non-functional requirements, it proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models (Figure~\ref{fig:piSOD-M}).
+The next subsections present the meta-models from higher to lower levels of abstraction.
+The presentation is exemplified by a running example.
+The transformations between models are presented in section~\ref{sec:mmrules}.
+
 \begin{figure}[h]
@@
 \includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
-\caption{$\pi$SOD-M Overview.}
+\caption{\pisodm Overview.}
 \label{fig:piSOD-M}
 \end{figure}
-
-The $\pi$-SODM  meta-models are organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
-These meta-models are explained next.
 
@@
 
-This level focusses on the highest-level view of the system, including its business and requirement specifications.
+This level focuses on the highest-level view of the system, including its business and requirement specifications.
 At this stage of the development, the structure and system processing details are still unknown or undetermined.  
-$\pi$-SODM uses the \textit{e$^3$value}~\cite{e3value} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.
+\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.
 
-The e$^3$value model identifies the value/information exchange between components of the system. 
-It defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need.
+% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+\subsubsection{E$^3$value meta-model}
+% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+
+The e$^3$value model of an application identifies the value/information exchange between components of the system. 
+It is a business model that represents a business case graphically as a set of value exchanges\footnote{See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
+The model is well suited to enhance the understanding of the environment in which an application is being developed. 
+It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
 A dependency path has a direction and consists of a sequence of linked dependency nodes.
-A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node\footnote{See Legend on Figure~\ref{fig:CIM:tpme3v}}. 
+It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}). 
 Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
 
-The BPMN model is a graphical representation that establishes the business process of the application through a high-level workflow. The next example illustrates the use of these models.
-
-\begin{example}[To Publish Music]\label{ex:toPublicMusic}
-To illustrate our proposal, we build the models for a ``To Publish Music'' problem. 
-This problem involves three external actors ({\em Spotify, Twitter} and {\em Facebook}), as well as the user and the application itself.
-
-The e$^3$value and BPMN models for this applications are shown in Figure~\ref{fig:CIM:tpm}.
-Notice that the models at this level show\dots 
-{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
-%\hfill\openbox
-\end{example}
-
-\begin{figure}\center
-\includegraphics[width=0.88\textwidth]{figs/e3value.pdf}
+\begin{figure}[htpb]
+\center
+\includegraphics[width=0.75\textwidth]{figs/e3value.pdf}
 \hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
@@
 
-\begin{figure}\center
+\begin{example}[To Publish Music]\label{ex:toPublicMusic}
+Let us consider the scenario ``To Publish Music'', used as a running example in this section:
+An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
+In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
+The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
+The following (external) services will be used by the application:
+\begin{itemize}
+\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
+\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
+\item Facebook and Twitter services export methods for  updating the status of a given user:
+\begin{itemize} 
+%\item{\sf\small get-Status ( usedid ): String ; 
+\item {\sf\small update-Status ( userid, new-status ): String}; 
+\end{itemize}
+\end{itemize}
+
+The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}. 
+The e$^3$value value model shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
+The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
+Finally, the private application interacts with Facebook and Twitter for updating the user's status.
+We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
+\end{example}
+
+The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
+It is also necessary to understand the business process of the application and the conditions in which the different steps of this process are executed input/out data and the dependencies among these steps.
+Therefore, our methodology proposes the BPMN meta-model as a tool for modeling this aspect of the application.
+
+%Note that the models at this level show\dots 
+%{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
+%\hfill\openbox
+% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+\subsubsection{BPMN meta-model}% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow. 
+The next example illustrates the use of this meta-model.
+
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
+Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
+It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
+Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
+\end{example}
+%
+\begin{figure}[htpb]
+\center
 \includegraphics[width=0.49\textwidth]{figs/SC.pdf}
@@
 The CIM level models are the basis for the development of the application. 
-The information presented at this level are (manually) refined into PIM-level models.
+The information presented at this level is (manually) refined into PIM-level models of the methodology \pisodm described next.
+Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.
 
@@
 
-This level focusses on the system functionality, hiding the details of any particular platform.
+This level focuses on the system functionality, hiding the details of any particular platform.
 The specification defines those parts of the system that do not change from one platform to another. 
-Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess} and \textit{$\pi$-ServiceComposition}.
+Our methodology defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess} and \textit{$\pi$-ServiceComposition}.
 
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements. 
-Non-functional requirements are defined as \textit{constraints} over actions and data.
+ % .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+\subsubsection{\textit{$\pi$-UseCase} meta-model}% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+
+Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
+The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-functional requirements.
+The notion of \textit{policy} is used to describe NFRs. 
+(In Figure~\ref{fig:CIM:usecasemetamodel}, we have highlighted the concepts related to NFRs).
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing non-functional requirements with  the concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc non-functional requirement}, {\sc non-functional attribute} and {\sc Constraint}.
+An {\sc End Consumer} is represented by an {\sc Actor}.
+An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
+{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
+ The {\sc Business Collaborator} concept is represented  through {\sc Packages}. 
+A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled. 
+Each {\sc Business Collaborator} combines the features described in each  {\sc Package}. 
+Thus, the services functions can be specified as being grouped into packages.
+
+ \begin{figure}[htpb]
+\center
+\includegraphics[width=0.9\textwidth]{figs/UseCaseMetaModel.pdf}
+\caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
+\end{figure}
+
+ The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented as part of description of {\sc Use Cases} and {\sc Constraints}. 
+A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has its name, description, and which ones should be checked during the execution of the application. 
+Each {\sc Constraint} is represented as a stereotyped ({\sf constraint}) use case. 
+There are three kinds of constraint:
+\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value''; 
+\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and 
+\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
+These concepts are present in the following example.
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
-{\color{red} Talk about this model for the example.}
+Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example application.
+We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
+
+The ``To Publish Music'' Business Service expects  the Facebook or Twitter user status to be changed every time a user starts listening a new song.
+Therefore, it is necessary to perform a social network authentication with the users data.
+Each social network uses different services and different forms of authentication.
+The authentication constraint is required to update a music status.
+The restriction is stereotyped as a value constraint, because the users id and password are verified.
+Figure \ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases.
+The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as value and business constraint, respectively.
+For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
+Another restriction is that the minimum payment value is 2 euros.
+%{\color{red} Talk about this model for the example.}
 \end{example}
 
-The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
-gathers the constraints described in the \textit{$\pi$-UseCase}.
+\begin{figure}[htpb]
+\center
+\includegraphics[width=0.9\textwidth]{figs/UseCase.pdf}
+\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
+\end{figure}
+
+ % .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+
+The \textit{$\pi$-ServiceProcess} meta-model extends the UML activity diagram with the concept of \textit{contract} to represent constraints over data and actions. This concept is used to model 
+groups  of  constraints described in the \textit{$\pi$-UseCase}. 
+As shown in Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-ServiceProcess} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Service Activity}, {\sc Action} and {\sc Constraint}.
+The portion of the figure enclosed by a dotted line defines those concepts related to the representation of NFRs.
+
+A specific \textit{$\pi$-ServiceProcess} model shows the set of logically related activities to be performed in a service-based application. 
+So, the activities of this model represent a behavior that is part of the business logic of the application. 
+A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
+A service activity represents an operation that is part of the execution flow, and it is modeled as an {\sc Action}.
+An activity contract represents the {\sc Non-functional Requirement}  that is also part of the execution flow of a service, identified  as a stereotyped activity ({\sf assertion}). 
+The {\sc Assertion}s associated to an {\sc Action} compose a {\sc Contract}.
+Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre-conditions and post-conditions.
+
+\begin{figure}[htpb]
+\center
+\includegraphics[width=0.9\textwidth]{figs/ServiceProcessMetaModel.pdf}
+\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
+\end{figure}
+
+%According to the meta-model, a \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts.
+It models the set of system activities and describe contracts for each activity or the process as a whole.
+Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action.
+This model defines a workflow for  modeling business logic and the behaviour of the application being developed.
+The behaviour is represented by each action and the restrictions by the assertion description.
+All assertions are described by stereotyped actions.
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
-{\color{red} Talk about this model for the example.}
+Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}.
+The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service.
+The buy music pre-conditions consist in verifying:
+(i) if the User data are correct; 
+(ii) if the User is already logged in Spotify; 
+(iii) if bank account information are correct and; 
+(iv) if there are enough funds in the bank account to cover the payment. 
+A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization.
+There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified.
+To update services, depending of each service, there may be different restrictions.
+As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter.
+In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook.
+As post-condition, the application ensures that the Facebook service sends a notification of success. 
+To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modeled as post-condition). 
+As a pre-condition for ``twitter update'' it is necessary that (i) the music format is correct and (ii) the twitter login and password are correct for the update.
 \end{example}
 
-The \textit{$\pi$-ServiceComposition} meta-model provides the concept  of \textit{Policy}
-to integrate contracts with similar non-functional requirements.
+\begin{figure}[htpb]
+\center
+\includegraphics[width=0.9\textwidth]{figs/ServiceProcess.pdf}
+\caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
+\end{figure}
+
+% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+
+As shown in Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model 
+provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
+The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
+to group contracts with similar non-functional requirements.
 For instance, security and privacy restrictions may be grouped into a security policy.
+ The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
+%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
+\begin{itemizedTrivlist}
+\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action. 
+An instance of this meta-class is graphically represented as a partition in the activity diagram. 
+A collaborator can be either internal or external to the system. 
+When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
+
+\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class \textsc{Action}. 
+A class action represents some type of transformation or processing. 
+There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
+
+\begin{figure}[t]
+\centering 
+\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition} 
+\caption{$\pi$-Service Composition Meta-model.}
+\label{fig:e-scomposition-metamodel}
+\end{figure}
+
+\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
+
+\item In order to represent constraint types associated to services compositions, we introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
+An {\em A-policy} is defined by attributes and rules. 
+The conditions of each rule are evaluated and, if they are not verified, the actions of the rule will be performed.
+The {\sc Rule} meta-class represents the types of event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
+An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
+\end{itemizedTrivlist}
+Instances of this meta-model are represented as UML activity diagrams. 
+\begin{figure}[t]%[htpb]
+\centering
+\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
+%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
+\caption{$\pi$-ServiceComposition Model for the ``To publish
+music'' business service.}
+\label{fig:servicecompositionmodel}
+\end{figure}
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
+To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
  We model a business process that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
@@
 Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
-{\color{red} To be completed!.}
+%{\color{red} To be completed!.}
+In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
+% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
+The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
+Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
+Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that read and update the user data. 
+%A call to such services must be part of the authentication protocol required by these services.
+In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
+In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
+{\sf\small OAuthPolicy} implements the open authentication protocol.
+The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
+This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
+The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
+Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
+{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
+The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
+The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
+On the notification of an event of that type, a credential is obtained using the username and password. 
+\hfill\openbox
 \end{example}
+
+%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
+%These artifacts will be used to generate the actual programs that will implement the application:
+Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
+The $\pi$-PEWS  meta-model is described in the next section.
 
@@
 
-This level focusses on the functionality, in the context of a particular implementation platform.
+This level focuses on the functionality, in the context of a particular implementation platform.
 Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
@@
 
-The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions. 
+The \textit{$\pi$-PEWS} meta-model provides concepts for modeling service compositions. 
 Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
@@
 PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
-The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03} or any other language that lets the service designer combine the methods or subprograms that implement each operation of a service, in order to achieve the desired application logic.
-Figure~\ref{fig:metamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
+The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}. 
+Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
 \begin{itemizedTrivlist}
@@
 A {\sc Compound Operation} is defined using an {\sc Operator}.
-The language defines operators to denote guarded operations ($[C]S$); sequential ($\ .
-\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetition.
+The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) repetition.
 
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Metamodel}
-\label{fig:metamodel}
+\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
+\caption{$\pi$-{\sc Pews} Meta-model.}
+\label{fig:PPEWSmetamodel}
 \end{figure}
 
-Figure~\ref{fig:metamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
+Figure~\ref{fig:PPEWSmetamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
 Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.  
@@
 
-Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
-The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.
-
-===================================
-
-The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
-provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
-The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
-Instances of this meta-model are represented as UML activity diagrams. 
-In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
- white boxes represent classes modelling functional ones.
-
-%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
-\begin{itemizedTrivlist}
-\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
-An instance of this meta-class is graphically represented as a partition in the activity diagram. 
-A collaborator can be either internal or external to the system. 
-When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
-
-\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
-A class action represents some type of transformation or processing. 
-There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-\begin{figure}[t]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
-\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
-\label{fig:e-scomposition-metamodel}
-\end{figure}
-
-\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
-
-\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
-An {\em A-policy} is defined by attributes and rules. 
-Intuitively, the conditions of each rule will be checked.
-In case of no compliance, the actions defined by the rule will be performed.
-The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
-An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
-\end{itemizedTrivlist}
-
-\begin{figure}[t]%[htpb]
-\centering
-\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
-
-{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
-
-\caption{$\pi$-SCM for the ``To publish music'' business service.}
-\label{fig:servicecompositionmodel}
-\end{figure}
-
-\begin{example}[To Publish Music]\label{ex:toPublicMusic}
-To illustrate the use of the $\pi$-Service Composition Meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
-In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
-% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
-The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
-Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
-%A call to such services must be part of the authentication protocol required by these services.
-In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
-
-{\sf\small OAuthPolicy} will implement the open authentication protocol.
-The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
-The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
-Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
-
-{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
-The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
-On the notification of an event of that type, a credential is obtained using the username and password. 
-\hfill\openbox
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic5}
+Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our scenario example.
+Notice that the code contains namespaces and definitions (obtained by using information about the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
 \end{example}
 
-We propose the use of rules and policies to model and associate non-functional properties to service compositions.
-These artifacts will be used to generate the actual programs that will implement the application:
-Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
-The $\pi$-PEWS  meta-model is described in the next section.
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
+\begin{figure} 
+\centering
+\includegraphics[width=0.8\textwidth]{figs/pi-pewsSpecification-toPublishMusic}
+\caption{$\pi$-PEWS Specific and Policy Representation.}
+\label{fig:Specific-Contract-Representation}
+\end{figure}

### 2014-05-07 JournalToSubmit2013/transformationRules.tex
Final version?
final version (V2)

@@
 \pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models.
-These rules are used to build a semi-automatic tool.
+These rules are used to build a semi-automatic tool (the \pisodm environment plugin for Eclipse).
 The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.
@@
 {\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} attribute set to false are transformed into {\sf Assertion}s;
-{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behaviour}s.
+{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
 
@@
 %(as shown in figure \ref{fig:transformation-example-include}). 
-The same rule is applied for the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
+The same rule is applied to the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
  \end{example}
@@
 
-Some elements of the $\pi$-service composition model are obtained from the $pi$-UseCase model: 
+Some elements of the $\pi$-service composition model are obtained from the $\pi$-UseCase model: 
 A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
-\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Funtional requirements} of the $\pi$-ServiceComposition model. 
-These requirements will be associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).
+\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Functional requirements} of the $\pi$-ServiceComposition model. 
+These requirements are associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).
 
@@
 Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
-
-The "securityLoginPolicy" consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
+The ``securityLoginPolicy'' consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
 The information about Facebook and Spotify (both of them {\sf Business
@@
 
-Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Preconditions}, {\sf Postcondition} and {\sf TimeRestrictions}.
+Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Pre-conditions}, {\sf Post-condition} and {\sf TimeRestrictions}.
 
@@
 \subsection{Implementation}
-We have provided tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
+We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
 

### 2014-09-10 SystematicMapping/background.tex
Systematic Mapping: outline

\subsection{Adding NFP to service compositions}

\subsection{Problem space and solution space}

\subsection{Models, methodologies and environments}

\subsection{Related work}

### 2014-09-10 FlyingPig-journal/conclusions.tex
New space for reviewing the methodology paper and updates for the SM paper

--- 2014-04-28 JournalToSubmit2013/conclusions.tex
@@
-This paper presented \pisodm for designing and developing reliable service-based applications. 
-\pisodm is an MDA methodology that extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
-These requirements are taken into account from the initial stages of the software development process.
+This paper presented \pisodm, an MDA methodology for designing and developing reliable service-based applications. 
+\pisodm extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
+These requirements are taken into account from the early stages of the software development process.
 Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.

### 2014-09-10 FlyingPig-journal/intro.tex
New space for reviewing the methodology paper and updates for the SM paper

--- 2014-04-28 JournalToSubmit2013/intro.tex
@@
 Yet,  they are partially or rarely methodologically derived from the specification, being usually added once the code has been implemented. 
-In consequence, the development process will not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
+In consequence, the development process does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
 

### 2014-09-10 FlyingPig-journal/modelingCrimes.tex
New space for reviewing the methodology paper and updates for the SM paper

--- 2013-10-07 JournalToSubmit2013/2-modeling.tex
@@
 %\end{figure}
+
+In order to introduce the context of our work consider the Tracking crime application where civil population and by police share information about criminality in given zones of an imaginary city. 
+Users signal crimes using twitter and police officers notify crimes  they have to deal with. 
+Some of this information, if it is not confidential, can be shared to the community of users using this application. 
+Users can track crimes in given zones and see information located in a map according to their privileges.  
+For example civilians can ask Locate the crimes done the last month 100 km from my current position that happened between 8:00 and 14:00.
+While police officers can ask Locate the regions in my sector where murders happened in the last month
+This information can come from the police database and from twitter posts. 
+The zones of the city are thereby according to their degree of criminality.
+
+ In order to provide these functions the application benefits from existing services that provide information,  storage and data visualization functions.
+Thus, the application is a service based application that implements a business process.
+ The business process is specified in terms of ordered tasks and define the application logic of a Service oriented Application.
+Tasks can be performed by a person or an entity.
+In our context, tasks are implemented by Services.
+Service is an application implemented by a provider that exports an API through a network (e.g., Internet)
+API is defined using an interface definition language (IDL, e.g., WSDL for Web services).
+ In our example the business procces can start with one of two tasks: Notify a crime, or track a crime.
+A notified crime can be then stored in a database.
+Tracked crimes are visualized in a map and then the used can ask for detailed information.
+They are implemented by four services: twitter and an adhoc police service for notifying crimes, Amazon as persistence service and google maps for visualizing and locating crimes in a map.
+Note that a task can be implemented by a service composition.
+This is the case of the task Notify crime in our example that enables to notify crimes through twitter or through the police notification service.
+
+Business processes have also associated rules and constraints that define their non functional requirements.  
+NFR represents the “semantics” and the conditions in which the tasks must be done. 
+In our example we have some constraints. 
+\begin{itemize}
+\item Twtter requires to be accessed through an authentication protocol, and the police crime notification service has a control  access protocol and allows only 3 tries to access it. 
+\item Only users with enough privileges can store crimes’ notifications. 
+\item Users can only track crimes notified by providers t that hey have authorization to contact; For example civil population cannot track all the crimes notified by the police. 
+\item If the google map is unavailable the results of a track request are delivered on text. 
+\item If a user tracks crimes that are provided by a service for which she does not have authorization then show an empty map. 
+\item Detailed information about a crime depends on the user privileges, on the police assignment regions. 
+\item The tracking crime app keeps track of the accesses to detailed information about crimes. 
+\end{itemize}
+
+Through the example we underlined that every application implements functional aspects that describe its application logic.
+Recall that an application logic refers to routines that perform the activities to reach the application objective.
+Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
+These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
+
+The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
+API does not necessarily export information about methods dependency (e.g., in the REST protocol);
+they do not share their state (stateless).
 

### 2014-09-10 FlyingPig-journal/modelingFlyingPig.tex
New space for reviewing the methodology paper and updates for the SM paper

--- 2014-04-28 JournalToSubmit2013/modelingFlyingPig.tex
@@
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-%\subsection{Platform-Specific Models}
-%
-%
-%This level focusses on the functionality, in the context of a particular implementation infrastructure.
-%Models at this level combine the platform-independent view with specific aspects of the application to be deployed. 
-%At the PSM level we have lower-level models that can be automatically translated into any service composition language. 
-%We have defined one meta-model at this level, called \textit{$\pi$-PEWS}.
-%
-%The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions, including non-functional features.
-%\textit{$\pi$-PEWS} models have two main parts:
-%\begin{itemizedTrivlist}
-%\item \textit{Service compositions (workflows)} formed by individual service operations and composite operations defined by the use of workflow operators, such as guarded operations ($[C]S$); sequential ($\ .
-\ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetitions.
-%
-%\item \textit{ECA Rules}, with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}. 
-%These rules are used to specify non-functional behaviour.
-%\end{itemizedTrivlist}
-%
-%The workflow constructors of \textit{$\pi$-PEWS} are similar to those of  most service orchestration languages, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
-%
-%
-%%%
-%%\begin{figure}[t]
-%%\centering
-%%\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
-%%\caption{$\pi$-{\sc Pews} Metamodel}
-%%\label{fig:metamodel}
-%%\end{figure}
-%%%
-%
-%\bigskip
-%The following example presents the \textit{$\pi$-PEWS} model for the ``To Publish Music'' application.
-%\begin{example}[To Publish Music (end)]\label{Colocar-o-label-correto-aqui}
-%The next code presents the \textit{$\pi$-PEWS}  model for our running example.\footnote{\color{red} Placido: Please change the pdf file as needed!}
-%
-%\begin{center}
-%\includegraphics[scale=0.7]{figs/pewsEspecificationContract.pdf}
-%\end{center}
-%
-%The initial part of the model consists of a workflow expression.
-%In the given code, the execution of \texttt{searchSong} is followed by \texttt{selectSong}.
-%At this point, the user may either buy, download and listen a song or just listen to a song (conditional workflow).
-%The last part of the workflow publishes the title of the song either on tweeter or facebook.
-%
-%ECA rules are specified as contracts~\cite{Meyer92}.
-%In our case, the \texttt{buySongContract} specifies conditions to be verified upon the execution of \texttt{buySong}.
-%This contract specifies pre and post-conditions, as well as time constraints.\footnote{\color{red} Placido: Please verify the pre-condition assertions!}
-%The precondition (EXPLAIN!)\dots
-%
-%In case of failure in verifying the precondition, the service \texttt{buySong} is retried.
-%
-%The postcondition ensures that the payment succeeds before .
-%
-%The time constraint indicates that the \texttt{sendConfirmation} service is invoked immediately after \texttt{buySong}.
-%\end{example}
-%
-%
-%
-%
-%Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
-%The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.
-%
-%
-%
-%
-%
-%
-%
-%The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
-%provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
-%The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
-%Instances of this meta-model are represented as UML activity diagrams. 
-%In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
-% white boxes represent classes modelling functional ones. 
-%
-%%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
-%\begin{itemizedTrivlist}
-%\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
-%An instance of this meta-class is graphically represented as a partition in the activity diagram. 
-%A collaborator can be either internal or external to the system. 
-%When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
-%
-%\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
-%A class action represents some type of transformation or processing. 
-%There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
-%\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
-%\label{fig:e-scomposition-metamodel}
-%\end{figure}
-%
-%\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
-%
-%\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-%We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
-%An {\em A-policy} is defined by attributes and rules. 
-%Intuitively, the conditions of each rule will be checked.
-%In case of no compliance, the actions defined by the rule will be performed.
-%The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
-%An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
-%\end{itemizedTrivlist}
-%
-%\begin{figure}[t]%[htpb]
-%\centering
-%\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
-%
-%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
-%
-%\caption{$\pi$-SCM for the ``To publish music'' business service.}
-%\label{fig:servicecompositionmodel}
-%\end{figure}
-%
-%\begin{example}[To Publish Music]\label{ex:toPublicMusic}
-%To illustrate the use of the $\pi$-Service Composition Meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
-%In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
-%% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
-%The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
-%Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-%Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
-%%A call to such services must be part of the authentication protocol required by these services.
-%In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-%In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
-%
-%{\sf\small OAuthPolicy} will implement the open authentication protocol.
-%The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
-%This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
-%The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
-%Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
-%
-%{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
-%The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
-%The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
-%On the notification of an event of that type, a credential is obtained using the username and password. 
-%\hfill\openbox
-%\end{example}
-%
-%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
-%These artifacts will be used to generate the actual programs that will implement the application:
-%Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
-%The $\pi$-PEWS  meta-model is described in the next section.
-
 We validated our methodology by developing a use-case concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
@@
 and 
-(iv) the risk assessment \textsl{Fee}.
+(iv) the risk assessment \textsl{fee}.
 
-The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emited by a particular company. 
+The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company. 
 Once this need has been declared, the value exchanges between ORCA and Broker are triggered. 
-The client company will provide ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
-ORCA will provide amendments (recommendations to change practices) and an evaluation report.
+The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
+Then, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.
 
@@
 The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
-Questionnaires are considered tasks that users will have to perform. 
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecure access to the premises of the company.
+Questionnaires are considered tasks to be performed by users. 
+Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
 They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
 
-Once tasks have been completed, they are stored and analyzed to generate a list of noncompliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
-During the process of analyzing a questionnaire, the answers of some questions can trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
+Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
+During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
 
@@
 \begin{numtrivlist}
-\item An acknowledgement is due less than 30 seconds after registering a task or demand for assessment. 
+\item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment. 
 \item The system should be able to deal with, at least, 200 users. 
 \item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests. 
-\item The proviledges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
-\item The privileges of Users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
+\item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
+\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
 \item All questionnaires need to be fully answered, in order to consider that a task is completed.
@@
 %Use cases 
-that describe the main functionalities of the system.
+that describe the main functionality of the system.
 The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.
 
-In our model, each $\pi$-use case may be associated to one or more (non-functional) constraints (depicted as coloured ovals in Figure~\ref{fig:piUseCaseModel}). 
+In our model, each $\pi$-use case may be associated to one or more (non-functional) constraints (depicted as colored ovals in Figure~\ref{fig:piUseCaseModel}). 
 The model defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}. 
@@
 Use cases are transformed into \textit{actions} and are represented by white boxes.
-The restrictions associated to each $\pi$-use case are transformed into \textit{assertions} (represented by coloured boxes) and may be decorated with pre- and post-conditions. 
+The restrictions associated to each $\pi$-use case are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions. 
 We can see that this model refines the concepts defined in the $\pi$-UseCase model.
@@
 %The previous model 
-is now  completed with the services that will provide some functions of the application  \FlyingPig\.
+is now  completed with the services that provides functions of the application  \FlyingPig\.
 The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.

### 2014-09-10 FlyingPig-journal/relworks.tex
New space for reviewing the methodology paper and updates for the SM paper

--- 2014-03-25 JournalToSubmit2013/relworks.tex
@@
 These requirements are usually specified by conditions, called Non-functional Properties.
-Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioural requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
-Most research on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
-In the case of service-based applications, non-functional requirements concern the application itself as well as its component services.
+Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
+Most research on NFRs focusses on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
+In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.
 
@@
 The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services.
-A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties as \textit{availability}, \textit{reliability} and \textit{access control}.
+A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties such as \textit{availability}, \textit{reliability} and \textit{access control}.
 
-Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolset for specifying and implementing web service compositions with support to several NFRs. 
+Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs. 
 The approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
@@
 In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs, 
-whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modelling is used for
+whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
 system specification, including NFRs. 
-Most authors focus their work on modelling service compositions, although a significant number of approaches is focused on the definition of NFR models.
-
 In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be 
@@
 \bigskip
-Most works agree on distinguishing three points of view, namely the point of view of the organization (or \textit{Business} view), of the individual service providers (or \textit{Service} view) and of the composition designer (or \textit{System} view).
+Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
 The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
-Service and System views are concerned with the implementation of the software solution: The Service level is concerned with the building blocks of the application.
+Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
 It may use web services provided by third party sources.
-The System level is concerned with the coordination of services, to implement the business logic.
+The System level deals with the coordination of services, to implement the business logic.
+Our proposal, shown in the next section, follows these ideas.

### 2014-09-10 SystematicMapping/background.tex
New space for reviewing the methodology paper and updates for the SM paper
no message

@@
+This section introduces the vocabulary and concepts related to nonfunctional properties of service based applications.
+
 \subsection{Adding NFP to service compositions}
+In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
+combined to build an application business logic.
+The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
+An important challenge of service-oriented development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
 
-\subsection{Problem space and solution space}
+Functional properties are not the only  aspect in the software development process.
+Non-functional properties, such as data privacy, exception handling, atomicity  and, data persistence, need to be addressed  to fit in the application.
+Adding non-functional properties and respecting services constraints while composing services is a complex task that implies programming  protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
+
+Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements. 
+Ideally, non-functional requirements should be considered in every phase of the software development.
+Yet,  they are partially or rarely methodologically derived from the specification, being usually added once the code has been implemented. 
+In consequence, the development process does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
+
+The literature stresses the need for methodologies and techniques for service oriented analysis and design 
+%since they are the cornerstone in the development of meaningful service based applications
+~\cite{Papazoglou2007}. 
+Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software~\cite{watson}.
+Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
+These models are successively refined from abstract specifications into actual computer programs.
+
+%\subsection{Problem space and solution space}
 
@@
 
+Non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not always fully considered from the early phases of the (service) software process.
+Most methods integrate them only after the application has been implemented.
+This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.
+
+The adoption of non-functional specifications from the early states of development can help the developer to produce applications that  can deal with the application context.
+
 \subsection{Related work}

### 2014-11-12 JournalToSubmit2014/conclusions.tex
new journal version

--- 2014-09-10 FlyingPig-journal/conclusions.tex
@@
-This paper presented \pisodm, an MDA methodology for designing and developing reliable service-based applications. 
+We presented \pisodm, an MDA methodology for designing and developing reliable service-based applications. 
 \pisodm extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
@@
 Our case study demonstrates the applicability of \pisodm.
-The case study was developed together with our industrial partner, GCP Global.
+This case study was developed together with our industrial partner, GCP Global.
 The Company is using \pisodm for the development of their product.

### 2014-11-12 JournalToSubmit2014/figs/piPewsModel.tex
new journal version

--- 2014-09-10 FlyingPig-journal/modelingFlyingPig.tex
@@
-We validated our methodology by developing a use-case concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
-Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security for accessing reserved areas (such as computing server installations).
-The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
-The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
-By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
+\begin{figure}[ht!]
+\tiny
+\centering
 
-Our goal is to model a service based application (called \FlyingPig), for providing risk assessment as a service.
-In order to provide this functionality, \FlyingPig\ would benefit from ORCA's legacy services providing storage, assessment and data visualization functions.
+\begin{lstlisting}[label=list:flyingPigPiPEWS,caption=pi-PEWS
+Specification: FlyingPig. ]
 
-In the following sections, we describe the results of applying \pisodm to develop the \FlyingPig\ risk assessment system.
-The models presented next were generated as a result of interacting with software developers at GCP Global.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.7\textwidth]{figs/3ValueModel.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
-\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
-\end{figure}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Computation-Independent Models (CIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
-It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
-%\footnote{\color{red} This sentence need to move to section 3.1!!}
-We identify two business actors: \textsl{ORCA} and \textsl{Broker}. 
-Brokers emit requests for risk assessment for one or more companies. 
-ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}. 
-The values exchanged between ORCA and the brokers are: 
-(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
-(ii) \textsl{Amendments} which are ORCA's recommendations, based on the data provided by the answers to questionnaires; 
-(iii) \textsl{Evaluation Reports} for the client companies; 
-and 
-(iv) the risk assessment \textsl{fee}.
-
-The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company. 
-Once this need has been declared, the value exchanges between ORCA and Broker are triggered. 
-The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
-Then, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
-\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
-\end{figure}
-
-Figure~\ref{fig:BPMNmodel} shows the BPMN model
-%\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.} 
-for the \FlyingPig\ scenario. The BPMN model for \FlyingPig\ is partitioned for better understanding the process in which the value exchanges occur.
-The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}. 
-Brokers have two lanes, the client \textsl{Company} and a \textsl{User}. 
-The user is a contact member of the company, who  coordinates the assessment process. 
-This process  involves other members of the company as well.
-
-The risk assessment process starts after a request from a company.
-This corresponds to the value model, in which the start stimulus triggers the whole process.
-The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
-Questionnaires are considered tasks to be performed by users. 
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
-
-Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
-During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
-
-Business processes also have associated rules and constraints to define their non-functional requirements (NFR):
-\begin{numtrivlist}
-\item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment. 
-\item The system should be able to deal with, at least, 200 users. 
-\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests. 
-\item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
-\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
-\item All questionnaires need to be fully answered, in order to consider that a task is completed.
-\item There is a time limit (in days) for each amendment required by the system.
-\end{numtrivlist}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform-Independent Models (PIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-%In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
-The \pisodm models of the PIM level  for the \FlyingPig\ scenario are presented next.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
-\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
-\end{figure}
-
-\paragraph{\underline{$\pi$-UseCase Model for \FlyingPig}}~
-
-The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application. 
-In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
-In the context of \FlyingPig, Company is the actor requesting a risk evaluation.
-A Broker is  responsible for coordinating  the evaluation process, assigning users to be in charge of tasks as well as delegating tasks. 
-A User, in this model is an actor who answers questionnaires according to the current situation  of the Company.
-The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
-
-Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Figure~\ref{fig:piUseCaseModel}). 
-%Use cases 
-that describe the main functionality of the system.
-The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.
-
-In our model, each $\pi$-use case may be associated to one or more (non-functional) constraints (depicted as colored ovals in Figure~\ref{fig:piUseCaseModel}). 
-The model defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}. 
-Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
-
-%In the case of \FlyingPig, as expressed above the model constraints:
-%{\color{red} Anything else here? --M}
-
-\begin{figure}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneralCut.png}
-\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
-\end{figure}
-
-\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
-
-The $\pi$-ServiceProcess model (Figure~\ref{fig:PiServiceProcessModel}) presents the workflow for \FlyingPig.
-The actions in this model were obtained by applying the $\pi$-use case transformation rules.
-% described in Section~\ref{sec:pewsmetamodel}.
-
-The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
-Use cases are transformed into \textit{actions} and are represented by white boxes.
-The restrictions associated to each $\pi$-use case are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions. 
-We can see that this model refines the concepts defined in the $\pi$-UseCase model.
-The assertions specify those non-functional requirements, as they are seen by the actors. 
-The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
-\caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
-\end{figure}
-
-\paragraph{\underline{$\pi$-ServiceComposition Model for \FlyingPig}}~
-
-The model presented in Figure~\ref{fig:PiServiceCompositionModel} 
-%offers a system-centric view of the application.
-%The previous model 
-is now  completed with the services that provides functions of the application  \FlyingPig\.
-The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
-These policies express the pre- and post-conditions of the previous model.
-They are associated to the actions of the system.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-%\subsection{Platform-Specific Model (PSM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\paragraph{\underline{$\pi$-PEWS Model for \FlyingPig}}~
-
-The  PSM model of our case study is obtained by transforming the $\pi$-service composition model into a workflow.
-Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
-
-%\begin{figure}[t]
-%\centering
-\begin{lstlisting}[basicstyle=\footnotesize,label=list:flyingPigPiPEWS,caption=pi-PEWS
-Specification: FlyingPig. ]!\label{fig:PiPEWSModel}! 
 //Namespaces specify service URI
@@
 service receiveRequest(R, Id) = createCompanySpace(R, Id)
+
 service generateNewInterface(Id, NULL) = ...
+
 service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
+
 service notifyResponsable((Id, Q); NULL) = ...
+
 service receiveAnswers((Id, T); P) =
-	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL)
-	. ((createReport(Id, P) . return(P)) + (callForAmendments(Id,T) . return(NULL)))
+								analizeAnswer((Id, T), NULL) 
+								. storeData((Id, T), NULL)
+								. (
+								  (createReport(Id, P) . return(P))
+								  + (callForAmendments(Id,T) . return(NULL))
+								  )
 
-//Workflow
+//Path
+
 receiveRequest(R, Id)
 ||
-(generateNewInterface(Id, NULL).createQuestionnaire(Id, Q).notifyResponsable((Id, Q); NULL))
+( generateNewInterface(Id, NULL)
+  . createQuestionnaire(Id, Q)
+  .  notifyResponsable((Id, Q); NULL
+ )
 ||
- (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
+ ( receiveAnswers((Id, T); P)
+ . [P != NULL] STOP
+ )*
+
+//Policies
+
 \end{lstlisting}
-%\caption{$\pi$-PEWS model for \FlyingPig.\label{fig:PiPEWSModel}}
-%\end{figure}
-
-%Contracts may be implemented as transversal concerns~\cite{aspects}.
-%Their conditions may be verified at specific point of the composition and their actions are executed if needed.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Lessons Learned}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-Through the example we underlined that every application implements functional aspects that describe its application logic.
-Recall that an application logic refers to routines that perform the activities to reach the application objective.
-Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
-These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
-
-The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
-API does not necessarily export information about methods dependency (e.g., in the REST protocol);
-they do not share their state (stateless).
-
-Given a set of services with their exported methods known in advance or provided by a  service directory, building service-based applications can be  a simple task that implies expressing an application logic as a service composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way.
-Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
+\label{fig:crimeInformationPEWS}  
+\end{figure}

### 2014-11-12 JournalToSubmit2014/intro.tex
new journal version

--- 2014-09-10 FlyingPig-journal/intro.tex
@@
 
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
+In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
 combined to build an application business logic.
 The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
-An important challenge of service-oriented development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
+An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
 
@@
 Ideally, non-functional requirements should be considered in every phase of the software development.
-Yet,  they are partially or rarely methodologically derived from the specification, being usually added once the code has been implemented. 
-In consequence, the development process does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.
+Yet,  they are partially or rarely methodologically derived from the specification, being usually added once the code has been im\-ple\-men\-ted. 
+In consequence, the development process does not fully preserve the compliance and re\-use expectations provided by the service oriented computing methods.
 
@@
 
-The literature stresses the need for methodologies and techniques for service oriented analysis and design 
+The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design 
 %since they are the cornerstone in the development of meaningful service based applications
 ~\cite{Papazoglou2007}. 
-Existing approaches argue that the convergence of model-driven software development, service orientation,   and  business processes improvement are key for developing accurate  software~\cite{watson}.
+Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing accurate  software~\cite{watson}.
 Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
@@
 
-Our work proposes  the $\pi$-\textit{Service Oriented Development Method} (\pisodm)
-to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements from the early stages of software development.
-\pisodm is aligned with the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.
+Our work proposes, $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements early in software development.
+Our proposal is aligned with the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.
 The technique proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
-Models are organized according to the guidelines for structuring the specifications of software systems of the Model Driven Architecture (MDA)~\cite{miller} approach. 
+Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller} approach. 
 The goals of  \pisodm are to:
@@
 This paper is organized as follows.
-Section~\ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
+Sec\-tion \ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
 Sections~\ref{sec:motivation}  and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules. 

### 2014-11-12 JournalToSubmit2014/main.tex
new journal version

\documentclass[twocolumn]{svjour3}

\usepackage{geometry}
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{amssymb} 
\usepackage{epstopdf}

\usepackage{amssymb}

%\usepackage{natbib}

%%% OUR MACROS %%%
\newcommand{\COMMENT}[1]{ }

\usepackage[usenames,dvipsnames]{xcolor}

\usepackage{amsmath}
%\usepackage[thmmarks,amsmath]{ntheorem}

\newcommand{\openbox}{\leavevmode
  \hbox to.77778em{%
  \hfil\vrule
  \vbox to.675em{\hrule width.6em\vfil\hrule}%
  \vrule\hfil}}

%\theoremstyle{plain}
%\theoremheaderfont{\normalfont\bfseries}
%\theorembodyfont{\normalfont}
%\theoremseparator{}
%\theoremindent0cm
%\theoremnumbering{arabic}
%\newtheorem{algo}{Algorithm}
%
%%\theoremstyle{plain}
%%%\theoremheaderfont{\normalfont\itshape}
%%\theoremheaderfont{\normalfont\bfseries}
%%\theorembodyfont{\normalfont}
%%\theoremseparator{}
%%\theoremindent0cm
%%\theoremnumbering{arabic}
%%\theoremsymbol{\ensuremath{\openbox}} 
%%\newtheorem{example}{Example}
%
%
%\theoremstyle{plain}
%\theoremheaderfont{\normalfont\bfseries}
%\theorembodyfont{\normalfont}
%\theoremseparator{.}
%\theoremindent0cm
%\theoremnumbering{arabic}
%\theoremsymbol{\ensuremath{\Box}} 
%\newtheorem{defi}{Definition}
%
%\theoremstyle{plain} 
%\theoremsymbol{\ensuremath{\Box}} 
%\theoremseparator{.} 
%\newtheorem{prop}{Property}

\def\FlyingPig{\textsl{FlyingPig}}

\newcounter{numberInTrivlist}

\newenvironment{numtrivlist}{\begin{list}{\rm \arabic{numberInTrivlist})} 
                                         {\usecounter{numberInTrivlist}
                                          \setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\newenvironment{itemizedTrivlist}{\begin{list}{\rm ~\hspace{2mm} $\bullet$\ } 
                                         {\setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\usepackage{listings}

\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
        true,false,rise,exception,extend,resume,to,return,function}}

\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}

\begin{document}

\title{\pisodm: A New Methodology for Building Service-Oriented Applications}\thanks{\textsl{Contact author:}\\
Martin A. Musicante\\
UFRN - Cx Postal 1647, \\
Campus Universitário Lagoa Nova \\
59078-970, Natal, RN, Brazil\\[2mm]
Telephone: +55 84 3215 3814 ext 1.208\\
Mobile phone: +55 84 9653 4331\\
Fax: +55 84 3215 3781}

\author{Khalid Belhajjame \and Valeria~de~Castro \and Umberto~Souza~da~Costa \and Javier~A.~Espinosa-Oviedo \and Martin~A.~Musicante \and Pl\'acido~A.~Souza~Neto \and Genoveva~Vargas-Solar \and Jos\'e-Luis~Zechinelli-Martini}

\institute{K. Belhajjame
               \at Universit\'e de Paris - Dauphine -- Paris, France
               \and
               V. de Castro
               \at Universidad Rey Juan Carlos -- M\'{o}stoles, Spain
               \and
               U.S.Costa \and M.A. Musicante
               \at Universidade Federal do Rio Grande do Norte -- Natal, Brazil
               \and
               J.A. Espinosa-Oviedo, J-L Zechinelli-Martini
               \at Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico
               \and 
               P. Souza Neto 
               \at Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil
               \and
               G. Vargas-Solar
               \at CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France}

\maketitle

\begin{abstract}

Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
The late inclusion of NFRs in the development may compromise the quality of the deployed application.
This paper presents \pisodm, a methodology and associated tools that 
\textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details; 
\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
The pro\-po\-sed solution has been utilized in the context of an industrial case study.
\end{abstract}

\keywords{MDA \and Non-Functional Requirements \and Service-based software process}

%% main text
%*********************************************************************************************************
\section{Introduction}
\label{sec:intro}
\input{intro}

\section{Related Works}
\label{sec:relworks}
\input{relworks}

\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}
\input{modelingWithPISODM}

\section{Transformation rules}\label{sec:mmrules}
\input{transformationRules}

\section{Applying \pisodm: The \FlyingPig\ use case}
\label{sec:flyingPig}

\input{modelingFlyingPig}

\section{Conclusions and future work}\label{sec:conclusions}
\input{conclusions}

\bibliographystyle{plain}
\bibliography{biblio}

\end{document}

### 2014-11-12 JournalToSubmit2014/modelingFlyingPig.tex
new journal version

--- 2014-09-10 FlyingPig-journal/modelingFlyingPig.tex
@@
 Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
-Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security for accessing reserved areas (such as computing server installations).
+Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
 The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
@@
 
-Our goal is to model a service based application (called \FlyingPig), for providing risk assessment as a service.
-In order to provide this functionality, \FlyingPig\ would benefit from ORCA's legacy services providing storage, assessment and data visualization functions.
+Our goal is to model a service based application (\FlyingPig), for providing risk assessment as a service.
+In order to provide this functionality, \FlyingPig\ benefits from ORCA's legacy services providing storage, assessment and data visualization functions.
 
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[width=0.7\textwidth]{figs/3ValueModel.pdf}
+\includegraphics[width=0.9\textwidth]{figs/3ValueModel.pdf}
 \hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
 \caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
-\end{figure}
+\end{figure*}
 
@@
 (i) \textsl{Questionnaire and Evidences} filled with information about the client company;
-(ii) \textsl{Amendments} which are ORCA's recommendations, based on the data provided by the answers to questionnaires; 
+(ii) \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires; 
 (iii) \textsl{Evaluation Reports} for the client companies; 
@@
 The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
-Then, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.
+In the next step, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.
 
-\begin{figure}[t]
+\begin{figure*}
 \centering
@@
 \caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
-\end{figure}
+\end{figure*}
 
@@
 This corresponds to the value model, in which the start stimulus triggers the whole process.
-The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
-Questionnaires are considered tasks to be performed by users. 
+The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
 Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
@@
 
-Business processes also have associated rules and constraints to define their non-functional requirements (NFR):
+Business processes also have rules and constraints to define their non-functional requirements (NFR):
 \begin{numtrivlist}
@@
 \item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
-\item All questionnaires need to be fully answered, in order to consider that a task is completed.
-\item There is a time limit (in days) for each amendment required by the system.
+\item All questionnaires need to be fully answered in order to consider that a task is completed.
+\item There is a time limit (in days) for amendments required by the system.
 \end{numtrivlist}
@@
 
-\begin{figure}[t]
+\begin{figure*}
 \centering
@@
 \caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
-\end{figure}
+\end{figure*}
 
@@
 
-Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Figure~\ref{fig:piUseCaseModel}). 
+Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Fig.~\ref{fig:piUseCaseModel}). 
 %Use cases 
@@
 
-In our model, each $\pi$-use case may be associated to one or more (non-functional) constraints (depicted as colored ovals in Figure~\ref{fig:piUseCaseModel}). 
-The model defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}. 
+Each $\pi$-use case may be associated to one or more (non-functional) constraints (coloured ovals in Figure~\ref{fig:piUseCaseModel}). 
+The mod\-el defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}. 
 Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
@@
 \caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
-\end{figure}
+\end{figure*}
 
@@
 
-The $\pi$-ServiceProcess model (Figure~\ref{fig:PiServiceProcessModel}) presents the workflow for \FlyingPig.
-The actions in this model were obtained by applying the $\pi$-use case transformation rules.
+The $\pi$-ServiceProcess model presents the workflow for \FlyingPig (Figure~\ref{fig:PiServiceProcessModel}).
+Actions in this model were obtained by applying the $\pi$-use case transformation rules.
 % described in Section~\ref{sec:pewsmetamodel}.
@@
 Use cases are transformed into \textit{actions} and are represented by white boxes.
-The restrictions associated to each $\pi$-use case are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions. 
+The restrictions associated to  $\pi$-use cases are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions. 
 We can see that this model refines the concepts defined in the $\pi$-UseCase model.
@@
 \caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
-\end{figure}
+\end{figure*}
 
@@
 %The previous model 
-is now  completed with the services that provides functions of the application  \FlyingPig\.
+is now  completed with the services that provide functions of the application  \FlyingPig\.
 The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
@@
 Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
+The model is given in Figure~\ref{fig:piPEWSFlyingPig}.
 
-%\begin{figure}[t]
-%\centering
-\begin{lstlisting}[basicstyle=\footnotesize,label=list:flyingPigPiPEWS,caption=pi-PEWS
-Specification: FlyingPig. ]!\label{fig:PiPEWSModel}! 
+\begin{figure*}
+\begin{verbatim}
 //Namespaces specify service URI
@@
  (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
-\end{lstlisting}
-%\caption{$\pi$-PEWS model for \FlyingPig.\label{fig:PiPEWSModel}}
-%\end{figure}
+\end{verbatim}
+\caption{$\pi$-PEWS Model for \FlyingPig}\label{fig:piPEWSFlyingPig}
+\end{figure*}
 
-%Contracts may be implemented as transversal concerns~\cite{aspects}.
-%Their conditions may be verified at specific point of the composition and their actions are executed if needed.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Lessons Learned}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 
@@
 
-Given a set of services with their exported methods known in advance or provided by a  service directory, building service-based applications can be  a simple task that implies expressing an application logic as a service composition.
-The challenge being  ensuring the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) today can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
+Given a set of services with their exported methods, building service-based applications may consist on expressing an application logic as a service composition. 
+During this task, we must ensure the compliance between the specification and the resulting application. 
+Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
 

### 2014-11-12 JournalToSubmit2014/modelingWithPISODM.tex
new journal version

--- 2014-05-07 JournalToSubmit2013/modelingWithPISODM.tex
@@
-This section presents \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
-\pisodm provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Platform Independent Models}) and PSM (\textit{Platform Specific Models}).
+In this section we present \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
+Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
 Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
@@
 
-The next subsections present the meta-models from higher to lower levels of abstraction.
+In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
 The presentation is exemplified by a running example.
@@
 
-\begin{figure}[h]
+\begin{figure*}
 \centering
@@
 \label{fig:piSOD-M}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[htpb]
+\begin{figure*}
 \center
@@
 \caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
-\end{figure}
+\end{figure*}
 
@@
 \begin{itemize}
-\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
+\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
 \begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item Facebook and Twitter services export methods for  updating the status of a given user:
+\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
 \begin{itemize} 
@@
 The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application. 
-It is also necessary to understand the business process of the application and the conditions in which the different steps of this process are executed input/out data and the dependencies among these steps.
-Therefore, our methodology proposes the BPMN meta-model as a tool for modeling this aspect of the application.
+However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed. 
+In order to face this shortcoming, our methodology proposes the BPMN meta-model as a tool for modelling this aspect of the application.
 
-%Note that the models at this level show\dots 
-%{\color{red} Valeria and Placido: Please explain the CIM models for this example.}
-%\hfill\openbox
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{BPMN meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
+\subsubsection{BPMN meta-model}
+
 BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow. 
@@
 %
-\begin{figure}[htpb]
+\begin{figure}
 \center
-\includegraphics[width=0.49\textwidth]{figs/SC.pdf}
+\includegraphics[width=0.45\textwidth]{figs/SC.pdf}
 \caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
@@
 The specification defines those parts of the system that do not change from one platform to another. 
-Our methodology defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess} and \textit{$\pi$-ServiceComposition}.
+Our methodology defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.
 
@@
 Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
-The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-functional requirements.
+The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
 The notion of \textit{policy} is used to describe NFRs. 
-(In Figure~\ref{fig:CIM:usecasemetamodel}, we have highlighted the concepts related to NFRs).
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing non-functional requirements with  the concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc non-functional requirement}, {\sc non-functional attribute} and {\sc Constraint}.
+(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, \ \ {\sc Non-Functional}\\ {\sc At\-tri\-bute} and {\sc Constraint}.
 An {\sc End Consumer} is represented by an {\sc Actor}. 
@@
 Each {\sc Business Collaborator} combines the features described in each  {\sc Package}. 
-Thus, the services functions can be specified as being grouped into packages.
+%Thus, the services functions can be specified as being grouped into packages.
 
- \begin{figure}[htpb]
+ \begin{figure*}
 \center
@@
 \caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
-\end{figure}
+\end{figure*}
 
- The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented as part of description of {\sc Use Cases} and {\sc Constraints}. 
-A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has its name, description, and which ones should be checked during the execution of the application. 
-Each {\sc Constraint} is represented as a stereotyped ({\sf constraint}) use case. 
+ The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts \ are \\ represented by {\sc Use Cases} and {\sc Constraints}. 
+A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
+Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint}) use case. 
 There are three kinds of constraint:
@@
 \textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
-These concepts are present in the following example.
+The use of these constraints is shown in the following example.
 
@@
 \caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
-\end{figure}
+\end{figure*}
 
@@
 The \textit{$\pi$-ServiceProcess} meta-model extends the UML activity diagram with the concept of \textit{contract} to represent constraints over data and actions. This concept is used to model 
-groups  of  constraints described in the \textit{$\pi$-UseCase}. 
-As shown in Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-ServiceProcess} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Service Activity}, {\sc Action} and {\sc Constraint}.
-The portion of the figure enclosed by a dotted line defines those concepts related to the representation of NFRs.
+groups  of  constraints  in the \textit{$\pi$-UseCase}. 
+In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, \ {\sc Activity}, \ {\sc Ser\-vice Acti\-vi\-ty}, \\ {\sc Action} and {\sc Constraint}.
+The dotted part of the figure defines those concepts related to the representation of NFRs.
 
@@
 \caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
-\end{figure}
+\end{figure*}
 
@@
 \caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
-\end{figure}
+\end{figure*}
 
@@
 
-As shown in Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model 
+In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model 
 provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
@@
 
-\begin{figure}[t]
+\begin{figure*}
 \centering 
@@
 \label{fig:e-scomposition-metamodel}
-\end{figure}
+\end{figure*}
 
-\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
+\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more execu\-ta\-ble nodes.
 
-\item In order to represent constraint types associated to services compositions, we introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
+\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
 An {\em A-policy} is defined by attributes and rules. 
-The conditions of each rule are evaluated and, if they are not verified, the actions of the rule will be performed.
-The {\sc Rule} meta-class represents the types of event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
+The conditions of each rule are evaluated.
+In case of failure, the actions of the rule will be performed.
+The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
 An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
@@
 Instances of this meta-model are represented as UML activity diagrams. 
-\begin{figure}[t]%[htpb]
+\begin{figure*}
 \centering
@@
 \label{fig:servicecompositionmodel}
-\end{figure}
+\end{figure*}
 
@@
 To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
- We model a business process that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
+ We model a business process with three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
 The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
@@
 On the notification of an event of that type, a credential is obtained using the username and password. 
-\hfill\openbox
 \end{example}
@@
 This level focuses on the functionality, in the context of a particular implementation platform.
-Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.
-At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
+Models at this level put together the platform-independent view \ with \ the \ specific \\ aspects of the platform to implement the system. 
+Models at the PSM level can be automatically translated into actual computer programs. 
 We have defined one meta-model at this level.
 
-The \textit{$\pi$-PEWS} meta-model provides concepts for modeling service compositions. 
+\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions. 
 Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
@@
 %
-\begin{figure}[t]
+\begin{figure*}
 \centering
@@
 \label{fig:PPEWSmetamodel}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure} 
+\begin{figure*} 
 \centering
@@
 \label{fig:Specific-Contract-Representation}
-\end{figure}
+\end{figure*}

### 2014-11-12 JournalToSubmit2014/relworks.tex
new journal version

--- 2014-09-10 FlyingPig-journal/relworks.tex
@@
 Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
-Most research on NFRs focusses on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
+
+Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
 In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.
 
-In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view, namely,  
+In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:   
 \textit{service level}, \textit{system level} and \textit{business level}.
@@
 In~\cite{XiaoCZBOLH08} the authors use the terms  
-\textit{non-functional attributes}, \textit{composition mo\-del entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
+\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
 The notion of non-functional attribute is used to describe NFRs of the abstract process model. 
@@
 Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs. 
-The approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
-Non-functional concern is a general term used to describe non-functional requirements (such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}). 
-Each concern is refined into a set of non-functional attributes.  
-Each non-functional attribute represents some behavior, to be refined into non-functional actions. 
-An example of non-functional action is \textit{encryption}, which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
+Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
+A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}. 
+Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where  
+a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions. 
+For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
 %Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.
@@
 Pastrana et al.~\cite{PastranaPK11} use a traditional design-by-contract approach~\cite{Meyer97}.
-They use the term \textit{contract} to describe non-functional requirements. 
-Contracts may have pre-conditions, post-conditions and invariants. 
-Each contract defines \textit{assertions} associated with \textit{quality properties}. 
+The authors use the term \textit{contract} to describe non-functional re\-quire\-ments.  
+Contracts may present pre-con\-di\-tions, post-conditions and invariants. 
+Also, a con\-tract may define \textit{assertions} associated with \textit{qual\-i\-ty properties}. 
 Each service may have as many associated \textit{contracts} as needed.
 
-Chollet et al.~\cite{CholletL09} associate (non-functional) \textit{quality properties} to 
+Chollet et al.~\cite{CholletL09} associate (non-func\-tion\-al) \textit{quality properties} to 
 (functional) activities. 
-The authors present a security meta-model that takes into account web service compositions. 
-The NFRs considered are \textit{authentication}, \textit{integrity} and \textit{confidentiality}. 
+They present a security meta-model that takes into account web service compositions. 
+In this work the non-functional requirements considered are  \textit{authentication}, \textit{integrity} and \textit{confidentiality}. 
 Each NFR is associated with a service activity.
@@
 \textit{characteristic}~\cite{DiamadopoulouMPS08}, 
-\textit{quality level}~\cite{ModicaTV09}, and
+\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
 \textit{value}~\cite{ThissenW06,BasinDL06}.
@@
 %The work  presents a development methodology, although this methodology is not centered on NFRs.
-The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development processes that take NFRs into account. 
+The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account. 
 In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs, 
-whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
+as  in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
 system specification, including NFRs. 
 In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be 
-annotated with functional and non-functional attributes (NFAs). 
-Functional and non-functional attributes are independently defined.
+annotated with functional as well as non-functional attributes (NFAs). 
+Functional and non-functional at\-tri\-butes are independently defined.
 They are attached to specific tasks later in the development of the application. 

### 2015-03-31 JournalToSubmit2015/conclusions.tex
first version for IJSEKE

--- 2014-11-12 JournalToSubmit2014/conclusions.tex
@@
-We presented \pisodm, an MDA methodology for designing and developing reliable service-based applications. 
+We presented \pisodm, an model-driven method for designing and developing reliable service-based applications.
 \pisodm extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
@@
 
-Our methodology includes two CIM-level models, three PIM-level models and one PSM-level model. 
+Our method includes two CIM-level models, three PIM-level models and one PSM-level model.
 We implemented the meta-models on the Eclipse platform and we validated the approach by using an industrially inspired use case.

### 2015-03-31 JournalToSubmit2015/intro.tex
first version for IJSEKE

--- 2014-11-12 JournalToSubmit2014/intro.tex
@@
 These models are successively refined from abstract specifications into actual computer programs.
+A recent literature review study concludes there are a very scarce attention to non-functional requirements in the development of service-oriented applications using model-driven methodologies~\cite{Ameller201542}.
 
@@
 
+The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
+In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.The models presented were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.
+
 This paper is organized as follows.
@@
 % Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology.
-Section~\ref{sec:flyingPig} describes the experiment that we conducted for validating our methodology and discusses lessons learned. 
+Section~\ref{sec:flyingPig} describes the \FlyingPig case study that we develop for validating our method and discusses lessons learned.
 Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.

### 2015-03-31 JournalToSubmit2015/main.tex
first version for IJSEKE

--- 2014-11-12 JournalToSubmit2014/main.tex
@@
-\documentclass[twocolumn]{svjour3}
+\documentclass{ws-ijseke}
+\usepackage{xspace}
 
-\usepackage{geometry}
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage{graphicx}
-\usepackage{xspace}
-\usepackage{amssymb} 
-\usepackage{epstopdf}
-
-\usepackage{amssymb}
-
-%\usepackage{natbib}
-
-%%% OUR MACROS %%%
-\newcommand{\COMMENT}[1]{ }
-
-\usepackage[usenames,dvipsnames]{xcolor}
-
-\usepackage{amsmath}
-%\usepackage[thmmarks,amsmath]{ntheorem}
-
-\newcommand{\openbox}{\leavevmode
-  \hbox to.77778em{%
-  \hfil\vrule
-  \vbox to.675em{\hrule width.6em\vfil\hrule}%
-  \vrule\hfil}}
-
-%\theoremstyle{plain}
-%\theoremheaderfont{\normalfont\bfseries}
-%\theorembodyfont{\normalfont}
-%\theoremseparator{}
-%\theoremindent0cm
-%\theoremnumbering{arabic}
-%\newtheorem{algo}{Algorithm}
-%
-%%\theoremstyle{plain}
-%%%\theoremheaderfont{\normalfont\itshape}
-%%\theoremheaderfont{\normalfont\bfseries}
-%%\theorembodyfont{\normalfont}
-%%\theoremseparator{}
-%%\theoremindent0cm
-%%\theoremnumbering{arabic}
-%%\theoremsymbol{\ensuremath{\openbox}} 
-%%\newtheorem{example}{Example}
-%
-%
-%\theoremstyle{plain}
-%\theoremheaderfont{\normalfont\bfseries}
-%\theorembodyfont{\normalfont}
-%\theoremseparator{.}
-%\theoremindent0cm
-%\theoremnumbering{arabic}
-%\theoremsymbol{\ensuremath{\Box}} 
-%\newtheorem{defi}{Definition}
-%
-%\theoremstyle{plain} 
-%\theoremsymbol{\ensuremath{\Box}} 
-%\theoremseparator{.} 
-%\newtheorem{prop}{Property}
-
+\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
 \def\FlyingPig{\textsl{FlyingPig}}
@@
 
-\usepackage{listings}
-
-\lstset{numbers=right, numbersep=5pt, numberstyle=\tiny, stepnumber=1,escapechar=\!,columns=fullflexible,
-        morekeywords={procedure,let,for,do,if,then,else,add,choose,end,while,
-        true,false,rise,exception,extend,resume,to,return,function}}
-
-\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
-
 \begin{document}
 
-\title{\pisodm: A New Methodology for Building Service-Oriented Applications}\thanks{\textsl{Contact author:}\\
-Martin A. Musicante\\
-UFRN - Cx Postal 1647, \\
-Campus Universitário Lagoa Nova \\
-59078-970, Natal, RN, Brazil\\[2mm]
-Telephone: +55 84 3215 3814 ext 1.208\\
-Mobile phone: +55 84 9653 4331\\
-Fax: +55 84 3215 3781}
+\markboth{Authors' Names}
+{Instructions for Typing Manuscripts (Paper's Title)}
 
-\author{Khalid Belhajjame \and Valeria~de~Castro \and Umberto~Souza~da~Costa \and Javier~A.~Espinosa-Oviedo \and Martin~A.~Musicante \and Pl\'acido~A.~Souza~Neto \and Genoveva~Vargas-Solar \and Jos\'e-Luis~Zechinelli-Martini}
+%%%%%%%%%%%%%%%%%%%%% Publisher's Area please ignore %%%%%%%%%%%%%%%
+%
+\catchline{}{}{}{}{}
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\institute{K. Belhajjame
-               \at Universit\'e de Paris - Dauphine -- Paris, France
-               \and
-               V. de Castro
-               \at Universidad Rey Juan Carlos -- M\'{o}stoles, Spain
-               \and
-               U.S.Costa \and M.A. Musicante
-               \at Universidade Federal do Rio Grande do Norte -- Natal, Brazil
-               \and
-               J.A. Espinosa-Oviedo, J-L Zechinelli-Martini
-               \at Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico
-               \and 
-               P. Souza Neto 
-               \at Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil
-               \and
-               G. Vargas-Solar
-               \at CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France}
+\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
+
+\author{Khalid Belhajjame}
+\address{Universit\'e de Paris - Dauphine -- Paris, France\\
+\email{kbelhajj@googlemail.com}
+}
+
+\author{Valeria~de~Castro} 
+\address{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
+\email{valeria.decastro@urjc.es}
+}
+
+\author{Umberto~Souza~da~Costa} 
+\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+\email{umberto@dimap.ufrn.br}
+}
+
+\author{Javier~A.~Espinosa-Oviedo} 
+\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
+\email{javiera.espinosa@gmail.com}
+}
+
+\author{Martin~A.~Musicante} 
+\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+\email{mam@dimap.ufrn.br}
+}
+
+\author{Pl\'acido~A.~Souza~Neto} 
+\address{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
+\email{placido.neto@ifrn.edu.br}
+}
+
+\author{Genoveva~Vargas-Solar} 
+\address{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
+\email{genoveva.vargas@imag.fr}
+}
+
+\author{Jos\'e-Luis~Zechinelli-Martini}
+\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
+\email{joseluis.zechinelli@udlap.mx}
+}
 
@@
 
+\begin{history}
+\received{(Day Month Year)}
+\revised{(Day Month Year)}
+\accepted{(Day Month Year)}
+%\comby{(xxxxxxxxxx)}
+\end{history}
+
 \begin{abstract}
-
 Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
 The late inclusion of NFRs in the development may compromise the quality of the deployed application.
-This paper presents \pisodm, a methodology and associated tools that 
+This paper presents \pisodm, a method and associated tools that
 \textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details;
 \textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
-The pro\-po\-sed solution has been utilized in the context of an industrial case study.
+The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
 \end{abstract}
@@
 
-\section{Applying \pisodm: The \FlyingPig\ use case}
+\section{Applying \pisodm: The \FlyingPig\ case study}
 \label{sec:flyingPig}

### 2015-03-31 JournalToSubmit2015/modelingFlyingPig.tex
first version for IJSEKE

--- 2014-11-12 JournalToSubmit2014/modelingFlyingPig.tex
@@
-We validated our methodology by developing a use-case concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
+To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
 Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.

### 2015-03-31 JournalToSubmit2015/modelingWithPISODM.tex
first version for IJSEKE

--- 2014-11-12 JournalToSubmit2014/modelingWithPISODM.tex
@@
-In this section we present \pisodm, an MDD-based methodology for building service compositions with non-functional requirements. 
+In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
 Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
@@
 However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
-In order to face this shortcoming, our methodology proposes the BPMN meta-model as a tool for modelling this aspect of the application.
+In order to face this shortcoming, our method proposes the BPMN meta-model as a tool for modelling this aspect of the application.
 
@@
 The CIM level models are the basis for the development of the application.
-The information presented at this level is (manually) refined into PIM-level models of the methodology \pisodm described next.
+The information presented at this level is (manually) refined into PIM-level models of the method \pisodm described next.
 Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.
@@
 The specification defines those parts of the system that do not change from one platform to another.
-Our methodology defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.
+Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.
 

### 2015-03-31 JournalToSubmit2015/relworks.tex
first version for IJSEKE

--- 2014-11-12 JournalToSubmit2014/relworks.tex
@@
 
+Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}.
+As Ameller et al.
+realize in a recent published literature review \textit{"Not many papers that presented a generic approach to deal with NFRs"}.
+Authors also remark in such study the lacks of industrial use case~\cite{Ameller201542}.
+Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support in addition to real need presented in a company related to the risk assessment industry.
+
 %Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
@@
 
-\bigskip
-Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
-The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
-Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
-It may use web services provided by third party sources.
-The System level deals with the coordination of services, to implement the business logic.
-Our proposal, shown in the next section, follows these ideas.
+%\bigskip
+%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
+%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
+%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
+%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
+%It may use web services provided by third party sources.
+%The System level deals with the coordination of services, to implement the business logic.
+%Our proposal, shown in the next section, follows these ideas.

### 2015-03-31 JournalToSubmit2015/transformationRules.tex
first version for IJSEKE

--- 2014-05-07 JournalToSubmit2013/transformationRules.tex
@@
 Constraints are transformed according to their type:
-{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} attribute set to false are transformed into {\sf Assertion}s;
+{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} at\-tri\-bute set to false are transformed into {\sf Assertion}s;
 {\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
 
-In order to transform constraints of type {\sf Value Constraint}, the designer must specify thresholds to be associated to the assertions of a contract.
+In order to transform constraints of type {\sf Value Constraint}, the designer specifies thresholds to be associated to the assertions of a contract.
 By default, value constraints are transformed into pre-conditions and business constraints are transformed into post-conditions.
@@
 
-The details of these transformations are not included here (due to space restrictions).
+Details of these transformations are not included here (due to space restrictions).
 They can be found in~\cite{SouzaNeto:2012}.
 
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1}
-The rules presented above have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} \\
+The transformation rules have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).
 
@@
 For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}.
-The ``download music'' use case  includes the payment process to buy the music. 
+The use case ``download music''  includes the payment process to buy the music. 
 Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
@@
 
-{\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their homonym concepts of the $\pi$-Service Composition model.
+{\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their hom\-onym concepts of the $\pi$-Service Composition model.
 
@@
 
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}\\
 Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
@@
 
-This section describes the PIM to PSM transformations from a $\pi$-Service composition model to a $\pi$-PEWS model. 
-We distinguish two groups of rules: \textit{(i)} those transforming service composition entities into workflows; and \textit{(ii)} those that transform  A-Policies into Contracts.
+This section describes the PIM to PSM transformations from a $\pi$-Service composition mod\-el to a $\pi$-PEWS model. 
+We distinguish two groups of rules: \textit{(i)} those transforming service composition entities into workflows; and \textit{(ii)} those transforming  A-Policies into Contracts.
 
-Single actions (represented by {\sf Action} and {\sf
+Single actions ({\sf Action} and {\sf
 Action:name}) are transformed into individual service {\sf Operations}.
-Complex actions (represented by {\sf ServiceActivity}  and  {\sf
+Complex actions (rep\-re\-sented by {\sf ServiceActivity}  and  {\sf
 ServiceActivity:name}) are transformed into (named) composite
 operations that defines a (named) workflow of the application.
-Composition patterns expressed using the operators {\sc\em merge, decision, fork and join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
+Composition patterns expressed using  {\em merge, decision, fork} and {\em join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
 
@@
 
-The A-Policies defined for the entities of a $\pi$-service composition model are transformed into {\sf A-policy} entities, named according to the names expressed in the source model.
-The transformation of the rules expressed in a $\pi$-service composition is guided by the event types associated to these rules. 
-The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into entities  {\sf Variable} with attributes {\sf Name} and {\sf Type} directly specified from the elements {\sf Variable:name} and {\sf Variable:type} of a $\pi$-service composition model.
+A-Policies defined for the entities of a $\pi$-service composition model generate {\sf A-policy} entities, named according to the names expres\-sed in the source model.
+The transformation of the rules expres\-sed in a $\pi$-service composition is guided by the event types associated to these rules. 
+The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into  {\sf Variable}, with  {\sf Name} and {\sf Type} attributes specified from the elements {\sf Variable: name} and {\sf Variable:type} of a $\pi$-service composition model.
 
@@
 We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
-
 Our tool is implemented as a series of Eclipse plug-ins: 
 \begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   for implementing the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
-Then, from these meta-models, we  developed plug-ins to support their graphical representation.
+\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
+From these meta-models, we  developed plug-ins to support their graphical representation.
 
@@
 
-\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} for implementing  the code generation plug-in for generating executable code. 
+\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
 It takes a $\pi$-PEWS model and generates the code to be executed by the {\em

### 2015-03-31 JournalToSubmit2015/transformations.tex
first version for IJSEKE

--- 2013-12-06 JournalToSubmit2013/transformations.tex
@@
 
-\begin{figure}[t]
-\centering{
-\includegraphics[width=0.96\textwidth]{figs/Mapping-1a.png}
-}
-\caption{ $\pi$-SCM to $\pi$-{\sc Pews} transformations.}
-\label{fig:transformationsA}
-\end{figure}
-
 \subsection{Transformation of A-policies and rules into $\pi$-{\sc Pews}}
@@
 
-As shown in Figures~\ref{fig:transformationsA} and~\ref{fig:transformationsB}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
+As shown in Figures~\ref{fig:transformationsB} and~\ref{fig:transformationsC}, for an event of type {\sc\em Pre} the corresponding transformed rule is of type {\sc Precondition}; for an event of type {\sc\em Post} the corresponding transformed rule is of type {\sc Postcondition}; finally, for an event of type {\sc\em TimeRestriction} the corresponding transformed rule is of type {\sc Time}.
 The condition inside a $\pi$-SCM rule ({\sc\em Rule:condition}) is transformed into a class {\sc\em Condition:expression} where the attributes of the expression are transformed into elements of type {\sc Attribute}.

### 2015-04-02 JournalToSubmit2015/modelingWithPISODM.tex
version to submit?

@@
 (Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, \ \ {\sc Non-Functional}\\ {\sc At\-tri\-bute} and {\sc Constraint}.
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
 An {\sc End Consumer} is represented by an {\sc Actor}.
@@
 
- The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts \ are \\ represented by {\sc Use Cases} and {\sc Constraints}.
+ The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
 A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
@@
 groups  of  constraints  in the \textit{$\pi$-UseCase}.
-In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, \ {\sc Activity}, \ {\sc Ser\-vice Acti\-vi\-ty}, \\ {\sc Action} and {\sc Constraint}.
+In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Ser\-vice Acti\-vi\-ty}, {\sc Action} and {\sc Constraint}.
 The dotted part of the figure defines those concepts related to the representation of NFRs.
@@
 This level focuses on the functionality, in the context of a particular implementation platform.
-Models at this level put together the platform-independent view \ with \ the \ specific \\ aspects of the platform to implement the system.
+Models at this level put together the platform-independent view with the specific aspects of the platform to implement the system.
 Models at the PSM level can be automatically translated into actual computer programs.

### 2015-04-02 JournalToSubmit2015/relworks.tex
version to submit?

@@
 As Ameller et al.
-realize in a recent published literature review \textit{"Not many papers that presented a generic approach to deal with NFRs"}.
-Authors also remark in such study the lacks of industrial use case~\cite{Ameller201542}.
-Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support in addition to real need presented in a company related to the risk assessment industry.
+realize in a recent published literature review that there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
+Authors also remark in such study the lack of industrial use cases~\cite{Ameller201542}.
+Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.
 

### 2015-04-07 JournalToSubmit2015/relworks.tex
Small changes on Section 2.

@@
-While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is preformed.
+While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
 NFRs include aspects such as performance, authentication and quality constraints.
-These requirements are usually specified by conditions, called Non-functional Properties.
+These requirements are usually specified by conditions, called non-functional properties.
 Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
@@
 As Ameller et al.
-realize in a recent published literature review that there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
-Authors also remark in such study the lack of industrial use cases~\cite{Ameller201542}.
+realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
+Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
 Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.

### 2015-04-10 JournalToSubmit2015/submitted/main.tex
version to submit

\documentclass{ws-ijseke}
\usepackage{xspace}

\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
\def\FlyingPig{\textsl{FlyingPig}\xspace}

\newcounter{numberInTrivlist}

\newenvironment{numtrivlist}{\begin{list}{\rm \arabic{numberInTrivlist})}
                                         {\usecounter{numberInTrivlist}
                                          \setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\newenvironment{itemizedTrivlist}{\begin{list}{\rm ~\hspace{2mm} $\bullet$\ }
                                         {\setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\begin{document}

\markboth{K.Belhajjame \textit{et al.}}
{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}

%%%%%%%%%%%%%%%%%%%%% Publisher's Area please ignore %%%%%%%%%%%%%%%
%
\catchline{}{}{}{}{}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}

\author{Khalid Belhajjame}
\address{Universit\'e de Paris - Dauphine -- Paris, France\\
\email{kbelhajj@googlemail.com}
}

\author{Valeria~de~Castro} 
\address{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
\email{valeria.decastro@urjc.es}
}

\author{Umberto~Souza~da~Costa} 
\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
\email{umberto@dimap.ufrn.br}
}

\author{Javier~A.~Espinosa-Oviedo} 
\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
\email{javiera.espinosa@gmail.com}
}

\author{Martin~A.~Musicante} 
\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
\email{mam@dimap.ufrn.br}
}

\author{Pl\'acido~A.~Souza~Neto} 
\address{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
\email{placido.neto@ifrn.edu.br}
}

\author{Genoveva~Vargas-Solar} 
\address{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
\email{genoveva.vargas@imag.fr}
}

\author{Jos\'e-Luis~Zechinelli-Martini}
\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
\email{joseluis.zechinelli@udlap.mx}
}

\maketitle

\begin{history}
\received{(Day Month Year)}
\revised{(Day Month Year)}
\accepted{(Day Month Year)}
%\comby{(xxxxxxxxxx)}
\end{history}

\begin{abstract}
Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
The late inclusion of NFRs in the development may compromise the quality of the deployed application.
This paper presents \pisodm, a method and associated tools that
\textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details;
\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
\end{abstract}

\keywords{MDA \and Non-Functional Requirements \and Service-based software process}

%% main text
%*********************************************************************************************************
\section{Introduction}
\label{sec:intro}

%Service oriented computing is at the origin of an evolution in the field of software development.
%Service oriented methods advocates for the construction of software systems formed by the composition of heterogeneous, loosely coupled modules.
%These modules (or services) communicate in order to achieve a common purpose.

%(Moreover, IT systems need to evolve according to the business needs.)
%Thus, organizations are seeking for mechanisms to bridge the gap between the actually developed systems and their business needs~\cite{bell}.

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
combined to build an application business logic.
The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.

Functional properties are not the only aspect to be considered in the software development process.
Non-functional properties, such as data privacy, exception handling, atomicity  and, data persistence, need to be addressed to fit in the application.
%Adding non-functional properties and respecting services constraints while composing services is a complex task that implies implementing protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
%Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements.
%
%Furthermore,  often they are  not fully considered in the specification process of existing services' oriented development methods. Rather,
%They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed.
Ideally, Non-Functional Requirements (NFR) should be considered in every phase of the software development.
Yet, they are partially or rarely methodologically derived from the specification, being usually added once the code has been im\-ple\-men\-ted.
In the context of service oriented computing, the late consideration of non-functional requirements in the development process does not fully preserve the compliance and re\-use expectations promoted by the service oriented paradigm.

%Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
%Most methods integrate them only after the application has been implemented.
%This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.

% song title disseminated in the walls of the user's Facebook and Twitter accounts.

%The adoption of non-functional specifications from the early states of development can help the developer to produce applications that  can deal with the application context.

The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
These models are successively refined from abstract specifications into actual computer programs. 
A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR. Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.

\bigskip
Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
\pisodm proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller} approach.
The goals of  \pisodm are to:
%SOD-M does not provide support for the specification of non-functional requirements, such as
%security, reliability, and efficiency.
\begin{trivlist}
\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
\item \textit{(iii)} Provide a general way of describing non-functional requirements.
\end{trivlist}

The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}. In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.

\bigskip
This paper is organized as follows.
Sec\-tion \ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
Sections~\ref{sec:motivation} and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules.
% Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology.
Section~\ref{sec:flyingPig} describes the \FlyingPig case study that we develop for validating our method and discusses lessons learned.
Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.

\section{Related Work}
\label{sec:relworks}
While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
NFRs include aspects such as performance, authentication and quality constraints.
These requirements are usually specified by conditions, called non-functional properties.
Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.

Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.

In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:
\textit{service level}, \textit{system level} and \textit{business level}.
%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}. The work also defines a \textit{web service quality model}, which considers non-functional properties.
In~\cite{XiaoCZBOLH08} the authors use the terms
\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
The notion of non-functional attribute is used to describe NFRs of the abstract process model.
In the lower level, the composition is annotated with non-functional attributes.

D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}.
\textit{Quality dimensions} are used to quantify an individual characteristic.
For instance, the quality category \textit{performance} groups characteristics such as
\textit{latency} and \textit{throughput}.
The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services. A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties such as \textit{availability}, \textit{reliability} and \textit{access control}.

Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions.
For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.

Pastrana et al.~\cite{PastranaPK11} use a traditional design-by-contract approach~\cite{Meyer97}.
The authors use the term \textit{contract} to describe non-functional re\-quire\-ments.
Contracts may present pre-con\-di\-tions, post-conditions and invariants.
Also, a con\-tract may define \textit{assertions} associated with \textit{qual\-i\-ty properties}.
Each service may have as many associated \textit{contracts} as needed.

Chollet et al.~\cite{CholletL09} associate (non-func\-tion\-al) \textit{quality properties} to
(functional) activities.
They present a security meta-model that takes into account web service compositions.
In this work the non-functional requirements considered are  \textit{authentication}, \textit{integrity} and \textit{confidentiality}.
Each NFR is associated with a service activity.

Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services.
Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}.
Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09},
\textit{property}~\cite{Fabra2011},
\textit{factor}~\cite{MohantyRP10,GutierrezRF10},
\textit{characteristic}~\cite{DiamadopoulouMPS08},
\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
\textit{value}~\cite{ThissenW06,BasinDL06}.

Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
However, distinct hierarchies and models are proposed for NFRs, according to different points of view.
We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.

Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}. As Ameller et al. realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}. Those authors also remark the lack of industrial use cases~\cite{Ameller201542}. 
Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.

%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
%The work  presents a development methodology, although this methodology is not centered on NFRs.
The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account.
In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs,
as  in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
system specification, including NFRs.
In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be
annotated with functional as well as non-functional attributes (NFAs).
Functional and non-functional at\-tri\-butes are independently defined.
They are attached to specific tasks later in the development of the application.
NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.

The proposal in~\cite{ThissenW06} presents steps for  selecting services
by taking QoS information into account. The proposed steps are:
\textit{(i)} identification of relevant QoS information;
\textit{(ii)} identification of basic composition patterns and
QoS aggregation rules for these patterns; and
\textit{(iii)} definition of a selection mechanism of services.
The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
\textit{availability}.

Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.

Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.

Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.
For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.

%\bigskip
%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
%It may use web services provided by third party sources.
%The System level deals with the coordination of services, to implement the business logic.
%Our proposal, shown in the next section, follows these ideas.

\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}

In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.

In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
The presentation is exemplified by a running example.
The transformations between models are presented in section~\ref{sec:mmrules}.

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
\caption{\pisodm Overview.}
\label{fig:piSOD-M}
\end{figure*}

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Computation Independent Models}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

This level focuses on the highest-level view of the system, including its business and requirement specifications.
At this stage of the development, the structure and system processing details are still unknown or undetermined.
\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{E$^3$value meta-model}
% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The e$^3$value model of an application identifies the value/information exchange between components of the system.
It is a business model that represents a business case graphically as a set of value exchanges\footnote{See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
The model is well suited to enhance the understanding of the environment in which an application is being developed.
It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}).
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

\begin{figure*}
\center
\includegraphics[width=0.75\textwidth]{figs/e3value.pdf}
\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
\end{figure*}

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
Let us consider the scenario ``To Publish Music'', used as a running example in this section:
An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
The following (external) services will be used by the application:
\begin{itemize}
\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
\begin{itemize}
%\item{\sf\small get-Status ( usedid ): String ;
\item {\sf\small update-Status ( userid, new-status ): String};
\end{itemize}
\end{itemize}

The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
The e$^3$value value model shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users. The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription). Finally, the private application interacts with Facebook and Twitter for updating the user's status.
We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
\end{example}

The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
In order to face this shortcoming, our method proposes the BPMN meta-model as a tool for modelling this aspect of the application.

\subsubsection{BPMN meta-model}

BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow.
The next example illustrates the use of this meta-model.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\end{example}
%
\begin{figure}
\center
\includegraphics[width=0.45\textwidth]{figs/SC.pdf}
\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
\end{figure}

The CIM level models are the basis for the development of the application.
The information presented at this level is (manually) refined into PIM-level models of the method \pisodm described next.
Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Platform Independent Models}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

This level focuses on the system functionality, hiding the details of any particular platform.
The specification defines those parts of the system that do not change from one platform to another.
Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-UseCase} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
The notion of \textit{policy} is used to describe NFRs.
(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
An {\sc End Consumer} is represented by an {\sc Actor}.
An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
 The {\sc Business Collaborator} concept is represented  through {\sc Packages}.
A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled.
Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
%Thus, the services functions can be specified as being grouped into packages.

 \begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{figs/UseCaseMetaModel.pdf}
\caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
\end{figure*}

 The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint}) use case.
There are three kinds of constraint:
\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value'';
\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
The use of these constraints is shown in the following example.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example application.
We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).

The ``To Publish Music'' Business Service expects  the Facebook or Twitter user status to be changed every time a user starts listening a new song.
Therefore, it is necessary to perform a social network authentication with the users data. Each social network uses different services and different forms of authentication. The authentication constraint is required to update a music status. The restriction is stereotyped as a value constraint, because the users id and password are verified.  Figure \ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases. The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as value and business constraint, respectively. For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
Another restriction is that the minimum payment value is 2 euros.
%{\color{red} Talk about this model for the example.}
\end{example}

\begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{figs/UseCase.pdf}
\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
\end{figure*}

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The \textit{$\pi$-ServiceProcess} meta-model extends the UML activity diagram with the concept of \textit{contract} to represent constraints over data and actions. This concept is used to model
groups  of  constraints  in the \textit{$\pi$-UseCase}.
In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Ser\-vice Acti\-vi\-ty}, {\sc Action} and {\sc Constraint}.
The dotted part of the figure defines those concepts related to the representation of NFRs.

A specific \textit{$\pi$-ServiceProcess} model shows the set of logically related activities to be performed in a service-based application.
So, the activities of this model represent a behavior that is part of the business logic of the application.
A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract. A service activity represents an operation that is part of the execution flow, and it is modeled as an {\sc Action}.
An activity contract represents the {\sc Non-functional Requirement}  that is also part of the execution flow of a service, identified  as a stereotyped activity ({\sf assertion}).
The {\sc Assertion}s associated to an {\sc Action} compose a {\sc Contract}. 
Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre-conditions and post-conditions.

\begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{figs/ServiceProcessMetaModel.pdf}
\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
\end{figure*}

%According to the meta-model, a \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts. It models the set of system activities and describe contracts for each activity or the process as a whole. Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action. This model defines a workflow for  modeling business logic and the behaviour of the application being developed. The behaviour is represented by each action and the restrictions by the assertion description. All assertions are described by stereotyped actions.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}. The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service. The buy music pre-conditions consist in verifying:
(i) if the User data are correct;
(ii) if the User is already logged in Spotify;
(iii) if bank account information are correct and;
(iv) if there are enough funds in the bank account to cover the payment.
A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization. There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. To update services, depending of each service, there may be different restrictions. As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter. In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook.
As post-condition, the application ensures that the Facebook service sends a notification of success.
To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modeled as post-condition).
As a pre-condition for ``twitter update'' it is necessary that (i) the music format is correct and (ii) the twitter login and password are correct for the update.
\end{example}

\begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{figs/ServiceProcess.pdf}
\caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
\end{figure*}

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
to group contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a security policy.
 The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action.
An instance of this meta-class is graphically represented as a partition in the activity diagram.
A collaborator can be either internal or external to the system.
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class \textsc{Action}.
A class action represents some type of transformation or processing.
There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition}
\caption{$\pi$-Service Composition Meta-model.}
\label{fig:e-scomposition-metamodel}
\end{figure*}

\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more execu\-ta\-ble nodes.

\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules.
The conditions of each rule are evaluated.
In case of failure, the actions of the rule will be performed.
The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.
\end{itemizedTrivlist}
Instances of this meta-model are represented as UML activity diagrams.
\begin{figure*}
\centering
\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
\caption{$\pi$-ServiceComposition Model for the ``To publish
music'' business service.}
\label{fig:servicecompositionmodel}
\end{figure*}

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}).
 We model a business process with three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
Both {\em Facebook} and {\em Twitter} require authentication.
Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
%{\color{red} To be completed!.}
In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that read and update the user data.
%A call to such services must be part of the authentication protocol required by these services.
In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
{\sf\small OAuthPolicy} implements the open authentication protocol.
The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
This variable is imported through the library {\sf\small OAuthPolicy.Token}.
The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol.
The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
On the notification of an event of that type, a credential is obtained using the username and password.
\end{example}

%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
%These artifacts will be used to generate the actual programs that will implement the application:
Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation.
The $\pi$-PEWS  meta-model is described in the next section.

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Platform Specific Models}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

This level focuses on the functionality, in the context of a particular implementation platform.
Models at this level put together the platform-independent view with the specific aspects of the platform to implement the system.
Models at the PSM level can be automatically translated into actual computer programs.
We have defined one meta-model at this level.

\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.

PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
\begin{itemizedTrivlist}
\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}.
A {\sc Compound Operation} is defined using an {\sc Operator}.
The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) repetition.

\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
A brief description of these classes is given next.
\end{itemizedTrivlist}
%
\begin{figure*}
\centering
\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
\caption{$\pi$-{\sc Pews} Meta-model.}
\label{fig:PPEWSmetamodel}
\end{figure*}

Figure~\ref{fig:PPEWSmetamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.
In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic5}
Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our scenario example.
Notice that the code contains namespaces and definitions (obtained by using information about the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
\end{example}

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{figs/pi-pewsSpecification-toPublishMusic}
\caption{$\pi$-PEWS Specific and Policy Representation.}
\label{fig:Specific-Contract-Representation}
\end{figure*}

\section{Transformation rules}\label{sec:mmrules}

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..

\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models. 
These rules are used to build a semi-automatic tool (the \pisodm environment plugin for Eclipse).
The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-UseCase to the $\pi$-ServiceProcess}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .

The refinement of a (composite) $\pi$-Use Case model into a $\pi$-Service Process model is driven by the principle of expressing a set of $\pi$-use cases (i.e., use cases with constraints)   in terms of  a business process.
The resulting model consists of actions related by a control flow and contracts specifying NFPs.

As defined by the $\pi$-ServiceProcess  meta-model (Figure~\ref{fig:CIM:serviceprocessmetamodel}) an {\sc Activity Service} consists of a composition of  entities of type {\sc Action}. 
Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
transformed into a  {\sf Fork node}~\cite{valeriaThesis,placidoPhDThesis2012}.
If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/35}}
%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
%\label{fig:transformationsUseCase-ServiceProcessRules}
%\end{figure}

A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
The set of resulting assertions  are grouped into a {\sf Contract}.
Constraints are transformed according to their type:
{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} at\-tri\-bute set to false are transformed into {\sf Assertion}s;
{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.

In order to transform constraints of type {\sf Value Constraint}, the designer specifies thresholds to be associated to the assertions of a contract.
By default, value constraints are transformed into pre-conditions and business constraints are transformed into post-conditions.

%A {\sf Requirement} and a {\sf Composite use case} in a  $\pi$-UseCase model are transformed respectively into a {\sf Service activity} and an {\sf Action} (see figures  \ref{fig:transformationsUseCase-ServiceProcessRules}-c  and d). 
%
%
%\begin{figure}
%\centering{
%\includegraphics[width=0.80\textwidth]{figs/37}}
%\caption{ Extended Transformation Examples}
%\label{fig:transformaton-examples}
%\end{figure}

%The transformations for {\sc Extend} and {\sc Include} dependencies  are not as simple as the previous transformations (figures \ref{fig:transformaton-examples}-a).

The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  
The generated workflow is composed by {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/36}}
%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
%\label{fig:transformationsUseCase-ServiceProcessRules}
%\end{figure}

{\sf Include} use case entities are transformed into an {\sf Action} sequence.
A {\sc Use case} element is transformed into an {\sf Action}. 
A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements.

Details of these transformations are not included here (due to space restrictions).
They can be found in~\cite{SouzaNeto:2012}.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} \\
The transformation rules have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).

The ``listen music'' use case is transformed into a Service Action. 
This Action  represents a Spotify service function that can be invoked to play the music. 
For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}. 
The use case ``download music''  includes the payment process to buy the music. 
Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
%(as shown in figure \ref{fig:transformation-example-include}). 
The same rule is applied to the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
 \end{example}

%\begin{figure}
%\centering{
%\includegraphics[width=0.76\textwidth]{figs/38}}
%\caption{ Include Transformation Example}
%\label{fig:transformation-example-include}
%\end{figure}

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .

The  principle of the transformation  of a  $\pi$-Service Process model into a $\pi$-Service Composition model is to group {\sf Contracts} into {\sf A-Policies} and {\sf Actions} into {\sf Service Activities}.

Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule}. 
{\sf Rules}  concerning the same NFP  are grouped into {\sf A-Policies}. 
Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute. 
If the {\sf Assertion} has a value type, the name and the attributes are transformed into {\sf Variables} in the target model.  
The {\sf Assertion: aProperty} attribute can have different transformations, according to: 
\textit{(i)} a {\sf Precondition} is transformed into {\sf Pre};
\textit{(ii)} a {\sf Post-Condition} is transformed into a {\sf  Post};
\textit{(iii)} a {\sf TimeRestriction} is transformed into {\sf Time}.

%
%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
%\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
%\label{fig:ServiceProcess-ServiceComposition-Rules}
%\end{figure}

Some elements of the $\pi$-service composition model are obtained from the $\pi$-UseCase model: 
A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Functional requirements} of the $\pi$-ServiceComposition model. 
These requirements are associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).

{\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their hom\-onym concepts of the $\pi$-Service Composition model.

Finally, {\sf Actions} are grouped into  a {\sf Business collaborator}.
%\begin{figure}
%\centering{
%\includegraphics[width=1\textwidth]{figs/ServiceProcess-ServiceComposition}}
%\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
%\label{fig:ServiceProcess-ServiceComposition}
%\end{figure}

%As the $\pi$-ServiceComposition model refines the $\pi$-ServiceProcess concepts at PIM level, a service previously defined as actions (source model) is refined as composition of those actions (target model) that are necessary to represent a business service, identifying who are the partners involved in the realization ({\sc Business collaborators}). 
%In addition, $\pi$-SOD-M defines a platform specific model based on web services composition. This model is explicitly indicates those actions which are (or will be, if not yet implemented) supported by web services.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}\\
Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
The ``securityLoginPolicy'' consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
The information about Facebook and Spotify (both of them {\sf Business
Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
model.
\end{example}

% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
\subsection{From $\pi$-ServiceComposition to $\pi$-PEWS}
% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .

This section describes the PIM to PSM transformations from a $\pi$-Service composition mod\-el to a $\pi$-PEWS model. 
We distinguish two groups of rules: \textit{(i)} those transforming service composition entities into workflows; and \textit{(ii)} those transforming  A-Policies into Contracts.

Single actions ({\sf Action} and {\sf
Action:name}) are transformed into individual service {\sf Operations}.
Complex actions (rep\-re\-sented by {\sf ServiceActivity}  and  {\sf
ServiceActivity:name}) are transformed into (named) composite
operations that defines a (named) workflow of the application.
Composition patterns expressed using  {\em merge, decision, fork} and {\em join} are transformed into their corresponding workflows of the $\pi$-PEWS model.

%\begin{figure}
%\centering{
%\includegraphics[width=0.96\textwidth]{figs/Table7}}
%\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
%\label{fig:ServiceComposition-Pews-Rules}
%\end{figure}

A-Policies defined for the entities of a $\pi$-service composition model generate {\sf A-policy} entities, named according to the names expres\-sed in the source model.
The transformation of the rules expres\-sed in a $\pi$-service composition is guided by the event types associated to these rules. 
The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into  {\sf Variable}, with  {\sf Name} and {\sf Type} attributes specified from the elements {\sf Variable: name} and {\sf Variable:type} of a $\pi$-service composition model.

Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Pre-conditions}, {\sf Post-condition} and {\sf TimeRestrictions}.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT6}
Figure \ref{fig:piSOD-M} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.
\end{example}

\subsection{Implementation}
We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
Our tool is implemented as a series of Eclipse plug-ins: 
\begin{itemize}
\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models. 
From these meta-models, we  developed plug-ins to support their graphical representation.

\item	 We used  ATL\footnote{http://eclipse.org/atl/. An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
for implementing the  mappings between models.

\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
A-Policy} based service composition execution environment.
\end{itemize}

\section{Applying \pisodm: The \FlyingPig\ case study}
\label{sec:flyingPig}

To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.

Our goal is to model a service based application (\FlyingPig), for providing risk assessment as a service.
In order to provide this functionality, \FlyingPig\ benefits from ORCA's legacy services providing storage, assessment and data visualization functions.

In the following sections, we describe the results of applying \pisodm to develop the \FlyingPig\ risk assessment system.
The models presented next were generated as a result of interacting with software developers at GCP Global.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figs/3ValueModel.pdf}
\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
\end{figure*}

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Computation-Independent Models (CIM)}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
%\footnote{\color{red} This sentence need to move to section 3.1!!}
We identify two business actors: \textsl{ORCA} and \textsl{Broker}.
Brokers emit requests for risk assessment for one or more companies.
ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}.
The values exchanged between ORCA and the brokers are:
(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
(ii) \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires;
(iii) \textsl{Evaluation Reports} for the client companies;
and
(iv) the risk assessment \textsl{fee}.

The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company.
Once this need has been declared, the value exchanges between ORCA and Broker are triggered.
The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
In the next step, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
\end{figure*}

Figure~\ref{fig:BPMNmodel} shows the BPMN model
%\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.}
for the \FlyingPig\ scenario. The BPMN model for \FlyingPig\ is partitioned for better understanding the process in which the value exchanges occur.
The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
Brokers have two lanes, the client \textsl{Company} and a \textsl{User}.
The user is a contact member of the company, who  coordinates the assessment process.
This process  involves other members of the company as well.

The risk assessment process starts after a request from a company.
This corresponds to the value model, in which the start stimulus triggers the whole process.
The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.

Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.

Business processes also have rules and constraints to define their non-functional requirements (NFR):
\begin{numtrivlist}
\item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment.
\item The system should be able to deal with, at least, 200 users.
\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests.
\item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
\item All questionnaires need to be fully answered in order to consider that a task is completed.
\item There is a time limit (in days) for amendments required by the system.
\end{numtrivlist}

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Platform-Independent Models (PIM)}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

%In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
The \pisodm models of the PIM level  for the \FlyingPig\ scenario are presented next.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
\end{figure*}

\paragraph{\underline{$\pi$-UseCase Model for \FlyingPig}}~

The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application.
In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
In the context of \FlyingPig, Company is the actor requesting a risk evaluation.
A Broker is  responsible for coordinating  the evaluation process, assigning users to be in charge of tasks as well as delegating tasks.
A User, in this model is an actor who answers questionnaires according to the current situation  of the Company.
The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.

Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Fig.~\ref{fig:piUseCaseModel}).
%Use cases
that describe the main functionality of the system.
The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.

Each $\pi$-use case may be associated to one or more (non-functional) constraints (coloured ovals in Figure~\ref{fig:piUseCaseModel}).
The mod\-el defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}.
Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.

%In the case of \FlyingPig, as expressed above the model constraints:
%{\color{red} Anything else here? --M}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneralCut.png}
\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
\end{figure*}

\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~

The $\pi$-ServiceProcess model presents the workflow for \FlyingPig (Figure~\ref{fig:PiServiceProcessModel}).
Actions in this model were obtained by applying the $\pi$-use case transformation rules.
% described in Section~\ref{sec:pewsmetamodel}.

The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
Use cases are transformed into \textit{actions} and are represented by white boxes.
The restrictions associated to  $\pi$-use cases are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions.
We can see that this model refines the concepts defined in the $\pi$-UseCase model.
The assertions specify those non-functional requirements, as they are seen by the actors.
The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
\caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
\end{figure*}

\paragraph{\underline{$\pi$-ServiceComposition Model for \FlyingPig}}~

The model presented in Figure~\ref{fig:PiServiceCompositionModel}
%offers a system-centric view of the application.
%The previous model
is now  completed with the services that provide functions of the application  \FlyingPig\.
The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
These policies express the pre- and post-conditions of the previous model.
They are associated to the actions of the system.

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
%\subsection{Platform-Specific Model (PSM)}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\paragraph{\underline{$\pi$-PEWS Model for \FlyingPig}}~

The  PSM model of our case study is obtained by transforming the $\pi$-service composition model into a workflow.
Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
The model is given in Figure~\ref{fig:piPEWSFlyingPig}.

\begin{figure*}
\begin{verbatim}
//Namespaces specify service URI
namespace orca = www.orca.mx/service.wsdl

//Operations
alias createCompanySpace = portType/createCompanySpace in orca
alias generateQuestionnaire = portType/generateQuestionnaire in orca
alias analizeAnswer = portType/analizeAnswer in orca
alias storeData = portType/storeData in orca
alias createReport = portType/createReport in orca
alias callForAmendments = portType/callForAmendments in orca

//Services
service receiveRequest(R, Id) = createCompanySpace(R, Id)
service generateNewInterface(Id, NULL) = ...
service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
service notifyResponsable((Id, Q); NULL) = ...
service receiveAnswers((Id, T); P) =
	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL)
	. ((createReport(Id, P) . return(P)) 
	    + (callForAmendments(Id,T) . return(NULL)))

//Workflow
receiveRequest(R, Id)
||
(generateNewInterface(Id, NULL)
 . createQuestionnaire(Id, Q) . notifyResponsable((Id, Q); NULL))
||
 (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
\end{verbatim}
\caption{$\pi$-PEWS Model for \FlyingPig}\label{fig:piPEWSFlyingPig}
\end{figure*}

\subsection{Lessons Learned}

Through the example we underlined that every application implements functional aspects that describe its application logic.
Recall that an application logic refers to routines that perform the activities to reach the application objective.
Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.

The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
API does not necessarily export information about methods dependency (e.g., in the REST protocol);
they do not share their state (stateless).

Given a set of services with their exported methods, building service-based applications may consist on expressing an application logic as a service composition.
During this task, we must ensure the compliance between the specification and the resulting application.
Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.

As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way. Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse. In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.

\section{Conclusions and future work}\label{sec:conclusions}
We presented \pisodm, a model-driven method for designing and developing reliable service-based applications.
\pisodm extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
These requirements are taken into account from the early stages of the software development process.
Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.

Our method includes two CIM-level models, three PIM-level models and one PSM-level model.
We implemented the meta-models on the Eclipse platform and we validated the approach by using an industrially inspired use case.

Our case study demonstrates the applicability of \pisodm.
This case study was developed together with our industrial partner, GCP Global.
The Company is using \pisodm for the development of their product.
The case study presented here is a simplified version of their application.

\bibliographystyle{plain}
\bibliography{biblio}

\end{document}

### 2015-04-22 JournalToSubmit2015/submitted/main.tex
version to submit (good one)
version to submit (chinese journal)

@@
-\documentclass{ws-ijseke}
+\documentclass[12pt,twoside]{article}
+\usepackage{Jcst}
+%Equations,theorems,lemmas,etc. are created using the traditional enviroment
+%
+
 \usepackage{xspace}
+
+\usepackage{amsmath}
+\usepackage[thmmarks,amsmath]{ntheorem}
+
+\newcommand{\openbox}{\leavevmode
+  \hbox to.77778em{%
+  \hfil\vrule
+  \vbox to.675em{\hrule width.6em\vfil\hrule}%
+  \vrule\hfil}}
+
+\theoremstyle{plain}
+\theoremheaderfont{\normalfont\bfseries}
+\theorembodyfont{\normalfont}
+\theoremseparator{}
+\theoremindent0cm
+\theoremnumbering{arabic}
+\newtheorem{algo}{Algorithm}
+
+\theoremstyle{plain}
+%\theoremheaderfont{\normalfont\itshape}
+\theoremheaderfont{\normalfont\bfseries}
+\theorembodyfont{\normalfont}
+\theoremseparator{}
+\theoremindent0cm
+\theoremnumbering{arabic}
+\theoremsymbol{\ensuremath{\openbox}} 
+\newtheorem{example}{Example}
 
@@
 
-\markboth{K.Belhajjame \textit{et al.}}
-{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
-
-%%%%%%%%%%%%%%%%%%%%% Publisher's Area please ignore %%%%%%%%%%%%%%%
-%
-\catchline{}{}{}{}{}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\setcounter{page}{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%Set Page Head%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\setpageinformation
+%{Head of the first page} {Running head of odd pages}
+{}
+{ }{ }{}{Apr.}{2015}
+{\pisodm: Building Service-Oriented Applications}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%===========================================================
+
+%\begin{CJK}{GBK}{song}
 
@@
 
-\author{Khalid Belhajjame}
-\address{Universit\'e de Paris - Dauphine -- Paris, France\\
-\email{kbelhajj@googlemail.com}
+\author{Khalid Belhajjame$^1$, Valeria~de~Castro$^2$, Umberto~Souza~da~Costa$^3$, Javier~A.~Espinosa-Oviedo$^4$, Martin~A.~Musicante$^{3,*}$, Pl\'acido~A.~Souza~Neto$^5$, Genoveva~Vargas-Solar$^6$, \\ Jos\'e-Luis~Zechinelli-Martini$^4$}
+
+\footnotetext{$^*$.Corresponding author.}
+\footnotetext{$^1$.Universit\'e de Paris - Dauphine -- Paris, France.
+\textsl{kbelhajj@googlemail.com}.
 }
-
-\author{Valeria~de~Castro} 
-\address{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-\email{valeria.decastro@urjc.es}
+\footnotetext{$^2$.Universidad Rey Juan Carlos -- M\'{o}stoles, Spain.
+\textsl{valeria.decastro@urjc.es}.
 }
-
-\author{Umberto~Souza~da~Costa} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{umberto@dimap.ufrn.br}
+\footnotetext{$^3$.Universidade Federal do Rio Grande do Norte -- Natal, Brazil.
+\textsl{\{umberto,mam\}@dimap.ufrn.br}.
 }
-
-\author{Javier~A.~Espinosa-Oviedo} 
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{javiera.espinosa@gmail.com}
+\footnotetext{$^4$.Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico.
+\textsl{javiera.espinosa@gmail.com}, \textsl{joseluis.zechinelli@udlap.mx}.
 }
-
-\author{Martin~A.~Musicante} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{mam@dimap.ufrn.br}
+\footnotetext{$^5$.Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil.
+\textsl{placido.neto@ifrn.edu.br}.
 }
-
-\author{Pl\'acido~A.~Souza~Neto} 
-\address{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
-\email{placido.neto@ifrn.edu.br}
+\footnotetext{$^6$.CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France.
+\textsl{genoveva.vargas@imag.fr}.
 }
 
-\author{Genoveva~Vargas-Solar} 
-\address{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
-\email{genoveva.vargas@imag.fr}
-}
-
-\author{Jos\'e-Luis~Zechinelli-Martini}
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{joseluis.zechinelli@udlap.mx}
-}
-
-\maketitle
-
-\begin{history}
-\received{(Day Month Year)}
-\revised{(Day Month Year)}
-\accepted{(Day Month Year)}
-%\comby{(xxxxxxxxxx)}
-\end{history}
+%\maketitle
 
@@
 
-\keywords{MDA \and Non-Functional Requirements \and Service-based software process}
+\keywords{MDA , Non-Functional Requirements, Service-based software process}
+
+\begin{multicols}{2}
+\normalsize
 
@@
 combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
+The selection of services is usually guided by the \textit{functional}$^7$\footnote{$^7$.Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
 An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
@@
 These models are successively refined from abstract specifications into actual computer programs. 
-A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
+A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}$^8$\footnote{$^8$.The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
 Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.
@@
 
-The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
+The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System$^9$\footnote{$^9$.The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
 In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
+\includegraphics[width=1.0\textwidth]{piSOD-M_process.pdf}
 \caption{\pisodm Overview.}
@@
 The e$^3$value model of an application identifies the value/information exchange between components of the system.
-It is a business model that represents a business case graphically as a set of value exchanges\footnote{See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
+It is a business model that represents a business case graphically as a set of value exchanges$^{10}$\footnote{$^{10}$.See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
 The model is well suited to enhance the understanding of the environment in which an application is being developed.
@@
 \center
-\includegraphics[width=0.75\textwidth]{figs/e3value.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
+\includegraphics[width=0.75\textwidth]{e3value.pdf}
+\hspace*{5cm}\includegraphics[width=0.4\textwidth]{3ValueKey.pdf}
 \caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
@@
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
-Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
+Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model$^{11}$\footnote{$^{11}$.Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
 It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
@@
 \center
-\includegraphics[width=0.45\textwidth]{figs/SC.pdf}
+\includegraphics[width=0.45\textwidth]{SC.pdf}
 \caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/UseCaseMetaModel.pdf}
+\includegraphics[width=0.9\textwidth]{UseCaseMetaModel.pdf}
 \caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/UseCase.pdf}
+\includegraphics[width=0.9\textwidth]{UseCase.pdf}
 \caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcessMetaModel.pdf}
+\includegraphics[width=0.9\textwidth]{ServiceProcessMetaModel.pdf}
 \caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcess.pdf}
+\includegraphics[width=0.9\textwidth]{ServiceProcess.pdf}
 \caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
@@
 In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
-provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
+provides meta-classes to represent workflows$^{12}$\footnote{$^{12}$.Workflows are transformed into implemented service compositions.} that model  business processes.
 The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
@@
 A collaborator can be either internal or external to the system.
-When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
+When the collaborator of the business is external to the system, the attribute {\sf IsExternal}$^{13}$\footnote{$^{13}$.We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
 
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition}
+\includegraphics[width=1.0\textwidth]{PiServiceComposition}
 \caption{$\pi$-Service Composition Meta-model.}
@@
 \centering
-\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
+\includegraphics[width=0.95\textwidth]{piServiceComposition-toPublishMusic}
 %{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
+\includegraphics[width=0.75\textwidth]{PEWSMetamodel}
 \caption{$\pi$-{\sc Pews} Meta-model.}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figs/pi-pewsSpecification-toPublishMusic}
+\includegraphics[width=0.8\textwidth]{pi-pewsSpecification-toPublishMusic}
 \caption{$\pi$-PEWS Specific and Policy Representation.}
@@
 %\centering{
-%\includegraphics[width=0.96\textwidth]{figs/35}}
+%\includegraphics[width=0.96\textwidth]{35}}
 %\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
@@
 %\centering{
-%\includegraphics[width=0.80\textwidth]{figs/37}}
+%\includegraphics[width=0.80\textwidth]{37}}
 %\caption{ Extended Transformation Examples}
@@
 %\centering{
-%\includegraphics[width=0.96\textwidth]{figs/36}}
+%\includegraphics[width=0.96\textwidth]{36}}
 %\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
@@
 
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} \\
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} 
 The transformation rules have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).
@@
 %\centering{
-%\includegraphics[width=0.76\textwidth]{figs/38}}
+%\includegraphics[width=0.76\textwidth]{38}}
 %\caption{ Include Transformation Example}
@@
 %\centering{
-%\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
+%\includegraphics[width=0.96\textwidth]{ExceptionalRules}}
 %\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
@@
 %\centering{
-%\includegraphics[width=1\textwidth]{figs/ServiceProcess-ServiceComposition}}
+%\includegraphics[width=1\textwidth]{ServiceProcess-ServiceComposition}}
 %\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
@@
 
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}\\
+\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
 Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
@@
 %\centering{
-%\includegraphics[width=0.96\textwidth]{figs/Table7}}
+%\includegraphics[width=0.96\textwidth]{Table7}}
 %\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
@@
 \begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
+\item 	We  used the Eclipse Modeling Framework (EMF)$^{14}$\footnote{$^{14}$.The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
 From these meta-models, we  developed plug-ins to support their graphical representation.
 
-\item	 We used  ATL\footnote{http://eclipse.org/atl/.
+\item	 We used  ATL$^{15}$\footnote{$^{15}$.http://eclipse.org/atl/.
 An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
@@
 
-\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
+\item 	We  used Acceleo$^{16}$\footnote{$^{16}$.http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
 It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
@@
 
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
+To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System$^{17}$\footnote{$^{17}$.The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
 Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs/3ValueModel.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
+\includegraphics[width=0.9\textwidth]{3ValueModel.pdf}
+\hspace*{5cm}\includegraphics[width=0.4\textwidth]{3ValueKey.pdf}
 \caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
+\includegraphics[width=1.0\textwidth]{BPMN_GCP.pdf}
 \caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
@@
 The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
+Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated$^{18}$\footnote{$^{18}$.Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
 They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
+\includegraphics[width=0.9\textwidth]{UseCaseGeneral.png}
 \caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneralCut.png}
+\includegraphics[width=1.0\textwidth]{ServiceProcessGeneralCut.png}
 \caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
+\includegraphics[width=1.0\textwidth]{ServiceCompositionGeneralCut}
 \caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
@@
 
+%\begin{thebibliography}{00}
+%
+%\bibitem{1}
+%
+%\bibitem{2}
+%\end{thebibliography}
+
+\end{multicols}
+%\end{CJK}
 \end{document}

### 2015-04-23 JournalToSubmit2015/submitted/main.tex
submitted to the chinese journal

@@
 
-\author{Khalid Belhajjame$^1$, Valeria~de~Castro$^2$, Umberto~Souza~da~Costa$^3$, Javier~A.~Espinosa-Oviedo$^4$, Martin~A.~Musicante$^{3,*}$, Pl\'acido~A.~Souza~Neto$^5$, Genoveva~Vargas-Solar$^6$, \\ Jos\'e-Luis~Zechinelli-Martini$^4$}
-
-\footnotetext{$^*$.Corresponding author.}
-\footnotetext{$^1$.Universit\'e de Paris - Dauphine -- Paris, France.
-\textsl{kbelhajj@googlemail.com}.
-}
-\footnotetext{$^2$.Universidad Rey Juan Carlos -- M\'{o}stoles, Spain.
-\textsl{valeria.decastro@urjc.es}.
-}
-\footnotetext{$^3$.Universidade Federal do Rio Grande do Norte -- Natal, Brazil.
-\textsl{\{umberto,mam\}@dimap.ufrn.br}.
-}
-\footnotetext{$^4$.Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico.
-\textsl{javiera.espinosa@gmail.com}, \textsl{joseluis.zechinelli@udlap.mx}.
-}
-\footnotetext{$^5$.Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil.
-\textsl{placido.neto@ifrn.edu.br}.
-}
-\footnotetext{$^6$.CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France.
-\textsl{genoveva.vargas@imag.fr}.
-}
+%\author{Khalid Belhajjame$^1$, Valeria~de~Castro$^2$, Umberto~Souza~da~Costa$^3$, Javier~A.~Espinosa-Oviedo$^4$, Martin~A.~Musicante$^{3,*}$, Pl\'acido~A.~Souza~Neto$^5$, Genoveva~Vargas-Solar$^6$, \\ Jos\'e-Luis~Zechinelli-Martini$^4$}
+
+%\footnotetext{$^*$.Corresponding author.}
+%\footnotetext{$^1$.Universit\'e de Paris - Dauphine -- Paris, France.
+%\textsl{kbelhajj@googlemail.com}.
+%}
+%\footnotetext{$^2$.Universidad Rey Juan Carlos -- M\'{o}stoles, Spain.
+%\textsl{valeria.decastro@urjc.es}.
+%}
+%\footnotetext{$^3$.Universidade Federal do Rio Grande do Norte -- Natal, Brazil.
+%\textsl{\{umberto,mam\}@dimap.ufrn.br}.
+%}
+%\footnotetext{$^4$.Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico.
+%\textsl{javiera.espinosa@gmail.com}, \textsl{joseluis.zechinelli@udlap.mx}.
+%}
+%\footnotetext{$^5$.Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil.
+%\textsl{placido.neto@ifrn.edu.br}.
+%}
+%\footnotetext{$^6$.CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France.
+%\textsl{genoveva.vargas@imag.fr}.
+%}
 

### 2015-04-23 JournalToSubmit2015/submitted/titlePage.tex
submitted to the chinese journal

--- 2015-04-22 JournalToSubmit2015/submitted/main.tex
@@
 
-%The adoption of non-functional specifications from the early states of development can help the developer to produce applications that  can deal with the application context.
-
-The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
-Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
-Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
-These models are successively refined from abstract specifications into actual computer programs. 
-A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}$^8$\footnote{$^8$.The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
-Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.
-
-\bigskip
-Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
-Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
-\pisodm proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
-Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller} approach.
-The goals of  \pisodm are to:
-%SOD-M does not provide support for the specification of non-functional requirements, such as
-%security, reliability, and efficiency.
-\begin{trivlist}
-\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
-\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
-\item \textit{(iii)} Provide a general way of describing non-functional requirements.
-\end{trivlist}
-
-The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System$^9$\footnote{$^9$.The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
-The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.
-
-\bigskip
-This paper is organized as follows.
-Sec\-tion \ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
-Sections~\ref{sec:motivation} and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules.
-% Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology.
-Section~\ref{sec:flyingPig} describes the \FlyingPig case study that we develop for validating our method and discusses lessons learned.
-Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.
-
-\section{Related Work}
-\label{sec:relworks}
-While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
-NFRs include aspects such as performance, authentication and quality constraints.
-These requirements are usually specified by conditions, called non-functional properties.
-Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
-
-Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
-In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.
-
-In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:
-\textit{service level}, \textit{system level} and \textit{business level}.
-%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
-%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
-%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}.
-The work also defines a \textit{web service quality model}, which considers non-functional properties.
-In~\cite{XiaoCZBOLH08} the authors use the terms
-\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
-The notion of non-functional attribute is used to describe NFRs of the abstract process model.
-In the lower level, the composition is annotated with non-functional attributes.
-
-D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}.
-\textit{Quality dimensions} are used to quantify an individual characteristic.
-For instance, the quality category \textit{performance} groups characteristics such as
-\textit{latency} and \textit{throughput}.
-The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services.
-A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties such as \textit{availability}, \textit{reliability} and \textit{access control}.
-
-Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
-Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
-A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
-Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
-a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions.
-For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
-%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.
-
-Pastrana et al.~\cite{PastranaPK11} use a traditional design-by-contract approach~\cite{Meyer97}.
-The authors use the term \textit{contract} to describe non-functional re\-quire\-ments.
-Contracts may present pre-con\-di\-tions, post-conditions and invariants.
-Also, a con\-tract may define \textit{assertions} associated with \textit{qual\-i\-ty properties}.
-Each service may have as many associated \textit{contracts} as needed.
-
-Chollet et al.~\cite{CholletL09} associate (non-func\-tion\-al) \textit{quality properties} to
-(functional) activities.
-They present a security meta-model that takes into account web service compositions.
-In this work the non-functional requirements considered are  \textit{authentication}, \textit{integrity} and \textit{confidentiality}.
-Each NFR is associated with a service activity.
-
-Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
-Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services.
-Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}.
-Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
-Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09},
-\textit{property}~\cite{Fabra2011},
-\textit{factor}~\cite{MohantyRP10,GutierrezRF10},
-\textit{characteristic}~\cite{DiamadopoulouMPS08},
-\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
-\textit{value}~\cite{ThissenW06,BasinDL06}.
-
-Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
-However, distinct hierarchies and models are proposed for NFRs, according to different points of view.
-We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.
-
-Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}.
-As Ameller et al.
-realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
-Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
-Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.
-
-%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
-%The work  presents a development methodology, although this methodology is not centered on NFRs.
-The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account.
-In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs,
-as  in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
-system specification, including NFRs.
-In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be
-annotated with functional as well as non-functional attributes (NFAs).
-Functional and non-functional at\-tri\-butes are independently defined.
-They are attached to specific tasks later in the development of the application.
-NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.
-
-The proposal in~\cite{ThissenW06} presents steps for  selecting services
-by taking QoS information into account. The proposed steps are:
-\textit{(i)} identification of relevant QoS information;
-\textit{(ii)} identification of basic composition patterns and
-QoS aggregation rules for these patterns; and
-\textit{(iii)} definition of a selection mechanism of services.
-The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
-\textit{availability}.
-
-Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
-The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.
-
-Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
-In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.
-
-Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.
-For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
-Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.
-
-%\bigskip
-%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
-%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
-%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
-%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
-%It may use web services provided by third party sources.
-%The System level deals with the coordination of services, to implement the business logic.
-%Our proposal, shown in the next section, follows these ideas.
-
-\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}
-
-In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
-Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
-Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
-The refinement process is driven by transformation rules specified between the meta-models.
-
-In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
-The presentation is exemplified by a running example.
-The transformations between models are presented in section~\ref{sec:mmrules}.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{piSOD-M_process.pdf}
-\caption{\pisodm Overview.}
-\label{fig:piSOD-M}
-\end{figure*}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Computation Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the highest-level view of the system, including its business and requirement specifications.
-At this stage of the development, the structure and system processing details are still unknown or undetermined.
-\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.
-
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{E$^3$value meta-model}
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-The e$^3$value model of an application identifies the value/information exchange between components of the system.
-It is a business model that represents a business case graphically as a set of value exchanges$^{10}$\footnote{$^{10}$.See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
-The model is well suited to enhance the understanding of the environment in which an application is being developed.
-It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
-A dependency path has a direction and consists of a sequence of linked dependency nodes.
-It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}).
-Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
-
-\begin{figure*}
-\center
-\includegraphics[width=0.75\textwidth]{e3value.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{3ValueKey.pdf}
-\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
-\end{figure*}
-
-\begin{example}[To Publish Music]\label{ex:toPublicMusic}
-Let us consider the scenario ``To Publish Music'', used as a running example in this section:
-An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
-The following (external) services will be used by the application:
-\begin{itemize}
-\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
-\begin{itemize}
-%\item{\sf\small get-Status ( usedid ): String ;
-\item {\sf\small update-Status ( userid, new-status ): String};
-\end{itemize}
-\end{itemize}
-
-The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
-The e$^3$value value model shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status.
-We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
-\end{example}
-
-The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
-However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
-In order to face this shortcoming, our method proposes the BPMN meta-model as a tool for modelling this aspect of the application.
-
-\subsubsection{BPMN meta-model}
-
-BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow.
-The next example illustrates the use of this meta-model.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
-Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model$^{11}$\footnote{$^{11}$.Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
-It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\end{example}
-%
-\begin{figure}
-\center
-\includegraphics[width=0.45\textwidth]{SC.pdf}
-\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
-\end{figure}
-
-The CIM level models are the basis for the development of the application.
-The information presented at this level is (manually) refined into PIM-level models of the method \pisodm described next.
-Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the system functionality, hiding the details of any particular platform.
-The specification defines those parts of the system that do not change from one platform to another.
-Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.
-
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-UseCase} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
-The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
-The notion of \textit{policy} is used to describe NFRs.
-(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
-An {\sc End Consumer} is represented by an {\sc Actor}.
-An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
-{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
- The {\sc Business Collaborator} concept is represented  through {\sc Packages}.
-A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled.
-Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
-%Thus, the services functions can be specified as being grouped into packages.
-
- \begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{UseCaseMetaModel.pdf}
-\caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
-\end{figure*}
-
- The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
-A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
-Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint}) use case.
-There are three kinds of constraint:
-\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value'';
-\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
-\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
-The use of these constraints is shown in the following example.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
-Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example application.
-We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
-
-The ``To Publish Music'' Business Service expects  the Facebook or Twitter user status to be changed every time a user starts listening a new song.
-Therefore, it is necessary to perform a social network authentication with the users data.
-Each social network uses different services and different forms of authentication.
-The authentication constraint is required to update a music status.
-The restriction is stereotyped as a value constraint, because the users id and password are verified.
-Figure \ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases.
-The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as value and business constraint, respectively.
-For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
-Another restriction is that the minimum payment value is 2 euros.
-%{\color{red} Talk about this model for the example.}
-\end{example}
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{UseCase.pdf}
-\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
-\end{figure*}
-
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-The \textit{$\pi$-ServiceProcess} meta-model extends the UML activity diagram with the concept of \textit{contract} to represent constraints over data and actions. This concept is used to model
-groups  of  constraints  in the \textit{$\pi$-UseCase}.
-In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Ser\-vice Acti\-vi\-ty}, {\sc Action} and {\sc Constraint}.
-The dotted part of the figure defines those concepts related to the representation of NFRs.
-
-A specific \textit{$\pi$-ServiceProcess} model shows the set of logically related activities to be performed in a service-based application.
-So, the activities of this model represent a behavior that is part of the business logic of the application.
-A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
-A service activity represents an operation that is part of the execution flow, and it is modeled as an {\sc Action}.
-An activity contract represents the {\sc Non-functional Requirement}  that is also part of the execution flow of a service, identified  as a stereotyped activity ({\sf assertion}).
-The {\sc Assertion}s associated to an {\sc Action} compose a {\sc Contract}. 
-Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre-conditions and post-conditions.
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{ServiceProcessMetaModel.pdf}
-\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
-\end{figure*}
-
-%According to the meta-model, a \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts.
-It models the set of system activities and describe contracts for each activity or the process as a whole.
-Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action.
-This model defines a workflow for  modeling business logic and the behaviour of the application being developed.
-The behaviour is represented by each action and the restrictions by the assertion description.
-All assertions are described by stereotyped actions.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
-Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}.
-The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service.
-The buy music pre-conditions consist in verifying:
-(i) if the User data are correct;
-(ii) if the User is already logged in Spotify;
-(iii) if bank account information are correct and;
-(iv) if there are enough funds in the bank account to cover the payment.
-A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization.
-There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified.
-To update services, depending of each service, there may be different restrictions.
-As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter.
-In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook.
-As post-condition, the application ensures that the Facebook service sends a notification of success.
-To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modeled as post-condition).
-As a pre-condition for ``twitter update'' it is necessary that (i) the music format is correct and (ii) the twitter login and password are correct for the update.
-\end{example}
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{ServiceProcess.pdf}
-\caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
-\end{figure*}
-
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
-provides meta-classes to represent workflows$^{12}$\footnote{$^{12}$.Workflows are transformed into implemented service compositions.} that model  business processes.
-The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
-to group contracts with similar non-functional requirements.
-For instance, security and privacy restrictions may be grouped into a security policy.
- The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
-%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
-\begin{itemizedTrivlist}
-\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action.
-An instance of this meta-class is graphically represented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system.
-When the collaborator of the business is external to the system, the attribute {\sf IsExternal}$^{13}$\footnote{$^{13}$.We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
-
-\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class \textsc{Action}.
-A class action represents some type of transformation or processing.
-There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{PiServiceComposition}
-\caption{$\pi$-Service Composition Meta-model.}
-\label{fig:e-scomposition-metamodel}
-\end{figure*}
-
-\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more execu\-ta\-ble nodes.
-
-\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
-An {\em A-policy} is defined by attributes and rules.
-The conditions of each rule are evaluated.
-In case of failure, the actions of the rule will be performed.
-The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
-An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.
-\end{itemizedTrivlist}
-Instances of this meta-model are represented as UML activity diagrams.
-\begin{figure*}
-\centering
-\includegraphics[width=0.95\textwidth]{piServiceComposition-toPublishMusic}
-%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
-\caption{$\pi$-ServiceComposition Model for the ``To publish
-music'' business service.}
-\label{fig:servicecompositionmodel}
-\end{figure*}
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
-To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}).
- We model a business process with three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
-The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
-Both {\em Facebook} and {\em Twitter} require authentication.
-Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
-%{\color{red} To be completed!.}
-In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
-% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
-Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that read and update the user data.
-%A call to such services must be part of the authentication protocol required by these services.
-In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
-{\sf\small OAuthPolicy} implements the open authentication protocol.
-The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
-This variable is imported through the library {\sf\small OAuthPolicy.Token}.
-The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
-Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
-{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol.
-The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password.
-\end{example}
-
-%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
-%These artifacts will be used to generate the actual programs that will implement the application:
-Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation.
-The $\pi$-PEWS  meta-model is described in the next section.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform Specific Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the functionality, in the context of a particular implementation platform.
-Models at this level put together the platform-independent view with the specific aspects of the platform to implement the system.
-Models at the PSM level can be automatically translated into actual computer programs.
-We have defined one meta-model at this level.
-
-\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
-Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
-
-PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
-The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
-Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
-\begin{itemizedTrivlist}
-\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}.
-A {\sc Compound Operation} is defined using an {\sc Operator}.
-The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) repetition.
-
-\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-A brief description of these classes is given next.
-\end{itemizedTrivlist}
-%
-\begin{figure*}
-\centering
-\includegraphics[width=0.75\textwidth]{PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Meta-model.}
-\label{fig:PPEWSmetamodel}
-\end{figure*}
-
-Figure~\ref{fig:PPEWSmetamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
-Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.
-In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic5}
-Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our scenario example.
-Notice that the code contains namespaces and definitions (obtained by using information about the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
-\end{example}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.8\textwidth]{pi-pewsSpecification-toPublishMusic}
-\caption{$\pi$-PEWS Specific and Policy Representation.}
-\label{fig:Specific-Contract-Representation}
-\end{figure*}
-
-\section{Transformation rules}\label{sec:mmrules}
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-
-\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models.
-These rules are used to build a semi-automatic tool (the \pisodm environment plugin for Eclipse).
-The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-UseCase to the $\pi$-ServiceProcess}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-The refinement of a (composite) $\pi$-Use Case model into a $\pi$-Service Process model is driven by the principle of expressing a set of $\pi$-use cases (i.e., use cases with constraints)   in terms of  a business process.
-The resulting model consists of actions related by a control flow and contracts specifying NFPs.
-
-As defined by the $\pi$-ServiceProcess  meta-model (Figure~\ref{fig:CIM:serviceprocessmetamodel}) an {\sc Activity Service} consists of a composition of  entities of type {\sc Action}. 
-Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
-Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
-transformed into a  {\sf Fork node}~\cite{valeriaThesis,placidoPhDThesis2012}.
-If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
-Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{35}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
-
-A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
-The set of resulting assertions  are grouped into a {\sf Contract}.
-Constraints are transformed according to their type:
-{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} at\-tri\-bute set to false are transformed into {\sf Assertion}s;
-{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
-
-In order to transform constraints of type {\sf Value Constraint}, the designer specifies thresholds to be associated to the assertions of a contract.
-By default, value constraints are transformed into pre-conditions and business constraints are transformed into post-conditions.
-
-%A {\sf Requirement} and a {\sf Composite use case} in a  $\pi$-UseCase model are transformed respectively into a {\sf Service activity} and an {\sf Action} (see figures  \ref{fig:transformationsUseCase-ServiceProcessRules}-c  and d).
-%
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.80\textwidth]{37}}
-%\caption{ Extended Transformation Examples}
-%\label{fig:transformaton-examples}
-%\end{figure}
-
-%The transformations for {\sc Extend} and {\sc Include} dependencies  are not as simple as the previous transformations (figures \ref{fig:transformaton-examples}-a).
-
-The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  
-The generated workflow is composed by {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{36}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
-
-{\sf Include} use case entities are transformed into an {\sf Action} sequence.
-A {\sc Use case} element is transformed into an {\sf Action}. 
-A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements.
-
-Details of these transformations are not included here (due to space restrictions).
-They can be found in~\cite{SouzaNeto:2012}.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} 
-The transformation rules have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).
-
-The ``listen music'' use case is transformed into a Service Action. 
-This Action  represents a Spotify service function that can be invoked to play the music. 
-For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}.
-The use case ``download music''  includes the payment process to buy the music. 
-Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
-This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
-%(as shown in figure \ref{fig:transformation-example-include}). 
-The same rule is applied to the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
- \end{example}
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.76\textwidth]{38}}
-%\caption{ Include Transformation Example}
-%\label{fig:transformation-example-include}
-%\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-The  principle of the transformation  of a  $\pi$-Service Process model into a $\pi$-Service Composition model is to group {\sf Contracts} into {\sf A-Policies} and {\sf Actions} into {\sf Service Activities}.
-
-Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule}. 
-{\sf Rules}  concerning the same NFP  are grouped into {\sf A-Policies}. 
-Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute. 
-If the {\sf Assertion} has a value type, the name and the attributes are transformed into {\sf Variables} in the target model.  
-The {\sf Assertion: aProperty} attribute can have different transformations, according to: 
-\textit{(i)} a {\sf Precondition} is transformed into {\sf Pre};
-\textit{(ii)} a {\sf Post-Condition} is transformed into a {\sf  Post};
-\textit{(iii)} a {\sf TimeRestriction} is transformed into {\sf Time}.
-
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{ExceptionalRules}}
-%\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
-%\label{fig:ServiceProcess-ServiceComposition-Rules}
-%\end{figure}
-
-Some elements of the $\pi$-service composition model are obtained from the $\pi$-UseCase model: 
-A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
-\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Functional requirements} of the $\pi$-ServiceComposition model. 
-These requirements are associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).
-
-{\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their hom\-onym concepts of the $\pi$-Service Composition model.
-
-Finally, {\sf Actions} are grouped into  a {\sf Business collaborator}.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=1\textwidth]{ServiceProcess-ServiceComposition}}
-%\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
-%\label{fig:ServiceProcess-ServiceComposition}
-%\end{figure}
-
-%As the $\pi$-ServiceComposition model refines the $\pi$-ServiceProcess concepts at PIM level, a service previously defined as actions (source model) is refined as composition of those actions (target model) that are necessary to represent a business service, identifying who are the partners involved in the realization ({\sc Business collaborators}).
-%In addition, $\pi$-SOD-M defines a platform specific model based on web services composition.
-This model is explicitly indicates those actions which are (or will be, if not yet implemented) supported by web services.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
-Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
-The ``securityLoginPolicy'' consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
-The information about Facebook and Spotify (both of them {\sf Business
-Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
-model.
-\end{example}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-ServiceComposition to $\pi$-PEWS}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-This section describes the PIM to PSM transformations from a $\pi$-Service composition mod\-el to a $\pi$-PEWS model. 
-We distinguish two groups of rules: \textit{(i)} those transforming service composition entities into workflows; and \textit{(ii)} those transforming  A-Policies into Contracts.
-
-Single actions ({\sf Action} and {\sf
-Action:name}) are transformed into individual service {\sf Operations}.
-Complex actions (rep\-re\-sented by {\sf ServiceActivity}  and  {\sf
-ServiceActivity:name}) are transformed into (named) composite
-operations that defines a (named) workflow of the application.
-Composition patterns expressed using  {\em merge, decision, fork} and {\em join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{Table7}}
-%\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
-%\label{fig:ServiceComposition-Pews-Rules}
-%\end{figure}
-
-A-Policies defined for the entities of a $\pi$-service composition model generate {\sf A-policy} entities, named according to the names expres\-sed in the source model.
-The transformation of the rules expres\-sed in a $\pi$-service composition is guided by the event types associated to these rules. 
-The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into  {\sf Variable}, with  {\sf Name} and {\sf Type} attributes specified from the elements {\sf Variable: name} and {\sf Variable:type} of a $\pi$-service composition model.
-
-Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Pre-conditions}, {\sf Post-condition} and {\sf TimeRestrictions}.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT6}
-Figure \ref{fig:piSOD-M} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.
-\end{example}
-
-\subsection{Implementation}
-We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
-Our tool is implemented as a series of Eclipse plug-ins: 
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF)$^{14}$\footnote{$^{14}$.The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
-From these meta-models, we  developed plug-ins to support their graphical representation.
-
-\item	 We used  ATL$^{15}$\footnote{$^{15}$.http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for implementing the  mappings between models.
-
-\item 	We  used Acceleo$^{16}$\footnote{$^{16}$.http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
-It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
-A-Policy} based service composition execution environment.
-\end{itemize}
-
-\section{Applying \pisodm: The \FlyingPig\ case study}
-\label{sec:flyingPig}
-
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System$^{17}$\footnote{$^{17}$.The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
-Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
-The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
-The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
-By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
-
-Our goal is to model a service based application (\FlyingPig), for providing risk assessment as a service.
-In order to provide this functionality, \FlyingPig\ benefits from ORCA's legacy services providing storage, assessment and data visualization functions.
-
-In the following sections, we describe the results of applying \pisodm to develop the \FlyingPig\ risk assessment system.
-The models presented next were generated as a result of interacting with software developers at GCP Global.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{3ValueModel.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{3ValueKey.pdf}
-\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
-\end{figure*}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Computation-Independent Models (CIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
-It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
-%\footnote{\color{red} This sentence need to move to section 3.1!!}
-We identify two business actors: \textsl{ORCA} and \textsl{Broker}.
-Brokers emit requests for risk assessment for one or more companies.
-ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}.
-The values exchanged between ORCA and the brokers are:
-(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
-(ii) \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires;
-(iii) \textsl{Evaluation Reports} for the client companies;
-and
-(iv) the risk assessment \textsl{fee}.
-
-The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company.
-Once this need has been declared, the value exchanges between ORCA and Broker are triggered.
-The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
-In the next step, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{BPMN_GCP.pdf}
-\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
-\end{figure*}
-
-Figure~\ref{fig:BPMNmodel} shows the BPMN model
-%\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.}
-for the \FlyingPig\ scenario. The BPMN model for \FlyingPig\ is partitioned for better understanding the process in which the value exchanges occur.
-The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
-Brokers have two lanes, the client \textsl{Company} and a \textsl{User}.
-The user is a contact member of the company, who  coordinates the assessment process.
-This process  involves other members of the company as well.
-
-The risk assessment process starts after a request from a company.
-This corresponds to the value model, in which the start stimulus triggers the whole process.
-The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated$^{18}$\footnote{$^{18}$.Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
-
-Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
-During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
-
-Business processes also have rules and constraints to define their non-functional requirements (NFR):
-\begin{numtrivlist}
-\item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment.
-\item The system should be able to deal with, at least, 200 users.
-\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests.
-\item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
-\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
-\item All questionnaires need to be fully answered in order to consider that a task is completed.
-\item There is a time limit (in days) for amendments required by the system.
-\end{numtrivlist}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform-Independent Models (PIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-%In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
-The \pisodm models of the PIM level  for the \FlyingPig\ scenario are presented next.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{UseCaseGeneral.png}
-\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-UseCase Model for \FlyingPig}}~
-
-The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application.
-In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
-In the context of \FlyingPig, Company is the actor requesting a risk evaluation.
-A Broker is  responsible for coordinating  the evaluation process, assigning users to be in charge of tasks as well as delegating tasks.
-A User, in this model is an actor who answers questionnaires according to the current situation  of the Company.
-The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
-
-Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Fig.~\ref{fig:piUseCaseModel}).
-%Use cases
-that describe the main functionality of the system.
-The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.
-
-Each $\pi$-use case may be associated to one or more (non-functional) constraints (coloured ovals in Figure~\ref{fig:piUseCaseModel}).
-The mod\-el defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}.
-Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
-
-%In the case of \FlyingPig, as expressed above the model constraints:
-%{\color{red} Anything else here? --M}
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{ServiceProcessGeneralCut.png}
-\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
-
-The $\pi$-ServiceProcess model presents the workflow for \FlyingPig (Figure~\ref{fig:PiServiceProcessModel}).
-Actions in this model were obtained by applying the $\pi$-use case transformation rules.
-% described in Section~\ref{sec:pewsmetamodel}.
-
-The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
-Use cases are transformed into \textit{actions} and are represented by white boxes.
-The restrictions associated to  $\pi$-use cases are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions.
-We can see that this model refines the concepts defined in the $\pi$-UseCase model.
-The assertions specify those non-functional requirements, as they are seen by the actors.
-The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{ServiceCompositionGeneralCut}
-\caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-ServiceComposition Model for \FlyingPig}}~
-
-The model presented in Figure~\ref{fig:PiServiceCompositionModel}
-%offers a system-centric view of the application.
-%The previous model
-is now  completed with the services that provide functions of the application  \FlyingPig\.
-The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
-These policies express the pre- and post-conditions of the previous model.
-They are associated to the actions of the system.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-%\subsection{Platform-Specific Model (PSM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\paragraph{\underline{$\pi$-PEWS Model for \FlyingPig}}~
-
-The  PSM model of our case study is obtained by transforming the $\pi$-service composition model into a workflow.
-Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
-The model is given in Figure~\ref{fig:piPEWSFlyingPig}.
-
-\begin{figure*}
-\begin{verbatim}
-//Namespaces specify service URI
-namespace orca = www.orca.mx/service.wsdl
-
-//Operations
-alias createCompanySpace = portType/createCompanySpace in orca
-alias generateQuestionnaire = portType/generateQuestionnaire in orca
-alias analizeAnswer = portType/analizeAnswer in orca
-alias storeData = portType/storeData in orca
-alias createReport = portType/createReport in orca
-alias callForAmendments = portType/callForAmendments in orca
-
-//Services
-service receiveRequest(R, Id) = createCompanySpace(R, Id)
-service generateNewInterface(Id, NULL) = ...
-service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
-service notifyResponsable((Id, Q); NULL) = ...
-service receiveAnswers((Id, T); P) =
-	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL)
-	. ((createReport(Id, P) . return(P)) 
-	    + (callForAmendments(Id,T) . return(NULL)))
-
-//Workflow
-receiveRequest(R, Id)
-||
-(generateNewInterface(Id, NULL)
- . createQuestionnaire(Id, Q) . notifyResponsable((Id, Q); NULL))
-||
- (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
-\end{verbatim}
-\caption{$\pi$-PEWS Model for \FlyingPig}\label{fig:piPEWSFlyingPig}
-\end{figure*}
-
-\subsection{Lessons Learned}
-
-Through the example we underlined that every application implements functional aspects that describe its application logic.
-Recall that an application logic refers to routines that perform the activities to reach the application objective.
-Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
-These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
-
-The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
-API does not necessarily export information about methods dependency (e.g., in the REST protocol);
-they do not share their state (stateless).
-
-Given a set of services with their exported methods, building service-based applications may consist on expressing an application logic as a service composition.
-During this task, we must ensure the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way.
-Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-\section{Conclusions and future work}\label{sec:conclusions}
-We presented \pisodm, a model-driven method for designing and developing reliable service-based applications.
-\pisodm extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
-These requirements are taken into account from the early stages of the software development process.
-Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.
-
-Our method includes two CIM-level models, three PIM-level models and one PSM-level model.
-We implemented the meta-models on the Eclipse platform and we validated the approach by using an industrially inspired use case.
-
-Our case study demonstrates the applicability of \pisodm.
-This case study was developed together with our industrial partner, GCP Global.
-The Company is using \pisodm for the development of their product.
-The case study presented here is a simplified version of their application.
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-%\begin{thebibliography}{00}
-%
-%\bibitem{1}
-%
-%\bibitem{2}
-%\end{thebibliography}
-
 \end{multicols}

### 2015-07-23 RITA2015/OldVersion/conclusions.tex
New folder for RITA submission.

--- 2015-04-10 JournalToSubmit2015/submitted/main.tex
@@
-\documentclass{ws-ijseke}
-\usepackage{xspace}
-
-\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
-\def\FlyingPig{\textsl{FlyingPig}\xspace}
-
-\newcounter{numberInTrivlist}
-
-\newenvironment{numtrivlist}{\begin{list}{\rm \arabic{numberInTrivlist})}
-                                         {\usecounter{numberInTrivlist}
-                                          \setlength{\leftmargin}{0pt}
-                                          \setlength{\rightmargin}{0pt}
-                                          \setlength{\itemindent}{12pt}
-                                          \setlength{\listparindent}{0pt}}}
-                            {\end{list}}
-
-\newenvironment{itemizedTrivlist}{\begin{list}{\rm ~\hspace{2mm} $\bullet$\ }
-                                         {\setlength{\leftmargin}{0pt}
-                                          \setlength{\rightmargin}{0pt}
-                                          \setlength{\itemindent}{12pt}
-                                          \setlength{\listparindent}{0pt}}}
-                            {\end{list}}
-
-\begin{document}
-
-\markboth{K.Belhajjame \textit{et al.}}
-{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
-
-%%%%%%%%%%%%%%%%%%%%% Publisher's Area please ignore %%%%%%%%%%%%%%%
-%
-\catchline{}{}{}{}{}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
-
-\author{Khalid Belhajjame}
-\address{Universit\'e de Paris - Dauphine -- Paris, France\\
-\email{kbelhajj@googlemail.com}
-}
-
-\author{Valeria~de~Castro} 
-\address{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-\email{valeria.decastro@urjc.es}
-}
-
-\author{Umberto~Souza~da~Costa} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{umberto@dimap.ufrn.br}
-}
-
-\author{Javier~A.~Espinosa-Oviedo} 
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{javiera.espinosa@gmail.com}
-}
-
-\author{Martin~A.~Musicante} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{mam@dimap.ufrn.br}
-}
-
-\author{Pl\'acido~A.~Souza~Neto} 
-\address{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
-\email{placido.neto@ifrn.edu.br}
-}
-
-\author{Genoveva~Vargas-Solar} 
-\address{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
-\email{genoveva.vargas@imag.fr}
-}
-
-\author{Jos\'e-Luis~Zechinelli-Martini}
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{joseluis.zechinelli@udlap.mx}
-}
-
-\maketitle
-
-\begin{history}
-\received{(Day Month Year)}
-\revised{(Day Month Year)}
-\accepted{(Day Month Year)}
-%\comby{(xxxxxxxxxx)}
-\end{history}
-
-\begin{abstract}
-Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
-The late inclusion of NFRs in the development may compromise the quality of the deployed application.
-This paper presents \pisodm, a method and associated tools that
-\textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details;
-\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
-The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
-\end{abstract}
-
-\keywords{MDA \and Non-Functional Requirements \and Service-based software process}
-
-%% main text
-%*********************************************************************************************************
-\section{Introduction}
-\label{sec:intro}
-
-%Service oriented computing is at the origin of an evolution in the field of software development.
-%Service oriented methods advocates for the construction of software systems formed by the composition of heterogeneous, loosely coupled modules.
-%These modules (or services) communicate in order to achieve a common purpose.
-
-%(Moreover, IT systems need to evolve according to the business needs.)
-%Thus, organizations are seeking for mechanisms to bridge the gap between the actually developed systems and their business needs~\cite{bell}.
-
-In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
-combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
-An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
-
-Functional properties are not the only aspect to be considered in the software development process.
-Non-functional properties, such as data privacy, exception handling, atomicity  and, data persistence, need to be addressed to fit in the application.
-%Adding non-functional properties and respecting services constraints while composing services is a complex task that implies implementing protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
-%Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements.
-%
-%Furthermore,  often they are  not fully considered in the specification process of existing services' oriented development methods. Rather,
-%They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed.
-Ideally, Non-Functional Requirements (NFR) should be considered in every phase of the software development.
-Yet, they are partially or rarely methodologically derived from the specification, being usually added once the code has been im\-ple\-men\-ted.
-In the context of service oriented computing, the late consideration of non-functional requirements in the development process does not fully preserve the compliance and re\-use expectations promoted by the service oriented paradigm.
-
-%Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
-%Most methods integrate them only after the application has been implemented.
-%This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.
-
-% song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-%The adoption of non-functional specifications from the early states of development can help the developer to produce applications that  can deal with the application context.
-
-The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
-Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
-Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
-These models are successively refined from abstract specifications into actual computer programs. 
-A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
-Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.
-
-\bigskip
-Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
-Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
-\pisodm proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
-Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller} approach.
-The goals of  \pisodm are to:
-%SOD-M does not provide support for the specification of non-functional requirements, such as
-%security, reliability, and efficiency.
-\begin{trivlist}
-\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
-\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
-\item \textit{(iii)} Provide a general way of describing non-functional requirements.
-\end{trivlist}
-
-The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
-The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.
-
-\bigskip
-This paper is organized as follows.
-Sec\-tion \ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
-Sections~\ref{sec:motivation} and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules.
-% Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology.
-Section~\ref{sec:flyingPig} describes the \FlyingPig case study that we develop for validating our method and discusses lessons learned.
-Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.
-
-\section{Related Work}
-\label{sec:relworks}
-While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
-NFRs include aspects such as performance, authentication and quality constraints.
-These requirements are usually specified by conditions, called non-functional properties.
-Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
-
-Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
-In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.
-
-In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:
-\textit{service level}, \textit{system level} and \textit{business level}.
-%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
-%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
-%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}.
-The work also defines a \textit{web service quality model}, which considers non-functional properties.
-In~\cite{XiaoCZBOLH08} the authors use the terms
-\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
-The notion of non-functional attribute is used to describe NFRs of the abstract process model.
-In the lower level, the composition is annotated with non-functional attributes.
-
-D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}.
-\textit{Quality dimensions} are used to quantify an individual characteristic.
-For instance, the quality category \textit{performance} groups characteristics such as
-\textit{latency} and \textit{throughput}.
-The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services.
-A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties such as \textit{availability}, \textit{reliability} and \textit{access control}.
-
-Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
-Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
-A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
-Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
-a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions.
-For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
-%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.
-
-Pastrana et al.~\cite{PastranaPK11} use a traditional design-by-contract approach~\cite{Meyer97}.
-The authors use the term \textit{contract} to describe non-functional re\-quire\-ments.
-Contracts may present pre-con\-di\-tions, post-conditions and invariants.
-Also, a con\-tract may define \textit{assertions} associated with \textit{qual\-i\-ty properties}.
-Each service may have as many associated \textit{contracts} as needed.
-
-Chollet et al.~\cite{CholletL09} associate (non-func\-tion\-al) \textit{quality properties} to
-(functional) activities.
-They present a security meta-model that takes into account web service compositions.
-In this work the non-functional requirements considered are  \textit{authentication}, \textit{integrity} and \textit{confidentiality}.
-Each NFR is associated with a service activity.
-
-Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
-Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services.
-Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}.
-Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
-Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09},
-\textit{property}~\cite{Fabra2011},
-\textit{factor}~\cite{MohantyRP10,GutierrezRF10},
-\textit{characteristic}~\cite{DiamadopoulouMPS08},
-\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
-\textit{value}~\cite{ThissenW06,BasinDL06}.
-
-Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
-However, distinct hierarchies and models are proposed for NFRs, according to different points of view.
-We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.
-
-Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}.
-As Ameller et al.
-realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
-Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
-Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.
-
-%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
-%The work  presents a development methodology, although this methodology is not centered on NFRs.
-The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account.
-In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs,
-as  in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
-system specification, including NFRs.
-In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be
-annotated with functional as well as non-functional attributes (NFAs).
-Functional and non-functional at\-tri\-butes are independently defined.
-They are attached to specific tasks later in the development of the application.
-NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.
-
-The proposal in~\cite{ThissenW06} presents steps for  selecting services
-by taking QoS information into account. The proposed steps are:
-\textit{(i)} identification of relevant QoS information;
-\textit{(ii)} identification of basic composition patterns and
-QoS aggregation rules for these patterns; and
-\textit{(iii)} definition of a selection mechanism of services.
-The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
-\textit{availability}.
-
-Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
-The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.
-
-Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
-In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.
-
-Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.
-For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
-Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.
-
-%\bigskip
-%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
-%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
-%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
-%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
-%It may use web services provided by third party sources.
-%The System level deals with the coordination of services, to implement the business logic.
-%Our proposal, shown in the next section, follows these ideas.
-
-\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}
-
-In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
-Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
-Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
-The refinement process is driven by transformation rules specified between the meta-models.
-
-In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
-The presentation is exemplified by a running example.
-The transformations between models are presented in section~\ref{sec:mmrules}.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
-\caption{\pisodm Overview.}
-\label{fig:piSOD-M}
-\end{figure*}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Computation Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the highest-level view of the system, including its business and requirement specifications.
-At this stage of the development, the structure and system processing details are still unknown or undetermined.
-\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.
-
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{E$^3$value meta-model}
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-The e$^3$value model of an application identifies the value/information exchange between components of the system.
-It is a business model that represents a business case graphically as a set of value exchanges\footnote{See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
-The model is well suited to enhance the understanding of the environment in which an application is being developed.
-It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
-A dependency path has a direction and consists of a sequence of linked dependency nodes.
-It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}).
-Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
-
-\begin{figure*}
-\center
-\includegraphics[width=0.75\textwidth]{figs/e3value.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
-\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
-\end{figure*}
-
-\begin{example}[To Publish Music]\label{ex:toPublicMusic}
-Let us consider the scenario ``To Publish Music'', used as a running example in this section:
-An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
-The following (external) services will be used by the application:
-\begin{itemize}
-\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
-\begin{itemize}
-%\item{\sf\small get-Status ( usedid ): String ;
-\item {\sf\small update-Status ( userid, new-status ): String};
-\end{itemize}
-\end{itemize}
-
-The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
-The e$^3$value value model shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status.
-We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
-\end{example}
-
-The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
-However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
-In order to face this shortcoming, our method proposes the BPMN meta-model as a tool for modelling this aspect of the application.
-
-\subsubsection{BPMN meta-model}
-
-BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow.
-The next example illustrates the use of this meta-model.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
-Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
-It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\end{example}
-%
-\begin{figure}
-\center
-\includegraphics[width=0.45\textwidth]{figs/SC.pdf}
-\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
-\end{figure}
-
-The CIM level models are the basis for the development of the application.
-The information presented at this level is (manually) refined into PIM-level models of the method \pisodm described next.
-Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the system functionality, hiding the details of any particular platform.
-The specification defines those parts of the system that do not change from one platform to another.
-Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.
-
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-UseCase} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
-The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
-The notion of \textit{policy} is used to describe NFRs.
-(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
-An {\sc End Consumer} is represented by an {\sc Actor}.
-An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
-{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
- The {\sc Business Collaborator} concept is represented  through {\sc Packages}.
-A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled.
-Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
-%Thus, the services functions can be specified as being grouped into packages.
-
- \begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/UseCaseMetaModel.pdf}
-\caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
-\end{figure*}
-
- The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
-A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
-Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint}) use case.
-There are three kinds of constraint:
-\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value'';
-\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
-\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
-The use of these constraints is shown in the following example.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
-Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example application.
-We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
-
-The ``To Publish Music'' Business Service expects  the Facebook or Twitter user status to be changed every time a user starts listening a new song.
-Therefore, it is necessary to perform a social network authentication with the users data.
-Each social network uses different services and different forms of authentication.
-The authentication constraint is required to update a music status.
-The restriction is stereotyped as a value constraint, because the users id and password are verified.
-Figure \ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases.
-The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as value and business constraint, respectively.
-For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
-Another restriction is that the minimum payment value is 2 euros.
-%{\color{red} Talk about this model for the example.}
-\end{example}
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/UseCase.pdf}
-\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
-\end{figure*}
-
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-The \textit{$\pi$-ServiceProcess} meta-model extends the UML activity diagram with the concept of \textit{contract} to represent constraints over data and actions. This concept is used to model
-groups  of  constraints  in the \textit{$\pi$-UseCase}.
-In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Ser\-vice Acti\-vi\-ty}, {\sc Action} and {\sc Constraint}.
-The dotted part of the figure defines those concepts related to the representation of NFRs.
-
-A specific \textit{$\pi$-ServiceProcess} model shows the set of logically related activities to be performed in a service-based application.
-So, the activities of this model represent a behavior that is part of the business logic of the application.
-A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
-A service activity represents an operation that is part of the execution flow, and it is modeled as an {\sc Action}.
-An activity contract represents the {\sc Non-functional Requirement}  that is also part of the execution flow of a service, identified  as a stereotyped activity ({\sf assertion}).
-The {\sc Assertion}s associated to an {\sc Action} compose a {\sc Contract}. 
-Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre-conditions and post-conditions.
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcessMetaModel.pdf}
-\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
-\end{figure*}
-
-%According to the meta-model, a \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts.
-It models the set of system activities and describe contracts for each activity or the process as a whole.
-Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action.
-This model defines a workflow for  modeling business logic and the behaviour of the application being developed.
-The behaviour is represented by each action and the restrictions by the assertion description.
-All assertions are described by stereotyped actions.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
-Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}.
-The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service.
-The buy music pre-conditions consist in verifying:
-(i) if the User data are correct;
-(ii) if the User is already logged in Spotify;
-(iii) if bank account information are correct and;
-(iv) if there are enough funds in the bank account to cover the payment.
-A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization.
-There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified.
-To update services, depending of each service, there may be different restrictions.
-As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter.
-In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook.
-As post-condition, the application ensures that the Facebook service sends a notification of success.
-To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modeled as post-condition).
-As a pre-condition for ``twitter update'' it is necessary that (i) the music format is correct and (ii) the twitter login and password are correct for the update.
-\end{example}
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcess.pdf}
-\caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
-\end{figure*}
-
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
-provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
-The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
-to group contracts with similar non-functional requirements.
-For instance, security and privacy restrictions may be grouped into a security policy.
- The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
-%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
-\begin{itemizedTrivlist}
-\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action.
-An instance of this meta-class is graphically represented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system.
-When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
-
-\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class \textsc{Action}.
-A class action represents some type of transformation or processing.
-There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition}
-\caption{$\pi$-Service Composition Meta-model.}
-\label{fig:e-scomposition-metamodel}
-\end{figure*}
-
-\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more execu\-ta\-ble nodes.
-
-\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
-An {\em A-policy} is defined by attributes and rules.
-The conditions of each rule are evaluated.
-In case of failure, the actions of the rule will be performed.
-The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
-An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.
-\end{itemizedTrivlist}
-Instances of this meta-model are represented as UML activity diagrams.
-\begin{figure*}
-\centering
-\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
-%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
-\caption{$\pi$-ServiceComposition Model for the ``To publish
-music'' business service.}
-\label{fig:servicecompositionmodel}
-\end{figure*}
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
-To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}).
- We model a business process with three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
-The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
-Both {\em Facebook} and {\em Twitter} require authentication.
-Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
-%{\color{red} To be completed!.}
-In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
-% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
-Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that read and update the user data.
-%A call to such services must be part of the authentication protocol required by these services.
-In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
-{\sf\small OAuthPolicy} implements the open authentication protocol.
-The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
-This variable is imported through the library {\sf\small OAuthPolicy.Token}.
-The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
-Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
-{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol.
-The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password.
-\end{example}
-
-%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
-%These artifacts will be used to generate the actual programs that will implement the application:
-Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation.
-The $\pi$-PEWS  meta-model is described in the next section.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform Specific Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the functionality, in the context of a particular implementation platform.
-Models at this level put together the platform-independent view with the specific aspects of the platform to implement the system.
-Models at the PSM level can be automatically translated into actual computer programs.
-We have defined one meta-model at this level.
-
-\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
-Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
-
-PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
-The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
-Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
-\begin{itemizedTrivlist}
-\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}.
-A {\sc Compound Operation} is defined using an {\sc Operator}.
-The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) repetition.
-
-\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-A brief description of these classes is given next.
-\end{itemizedTrivlist}
-%
-\begin{figure*}
-\centering
-\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Meta-model.}
-\label{fig:PPEWSmetamodel}
-\end{figure*}
-
-Figure~\ref{fig:PPEWSmetamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
-Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.
-In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic5}
-Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our scenario example.
-Notice that the code contains namespaces and definitions (obtained by using information about the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
-\end{example}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.8\textwidth]{figs/pi-pewsSpecification-toPublishMusic}
-\caption{$\pi$-PEWS Specific and Policy Representation.}
-\label{fig:Specific-Contract-Representation}
-\end{figure*}
-
-\section{Transformation rules}\label{sec:mmrules}
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-
-\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models.
-These rules are used to build a semi-automatic tool (the \pisodm environment plugin for Eclipse).
-The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-UseCase to the $\pi$-ServiceProcess}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-The refinement of a (composite) $\pi$-Use Case model into a $\pi$-Service Process model is driven by the principle of expressing a set of $\pi$-use cases (i.e., use cases with constraints)   in terms of  a business process.
-The resulting model consists of actions related by a control flow and contracts specifying NFPs.
-
-As defined by the $\pi$-ServiceProcess  meta-model (Figure~\ref{fig:CIM:serviceprocessmetamodel}) an {\sc Activity Service} consists of a composition of  entities of type {\sc Action}. 
-Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
-Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
-transformed into a  {\sf Fork node}~\cite{valeriaThesis,placidoPhDThesis2012}.
-If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
-Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/35}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
-
-A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
-The set of resulting assertions  are grouped into a {\sf Contract}.
-Constraints are transformed according to their type:
-{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} at\-tri\-bute set to false are transformed into {\sf Assertion}s;
-{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
-
-In order to transform constraints of type {\sf Value Constraint}, the designer specifies thresholds to be associated to the assertions of a contract.
-By default, value constraints are transformed into pre-conditions and business constraints are transformed into post-conditions.
-
-%A {\sf Requirement} and a {\sf Composite use case} in a  $\pi$-UseCase model are transformed respectively into a {\sf Service activity} and an {\sf Action} (see figures  \ref{fig:transformationsUseCase-ServiceProcessRules}-c  and d).
-%
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.80\textwidth]{figs/37}}
-%\caption{ Extended Transformation Examples}
-%\label{fig:transformaton-examples}
-%\end{figure}
-
-%The transformations for {\sc Extend} and {\sc Include} dependencies  are not as simple as the previous transformations (figures \ref{fig:transformaton-examples}-a).
-
-The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  
-The generated workflow is composed by {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/36}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
-
-{\sf Include} use case entities are transformed into an {\sf Action} sequence.
-A {\sc Use case} element is transformed into an {\sf Action}. 
-A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements.
-
-Details of these transformations are not included here (due to space restrictions).
-They can be found in~\cite{SouzaNeto:2012}.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} \\
-The transformation rules have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).
-
-The ``listen music'' use case is transformed into a Service Action. 
-This Action  represents a Spotify service function that can be invoked to play the music. 
-For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}.
-The use case ``download music''  includes the payment process to buy the music. 
-Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
-This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
-%(as shown in figure \ref{fig:transformation-example-include}). 
-The same rule is applied to the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
- \end{example}
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.76\textwidth]{figs/38}}
-%\caption{ Include Transformation Example}
-%\label{fig:transformation-example-include}
-%\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-The  principle of the transformation  of a  $\pi$-Service Process model into a $\pi$-Service Composition model is to group {\sf Contracts} into {\sf A-Policies} and {\sf Actions} into {\sf Service Activities}.
-
-Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule}. 
-{\sf Rules}  concerning the same NFP  are grouped into {\sf A-Policies}. 
-Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute. 
-If the {\sf Assertion} has a value type, the name and the attributes are transformed into {\sf Variables} in the target model.  
-The {\sf Assertion: aProperty} attribute can have different transformations, according to: 
-\textit{(i)} a {\sf Precondition} is transformed into {\sf Pre};
-\textit{(ii)} a {\sf Post-Condition} is transformed into a {\sf  Post};
-\textit{(iii)} a {\sf TimeRestriction} is transformed into {\sf Time}.
-
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
-%\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
-%\label{fig:ServiceProcess-ServiceComposition-Rules}
-%\end{figure}
-
-Some elements of the $\pi$-service composition model are obtained from the $\pi$-UseCase model: 
-A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
-\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Functional requirements} of the $\pi$-ServiceComposition model. 
-These requirements are associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).
-
-{\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their hom\-onym concepts of the $\pi$-Service Composition model.
-
-Finally, {\sf Actions} are grouped into  a {\sf Business collaborator}.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=1\textwidth]{figs/ServiceProcess-ServiceComposition}}
-%\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
-%\label{fig:ServiceProcess-ServiceComposition}
-%\end{figure}
-
-%As the $\pi$-ServiceComposition model refines the $\pi$-ServiceProcess concepts at PIM level, a service previously defined as actions (source model) is refined as composition of those actions (target model) that are necessary to represent a business service, identifying who are the partners involved in the realization ({\sc Business collaborators}).
-%In addition, $\pi$-SOD-M defines a platform specific model based on web services composition.
-This model is explicitly indicates those actions which are (or will be, if not yet implemented) supported by web services.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}\\
-Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
-The ``securityLoginPolicy'' consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
-The information about Facebook and Spotify (both of them {\sf Business
-Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
-model.
-\end{example}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-ServiceComposition to $\pi$-PEWS}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-This section describes the PIM to PSM transformations from a $\pi$-Service composition mod\-el to a $\pi$-PEWS model. 
-We distinguish two groups of rules: \textit{(i)} those transforming service composition entities into workflows; and \textit{(ii)} those transforming  A-Policies into Contracts.
-
-Single actions ({\sf Action} and {\sf
-Action:name}) are transformed into individual service {\sf Operations}.
-Complex actions (rep\-re\-sented by {\sf ServiceActivity}  and  {\sf
-ServiceActivity:name}) are transformed into (named) composite
-operations that defines a (named) workflow of the application.
-Composition patterns expressed using  {\em merge, decision, fork} and {\em join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/Table7}}
-%\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
-%\label{fig:ServiceComposition-Pews-Rules}
-%\end{figure}
-
-A-Policies defined for the entities of a $\pi$-service composition model generate {\sf A-policy} entities, named according to the names expres\-sed in the source model.
-The transformation of the rules expres\-sed in a $\pi$-service composition is guided by the event types associated to these rules. 
-The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into  {\sf Variable}, with  {\sf Name} and {\sf Type} attributes specified from the elements {\sf Variable: name} and {\sf Variable:type} of a $\pi$-service composition model.
-
-Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Pre-conditions}, {\sf Post-condition} and {\sf TimeRestrictions}.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT6}
-Figure \ref{fig:piSOD-M} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.
-\end{example}
-
-\subsection{Implementation}
-We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
-Our tool is implemented as a series of Eclipse plug-ins: 
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
-From these meta-models, we  developed plug-ins to support their graphical representation.
-
-\item	 We used  ATL\footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for implementing the  mappings between models.
-
-\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
-It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
-A-Policy} based service composition execution environment.
-\end{itemize}
-
-\section{Applying \pisodm: The \FlyingPig\ case study}
-\label{sec:flyingPig}
-
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
-Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
-The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
-The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
-By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
-
-Our goal is to model a service based application (\FlyingPig), for providing risk assessment as a service.
-In order to provide this functionality, \FlyingPig\ benefits from ORCA's legacy services providing storage, assessment and data visualization functions.
-
-In the following sections, we describe the results of applying \pisodm to develop the \FlyingPig\ risk assessment system.
-The models presented next were generated as a result of interacting with software developers at GCP Global.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figs/3ValueModel.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
-\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
-\end{figure*}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Computation-Independent Models (CIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
-It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
-%\footnote{\color{red} This sentence need to move to section 3.1!!}
-We identify two business actors: \textsl{ORCA} and \textsl{Broker}.
-Brokers emit requests for risk assessment for one or more companies.
-ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}.
-The values exchanged between ORCA and the brokers are:
-(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
-(ii) \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires;
-(iii) \textsl{Evaluation Reports} for the client companies;
-and
-(iv) the risk assessment \textsl{fee}.
-
-The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company.
-Once this need has been declared, the value exchanges between ORCA and Broker are triggered.
-The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
-In the next step, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
-\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
-\end{figure*}
-
-Figure~\ref{fig:BPMNmodel} shows the BPMN model
-%\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.}
-for the \FlyingPig\ scenario. The BPMN model for \FlyingPig\ is partitioned for better understanding the process in which the value exchanges occur.
-The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
-Brokers have two lanes, the client \textsl{Company} and a \textsl{User}.
-The user is a contact member of the company, who  coordinates the assessment process.
-This process  involves other members of the company as well.
-
-The risk assessment process starts after a request from a company.
-This corresponds to the value model, in which the start stimulus triggers the whole process.
-The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
-
-Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
-During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
-
-Business processes also have rules and constraints to define their non-functional requirements (NFR):
-\begin{numtrivlist}
-\item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment.
-\item The system should be able to deal with, at least, 200 users.
-\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests.
-\item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
-\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
-\item All questionnaires need to be fully answered in order to consider that a task is completed.
-\item There is a time limit (in days) for amendments required by the system.
-\end{numtrivlist}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform-Independent Models (PIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-%In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
-The \pisodm models of the PIM level  for the \FlyingPig\ scenario are presented next.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
-\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-UseCase Model for \FlyingPig}}~
-
-The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application.
-In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
-In the context of \FlyingPig, Company is the actor requesting a risk evaluation.
-A Broker is  responsible for coordinating  the evaluation process, assigning users to be in charge of tasks as well as delegating tasks.
-A User, in this model is an actor who answers questionnaires according to the current situation  of the Company.
-The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
-
-Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Fig.~\ref{fig:piUseCaseModel}).
-%Use cases
-that describe the main functionality of the system.
-The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.
-
-Each $\pi$-use case may be associated to one or more (non-functional) constraints (coloured ovals in Figure~\ref{fig:piUseCaseModel}).
-The mod\-el defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}.
-Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
-
-%In the case of \FlyingPig, as expressed above the model constraints:
-%{\color{red} Anything else here? --M}
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneralCut.png}
-\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
-
-The $\pi$-ServiceProcess model presents the workflow for \FlyingPig (Figure~\ref{fig:PiServiceProcessModel}).
-Actions in this model were obtained by applying the $\pi$-use case transformation rules.
-% described in Section~\ref{sec:pewsmetamodel}.
-
-The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
-Use cases are transformed into \textit{actions} and are represented by white boxes.
-The restrictions associated to  $\pi$-use cases are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions.
-We can see that this model refines the concepts defined in the $\pi$-UseCase model.
-The assertions specify those non-functional requirements, as they are seen by the actors.
-The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
-\caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-ServiceComposition Model for \FlyingPig}}~
-
-The model presented in Figure~\ref{fig:PiServiceCompositionModel}
-%offers a system-centric view of the application.
-%The previous model
-is now  completed with the services that provide functions of the application  \FlyingPig\.
-The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
-These policies express the pre- and post-conditions of the previous model.
-They are associated to the actions of the system.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-%\subsection{Platform-Specific Model (PSM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\paragraph{\underline{$\pi$-PEWS Model for \FlyingPig}}~
-
-The  PSM model of our case study is obtained by transforming the $\pi$-service composition model into a workflow.
-Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
-The model is given in Figure~\ref{fig:piPEWSFlyingPig}.
-
-\begin{figure*}
-\begin{verbatim}
-//Namespaces specify service URI
-namespace orca = www.orca.mx/service.wsdl
-
-//Operations
-alias createCompanySpace = portType/createCompanySpace in orca
-alias generateQuestionnaire = portType/generateQuestionnaire in orca
-alias analizeAnswer = portType/analizeAnswer in orca
-alias storeData = portType/storeData in orca
-alias createReport = portType/createReport in orca
-alias callForAmendments = portType/callForAmendments in orca
-
-//Services
-service receiveRequest(R, Id) = createCompanySpace(R, Id)
-service generateNewInterface(Id, NULL) = ...
-service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
-service notifyResponsable((Id, Q); NULL) = ...
-service receiveAnswers((Id, T); P) =
-	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL)
-	. ((createReport(Id, P) . return(P)) 
-	    + (callForAmendments(Id,T) . return(NULL)))
-
-//Workflow
-receiveRequest(R, Id)
-||
-(generateNewInterface(Id, NULL)
- . createQuestionnaire(Id, Q) . notifyResponsable((Id, Q); NULL))
-||
- (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
-\end{verbatim}
-\caption{$\pi$-PEWS Model for \FlyingPig}\label{fig:piPEWSFlyingPig}
-\end{figure*}
-
-\subsection{Lessons Learned}
-
-Through the example we underlined that every application implements functional aspects that describe its application logic.
-Recall that an application logic refers to routines that perform the activities to reach the application objective.
-Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
-These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
-
-The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
-API does not necessarily export information about methods dependency (e.g., in the REST protocol);
-they do not share their state (stateless).
-
-Given a set of services with their exported methods, building service-based applications may consist on expressing an application logic as a service composition.
-During this task, we must ensure the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way.
-Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-\section{Conclusions and future work}\label{sec:conclusions}
 We presented \pisodm, a model-driven method for designing and developing reliable service-based applications.
@@
 The case study presented here is a simplified version of their application.
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-\end{document}

### 2015-07-23 RITA2015/OldVersion/intro.tex
New folder for RITA submission.

--- 2015-04-10 JournalToSubmit2015/submitted/main.tex
@@
-\documentclass{ws-ijseke}
-\usepackage{xspace}
-
-\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
-\def\FlyingPig{\textsl{FlyingPig}\xspace}
-
-\newcounter{numberInTrivlist}
-
-\newenvironment{numtrivlist}{\begin{list}{\rm \arabic{numberInTrivlist})}
-                                         {\usecounter{numberInTrivlist}
-                                          \setlength{\leftmargin}{0pt}
-                                          \setlength{\rightmargin}{0pt}
-                                          \setlength{\itemindent}{12pt}
-                                          \setlength{\listparindent}{0pt}}}
-                            {\end{list}}
-
-\newenvironment{itemizedTrivlist}{\begin{list}{\rm ~\hspace{2mm} $\bullet$\ }
-                                         {\setlength{\leftmargin}{0pt}
-                                          \setlength{\rightmargin}{0pt}
-                                          \setlength{\itemindent}{12pt}
-                                          \setlength{\listparindent}{0pt}}}
-                            {\end{list}}
-
-\begin{document}
-
-\markboth{K.Belhajjame \textit{et al.}}
-{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
-
-%%%%%%%%%%%%%%%%%%%%% Publisher's Area please ignore %%%%%%%%%%%%%%%
-%
-\catchline{}{}{}{}{}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
-
-\author{Khalid Belhajjame}
-\address{Universit\'e de Paris - Dauphine -- Paris, France\\
-\email{kbelhajj@googlemail.com}
-}
-
-\author{Valeria~de~Castro} 
-\address{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-\email{valeria.decastro@urjc.es}
-}
-
-\author{Umberto~Souza~da~Costa} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{umberto@dimap.ufrn.br}
-}
-
-\author{Javier~A.~Espinosa-Oviedo} 
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{javiera.espinosa@gmail.com}
-}
-
-\author{Martin~A.~Musicante} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{mam@dimap.ufrn.br}
-}
-
-\author{Pl\'acido~A.~Souza~Neto} 
-\address{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
-\email{placido.neto@ifrn.edu.br}
-}
-
-\author{Genoveva~Vargas-Solar} 
-\address{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
-\email{genoveva.vargas@imag.fr}
-}
-
-\author{Jos\'e-Luis~Zechinelli-Martini}
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{joseluis.zechinelli@udlap.mx}
-}
-
-\maketitle
-
-\begin{history}
-\received{(Day Month Year)}
-\revised{(Day Month Year)}
-\accepted{(Day Month Year)}
-%\comby{(xxxxxxxxxx)}
-\end{history}
-
-\begin{abstract}
-Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
-The late inclusion of NFRs in the development may compromise the quality of the deployed application.
-This paper presents \pisodm, a method and associated tools that
-\textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details;
-\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
-The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
-\end{abstract}
-
-\keywords{MDA \and Non-Functional Requirements \and Service-based software process}
-
-%% main text
-%*********************************************************************************************************
-\section{Introduction}
-\label{sec:intro}
-
 %Service oriented computing is at the origin of an evolution in the field of software development.
@@
 Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.
-
-\section{Related Work}
-\label{sec:relworks}
-While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
-NFRs include aspects such as performance, authentication and quality constraints.
-These requirements are usually specified by conditions, called non-functional properties.
-Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
-
-Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
-In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.
-
-In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:
-\textit{service level}, \textit{system level} and \textit{business level}.
-%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
-%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
-%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}.
-The work also defines a \textit{web service quality model}, which considers non-functional properties.
-In~\cite{XiaoCZBOLH08} the authors use the terms
-\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
-The notion of non-functional attribute is used to describe NFRs of the abstract process model.
-In the lower level, the composition is annotated with non-functional attributes.
-
-D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}.
-\textit{Quality dimensions} are used to quantify an individual characteristic.
-For instance, the quality category \textit{performance} groups characteristics such as
-\textit{latency} and \textit{throughput}.
-The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services.
-A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties such as \textit{availability}, \textit{reliability} and \textit{access control}.
-
-Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
-Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
-A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
-Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
-a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions.
-For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
-%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.
-
-Pastrana et al.~\cite{PastranaPK11} use a traditional design-by-contract approach~\cite{Meyer97}.
-The authors use the term \textit{contract} to describe non-functional re\-quire\-ments.
-Contracts may present pre-con\-di\-tions, post-conditions and invariants.
-Also, a con\-tract may define \textit{assertions} associated with \textit{qual\-i\-ty properties}.
-Each service may have as many associated \textit{contracts} as needed.
-
-Chollet et al.~\cite{CholletL09} associate (non-func\-tion\-al) \textit{quality properties} to
-(functional) activities.
-They present a security meta-model that takes into account web service compositions.
-In this work the non-functional requirements considered are  \textit{authentication}, \textit{integrity} and \textit{confidentiality}.
-Each NFR is associated with a service activity.
-
-Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
-Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services.
-Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}.
-Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
-Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09},
-\textit{property}~\cite{Fabra2011},
-\textit{factor}~\cite{MohantyRP10,GutierrezRF10},
-\textit{characteristic}~\cite{DiamadopoulouMPS08},
-\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
-\textit{value}~\cite{ThissenW06,BasinDL06}.
-
-Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
-However, distinct hierarchies and models are proposed for NFRs, according to different points of view.
-We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.
-
-Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}.
-As Ameller et al.
-realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
-Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
-Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.
-
-%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
-%The work  presents a development methodology, although this methodology is not centered on NFRs.
-The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account.
-In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs,
-as  in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
-system specification, including NFRs.
-In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be
-annotated with functional as well as non-functional attributes (NFAs).
-Functional and non-functional at\-tri\-butes are independently defined.
-They are attached to specific tasks later in the development of the application.
-NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.
-
-The proposal in~\cite{ThissenW06} presents steps for  selecting services
-by taking QoS information into account. The proposed steps are:
-\textit{(i)} identification of relevant QoS information;
-\textit{(ii)} identification of basic composition patterns and
-QoS aggregation rules for these patterns; and
-\textit{(iii)} definition of a selection mechanism of services.
-The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
-\textit{availability}.
-
-Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
-The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.
-
-Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
-In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.
-
-Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.
-For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
-Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.
-
-%\bigskip
-%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
-%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
-%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
-%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
-%It may use web services provided by third party sources.
-%The System level deals with the coordination of services, to implement the business logic.
-%Our proposal, shown in the next section, follows these ideas.
-
-\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}
-
-In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
-Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
-Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
-The refinement process is driven by transformation rules specified between the meta-models.
-
-In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
-The presentation is exemplified by a running example.
-The transformations between models are presented in section~\ref{sec:mmrules}.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
-\caption{\pisodm Overview.}
-\label{fig:piSOD-M}
-\end{figure*}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Computation Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the highest-level view of the system, including its business and requirement specifications.
-At this stage of the development, the structure and system processing details are still unknown or undetermined.
-\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.
-
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{E$^3$value meta-model}
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-The e$^3$value model of an application identifies the value/information exchange between components of the system.
-It is a business model that represents a business case graphically as a set of value exchanges\footnote{See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
-The model is well suited to enhance the understanding of the environment in which an application is being developed.
-It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
-A dependency path has a direction and consists of a sequence of linked dependency nodes.
-It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}).
-Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
-
-\begin{figure*}
-\center
-\includegraphics[width=0.75\textwidth]{figs/e3value.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
-\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
-\end{figure*}
-
-\begin{example}[To Publish Music]\label{ex:toPublicMusic}
-Let us consider the scenario ``To Publish Music'', used as a running example in this section:
-An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
-In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
-The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
-The following (external) services will be used by the application:
-\begin{itemize}
-\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
-\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
-\begin{itemize}
-%\item{\sf\small get-Status ( usedid ): String ;
-\item {\sf\small update-Status ( userid, new-status ): String};
-\end{itemize}
-\end{itemize}
-
-The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
-The e$^3$value value model shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
-Finally, the private application interacts with Facebook and Twitter for updating the user's status.
-We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
-\end{example}
-
-The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
-However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
-In order to face this shortcoming, our method proposes the BPMN meta-model as a tool for modelling this aspect of the application.
-
-\subsubsection{BPMN meta-model}
-
-BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow.
-The next example illustrates the use of this meta-model.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
-Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
-It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
-Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
-\end{example}
-%
-\begin{figure}
-\center
-\includegraphics[width=0.45\textwidth]{figs/SC.pdf}
-\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
-\end{figure}
-
-The CIM level models are the basis for the development of the application.
-The information presented at this level is (manually) refined into PIM-level models of the method \pisodm described next.
-Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the system functionality, hiding the details of any particular platform.
-The specification defines those parts of the system that do not change from one platform to another.
-Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.
-
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-UseCase} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
-The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
-The notion of \textit{policy} is used to describe NFRs.
-(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
-An {\sc End Consumer} is represented by an {\sc Actor}.
-An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
-{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
- The {\sc Business Collaborator} concept is represented  through {\sc Packages}.
-A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled.
-Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
-%Thus, the services functions can be specified as being grouped into packages.
-
- \begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/UseCaseMetaModel.pdf}
-\caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
-\end{figure*}
-
- The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
-A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
-Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint}) use case.
-There are three kinds of constraint:
-\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value'';
-\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
-\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
-The use of these constraints is shown in the following example.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
-Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example application.
-We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
-
-The ``To Publish Music'' Business Service expects  the Facebook or Twitter user status to be changed every time a user starts listening a new song.
-Therefore, it is necessary to perform a social network authentication with the users data.
-Each social network uses different services and different forms of authentication.
-The authentication constraint is required to update a music status.
-The restriction is stereotyped as a value constraint, because the users id and password are verified.
-Figure \ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases.
-The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as value and business constraint, respectively.
-For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
-Another restriction is that the minimum payment value is 2 euros.
-%{\color{red} Talk about this model for the example.}
-\end{example}
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/UseCase.pdf}
-\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
-\end{figure*}
-
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-The \textit{$\pi$-ServiceProcess} meta-model extends the UML activity diagram with the concept of \textit{contract} to represent constraints over data and actions. This concept is used to model
-groups  of  constraints  in the \textit{$\pi$-UseCase}.
-In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Ser\-vice Acti\-vi\-ty}, {\sc Action} and {\sc Constraint}.
-The dotted part of the figure defines those concepts related to the representation of NFRs.
-
-A specific \textit{$\pi$-ServiceProcess} model shows the set of logically related activities to be performed in a service-based application.
-So, the activities of this model represent a behavior that is part of the business logic of the application.
-A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
-A service activity represents an operation that is part of the execution flow, and it is modeled as an {\sc Action}.
-An activity contract represents the {\sc Non-functional Requirement}  that is also part of the execution flow of a service, identified  as a stereotyped activity ({\sf assertion}).
-The {\sc Assertion}s associated to an {\sc Action} compose a {\sc Contract}. 
-Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre-conditions and post-conditions.
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcessMetaModel.pdf}
-\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
-\end{figure*}
-
-%According to the meta-model, a \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts.
-It models the set of system activities and describe contracts for each activity or the process as a whole.
-Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action.
-This model defines a workflow for  modeling business logic and the behaviour of the application being developed.
-The behaviour is represented by each action and the restrictions by the assertion description.
-All assertions are described by stereotyped actions.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
-Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}.
-The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service.
-The buy music pre-conditions consist in verifying:
-(i) if the User data are correct;
-(ii) if the User is already logged in Spotify;
-(iii) if bank account information are correct and;
-(iv) if there are enough funds in the bank account to cover the payment.
-A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization.
-There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified.
-To update services, depending of each service, there may be different restrictions.
-As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter.
-In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook.
-As post-condition, the application ensures that the Facebook service sends a notification of success.
-To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modeled as post-condition).
-As a pre-condition for ``twitter update'' it is necessary that (i) the music format is correct and (ii) the twitter login and password are correct for the update.
-\end{example}
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcess.pdf}
-\caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
-\end{figure*}
-
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
-provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
-The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
-to group contracts with similar non-functional requirements.
-For instance, security and privacy restrictions may be grouped into a security policy.
- The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
-%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
-\begin{itemizedTrivlist}
-\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action.
-An instance of this meta-class is graphically represented as a partition in the activity diagram.
-A collaborator can be either internal or external to the system.
-When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
-
-\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class \textsc{Action}.
-A class action represents some type of transformation or processing.
-There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition}
-\caption{$\pi$-Service Composition Meta-model.}
-\label{fig:e-scomposition-metamodel}
-\end{figure*}
-
-\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more execu\-ta\-ble nodes.
-
-\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
-We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
-An {\em A-policy} is defined by attributes and rules.
-The conditions of each rule are evaluated.
-In case of failure, the actions of the rule will be performed.
-The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
-An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.
-\end{itemizedTrivlist}
-Instances of this meta-model are represented as UML activity diagrams.
-\begin{figure*}
-\centering
-\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
-%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
-\caption{$\pi$-ServiceComposition Model for the ``To publish
-music'' business service.}
-\label{fig:servicecompositionmodel}
-\end{figure*}
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
-To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}).
- We model a business process with three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
-The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
-Both {\em Facebook} and {\em Twitter} require authentication.
-Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
-%{\color{red} To be completed!.}
-In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
-% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
-The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
-Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that read and update the user data.
-%A call to such services must be part of the authentication protocol required by these services.
-In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
-In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
-{\sf\small OAuthPolicy} implements the open authentication protocol.
-The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
-This variable is imported through the library {\sf\small OAuthPolicy.Token}.
-The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
-Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
-{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol.
-The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
-The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
-On the notification of an event of that type, a credential is obtained using the username and password.
-\end{example}
-
-%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
-%These artifacts will be used to generate the actual programs that will implement the application:
-Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation.
-The $\pi$-PEWS  meta-model is described in the next section.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform Specific Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-This level focuses on the functionality, in the context of a particular implementation platform.
-Models at this level put together the platform-independent view with the specific aspects of the platform to implement the system.
-Models at the PSM level can be automatically translated into actual computer programs.
-We have defined one meta-model at this level.
-
-\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
-Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
-
-PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
-The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
-Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
-\begin{itemizedTrivlist}
-\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
-A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}.
-A {\sc Compound Operation} is defined using an {\sc Operator}.
-The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) repetition.
-
-\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-A brief description of these classes is given next.
-\end{itemizedTrivlist}
-%
-\begin{figure*}
-\centering
-\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Meta-model.}
-\label{fig:PPEWSmetamodel}
-\end{figure*}
-
-Figure~\ref{fig:PPEWSmetamodel} shows that each {\sc A-Policy} is associated to a {\sc Scope} that can be either an {\sc Operation} (e.g., an authentication protocol associated to a method exported by a service),  an {\sc Operator} (e.g., a temporal constraint associated to a sequence of operators) or a {\sc Path}.
-Each {\sc A-Policy} groups a set of ECA rules with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}.
-In this way, an {\em A-policy} represents a set of reactions to be possibly executed when one or several events are notified.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic5}
-Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our scenario example.
-Notice that the code contains namespaces and definitions (obtained by using information about the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
-\end{example}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.8\textwidth]{figs/pi-pewsSpecification-toPublishMusic}
-\caption{$\pi$-PEWS Specific and Policy Representation.}
-\label{fig:Specific-Contract-Representation}
-\end{figure*}
-
-\section{Transformation rules}\label{sec:mmrules}
-
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-%\subsubsection{$\pi$-{\sc Pews}  meta-model}\label{sec:pewsmetamodel}
-%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..
-
-\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models.
-These rules are used to build a semi-automatic tool (the \pisodm environment plugin for Eclipse).
-The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-UseCase to the $\pi$-ServiceProcess}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-The refinement of a (composite) $\pi$-Use Case model into a $\pi$-Service Process model is driven by the principle of expressing a set of $\pi$-use cases (i.e., use cases with constraints)   in terms of  a business process.
-The resulting model consists of actions related by a control flow and contracts specifying NFPs.
-
-As defined by the $\pi$-ServiceProcess  meta-model (Figure~\ref{fig:CIM:serviceprocessmetamodel}) an {\sc Activity Service} consists of a composition of  entities of type {\sc Action}. 
-Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
-Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
-transformed into a  {\sf Fork node}~\cite{valeriaThesis,placidoPhDThesis2012}.
-If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
-Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/35}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
-
-A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
-The set of resulting assertions  are grouped into a {\sf Contract}.
-Constraints are transformed according to their type:
-{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} at\-tri\-bute set to false are transformed into {\sf Assertion}s;
-{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
-
-In order to transform constraints of type {\sf Value Constraint}, the designer specifies thresholds to be associated to the assertions of a contract.
-By default, value constraints are transformed into pre-conditions and business constraints are transformed into post-conditions.
-
-%A {\sf Requirement} and a {\sf Composite use case} in a  $\pi$-UseCase model are transformed respectively into a {\sf Service activity} and an {\sf Action} (see figures  \ref{fig:transformationsUseCase-ServiceProcessRules}-c  and d).
-%
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.80\textwidth]{figs/37}}
-%\caption{ Extended Transformation Examples}
-%\label{fig:transformaton-examples}
-%\end{figure}
-
-%The transformations for {\sc Extend} and {\sc Include} dependencies  are not as simple as the previous transformations (figures \ref{fig:transformaton-examples}-a).
-
-The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  
-The generated workflow is composed by {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/36}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
-
-{\sf Include} use case entities are transformed into an {\sf Action} sequence.
-A {\sc Use case} element is transformed into an {\sf Action}. 
-A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements.
-
-Details of these transformations are not included here (due to space restrictions).
-They can be found in~\cite{SouzaNeto:2012}.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} \\
-The transformation rules have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).
-
-The ``listen music'' use case is transformed into a Service Action. 
-This Action  represents a Spotify service function that can be invoked to play the music. 
-For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}.
-The use case ``download music''  includes the payment process to buy the music. 
-Thus, these use cases  are transformed into {\sf Actions}, and a {\sf Service Activity} that aggregates these {\sf Actions}.   
-This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
-%(as shown in figure \ref{fig:transformation-example-include}). 
-The same rule is applied to the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
- \end{example}
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.76\textwidth]{figs/38}}
-%\caption{ Include Transformation Example}
-%\label{fig:transformation-example-include}
-%\end{figure}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-The  principle of the transformation  of a  $\pi$-Service Process model into a $\pi$-Service Composition model is to group {\sf Contracts} into {\sf A-Policies} and {\sf Actions} into {\sf Service Activities}.
-
-Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule}. 
-{\sf Rules}  concerning the same NFP  are grouped into {\sf A-Policies}. 
-Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule:Condition} attribute. 
-If the {\sf Assertion} has a value type, the name and the attributes are transformed into {\sf Variables} in the target model.  
-The {\sf Assertion: aProperty} attribute can have different transformations, according to: 
-\textit{(i)} a {\sf Precondition} is transformed into {\sf Pre};
-\textit{(ii)} a {\sf Post-Condition} is transformed into a {\sf  Post};
-\textit{(iii)} a {\sf TimeRestriction} is transformed into {\sf Time}.
-
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/ExceptionalRules}}
-%\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
-%\label{fig:ServiceProcess-ServiceComposition-Rules}
-%\end{figure}
-
-Some elements of the $\pi$-service composition model are obtained from the $\pi$-UseCase model: 
-A {\sf Package} in the $\pi$-use case model is transformed into a {\sf Business Collaborator};
-\textsf{Non-Functional Attributes} of the $\pi$-use case model are grouped into \textsf{Non-Functional requirements} of the $\pi$-ServiceComposition model. 
-These requirements are associated to a \textsf{Policy} (from the $\pi$-ServiceProcess model).
-
-{\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their hom\-onym concepts of the $\pi$-Service Composition model.
-
-Finally, {\sf Actions} are grouped into  a {\sf Business collaborator}.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=1\textwidth]{figs/ServiceProcess-ServiceComposition}}
-%\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
-%\label{fig:ServiceProcess-ServiceComposition}
-%\end{figure}
-
-%As the $\pi$-ServiceComposition model refines the $\pi$-ServiceProcess concepts at PIM level, a service previously defined as actions (source model) is refined as composition of those actions (target model) that are necessary to represent a business service, identifying who are the partners involved in the realization ({\sc Business collaborators}).
-%In addition, $\pi$-SOD-M defines a platform specific model based on web services composition.
-This model is explicitly indicates those actions which are (or will be, if not yet implemented) supported by web services.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}\\
-Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
-The ``securityLoginPolicy'' consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
-The information about Facebook and Spotify (both of them {\sf Business
-Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
-model.
-\end{example}
-
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-ServiceComposition to $\pi$-PEWS}
-% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-
-This section describes the PIM to PSM transformations from a $\pi$-Service composition mod\-el to a $\pi$-PEWS model. 
-We distinguish two groups of rules: \textit{(i)} those transforming service composition entities into workflows; and \textit{(ii)} those transforming  A-Policies into Contracts.
-
-Single actions ({\sf Action} and {\sf
-Action:name}) are transformed into individual service {\sf Operations}.
-Complex actions (rep\-re\-sented by {\sf ServiceActivity}  and  {\sf
-ServiceActivity:name}) are transformed into (named) composite
-operations that defines a (named) workflow of the application.
-Composition patterns expressed using  {\em merge, decision, fork} and {\em join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{figs/Table7}}
-%\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
-%\label{fig:ServiceComposition-Pews-Rules}
-%\end{figure}
-
-A-Policies defined for the entities of a $\pi$-service composition model generate {\sf A-policy} entities, named according to the names expres\-sed in the source model.
-The transformation of the rules expres\-sed in a $\pi$-service composition is guided by the event types associated to these rules. 
-The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into  {\sf Variable}, with  {\sf Name} and {\sf Type} attributes specified from the elements {\sf Variable: name} and {\sf Variable:type} of a $\pi$-service composition model.
-
-Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Pre-conditions}, {\sf Post-condition} and {\sf TimeRestrictions}.
-
-\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT6}
-Figure \ref{fig:piSOD-M} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.
-\end{example}
-
-\subsection{Implementation}
-We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
-Our tool is implemented as a series of Eclipse plug-ins: 
-\begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
-From these meta-models, we  developed plug-ins to support their graphical representation.
-
-\item	 We used  ATL\footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
-for implementing the  mappings between models.
-
-\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
-It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
-A-Policy} based service composition execution environment.
-\end{itemize}
-
-\section{Applying \pisodm: The \FlyingPig\ case study}
-\label{sec:flyingPig}
-
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
-Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
-The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
-The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
-By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.
-
-Our goal is to model a service based application (\FlyingPig), for providing risk assessment as a service.
-In order to provide this functionality, \FlyingPig\ benefits from ORCA's legacy services providing storage, assessment and data visualization functions.
-
-In the following sections, we describe the results of applying \pisodm to develop the \FlyingPig\ risk assessment system.
-The models presented next were generated as a result of interacting with software developers at GCP Global.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figs/3ValueModel.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
-\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
-\end{figure*}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Computation-Independent Models (CIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
-It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
-%\footnote{\color{red} This sentence need to move to section 3.1!!}
-We identify two business actors: \textsl{ORCA} and \textsl{Broker}.
-Brokers emit requests for risk assessment for one or more companies.
-ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}.
-The values exchanged between ORCA and the brokers are:
-(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
-(ii) \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires;
-(iii) \textsl{Evaluation Reports} for the client companies;
-and
-(iv) the risk assessment \textsl{fee}.
-
-The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company.
-Once this need has been declared, the value exchanges between ORCA and Broker are triggered.
-The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
-In the next step, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
-\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
-\end{figure*}
-
-Figure~\ref{fig:BPMNmodel} shows the BPMN model
-%\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.}
-for the \FlyingPig\ scenario. The BPMN model for \FlyingPig\ is partitioned for better understanding the process in which the value exchanges occur.
-The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
-Brokers have two lanes, the client \textsl{Company} and a \textsl{User}.
-The user is a contact member of the company, who  coordinates the assessment process.
-This process  involves other members of the company as well.
-
-The risk assessment process starts after a request from a company.
-This corresponds to the value model, in which the start stimulus triggers the whole process.
-The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
-
-Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
-During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
-
-Business processes also have rules and constraints to define their non-functional requirements (NFR):
-\begin{numtrivlist}
-\item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment.
-\item The system should be able to deal with, at least, 200 users.
-\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests.
-\item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
-\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
-\item All questionnaires need to be fully answered in order to consider that a task is completed.
-\item There is a time limit (in days) for amendments required by the system.
-\end{numtrivlist}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\subsection{Platform-Independent Models (PIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-
-%In Section~\ref{sec:modelingWithPISODM} we defined three models at the PIM level.
-The \pisodm models of the PIM level  for the \FlyingPig\ scenario are presented next.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
-\caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-UseCase Model for \FlyingPig}}~
-
-The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application.
-In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
-In the context of \FlyingPig, Company is the actor requesting a risk evaluation.
-A Broker is  responsible for coordinating  the evaluation process, assigning users to be in charge of tasks as well as delegating tasks.
-A User, in this model is an actor who answers questionnaires according to the current situation  of the Company.
-The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
-
-Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Fig.~\ref{fig:piUseCaseModel}).
-%Use cases
-that describe the main functionality of the system.
-The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.
-
-Each $\pi$-use case may be associated to one or more (non-functional) constraints (coloured ovals in Figure~\ref{fig:piUseCaseModel}).
-The mod\-el defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}.
-Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
-
-%In the case of \FlyingPig, as expressed above the model constraints:
-%{\color{red} Anything else here? --M}
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneralCut.png}
-\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
-
-The $\pi$-ServiceProcess model presents the workflow for \FlyingPig (Figure~\ref{fig:PiServiceProcessModel}).
-Actions in this model were obtained by applying the $\pi$-use case transformation rules.
-% described in Section~\ref{sec:pewsmetamodel}.
-
-The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
-Use cases are transformed into \textit{actions} and are represented by white boxes.
-The restrictions associated to  $\pi$-use cases are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions.
-We can see that this model refines the concepts defined in the $\pi$-UseCase model.
-The assertions specify those non-functional requirements, as they are seen by the actors.
-The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
-\caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-ServiceComposition Model for \FlyingPig}}~
-
-The model presented in Figure~\ref{fig:PiServiceCompositionModel}
-%offers a system-centric view of the application.
-%The previous model
-is now  completed with the services that provide functions of the application  \FlyingPig\.
-The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
-These policies express the pre- and post-conditions of the previous model.
-They are associated to the actions of the system.
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-%\subsection{Platform-Specific Model (PSM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\paragraph{\underline{$\pi$-PEWS Model for \FlyingPig}}~
-
-The  PSM model of our case study is obtained by transforming the $\pi$-service composition model into a workflow.
-Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
-The model is given in Figure~\ref{fig:piPEWSFlyingPig}.
-
-\begin{figure*}
-\begin{verbatim}
-//Namespaces specify service URI
-namespace orca = www.orca.mx/service.wsdl
-
-//Operations
-alias createCompanySpace = portType/createCompanySpace in orca
-alias generateQuestionnaire = portType/generateQuestionnaire in orca
-alias analizeAnswer = portType/analizeAnswer in orca
-alias storeData = portType/storeData in orca
-alias createReport = portType/createReport in orca
-alias callForAmendments = portType/callForAmendments in orca
-
-//Services
-service receiveRequest(R, Id) = createCompanySpace(R, Id)
-service generateNewInterface(Id, NULL) = ...
-service createQuestionnaire(Id, Q) = generateQuestionnaire(Id, Q)
-service notifyResponsable((Id, Q); NULL) = ...
-service receiveAnswers((Id, T); P) =
-	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL)
-	. ((createReport(Id, P) . return(P)) 
-	    + (callForAmendments(Id,T) . return(NULL)))
-
-//Workflow
-receiveRequest(R, Id)
-||
-(generateNewInterface(Id, NULL)
- . createQuestionnaire(Id, Q) . notifyResponsable((Id, Q); NULL))
-||
- (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
-\end{verbatim}
-\caption{$\pi$-PEWS Model for \FlyingPig}\label{fig:piPEWSFlyingPig}
-\end{figure*}
-
-\subsection{Lessons Learned}
-
-Through the example we underlined that every application implements functional aspects that describe its application logic.
-Recall that an application logic refers to routines that perform the activities to reach the application objective.
-Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
-These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
-
-The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
-API does not necessarily export information about methods dependency (e.g., in the REST protocol);
-they do not share their state (stateless).
-
-Given a set of services with their exported methods, building service-based applications may consist on expressing an application logic as a service composition.
-During this task, we must ensure the compliance between the specification and the resulting application.
-Software engineering methods (e.g., \cite{2,decastro1,PapazoglouH06}) can help to ensure this compliance, particularly when information systems include several sometimes complex business processes calling Web services or legacy applications exported as services.
-
-As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way.
-Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-\section{Conclusions and future work}\label{sec:conclusions}
-We presented \pisodm, a model-driven method for designing and developing reliable service-based applications.
-\pisodm extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
-These requirements are taken into account from the early stages of the software development process.
-Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.
-
-Our method includes two CIM-level models, three PIM-level models and one PSM-level model.
-We implemented the meta-models on the Eclipse platform and we validated the approach by using an industrially inspired use case.
-
-Our case study demonstrates the applicability of \pisodm.
-This case study was developed together with our industrial partner, GCP Global.
-The Company is using \pisodm for the development of their product.
-The case study presented here is a simplified version of their application.
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-\end{document}

### 2015-07-24 RITA2015/RITA_Template/RITAexemplo.tex
Article in the RITA format. To be refined...

%
% exemplo de artigo usando o modelo RITA.cls
%
\documentclass{RITA}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage[brazilian]{babel}

%
% preâmbulo
%
\title{Revista RITA: Instruções para Preparação de Documentos em \LaTeX}
\author{
Maria Lúcia B. Lisbôa
\footnote{Instituto de Informática, UFRGS, Caixa Postal~15064\\
  \texttt{\{llisboa,carla@inf.ufrgs.br\}}}
\and
Carla M. Dal~Sasso Freitas
\footnotemark[1]
\and
Outro Autor
\footnote{Endereço de Outro Autor\\
  \texttt{\{outroautor@email.br\}}}
}
\RITAvolume{VIII}
\RITAnumber{1}
\RITAyear{2001}

%
% início do texto
%
\begin{document}
\maketitle

\begin{abstract}
As instruções aqui contidas objetivam auxiliar os autores na
preparação de documentos para impressão na Revista RITA. Os estilos
mencionados encontram-se definidos em um modelo denominado Rita.dot. O
resumo deve ser escrito na mesma língua do texto (Português, Inglês ou
Espanhol) e descreve o conteúdo do texto em cerca de 150-200
palavras. Font Times, tamanho 10, margens laterais reduzidas em 1 cm
de cada lado, com duas linhas em branco antes e depois do
resumo. Formato abstract. Esta é a primeira versão das instruções e
dos formatos e, portanto, sujeita a incorreções e omissões. Sugestões
de melhorias são muito benvindas: envie mensagem para as editoras.
\end{abstract}

\begin{abstractinenglish}
Resumo traduzido para a língua inglesa.
\end{abstractinenglish}

\section{Usando o Modelo Rita}
O modelo Rita é um arquivo de estilos, com a extensão .dot, que pode ser armazenado na pasta de modelos do Office, juntamente com o modelo Normal.dot, para torná-lo visível automaticamente. Ou ainda você pode alterar o caminho de busca usando Ferramentas/Opções/Arquivos/Modelos do usuário...modificar.

Usando o modelo para criar um novo documento: abrir um novo documento  (Arquivo/Novo) e selecione em ModeloRita.  O novo documento terá a barra de ferramentas de estilos Rita automaticamente visível (ver Exibir/Barra de Ferramentas/ModeloRita).

Aplicando o modelo em um documento existente: abrir um arquivo existente (Arquivo/Abrir) e anexar o modelo através de Ferramentas/Modelos e Suplementos/Adicionar... ModeloRita.dot e após Ferramentas/Modelos e Suplementos/Anexar... ModeloRita.dot. Percorrer o documento original aplicando os estilos Rita simplesmente clicando na barra de ferramentas ou usando a seleção de estilo via barra padrão.

Se você tiver dificuldades em usar a nossa folha de estilos, use a folha normal de estilos procurando seguir os formatos descritos neste documento e usando nomes de estilos iguais aos aqui usados.

\section{Seção}
O título de seção deve ser em negrito, numerado automaticamente (1, 1.1,..), Times 12, com 14 espaços antes e 12 após. Este espaçamento pode ser ligeiramente alterado quando houver necessidade de completar uma página.

A preparação do documento deve ser cuidadosa, principalmente quanto a seu conteúdo técnico e correção gramatical. Alertamos os autores que não dispomos de pessoal especializado para fazer correção de textos nas três línguas aceitas (Português, Inglês, Espanhol).

A primeira linha de cada parágrafo possui indentação de 1 cm e as linhas observam espaçamento simples. Cada parágrafo  deve ser separado do seguinte por  6  espaços. Este espaçamento pode ser ligeiramente alterado quando houver necessidade de completar uma página. Estilo text1.

\subsection{Subseção (estilo heading2)}
Cada subseção deve ser em negrito, Times 10, com 12 espaços antes e 8 após. Este espaçamento pode ser ligeiramente alterado quando houver necessidade de completar uma página. Segue-se um parágrafo no estilo text1.

\section{Preparação dos originais}
O texto pode ser preparado usando LaTeX (ou TeX) ou Word. Para Latex, por favor veja as instruções contidas no documento RitaLatex.doc. Na preparação com Word, deve ser usada a folha de estilos Rita.dot, disponível na página da Revista.

Ao ler estas instruções, observe na barra de ferramentas o formato correspondente. Use estes mesmos formatos ao preparar seu manuscrito. Por favor, procure seguir as instruções para que a Revista RITA possua uma aparência uniforme.

\subsection{Área de impressão}
A área de impressão é de 13cmX17cm sendo o tamanho da página definido como 16cm $×$ 23cm. O texto deve ocupar a linha inteira e as figuras não podem ultrapassar as margens definidas. Procure preencher toda a área de impressão da página de modo não deixar mais do que 2 cm em branco no final de cada página. Faça seu maior esforço, movimentando figuras e texto, para completar cada página.

\subsection{Títulos}
Os títulos, incluindo o título principal, devem iniciar com letra maiúscula, sendo as demais letras minúsculas e todas as demais palavras do título devem iniciar por letra maiúscula. Por exemplo: " Revista RITA: Instruções para Preparação de Documentos em Word".

\subsection{Cabeçalhos, rodapé e numeração de páginas}
O cabeçalho contém o título do artigo e o rodapé contém a identificação da revista, seu volume e número e a numeração de páginas. Tanto cabeçalho quanto rodapé são espelhados: páginas de numeração ímpar possuem cabeçalho e rodapé alinhados à direita e as de numeração par, à esquerda. Formatos evenhead (caceçalho par), oddhead (cabeçalho ímpar), evenfoot(rodapé par) e oddfoot (rodapé impar). A primeira página não contém cabeçalho; apenas rodapé direito, por iniciar sempre em numeração ímpar.

\subsection{Legendas de tabelas e de figuras}
As tabelas devem ser numeradas a partir de 1 e indicadas por Tabela <num>. <legenda> no topo da mesma, conforme exemplo mostrado na Tabela~\ref{tab:exemplo}. No texto, sempre faça referência a Tabela <num> ( não usar as expressões acima, abaixo, a seguir, etc).  Proceda do mesmo modo com as figuras, lembrando que a legenda da figura fica abaixo da mesma.

\begin{table}
\begin{center}
\caption{Formatos de títulos}
\label{tab:exemplo}
\begin{tabular}{lll}
        \hline
	Item (alinhamento) & Exemplo & Font e estilo\\
	\hline
	Título (esquerdo) & \Large\textbf{Título do Artigo} & 14~pontos, negrito\\
	Seção (esquerdo) & \large\textbf{1~Introdução} & 12~pontos,
	negrito\\
	$\cdots$ & $\cdots$ & $\cdots$\\
	\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Figuras e fotografias}
A impressão da Revista RITA é feita em tons de branco e preto. Portanto, evite fazer uso de fotografias coloridas, a menos que, quando transformadas em tons de cinza seus detalhes continuem visíveis.

As figuras devem ser integradas no texto, centralizadas de acordo com as margens. Para testar a visibilidade dos detalhes de suas figuras, por favor, faça a geração de um arquivo imagem de impressão (postscript) e observem se todos os detalhes estão perfeitamente visíveis e os textos legíveis.  As figuras devem ser numeradas e todas devem ter uma legenda explicativa. Centralize a legenda entre as margens, use font Times 9 , e mantenha uma distância de cerca de 6 pontos antes e 12 pontos após (formato legend).

Tenha especial cuidado com figuras feitas diretamente com as ferramentas Word. Permita que elas flutuem sobre o texto, sem âncoras. Se houver necessidade de inserir uma quebra de página antes da figura, certifique-se que a página anterior esteja completa. Se a figura ocupar uma página completa, certifique-se que esta não ultrapasse as margens. Evite colocar figuras e tabelas no formato paisagem.

\subsection{Fórmulas e equações}
Equações e fórmulas deve ser colocadas em uma nova linha,
centralizadas e numeradas consecutivamente para fins de referência,
como pode ser observado na Equação~\ref{eq:exemplo}.

\begin{equation}
x + y = z
\label{eq:exemplo}
\end{equation}

\subsection{Program Code}
Listagens de código de programas devem usar um outro estilo de font, por exemplo, Courier 10, para que recebam destaque.  As listagens não são consideradas figuras, de modo que não necessitam ter legenda. Para fins de referência, as linhas do código podem  ser numeradas. Para numerar a tabela em Word, selecione a primeira célula da tabela, escolha no menu Formatar a opção Numeração (marcadores e numeração) e, na caixa de diálogos exibida selecione a opção Numerada. Ao fazer isto, a partir da primeira linha numerada, a numeração será feita automaticamente a cada troca de linha da tabela (Dica: para inserir uma nova linha, posicione o cursor no final da tabela e pressione Enter).

Por exemplo, o código a seguir mostra uma classe Java, onde a linha 6 inicia um comando que se estende por diversas linhas. Estilo progcode (tabela de duas colunas, com números na coluna 1).

\begin{verbatim}
 1. import java.util.Random;
 2. class Aleatorios {
 3.   public static void main (String[] args) {
 4.     Random qq=new Random();
 5.     for (int k=1;k<10;k++)
 6.       System.out.println("Usando o método nextInT() da classe Random: " +
 7.         qq.nextInt(100) + "\nUsando o método Math.random(): " +
 8.         Math.random());
 9.   }
10. }
\end{verbatim}

\subsection{Notas de rodapé}
As notas de rodapé são usadas na primeira página para a identificação dos autores e, ao longo do texto, podem ser usadas para esclarecimentos.

\subsection{Citações e referências e agradecimentos}
Ao longo do texto, as citações são feitas através de números consecutivos entre colchetes de acordo com a ordem de citação, como em [1],[2],[3]. Ao final, a lista de referências deve ter o nome de Referências (negrito, font Times 12, alinhado à esquerda), sem numeração de seção. Não colocar quebra de página antes. As referências devem ser numeradas, seguindo o texto. Ver exemplo a seguir.

\subsection{Conferência final}
Ao submeter seu trabalho à Revista RITA, verifique se você possui o seguinte:
Latex: seu arquivo fonte (arquivos TEX para o texto e PS ou EPS para as figuras); arquivos ou macros adicionais que você tenha usado; e o arquivo final DVI. Se possível, prepara um arquivo PDF para a submissão inicial.
Word: seu arquivo fonte (arquivo  DOC) e arquivo em imagem de impressão PS ou PDF.

Para a submissão inicial enviar apenas o arquivo DVI, ou PS ou PDF.

\end{document}

### 2015-07-24 RITA2015/main.tex
Article in the RITA format. To be refined...

--- 2015-04-10 JournalToSubmit2015/submitted/main.tex
@@
-\documentclass{ws-ijseke}
+\documentclass{RITA}
+\usepackage[T1]{fontenc}
+%\usepackage[latin1]{inputenc}
+\usepackage[brazilian]{babel}
+
+% Our packages
 \usepackage{xspace}
+\usepackage{amsmath}
+\usepackage[thmmarks,amsmath]{ntheorem}
+\usepackage{graphicx}
+
+% Our definitions
+\newcommand{\openbox}{\leavevmode
+  \hbox to.77778em{%
+  \hfil\vrule
+  \vbox to.675em{\hrule width.6em\vfil\hrule}%
+  \vrule\hfil}}
+
+\theoremstyle{plain}
+\theoremheaderfont{\normalfont\bfseries}
+\theorembodyfont{\normalfont}
+\theoremseparator{}
+\theoremindent0cm
+\theoremnumbering{arabic}
+\newtheorem{algo}{Algorithm}
+
+\theoremstyle{plain}
+\theoremheaderfont{\normalfont\bfseries}
+\theorembodyfont{\normalfont}
+\theoremseparator{}
+\theoremindent0cm
+\theoremnumbering{arabic}
+\theoremsymbol{\ensuremath{\openbox}} 
+\newtheorem{example}{Example}
 
@@
 
+% RITA Preamble
+
+\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
+\author{
+Khalid Belhajjame
+\footnote{Universit\'e de Paris - Dauphine -- Paris, France\\
+  \texttt{\{kbelhajj@googlemail.com\}}}
+\and
+Valeria~de~Castro
+\footnote{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
+  \texttt{\{valeria.decastro@urjc.es\}}}
+\and
+Umberto~Souza~da~Costa
+\footnote{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+  \texttt{\{\{umberto,mam\}@dimap.ufrn.br\}}}
+\and
+Javier~A.~Espinosa-Oviedo
+\footnote{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\
+  \texttt{\{javiera.espinosa@gmail.com, joseluis.zechinelli@udlap.mx\}}}  
+\and
+Martin~A.~Musicante
+\footnotemark[3]
+\and
+Pl\'acido~A.~Souza~Neto
+\footnote{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\
+  \texttt{\{placido.neto@ifrn.edu.br\}}}  
+\and
+Genoveva~Vargas-Solar
+\footnote{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\
+  \texttt{\{genoveva.vargas@imag.fr\}}}  
+\and  
+Jos\'e-Luis~Zechinelli-Martini
+\footnotemark[4]
+}
+\RITAvolume{VIII}
+\RITAnumber{1}
+\RITAyear{2001}
+
 \begin{document}
-
-\markboth{K.Belhajjame \textit{et al.}}
-{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
-
-%%%%%%%%%%%%%%%%%%%%% Publisher's Area please ignore %%%%%%%%%%%%%%%
+\maketitle
+
+%\begin{abstract}
 %
-\catchline{}{}{}{}{}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
-
-\author{Khalid Belhajjame}
-\address{Universit\'e de Paris - Dauphine -- Paris, France\\
-\email{kbelhajj@googlemail.com}
-}
-
-\author{Valeria~de~Castro} 
-\address{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-\email{valeria.decastro@urjc.es}
-}
-
-\author{Umberto~Souza~da~Costa} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{umberto@dimap.ufrn.br}
-}
-
-\author{Javier~A.~Espinosa-Oviedo} 
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{javiera.espinosa@gmail.com}
-}
-
-\author{Martin~A.~Musicante} 
-\address{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-\email{mam@dimap.ufrn.br}
-}
-
-\author{Pl\'acido~A.~Souza~Neto} 
-\address{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
-\email{placido.neto@ifrn.edu.br}
-}
-
-\author{Genoveva~Vargas-Solar} 
-\address{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
-\email{genoveva.vargas@imag.fr}
-}
-
-\author{Jos\'e-Luis~Zechinelli-Martini}
-\address{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-\email{joseluis.zechinelli@udlap.mx}
-}
-
-\maketitle
-
-\begin{history}
-\received{(Day Month Year)}
-\revised{(Day Month Year)}
-\accepted{(Day Month Year)}
-%\comby{(xxxxxxxxxx)}
-\end{history}
-
-\begin{abstract}
+%\end{abstract}
+
+\begin{abstractinenglish}
 Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
@@
 The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
-\end{abstract}
-
-\keywords{MDA \and Non-Functional Requirements \and Service-based software process}
-
-%% main text
-%*********************************************************************************************************
+\end{abstractinenglish}
+
+%\keywords{MDA , Non-Functional Requirements, Service-based software process}
+
 \section{Introduction}
@@
 combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
+The selection of services is usually guided by the \textit{functional}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
 An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/piSOD-M_process.pdf}
+\includegraphics[width=1.0\textwidth]{./figures/piSOD-M_process.pdf}
 \caption{\pisodm Overview.}
@@
 \center
-\includegraphics[width=0.75\textwidth]{figs/e3value.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
+\includegraphics[width=0.75\textwidth]{./figures/e3value.pdf}
+\hspace*{5cm}\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}
 \caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
@@
 \center
-\includegraphics[width=0.45\textwidth]{figs/SC.pdf}
+\includegraphics[width=0.45\textwidth]{./figures/SC.pdf}
 \caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/UseCaseMetaModel.pdf}
+\includegraphics[width=0.9\textwidth]{./figures/UseCaseMetaModel.pdf}
 \caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/UseCase.pdf}
+\includegraphics[width=0.9\textwidth]{./figures/UseCase.pdf}
 \caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcessMetaModel.pdf}
+\includegraphics[width=0.9\textwidth]{./figures/ServiceProcessMetaModel.pdf}
 \caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
@@
 \center
-\includegraphics[width=0.9\textwidth]{figs/ServiceProcess.pdf}
+\includegraphics[width=0.9\textwidth]{./figures/ServiceProcess.pdf}
 \caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/PiServiceComposition}
+\includegraphics[width=1.0\textwidth]{./figures/PiServiceComposition}
 \caption{$\pi$-Service Composition Meta-model.}
@@
 \centering
-\includegraphics[width=0.95\textwidth]{figs/piServiceComposition-toPublishMusic}
+\includegraphics[width=0.95\textwidth]{./figures/piServiceComposition-toPublishMusic}
 %{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{figs/PEWSMetamodel}
+\includegraphics[width=0.75\textwidth]{./figures/PEWSMetamodel}
 \caption{$\pi$-{\sc Pews} Meta-model.}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figs/pi-pewsSpecification-toPublishMusic}
+\includegraphics[width=0.8\textwidth]{./figures/pi-pewsSpecification-toPublishMusic}
 \caption{$\pi$-PEWS Specific and Policy Representation.}
@@
 \begin{itemize}
-\item 	We  used the Eclipse Modeling Framework (EMF)\footnote {The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
+\item 	We  used the Eclipse Modeling Framework (EMF)\footnote{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
 From these meta-models, we  developed plug-ins to support their graphical representation.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs/3ValueModel.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{figs/3ValueKey.pdf}
+\includegraphics[width=0.9\textwidth]{./figures/3ValueModel.pdf}
+\hspace*{5cm}\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}
 \caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/BPMN_GCP.pdf}
+\includegraphics[width=1.0\textwidth]{./figures/BPMN_GCP.pdf}
 \caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs/UseCaseGeneral.png}
+\includegraphics[width=0.9\textwidth]{./figures/UseCaseGeneral.png}
 \caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceProcessGeneralCut.png}
+\includegraphics[width=1.0\textwidth]{./figures/ServiceProcessGeneralCut.png}
 \caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
@@
 \centering
-\includegraphics[width=1.0\textwidth]{figs/ServiceCompositionGeneralCut}
+\includegraphics[width=1.0\textwidth]{./figures/ServiceCompositionGeneralCut}
 \caption{$\pi$-ServiceComposition model for \FlyingPig.\label{fig:PiServiceCompositionModel}}

### 2015-07-28 RITA2015/main.tex
Compressed version -- 29 Pages.

@@
 %\usepackage[latin1]{inputenc}
-\usepackage[brazilian]{babel}
-
-% Our packages
+\usepackage[english]{babel}
+
+% Our packages and definitions
 \usepackage{xspace}
-\usepackage{amsmath}
-\usepackage[thmmarks,amsmath]{ntheorem}
 \usepackage{graphicx}
-
-% Our definitions
-\newcommand{\openbox}{\leavevmode
-  \hbox to.77778em{%
-  \hfil\vrule
-  \vbox to.675em{\hrule width.6em\vfil\hrule}%
-  \vrule\hfil}}
-
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{}
-\theoremindent0cm
-\theoremnumbering{arabic}
-\newtheorem{algo}{Algorithm}
-
-\theoremstyle{plain}
-\theoremheaderfont{\normalfont\bfseries}
-\theorembodyfont{\normalfont}
-\theoremseparator{}
-\theoremindent0cm
-\theoremnumbering{arabic}
-\theoremsymbol{\ensuremath{\openbox}} 
-\newtheorem{example}{Example}
+\usepackage{paralist} % for compact lists: compactitem, compactenum, compactdesc
+\usepackage{lipsum}  % for the example environment
+\usepackage{amssymb} % for the \blacksquare symbol
+\usepackage{subfigure}
+
+\newcounter{examplecounter}
+\newenvironment{example}{%\begin{quote}%
+    \refstepcounter{examplecounter}%
+  \textbf{Example \arabic{examplecounter}}%
+  \quad
+}{%
+%\end{quote}%
+$\blacksquare$ 
+}
+
+\let\OLDthebibliography\thebibliography
+\renewcommand\thebibliography[1]{
+  \OLDthebibliography{#1}
+  \setlength{\parskip}{0pt}
+  \setlength{\itemsep}{0pt plus 0.7ex}
+}
 
@@
 
-\newcounter{numberInTrivlist}
-
-\newenvironment{numtrivlist}{\begin{list}{\rm \arabic{numberInTrivlist})}
-                                         {\usecounter{numberInTrivlist}
-                                          \setlength{\leftmargin}{0pt}
-                                          \setlength{\rightmargin}{0pt}
-                                          \setlength{\itemindent}{12pt}
-                                          \setlength{\listparindent}{0pt}}}
-                            {\end{list}}
-
-\newenvironment{itemizedTrivlist}{\begin{list}{\rm ~\hspace{2mm} $\bullet$\ }
-                                         {\setlength{\leftmargin}{0pt}
-                                          \setlength{\rightmargin}{0pt}
-                                          \setlength{\itemindent}{12pt}
-                                          \setlength{\listparindent}{0pt}}}
-                            {\end{list}}
-
 % RITA Preamble
 
-\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
+\title{\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements}
 \author{
@@
 }
-\RITAvolume{VIII}
-\RITAnumber{1}
-\RITAyear{2001}
+\RITAvolume{22}
+\RITAnumber{2}
+\RITAyear{2015}
 
@@
 This paper presents \pisodm, a method and associated tools that
-\textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details;
+\textit{(i)}  allows the early specification of NFRs in a principled way: users are abstracted away from low level details;
 \textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
-The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
+Our solution has been used in the context of an industrial and real case study.
 \end{abstractinenglish}
@@
 
-%Service oriented computing is at the origin of an evolution in the field of software development.
-%Service oriented methods advocates for the construction of software systems formed by the composition of heterogeneous, loosely coupled modules.
-%These modules (or services) communicate in order to achieve a common purpose.
-
-%(Moreover, IT systems need to evolve according to the business needs.)
-%Thus, organizations are seeking for mechanisms to bridge the gap between the actually developed systems and their business needs~\cite{bell}.
-
 In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
 combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
+The selection of services is usually guided by the \textit{functional}\footnote{Functional properties of a computer system are characterized by the effects it produces when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
 An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
@@
 Non-functional properties, such as data privacy, exception handling, atomicity  and, data persistence, need to be addressed to fit in the application.
-%Adding non-functional properties and respecting services constraints while composing services is a complex task that implies implementing protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
-%Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements.
-%
-%Furthermore,  often they are  not fully considered in the specification process of existing services' oriented development methods. Rather,
-%They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed.
 Ideally, Non-Functional Requirements (NFR) should be considered in every phase of the software development.
@@
 
-%Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
-%Most methods integrate them only after the application has been implemented.
-%This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.
-
-% song title disseminated in the walls of the user's Facebook and Twitter accounts.
-
-%The adoption of non-functional specifications from the early states of development can help the developer to produce applications that  can deal with the application context.
-
-The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
-Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
+The literature emphasizes the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
+Existing approaches assert that the convergence of model-driven soft\-ware development, service orientation  and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
 Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
 These models are successively refined from abstract specifications into actual computer programs. 
-A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
+A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
 Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.
 
-\bigskip
 Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
@@
 \pisodm proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
-Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller} approach.
+Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller}.
 The goals of  \pisodm are to:
-%SOD-M does not provide support for the specification of non-functional requirements, such as
-%security, reliability, and efficiency.
-\begin{trivlist}
-\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
-\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
-\item \textit{(iii)} Provide a general way of describing non-functional requirements.
-\end{trivlist}
+
+\begin{compactenum}[i]
+\item Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
+\item Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
+\item Provide a general way of describing non-functional requirements.
+\end{compactenum}
 
@@
 
-\bigskip
 This paper is organized as follows.
 Sec\-tion \ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
-Sections~\ref{sec:motivation} and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules.
+Sections~\ref{sec:motivation} and~\ref{sec:mmrules} introduce respectively the meta-models and transformation rules of \pisodm.
 % Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology.
@@
 Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
-
-Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
-In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.
-
-In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:
-\textit{service level}, \textit{system level} and \textit{business level}.
-%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
-%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
-%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}.
-The work also defines a \textit{web service quality model}, which considers non-functional properties.
-In~\cite{XiaoCZBOLH08} the authors use the terms
+%
+Most research  on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
+In service-based applications, NFRs are related to the application itself as well as to  its component services.
+
+In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three perspectives:
+\textit{service}, \textit{system} and \textit{business levels}.
+In~\cite{XiaoCZBOLH08}  authors use the terms
 \textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
@@
 Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
-A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
+A non-func\-tion\-al concern is a general term used to describe NFRs, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
 Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
@@
 
-Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
-However, distinct hierarchies and models are proposed for NFRs, according to different points of view.
+Despite of the different notations found in the literature for classifying NFRs, some NFRs are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
+However, distinct hierarchies and models are proposed for NFRs, according to different perspectives.
 We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.
 
-Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}.
+Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010}.
 As Ameller et al.
@@
 Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
-Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.
+Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.
 
@@
 
-The proposal in~\cite{ThissenW06} presents steps for  selecting services
-by taking QoS information into account. The proposed steps are:
+The proposal in~\cite{ThissenW06} presents steps to  select services
+by taking QoS information into account:
 \textit{(i)} identification of relevant QoS information;
@@
 \textit{(iii)} definition of a selection mechanism of services.
-The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
+The authors consider  \textit{performance}, \textit{cost}, \textit{reliability} and
 \textit{availability}.
 
-Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
-The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.
-
-Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
+Karunamurthy et al.~\cite{Karunamurthy2012787} define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}, as  \textit{non-function parameters}.
+The \textit{Non-Func\-tion\-al Specification Language} 
+%(NFSL)
+ is proposed as a domain specific language 
+ %(DSL) 
+ to express \textit{non-function parameters}.
+
+Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe NFRs such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
 In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.
@@
 For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
-Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.
-
-%\bigskip
-%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
-%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
-%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
-%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
-%It may use web services provided by third party sources.
-%The System level deals with the coordination of services, to implement the business logic.
-%Our proposal, shown in the next section, follows these ideas.
-
-\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}
-
-In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
+Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are to be specified as constraints.
+
+\section{Modeling Reliable Service Compositions with \pisodm}\label{sec:motivation}
+
+In this section we present \pisodm, an MDD-based method for building service compositions with NFRs.
 Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
@@
 The refinement process is driven by transformation rules specified between the meta-models.
-
-In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
-The presentation is exemplified by a running example.
+%
+The me\-ta-models are presented from higher to lower levels of abstraction along with a running example.
 The transformations between models are presented in section~\ref{sec:mmrules}.
 
-\begin{figure*}
+\begin{figure*}[h]
 \centering
-\includegraphics[width=1.0\textwidth]{./figures/piSOD-M_process.pdf}
+\includegraphics[width=0.75\textwidth]{./figures/piSOD-M_process.pdf}
 \caption{\pisodm Overview.}
@@
 .
-\subsubsection{E$^3$value meta-model}
+\subsubsection{E$^3$value Meta-Model}
 % .
@@
 
-The e$^3$value model of an application identifies the value/information exchange between components of the system.
-It is a business model that represents a business case graphically as a set of value exchanges\footnote{See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
+The e$^3$value model  identifies the value/in\-for\-ma\-tion  exchange between system components (see Figure~\ref{fig:CIM:tpme3v}).
+This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
 The model is well suited to enhance the understanding of the environment in which an application is being developed.
@@
 A dependency path has a direction and consists of a sequence of linked dependency nodes.
-It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}).
+It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node.
 Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).
 
-\begin{figure*}
+\begin{figure*}[h]
 \center
-\includegraphics[width=0.75\textwidth]{./figures/e3value.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}
+\subfigure{\includegraphics[width=0.55\textwidth]{./figures/e3value.pdf}}
+~ 
+\subfigure{\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}}
 \caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
@@
 The following (external) services will be used by the application:
-\begin{itemize}
+
+\begin{compactitem}
 \item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
-\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
+ {\sf\small get-Last-Song ( userid ): String}; 
 \item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
-\begin{itemize}
-%\item{\sf\small get-Status ( usedid ): String ;
-\item {\sf\small update-Status ( userid, new-status ): String};
-\end{itemize}
-\end{itemize}
+ {\sf\small update-Status ( userid, new-status ): String};
+\end{compactitem}
 
@@
 
-\subsubsection{BPMN meta-model}
+\subsubsection{BPMN Meta-Model}
 
@@
 %
-\begin{figure}
+\begin{figure}[h]
 \center
-\includegraphics[width=0.45\textwidth]{./figures/SC.pdf}
+\includegraphics[width=0.4\textwidth]{./figures/SC.pdf}
 \caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
@@
 .
-\subsubsection{\textit{$\pi$-UseCase} meta-model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-
-Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
-The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
+\subsubsection{\textit{$\pi$-UseCase} Meta-Model}% .
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+.
+
+Figure~\ref{fig:CIM:usecasemetamodel} shows the \textit{$\pi$-UseCase} meta-model.
+This model is  to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
 The notion of \textit{policy} is used to describe NFRs.
 (Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
 An {\sc End Consumer} is represented by an {\sc Actor}.
-An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
+An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), as a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
 {\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
- The {\sc Business Collaborator} concept is represented  through {\sc Packages}.
-A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled.
+ The {\sc Business Collaborator} concept is represented  by {\sc Packages}.
+A {\sc Business Collaborator} denotes an external service or a system that interacts with the application that is being modeled.
 Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
@@
 
- \begin{figure*}
+ \begin{figure*}[h]
 \center
-\includegraphics[width=0.9\textwidth]{./figures/UseCaseMetaModel.pdf}
+\includegraphics[width=0.8\textwidth]{./figures/UseCaseMetaModel.pdf}
 \caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
@@
 \textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
-The use of these constraints is shown in the following example.
+These constraints are shown in the next example.
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
-Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example application.
+Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example.
 We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
+
+\begin{figure*}[h]
+\center
+\includegraphics[width=0.8\textwidth]{./figures/UseCase.pdf}
+\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
+\end{figure*}
 
@@
 For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
-Another restriction is that the minimum payment value is 2 euros.
+Minimum payment value is 2 euros.
 %{\color{red} Talk about this model for the example.}
@@
 
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{./figures/UseCase.pdf}
-\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
-\end{figure*}
-
  % .
@@
 .
-\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .
+\subsubsection{\textit{$\pi$-ServiceProcess} Meta-Model}% .
 .
@@
 
-A specific \textit{$\pi$-ServiceProcess} model shows the set of logically related activities to be performed in a service-based application.
-So, the activities of this model represent a behavior that is part of the business logic of the application.
+\begin{figure*}[h]
+\center
+\includegraphics[width=0.8\textwidth]{./figures/ServiceProcessMetaModel.pdf}
+\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
+\end{figure*}
+
+This model shows the set of logically related activities to be performed in a service-based application.
+The activities of this model represent a behavior that is part of the business logic of the application.
 A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
@@
 The {\sc Assertion}s associated to an {\sc Action} compose a {\sc Contract}. 
-Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre-conditions and post-conditions.
-
-\begin{figure*}
-\center
-\includegraphics[width=0.9\textwidth]{./figures/ServiceProcessMetaModel.pdf}
-\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
-\end{figure*}
-
-%According to the meta-model, a \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts.
-It models the set of system activities and describe contracts for each activity or the process as a whole.
-Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action.
-This model defines a workflow for  modeling business logic and the behaviour of the application being developed.
-The behaviour is represented by each action and the restrictions by the assertion description.
-All assertions are described by stereotyped actions.
+Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre- and post-conditions.
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
-Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}.
+Considering the example scenario, the contract based process of activities  is shown in Figure \ref{fig:CIM:serviceprocess}.
 The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service.
@@
 
-\begin{figure*}
+\begin{figure*}[h]
 \center
-\includegraphics[width=0.9\textwidth]{./figures/ServiceProcess.pdf}
+\includegraphics[width=0.8\textwidth]{./figures/ServiceProcess.pdf}
 \caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
@@
 .
-\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .
+\subsubsection{\textit{$\pi$-ServiceComposition} Meta-Model}% .
 .
@@
 For instance, security and privacy restrictions may be grouped into a security policy.
- The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
-%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
-\begin{itemizedTrivlist}
-\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action.
-An instance of this meta-class is graphically represented as a partition in the activity diagram.
+ This meta-model defines:
+
+\begin{compactitem}
+
+\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  action.
+An instance of this meta-class is graphically a partition in the activity diagram.
 A collaborator can be either internal or external to the system.
@@
 
-\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class \textsc{Action}.
+\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented as a class activity instance of the meta-class \textsc{Action}.
 A class action represents some type of transformation or processing.
@@
 
-\begin{figure*}
+\begin{figure*}[h]
 \centering
-\includegraphics[width=1.0\textwidth]{./figures/PiServiceComposition}
+\includegraphics[width=0.8\textwidth]{./figures/PiServiceComposition}
 \caption{$\pi$-Service Composition Meta-model.}
@@
 
-\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more execu\-ta\-ble nodes.
+\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service,  composed by  execu\-ta\-ble nodes.
 
@@
 An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.
-\end{itemizedTrivlist}
-Instances of this meta-model are represented as UML activity diagrams.
-\begin{figure*}
+
+\end{compactitem}
+
+\begin{figure*}[h]
 \centering
@@
 %{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
-\caption{$\pi$-ServiceComposition Model for the ``To publish
-music'' business service.}
+\caption{$\pi$-ServiceComposition Model for the ``To publish music'' business service.}
 \label{fig:servicecompositionmodel}
@@
 
-%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
-%These artifacts will be used to generate the actual programs that will implement the application:
-Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation.
-The $\pi$-PEWS  meta-model is described in the next section.
+Once the $\pi$-Service Composition Model is defined,  it can be transformed into a lower level model (in our case, $\pi$-PEWS) to support code generation as described in the next section.
 
@@
 Models at this level put together the platform-independent view with the specific aspects of the platform to implement the system.
-Models at the PSM level can be automatically translated into actual computer programs.
+These models  can be automatically translated into actual computer programs.
 We have defined one meta-model at this level.
@@
 Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
-
+%
 PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
@@
 Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:
-\begin{itemizedTrivlist}
-\item Service compositions: {\sc Namespace} representing the interface exported by a service, {\sc Operation} that represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path} for representing service compositions.
+
+\begin{compactitem}
+\item Service compositions: {\sc Namespace} represents the interface exported by a service, {\sc Operation} represents a call to a service method, {\sc CompositeOperation}, {\sc Operator} and {\sc Path}  denote service compositions.
 A {\sc Path} can be an {\sc Operation} or a {\sc Compound Operation}.
@@
 The language defines operators to denote guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) repetition.
-
-\item {\em A-Policies} that can be associated to a service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
-A brief description of these classes is given next.
-\end{itemizedTrivlist}
-%
-\begin{figure*}
-\centering
-\includegraphics[width=0.75\textwidth]{./figures/PEWSMetamodel}
-\caption{$\pi$-{\sc Pews} Meta-model.}
-\label{fig:PPEWSmetamodel}
-\end{figure*}
+\item {\em A-Policies} that can be associated to service compositions:  {\sc A-Policy}, {\sc Rule}, {\sc Event}, {\sc Condition}, {\sc Action}, {\sc State}, and {\sc Scope}.
+\end{compactitem}
 
@@
 
+\begin{figure*}[h]
+\centering
+\includegraphics[width=0.7\textwidth]{./figures/PEWSMetamodel}
+\caption{$\pi$-{\sc Pews} Meta-model.}
+\label{fig:PPEWSmetamodel}
+\end{figure*}
+
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic5}
-Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our scenario example.
-Notice that the code contains namespaces and definitions (obtained by using information about the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
+Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$-service composition model of our example.
+The code contains namespaces and definitions (obtained from the business collaborators of the $\pi$-SCM model), a workflow expression \textit{(Path)} containing operation calls and contracts (derived from the \textit{A-Policies}).
 \end{example}
 
-\begin{figure*}
+\begin{figure*}[h]
 \centering
-\includegraphics[width=0.8\textwidth]{./figures/pi-pewsSpecification-toPublishMusic}
+\includegraphics[width=0.5\textwidth]{./figures/pi-pewsSpecification-toPublishMusic}
 \caption{$\pi$-PEWS Specific and Policy Representation.}
@@
 
-\section{Transformation rules}\label{sec:mmrules}
+\section{Transformation Rules}\label{sec:mmrules}
 
@@
 \pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models.
-These rules are used to build a semi-automatic tool (the \pisodm environment plugin for Eclipse).
-The transformations from CIM to PIM level models is not automatized, due to the informal nature of CIM level representations.
+These rules are used to build a semi-automatic tool (the \pisodm  plugin for Eclipse).
+The transformations from CIM to PIM level models are not automatized, due to the informal nature of CIM level.
 
@@
 Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.   
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{35}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess transformation rules}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
-
+%
 A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
@@
 
-%A {\sf Requirement} and a {\sf Composite use case} in a  $\pi$-UseCase model are transformed respectively into a {\sf Service activity} and an {\sf Action} (see figures  \ref{fig:transformationsUseCase-ServiceProcessRules}-c  and d).
-%
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.80\textwidth]{37}}
-%\caption{ Extended Transformation Examples}
-%\label{fig:transformaton-examples}
-%\end{figure}
-
-%The transformations for {\sc Extend} and {\sc Include} dependencies  are not as simple as the previous transformations (figures \ref{fig:transformaton-examples}-a).
-
 The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  
 The generated workflow is composed by {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{36}}
-%\caption{ $\pi$-UseCase to $\pi$-ServiceProcess Model Transformation Rules (2)}
-%\label{fig:transformationsUseCase-ServiceProcessRules}
-%\end{figure}
 
@@
 A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements. 
-
-Details of these transformations are not included here (due to space restrictions).
-They can be found in~\cite{SouzaNeto:2012}.
+%
+Details of these transformations can be found in~\cite{SouzaNeto:2012}.
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT1} 
-The transformation rules have been applied to the model in Figure~\ref{fig:piUseCaseModel}, in order to obtain the $\pi$-Service Process model for our running example (Figure~\ref{fig:CIM:serviceprocess}).
-
-The ``listen music'' use case is transformed into a Service Action. 
-This Action  represents a Spotify service function that can be invoked to play the music. 
+The transformation rules have been applied to the model in Figure~\ref{fig:CIM:piusecasetpm} to obtain the $\pi$-Service Process model for our example (Figure~\ref{fig:CIM:serviceprocess}).
+
+The ``listen music'' use case is transformed into a Service Action that  represents a Spotify  function to be invoked to play music. 
 For the ``publish music'' use case,  constraints are transformed into a set of assertions that are grouped into a Contract ({\sf ``publishMusicContract''}) associated to the Action {\sf ``publishMusic''}.
@@
 This \textsf{Service Activity} is transformed into a sequence flow on the $\pi$-service process model.
-%(as shown in figure \ref{fig:transformation-example-include}). 
 The same rule is applied to the ``publish music'' use case, which has two extended use cases, to publish on Twitter and Facebook.
  \end{example}
-
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.76\textwidth]{38}}
-%\caption{ Include Transformation Example}
-%\label{fig:transformation-example-include}
-%\end{figure}
 
@@
 
-The  principle of the transformation  of a  $\pi$-Service Process model into a $\pi$-Service Composition model is to group {\sf Contracts} into {\sf A-Policies} and {\sf Actions} into {\sf Service Activities}.
-
+The transformation  of a  $\pi$-Service Process model into a $\pi$-Service Composition model groups {\sf Contracts} into {\sf A-Policies} and {\sf Actions} into {\sf Service Activities}.   
 Each {\sf Assertion} of a {\sf Contract} is transformed into a {\sf Rule}. 
@@
 
-%
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{ExceptionalRules}}
-%\caption{$\pi$-ServiceProcess to $\pi$-ServiceComposition Model Transformation Rules}
-%\label{fig:ServiceProcess-ServiceComposition-Rules}
-%\end{figure}
-
 Some elements of the $\pi$-service composition model are obtained from the $\pi$-UseCase model: 
@@
 {\sf Actions} and {\sf Service activity} of a $\pi$-Service Process model are transformed into their hom\-onym concepts of the $\pi$-Service Composition model.
-
 Finally, {\sf Actions} are grouped into  a {\sf Business collaborator}.
-%\begin{figure}
-%\centering{
-%\includegraphics[width=1\textwidth]{ServiceProcess-ServiceComposition}}
-%\caption{ Transformation Rules: From $\pi$-ServiceProcess to $\pi$-ServiceComposition}
-%\label{fig:ServiceProcess-ServiceComposition}
-%\end{figure}
-
-%As the $\pi$-ServiceComposition model refines the $\pi$-ServiceProcess concepts at PIM level, a service previously defined as actions (source model) is refined as composition of those actions (target model) that are necessary to represent a business service, identifying who are the partners involved in the realization ({\sc Business collaborators}).
-%In addition, $\pi$-SOD-M defines a platform specific model based on web services composition.
-This model is explicitly indicates those actions which are (or will be, if not yet implemented) supported by web services.
 
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT5}
-Considering the example scenario, the model in Figure~\ref{fig:PiServiceCompositionModel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
+Considering the example scenario, the model in Figure~\ref{fig:servicecompositionmodel} was obtained by applying the rules above to the model depicted by Figure~\ref{fig:CIM:serviceprocess}.  
 The ``securityLoginPolicy'' consists of a set of {\sf Rules} that were transformed from the {\sf Assertions} in $\pi$-service process model. 
-The information about Facebook and Spotify (both of them {\sf Business
-Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
+The information about Facebook and Spotify (both of them {\sf Business Collaborators}) come from entity of type {\sc Package} in the $\pi$-use case
 model.
@@
 
-%\begin{figure}
-%\centering{
-%\includegraphics[width=0.96\textwidth]{Table7}}
-%\caption{Transformation Rules: From $\pi$-ServiceComposition to $\pi$-PEWS}
-%\label{fig:ServiceComposition-Pews-Rules}
-%\end{figure}
-
 A-Policies defined for the entities of a $\pi$-service composition model generate {\sf A-policy} entities, named according to the names expres\-sed in the source model.
@@
 The variables associated to an A-Policy expressed in a $\pi$-service composition model as {\sf $<$Variable:name, Variable:type$>$} are transformed into  {\sf Variable}, with  {\sf Name} and {\sf Type} attributes specified from the elements {\sf Variable: name} and {\sf Variable:type} of a $\pi$-service composition model.
-
+%
 Events of type {\sf Pre}, {\sf Post} and {\sf Time} generate, respectively, {\sf Pre-conditions}, {\sf Post-condition} and {\sf TimeRestrictions}.
@@
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicT6}
-Figure \ref{fig:piSOD-M} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model  of our scenario example.
+Figure \ref{fig:Specific-Contract-Representation} shows the $\pi$-PEWS code resulting from the $\pi$- service composition model (Figure~\ref{fig:servicecompositionmodel}) of our scenario example.
 \end{example}
@@
 Our tool is implemented as a series of Eclipse plug-ins: 
-\begin{itemize}
+\begin{compactitem}
 \item 	We  used the Eclipse Modeling Framework (EMF)\footnote{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
@@
 A-Policy} based service composition execution environment.
-\end{itemize}
-
-\section{Applying \pisodm: The \FlyingPig\ case study}
+\end{compactitem}
+
+\section{Applying \pisodm: The \FlyingPig\ Case Study}
 \label{sec:flyingPig}
@@
 To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
-Risk assessment is implemented by an interactive business process based on the exchange of a series of questionnaires intended to evaluate the risks implied by the client business practice.
+Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
 Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
-The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes of the company, as well as to propose amendments to these practices.
+The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes, as well as to propose amendments to these practices.
 The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
@@
 In order to provide this functionality, \FlyingPig\ benefits from ORCA's legacy services providing storage, assessment and data visualization functions.
-
-In the following sections, we describe the results of applying \pisodm to develop the \FlyingPig\ risk assessment system.
-The models presented next were generated as a result of interacting with software developers at GCP Global.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.9\textwidth]{./figures/3ValueModel.pdf}
-\hspace*{5cm}\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}
-\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
-\end{figure*}
-
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%
+In the next sections we apply \pisodm to develop the \FlyingPig\ risk assessment system.
+The models presented were generated as a result of interacting with software developers at GCP Global.
+
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -.fig:E3valuemodel . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
 \subsection{Computation-Independent Models (CIM)}
@@
 It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
-%\footnote{\color{red} This sentence need to move to section 3.1!!}
 We identify two business actors: \textsl{ORCA} and \textsl{Broker}.
@@
 (iv) the risk assessment \textsl{fee}.
-
+%
 The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company.
@@
 
-\begin{figure*}
+\begin{figure*}[h]
 \centering
-\includegraphics[width=1.0\textwidth]{./figures/BPMN_GCP.pdf}
-\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
+\includegraphics[width=0.45\textwidth]{./figures/3ValueModel.pdf}
+%\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}
+\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
 \end{figure*}
@@
 Figure~\ref{fig:BPMNmodel} shows the BPMN model
-%\footnote{\color{red} Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/.}
-for the \FlyingPig\ scenario. The BPMN model for \FlyingPig\ is partitioned for better understanding the process in which the value exchanges occur.
+for the \FlyingPig\ scenario. This model is partitioned for better understanding the value exchanging process.
 The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
@@
 This process  involves other members of the company as well.
-
+%
 The risk assessment process starts after a request from a company.
@@
 
+\begin{figure*}[h]
+\centering
+\includegraphics[width=0.6\textwidth]{./figures/BPMN_GCP.pdf}
+\caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
+\end{figure*}
+
 Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
 During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.
-
 Business processes also have rules and constraints to define their non-functional requirements (NFR):
-\begin{numtrivlist}
+
+\begin{compactenum}
 \item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment.
@@
 \item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
-\item All questionnaires need to be fully answered in order to consider that a task is completed.
+\item All questionnaires need to be fully answered in order to consider a task as completed.
 \item There is a time limit (in days) for amendments required by the system.
-\end{numtrivlist}
+\end{compactenum}
 
@@
 
-\begin{figure*}
+\subsubsection{$\pi$-UseCase Model for \FlyingPig}
+
+The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints of the \FlyingPig\ application.
+In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
+Company is the actor requesting a risk evaluation as a Broker is  responsible for coordinating  the evaluation process, assigning users to tasks as well as delegating tasks.
+A User is an actor who answers questionnaires according to the current situation  of the Company.
+The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
+Each actor is associated to $\pi$-UseCases (white ovals in Fig.~\ref{fig:piUseCaseModel})
+that describe the main functionality of the system.
+The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-UseCases.
+%%
+Each $\pi$-UseCase may be associated to (non-functional) constraints (coloured ovals in Fig.~\ref{fig:piUseCaseModel}).
+Three types of constraints are defined: \textit{value}, \textit{business} or \textit{exceptional behavior}.
+Constraints are identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
+
+\begin{figure*}[h]
 \centering
-\includegraphics[width=0.9\textwidth]{./figures/UseCaseGeneral.png}
+\includegraphics[width=0.55\textwidth]{./figures/UseCaseGeneral.png}
 \caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
@@
 
-\paragraph{\underline{$\pi$-UseCase Model for \FlyingPig}}~
-
-The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints for the \FlyingPig\ application.
-In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
-In the context of \FlyingPig, Company is the actor requesting a risk evaluation.
-A Broker is  responsible for coordinating  the evaluation process, assigning users to be in charge of tasks as well as delegating tasks.
-A User, in this model is an actor who answers questionnaires according to the current situation  of the Company.
-The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
-
-Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Fig.~\ref{fig:piUseCaseModel}).
-%Use cases
-that describe the main functionality of the system.
-The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.
-
-Each $\pi$-use case may be associated to one or more (non-functional) constraints (coloured ovals in Figure~\ref{fig:piUseCaseModel}).
-The mod\-el defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}.
-Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
-
-%In the case of \FlyingPig, as expressed above the model constraints:
-%{\color{red} Anything else here? --M}
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{./figures/ServiceProcessGeneralCut.png}
-\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
-\end{figure*}
-
-\paragraph{\underline{$\pi$-ServiceProcess Model for \FlyingPig}}~
+\subsubsection{$\pi$-ServiceProcess Model for \FlyingPig}
 
@@
 Actions in this model were obtained by applying the $\pi$-use case transformation rules.
-% described in Section~\ref{sec:pewsmetamodel}.
-
 The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
@@
 
-\begin{figure*}
+\begin{figure*}[h]
+\centering
+\includegraphics[width=1.0\textwidth]{./figures/ServiceProcessGeneralCut.png}
+\caption{$\pi$-ServiceProcess model for \FlyingPig.\label{fig:PiServiceProcessModel}}
+\end{figure*}
+
+\subsubsection{$\pi$-ServiceComposition Model for \FlyingPig}
+
+The model in Figure~\ref{fig:PiServiceCompositionModel}
+shows  the services that provide the functions of \FlyingPig.
+The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
+These policies express the pre- and post-conditions of the previous model, being  associated to the actions of the system.
+
+\begin{figure*}[h]
 \centering
@@
 
-\paragraph{\underline{$\pi$-ServiceComposition Model for \FlyingPig}}~
-
-The model presented in Figure~\ref{fig:PiServiceCompositionModel}
-%offers a system-centric view of the application.
-%The previous model
-is now  completed with the services that provide functions of the application  \FlyingPig\.
-The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
-These policies express the pre- and post-conditions of the previous model.
-They are associated to the actions of the system.
-
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
@@
 %. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-\paragraph{\underline{$\pi$-PEWS Model for \FlyingPig}}~
-
-The  PSM model of our case study is obtained by transforming the $\pi$-service composition model into a workflow.
-Notice that the workflow for \FlyingPig\ is implemented by using the BPEL constructors~\cite{BPEL}, adding the notion of \textit{policy}.
-The model is given in Figure~\ref{fig:piPEWSFlyingPig}.
-
-\begin{figure*}
+\subsubsection{$\pi$-PEWS Model for \FlyingPig}
+
+The  PSM of our case study is given in Figure~\ref{fig:piPEWSFlyingPig}. This model is obtained by transforming the $\pi$-service composition model into a workflow.
+Note that this workflow is implemented by using  BPEL constructors~\cite{BPEL} and \textit{policy} notions.
+
+\begin{figure*}[h]
+\begin{scriptsize}
 \begin{verbatim}
@@
 namespace orca = www.orca.mx/service.wsdl
-
 //Operations
@@
 alias callForAmendments = portType/callForAmendments in orca
-
 //Services
@@
 service receiveAnswers((Id, T); P) =
-	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL)
-	. ((createReport(Id, P) . return(P)) 
-	    + (callForAmendments(Id,T) . return(NULL)))
-
+	analizeAnswer((Id, T), NULL) . storeData((Id, T), NULL) . 
+	((createReport(Id, P) . return(P)) + (callForAmendments(Id,T) . return(NULL)))
 //Workflow
 receiveRequest(R, Id)
-||
-(generateNewInterface(Id, NULL)
+|| (generateNewInterface(Id, NULL)
  . createQuestionnaire(Id, Q) . notifyResponsable((Id, Q); NULL))
-||
- (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
+|| (receiveAnswers((Id, T); P) . [P != NULL] STOP)*	
 \end{verbatim}
-\caption{$\pi$-PEWS Model for \FlyingPig}\label{fig:piPEWSFlyingPig}
+\caption{$\pi$-PEWS Model for \FlyingPig.}\label{fig:piPEWSFlyingPig}
+\end{scriptsize}
 \end{figure*}
@@
 These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
-
+%
 The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
@@
 As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, our work specifies policies for a services composition in an orthogonal way.
+In contrast to these approaches, we specify policies for a services composition in an orthogonal way.
 Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
-In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
-
-\section{Conclusions and future work}\label{sec:conclusions}
+In our approach,  the policies defined for a given application  can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
+
+\section{Conclusions and Future Work}\label{sec:conclusions}
 We presented \pisodm, a model-driven method for designing and developing reliable service-based applications.
@@
 Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.
-
+%%
 Our method includes two CIM-level models, three PIM-level models and one PSM-level model.
@@
 
-Our case study demonstrates the applicability of \pisodm.
-This case study was developed together with our industrial partner, GCP Global.
+Our case study was developed together with our industrial partner GCP Global to demonstrate the applicability of \pisodm.
 The Company is using \pisodm for the development of their product.
-The case study presented here is a simplified version of their application.
+The case study presented is a simplified version of their application.
 

### 2016-04-08 JournalPISODM2016/main.tex
Version for the new journal - same article as RITA

--- 2015-07-28 RITA2015/main.tex
@@
-\documentclass{RITA}
-\usepackage[T1]{fontenc}
-%\usepackage[latin1]{inputenc}
-\usepackage[english]{babel}
-
-% Our packages and definitions
+%%%%%%%%%%%%%%%%%%%%%%
+\documentclass{singlecol-new}
+%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{natbib,stfloats}
+\usepackage{mathrsfs}
+
+\def\newblock{\hskip .11em plus .33em minus .07em}
+
+\theoremstyle{TH}{
+\newtheorem{lemma}{Lemma}
+\newtheorem{theorem}[lemma]{Theorem}
+\newtheorem{corrolary}[lemma]{Corrolary}
+\newtheorem{conjecture}[lemma]{Conjecture}
+\newtheorem{proposition}[lemma]{Proposition}
+\newtheorem{claim}[lemma]{Claim}
+\newtheorem{stheorem}[lemma]{Wrong Theorem}
+\newtheorem{algorithm}{Algorithm}
+}
+
+\theoremstyle{THrm}{
+\newtheorem{definition}{Definition}[section]
+\newtheorem{question}{Question}[section]
+\newtheorem{remark}{Remark}
+\newtheorem{scheme}{Scheme}
+\newtheorem{example}{Example}
+}
+
+\theoremstyle{THhit}{
+\newtheorem{case}{Case}[section]
+}
+
+%%%% Our commands %%%%
+
 \usepackage{xspace}
+\usepackage{paralist} % for compact lists: compactitem, compactenum, compactdesc
 \usepackage{graphicx}
-\usepackage{paralist} % for compact lists: compactitem, compactenum, compactdesc
-\usepackage{lipsum}  % for the example environment
-\usepackage{amssymb} % for the \blacksquare symbol
 \usepackage{subfigure}
-
-\newcounter{examplecounter}
-\newenvironment{example}{%\begin{quote}%
-    \refstepcounter{examplecounter}%
-  \textbf{Example \arabic{examplecounter}}%
-  \quad
-}{%
-%\end{quote}%
-$\blacksquare$ 
-}
-
-\let\OLDthebibliography\thebibliography
-\renewcommand\thebibliography[1]{
-  \OLDthebibliography{#1}
-  \setlength{\parskip}{0pt}
-  \setlength{\itemsep}{0pt plus 0.7ex}
-}
 
@@
 
-% RITA Preamble
+\makeatletter
+\def\theequation{\arabic{equation}}
+
+%\JOURNALNAME{\TEN{\it Int. J. System Control and Information
+%Processing,
+%Vol. \theVOL, No. \theISSUE, \thePUBYEAR\hfill\thepage}}%
+%
+%\def\BottomCatch{%
+%\vskip -10pt
+%\thispagestyle{empty}%
+%\begin{table}[b]%
+%\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcr@{}}%
+%\\[-12pt]
+%Copyright \copyright\ 2012 Inderscience Enterprises Ltd. & &%
+%\end{tabular*}%
+%\vskip -30pt%
+%%%\vskip -35pt%
+%\end{table}%
+%}
+\makeatother
+
+%%%%%%%%%%%%%%%%%
+\begin{document}%
+%%%%%%%%%%%%%%%%%
+
+\setcounter{page}{1}
+
+\LRH{K. Belhajjame et~al.}
+
+\RRH{\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements}
+
+\VOL{x}
+
+\ISSUE{x}
+
+\PUBYEAR{xxxx}
+
+\BottomCatch
+
+%\CLline
+
+\PUBYEAR{2016}
+
+\subtitle{}
 
 \title{\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements}
-\author{
-Khalid Belhajjame
-\footnote{Universit\'e de Paris - Dauphine -- Paris, France\\
-  \texttt{\{kbelhajj@googlemail.com\}}}
-\and
-Valeria~de~Castro
-\footnote{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-  \texttt{\{valeria.decastro@urjc.es\}}}
-\and
-Umberto~Souza~da~Costa
-\footnote{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-  \texttt{\{\{umberto,mam\}@dimap.ufrn.br\}}}
-\and
-Javier~A.~Espinosa-Oviedo
-\footnote{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\
-  \texttt{\{javiera.espinosa@gmail.com, joseluis.zechinelli@udlap.mx\}}}  
-\and
-Martin~A.~Musicante
-\footnotemark[3]
-\and
-Pl\'acido~A.~Souza~Neto
-\footnote{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\
-  \texttt{\{placido.neto@ifrn.edu.br\}}}  
-\and
-Genoveva~Vargas-Solar
-\footnote{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\
-  \texttt{\{genoveva.vargas@imag.fr\}}}  
-\and  
-Jos\'e-Luis~Zechinelli-Martini
-\footnotemark[4]
-}
-\RITAvolume{22}
-\RITAnumber{2}
-\RITAyear{2015}
-
-\begin{document}
-\maketitle
-
-%\begin{abstract}
-%
-%\end{abstract}
-
-\begin{abstractinenglish}
+
+%
+\authorA{Khalid Belhajjame}
+%
+\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
+  E-mail: kbelhajj@googlemail.com}
+%
+%
+\authorB{Valeria~de~Castro}
+\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
+  E-mail: valeria.decastro@urjc.es}
+\authorC{Umberto~Souza~da~Costa}
+\affC{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+  E-mail: umberto@dimap.ufrn.br}
+
+\authorD{Javier~A.~Espinosa-Oviedo}
+\affD{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
+  E-mail: javiera.espinosa@gmail.com}
+
+\authorE{Martin~A.~Musicante}
+\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+  E-mail: mam@dimap.ufrn.br}
+
+\authorF{Pl\'acido~A.~Souza~Neto}
+\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
+  E-mail: placido.neto@ifrn.edu.br}
+
+\authorG{Genoveva~Vargas-Solar}
+\affG{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
+  E-mail: genoveva.vargas@imag.fr}
+
+\authorH{Jos\'e-Luis~Zechinelli-Martini}
+\affH{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
+  E-mail: joseluis.zechinelli@udlap.mx}
+
+%
+%\authorA{Fusheng Wang\footnote{Work done while working at Siemens Corporate Research.} }
+%
+%\affA{Department of Biomedical Informatics, Emory University
+%\newline
+%36 Eagle Row, Ste 589, Atlanta, GA 30322, USA}
+%
+%
+%
+%\authorB{\footnotesize Cristobal Vergara-Niedermayr\footnote{Work done while working at Siemens Corporate Research.}}
+%\affB{Oracle \newline
+ %New Jersey, USA}
+%
+%
+
+\begin{abstract}
 Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
@@
 Our solution has been used in the context of an industrial and real case study.
-\end{abstractinenglish}
-
-%\keywords{MDA , Non-Functional Requirements, Service-based software process}
+\end{abstract}
+
+\KEYWORD{MDA , Non-Functional Requirements, Service-based software process.}
+
+%\REF{to this paper should be made as follows: Rodr\'{\i}guez
+%Bol\'{\i}var, M.P. and Sen\'{e}s Garc\'{\i}a, B. (xxxx) `The
+%corporate environmental disclosures on the internet: the case of
+%IBEX 35 Spanish companies', {\it International Journal of Metadata,
+%Semantics and Ontologies}, Vol. x, No. x, pp.xxx\textendash xxx.}
+
+\begin{bio}
+Khalid Belhajjame\dots Bio here.
+
+\noindent Valeria~de~Castro\dots Bio here.
+
+\noindent Umberto~Souza~da~Costa\dots Bio here.
+
+\noindent Javier~A.~Espinosa-Oviedo\dots Bio here.
+
+\noindent Martin~A.~Musicante\dots Bio here.
+
+\noindent Pl\'acido~A.~Souza~Neto\dots Bio here.
+
+\noindent Genoveva~Vargas-Solar\dots Bio here.
+
+\noindent Jos\'e-Luis~Zechinelli-Martini\dots Bio here.
+\end{bio}
+
+\maketitle
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%
+%\documentclass{RITA}
+%\usepackage[T1]{fontenc}
+%%\usepackage[latin1]{inputenc}
+%\usepackage[english]{babel}
+%
+%% Our packages and definitions
+%\usepackage{xspace}
+%\usepackage{graphicx}
+%\usepackage{paralist} % for compact lists: compactitem, compactenum, compactdesc
+%\usepackage{lipsum}  % for the example environment
+%\usepackage{amssymb} % for the \blacksquare symbol
+%\usepackage{subfigure}
+%
+%\newcounter{examplecounter}
+%\newenvironment{example}{%\begin{quote}%
+%    \refstepcounter{examplecounter}%
+%  \textbf{Example \arabic{examplecounter}}%
+%  \quad
+%}{%
+%%\end{quote}%
+%$\blacksquare$  
+%}
+%
+%\let\OLDthebibliography\thebibliography
+%\renewcommand\thebibliography[1]{
+%  \OLDthebibliography{#1}
+%  \setlength{\parskip}{0pt}
+%  \setlength{\itemsep}{0pt plus 0.7ex}
+%}
+%
+%\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
+%\def\FlyingPig{\textsl{FlyingPig}\xspace}
+%
+%% RITA Preamble
+%
+%\title{\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements}
+%\author{
+%Khalid Belhajjame
+%\footnote{Universit\'e de Paris - Dauphine -- Paris, France\\
+%  \texttt{\{kbelhajj@googlemail.com\}}}
+%\and
+%Valeria~de~Castro
+%\footnote{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
+%  \texttt{\{valeria.decastro@urjc.es\}}}
+%\and
+%Umberto~Souza~da~Costa
+%\footnote{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+%  \texttt{\{\{umberto,mam\}@dimap.ufrn.br\}}}
+%\and
+%Javier~A.~Espinosa-Oviedo
+%\footnote{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\
+%  \texttt{\{javiera.espinosa@gmail.com, joseluis.zechinelli@udlap.mx\}}}  
+%\and
+%Martin~A.~Musicante
+%\footnotemark[3]
+%\and
+%Pl\'acido~A.~Souza~Neto
+%\footnote{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\
+%  \texttt{\{placido.neto@ifrn.edu.br\}}}  
+%\and
+%Genoveva~Vargas-Solar
+%\footnote{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\
+%  \texttt{\{genoveva.vargas@imag.fr\}}}  
+%\and  
+%Jos\'e-Luis~Zechinelli-Martini
+%\footnotemark[4]
+%}
+%\RITAvolume{22}
+%\RITAnumber{2}
+%\RITAyear{2015}
+%
+%\begin{document}
+%\maketitle
+%
+%%\begin{abstract}
+%%
+%%\end{abstract}
+%
+%\begin{abstractinenglish}
+%
+%\end{abstractinenglish}
+%
+%%\keywords{}
 
@@
 
-\begin{figure*}[h]
+\begin{figure}[h]
 \centering
@@
 \label{fig:piSOD-M}
-\end{figure*}
+\end{figure}
 

### 2016-04-12 JournalPISODM2016/main.tex
new version

@@
 \newtheorem{scheme}{Scheme}
-\newtheorem{example}{Example}
 }
@@
 \newtheorem{case}{Case}[section]
+}
+
+\theoremstyle{THhsl}{
+\newtheorem{example}{Example}
 }
@@
 \def\FlyingPig{\textsl{FlyingPig}\xspace}
+
+\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
+\newcommand{\imartin}[1]{\textcolor{blue}{\textbf{Martin: }#1}}
+
+\newcommand{\umberto}[1]{\marginpar{\textcolor{orange}{\textbf{Umberto: }#1}}}
+\newcommand{\iumberto}[1]{\textcolor{orange}{\textbf{Umberto: }#1}}
 
@@
 Khalid Belhajjame\dots Bio here.
+\vs{9}
 
 \noindent Valeria~de~Castro\dots Bio here.
-
-\noindent Umberto~Souza~da~Costa\dots Bio here.
+\vs{9}
+
+\noindent Umberto Costa received the BS degree in Computer Science from Federal University of Rio Grande do Norte in 1998. 
+He obtained the MS degree in Systems and Computation from Federal University of Rio Grande do Norte in 2000. 
+He received the PhD degree in Computer Science from University Federal de Minas Gerais in 2005. He is currently an associate professor in the Department of 
+Informatics and Applied Mathematics (DIMAp) at Federal University of Rio Grande do Norte. His academic interests include Programming Languages, 
+Service-Oriented Computing and Cloud Computing.
+\vs{9}
 
 \noindent Javier~A.~Espinosa-Oviedo\dots Bio here.
-
-\noindent Martin~A.~Musicante\dots Bio here.
+\vs{9}
+
+\noindent Martin A. Musicante received his B.Sc at ESLAI, Argentina, in 1988 and
+his M.Sc and PhD in Computer Science at Universidade Federal de
+Pernambuco, Brazil, in 1990 and 1996, respectively. He is an associate
+professor at Universidade Federal do Rio Grande do Norte-UFRN (Natal,
+Brazil). He is part of the graduate program in Computer Science at
+UFRN and holder of a research grant from CNPq (level 2). Martin was
+associate researcher at LI - Universit\'e Fran\c{c}ois Rabelais Tours
+(2002-2011) and at LIFO - Universit\'e d'Orl\'eans, France, since 2008.
+Martin A. Musicante's main current research topics are big data
+processing languages, Web Services and cloud computing.
+\vs{9}
 
 \noindent Pl\'acido~A.~Souza~Neto\dots Bio here.
+\vs{9}
 
 \noindent Genoveva~Vargas-Solar\dots Bio here.
+\vs{9}
 
@@
 Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller}.
-The goals of  \pisodm are to:
+The goals of  \pisodm are:
 
 \begin{compactenum}[i]
-\item Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
-\item Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
-\item Provide a general way of describing non-functional requirements.
+\item To improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
+\item To reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
+\item To provide a general way of describing non-functional requirements.
 \end{compactenum}
@@
 Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
-Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
+Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Attributes} and \textit{Non-Functional Actions} are used at each level (in decreasing order of abstraction).
 A non-func\-tion\-al concern is a general term used to describe NFRs, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
@@
 Each NFR is associated with a service activity.
-
-Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
-Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services.
-Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}.
-Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
-Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09},
-\textit{property}~\cite{Fabra2011},
-\textit{factor}~\cite{MohantyRP10,GutierrezRF10},
-\textit{characteristic}~\cite{DiamadopoulouMPS08},
-\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
-\textit{value}~\cite{ThissenW06,BasinDL06}.
-
-Despite of the different notations found in the literature for classifying NFRs, some NFRs are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
-However, distinct hierarchies and models are proposed for NFRs, according to different perspectives.
-We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.
-
-Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010}.
-As Ameller et al.
-realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
-Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
-Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.
 
@@
 
+Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
+Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services.
+Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}.
+Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
+Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09},
+\textit{property}~\cite{Fabra2011},
+\textit{factor}~\cite{MohantyRP10,GutierrezRF10},
+\textit{characteristic}~\cite{DiamadopoulouMPS08},
+\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
+\textit{value}~\cite{ThissenW06,BasinDL06}.
+
+Despite of the different notations found in the literature for classifying NFRs, some NFRs are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
+However, distinct hierarchies and models are proposed for NFRs, according to different perspectives.
+We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.
+
+Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010,SouzaNeto201684}.
+As Ameller et al.
+realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
+Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
+Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.
+
 \section{Modeling Reliable Service Compositions with \pisodm}\label{sec:motivation}
@@
 %
-The me\-ta-models are presented from higher to lower levels of abstraction along with a running example.
-The transformations between models are presented in section~\ref{sec:mmrules}.
+The purpose of this section is to present the \pisodm me\-ta-models, from higher to lower level of abstraction, along with a running example.
+The transformations between models are presented next, in section~\ref{sec:mmrules}.
 
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./figures/piSOD-M_process.pdf}
+\includegraphics[width=1\textwidth]{./figures/piSOD-M_process.pdf}
 \caption{\pisodm Overview.}
@@
 \center
-\subfigure{\includegraphics[width=0.55\textwidth]{./figures/e3value.pdf}}
-~ 
-\subfigure{\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}}
+\subfigure{\includegraphics[width=1\textwidth]{./figures/e3value.pdf}}
+
+\subfigure{\includegraphics[width=0.7\textwidth]{./figures/3ValueKey.pdf}}
 \caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
@@
 The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
-The e$^3$value value model shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users.
-The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription).
+The e$^3$value value model shows Spotify and a private application (which is also a service) that provide services for listening and publishing information about music being listened by users. 
+The private application interacts with Spotify for obtaining information about the flow of music being listened by a user in return of a fee for a premium subscription. 
 Finally, the private application interacts with Facebook and Twitter for updating the user's status.
@@
 However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
-In order to face this shortcoming, our method proposes the BPMN meta-model as a tool for modelling this aspect of the application.
+In order to face this shortcoming, our method proposes the use of the BPMN meta-model as a tool for modelling this aspect of the application.
 
@@
 
-BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow.
+BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}.
 The next example illustrates the use of this meta-model.
@@
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
-Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
+Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model of the scenario.
 It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
@@
 \center
-\includegraphics[width=0.4\textwidth]{./figures/SC.pdf}
+\includegraphics[width=0.7\textwidth]{./figures/SC.pdf}
 \caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
@@
 
-The CIM level models are the basis for the development of the application.
-The information presented at this level is (manually) refined into PIM-level models of the method \pisodm described next.
-Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.
+CIM level models are the basis for the development of the application.
+The information presented at this level is (manually) refined into PIM-level models of \pisodm, as described next.
+Notice that the models are (informally) used to describe both functional and non-functional requirements.
 
@@
 
-This level focuses on the system functionality, hiding the details of any particular platform.
+This level focuses on the system functionality, abstracting the details of any particular platform.
 The specification defines those parts of the system that do not change from one platform to another.
@@
 (Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}\footnote{We use the {\sc small caps} font for referring to classes of a meta-model.}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
 An {\sc End Consumer} is represented by an {\sc Actor}.
@@
 \center
-\includegraphics[width=0.8\textwidth]{./figures/UseCaseMetaModel.pdf}
+\includegraphics[width=1\textwidth]{./figures/UseCaseMetaModel.pdf}
 \caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
@@
 A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
-Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint}) use case.
-There are three kinds of constraint:
-\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value'';
-\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
-\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
-These constraints are shown in the next example.
+Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint})\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} use case.
+There are three kinds of constraints:
+\textit{(i)} A data pair ({\sc Value Constraint}), represented as the stereotype ``value'';
+\textit{(ii)} Business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
+\textit{(iii)} An {\sc Exceptional Behavior} constraint, represented by the stereotype {\sf ``exceptional\_behavior''}.
+The use of these constraints is shown in the next example.
 
@@
 Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example.
-We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
+We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the \textit{(i)} conditions imposed by services usage -- for example, the fact that both Facebook and Twitter require authentication in order to call their methods for updating the wall; \textit{(ii)}  conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
 
@@
 \center
-\includegraphics[width=0.8\textwidth]{./figures/UseCase.pdf}
+\includegraphics[width=1\textwidth]{./figures/UseCase.pdf}
 \caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
@@
 The authentication constraint is required to update a music status.
-The restriction is stereotyped as a value constraint, because the users id and password are verified.
-Figure \ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases.
+The restriction is stereotyped as a value constraint, because the user id and password are verified.  
+Figure~\ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases. 
 The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as value and business constraint, respectively. 
-For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
+For payment, the user must provide the data of payment card or PayPal account login and password, represented by a value stereotype.
 Minimum payment value is 2 euros.
@@
 \center
-\includegraphics[width=0.8\textwidth]{./figures/ServiceProcessMetaModel.pdf}
+\includegraphics[width=1\textwidth]{./figures/ServiceProcessMetaModel.pdf}
 \caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
@@
 \begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
-Considering the example scenario, the contract based process of activities  is shown in Figure \ref{fig:CIM:serviceprocess}.
+Considering the example scenario, the contract based process of activities  is shown in Figure~\ref{fig:CIM:serviceprocess}.
 The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service.
 The buy music pre-conditions consist in verifying:
-(i) if the User data are correct;
-(ii) if the User is already logged in Spotify;
-(iii) if bank account information are correct and;
-(iv) if there are enough funds in the bank account to cover the payment.
-A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization.
+\textit{(i)} if the User data are correct;
+\textit{(ii)} if the User is already logged in Spotify;
+\textit{(iii)} if bank account information are correct and;
+\textit{(iv)} if there are enough funds in the bank account to cover the payment.
+A post-condition ensures the complete transaction and verifies if a notification about the payment authorization was sent both to the user and to Spotify. 
 There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. 
@@
 \center
-\includegraphics[width=0.8\textwidth]{./figures/ServiceProcess.pdf}
+\includegraphics[width=1\textwidth]{./figures/ServiceProcess.pdf}
 \caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
@@
 In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
-provides meta-classes to represent workflows\footnote{Workflows are transformed into implemented service compositions.} that model  business processes.
+provides meta-classes to represent workflows%\footnote{Workflows will be transformed into (implemented) service compositions.} 
+ that model  business processes.
 The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
 to group contracts with similar non-functional requirements.
-For instance, security and privacy restrictions may be grouped into a security policy.
+For instance, security and privacy restrictions may be grouped into a single policy.
  This meta-model defines:
@@
 \begin{compactitem}
-
 \item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  action.
@@
 A collaborator can be either internal or external to the system.
-When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
+When the collaborator of the business is external to the system, the attribute {\sf IsExternal} of the collaborator is set to \textbf{true}.
 
@@
 A class action represents some type of transformation or processing.
-There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
+There are two types of actions: \textit{(i)} a WebService (attribute Type is {\sf WS}); and \textit{(ii)} a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{./figures/PiServiceComposition}
+\includegraphics[width=1\textwidth]{./figures/PiServiceComposition}
 \caption{$\pi$-Service Composition Meta-model.}
@@
 
-\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue\martin{Placido: Mudar as cores...
+} meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
 We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
@@
 \centering
-\includegraphics[width=0.95\textwidth]{./figures/piServiceComposition-toPublishMusic}
+\includegraphics[width=1\textwidth]{./figures/piServiceComposition-toPublishMusic}
 %{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
@@
 Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
-Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that read and update the user data.
+Both {\em Facebook} and {\em Twitter} services require authentication in order to execute methods that read and update the user data.
 %A call to such services must be part of the authentication protocol required by these services.
@@
 In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
-{\sf\small OAuthPolicy} implements the open authentication protocol.
 The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
 This variable is imported through the library {\sf\small OAuthPolicy.Token}.
-The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed.
+The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: ({\sf\small R$_1$}): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and ({\sf\small R$_2$}): if the token has expired, then it is renewed.
 Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
@@
 \textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
-Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
+Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{Placido2010LTPD}.
 %
@@
 \centering
-\includegraphics[width=0.7\textwidth]{./figures/PEWSMetamodel}
+\includegraphics[width=1\textwidth]{./figures/PEWSMetamodel}
 \caption{$\pi$-{\sc Pews} Meta-model.}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{./figures/pi-pewsSpecification-toPublishMusic}
+\includegraphics[width=1\textwidth]{./figures/pi-pewsSpecification-toPublishMusic}
 \caption{$\pi$-PEWS Specific and Policy Representation.}
@@
 
-\pisodm  proposes intra-level transformation rules between $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules between $\pi$-service composition to $\pi$-PEWS models.
-These rules are used to build a semi-automatic tool (the \pisodm  plugin for Eclipse).
-The transformations from CIM to PIM level models are not automatized, due to the informal nature of CIM level.
+\pisodm  proposes intra-level transformation rules from $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules from $\pi$-service composition to $\pi$-PEWS models.
+These rules were implemented into a semi-automatic tool (the \pisodm  plugin for Eclipse).
+The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level.
 
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
-\subsection{From $\pi$-UseCase to the $\pi$-ServiceProcess}
+\subsection{From $\pi$-UseCase to $\pi$-ServiceProcess}
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
@@
 Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.   
+\martin{Nao entendo muito o algoritmo proposto aqui e o exemplo eh vago.}
+
 %
@@
 {\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
+\martin{E as outras constraints?}
 
@@
 %
-Details of these transformations can be found in~\cite{SouzaNeto:2012}.
+Details of these transformations can be found in~\cite{placidoPhDThesis2012}.
 
@@
 % _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
+
+\martin{Nesta secao e na seguinte tem transformacoes de constraints into apolicies e o inverso fica muito confuso.}
 
@@
 ServiceActivity:name}) are transformed into (named) composite
-operations that defines a (named) workflow of the application.
+operations that define a (named) workflow of the application.
 Composition patterns expressed using  {\em merge, decision, fork} and {\em join} are transformed into their corresponding workflows of the $\pi$-PEWS model.
@@
 
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
+To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.\martin{Is this a case study or a proof of concept?}
 Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
@@
 The values exchanged between ORCA and the brokers are:
-(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
-(ii) \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires;
-(iii) \textsl{Evaluation Reports} for the client companies;
+\textit{(i)} \textsl{Questionnaire and Evidences} filled with information about the client company;
+\textit{(ii)} \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires;
+\textit{(iii)} \textsl{Evaluation Reports} for the client companies;
 and
-(iv) the risk assessment \textsl{fee}.
+\textit{(iv)} the risk assessment \textsl{fee}.
 %
@@
 
-\begin{figure*}[h]
+\begin{figure}[h]
 \centering
-\includegraphics[width=0.45\textwidth]{./figures/3ValueModel.pdf}
+\includegraphics[width=1\textwidth]{./figures/3ValueModel.pdf}
 %\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}
-\caption{E3value model for \FlyingPig.\label{fig:E3valuemodel}}
-\end{figure*}
+\caption{E$^3$value model for \FlyingPig.\label{fig:E3valuemodel}}
+\end{figure}
 
@@
 Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.
-
-\begin{figure*}[h]
+They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}.
+
+\begin{figure}[h]
 \centering
-\includegraphics[width=0.6\textwidth]{./figures/BPMN_GCP.pdf}
+\includegraphics[width=1\textwidth]{./figures/BPMN_GCP.pdf}
 \caption{BPMN model for \FlyingPig.\label{fig:BPMNmodel}}
-\end{figure*}
+\end{figure}
 
@@
 The \pisodm models of the PIM level  for the \FlyingPig\ scenario are presented next.
+These models were manually obtained from the CIM level models.
 
@@
 %%
-Each $\pi$-UseCase may be associated to (non-functional) constraints (coloured ovals in Fig.~\ref{fig:piUseCaseModel}).
+Each $\pi$-UseCase may be associated to (non-functional) constraints (coloured ovals in Fig.~\ref{fig:piUseCaseModel}).\martin{colored?}
 Three types of constraints are defined: \textit{value}, \textit{business} or \textit{exceptional behavior}.
@@
 
-\begin{figure*}[h]
+\begin{figure}[h]
 \centering
-\includegraphics[width=0.55\textwidth]{./figures/UseCaseGeneral.png}
+\includegraphics[width=1\textwidth]{./figures/UseCaseGeneral.png}
 \caption{$\pi$-UseCase model for \FlyingPig.\label{fig:piUseCaseModel}}
-\end{figure*}
+\end{figure}
 

### 2016-11-29 JournalPISODM2016/submitted/main.tex
journal version

--- 2016-04-12 JournalPISODM2016/main.tex
@@
 
-\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
-\newcommand{\imartin}[1]{\textcolor{blue}{\textbf{Martin: }#1}}
-
-\newcommand{\umberto}[1]{\marginpar{\textcolor{orange}{\textbf{Umberto: }#1}}}
-\newcommand{\iumberto}[1]{\textcolor{orange}{\textbf{Umberto: }#1}}
+%\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
+%\newcommand{\imartin}[1]{\textcolor{blue}{\textbf{Martin: }#1}}
+%
+%\newcommand{\umberto}[1]{\marginpar{\textcolor{orange}{\textbf{Umberto: }#1}}}
+%\newcommand{\iumberto}[1]{\textcolor{orange}{\textbf{Umberto: }#1}}
 
@@
 
-\LRH{K. Belhajjame et~al.}
+\LRH{  }%K. Belhajjame et~al.}
 
@@
 
-%
-\authorA{Khalid Belhajjame}
-%
-\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
-  E-mail: kbelhajj@googlemail.com}
-%
-%
-\authorB{Valeria~de~Castro}
-\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-  E-mail: valeria.decastro@urjc.es}
-\authorC{Umberto~Souza~da~Costa}
-\affC{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-  E-mail: umberto@dimap.ufrn.br}
-
-\authorD{Javier~A.~Espinosa-Oviedo}
-\affD{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-  E-mail: javiera.espinosa@gmail.com}
-
-\authorE{Martin~A.~Musicante}
-\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-  E-mail: mam@dimap.ufrn.br}
-
-\authorF{Pl\'acido~A.~Souza~Neto}
-\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
-  E-mail: placido.neto@ifrn.edu.br}
-
-\authorG{Genoveva~Vargas-Solar}
-\affG{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
-  E-mail: genoveva.vargas@imag.fr}
-
-\authorH{Jos\'e-Luis~Zechinelli-Martini}
-\affH{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-  E-mail: joseluis.zechinelli@udlap.mx}
-
-%
-%\authorA{Fusheng Wang\footnote{Work done while working at Siemens Corporate Research.} }
-%
-%\affA{Department of Biomedical Informatics, Emory University
-%\newline
-%36 Eagle Row, Ste 589, Atlanta, GA 30322, USA}
-%
-%
-%
-%\authorB{\footnotesize Cristobal Vergara-Niedermayr\footnote{Work done while working at Siemens Corporate Research.}}
-%\affB{Oracle \newline
- %New Jersey, USA}
-%
+%%
+%\authorA{Khalid Belhajjame}
+%%
+%\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
+%  E-mail: kbelhajj@googlemail.com}
+%%
+%%
+%\authorB{Valeria~de~Castro}
+%\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
+%  E-mail: valeria.decastro@urjc.es}
+%\authorC{Umberto~Souza~da~Costa}
+%\affC{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+%  E-mail: umberto@dimap.ufrn.br}
+%
+%\authorD{Javier~A.~Espinosa-Oviedo}
+%\affD{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
+%  E-mail: javiera.espinosa@gmail.com}
+%    
+%\authorE{Martin~A.~Musicante}
+%\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+%  E-mail: mam@dimap.ufrn.br}
+%
+%\authorF{Pl\'acido~A.~Souza~Neto}
+%\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
+%  E-mail: placido.neto@ifrn.edu.br}  
+%
+%\authorG{Genoveva~Vargas-Solar}
+%\affG{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
+%  E-mail: genoveva.vargas@imag.fr}  
+%
+%\authorH{Jos\'e-Luis~Zechinelli-Martini}
+%\affH{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
+%  E-mail: joseluis.zechinelli@udlap.mx}
+%
+%
+%%
+%%\authorA{Fusheng Wang\footnote{Work done while working at Siemens Corporate Research.} }
+%%
+%%\affA{Department of Biomedical Informatics, Emory University
+%%\newline
+%%36 Eagle Row, Ste 589, Atlanta, GA 30322, USA}
+%%
+%%
+%%
+%%\authorB{\footnotesize Cristobal Vergara-Niedermayr\footnote{Work done while working at Siemens Corporate Research.}}
+%%\affB{Oracle \newline
+% %New Jersey, USA}
+%%
 %
@@
 
-\begin{bio}
-Khalid Belhajjame\dots Bio here.
-\vs{9}
-
-\noindent Valeria~de~Castro\dots Bio here.
-\vs{9}
-
-\noindent Umberto Costa received the BS degree in Computer Science from Federal University of Rio Grande do Norte in 1998. 
-He obtained the MS degree in Systems and Computation from Federal University of Rio Grande do Norte in 2000. 
-He received the PhD degree in Computer Science from University Federal de Minas Gerais in 2005. He is currently an associate professor in the Department of 
-Informatics and Applied Mathematics (DIMAp) at Federal University of Rio Grande do Norte. His academic interests include Programming Languages, 
-Service-Oriented Computing and Cloud Computing.
-\vs{9}
-
-\noindent Javier~A.~Espinosa-Oviedo\dots Bio here.
-\vs{9}
-
-\noindent Martin A. Musicante received his B.Sc at ESLAI, Argentina, in 1988 and
-his M.Sc and PhD in Computer Science at Universidade Federal de
-Pernambuco, Brazil, in 1990 and 1996, respectively. He is an associate
-professor at Universidade Federal do Rio Grande do Norte-UFRN (Natal,
-Brazil). He is part of the graduate program in Computer Science at
-UFRN and holder of a research grant from CNPq (level 2). Martin was
-associate researcher at LI - Universit\'e Fran\c{c}ois Rabelais Tours
-(2002-2011) and at LIFO - Universit\'e d'Orl\'eans, France, since 2008.
-Martin A. Musicante's main current research topics are big data
-processing languages, Web Services and cloud computing.
-\vs{9}
-
-\noindent Pl\'acido~A.~Souza~Neto\dots Bio here.
-\vs{9}
-
-\noindent Genoveva~Vargas-Solar\dots Bio here.
-\vs{9}
-
-\noindent Jos\'e-Luis~Zechinelli-Martini\dots Bio here.
-\end{bio}
+%\begin{bio}
+%Khalid Belhajjame\dots Bio here.
+%\vs{9}
+%
+%\noindent Valeria~de~Castro\dots Bio here.
+%\vs{9}
+%
+%\noindent Umberto Costa received the BS degree in Computer Science from Federal University of Rio Grande do Norte in 1998. 
+%He obtained the MS degree in Systems and Computation from Federal University of Rio Grande do Norte in 2000. 
+%He received the PhD degree in Computer Science from University Federal de Minas Gerais in 2005. He is currently an associate professor in the Department of 
+%Informatics and Applied Mathematics (DIMAp) at Federal University of Rio Grande do Norte. His academic interests include Programming Languages, 
+%Service-Oriented Computing and Cloud Computing.
+%\vs{9}
+%
+%\noindent Javier~A.~Espinosa-Oviedo\dots Bio here.
+%\vs{9}
+%
+%\noindent Martin A. Musicante received his B.Sc at ESLAI, Argentina, in 1988 and
+%his M.Sc and PhD in Computer Science at Universidade Federal de
+%Pernambuco, Brazil, in 1990 and 1996, respectively. He is an associate
+%professor at Universidade Federal do Rio Grande do Norte-UFRN (Natal,
+%Brazil). He is part of the graduate program in Computer Science at
+%UFRN and holder of a research grant from CNPq (level 2). Martin was
+%associate researcher at LI - Universit\'e Fran\c{c}ois Rabelais Tours
+%(2002-2011) and at LIFO - Universit\'e d'Orl\'eans, France, since 2008.
+%Martin A. Musicante's main current research topics are big data
+%processing languages, Web Services and cloud computing.
+%\vs{9}
+%
+%\noindent Pl\'acido~A.~Souza~Neto\dots Bio here.
+%\vs{9}
+%
+%\noindent Genoveva~Vargas-Solar\dots Bio here.
+%\vs{9}
+%
+%\noindent Jos\'e-Luis~Zechinelli-Martini\dots Bio here.
+%\end{bio}
 
@@
 
-Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010,SouzaNeto201684}.
+Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010}.%,SouzaNeto201684}.
 As Ameller et al.
@@
 
-\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue\martin{Placido: Mudar as cores...
-} meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
+\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
+%\martin{Placido: Mudar as cores... } 
+ meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
 We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
@@
 \textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
-Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{Placido2010LTPD}.
-%
-PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
+Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03}.% or PEWS~\cite{Placido2010LTPD}.
+%
+PEWS~\cite{BHM06}%,Placido2010LTPD}
+ is a notation to express service compositions.
 The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
@@
 Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
-transformed into a  {\sf Fork node}~\cite{valeriaThesis,placidoPhDThesis2012}.
+transformed into a  {\sf Fork node}~\cite{valeriaThesis}.%,placidoPhDThesis2012}.
 If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
 Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.   
-\martin{Nao entendo muito o algoritmo proposto aqui e o exemplo eh vago.}
+%\martin{Nao entendo muito o algoritmo proposto aqui e o exemplo eh vago.}
 
@@
 {\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
-\martin{E as outras constraints?}
+%\martin{E as outras constraints?}
 
@@
 %
-Details of these transformations can be found in~\cite{placidoPhDThesis2012}.
+%Details of these transformations can be found in~\cite{placidoPhDThesis2012}.
 
@@
 
-\martin{Nesta secao e na seguinte tem transformacoes de constraints into apolicies e o inverso fica muito confuso.}
+%\martin{Nesta secao e na seguinte tem transformacoes de constraints into apolicies e o inverso fica muito confuso.}
 
@@
 
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.\martin{Is this a case study or a proof of concept?}
+To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.%\martin{Is this a case study or a proof of concept?}
 Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
@@
 %%
-Each $\pi$-UseCase may be associated to (non-functional) constraints (coloured ovals in Fig.~\ref{fig:piUseCaseModel}).\martin{colored?}
+Each $\pi$-UseCase may be associated to (non-functional) constraints (coloured ovals in Fig.~\ref{fig:piUseCaseModel}).%\martin{colored?}
 Three types of constraints are defined: \textit{value}, \textit{business} or \textit{exceptional behavior}.

### 2017-09-29 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex
initial version 2 + answers to referees

\documentclass[12pt,a4wide]{article}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb} 
%\usepackage{xcolor}
\usepackage{color,soul}

%%%% OUR MACROS %%%%
\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
\newcommand{\imartin}[1]{\textcolor{magenta}{\textbf{Martin: }#1}}

\newcommand{\umberto}[1]{\marginpar{\textcolor{blue}{\textbf{Umberto: }#1}}}
\newcommand{\iumberto}[1]{\textcolor{blue}{\textbf{Umberto: }#1}}

\newcommand{\placido}[1]{\marginpar{\textcolor{purple}{\textbf{Placido: }#1}}}
\newcommand{\iplacido}[1]{\textcolor{purple}{\textbf{Placido: }#1}}
%%%%%%%%%%%

\title{$\pi$SOD-M: Building SOC Applications in the Presence of Non-Functional Requirements\\[3mm]
\textit{\large (List of changes and answers to the reviewers' observations)}}
%\author{Pl\'acido A. Souza Neto \and Genoveva Vargas-Solar \and Umberto Souza da Costa \and Martin A. Musicante}
\date{\today}

\begin{document}

\maketitle

\paragraph*{Editor's remarks:}
\begin{quotation}\sf\footnotesize

We have received the review reports for your paper "pisodm: Building SOC Applications in the Presence of Non-Functional Requirements".

We require now that you implement in your submission the following recommendations made by the reviewers:

\end{quotation}

\noindent 
All the comments have been taken into account and they are addressed in the paper. 
Our answers to the reviewers (below) explain how this has been done.

The text included in the new version is \hl{highlighted} for a better identification of the changes to the document.

\paragraph*{Reviewers' comments:}

\paragraph*{Reviewer A:} 
\begin{quotation}\sf\footnotesize
\noindent 
First of all, on the form: \\
1.	Footnotes do not appear in the paper. 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

2.	On page 21, title 5.2.4 you have a missing reference. 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

3.	On page 23, second paragraph, you have an incomplete sentence: `` \dots are implemented according a the knowledge\dots ''
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

\noindent On the content:

1.	On Section 2, the related works are cited as such. This section lucks of an evaluation of these works, and of the motivation of your proposed method. 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

2.	This method can be fully automated? Is it appropriate to non-IT professionals? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

3.	While designing the application to develop using your models, how is the selection of services done? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

4.	The paper validates the method with an industrial use case, but it lacks of a real evaluation that shows how the proposed method improves the existing once. For instance, evaluating the satisfaction of the users of the method, and comparing their satisfaction with the use of another development method. 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

5.	How the integration and the description of the constraints (Non-functional properties) improves the development process and the developed application? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

6.	These non-functional requirements can include other types of constraints? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

7.	On page 13, you specify that the model can be automatically translated into actual computer programs. Is this actually developed? If not, how is that can be possible? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\paragraph*{Reviewer B:} 
\begin{quotation}\sf\footnotesize

Changes which must be made before publication: 
+ Footnotes do not appear 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

+ Some references are not inserted [?] 
\end{quotation}

\noindent 
Fixed in the new version (Section 5.2.4).

\begin{quotation}\sf\footnotesize

+ Page 4, section 3.1.1: what is "value activities"? operations on values? 
\end{quotation}

\noindent 
We added a footnote explaning the value activity concept (according to~\cite{Gordijn02valuebased}).

\begin{quotation}\sf\footnotesize

+ Page 5 , paragraph following items: e3 value value model... 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

+ Page 22 paragraph on the bottom of the page: a too long sentence; I do not understand anything in this paragraph. 
\end{quotation}

\noindent 
We have rewritten the paragraph.

\begin{quotation}\sf\footnotesize

+ Page 23: service composition instead of services composition 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

+ The method is an extension of SOD-M, but I was wondering why the new name is piSOD-M. If an explanation exists, it would be interesting to indicate (since there is nothing to do with pi-calculs sometimes used in the WS world) 
\end{quotation}

\noindent 
There is no specific motivation behind the name. 
It just sounded appropriate for an extension of the existing SOD-M approach.

\begin{quotation}\sf\footnotesize

+ Section 6 does not mention any future work (except on its title!)

\end{quotation}

\noindent 
We have added a final paragraph describing directions of future work.

\section*{Final remarks}

We would like to acknowledge the anonymous referees for their useful and constructive critics to our paper.

\bibliographystyle{plain}
 \bibliography{biblio} 
\end{document}

### 2017-09-29 JournalPISODM2016/SubmittedV2/main.tex
initial version 2
initial version 2 + answers to referees

--- 2016-11-29 JournalPISODM2016/submitted/main.tex
@@
 \usepackage{subfigure}
+
+\usepackage{color,soul} %for color text and highlighting
 
@@
 
-\KEYWORD{MDA , Non-Functional Requirements, Service-based software process.}
+\KEYWORD{MDA, Non-Functional Requirements, Service-based software process.}
 
@@
 combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional}\footnote{Functional properties of a computer system are characterized by the effects it produces when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
+The selection of services is usually guided by the \textit{functional}\footnotemark \footnotetext{\hl{Functional properties of a computer system are characterized by the effects it produces when given a defined input.}} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
 An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
@@
 These models are successively refined from abstract specifications into actual computer programs. 
-A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
-Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.
+A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnotemark \footnotetext{\hl{The study shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
+Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}}.
 
@@
 
-The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
+The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{\hl{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}}.
 In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
@@
 The e$^3$value model  identifies the value/in\-for\-ma\-tion  exchange between system components (see Figure~\ref{fig:CIM:tpme3v}).
-This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
+This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes)\footnotemark \footnotetext{\hl{According to}~\cite{Gordijn02valuebased}\hl{, a value activity is defined as an activity expected to be profitable to one or more actors.}} performed by business actors (squared boxes).
 The model is well suited to enhance the understanding of the environment in which an application is being developed.
@@
 The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
-The e$^3$value value model shows Spotify and a private application (which is also a service) that provide services for listening and publishing information about music being listened by users. 
+The e$^3$\hl{value model} shows Spotify and a private application (which is also a service) that provide services for listening and publishing information about music being listened by users. 
 The private application interacts with Spotify for obtaining information about the flow of music being listened by a user in return of a fee for a premium subscription. 
@@
 However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
-In order to face this shortcoming, our method proposes the use of the BPMN meta-model as a tool for modelling this aspect of the application.
+In order to face this shortcoming, our method proposes the use of the BPMN meta-model as a tool for modeling this aspect of the application.
 
@@
 
-BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}.
+BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow\footnotemark \footnotetext{\hl{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}}.
 The next example illustrates the use of this meta-model.
@@
 (Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}\footnote{We use the {\sc small caps} font for referring to classes of a meta-model.}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}\footnotemark \footnotetext{\hl{We use the 
+%{\sc 
+small caps
+%} 
+font for referring to classes of a meta-model.}}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
 An {\sc End Consumer} is represented by an {\sc Actor}.
@@
 Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
-%Thus, the services functions can be specified as being grouped into packages.
+%Thus, the service functions can be specified as being grouped into packages.
 
@@
 A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
-Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint})\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} use case.
+Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint})\footnotemark \footnotetext{\hl{We use the 
+%{\sf 
+sans serif
+%}
+ font for referring to classes defined using a meta-model.}} use case.
 There are three kinds of constraints:
@@
 Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example.
-We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the \textit{(i)} conditions imposed by services usage -- for example, the fact that both Facebook and Twitter require authentication in order to call their methods for updating the wall; \textit{(ii)}  conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
+We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the \textit{(i)} conditions imposed by service usage -- for example, the fact that both Facebook and Twitter require authentication in order to call their methods for updating the wall; \textit{(ii)}  conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).
 
@@
 In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
-provides meta-classes to represent workflows%\footnote{Workflows will be transformed into (implemented) service compositions.} 
+provides meta-classes to represent workflows%\footnotemark \footnotetext{\hl{Workflows will be transformed into (implemented) service compositions.}} 
  that model  business processes.
@@
 
-\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
+\item In order to represent constraint types associated to \hl{service compositions}, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
 %\martin{Placido: Mudar as cores... } 
@@
 In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
-% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
+% \footnotemark \footnotetext{\hl{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}}).
 The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
@@
 \textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
-Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03}.% or PEWS~\cite{Placido2010LTPD}.
+Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03,ws-bpel2}.% or PEWS~\cite{Placido2010LTPD}.
 %
@@
 \begin{compactitem}
-\item 	We  used the Eclipse Modeling Framework (EMF)\footnote{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
+\item 	We  used the Eclipse Modeling Framework (EMF)\footnotemark \footnotetext{\hl{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
 From these meta-models, we  developed plug-ins to support their graphical representation.
 
-\item	 We used  ATL\footnote{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
+\item	 We used  ATL\footnotemark \footnotetext{\hl{http://eclipse.org/atl/.
+An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}}
 for implementing the  mappings between models.
 
-\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
+\item 	We  used Acceleo\footnotemark \footnotetext{\hl{http://www.acceleo.org/pages/home/en}} to implement  the code generation plug-in to generate executable code. 
 It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
@@
 
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.%\martin{Is this a case study or a proof of concept?}
+To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{\hl{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}}.%\martin{Is this a case study or a proof of concept?}
 Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
@@
 The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}.
+Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnotemark \footnotetext{\hl{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
+They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}}.
 
@@
 The  PSM of our case study is given in Figure~\ref{fig:piPEWSFlyingPig}. This model is obtained by transforming the $\pi$-service composition model into a workflow.
-Note that this workflow is implemented by using  BPEL constructors~\cite{BPEL} and \textit{policy} notions.
+Note that this workflow is implemented by using  \hl{BPEL constructors}~\cite{ws-bpel2} and \textit{policy} notions.
 
@@
 Recall that an application logic refers to routines that perform the activities to reach the application objective.
-Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
-These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
-%
-The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
-API does not necessarily export information about methods dependency (e.g., in the REST protocol);
-they do not share their state (stateless).
+Also there are non-functional properties derived from NFR. 
+They refer to strategies to be considered for the application execution such as security, isolation, adaptability, atomicity, and more.
+These non-functional properties must be ensured at execution time.
+
+\hl{In the context of service-oriented applications, ensuring non-functional properties is challenging owing to the nature of their components.
+Those components are defined by APIs which do not necessarily export information about their internals.}
 
@@
 As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, we specify policies for a services composition in an orthogonal way.
-Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
+In contrast to these approaches, we specify policies for a \hl{service composition} in an orthogonal way.
+Besides, these approaches suppose that non-functional properties are implemented \hl{according to the knowledge} that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
 In our approach,  the policies defined for a given application  can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
@@
 
+\hl{In the future, we expect to:}
+\textit{(i)} \hl{select metrics for assessing the efficacy of} \pisodm;
+%
+\textit{(ii)} \hl{use the feedback provided by our industrial partner to improve the methodology;}
+%
+\textit{(iii)} \hl{assess the use of} \pisodm \hl{in other contexts, in order to investigate the generality of the approach. }
+
 \bibliographystyle{plain}

### 2017-10-04 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex
Answer for the questions 1-3 and 6-7 to the Reviewer A.

@@
 \noindent 
-\hl{TO BE COMPLETED.}
-
+We added 2 paragraphs at the end of Section 2: (i) considering the evaluation of the related work; and (ii) also motivating our proposed method. 
 \begin{quotation}\sf\footnotesize
@@
 \noindent 
-\hl{TO BE COMPLETED.}
+We highlighted at the beginning of section 4, that the method is semi automatic, and we added a comment in as footnote explaining the reasons the method is not completely automatic.
 
@@
 \noindent 
-\hl{TO BE COMPLETED.}
+We highlighted in section 3.2.2, how the service is chosen.
+To be chosen and executed the service must fit the contract (constraints).
+If the service fits the contract it will used in the execution flow.
 
@@
 \noindent 
-\hl{TO BE COMPLETED.}
+As the non-functional requirements concept used in our method are related to business rules (constraints / contract rules) associated to the behavior of the application (services), it is possible include different types of constraints.
+The constraints are defined as pre- and post-conditions relates to services.
+So, during the design process, it is possible to include different kind of contracts related to services.
+Figure 4 ($\pi$-UseCase Meta-Model.) presents the Policy View Concepts used to design the constraints.
 
@@
 \noindent 
-\hl{TO BE COMPLETED.}
+Yes, it is automatically translated into $\pi$-PEWS \cite{MPC08,BaCAM05,BHM06,BHM06rep} specification programs.
+It is possible because of the PIM to PSM transformation rules described in sections 4.3 and 4.4 (using Acceleo specification programs) and also described in \cite{placidoPhDThesis2012}.
 

### 2017-10-04 JournalPISODM2016/SubmittedV2/main.tex
Answer for the questions 1-3 and 6-7 to the Reviewer A.

@@
 Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
-Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.
+
+\hl{
+To the best of our knowledge, and based on the related work analysis, there are no proposals that define a service-oriented approach for the whole development of systems, considering
+non-functional requirements. Although the efforts that have been made to support the new
+technological proposals for the Web such as web services, in general, these
+development approaches focus their processes on traditional software
+engineering, with emphasis on the functional aspects of the
+application.} Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.
+\hl{Our approach targets to deal with characteristics of the 3
+areas\footnote{(i) Non-Functional Requirements; (ii) Methodology for Service-Oriented Development;
+and (iii) Model-Driven Engineering} in order to propose a method for
+developing service-oriented applications considering non-functional aspects.
+
+The motivation of our method is to target the specification of non-functional aspects in the early stages of development of service-oriented  applications.
+Our approach specifies \textit{policies} for modeling the non-functional requirements of a service separating the specification of non-functional requirements from the main functionalities of the application.
+This is interesting because once the policies for a given application have been defined they can be reused and/or specialized for another application with similar requirements.
+}
 
@@
 \includegraphics[width=1\textwidth]{./figures/UseCaseMetaModel.pdf}
-\caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
+\caption{\label{fig:CIM:usecasemetamodel} $\pi$-\hl{UseCase Meta-Model}.}
 \end{figure*}
@@
 
-This model shows the set of logically related activities to be performed in a service-based application.
-The activities of this model represent a behavior that is part of the business logic of the application.
+\hl{This model shows the set of logically related activities to be performed in a service-based application.
+The activities of this model represent a behavior that is part of the business logic of the application.}
 A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
@@
 Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre- and post-conditions.
+\hl{The service that fits the {\sc Contract} and its {\sc Assertions} will be chosen and used in the execution flow.}
 
@@
 \pisodm  proposes intra-level transformation rules from $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules from $\pi$-service composition to $\pi$-PEWS models.
-These rules were implemented into a semi-automatic tool (the \pisodm  plugin for Eclipse).
+\hl{These rules were implemented into a semi-automatic tool} (the \pisodm  plugin\footnotemark \footnotetext{\hl{At each stage of transformation it is possible to include new constraints.
+In the more abstract stages of transformation there is complete automation in the models (i.e., from $\pi$-UseCase to $\pi$-ServiceProcess, and from $\pi$-ServiceProcess to $\pi$-ServiceComposition).
+On the other hand, to the transformation to specific platforms (i.e.
+from $\pi$-ServiceComposition to $\pi$-PEWS), a review by the analyst is required after the generation of the specification code.
+Thus, the method is more appropriate for those who have prior knowledge of Information Technology.}} for Eclipse).
 The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level.
@@
 
-\item 	We  used Acceleo\footnotemark \footnotetext{\hl{http://www.acceleo.org/pages/home/en}} to implement  the code generation plug-in to generate executable code. 
+\item 	\hl{We  used Acceleo}\footnotemark \footnotetext{\hl{http://www.acceleo.org/pages/home/en}} \hl{to implement  the code generation plug-in to generate executable code.} 
 It takes a $\pi$-PEWS model and generates the code to be executed by the {\em

### 2017-10-06 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex
Questions 4 and 5

@@
 \newcommand{\iplacido}[1]{\textcolor{purple}{\textbf{Placido: }#1}}
+
+\newcommand{\genov}[1]{\marginpar{\textcolor{red}{\textbf{Genoveva: }#1}}}
+\newcommand{\igenov}[1]{\textcolor{red}{\textbf{Genoveva: }#1}}
+
 %%%%%%%%%%%
@@
 To be chosen and executed the service must fit the contract (constraints).
-If the service fits the contract it will used in the execution flow.
+If the service fits the contract it will be used in the execution flow.
 
@@
 \noindent 
-\hl{TO BE COMPLETED.}
+\hl{We intend to evaluate our solution as suggested by the reviewer in our future work.
+That said, it is worth mentioning that the effectiveness of the solution proposed was showcased in the paper using a real-world application that underlines the features of our solution.
+In page 24 Section 5.3 we discussed about the importance of the use case for validating the methodology and about the perspectives of evaluating (i) time to market when a new module will be designed and implemented in the system (ii) maintenability of the new module having a separated and clear vision of business logic and non-functional properties.}
 
@@
 \noindent 
-\hl{TO BE COMPLETED.}
+\hl{ We believe that across the paper we have shown that decoupling  the description (specification) of non functional properties from the functional aspects, provides developers with the means to focus on the logic of their application.
+This isolates them from the nitty-gritty aspects of non-functional aspects.
+Furthermore, it improves modularity by allowing the same application to be run under different non-functional settings, if needed.
+}
 

### 2017-10-06 JournalPISODM2016/SubmittedV2/main.tex
Questions 4 and 5

@@
 
+%**********************************************************************************
 \section{Modeling Reliable Service Compositions with \pisodm}\label{sec:motivation}
+%**********************************************************************************
 
@@
 
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - 
 \subsection{Computation Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . -
 
@@
 
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   
 \subsubsection{E$^3$value Meta-Model}
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
 
@@
 
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
 \subsubsection{BPMN Meta-Model}
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
 
@@
 
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - 
 \subsection{Platform Independent Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . -
 
@@
 
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-UseCase} Meta-Model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
+ % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
+\subsubsection{\textit{$\pi$-UseCase} Meta-Model}
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
 
@@
 
- % .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceProcess} Meta-Model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
+\subsubsection{\textit{$\pi$-ServiceProcess} Meta-Model}
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
 
@@
 \hl{This model shows the set of logically related activities to be performed in a service-based application.
-The activities of this model represent a behavior that is part of the business logic of the application.}
+The activities of this model represent a behavior that can  implement the business logic of the application.}
 A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
@@
 Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre- and post-conditions.
-\hl{The service that fits the {\sc Contract} and its {\sc Assertions} will be chosen and used in the execution flow.}
+\hl{The service that fulfils the {\sc Contract} and its {\sc Assertions} will be chosen and used in the execution flow.}
 
@@
 
-% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-\subsubsection{\textit{$\pi$-ServiceComposition} Meta-Model}% .
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
-.
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
+\subsubsection{\textit{$\pi$-ServiceComposition} Meta-Model}
+% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .
 
@@
 
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - 
 \subsection{Platform Specific Models}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . -
 
@@
 
+% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
 \subsection{Implementation}
+% _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ . _ .
+
 We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
@@
 
+%*********************************************************************************
 \section{Applying \pisodm: The \FlyingPig\ Case Study}
 \label{sec:flyingPig}
+%*********************************************************************************
 
@@
 
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -.fig:E3valuemodel . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - 
 \subsection{Computation-Independent Models (CIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . -
 
@@
 
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - 
 \subsection{Platform-Independent Models (PIM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . -
 
@@
 
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 \subsubsection{$\pi$-UseCase Model for \FlyingPig}
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 
@@
 
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 \subsubsection{$\pi$-ServiceProcess Model for \FlyingPig}
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 
@@
 
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 \subsubsection{$\pi$-ServiceComposition Model for \FlyingPig}
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 
@@
 
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
-%\subsection{Platform-Specific Model (PSM)}
-%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 \subsubsection{$\pi$-PEWS Model for \FlyingPig}
+%. .  .  .  .  .  .  .  .  .  .  .  .   .  .  .  .  .  .  .  .  .  .  .  .
 
@@
 
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - 
 \subsection{Lessons Learned}
+%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . -
 
@@
 
-\hl{In the context of service-oriented applications, ensuring non-functional properties is challenging owing to the nature of their components.
+\hl{In the context of service-oriented applications, ensuring non-functional properties is challenging due to the nature of their components.
 Those components are defined by APIs which do not necessarily export information about their internals.}
@@
 
+\hl{Wrapping up, the industrial use case presented here  provided a real validation context for testing the usability of our methodology.
+The software units acting as services providig specific functions and constraints that were combined to build the business logic, were representative of the kind of target systems of our methodology.
+The different types of business rules could be  modelled as first class citizens and clearly assciated to the business logic of the system.
+Thereby, we could show  how non-functional properties can be associated both to the components (i.e., services)  and to the global business logic of the system.
+The methodology allowed to derive these non-functional properties thanks to the specialization and transformation of the different meta-models.
+
+We learned and showed that decoupling the description (specification) of non functional properties from the functional aspects, allows developers  to focus on the logic of their application, insulating them from the nitty-gritty aspects of non-functional aspects.
+Furthermore, it improves modularity by allowing the same application to be run under different non-functional settings, if needed.
+This can be interesting if business rules change or if constraints associated to services evolve.
+
+Finally, the system FlyingPig has been incrementaly built using classis software engineering techniques.
+It is difficult to compare the initial time to market versus the time to market using our methodology.
+The company could not estimate a realistic value of this measure.
+We intend to evaluate the quality of experience of programmers, when they will have to build a new module.
+The quality of experience will be given by measuring time to market and maintenability which are the two measures important to the company.
+}
+
+%****************************************************************************
 \section{Conclusions and Future Work}\label{sec:conclusions}
+%****************************************************************************
+
 We presented \pisodm, a model-driven method for designing and developing reliable service-based applications.

### 2017-10-06 JournalPISODM2016/SubmittedV2/main.tex
Questions 4 and 5 and other typos solved

@@
 \begin{abstract}
-Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
-The late inclusion of NFRs in the development may compromise the quality of the deployed application.
+Specifying non-functional requirements (NFRs) is a complex task, being usually addressed during the latter phases of the software development process.
+The late inclusion of NFRs during the software development  may compromise the quality of the resulting application.
 This paper presents \pisodm, a method and associated tools that
-\textit{(i)}  allows the early specification of NFRs in a principled way: users are abstracted away from low level details;
-\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
+\textit{(i)}  allow the early specification of NFRs allowing users to them in an abstract way without having to  care about low level details;
+\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible for reducing the specification and programming effort of the business application logic and its associated NFRs.
 Our solution has been used in the context of an industrial and real case study.
@@
 combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional}\footnotemark \footnotetext{\hl{Functional properties of a computer system are characterized by the effects it produces when given a defined input.}} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
+The selection of services is usually guided by the \textit{functional}\footnotemark \footnotetext{\hl{Functional properties of a computer system are characterized by the effects they produce in the behavior of the system given a specific input.}} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
 An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
@@
 The literature emphasizes the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
-Existing approaches assert that the convergence of model-driven soft\-ware development, service orientation  and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
+Existing approaches note that the convergence of model-driven soft\-ware development, service orientation  and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
 Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
@@
 
-Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
+Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support the specification of non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
 Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
@@
 \begin{compactenum}[i]
-\item To improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
-\item To reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
+\item To improve the development process by providing an abstract view of the application and helping to ensure the conformance to its specification.
+\item To reduce the programming effort through the semi-automatic generation of  models for the application,  to produce concrete implementations from higher-level models.
 \item To provide a general way of describing non-functional requirements.
@@
 In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
-The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.
+The models presented here were generated as a result of interacting with software developers at GCP Global and trying to fulfil  all the non-functional requirements present in the application.
 
@@
 
+%******************************************************************************
 \section{Related Work}
 \label{sec:relworks}
+%******************************************************************************
+
 While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
@@
 
-Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010}.%,SouzaNeto201684}.
+Besides the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010}.%,SouzaNeto201684}.
 As Ameller et al.
-realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
+notes in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}.
 Those authors also remark the lack of industrial use cases~\cite{Ameller201542}.
@@
 \hl{
-To the best of our knowledge, and based on the related work analysis, there are no proposals that define a service-oriented approach for the whole development of systems, considering
-non-functional requirements. Although the efforts that have been made to support the new
-technological proposals for the Web such as web services, in general, these
-development approaches focus their processes on traditional software
-engineering, with emphasis on the functional aspects of the
-application.} Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.
-\hl{Our approach targets to deal with characteristics of the 3
+To the best of our knowledge, and based on the related work analysis, there are no proposals that define a service-oriented approach for the whole development process of systems, considering
+non-functional requirements. Despite the efforts  made to support the new
+technological proposals for the Web, such as web services,  these
+development approaches address their development processes according to traditional software
+engineering, with emphasis on the functional aspect of the
+application.}
+
+Our method for developing service-oriented applications aims to overcome  limitations of existing  proposals by dealing with NFRs.  \hl{Our approach targets the characteristics of  3
 areas\footnote{(i) Non-Functional Requirements; (ii) Methodology for Service-Oriented Development;
-and (iii) Model-Driven Engineering} in order to propose a method for
+and (iii) Model-Driven Engineering} to propose a method for
 developing service-oriented applications considering non-functional aspects.
-
 The motivation of our method is to target the specification of non-functional aspects in the early stages of development of service-oriented  applications.
-Our approach specifies \textit{policies} for modeling the non-functional requirements of a service separating the specification of non-functional requirements from the main functionalities of the application.
+Our approach specifies \textit{policies} for modeling the non-functional requirements of a service.
+It separates the specification of non-functional requirements from the main functionalities of the application.
 This is interesting because once the policies for a given application have been defined they can be reused and/or specialized for another application with similar requirements.
+In order to show these characteristics this paper introduces a validation use case, which correponds to  a service based application provided  by the company ORCA.
+The application is related to the risk assessment industry.
 }
@@
 The notion of \textit{policy} is used to describe NFRs.
-(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
+Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.
 The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}\footnotemark \footnotetext{\hl{We use the 
@@
 In the more abstract stages of transformation there is complete automation in the models (i.e., from $\pi$-UseCase to $\pi$-ServiceProcess, and from $\pi$-ServiceProcess to $\pi$-ServiceComposition).
-On the other hand, to the transformation to specific platforms (i.e.
+On the other hand, in the case of the transformation to specific platforms (i.e.
 from $\pi$-ServiceComposition to $\pi$-PEWS), a review by the analyst is required after the generation of the specification code.
-Thus, the method is more appropriate for those who have prior knowledge of Information Technology.}} for Eclipse).
+Thus, the method is more appropriate for those who have background knowledge in Information Technology.}} for Eclipse).
 The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level.
@@
 
-\hl{Wrapping up, the industrial use case presented here  provided a real validation context for testing the usability of our methodology.
-The software units acting as services providig specific functions and constraints that were combined to build the business logic, were representative of the kind of target systems of our methodology.
-The different types of business rules could be  modelled as first class citizens and clearly assciated to the business logic of the system.
+\hl{Wrapping up, the industrial use case presented here  provided a real validation context for testing the usability of our method $\pi$SOD-M.
+The software units acting as services provide specific functions and constraints that were combined to build the business logic of the FlyingPig application.
+This application is  representative of the kind of target systems of $\pi$SOD-M.
+The different types of business rules could be  modelled as first class citizens and clearly associated to the business logic of the system.
 Thereby, we could show  how non-functional properties can be associated both to the components (i.e., services)  and to the global business logic of the system.
-The methodology allowed to derive these non-functional properties thanks to the specialization and transformation of the different meta-models.
-
-We learned and showed that decoupling the description (specification) of non functional properties from the functional aspects, allows developers  to focus on the logic of their application, insulating them from the nitty-gritty aspects of non-functional aspects.
+$\pi$SOD-M allowed to derive these non-functional properties thanks to the specialization and transformation of the different meta-models.
+
+We learned and showed that decoupling the description (specification) of non functional properties from the functional aspects, allows developers  to focus on the logic of their application, isolating them from the   non-functional aspects.
 Furthermore, it improves modularity by allowing the same application to be run under different non-functional settings, if needed.
@@
 
-Finally, the system FlyingPig has been incrementaly built using classis software engineering techniques.
-It is difficult to compare the initial time to market versus the time to market using our methodology.
-The company could not estimate a realistic value of this measure.
-We intend to evaluate the quality of experience of programmers, when they will have to build a new module.
-The quality of experience will be given by measuring time to market and maintenability which are the two measures important to the company.
+Finally, the system FlyingPig had been incrementally built using classic software engineering techniques before applying $\pi$SOD-M.
+It is difficult to compare the initial time to market versus the time to market using our method.
+The company could not estimate a realistic value of this measure for the first version of the application.
+We intend to evaluate the quality of experience of programmers, when they will have to build a new version of FlyingPig.
+The quality of experience related to $\pi$SOD-M will be given by measuring time to market and maintainability which are the two measures important to the company.
 }

### 2017-10-14 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex
Valeria's comments

@@
 \noindent 
-We highlighted at the beginning of section 4, that the method is semi automatic, and we added a comment in as footnote explaining the reasons the method is not completely automatic.
+We highlighted at the beginning of section 4 (page 15), that the method is semi automatic. We explain the reasons the method is not completely automatic.
 
@@
 That said, it is worth mentioning that the effectiveness of the solution proposed was showcased in the paper using a real-world application that underlines the features of our solution.
-In page 24 Section 5.3 we discussed about the importance of the use case for validating the methodology and about the perspectives of evaluating (i) time to market when a new module will be designed and implemented in the system (ii) maintenability of the new module having a separated and clear vision of business logic and non-functional properties.}
+In page 24 Section 5.3  we discussed about the importance of the use case for validating the methodology and about the perspectives of evaluating (i) time to market when a new module will be designed and implemented in the system (ii) maintainability of the new module having a separated and clear vision of business logic and non-functional properties.}
 
@@
 Furthermore, it improves modularity by allowing the same application to be run under different non-functional settings, if needed.
+We explained this on page 24 paragraph starting with "We learned ..."
 }
@@
 \noindent 
-As the non-functional requirements concept used in our method are related to business rules (constraints / contract rules) associated to the behavior of the application (services), it is possible include different types of constraints.
+As the non-functional requirements concept used in our method is related to the following types of constraints (page 9): (i) business rules (constraints / contract rules) associated to the behaviour of the application (services); (ii) Value Constraint related to the data types associated to attributes (like in the relational model); (iii) exceptional behaviour associated to the behaviour of the application when an exception is produced.
+
+%it is possible include different types of constraints.
 The constraints are defined as pre- and post-conditions relates to services.
 So, during the design process, it is possible to include different kind of contracts related to services.
+
 Figure 4 ($\pi$-UseCase Meta-Model.) presents the Policy View Concepts used to design the constraints.
@@
 \noindent 
-Yes, it is automatically translated into $\pi$-PEWS \cite{MPC08,BaCAM05,BHM06,BHM06rep} specification programs.
+%Yes, it is automatically translated into $\pi$-PEWS \cite{MPC08,BaCAM05,BHM06,BHM06rep} specification programs.
 It is possible because of the PIM to PSM transformation rules described in sections 4.3 and 4.4 (using Acceleo specification programs) and also described in \cite{placidoPhDThesis2012}.
+
+ Yes, at this time the $\pi$-PEWS model can be automatically translated into a $\pi$-PEWS executable program (this is already implemented using Acceleo platform).
 
@@
 \noindent 
-There is no specific motivation behind the name. 
-It just sounded appropriate for an extension of the existing SOD-M approach.
+The motivation behind the name has nothing to do with $\pi$ - calculus which is indeed used for formalizing service coordinations.
+$\pi$ comes from Policy which is the central notion to represent  NFR's and the way they are enforced at runtime.
+As you have seen, NFR are at some point implemented by sets of Event - Reaction rules.
+In the domain of active databases sets of rules could define a Policy.
+$\pi$ stands of the pronunciation of the first letter of the word "policy" which is /p/.
+The explanation of the name appears in a footnote of page 1 as explanation, the first time the name $\pi$-SOD-M appears in the text.
+
+%It just sounded appropriate for an extension of the existing SOD-M approach.
 

### 2017-10-14 JournalPISODM2016/SubmittedV2/main.tex
Valeria's comments

@@
 The late inclusion of NFRs during the software development  may compromise the quality of the resulting application.
-This paper presents \pisodm, a method and associated tools that
+This paper presents \pisodm \footnote{\hl{The letter $\pi$  of the name of the method comes from Policy which is the central notion to represent  NFR's and the way they are enforced at runtime.
+$\pi$ stands of the pronunciation of the first letter of the word "policy" which is /p/.}}, a method and associated tools that
 \textit{(i)}  allow the early specification of NFRs allowing users to them in an abstract way without having to  care about low level details;
@@
 Figure~\ref{fig:CIM:usecasemetamodel} shows the \textit{$\pi$-UseCase} meta-model.
-This model is  to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
+This model is  used to give the first representation of the application in terms of functionality, and of its non-func\-tion\-al requirements.
 The notion of \textit{policy} is used to describe NFRs.
@@
  The {\sc Business Collaborator} concept is represented  by {\sc Packages}.
-A {\sc Business Collaborator} denotes an external service or a system that interacts with the application that is being modeled.
+A {\sc Business Collaborator} denotes an external service or a system that interacts with the application that is being modelled.
 Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
@@
  font for referring to classes defined using a meta-model.}} use case.
-There are three kinds of constraints:
+
+\hl{There are three kinds of constraints:}
 \textit{(i)} A data pair ({\sc Value Constraint}), represented as the stereotype ``value'';
 \textit{(ii)} Business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
-\textit{(iii)} An {\sc Exceptional Behavior} constraint, represented by the stereotype {\sf ``exceptional\_behavior''}.
+\textit{(iii)} An {\sc Exceptional Behaviour} constraint, represented by the stereotype {\sf ``exceptional\_behavior''}.
 The use of these constraints is shown in the next example.
@@
 \pisodm  proposes intra-level transformation rules from $\pi$-use case to $\pi$-service process models, $\pi$-service process to $\pi$-service composition  models, as well as inter-level transformation rules from $\pi$-service composition to $\pi$-PEWS models.
-\hl{These rules were implemented into a semi-automatic tool} (the \pisodm  plugin\footnotemark \footnotetext{\hl{At each stage of transformation it is possible to include new constraints.
-In the more abstract stages of transformation there is complete automation in the models (i.e., from $\pi$-UseCase to $\pi$-ServiceProcess, and from $\pi$-ServiceProcess to $\pi$-ServiceComposition).
-On the other hand, in the case of the transformation to specific platforms (i.e.
-from $\pi$-ServiceComposition to $\pi$-PEWS), a review by the analyst is required after the generation of the specification code.
-Thus, the method is more appropriate for those who have background knowledge in Information Technology.}} for Eclipse).
+
+\hl{These rules were implemented into a semi-automatic tool, the $\pi$-SOD-M  plug in for Eclipse. At each stage of transformation it is possible to include new constraints.} 
+\hl{Transformation in the more abstract stages is completely automatic  (i.e., from $\pi$-UseCase to $\pi$-ServiceProcess, and from $\pi$-ServiceProcess to $\pi$-ServiceComposition).
+In contrast, the transformation to specific platforms (i.e.
+from $\pi$-ServiceComposition to $\pi$-PEWS), the analyst must intervene and check the generation of the specification code because there are programming details that can be interpreted in different manners depending on the target platform.
+In this sense, the method is more appropriate for those who have background knowledge in Information Technology.}.
 The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level.
@@
 We intend to evaluate the quality of experience of programmers, when they will have to build a new version of FlyingPig.
-The quality of experience related to $\pi$SOD-M will be given by measuring time to market and maintainability which are the two measures important to the company.
+The quality of experience related to $\pi$SOD-M will be given by measuring (i) time to market when a new module will be designed and implemented in the system (ii) maintainability of the new module having a separated and clear vision of business logic and non-functional properties.
+Both two measures  are important to the company.
 }

### 2017-10-15 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex
update the answer-to-the-reviewers latex document

@@
 \noindent 
-All the comments have been taken into account and they are addressed in the paper. 
+We have taken into consideration all the comments made by the reviewers and addressed in the new version of the paper. 
 Our answers to the reviewers (below) explain how this has been done.
@@
 We highlighted in section 3.2.2, how the service is chosen.
-To be chosen and executed the service must fit the contract (constraints).
-If the service fits the contract it will be used in the execution flow.
+In order for a service to be selected, it needs to fulfill contract (constraints).
+If a service fits the contract, then it will be used in the execution flow.
 
@@
 
-4.	The paper validates the method with an industrial use case, but it lacks of a real evaluation that shows how the proposed method improves the existing once.
+4.
+The paper validates the method with an industrial use case, but it lacks of a real evaluation that shows how the proposed method improves the existing once.
 For instance, evaluating the satisfaction of the users of the method, and comparing their satisfaction with the use of another development method.
@@
 \noindent 
-\hl{We intend to evaluate our solution as suggested by the reviewer in our future work.
+We intend to evaluate our solution as suggested by the reviewer in our future work.
 That said, it is worth mentioning that the effectiveness of the solution proposed was showcased in the paper using a real-world application that underlines the features of our solution.
-In page 24 Section 5.3  we discussed about the importance of the use case for validating the methodology and about the perspectives of evaluating (i) time to market when a new module will be designed and implemented in the system (ii) maintainability of the new module having a separated and clear vision of business logic and non-functional properties.}
+In page 24 Section 5.3  we discussed about the importance of the use case for validating the methodology and about the perspectives of evaluating (i) time to market when a new module will be designed and implemented in the system (ii) maintainability of the new module having a separated and clear vision of business logic and non-functional properties.
 
@@
 \noindent 
-\hl{ We believe that across the paper we have shown that decoupling  the description (specification) of non functional properties from the functional aspects, provides developers with the means to focus on the logic of their application.
+Decoupling the description (specification) of non functional properties from the functional aspects, provides developers with the means to focus on the logic of their application.
 This isolates them from the nitty-gritty aspects of non-functional aspects.
@@
 We explained this on page 24 paragraph starting with "We learned ..."
-}
 
@@
 \noindent 
-As the non-functional requirements concept used in our method is related to the following types of constraints (page 9): (i) business rules (constraints / contract rules) associated to the behaviour of the application (services); (ii) Value Constraint related to the data types associated to attributes (like in the relational model); (iii) exceptional behaviour associated to the behaviour of the application when an exception is produced.
+The non-functional requirements concept used in our method caters for the following types of constraints (page 9): (i) business rules (constraints / contract rules) associated to the behaviour of the application (services); (ii) Value Constraint related to the data types associated to attributes (like in the relational model); (iii) exceptional behaviour associated to the behaviour of the application when an exception is produced.
 
@@
 
- Yes, at this time the $\pi$-PEWS model can be automatically translated into a $\pi$-PEWS executable program (this is already implemented using Acceleo platform).
+Yes, that is the case.The $\pi$-PEWS model can be automatically translated into a $\pi$-PEWS executable program. We implemented this feature using the Acceleo platform.
 
@@
 \noindent 
-We added a footnote explaning the value activity concept (according to~\cite{Gordijn02valuebased}).
+We added a footnote explaining the value activity concept (according to~\cite{Gordijn02valuebased}).
 
@@
 + The method is an extension of SOD-M, but I was wondering why the new name is piSOD-M.
-If an explanation exists, it would be interesting to indicate (since there is nothing to do with pi-calculs sometimes used in the WS world)
+If an explanation exists, it would be interesting to indicate (since there is nothing to do with pi-calculus sometimes used in the WS world)
 \end{quotation}
@@
 $\pi$ comes from Policy which is the central notion to represent  NFR's and the way they are enforced at runtime.
-As you have seen, NFR are at some point implemented by sets of Event - Reaction rules.
+As you have seen, NFR is at some point implemented by sets of Event - Reaction rules.
 In the domain of active databases sets of rules could define a Policy.
 $\pi$ stands of the pronunciation of the first letter of the word "policy" which is /p/.
-The explanation of the name appears in a footnote of page 1 as explanation, the first time the name $\pi$-SOD-M appears in the text.
+The explanation of the name appears in a footnote of page 1, the first time the name $\pi$-SOD-M appears in the text.
 
@@
 
-We would like to acknowledge the anonymous referees for their useful and constructive critics to our paper.
+We would like to thank the anonymous referees for their useful and constructive critics to our paper.
 

### 2017-10-16 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex
update the answer-to-the-reviewers latex document
Missing answers highlighted by Khalid

@@
 \noindent 
-We have taken into consideration all the comments made by the reviewers and addressed in the new version of the paper. 
+We have taken into consideration all the comments made by the reviewers, and addressed  them in the new version of the paper. 
 Our answers to the reviewers (below) explain how this has been done.
@@
 \noindent 
-We highlighted at the beginning of section 4 (page 15), that the method is semi automatic. We explain the reasons the method is not completely automatic.
+We highlighted at the beginning of section 4 (page 15), that the method is semi automatic.
+Transformation in the more abstract stages is completely automatic  (i.e., from $\pi$-UseCase to $\pi$-ServiceProcess, and from $\pi$-ServiceProcess to $\pi$-ServiceComposition).
+In contrast, the transformation to specific platforms (i.e.
+from $\pi$-ServiceComposition to $\pi$-PEWS), the analyst must intervene and check the generation of the specification code because there are programming details that can be interpreted in different manners depending on the target platform.
+In this sense, the method is more appropriate for those who have background knowledge in Information Technology.
+The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level..
 

### 2017-10-16 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex
Final Version 2 (submitted)

@@
 In this sense, the method is more appropriate for those who have background knowledge in Information Technology.
-The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level..
+The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level.
+At the end of section 3.2.2 we mention that the selection of services that match the specification requires human intervention.
 
@@
 If a service fits the contract, then it will be used in the execution flow.
+As it is noted in section 3.2.2, this part of the process is not automatic. 
+Choosing the right service depends on the verification of a property that is done manually in the current version of the tool.
 
@@
 \noindent 
-The motivation behind the name has nothing to do with $\pi$ - calculus which is indeed used for formalizing service coordinations.
+The motivation behind the name is not related to the $\pi$-calculus which is indeed used for formalizing service coordinations. 
 $\pi$ comes from Policy which is the central notion to represent  NFR's and the way they are enforced at runtime.
@@
 In the domain of active databases sets of rules could define a Policy.
-$\pi$ stands of the pronunciation of the first letter of the word "policy" which is /p/.
-The explanation of the name appears in a footnote of page 1, the first time the name $\pi$-SOD-M appears in the text.
+$\pi$ stands of the pronunciation of the first letter of the word ``policy'' which is /p/.
+The explanation of the name now appears in a footnote of page 2, the first time the name $\pi$-SOD-M appears in the text.
 

### 2017-10-16 JournalPISODM2016/SubmittedV2/main.tex
Missing answers highlighted by Khalid
Final Version 2 (submitted)

@@
 
-Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support the specification of non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
+Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support the specification of non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process\footnotemark.
+\footnotetext{\hl{The letter} $\pi$ \hl{stands of the pronunciation of the first letter of the word ``policy'' which is a central concept in the representation of Non-Functional Requirementsd.}}
 Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
@@
 Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre- and post-conditions. 
-\hl{The service that fulfils the {\sc Contract} and its {\sc Assertions} will be chosen and used in the execution flow.}
+\hl{The service that fulfills the {\sc Contract} and its {\sc Assertions} will be chosen and used in the execution flow.
+This part of the process requires the verification of some logic conditions.
+In the present version of the tool, this step requires human intervention.}
 
@@
 
+%******************************************************************************
 \section{Transformation Rules}\label{sec:mmrules}
+%******************************************************************************
 

### 2017-11-22 JournalPISODM2016/FinalVersionNov2017/latex-single-column/single column/sample (single col).tex
final version - init

%%%%%%%%%%%%%%%%%%%%%%
\documentclass{singlecol-new}
%%%%%%%%%%%%%%%%%%%%%%

\usepackage{natbib,stfloats}
\usepackage{mathrsfs}

\def\newblock{\hskip .11em plus .33em minus .07em}

\theoremstyle{TH}{
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}[lemma]{Theorem}
\newtheorem{corrolary}[lemma]{Corrolary}
\newtheorem{conjecture}[lemma]{Conjecture}
\newtheorem{proposition}[lemma]{Proposition}
\newtheorem{claim}[lemma]{Claim}
\newtheorem{stheorem}[lemma]{Wrong Theorem}
\newtheorem{algorithm}{Algorithm}
}

\theoremstyle{THrm}{
\newtheorem{definition}{Definition}[section]
\newtheorem{question}{Question}[section]
\newtheorem{remark}{Remark}
\newtheorem{scheme}{Scheme}
}

\theoremstyle{THhit}{
\newtheorem{case}{Case}[section]
}

\makeatletter
\def\theequation{\arabic{equation}}

%\JOURNALNAME{\TEN{\it Int. J. System Control and Information
%Processing,
%Vol. \theVOL, No. \theISSUE, \thePUBYEAR\hfill\thepage}}%
%
%\def\BottomCatch{%
%\vskip -10pt
%\thispagestyle{empty}%
%\begin{table}[b]%
%\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcr@{}}%
%\\[-12pt]
%Copyright \copyright\ 2012 Inderscience Enterprises Ltd. & &%
%\end{tabular*}%
%\vskip -30pt%
%%%\vskip -35pt%
%\end{table}%
%}
\makeatother

%%%%%%%%%%%%%%%%%
\begin{document}%
%%%%%%%%%%%%%%%%%

\setcounter{page}{1}

\LRH{F. Wang et~al.}

\RRH{Metadata Based Management and Sharing of Distributed Biomedical
Data}

\VOL{x}

\ISSUE{x}

\PUBYEAR{xxxx}

\BottomCatch

%\CLline

\PUBYEAR{2012}

\subtitle{}

\title{Metadata Based Management and Sharing of Distributed Biomedical Data}

%
\authorA{Fusheng Wang}
%
\affA{Department of Biomedical Informatics,\\ Emory University,\\ Atlanta, GA, USA \\
Fax: +1 \qquad E-mail: fusheng.wang@emory.edu}
%
%
\authorB{Fusheng Wang}
\affB{Department of Biomedical Informatics,\\ Emory University,\\ Atlanta, GA, USA \\
Fax: +1 \qquad E-mail: fusheng.wang@emory.edu}
%

%
%\authorA{Fusheng Wang\footnote{Work done while working at Siemens Corporate Research.} }
%
%\affA{Department of Biomedical Informatics, Emory University
%\newline
%36 Eagle Row, Ste 589, Atlanta, GA 30322, USA}
%
%
%
%\authorB{\footnotesize Cristobal Vergara-Niedermayr\footnote{Work done while working at Siemens Corporate Research.}}
%\affB{Oracle \newline
 %New Jersey, USA}
%
%
\authorC{Peiya Liu}
\affC{Department of Integrated Data Systems, Siemens Corporate
Research \newline 755 College Road East, Princeton 08540, USA}

\begin{abstract}
Biomedical research becomes reliant on multi-disciplinary,
multi-institutional collaboration, and data sharing is becoming increasingly important for researchers to reuse experiments, pool expertise and validate approaches. However, there are many hurdles for data sharing, including the unwillingness to share,  lack of flexible data model for providing context information for shared data,  difficulty to share syntactically and semantically consistent data across distributed institutions, and expensive cost to provide tools to share the data.  In our work,  we develop a Web-based collaborative biomedical data sharing platform {\em SciPort} to support biomedical data sharing across distributed organizations. SciPort provides a generic metadata model for researchers to flexibly customize and organize the data. To enable convenient data sharing, SciPort provides a central server based data sharing architecture, where data can be shared by one click through publishing metadata to the central server. To enable consistent data sharing, SciPort provides collaborative distributed schema management across distributed sites. To enable semantic consistency for data sharing, SciPort provides semantic tagging through controlled vocabularies. SciPort is lightweight and can be easily deployed for building data sharing communities for biomedical research.
\end{abstract}

\KEYWORD{Metadata; Scientific Data Management; Data Sharing;  Data Integration; Computer Supported Collaborative Work.}

\REF{to this paper should be made as follows: Rodr\'{\i}guez
Bol\'{\i}var, M.P. and Sen\'{e}s Garc\'{\i}a, B. (xxxx) `The
corporate environmental disclosures on the internet: the case of
IBEX 35 Spanish companies', {\it International Journal of Metadata,
Semantics and Ontologies}, Vol. x, No. x, pp.xxx\textendash xxx.}

\begin{bio}
Manuel Pedro Rodr\'iguez Bol\'ivar received his PhD in Accounting at
the University of Granada. He is a Lecturer at the Department of
Accounting and Finance, University of Granada. His research
interests include issues related to conceptual frameworks of
accounting, diffusion of financial information on Internet, Balanced
Scorecard applications and environmental accounting. He is author of
a great deal of research studies published at national and
international journals, conference proceedings as well as book
chapters, one of which has been edited by Kluwer Academic
Publishers.\vs{9}

\noindent Bel\'en Sen\'es Garc\'ia received her PhD in Accounting at
the University of Granada. She is a Lecturer at the Department of
Accounting and Finance, University of Granada. Her research
interests are related to cultural, institutional and historic
accounting and in environmental accounting. She has published
research papers at national and international journals, conference
proceedings as well as chapters of books.\vs{8}

\noindent Both authors have published a book about environmental
accounting edited by the Institute of Accounting and Auditing,
Ministry of Economic Affairs, in Spain in October 2003.
\end{bio}

\maketitle

 \section{Introduction}

%The needs for data sharing
With increased complexity of scientific problems, biomedical
research is increasingly a collaborative effort across multiple
institutions and disciplines.  Data sharing is becoming critical for
validating approaches and ensuring that future research can build on
previous efforts and discoveries. As a result, data sharing is often
required by scientific funding agencies to share the data produced
in grant projects. For example,  the National Institutes of Health
(NIH) of US requires data sharing for NIH funded projects of
\$500,000 or more in direct costs in any one year.

To support large scale collaborative biomedical research, NIH
provides large-scale collaborative project awards  for a team of
independently funded investigators to synergize and integrate their
efforts, and the awards mandate the research results and data to be
shared \cite{nih03datasharing,nih08policy}.  The Network for
Translational Research (NTR): Optical Imaging in Multimodality
Platforms\cite{ntr} is one of such collaborative projects on the
development, optimization, and validation of imaging methods and
protocols for rapid translation to clinical environments.  It
requires not only managing the complex scientific research results,
but also sharing the data across hundreds of research collaborators.
As another example, Siemens Healthcare has research collaborations
with hundreds of research sites distributed across the US, each
providing Siemens marketing support by periodically delivering white
papers, case reports, clinic methods, clinic protocols,
state-of-the-art images, etc. In the past, there were no convenient
methods for research partners to share data with Siemens, and mostly
data were delivered through media such as emails, CDs and hard
copies. This made it very difficult to organize, query and integrate
the shared data.

Sharing biomedical research data is important but difficult
\cite{piwowar08culture,birnholtz03data}. One major difficulty is the
unwillingness to share, as investigators may restrict access to data
to maximize the professional development and economic benefit
\cite{bekelman03}. With increased awareness of this issue, the
government and funding agencies are developing  strategies and
policies to promote and enforce the sharing of data
\cite{piwowar08culture,dsic,cabigconnected}. While the social issue
of data sharing is challenging, this paper will mainly focus on the
development of a biomedical data sharing system which makes data
sharing becoming flexible and easy for humans on collaborative
research.

\subsection{Complexity of Biomedical Data}

Biomedical data can be in heterogeneous formats such as structured
data, standard based medical images such as DICOM \cite{dicom}, non
standard medical images such as optical images, raw equipment data,
spreadsheets, PDF files,  XML documents, and many others. Thus, the
data are  a mix of structured data (often deeply hierarchical) and
files. Meanwhile, the data structures of biomedical research data
can be very complex, ranging from different primitive data types, to
lists and tables, from flat structures to deep nested structures,
and so on.  It is often difficult for investigators to create a
sophisticated schema that can capture enough context information or
metadata information, if the data types do not have a public,
centralized, and well-recognized database. During our interviews
with more than a dozen biomedical research investigators, we found
that major investigators use spreadsheet like tools and manage their
data through operating system folder organized files. To reduce the
cost and complexity for data sharing, a data sharing system needs to
provide an extensible architecture that can be easily customized by
researchers with their own metadata models, even without programming
skills.

\subsection{Management of Data Sharing}

Even with willingness to share data, in many cases an investigator
would like to control when to share data, and who can access the
data. An investigator may want to keep unpublished data only to his
or her own group or collaborative partners, and selectively publish
a subset of data for sharing. An investigator may also want to stop
sharing certain data at any time, for example, when he or she finds
a quality problem of the data. An investigator may also have new
results added on existing shared data and want to keep the data
current. A manageable data sharing system with convenient operations
can potentially increase the willingness for users to share, as the
effort for data sharing can be much reduced.

\subsection{Data Sharing Architecture}

There are three major types of data sharing architectures: i)
Centralized, multiple datasets hosted at a single location in a
common schema. For example, the Cancer Genome Atlas (TCGA) Data
Portal \cite{tcga} manages genome related data; National Biomedical
Imaging Archive \cite{nbia} manages DICOM based medical research
data. ii) Federated, with a virtual view of physically separate
datasets. For example, Cancer Biomedical Informatics Grid
(caBIG{\small \textregistered})  \cite{cabig} is a Grid based data
federation infrastructure that supports a CQL query language across
distributed data sources.  iii) Distributed, physically and
virtually separate datasets.  Centralized approach is often limited
to common data types. Biomedical research, however,  generates
complex data and often new data types.  Distributed approach is
often difficult to retrieve, interpret and aggregate results, and
lacks data consistency between research sites.  While caGrid is
becoming widely used in biomedical research community, caGrid itself
has a complex infrastructure and the effort is significant.

Considering the high dynamic nature of biomedical research and the
need of  cost effective data sharing, we develop a hybrid
architecture which combines the benefits of the centralized approach
and the distributed approach. In this approach, a data sharing
central server is provided for multiple distributed data sources,
and stores only published metadata (not raw data or images with
large sizes) from distributed data sources. Users can have flexible
management of data sharing through publishing or unpublishing data
with  a simple operation. The metadata contain context information
of original data sources (including raw data) which are still
managed at distributed research sites. The central server provides
an integrated view of all shared data, and shared data can be easily
aggregated and retrieved from the central server. This architecture
not only is lightweight, but also provides support of consistent
data sharing through collaboratively managing schemas and
semantically tagging of data.

\subsection{Inconsistency of Shared Data}

Another difficulty on sharing data is the incompatible
representation of data between sharing partners. Each site may use
different schemas (e.g., data templates) and represent data in
different formats; and even if multiple partners choose to use the
same schema, the schema may evolve as research progresses. For
example, a field name may be changed or additional constraints may
be added in a schema. Data generated from different versions of a
schema  may not match when searched together. For example, if a
field named ``age'' is changed to ``patient\_age'',  these two field
names will lead to structurally different fields and can not be
searched together as a single field.   Semantic consistency is on
using the same vocabulary to represent knowledge. For example,  one
investigator may use ``Malignant Tumor'' to describe a lesion, and
another investigator  may use `cancer'.    While semantics
annotations can be added to data to enhance semantic
interoperability between shared data, there is a lack of integrated
tools and architecture that can support semantic annotations from
controlled vocabularies.

\subsection{Our Contributions}

To meet the requirements of a data sharing system and overcome the
challenges, we develop SciPort, a Web-based collaborative biomedical
data sharing platform to support collaborative biomedical research.
SciPort brings together the following essential components to
support biomedical data sharing:

\begin{itemize}

\item Generic biomedical experiment metadata model and XML based biomedical data management for research sites to flexibly customize and organize their data (Section
\ref{sec:datamanagementoverview});

\item Hybrid, lightweight data sharing architecture through a Central
Server, and data sharing can be conveniently managed by investigators (Section \ref{sec:sharing});

\item Easy sharing of schemas between distributed research partners (Section \ref{sec:sharingschema});

\item Collaborative management of schemas and their evolution between distributed research sites to maintain syntactic data consistency  (Section \ref{sec:collaborativeschema});

\item Semantic tagging of data to achieve semantically consistent data sharing (Section \ref{sec:tag}).
\end{itemize}

This paper has been significantly extended from previous work
\cite{wang08com}, which discussed preliminary work and the technique
aspect on distributed scientific data management. The work presents
here represents latest results and software development, and has a
major new focus on the human and sharing aspects of biomedical data.
The paper also covers new work on metadata modeling for data
sharing, collaborative tagging for data sharing, user experience on
real world deployment, and so on.

\section{Generic Biomedical Experiment Metadata Modeling and Management} \label{sec:datamanagementoverview}

As discussed in \cite{birnholtz03data}, metadata models serve as the
abstraction of data, and are essential for understanding shared
data. However, metadata  are often incomplete, and important context
information of data \cite{chin04context} is often ignored. However,
many domain specific applications or data models are specific to
certain types of data and limited to a small set of context
information.  To precisely capture as sufficient metadata and
context information as possible for biomedical data sharing,
researchers themselves should be also to customize their own
metadata models easily to meet their research need and data sharing
need.

SciPort was first developed as a biomedical data management system  with
metadata  modeling, authoring, management, viewing, searching and
exchange.

%SciPort takes an XML based approach for data modeling, schema representation, and storage  and queries.

\subsection{Generic Meta Data Modeling for Biomedical Data}

To meaningfuly represent each dataset, we develop a {\em SciPort document} model to represent the metadata and context information of the data.
A SciPort Document can represent both (nested) structured data, files and images.

A SciPort document includes several objects: i) Primitive Data
Types/Fields. Primitive data types are used to represent structural
data, including {\em integer}, {\em float}, {\em date}, {\em text},
and Web-based data types such as {\em textarea}, {\em radiobutton},
{\em checkbox}, {\em URL}, etc.; ii) File. Files can be linked to a
document through the file object; iii) \textit{Reference.}  A
reference type links to another SciPort document; iv) Group. A group
is similar to a table, which aggregates a collection of fields or
nested groups.  There can be multiple instances for a group, like
rows of a table; v)Category. A category relates a list of fields,
e.g., ``patient data'' category, ``experiment data'' category, etc.
Categories are used only at the top level of the content, and
categories are not nested.

Figure \ref{fig:sampledocument} shows an example SciPort document
that describes image annotations on top of medical images.  This
document captures generic information such as the tile, description,
author, creation date and modification date; patient information
such as the age and gender; annotation data such as a number of
annotated tumors marked up on top of the image, and the link to the
image file from which annotations were generated.  Note that in the
example, we can easily represent nested complex information such as
multiple tumors, multiple spatial coordinates and different  data
types such as images and files.

\begin{figure*}
\caption{A Sample SciPort Document}\label{fig:sampledocument}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F1.eps}}
\end{figure*}

\subsection{XML Based Implementation and the Benefits}

The hierarchical nature of the data model fits perfectly with the
tree based XML data model, and we take an XML based approach to
implement the data model -- we call it \textit{SciPort Exchange
Document}.  Users can also easily define their own schemas which are
internally represented with an XML-based schema definition language.
Besides, To provide an intuitive way to present data,  we also
develop a hierarchical model based on XML to organize biomedical
data, thus documents can be quickly browsed and identified through
the hierarchy. For example, we can define a hierarchy with levels of
``site''$\rightarrow$  ``patient'' $\rightarrow$ ``measurement'',
and attach documents at different folder levels. SciPort also
provides fine grained access control  at folder level for the data
in the database \cite{wang09security}.

The self-describing and rich structure of XML makes it possible to
represent arbitrary complex biomedical research data. By modeling
the data as XML documents, we can take advantage of native XML
database technologies to manage biomedical data, thus we can avoid
complex data model and query translation between XML and RDBMS. This
is especially beneficial since users can define arbitrary structured
and nested data formats for their data.  Moreover, powerful queries
can be supported directly on XML databases with the standard XML
query language XQuery \cite{xquery}.

A salient characteristic of SciPort is that the system is highly
adaptable.  SciPort provides a Web-based schema authoring tool to
easily create complex hierarchical metadata schema  model without
any need of programming.  The organizational hierarchy is also
customizable with its hierarchy authoring tool. This allows users to
configure their own biomedical data repository without requiring
expensive and time-consuming services or software development
effort.

\section{Sharing Distributed Biomedical Data} \label{sec:sharing}

While investigators can share their data simply through giving login
information to collaborators, a more systematic approach for sharing
data across research consortia or networks can provide more
flexibility and benefits. SciPort comes with comprehensive data
sharing capabilities: i) Convenience: data sharing is performed by a
single action and data can be selectively shared; ii) Ownership of
data: researchers own and manage their data by their own;  iii)
Flexible sharing control: data sharing can also be revoked by
researchers at any time;  iv) Up-to-date of shared data. As data are
updated or removed, corresponding shared data also need to be
synchronized accordingly to stay current;  v) Consistently
aggregated shared data from distributed sites; and vi) Lightweight.
Sharing is manipulated through metadata, no copy of large volume
data is needed.

These sharing capabilities are implemented through a lightweight,
central server based approach, as discussed next.

\subsection{Sharing Data through a Central Server}

SciPort provides a distributed architecture to share and integrate
data through a Central Server (Figure \ref{fig:integration}). In
this architecture, each research site will have its own Local
SciPort Server which itself functions as an independent Server for
data collection and management. In addition, there will be an
additional Central  Server upon which Local Servers are able to
selectively publish their data (structured documents) (Figure
\ref{fig:docpublish}). Images/files, which are often the major
source of data volume, are still stored on corresponding Local
Servers but are linked from the published documents on the Central
Server. Once a user on the Central Server begins to download a
document from the Central Server, actual data files are downloaded
from the corresponding Local Server that holds the data.

Thus, the Central Server provides a global view of shared data across all
distributed sites, and can also be used as a hub for sharing schemas among
multiple sites. Since data are shared through the metadata (SciPort
Documents), the integration is lightweight. Users on the Central Server will
only have read access to the data.

Figure \ref{fig:integration} illustrates an example SciPort sharing
architecture formed by four Local Servers at four universities: UCI,
UCSF, Dartmouth and Penn. Each Local Server is used for data
collection and management of clinical trial data at its local
institution. Since these clinical trials are under the same research
consortium, they would share their data together by publishing their
data (documents) to the Central Server located at UCI. Members at
NTR research consortia are granted read access on such shared data
through the Central Server. Once the user identifies a data set from
the Central Server and wants to download the data, the user will be
redirected to the corresponding hosting Local Server to download the
data to the client.

\begin{figure*}[t]%2
\caption{The Central Server Based Architecture for Data
Sharing}\label{fig:integration}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F2.eps}}
\end{figure*}

\begin{figure*}[t]%3
\caption{An Example of Publishing an Existing Document}
\label{fig:docpublish}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F3.eps}}
\end{figure*}

\subsection{Data Synchronization}

Shared data may become outdated as new results or analysis are
generated on the data. SciPort provides automated data
synchronization on shared data through synchronization enforcement
in the following operations: i) Create. When a document is created,
the author or publisher has the option to publish this document
(Figure \ref{fig:editpublish}). Once the document is   published, a
``published'' status is added to the document. A user can   also set
up an automatic publishing flag so all new documents will be
automatically published; ii) Update. When an update is performed on
a published document, the document will automatically be republished
to the Central Server; iii) Delete. When a published document is
deleted, it will also be automatically removed from the Central
Server;  iv) Unpublish. A user can stop sharing a document by
selecting the ``unpublish'' operation.  Unpublished documents will
be removed from the Central Server.

\begin{figure*}[t]%4
\caption{An Example of Publishing a New Document}
\label{fig:editpublish}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F4.eps}}
\end{figure*}

\subsection{Security and Trust between Local Servers and the Central Server}

\textbf{Server Verification.} The trust between Local Servers and
the Central Server is implemented through security tokens. For a
Local Server to be accepted into the network, it will be granted a
security token to access the Central Server services. The token will
be imported at the setup step. When a Local Server tries to connect
to the Central Server, the Central Server verifies if the token
matches.

\textbf{Single Sign-on and Security.} One issue for sharing data
from distributed databases is that it is not feasible for Central
Server users to login to every distributed database. When a user
publishes a document, the user already grants the read access of the
document (including the files linked to the document) to the users
on the Central Server, thus another authentication is unnecessary.
Therefore, users on the Central Server should be able to
automatically access shared data from a Local Server in a
transparent way. To support this, the Local Server Document Access
Control Manager has to make sure that the remote download requests
really come from Central Server users who are currently logging on.
We develop a single sign-on method to guarantee the security of the
data sharing, by verifying if an incoming data request to a Local
Server comes from the Central Server and if the current request user
is currently logged in on the Central Server.

\subsection{Sharing Data in Multiple Data Networks}

Data can  be shared not only in a single data network through one
Central Server, but also in multiple data networks through multiple
Central Servers. One organization may want to share the same data in
multiple networks, as demonstrated in an example (Figure
\ref{fig:multiserverintegration}).  There are two networks,  one
centered at UCI and another centered at Stanford. UCI is
collaborating with both networks and needs to share data with both
networks.  UCI will be granted as a partner site and its Local
Server will be configured for both networks. A document can then be
published to any of the Central Servers or both.  This sharing
architecture makes it possible for very flexible data sharing across
different research networks.

\begin{figure*}[t]%5
\caption{Sharing Data in Multiple Data Networks}
\label{fig:multiserverintegration}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F5.eps}}
\end{figure*}

Besides data sharing, SciPort also provides sharing of schemas
between distributed research sites (Figure
\ref{fig:collaborativearchitecture}), discussed in Section
\ref{sec:sharingschema}.

\begin{figure*}[t]%6
\caption{Types of Information Sharing in SciPort}
\label{fig:collaborativearchitecture}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F6.eps}}
\end{figure*}

\subsection{The Benefits of Our Hybrid Data Sharing Architecture}

The architecture of SciPort is a hybrid of centralized approach and
distributed approach.  Centralized approach can provide high
visibility, easy retrieval, easy aggregation within the repository,
but suffers from limited data types or flexibility of customizing
new data types. Distributed approach allows users to flexibly
manage, customize and control their data and data sharing,  but
often suffers from low visibility, difficulty on retrieval,
interpretation, and aggregation, and lacks of data consistency. Our
hybrid approach benefits from both. The Central Server provides high
data visibility, easy retrieval and aggregation of all shared data
from distributed sites.   SciPort allows users to conveniently
customize their data types through creating and updating schemas.
Users can flexibly manage their data sharing through simple
publishing or unpulishing operations.  Data consistency is
maintained through collaborative schema sharing and management
(discussed in next two Sections). The hybrid approach is also
lightweight, as it does not store directly original data such as raw
data or medical images -- which can have high data volumes, but only
metadata or structured documents that describe and abstract the
data.

\section{Sharing Schemas} \label{sec:sharingschema}

Schemas are used to define structures and constraints of documents.
The former includes a mix of (possibly nested) object types defined
in the data model, and the latter includes i) number of instance
constraints for file and group types, ii) minimal and maximal value
constraints; and iii) controlled values.  Schemas are an essential
component  since they are used for i) data validation; ii) document
authoring form generation; iii) data presentation -- templates are
defined based on schemas; and iv) search form generation.

Sharing schemas are critical for sharing data, since the Central
Server glues data together using shared schemas to present and
search data. How to keep data from multiple Local Servers coherent
is also dependent on at what level and how schemas can be shared
consistently.

\subsection{Publishing Schemas}

Schemas can be shared by publishing them to the Central Server. From
a  Local Server {\em Schema Management} menu, schemas created on the
Local Server can be selectively published to target Central Servers,
as shown in the example in Figure \ref{fig:schemapublish}. When a
new document is being published to the Central Server, the
availability of the corresponding schema on the Central Server will
be checked. If the schema is not present, then the schema will also
be published together with the document.

The {\em owner of a schema} is defined as the Local Server on which
the schema is first created.  A schema is identified by its owner
and schema ID.  SciPort also provides comprehensive access control
management \cite{wang09security}, and two roles are related to
schema management: i) {\em organizer} role with privileges to author
and update schemas and ii) {\em publisher} role with  privileges to
publish documents and schemas.

Once a schema is published, it can be shared through the Central
Server. Other Local Servers can reuse schemas by importing schemas
from the Central Server, as shown in Figure \ref{fig:schemaimport}.
A schema can be unpublished from a Central Server by a Local Server,
thus the schema is not available on the Central Server for further
sharing.  Users on the Central Server with an ``organizer'' role can
remove a schema from the Central Server if no document on Central
Server is depending on this schema. This can be used to clean up
non-used schemas.

\begin{figure*}[t]%7
\caption{An Example of Publishing Schemas from a Local Server}
\label{fig:schemapublish}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F7.eps}}
\end{figure*}

\begin{figure*}[t]%8
\caption{An Example of Importing Schemas to a Local Server}
\label{fig:schemaimport}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F8.eps}}
\end{figure*}

\subsection{Three Scenarios of Schemas Sharing}

Based on the use cases, there are three typical scenarios of schema sharing:

\textbf{Static Schema} A schema is fixed and will not be changed.
For example, some common standard based schemas are not likely to
change. Once a schema is authored and changed to the final status,
it can be published to the Central Server to be shared by every
Local Server (Figure \ref{fig:fixeddistributedschema}). This is the
simplest scenario, as schemas can be created once and shared
directly through the Central Server.

\textbf{Uniform Evolving Schema} A Schema can be changed and a uniform
version is shared by all Local Servers, thus data consistency is maintained
across all sites.  The schema is owned by its original creating Local Server,
and the owner can make certain changes.

\textbf{Multiform Evolving Schema} A ``seed'' schema is first
created and shared as public -- every Local Server becomes the owner
and can update the schema. The Central Server maintains a version of
the schema that conditionally merges updates from Local Servers,
thus all documents published on the Central Server will be
compatible under this schema.

\begin{figure*}[t]%9
\caption{Static Schemas Sharing} \label{fig:fixeddistributedschema}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F9.eps}}
\end{figure*}

\begin{figure*}[t]%10
\caption{Uniform Schema Sharing} \label{fig:distributedschema}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F10.eps}}
\end{figure*}

\begin{figure*}[t]%11
\caption{Multiform Schema Sharing}
\label{fig:publicdistributedschema}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F11.eps}}
\end{figure*}

Next, we will discuss how to manage schemas and their evolution for the last
two scenarios (static schema management is straightforward and ignored here.)

\section{Collaborative Management of Schemas for Consistent Data Sharing}
\label{sec:collaborativeschema}

Since schemas can be shared and used across multiple distributed
data sources, one challenging issue is how to manage schema
evolution while keeping shared documents consistent and compatible
with their schemas in a distributed environment. To solve this
problem, we first define the following favorable rules for schema
management and sharing in a distributed environment:
\begin{itemize}

\item {\em Minimal administration}. Human based manual schema
management can be difficult, especially for schemas that may be
updated by different sites.  To minimize human effort for managing
schemas across distributed sites, it would be ideal if the data
sharing system can facilitate the management of schemas, for
example, relying on an information exchange hub  based on the
Central Server.

\item {\em Data consistency}.  Schema evolution has to be backward compatible otherwise the integrity of documents will be broken.

\item {\em Control of schemas}. To prevent arbitrary update of
schemas, by default, only the owner (the user who creates the
initial schema on a Local Server) of a schema can update a schema.

\item {\em Current of Schemas}: The Central Server always has the
up-to-date version of a schema if that schema is shared by the
owner, to guarantee no outdated schemas are shared across sites.

\item {\em Sharing Maximization}: Since shared schemas on the Central
Server can be unpublished by local database users and removed from
the Central Server, to promote sharing, only the last time publisher
of this schema can unpublish a schema.

\end{itemize}

These rules will help to automate schema management in a distributed
environment, minimize conflicts of updates, maintain coherent shared
data on the Central Server, and reduce the effort from humans.  To
achieve this, we enforce the rules for schema operations on both
Local Servers and Central Servers. Schema operations on Local
Servers include create, update, delete, import, publish, and
unpublish, and schema operations on Central Servers include remove.

\subsection{Uniform Schema Management}

In this scenario, a Schema can be changed and a uniform version is
shared by all Local Servers, thus data consistency is maintained
across all sites and the Central Server.   Next we discuss how
schema management rules are enforced in each schema operation.

\textbf{Create and Delete}

A schema can be created on a Local Server if the user has the
organizer role. This Local Server will become the owner of this
schema. A schema can be  deleted  if the user has the organizer
role, and there are no documents using this schema on this Local
Server.  If the Local Server is the owner of the schema, and the
schema is never published,  deleting a schema will eliminate the
schema forever. If the schema was once published, it may still be
alive and used on other Local Servers.

\textbf{Update}

Incompatible update of a schema is the one that can lead to
inconsistency between the new schema version and existing documents.
Incompatible updates are not allowed unless the following conditions
are met:  i) only the owner of a schema can make updates to the
schema; ii) there are no existing documents created based on  this
schema on this Local Server; and iii) the schema was never
published, i.e., there will no other Local Servers using this schema
to create documents.
%A publishing status is associated with each schema.

Compatible update of a schema will not lead to inconsistency between
the new schema version and existing documents. The following
conditions are required for compatible update.   i){\em Ownership.}
The Local Server is the owner of the schema, and the user is the
organizer on this Local Server; ii){\em Field Containment.}  All the
fields in the last schema are present in the  new schema and belong
to the same category or group, and new categories and fields added.
iii){\em Type Compatibility.} All the fields in the new schema have
the same type   or a compatible type, i.e., a more general type.
iv){\em Relaxed Value Range Constraints.} No new value constraints
are permitted  for  existing fields which do not have any
constraint. Value constraints can be updated with more relaxed
ranges. Constraints on new fields are permitted. v){\em Relaxed
Controlled Values.} For field with controlled values, the extent is
enlarged with more options. vi){\em Relaxed Constraints on Number of
Instances.} For group field or file field, there can be a constraint
on the minimal and/or maximal number of instances.  No instance
number constraint is permitted on existing fields, and instance
number constraint can be updated with more relaxed range. Change of
a field's order within its sibling is not considered incompatible.

Once a schema is updated on the Local Server, it will be automatically
republished to the Central Server (if any) onto which the schema has been
published. This will ensure the Central Server always maintains up-to-date
versions of schemas.

\textbf{Publish and Unpublish}

A schema can be published to one or multiple Central Servers if the
user has a publisher role and there is no newer version of this
schema on the Central Server.  There are three scenarios of schema
publishing: i) Schemas can be manually  published from the schema
publishing interfaces; ii) The schema of a document is automatically
published when a document is published. When a document is
published, the Local Server will check the Central Server if the
schema is available or up-to-date. Otherwise the schema is
republished; iii) If a schema was published and is then updated with
compatibility, the new version schema will be automatically
republished on the Central  Server and replace the last version.
This will keep the schemas on the  Central Server up-to-date.

A schema can be unpublished by a Local Server with the following conditions:
i) the user has the ``publish'' role on the Local Server; ii) there is no
document associated with it on the Central Server, and iii) the Local Server
is the last publisher of the schema. The last condition is necessary,
otherwise if the schema is used at multiple Local Servers, every Local Server
can easily stop the publishing which can be against the sharing goal of the
last publisher.

\textbf{Import}

A Local Server can import a shared schema from the Central Server if
the user has the organizer role. When there is a new version of a
schema on the Central Server, the new version schema will be
automatically detected by a Local Server when there is a document
being published from that Local Server.

\textbf{Remove from Central Server}

Users on the Central Server with the organizer role can remove a
schema from the Central Server if no document on Central Server is
using this schema. This can be used to cleanup non-used schemas.

\textbf{Update, or create?}  As a schema  keeps on evolving, the
number of updates on a schema can be many. When it reaches certain
threshold, the latest schema may be very different from the original
schema, the data between the two schemas may be very different, and
the data consistency does not make much sense any more.  In this
case, it may be desirable to create a new schema.  SciPort provides
a functionality to view change history of a schema, and users can
create a new schema from an existing schema in the Schema Editor
tool.

\textbf{An Example of Uniform Schema Management}

A user with organizer privilege on a Local Server L1 creates a
schema S(V1) (Figure \ref{fig:distributedschema}). The user may
later find the schema not accurate, and makes changes to the schema.
Since no document has been created and the schema has never been
shared yet, the user may make arbitrary changes, including
incompatible updates. Once the user has a stable usable version of
the schema, the user begins to author documents based on this
schema, and publishes documents to the Central Server C. Schema
S(V1) will be automatically published to the Central Server. After
some time, the user may need more information for their data or
adjust existing fields, and need to update schema S. Since there are
existing documents using the schema, and the schema was also
published, the user can make only compatible update to schema S(V1).
(If compatible update is not sufficient, the user has to create a
new schema.) Once schema S(V1) is updated as S(V2), it will be
automatically propagated to the Central Server to replace the last
version S(V1). A user with organizer privilege on a Local Server L2
imports schema S(V1) after S(V1) is published on the Central Server,
and documents then are created on L2 on this schema. Later S(V2)
replaces S(V1) on the Central Server, which is detected when a
document of schema S(V1) is published to C. The user at L2 will be
prompted to synchronize the schema, and S(V1) is replaced by S(V2)
on L2.

\subsection{Distributed Multiform Schema Sharing} \label{sec:multiformschema}

Uniform schema sharing provides data compatibility through a single
uniform version of schemas across all servers, and maintains the
ownership of schemas and provides controlled schema evolution. There
can also be cases that multiple sites need to adapt certain schema
to their own needs, and a uniform schema may not be feasible. To
provide flexible schema evolution at each Local Server and support
data compatibility on the Central Server, we provide a multiform
schema sharing approach.

As shown in Figure \ref{fig:publicdistributedschema}, a ``seed'' schema {\em
S} will be first created and made as public owned, and then published onto the
Central Server. Each Local Server will be able to import this schema, and
adjust the schema for its local use ({\em S$_{L1}$}, {\em S$_{L2}$}, etc.)
Schema updates from each site will be conditionally merged on the Central
Server as {\em S$_M$}. The goal of the merging  is to keep the schema version
on the Central Server mostly relaxed, through the following merging
conditions:

\begin{itemize}
  \item If the changes are incremental structural changes, they will be merged. These include
  adding of new fields or new categories;
  \item If the constraints on the Central Server are more permissive
   than the new or updated constraints that the modification suggests,
   then the suggested changes are ignored by the Central Server;
  \item Structural removal changes such as field removal or category removal
  will be ignored and not merged.
\end{itemize}

In this way, each Local Server will maintain its local schema evolution while
the Central Server will maintain a merged schema for shared data
compatibility.

\section{Collaborative Semantically Tagging of Data}\label{sec:tag}

Recently tagging has become a popular method to enable users to add
keywords to Internet resources, thus to improve search and personal
organization \cite{marlow06tag}. In SciPort, data are hierarchically
organized based on a tree based organizational structure. In
practice, it can be very helpful that users can provide additional
classification of data, by assigning tags from a controlled
vocabulary to documents. Each document can be flexibly annotated
with one or more semantic tags, and tags themselves can be shared by
users at both Local Servers and the Central Server.

Semantically tagging adds additional semantics to documents, which
not only provides semantics based query support, but also enhances
semantic interoperability of shared data from multiple distributed
sites.

\subsection{Tagging from Controlled Vocabulary}

Free tagging is used by most Web-based tagging systems, where users
can define arbitrary tags. One issue for free tagging is that since
there is no common vocabulary among these tags, there can be
semantic mismatch between tags.

Instead, SciPort chooses to annotate data using semantic tags coming
from predefined ontology or controlled vocabulary, such as NCI
Enterprise Vocabulary Services (EVS) \cite{ncievs}. By standardizing
tags using such controlled vocabulary, the system can then provide a
controlled  set of semantic tags. Thus data can be categorized into
multiple semantic groups, which makes it possible to express queries
based on common semantics. For instance, when a user authors a
document of a study on breast cancer,  he assigns  a tag
``Stage\_II\_Breast\_Cancer'' (ID: 18077), and another tag
``Cancer\_Risk'' (ID: 7768). A problem arises that expensive lookup
through vocabulary may simply prevent users from using it.

Next we show that with a collaborative tag management system, we can
provide automatic tag lookup in a controlled vocabulary repository
by caching previously retrieved tags in the Central Server with Ajax
technology.

\begin{figure*}[t] \center%12
\caption{Collaborative Tag Management} \label{collaborativeTags}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F12.eps}}
\end{figure*}

\begin{figure*}[t] \center%13
\caption{An Example of Automatic Tag Lookup} \label{taglookup}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F13.eps}}
\end{figure*}

\subsection{Collaborative Semantic Tag Management} \label{sec:coltag}

Caching tags is to exploit locality inherent to the subset of the
vocabulary that is used within a group of researchers. Since a
collaborative research consortium often focuses on solving a single
significant problem, the vocabulary is quite smaller than the
standardized vocabulary. For instance, the NCI Thesaurus
\cite{ncithesaurus} is 77MB in size.  In SciPort, we provide a
cached vocabulary repository on the Central Server, thus
previous-retrieved tags from the standardized vocabulary is shared
among all users.   Cached vocabulary on the Central Server makes it
very efficient to search for a tag in the vocabulary: instead of
searching for a tag at a remote large vocabulary all the time,
previous used tags in the tag cache repository can be searched
first.

Figure \ref{collaborativeTags} shows the architecture of
collaborative tag management. At each Local Server, there is a tag
repository that manages all tags on this Local Server; there is a
remote NCI controlled vocabulary database from which users can
dynamically search and retrieve tags. Once a tag is defined on the
Local Server, it will  be automatically cached at the Central Server
tag repository. When a user wants to associate a tag to its data, he
can dynamically search and select the tag, through automatic tag
lookup, as discussed next.

\subsection{Semantic Annotation of Documents with Tags}
By taking advantage of the Ajax technology, we provide automatic tag
lookup while a user wants to add a tag. As the user is typing in a
keyword on a web page to search for a tag, there will be an
automatic tag lookup from three resources: the local tag repository,
the cached/shared tag repository on the Central Server, and the NCI
controlled vocabulary. Implemented through Ajax, the lookup is
performed in the background, and retrieved tags are displayed in the
order of local tags, cached tags, and remote tags.  The lookup will
dynamically load a dropdown list of tags from which the user can
pick up (Figure \ref{taglookup}). By dynamically sending
asynchronous tag queries to tag repositories, users can immediately
select a desired tag to label his data, instead of opening multiple
browser windows to do separate searches.  With this technology, we
are able to support using tags from a controlled vocabulary or
shared repository by providing a convenient interface.

\subsection{Semantically Querying and Browsing Data through Semantic Tags}

Once documents are annotated with semantic tags, it is now possible
for users to semantically browse data by clicking on a semantic tag,
or use tagged fields as additional constraints for specifying
queries.  For example, a user might want to  search all documents
tagged with ``Stage\_II\_Breast\_Cancer'', and a type of ``Breast
Cancer'' can generate a drop down list of semantic tags and the user
can quickly use the right tag to specify queries.  Semantic
annotation can also help to classify data. Since semantic tags
coming from an ontology are hierarchically organized in the
ontology,  it is possible to generate an ontology tree based view of
all documents based on annotated ontology concepts on documents.

\begin{figure*}[t] \center %14
\caption{A Screenshot of a SciPort Local Server Deployed at
University of California, Irvine} \label{sciportfrontpage}
%\centerline{\epsffile{E:/Inderscience/LATEX-FILES/IJMSO/F14.eps}}
\end{figure*}

\section{Implementation}

\subsection{Software}

SciPort is built with J2EE and XML, running on Apache Tomcat servers
and Oracle Berkeley DB XML database server (open source) or IBM DB2
XML database. The system is OS neutral and has been tested on
Windows, Linux, and MacOS. It uses standard protocols, including
XML, XSLT, XPath and XQuery, and Web Services. SciPort is a
lightweight application and can be easily deployed and customized
by users.

SciPort is a rich Web-based application, thus it is possible for
users to use it at any place and at any time. SciPort supports major
Web browsers including Internet Explorer, Firefox, Opera and Safari.
Taking advantage of Web 2.0 technologies such as Ajax, SciPort
provides rich application capabilities with smooth user experience.

A salient characteristic of SciPort is that the system is highly
adaptable. Through customizable metadata schemas, hierarchy
organization, and sharing architecture, SciPort can be easily setup
for managing biomedical research data and providing a data sharing
network.

Figure \ref{sciportfrontpage} shows the screenshot of the front page
of a Local Server deployed at University of California, Irvine,
while the Central Server is located at Siemens Corporate Research
located at Princeton, New Jersey.

\subsection{Deployment}

SciPort was initially adopted by the Network for Translational
Research on Optical Imaging (NTROI) for collecting and sharing data
across several institutions (University of California, Irvine,
University of Pennsylvania, Datmouth, Stanford and University of
California, San Francisco. It is again adopted by the second cycle
of Network for Translational Research (Washington University,
University of Texas Health Science Center, Houston, University of
Michigan, and Stanford University).  SciPort is also used for
sharing pathology images and their annotations for a large scale
prostate cancer multi-modality study at University of Pennsylvania,
the State University of New Jersey, Rutgers, and Siemens Corporate
Research.  SciPort was also adopted for supporting Siemens
Healthcare for collecting and sharing large scale biomedical
research data from university and hospital research partners.

\section{Discussion}

The design and development of SciPort has been an iterative process
and driven by numerous discussions with biomedical researchers and
users. During the process, we have learned many lessons.

\textbf{Usability.}  A critical requirement for software tools from
biomedical researchers is the usability of the system. This includes
intuitive user interfaces and workflows for data authoring, sharing
and querying, and the ease of system setup. When the initial version
of SciPort was released, the pilot users complained about multiple
clicks needed to navigate through multiple web pages, and showed
strong desires of simpler interfaces and fast response. Especially,
a single click data publishing is desired, otherwise users gradually
get frustrated to share their data.   The Web based user interfaces
then evolved from static web pages (jumping from one page to
another) based interface to Ajax based interface, and then another
Flex based interface is developed to provide even smoother user
experience. Our conclusion is that smooth user experience of data
sharing tools could much improve the interest for researchers to
share their data.

\textbf{Generalization and Customization.} Although SciPort is
currently a software platform and is not tied to specific
applications, the initial version of SciPort (Version 1.0) was
designed and developed for one single project with static data
forms. Soon we realized that even for a single project, the schemas
were dynamic, and new data forms were occasionally required. And
researchers from other projects were using totally different data
forms. Besides, the data are heterogenous, ranging from structural
data to medical images, to  files, etc. This motivated us to
redesign SciPort as a platform that could provide easy and flexible
customization. This is challenging as schema evolution is very
difficult to support in traditional relational database management
systems based on tables. The emergence of native XML database
technologies makes it possible to support customizable schemas, as
native XML DBMSs have a significant advantage on supporting schema
evolution. Our data models, queries and data management are fully
based on XML.

\textbf{Integration with Other Systems.} In the past, we also worked
on extending SciPort so data managed in SciPort could be integrated
into other infrastructures such as caBIG. We developed a
SciPort-NBIA \cite{nbia} bridge to support such integration.

\section{Related Work}

Computer supported collaborative work has been increasingly used to
support research collaborations. A review of collaborative systems
is presented in \cite{bafoutsou2002281}, and a review of taxonomy of
collaboratory types is discussed \cite{zimmerman07taxonomy}. SciPort
belongs to the Community Data Systems based on this classification.
Lee et al review collaboratory concepts relevant for collaborative
biomedical research \cite{lee09} and analyze the major challenges.
In \cite{myneni10col}, a collaborative information management system
is discussed.  A context-based sharing system is proposed in
\cite{chin04context} to support more data-centric collaboration than
tools oriented ones, and this motivates our metadata oriented
approach for data modeling and sharing in SciPort.

With the increasing collaboration of scientific research,
collaborative cyberinfrastructures have been researched and
developed.  Grid-based systems(such as caBIG{\small \textregistered}
--cancer Biomedical Informatics Grid \cite{cabig}, Biomedical
Informatics Research Network (BIRN) \cite{birn}) provide
infrastructures to integrate existing computing and data resources.
They rely on a top down common data structures. The iPlant
Collaborative (iPC) \cite{pscic}  is a cyberinfrastructure project
recently funded by US NSF. iPlant has a more focus on the human side
of the infrastructure. myGrid \cite{mygrid} is a suite of tools for
e-Science, with a focus on workflows.   A review on
cyberinfrastructure systems for the biological sciences  is
discussed in \cite{stein08}.  Grid based systems are more used on
sharing computing and storage resources, P2P is more used on sharing
data \cite{foster03p2p2grid}.  MIRC \cite{mirc} is an example of P2P
based data system for authoring and sharing radiology  teaching
files.

While Grid based systems are more used on sharing computing and storage
resources, P2P is more used on sharing data \cite{foster03p2p2grid}.  MIRC
\cite{mirc} is a popular pure-P2P based system for authoring and sharing
teaching files.

A publish and subscribe architecture for distributed metadata management is
discussed in \cite{keidl02metadata}, which focuses on the synchronization
problems. In \cite{taylor06reconcile}, an approach of bottom-up collaborative
data sharing is proposed, where each group independently manages and extends
their data, and  the groups compare and reconcile their changes eventually
while tolerating disagreement. Our approach takes an approach in between the
bottom-up and top-down approaches, where each group manages their data, but
also achieves as much agreement on schemas as possible through controlled
schema evolution.

In \cite{rader08tag}, influence on tag choices is analyzed from
del.icio.us. Our approach focuses on semantic tagging from a
controlled vocabulary instead of free tags.  In \cite{pike07},
different approaches for representing scientific knowledge is
discussed.

Extensive work has been done in data integration and schema
integration \cite{halevy06integration,doan05semantic}. In
\cite{beynon97}, a collaborative schema integration system is
discussed for database design. Our system takes a proactive approach
where schema and data consistency is enforced during data authoring
and schema authoring.

\section{Conclusion}

Contemporary biomedical research is moving towards
multi-disciplinary, multi-institutional collaboration. These lead to
strong demand for tools and systems to share biomedical data. This
drives the development of SciPort -- a Web-based data sharing
platform for collaborative biomedical research. To support
meaningful abstraction of data, SciPort provides a generic metadata
model for users to conveniently define their own metadata schemas.
SciPort provides an innovative lightweight hybrid data sharing
architecture which combines the benefits of the centralized approach
and the distributed approach. Investigators are able to flexibly
manage their data and schema sharing, and quickly build data sharing
networks. To enable data consistency for data sharing, SciPort
provides comprehensive collaborative distributed schema management.
Through semantic tagging, SciPort further enhances semantic
interoperability of shared data from distributed sites.

%\section*{Acknowledgement}
%The project is funded in part by the National Institutes of Health, under Grant No. 5R01CA136535.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\bibliography{ijmso}
%\bibliographystyle{unsrt}
%\bibliographystyle{alpha}

\begin{thebibliography}{10}

\bibitem[\protect\citeauthoryear{Myneni and Patel}{2010}]{myneni10col}
Myneni, S. and  Patel, V.L. (2010) 'Organization of biomedical data
for collaborative scientific research: a research information
management system', {\it International Journal of Information
Management}, Vol. 30, No. 3, pp.256--264

\bibitem[\protect\citeauthoryear{Bekelman et al.}{2003}]{bekelman03}
Bekelman, J.E., Li, Y. and Gross, C.P. (2003) 'Scope and impact of
financial conflicts of interest in biomedical research: a systematic
review', {\it JAMA}, Vol. 289, No. 19, pp.454--65

\bibitem[\protect\citeauthoryear{Stein}{2008}]{stein08}
Stein, L.D. {2008} 'Towards a cyberinfrastructure for the biological
sciences: progress, visions and challenges', {\it Nature Reviews
Genetics}, Vol. 9, No. 9, pp.678--688.

\bibitem[\protect\citeauthoryear{NIH Statement on Sharing Scientific Research
Data}{http://grants.nih.gov/grants/policy/data\_sharing/}]{nih03datasharing}
NIH Statement on Sharing Scientific Research Data,
http://grants.nih.gov/grants/policy/data\_sharing/

\bibitem[\protect\citeauthoryear{Policy for sharing of data obtained in NIH supported or conducted
genome-wide association studies}{2008}]{nih08policy} Policy for
sharing of data obtained in NIH supported or conducted genome-wide
association studies,
http://grants.nih.gov/grants/guide/notice-files/not-od-07-088.html

\bibitem[\protect\citeauthoryear{Network for Translational Research (NTR): Optical Imaging in
Multimodality
Platforms}{http://imaging.cancer.gov/programsandresources/speci\break
alizedinitiatives/ntroi}]{ntr} Network for Translational Research
(NTR): Optical Imaging in Multimodality Platforms,
http://imaging.cancer.gov/programsandresources/specializedinitiatives/ntroi

\bibitem[\protect\citeauthoryear{Piwowar~et~al.}{2008}]{piwowar08culture}
Piwowar, H., Becich, M., Bilofsky, H. and Crowley, R. (2008) `PLoS
medicine', {\it Sept, No. 9, Towards a Data Sharing Culture:
Recommendations for Leadership from Academic Health Centers},
Vol.~5.

\bibitem[\protect\citeauthoryear{Birnholtz et al.}{2003}]{birnholtz03data}
Birnholtz, J.P. and Bietz, M.J. (2003) `Data at work: supporting
sharing in science and engineering', {\it GROUP}, pp.339--348.

\bibitem[\protect\citeauthoryear{Data Sharing \& Intellectual Capital (DSIC) Workspace}
{https://cabig.nci.nih.gov/working\_groups/DSIC\_SLWG}]{dsic} Data
Sharing \& Intellectual Capital (DSIC) Workspace,
https://cabig.nci.nih.gov/working\_groups/DSIC\_SLWG

\bibitem[\protect\citeauthoryear{Getting Connected with
caBIG}{https://cabig.nci.nih.gov/getting\_connected/}]{cabigconnected}
Getting Connected with caBIG, https://cabig.nci.nih.gov/\break
getting\_connected/

\bibitem[\protect\citeauthoryear{Digital Imaging and Communications in Medicine
(DICOM)}{http://medical.nema.org/}]{dicom} Digital Imaging and
Communications in Medicine (DICOM), http://medical.nema.org/

\bibitem[\protect\citeauthoryear{The Cancer Genome Atlas (TCGA) Data Portal}{http://cancergenome.nih.gov/dataportal}]{tcga}
The Cancer Genome Atlas (TCGA) Data Portal,
http://cancergenome.nih.gov/dataportal

\bibitem[\protect\citeauthoryear{National Biomedical Imaging
Archive}{https://cabig.nci.nih.gov/tools/NCIA}]{nbia} National
Biomedical Imaging Archive, https://cabig.nci.nih.gov/tools/NCIA

\bibitem[\protect\citeauthoryear{caBIG: cancer Biomedical Informatics Grid}{http://caBIG.nci.nih.gov/}]{cabig}
caBIG: cancer Biomedical Informatics Grid, http://caBIG.nci.nih.gov/

\bibitem[\protect\citeauthoryear{Biomedical Informatics Research Network}{
http://www.nbirn.net/}]{birn}
Biomedical Informatics Research Network, http://www.nbirn.net/

\bibitem[\protect\citeauthoryear{Cyberinfrastructure for the Biological Sciences: Plant Science
Cyberinfrastructure Collaborative
(PSCIC)}{http://www.nsf.gov/pubs/2006/nsf06594/nsf06594.htm}]{pscic}
Cyberinfrastructure for the Biological Sciences: Plant Science
Cyberinfrastructure Collaborative (PSCIC),
http://www.nsf.gov/pubs/2006/nsf06594/nsf06594.htm

\bibitem[\protect\citeauthoryear{MIRC}{http://mirc.rsna.org}]{mirc}
MIRC, http://mirc.rsna.org

\bibitem[\protect\citeauthoryear{Foster and
Iamnitchi}{2003}]{foster03p2p2grid} Foster, I. and Iamnitchi, A.
(2003) `On death, taxes, and the convergence of peer-to-peer and
grid computing', {\it IPTPS'03}.

\bibitem[\protect\citeauthoryear{Keidl et al.}{2002}]{keidl02metadata}
Keidl, M., Kreutz, A., Kemper, A. and Kossmann, D. (2002) `A publish
{\&} subscribe architecture for distributed metadata management',
{\it ICDE}.

\bibitem[\protect\citeauthoryear{Taylor and Ives}{2006}]{taylor06reconcile}
Taylor, N.E. and Ives, Z.G. (2006) `Reconciling while tolerating
disagreement in collaborative data sharing', {\it SIGMOD}.

\bibitem[\protect\citeauthoryear{Rader and Wash}{2008}]{rader08tag}
Rader, E.J. and Wash, R. (2008) {\it CSCW}, pp.239-248, Influences
on tag choices in del.icio.us.

\bibitem[\protect\citeauthoryear{Halevy et~al.}{2006}]{halevy06integration}
Halevy, A., Rajaraman, A. and Ordille, J. (2006) {\it VLDB, Data
integration: the teenage years},
http://portal.acm.org/citation.cfm?id=1182635.1164130

\bibitem[\protect\citeauthoryear{Doan and Halevy}{2005}]{doan05semantic}
Doan, A. and Halevy, A.Y. (2005) Semantic Integration Research in
the Database Community: A Brief Survey, {\it AI Magazine}, Vol. 26,
No. 1, pp.83--94.

\bibitem[\protect\citeauthoryear{Beynon-Davies et~al.}{1997}]{beynon97}
Beynon-Davies, P., Bonde, L., McPhee, D. and Jones, C.B. (1997) `A
collaborative schema integration system', {\it Comput. Supported
Coop. Work}, Vol. 6, No. 1, Norwell, MA, USA, pp.1--18, issn =
0925-9724.

\bibitem[\protect\citeauthoryear{Wang and Vergara-Niedermayr}{2008}]{wang08com}
Wang, F. and Vergara-Niedermayr, C. (2008) `Collaboratively Sharing
Scientific Data', {\it CollaborateCom}, pp.805--823.

\bibitem[\protect\citeauthoryear{Chin and Lansing}{2004}]{chin04context} Chin Jr., G. and Lansing,
C.S. (2008) `Capturing and Supporting Contexts for Scientific Data
Sharing via the Biological Sciences Collaboratory', {\it CSCW}, ISBN
1-58113-810-5.

\bibitem[\protect\citeauthoryear{W3C XML Query (XQuery)}{http://www.w3.org/XML/Query}]{xquery}
W3C XML Query (XQuery), http://www.w3.org/XML/Query

\bibitem[\protect\citeauthoryear{Wang et al.}{2009}]{wang09security}
Wang, F., Hussels, P. and Liu, P. (2009) `Securely and flexibly
sharing a biomedical data management system', {\it SPIE}.

\bibitem[\protect\citeauthoryear{Marlow et al.}{2006}]{marlow06tag}
Marlow, C., Naaman, M., Boyd, D. and Davis, M. (2006) `Position
Paper, Tagging, Taxonomy, Flickr, Article, ToRead', {\it
Collaborative Web Tagging Workshop}.

\bibitem[\protect\citeauthoryear{NCI Enterprise Vocabulary Services (EVS)}{http://ncicb.nci.nih.gov/NCICB /infrastructure /cacore\_overview
/vocabulary}]{ncievs} NCI Enterprise Vocabulary Services (EVS),
http://ncicb.nci.nih.gov/NCICB /infrastructure /cacore\_overview
/vocabulary

\bibitem[\protect\citeauthoryear{NCI Enterprise Vocabulary Services (EVS)}{http://ncicb.nci.nih.gov/NCICB /infrastructure /cacore\_overview
/vocabulary}]{ncithesaurus} NCI Enterprise Vocabulary Services
(EVS), http://ncicb.nci.nih.gov/NCICB /infrastructure
/cacore\_overview /vocabulary

\bibitem[\protect\citeauthoryear{Bafoutsou and Mentzas}{2002}]{bafoutsou2002281}
Bafoutsou, G. and Mentzas, G. (2002) `Review and functional
classification of collaborative systems', {\it International Journal
of Information Management}, Vol. 22, No. 4, pp.281--305.

\bibitem[\protect\citeauthoryear{Bos et al.}{2007}]{zimmerman07taxonomy}
Bos, N., Zimmerman, A., Olson, J., Yew, J., Yerkie, J. and Dahl, E.
et al. (2007) `From shared databases to communities of practice: A
taxonomy of collaboratories', {\it Journal of Computer-Mediated
Communication},  Vol. 12, No. 2.

\bibitem[\protect\citeauthoryear{Myneni and Patel}{2010}]{myneni10col}
Myneni, S. and  Patel, V.L. (2010) `Organization of biomedical data
for collaborative scientific research: A research information
management system', {\it International Journal of Information
Management}, Vol. 30, No. 3, June, pp.256--264.

\bibitem[\protect\citeauthoryear{myGrid}{http://www.mygrid.org.uk/}]{misc{mygrid}}
myGrid, http://www.mygrid.org.uk/

\bibitem[\protect\citeauthoryear{Pike and Gahegan}{2007}]{pike07}
Pike, W. and Gahegan, M. (2007) `Beyond ontologies: Toward situated
representations of scientific knowledge', {\it International Journal
of Human-Computer Studies}, Vol. 65, No. 7, pp.674--688.
\end{thebibliography}
\end{document}

### 2017-11-23 JournalPISODM2016/FinalVersionNov2017/main.tex
working on the final version

--- 2017-10-14 JournalPISODM2016/SubmittedV2/main.tex
@@
 
-%\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
+\newcommand{\mm}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
 %\newcommand{\imartin}[1]{\textcolor{blue}{\textbf{Martin: }#1}}
@@
 
-%\JOURNALNAME{\TEN{\it Int. J. System Control and Information
-%Processing,
-%Vol. \theVOL, No. \theISSUE, \thePUBYEAR\hfill\thepage}}%
-%
-%\def\BottomCatch{%
-%\vskip -10pt
-%\thispagestyle{empty}%
-%\begin{table}[b]%
-%\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcr@{}}%
-%\\[-12pt]
-%Copyright \copyright\ 2012 Inderscience Enterprises Ltd. & &%
-%\end{tabular*}%
-%\vskip -30pt%
-%%%\vskip -35pt%
-%\end{table}%
-%}
+\JOURNALNAME{\TEN{\it Int. J. Web and Grid Services,
+Vol. \theVOL, No. \theISSUE, \thePUBYEAR\hfill\thepage}}%
+
+\def\BottomCatch{%
+\vskip -10pt
+\thispagestyle{empty}%
+\begin{table}[b]%
+\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcr@{}}%
+\\[-12pt]
+Copyright \copyright\ 2017 Inderscience Enterprises Ltd. & &%
+\end{tabular*}%
+\vskip -30pt%
+%%\vskip -35pt%
+\end{table}%
+}
 \makeatother
@@
 
-\LRH{  }%K. Belhajjame et~al.}
-
-\RRH{\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements}
+\LRH{K. Belhajjame et~al.}
+
+\RRH{\pisodm: Building SOC Applications in the Presence of NFRs}
 
@@
 
-\PUBYEAR{2016}
+\PUBYEAR{2017}
 
@@
 
-%%
-%\authorA{Khalid Belhajjame}
-%%
-%\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
-%  E-mail: kbelhajj@googlemail.com}
-%%
-%%
-%\authorB{Valeria~de~Castro}
-%\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-%  E-mail: valeria.decastro@urjc.es}
-%\authorC{Umberto~Souza~da~Costa}
-%\affC{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-%  E-mail: umberto@dimap.ufrn.br}
-%
-%\authorD{Javier~A.~Espinosa-Oviedo}
-%\affD{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-%  E-mail: javiera.espinosa@gmail.com}
-%    
-%\authorE{Martin~A.~Musicante}
-%\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-%  E-mail: mam@dimap.ufrn.br}
-%
-%\authorF{Pl\'acido~A.~Souza~Neto}
-%\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
-%  E-mail: placido.neto@ifrn.edu.br}  
-%
-%\authorG{Genoveva~Vargas-Solar}
-%\affG{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
-%  E-mail: genoveva.vargas@imag.fr}  
-%
-%\authorH{Jos\'e-Luis~Zechinelli-Martini}
-%\affH{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
-%  E-mail: joseluis.zechinelli@udlap.mx}
-%
-%
-%%
-%%\authorA{Fusheng Wang\footnote{Work done while working at Siemens Corporate Research.} }
-%%
-%%\affA{Department of Biomedical Informatics, Emory University
-%%\newline
-%%36 Eagle Row, Ste 589, Atlanta, GA 30322, USA}
-%%
-%%
-%%
-%%\authorB{\footnotesize Cristobal Vergara-Niedermayr\footnote{Work done while working at Siemens Corporate Research.}}
-%%\affB{Oracle \newline
-% %New Jersey, USA}
-%%
-%
+%
+\authorA{\sf Khalid Belhajjame}
+\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
+  E-mail: kbelhajj@googlemail.com}
+
+\authorB{\sf Valeria~de~Castro}
+\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
+  E-mail: valeria.decastro@urjc.es}
+
+\authorC{\sf Javier~A.~Espinosa-Oviedo}
+\affC{Barcelona Supercomputing Centre (BSC), Spain and
+	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico\\
+  E-mail: javiera.espinosa@gmail.com}
+
+\authorD{\sf Martin~A.~Musicante}
+\affD{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+  E-mail: mam@dimap.ufrn.br}
+
+\authorE{\sf Umberto~Souza~da~Costa}
+\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+  E-mail: umberto@dimap.ufrn.br}
+
+\authorF{\sf Pl\'acido~A.~Souza~Neto}
+\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
+  E-mail: placido.neto@ifrn.edu.br}
+
+\authorG{\sf Genoveva~Vargas-Solar}
+\affG{Univ. Grenoble Alpes, CNRS, Institute of Engineering Univ. Grenoble Alpes (Grenoble INP), LIG, 38000 Grenoble, France
+	\\
+  E-mail: genoveva.vargas@imag.fr}
+
+\authorH{\sf Jos\'e-Luis~Zechinelli-Martini}
+\affH{Universidad de las Am\'ericas-Puebla -- Cholula, Mexico and
+	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico\\
+  E-mail: joseluis.zechinelli@udlap.mx}
 
@@
 The late inclusion of NFRs during the software development  may compromise the quality of the resulting application.
-This paper presents \pisodm \footnote{\hl{The letter $\pi$  of the name of the method comes from Policy which is the central notion to represent  NFR's and the way they are enforced at runtime.
-$\pi$ stands of the pronunciation of the first letter of the word "policy" which is /p/.}}, a method and associated tools that
+This paper presents \pisodm, a method and associated tools that
 \textit{(i)}  allow the early specification of NFRs allowing users to them in an abstract way without having to  care about low level details;
@@
 
-%\REF{to this paper should be made as follows: Rodr\'{\i}guez
-%Bol\'{\i}var, M.P. and Sen\'{e}s Garc\'{\i}a, B. (xxxx) `The
-%corporate environmental disclosures on the internet: the case of
-%IBEX 35 Spanish companies', {\it International Journal of Metadata,
-%Semantics and Ontologies}, Vol. x, No. x, pp.xxx\textendash xxx.}
-
-%\begin{bio}
-%Khalid Belhajjame\dots Bio here.
-%\vs{9}
-%
-%\noindent Valeria~de~Castro\dots Bio here.
-%\vs{9}
-%
-%\noindent Umberto Costa received the BS degree in Computer Science from Federal University of Rio Grande do Norte in 1998. 
-%He obtained the MS degree in Systems and Computation from Federal University of Rio Grande do Norte in 2000. 
-%He received the PhD degree in Computer Science from University Federal de Minas Gerais in 2005. He is currently an associate professor in the Department of 
-%Informatics and Applied Mathematics (DIMAp) at Federal University of Rio Grande do Norte. His academic interests include Programming Languages, 
-%Service-Oriented Computing and Cloud Computing.
-%\vs{9}
-%
-%\noindent Javier~A.~Espinosa-Oviedo\dots Bio here.
-%\vs{9}
-%
-%\noindent Martin A. Musicante received his B.Sc at ESLAI, Argentina, in 1988 and
-%his M.Sc and PhD in Computer Science at Universidade Federal de
-%Pernambuco, Brazil, in 1990 and 1996, respectively. He is an associate
-%professor at Universidade Federal do Rio Grande do Norte-UFRN (Natal,
-%Brazil). He is part of the graduate program in Computer Science at
-%UFRN and holder of a research grant from CNPq (level 2). Martin was
-%associate researcher at LI - Universit\'e Fran\c{c}ois Rabelais Tours
-%(2002-2011) and at LIFO - Universit\'e d'Orl\'eans, France, since 2008.
-%Martin A. Musicante's main current research topics are big data
-%processing languages, Web Services and cloud computing.
-%\vs{9}
-%
-%\noindent Pl\'acido~A.~Souza~Neto\dots Bio here.
-%\vs{9}
-%
-%\noindent Genoveva~Vargas-Solar\dots Bio here.
-%\vs{9}
-%
-%\noindent Jos\'e-Luis~Zechinelli-Martini\dots Bio here.
-%\end{bio}
+\REF{to this paper should be made as follows: Rodr\'{\i}guez
+Khalid Belhajjame, Valeria~de~Castro, Javier~A.~Espinosa-Oviedo, Martin~A.~Musicante, Umberto~Souza~da~Costa, Pl\'acido~A.~Souza~Neto, Genoveva~Vargas-Solar, Jos\'e-Luis~Zechinelli-Martini (xxxx) `\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements', {\it International Journal of Web and Grid Services}, Vol.
+x, No.
+x, pp.xxx\textendash xxx.}
+
+\begin{bio}
+Khalid Belhajjame is a lecturer (Maitre de Conférences) at the University Paris-Dauphine, where he is a member of the LAMSADE.
+His research interests lie in the areas of information and knowledge management.
+He has made key contributions to the areas of pay-as-you data integration, e-Science, scientific workflow management, provenance tracking and exploitation, and semantic web services.
+\vs{9}
+
+\noindent Valeria~de~Castro is associate professor in the Department of Computing Languages and Systems at the Rey Juan Carlos University sited in Madrid, Spain.
+She received her M.Sc.
+degree in Information Systems from the National Technological University, Argentine in 2003 and she got her Ph.D.
+in Computer Science by the Rey Juan Carlos University in 2007.
+She is member of Kybele Research Group where leads the Service Engineering research line.
+Her research interests include Services Engineering, Service Oriented Development, Web Engineering and Model Driven Engineering.
+She has co-authored several publications in national and international conferences and journals.
+She has participated in several research projects and has lead the main researcher at the Spanish Network of Service Science from 2010 to 2014.
+\vs{9}
+
+\noindent Javier Alfonso Espinosa-Oviedo is a computer scientist specialized in the domains of databases and distributed systems.
+His research focus mainly on large-scale data management and processing based on service oriented architectures.
+He currently works as postdoctoral researcher at the Barcelona Supercomputing Centre (BSC), in the Extreme Computing Group, and is researcher at the French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), in the Database Management Group.
+His research interests include: Stream Processing, Urban Computing, Digital Humanities, Data Visualization.
+He has participated in several national and European projects (CASES EU-FP7; S2EUNET FP7-IRSES; POLIWEB PEPS CNRS), and has been invited as visiting researcher in different institutions (POLITO, Italy; UdelaR, Uruguay; NUAA, China; UDLAP, Mexico) where he has lectured on selected topics of data management.
+\vs{9}
+
+\noindent Martin A.
+Musicante received his B.Sc at ESLAI, Argentina, in 1988 and his M.Sc and PhD in Computer Science at Universidade Federal de Pernambuco, Brazil, in 1990 and 1996, respectively.
+He is full professor at Universidade Federal do Rio Grande do Norte-UFRN (Natal, Brazil).
+He is part of the graduate program in Computer Science at UFRN.
+Martin was associate researcher at LI - Université François Rabelais Tours (2002-2011) and at LIFO - Université d'Orléans, France, since 2008.
+\vs{9}
+
+\noindent Umberto Souza da Costa received the BS degree in Computer Science from Federal University of Rio Grande do Norte in 1998.
+He obtained the MS degree in Systems and Computation from Federal University of Rio Grande do Norte in 2000.
+He received the PhD degree in Computer Science from University Federal de Minas Gerais in 2005.
+He is currently an associate professor in the Department of Informatics and Applied Mathematics (DIMAp) at Federal University of Rio Grande do Norte.
+His academic interests include Programming Languages, Service-Oriented Computing and Cloud Computing.
+\vs{9}
+
+\noindent Pl\'acido~A.~Souza~Neto received his B.Sc at CEFET-RN, Brazil, in 2005 and his M.Sc and PhD in Computer Science at Universidade Federal de Rio Grande do Norte, Brazil, in 2007 and 2012, respectively.
+He is an associate professor at Instituto Federal do Rio Grande do Norte-IFRN (Natal, Brazil).
+He was part of LIG (Grenoble - France) between 2014-2015 for as a post-doc fellow.
+He is part of the graduate program in Software Analysis and Development at IFRN.
+Plácido A.
+Souza Neto main current research topics are Software Engineering, Data Analysis, Model Driven Development and Web Services.
+\vs{9}
+
+\noindent Genoveva Vargas-Solar (http://www.vargas-solar.com) is senior scientist of the French Council of Scientific Research (CNRS) and member of the HADAS group of the Informatics Laboratory of Grenoble, France.
+She is deputy director the Franco-Mexican Laboratory of Informatics and Automatic Control (LAFMIA).
+She is regular member of the Mexican Academia of Computing Science.
+Her research interests in Computer Science concern distributed and heterogeneous databases, reflexive systems and service based database systems.
+Particularly, she contributes to the construction of service based database management systems.
+The purpose is to provide methodologies, algorithms and tools for integrating, deploying and executing a service composition for programing data management functions.
+Data services composition must respect quality of service criteria (e.g., security, reliability, fault tolerance, evolution and dynamic adaptability) and behaviour properties (e.g., transactional execution) adapted to applications requirements.
+She conducts fundamental and applied research activities for addressing these challenges.
+\vs{9}
+
+\noindent Jos\'e-Luis~Zechinelli-Martini is Associate professor of the Department of Computing, Electronics and Mecatronics at the Universidad de las Américas Puebla since 2002 and he is currently senior researcher at LAFMIA.
+He is head of the Data and Knowledge Management Group (DBKM) of LAFMIA.
+In 2001 he obtained his Ph.D.
+in Computer Science and in 1997 he obtained his Master Degree in Computer Science at University Joseph Fourier, Grenoble, France.
+In 1993 he obtained his diploma on Computer Systems Engineering at UDLAP.
+Between 2003 and 2006 he was director of the CENTIA y coordinator of the Ph.D program on Computer Science.
+His research project concerns on the integration of heterogeneous data services accessible in heterogeneous networks.
+\end{bio}
 
 \maketitle
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%
-%\documentclass{RITA}
-%\usepackage[T1]{fontenc}
-%%\usepackage[latin1]{inputenc}
-%\usepackage[english]{babel}
-%
-%% Our packages and definitions
-%\usepackage{xspace}
-%\usepackage{graphicx}
-%\usepackage{paralist} % for compact lists: compactitem, compactenum, compactdesc
-%\usepackage{lipsum}  % for the example environment
-%\usepackage{amssymb} % for the \blacksquare symbol
-%\usepackage{subfigure}
-%
-%\newcounter{examplecounter}
-%\newenvironment{example}{%\begin{quote}%
-%    \refstepcounter{examplecounter}%
-%  \textbf{Example \arabic{examplecounter}}%
-%  \quad
-%}{%
-%%\end{quote}%
-%$\blacksquare$  
-%}
-%
-%\let\OLDthebibliography\thebibliography
-%\renewcommand\thebibliography[1]{
-%  \OLDthebibliography{#1}
-%  \setlength{\parskip}{0pt}
-%  \setlength{\itemsep}{0pt plus 0.7ex}
-%}
-%
-%\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
-%\def\FlyingPig{\textsl{FlyingPig}\xspace}
-%
-%% RITA Preamble
-%
-%\title{\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements}
-%\author{
-%Khalid Belhajjame
-%\footnote{Universit\'e de Paris - Dauphine -- Paris, France\\
-%  \texttt{\{kbelhajj@googlemail.com\}}}
-%\and
-%Valeria~de~Castro
-%\footnote{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
-%  \texttt{\{valeria.decastro@urjc.es\}}}
-%\and
-%Umberto~Souza~da~Costa
-%\footnote{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-%  \texttt{\{\{umberto,mam\}@dimap.ufrn.br\}}}
-%\and
-%Javier~A.~Espinosa-Oviedo
-%\footnote{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\
-%  \texttt{\{javiera.espinosa@gmail.com, joseluis.zechinelli@udlap.mx\}}}  
-%\and
-%Martin~A.~Musicante
-%\footnotemark[3]
-%\and
-%Pl\'acido~A.~Souza~Neto
-%\footnote{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\
-%  \texttt{\{placido.neto@ifrn.edu.br\}}}  
-%\and
-%Genoveva~Vargas-Solar
-%\footnote{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\
-%  \texttt{\{genoveva.vargas@imag.fr\}}}  
-%\and  
-%Jos\'e-Luis~Zechinelli-Martini
-%\footnotemark[4]
-%}
-%\RITAvolume{22}
-%\RITAnumber{2}
-%\RITAyear{2015}
-%
-%\begin{document}
-%\maketitle
-%
-%%\begin{abstract}
-%%
-%%\end{abstract}
-%
-%\begin{abstractinenglish}
-%
-%\end{abstractinenglish}
-%
-%%\keywords{}
 
@@
 combined to build an application business logic.
-The selection of services is usually guided by the \textit{functional}\footnotemark \footnotetext{\hl{Functional properties of a computer system are characterized by the effects they produce in the behavior of the system given a specific input.}} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
+The selection of services is usually guided by the \textit{functional}\footnotemark \footnotetext{Functional properties of a computer system are characterized by the effects they produce in the behavior of the system given a specific input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
 An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.
@@
 These models are successively refined from abstract specifications into actual computer programs. 
-A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnotemark \footnotetext{\hl{The study shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
-Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}}.
-
-Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support the specification of non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
+A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnotemark \footnotetext{The study shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR.
+Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.
+
+Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support the specification of non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process\footnotemark \footnotetext{The letter $\pi$  of the name of the method comes from \textit{Policy} which is a central notion to represent  NFR's and the way they are enforced at runtime.
+$\pi$ stands of the pronunciation of the first letter of the word ``policy'' which is /p/.}
 Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
@@
 
-The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{\hl{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}}.
+The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.
 In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
@@
 
-\hl{
 To the best of our knowledge, and based on the related work analysis, there are no proposals that define a service-oriented approach for the whole development process of systems, considering
@@
 engineering, with emphasis on the functional aspect of the
-application.}
-
-Our method for developing service-oriented applications aims to overcome  limitations of existing  proposals by dealing with NFRs.  \hl{Our approach targets the characteristics of  3
+application.
+
+Our method for developing service-oriented applications aims to overcome  limitations of existing  proposals by dealing with NFRs.  
+Our approach targets the characteristics of  3
 areas\footnote{(i) Non-Functional Requirements; (ii) Methodology for Service-Oriented Development;
@@
 The application is related to the risk assessment industry.
-}
 
@@
 The e$^3$value model  identifies the value/in\-for\-ma\-tion  exchange between system components (see Figure~\ref{fig:CIM:tpme3v}).
-This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes)\footnotemark \footnotetext{\hl{According to}~\cite{Gordijn02valuebased}\hl{, a value activity is defined as an activity expected to be profitable to one or more actors.}} performed by business actors (squared boxes).
+This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes)\footnotemark \footnotetext{According to~\cite{Gordijn02valuebased}, a value activity is defined as an activity expected to be profitable to one or more actors.} performed by business actors (squared boxes).
 The model is well suited to enhance the understanding of the environment in which an application is being developed.
@@
 The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
-The e$^3$\hl{value model} shows Spotify and a private application (which is also a service) that provide services for listening and publishing information about music being listened by users. 
+The e$^3${value model} shows Spotify and a private application (which is also a service) that provide services for listening and publishing information about music being listened by users. 
 The private application interacts with Spotify for obtaining information about the flow of music being listened by a user in return of a fee for a premium subscription. 
@@
 
-BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow\footnotemark \footnotetext{\hl{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}}.
+BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow\footnotemark \footnotetext{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}.
 The next example illustrates the use of this meta-model.
@@
 Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}\footnotemark \footnotetext{\hl{We use the 
-%{\sc 
-small caps
-%} 
-font for referring to classes of a meta-model.}}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  \textsc{Business service}\footnotemark 
+\mm{Small caps font does not appear as such!}\footnotetext{We use the 
+{\sc small caps} font for referring to classes of a meta-model.}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
 An {\sc End Consumer} is represented by an {\sc Actor}.
@@
 \includegraphics[width=1\textwidth]{./figures/UseCaseMetaModel.pdf}
-\caption{\label{fig:CIM:usecasemetamodel} $\pi$-\hl{UseCase Meta-Model}.}
+\caption{\label{fig:CIM:usecasemetamodel} $\pi$-{UseCase Meta-Model}.}
 \end{figure*}
@@
 A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
-Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint})\footnotemark \footnotetext{\hl{We use the 
-%{\sf 
-sans serif
-%}
- font for referring to classes defined using a meta-model.}} use case.
-
-\hl{There are three kinds of constraints:}
+Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint})\footnotemark \footnotetext{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} use case.
+
+{There are three kinds of constraints:}
 \textit{(i)} A data pair ({\sc Value Constraint}), represented as the stereotype ``value'';
@@
 
-\hl{This model shows the set of logically related activities to be performed in a service-based application.
-The activities of this model represent a behavior that can  implement the business logic of the application.}
+This model shows the set of logically related activities to be performed in a service-based application.
+The activities of this model represent a behavior that can  implement the business logic of the application.
 A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract.
@@
 Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre- and post-conditions. 
-\hl{The service that fulfils the {\sc Contract} and its {\sc Assertions} will be chosen and used in the execution flow.}
+The service that fulfills the {\sc Contract} and its {\sc Assertions} will be chosen and used in the execution flow.
+This part of the process requires the verification of some logic conditions.
+In the present version of the tool, this step requires human intervention.
 
@@
 
-\item In order to represent constraint types associated to \hl{service compositions}, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
+\item In order to represent constraint types associated to {service compositions}, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
 %\martin{Placido: Mudar as cores... } 
@@
 
-\hl{These rules were implemented into a semi-automatic tool, the $\pi$-SOD-M  plug in for Eclipse. At each stage of transformation it is possible to include new constraints.} 
-\hl{Transformation in the more abstract stages is completely automatic  (i.e., from $\pi$-UseCase to $\pi$-ServiceProcess, and from $\pi$-ServiceProcess to $\pi$-ServiceComposition).
+These rules were implemented into a semi-automatic tool, the $\pi$-SOD-M  plug in for Eclipse. At each stage of transformation it is possible to include new constraints.
+Transformation in the more abstract stages is completely automatic  (i.e., from $\pi$-UseCase to $\pi$-ServiceProcess, and from $\pi$-ServiceProcess to $\pi$-ServiceComposition).
 In contrast, the transformation to specific platforms (i.e.
 from $\pi$-ServiceComposition to $\pi$-PEWS), the analyst must intervene and check the generation of the specification code because there are programming details that can be interpreted in different manners depending on the target platform.
-In this sense, the method is more appropriate for those who have background knowledge in Information Technology.}.
+In this sense, the method is more appropriate for those who have background knowledge in Information Technology.
 The transformations from CIM to PIM level models are not automatized, due to the informal nature of the CIM level.
@@
 \begin{compactitem}
-\item 	We  used the Eclipse Modeling Framework (EMF)\footnotemark \footnotetext{\hl{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
+\item 	We  used the Eclipse Modeling Framework (EMF)\footnotemark \footnotetext{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}  to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models.
 From these meta-models, we  developed plug-ins to support their graphical representation.
 
-\item	 We used  ATL\footnotemark \footnotetext{\hl{http://eclipse.org/atl/.
-An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}}
+\item	 We used  ATL\footnotemark \footnotetext{http://eclipse.org/atl/.
+An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
 for implementing the  mappings between models.
 
-\item 	\hl{We  used Acceleo}\footnotemark \footnotetext{\hl{http://www.acceleo.org/pages/home/en}} \hl{to implement  the code generation plug-in to generate executable code.} 
+\item 	We  used Acceleo\footnotemark \footnotetext{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code.
 It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
@@
 
-To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{\hl{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}}.%\martin{Is this a case study or a proof of concept?}
+To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.%\martin{Is this a case study or a proof of concept?}
 Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
@@
 The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
-Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnotemark \footnotetext{\hl{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}}.
+Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnotemark \footnotetext{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
+They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}.
 
@@
 The  PSM of our case study is given in Figure~\ref{fig:piPEWSFlyingPig}. This model is obtained by transforming the $\pi$-service composition model into a workflow.
-Note that this workflow is implemented by using  \hl{BPEL constructors}~\cite{ws-bpel2} and \textit{policy} notions.
+Note that this workflow is implemented by using  {BPEL constructors}~\cite{ws-bpel2} and \textit{policy} notions.
 
@@
 
-\hl{In the context of service-oriented applications, ensuring non-functional properties is challenging due to the nature of their components.
-Those components are defined by APIs which do not necessarily export information about their internals.}
+In the context of service-oriented applications, ensuring non-functional properties is challenging due to the nature of their components.
+Those components are defined by APIs which do not necessarily export information about their internals.
 
@@
 As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
-In contrast to these approaches, we specify policies for a \hl{service composition} in an orthogonal way.
-Besides, these approaches suppose that non-functional properties are implemented \hl{according to the knowledge} that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse.
+In contrast to these approaches, we specify policies for a {service composition} in an orthogonal way.
+Besides, these approaches suppose that non-functional properties are implemented {according to the knowledge} that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to \textit{ad-hoc} solutions that can be difficult to reuse.
 In our approach,  the policies defined for a given application  can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.
 
-\hl{Wrapping up, the industrial use case presented here  provided a real validation context for testing the usability of our method $\pi$SOD-M.
+Wrapping up, the industrial use case presented here  provided a real validation context for testing the usability of our method $\pi$SOD-M.
 The software units acting as services provide specific functions and constraints that were combined to build the business logic of the FlyingPig application.
@@
 Both two measures  are important to the company.
-}
 
@@
 
-\hl{In the future, we expect to:}
-\textit{(i)} \hl{select metrics for assessing the efficacy of} \pisodm;
-%
-\textit{(ii)} \hl{use the feedback provided by our industrial partner to improve the methodology;}
-%
-\textit{(iii)} \hl{assess the use of} \pisodm \hl{in other contexts, in order to investigate the generality of the approach. }
+In the future, we expect to:
+\textit{(i)} select metrics for assessing the efficacy of \pisodm;
+%
+\textit{(ii)} use the feedback provided by our industrial partner to improve the methodology;
+%
+\textit{(iii)} assess the use of \pisodm in other contexts, in order to investigate the generality of the approach.
 

### 2017-11-23 JournalPISODM2016/FinalVersionNov2017/main.tex
no message

@@
 \usepackage{mathrsfs}
+
+\usepackage[T1]{fontenc}
 
@@
 
-\newcommand{\mm}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
+%\newcommand{\mm}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
 %\newcommand{\imartin}[1]{\textcolor{blue}{\textbf{Martin: }#1}}
@@
 \\[-12pt]
-Copyright \copyright\ 2017 Inderscience Enterprises Ltd. & &%
+Copyright \textcopyright\ 2017 Inderscience Enterprises Ltd. & &%
 \end{tabular*}%
@@
 \authorC{\sf Javier~A.~Espinosa-Oviedo}
-\affC{Barcelona Supercomputing Centre (BSC), Spain and
-	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico\\
-  E-mail: javiera.espinosa@gmail.com}
+\affC{Barcelona Supercomputing Center (BSC), Spain and\\
+	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico/France\\
+  E-mail: javiera.espinosa@bsc.es}
 
@@
 \authorG{\sf Genoveva~Vargas-Solar}
-\affG{Univ. Grenoble Alpes, CNRS, Institute of Engineering Univ. Grenoble Alpes (Grenoble INP), LIG, 38000 Grenoble, France
+\affG{Univ. Grenoble Alpes, CNRS, Institute of Engineering Univ. Grenoble Alpes (Grenoble INP), LIG, 38000 Grenoble, France and
+	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico/France
 	\\
@@
 \affH{Universidad de las Am\'ericas-Puebla -- Cholula, Mexico and
-	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico\\
+	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico/France\\
   E-mail: joseluis.zechinelli@udlap.mx}
@@
 
-\REF{to this paper should be made as follows: Rodr\'{\i}guez
-Khalid Belhajjame, Valeria~de~Castro, Javier~A.~Espinosa-Oviedo, Martin~A.~Musicante, Umberto~Souza~da~Costa, Pl\'acido~A.~Souza~Neto, Genoveva~Vargas-Solar, Jos\'e-Luis~Zechinelli-Martini (xxxx) `\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements', {\it International Journal of Web and Grid Services}, Vol.
+\REF{to this paper should be made as follows: Khalid Belhajjame, Valeria~de~Castro, Javier~A.~Espinosa-Oviedo, Martin~A.~Musicante, Umberto~Souza~da~Costa, Pl\'acido~A.~Souza~Neto, Genoveva~Vargas-Solar, Jos\'e-Luis~Zechinelli-Martini (xxxx) `\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements', {\it International Journal of Web and Grid Services}, Vol.
 x, No.
@@
 \begin{bio}
-Khalid Belhajjame is a lecturer (Maitre de Conférences) at the University Paris-Dauphine, where he is a member of the LAMSADE.
+Khalid Belhajjame is a lecturer (Maitre de Conf\'erences) at the University Paris-Dauphine, where he is a member of the LAMSADE.
 His research interests lie in the areas of information and knowledge management.
@@
 \noindent Javier Alfonso Espinosa-Oviedo is a computer scientist specialized in the domains of databases and distributed systems.
-His research focus mainly on large-scale data management and processing based on service oriented architectures.
+His research focuses mainly on large-scale data management and processing based on service oriented architectures.
 He currently works as postdoctoral researcher at the Barcelona Supercomputing Centre (BSC), in the Extreme Computing Group, and is researcher at the French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), in the Database Management Group.
@@
 He is part of the graduate program in Computer Science at UFRN.
-Martin was associate researcher at LI - Université François Rabelais Tours (2002-2011) and at LIFO - Université d'Orléans, France, since 2008.
+Martin was associate researcher at LI - Universit\'e Fran\c{c}ois Rabelais Tours (2002-2011) and at LIFO - Universit\'e d'Orl\'eans, France, since 2008.
 \vs{9}
@@
 He is part of the graduate program in Software Analysis and Development at IFRN.
-Plácido A.
-Souza Neto main current research topics are Software Engineering, Data Analysis, Model Driven Development and Web Services.
+Pl\'acido~A.~Souza~Neto's main current research topics are Software Engineering, Data Analysis, Model Driven Development and Web Services.
 \vs{9}
@@
 
-\noindent Jos\'e-Luis~Zechinelli-Martini is Associate professor of the Department of Computing, Electronics and Mecatronics at the Universidad de las Américas Puebla since 2002 and he is currently senior researcher at LAFMIA.
+\noindent Jos\'e-Luis~Zechinelli-Martini is titular professor of the Department of Computing, Electronics and Mecatronics at the Universidad de las Américas Puebla (UDLAP) since 2002 and he is currently senior researcher at LAFMIA.
 He is head of the Data and Knowledge Management Group (DBKM) of LAFMIA.
@@
 In 1993 he obtained his diploma on Computer Systems Engineering at UDLAP.
-Between 2003 and 2006 he was director of the CENTIA y coordinator of the Ph.D program on Computer Science.
-His research project concerns on the integration of heterogeneous data services accessible in heterogeneous networks.
+Between 2003 and 2006 he was director of the CENTIA and head of the PhD program on Computer Science.
+His research project concerns on the integration of Big Data collections on different infrastructures and the specification of spatio-temporal and visualization query languages for retrieving multimedia and multiform data from distributed services.
 \end{bio}
@@
 
-\begin{compactenum}[i]
+\begin{compactenum}[i.]
 \item To improve the development process by providing an abstract view of the application and helping to ensure the conformance to its specification.
@@
 In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
-The models presented here were generated as a result of interacting with software developers at GCP Global and trying to fulfil  all the non-functional requirements present in the application.
+The models presented here were generated as a result of interacting with software developers at GCP Global and trying to fulfill  all the non-functional requirements present in the application.
 
@@
 
-Despite of the different notations found in the literature for classifying NFRs, some NFRs are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
+Despite the different notations found in the literature for classifying NFRs, some NFRs are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
 However, distinct hierarchies and models are proposed for NFRs, according to different perspectives.
@@
 Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.
-The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  \textsc{Business service}\footnotemark 
-\mm{Small caps font does not appear as such!}\footnotetext{We use the 
+The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  \textsc{Business service}\footnotemark \footnotetext{We use the 
 {\sc small caps} font for referring to classes of a meta-model.}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
@@
 \textit{(ii)} if the User is already logged in Spotify;
-\textit{(iii)} if bank account information are correct and;
+\textit{(iii)} if bank account information is correct and;
 \textit{(iv)} if there are enough funds in the bank account to cover the payment.
@@
 There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. 
-To update services, depending of each service, there may be different restrictions.
+To update services, depending on each service, there may be different restrictions.
 As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter.
@@
 
-\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service,  composed by  execu\-ta\-ble nodes.
+\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service,  composed of  execu\-ta\-ble nodes.
 
@@
 The relationships of type  {\sc Extend} and {\sc Include}  determine the way the business process is expressed as a workflow.  
-The generated workflow is composed by {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.
+The generated workflow is composed of {\sf Fork} and {\sf Join} nodes,  {\sc Control flow} constructors, as well as entities of type {\sc Action}.
 
@@
 The values exchanged between ORCA and the brokers are:
-\textit{(i)} \textsl{Questionnaire and Evidences} filled with information about the client company;
+\textit{(i)} \textsl{Questionnaire and Evidence} filled with information about the client company;
 \textit{(ii)} \textsl{Amendments} which are ORCA's rec\-om\-men\-da\-tions, based on the data provided by the answers to questionnaires;
@@
 Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnotemark \footnotetext{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
-They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}.
+They can be also immaterial restrictions such as the protocol used for accessing data on the company's computer server.}.
 
@@
 \item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
-\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
+\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidence} $\pi$-use case.
 \item All questionnaires need to be fully answered in order to consider a task as completed.
@@
 
+\section*{Acknowledgements}
+This research has been partially financed by the Government of the Autonomous Community of Madrid under the SICOMORo-CM project (S2013/ICE-3006), and by the ELASTIC project (TIN2014-52938-C2-1-R), financed by the Spanish Governments Ministry of Science and Innovation.
+This work is partially supported by INES, grant CNPq/465614/2014-0 (Brazil).
+
 \bibliographystyle{plain}

### 2018-01-18 JournalPISODM2016/ProofJan2018/latex-single-column/Appendix/sample (single column).tex
corrected proof

%%%%%%%%%%%%%%%%%%%%%%
\documentclass{singlecol-new}
%%%%%%%%%%%%%%%%%%%%%%

\usepackage{natbib,stfloats}
\usepackage{mathrsfs,amsmath,upgreek}

\def\newblock{\hskip .11em plus .33em minus .07em}

\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\newcommand{\nn}{\nonumber}

\theoremstyle{TH}{
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}[lemma]{Theorem}
\newtheorem{corrolary}[lemma]{Corrolary}
\newtheorem{conjecture}[lemma]{Conjecture}
\newtheorem{proposition}[lemma]{Proposition}
\newtheorem{claim}[lemma]{Claim}
\newtheorem{stheorem}[lemma]{Wrong Theorem}
\newtheorem{algorithm}{Algorithm}
}

\theoremstyle{THrm}{
\newtheorem{definition}{Definition}
\newtheorem{question}{Question}
\newtheorem{remark}{Remark}
\newtheorem{scheme}{Scheme}
}

\theoremstyle{THhit}{
\newtheorem{case}{Case}[section]
}

\makeatletter

\def\Reals{\mathbb{R}}
\def\Ints{\mathbb{Z}}
\def\Nats{\mathbb{N}}

\def\theequation{\arabic{equation}}

\def\tc{\textcolor{red}}

\def\BottomCatch{%
\vskip -10pt
\thispagestyle{empty}%
\begin{table}[b]%
\NINE\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcr@{}}%
\\[-12pt]
Copyright \copyright\ 2012 Inderscience Enterprises Ltd. & &%
\end{tabular*}%
\vskip -30pt%
%%\vskip -35pt%
\end{table}%
%\def\tc{\textcolor}{red}
} \makeatother

%%%%%%%%%%%%%%%%%
\begin{document}%
%%%%%%%%%%%%%%%%%

\thispagestyle{plain}

\setcounter{page}{1}

\LRH{xxxx}

\RRH{xxxx}

\VOL{x}

\ISSUE{x}

\PUBYEAR{xxxx}

%\BottomCatch

%\CLline

%\subtitle{}

\title{xxxx}

\authorA{xxxx}

\affA{xxxx*\\
*Corresponding author}

\authorB{xxxx}
\affB{xxxx}

\begin{abstract}
xxxxx
\end{abstract}

\KEYWORD{xxxx}

\REF{to this paper should be made as follows: xxxx (xxxx) `xxxx',
{\it xxxx}, Vol.~x, No.~x, pp.xxx--xxx.}

\begin{bio}
\tc{AUTHOR PLEASE SUPPLY CAREER HISTORY OF NO MORE THAN 100 WORDS
FOR EACH AUTHOR.}
\end{bio}

\maketitle

\section{Introduction}
\label{sec-introduction}

\section{Performance evaluation}
\label{sec-performance}

\section{Related work}
\label{sec-related}

\section{Conclusions}
\label{sec-conclusion}

\section*{Acknowledgements}

\begin{thebibliography}{99}

\bibitem[\protect\citeauthoryear{xxxx}{2011}]{Zhang11}
xxxx (2011) `Reliable packets delivery over segment-based multi-path
in wireless ad hoc networks', {\it Pervasive Computing and
Applications (ICPCA), 2011 6th International Conference on},
pp.300-–306. \tc{AUTHOR PLEASE SUPPLY LOCATION.}

\end{thebibliography}

\section*{Appendx}

We have proposed the CSD scheme that makes use of multiple bridge
nodes within the close neighbourhood of the source and insecure
neighbours. These bridge nodes cooperatively help to deliver secret
link key information from the source towards the insecure
neighbours. In the delivery process, no secret information is
disclosed to any node en route. These routers simply forward the
encrypted data. Even the bridge nodes only get to know partial
information about the secrets that they help to deliver.

\end{document}

%\def\notesname{Note}
%
%\theendnotes

%\section*{Query}
%
%\tc{AQ1: AUTHOR PLEASE CITE FIGURE 8 IN TEXT.}

\end{document}

### 2018-01-18 JournalPISODM2016/ProofJan2018/main.tex
corrected proof

--- 2017-11-23 JournalPISODM2016/FinalVersionNov2017/main.tex
@@
 \authorA{\sf Khalid Belhajjame}
-\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
+\affA{University Paris-Dauphine, PSL Research University, \\
+CNRS, UMR 7243, Centre Lamsade, 75016 Paris, France\\
   E-mail: kbelhajj@googlemail.com}
@@
 \authorB{\sf Valeria~de~Castro}
-\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
+\affB{Kybele Research Group, Rey Juan Carlos University, Madrid, Spain\\
   E-mail: valeria.decastro@urjc.es}
@@
 \authorC{\sf Javier~A.~Espinosa-Oviedo}
-\affC{Barcelona Supercomputing Center (BSC), Spain and\\
-	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico/France\\
+\affC{Barcelona Supercomputing Center (BSC), Barcelona, Spain
+ and French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Puebla, Mexico/ Grenoble, France\\
   E-mail: javiera.espinosa@bsc.es}
 
-\authorD{\sf Martin~A.~Musicante}
-\affD{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
-  E-mail: mam@dimap.ufrn.br}
-
-\authorE{\sf Umberto~Souza~da~Costa}
-\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+\authorD{\sf Martin A. Musicante and \\ Umberto Souza da Costa}
+\affD{Department of Computer Science (DIMAp), \\ 
+Federal University of Rio Grande do Norte, Natal, Brazil\\
+  E-mail: mam@dimap.ufrn.br\\
   E-mail: umberto@dimap.ufrn.br}
 
-\authorF{\sf Pl\'acido~A.~Souza~Neto}
-\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
-  E-mail: placido.neto@ifrn.edu.br}
-
-\authorG{\sf Genoveva~Vargas-Solar}
-\affG{Univ. Grenoble Alpes, CNRS, Institute of Engineering Univ. Grenoble Alpes (Grenoble INP), LIG, 38000 Grenoble, France and
-	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico/France
+%\authorE{\sf Umberto~Souza~da~Costa}
+%\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
+%  E-mail: umberto@dimap.ufrn.br}
+
+\authorE{\sf Pl\'acido~A.~Souza~Neto$^*$}
+\affE{Federal Institute for Education, Science and Technology \\
+of Rio Grande do Norte, Natal, Brazil\\
+  E-mail: placido.neto@ifrn.edu.br\\
+  *Corresponding author}
+
+\authorF{\sf Genoveva~Vargas-Solar}
+\affF{Univ. Grenoble Alpes, CNRS, Institute of Engineering Univ. Grenoble Alpes (Grenoble INP), LIG, 38000 Grenoble, France and
+	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Puebla, Mexico/ Grenoble, France
 	\\
@@
 
-\authorH{\sf Jos\'e-Luis~Zechinelli-Martini}
-\affH{Universidad de las Am\'ericas-Puebla -- Cholula, Mexico and
-	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Mexico/France\\
+\authorG{\sf Jos\'e-Luis~Zechinelli-Martini}
+\affG{University of the Americas-Puebla, Cholula, Mexico and
+	French-Mexican Laboratory of Informatics and Automatic Control (LAFMIA), Puebla, Mexico/ Grenoble, France\\
   E-mail: joseluis.zechinelli@udlap.mx}
@@
 His research project concerns on the integration of Big Data collections on different infrastructures and the specification of spatio-temporal and visualization query languages for retrieving multimedia and multiform data from distributed services.
+
+\bigskip
+
+\noindent This paper is a revised and expanded version of a paper entitled ``Supporting Non-functional Requirements in Services Software Development Process: An MDD Approach'' presented at SOFSEM 2014, Nov\'y Smokovec, High Tatras, Slovakia, January 25-30, 2014~\cite{Placido2014}.
 \end{bio}
